\subsubsection[The Westminster Confession of Faith of 1647]{{The Westminster Confession of
Faith}


\index{Faith:saving ``f10!}\section{Chapter 1 -- Of the Holy Scripture}
\index{Scripture:holy ``f10!}1.1\ \ Although the light of nature and the works of creation\index{God's
attributes:creator:works of ``f10!} and providence\index{Providence:providence ``f10!} do so far manifest
the\index{Providence:manifests goodness of God ``f10!} goodness\index{God's attributes:goodness ``f10!}, wisdom, and
power\index{God's attributes:almighty power ``f10!} of God, as to leave men unexcusable\index{Unbelievers:unexcusable
``f10!};\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg.
``f30!}; Romans 1:19-20\index{45 Romans:ch.   1, v.  19-20, pg. ``f30!}; Psalm 19:1-3\index{19 Psalms:ch.  19, v.  
1-3, pg. ``f30!}; Romans 1:32\index{45 Romans:ch.   1, v.  32, pg. ``f30!} with Romans 2:1.} yet are they not
sufficient to\index{Covenant:OT ordinances:sufficient for that time ``f10!} give that knowledge of
God\index{Revelation:natural:not sufficient knowledge of God ``f10!} and of His will, which is necessary unto
salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Corinthians 1:2\index{46 1 Corinthians:ch.  
1, v.   2, pg. ``f30!}1\index{46 1 Corinthians:ch.   1, v.  21, pg. ``f30!}; 1 Corinthians 2:13\index{46 1
Corinthians:ch.   2, v.  13, pg. ``f30!}{}-14\index{46 1 Corinthians:ch.   2, v.  13-14, pg. ``f30!}.} Therefore it
pleased\index{Providence:at God's pleasure ``f10!} the Lord, at sundry times, and in divers
manners\index{Revelation:sundry times and divers manners ``f10!}, to reveal Himself, and to declare
that\index{Revelation:special declaration ``f10!} His will unto His Church\index{Church:Christ's
``f10!};\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}.} and afterwards, for the better
preserving\index{Church:to preserve truth ``f10!} and propagating\index{Scripture:for propagating truth ``f10!} of the
truth, and\index{Oaths and Vows:assertion of truth ``f10!} for the more sure establishment\index{Scripture:for
establishment and comfort of Church ``f10!} and comfort of the Church against\index{Church:institution ``f10!} the
corruption of the\index{Sin:original:corrupt nature:flesh ``f10!} flesh, and the\index{flesh, sinful nature ``f10!}
malice of\index{Satan:malice of ``f10!} Satan and of\index{Satan:tempter ``f10!} the world, to commit the same wholly
unto writing:\index{Scripture:written ``f10!}\footnote{\ Proverbs 22:1\index{20 Proverbs:ch.  22, v.   1, pg.
``f30!}9-21\index{20 Proverbs:ch.  22, v.  19-21, pg. ``f30!}; Luke 1:3-4\index{42 Luke:ch.   1, v.   3-4, pg. ``f30!};
Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}; Matthew 4:4\index{40 Matthew:ch.   4, v.   4, pg. ``f30!}, 7,
10\index{40 Matthew:ch.   4, v.   4, 7, 10, pg. ``f30!}; Isaiah 8:19-20\index{23 Isaiah:ch.   8, v.  19-20, pg.
``f30!}.} which makes the Holy Scripture\index{Scripture:holy ``f10!} to be most necessary;\footnote{\ 2 Timothy
3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}; 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg. ``f30!}.}
those former ways of God's revealing His\index{Revelation:God's will ``f10!} will unto His people being now
ceased\index{Scripture:other forms of revelation have ceased ``f10!}.\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1,
v.   1, pg. ``f30!}{}-2\index{58 Hebrews:ch.   1, v.   1-2, pg. ``f30!}.}


\bigskip


1.2\ \ Under the name of Holy Scripture\index{Scripture:holy ``f10!}, or the Word of God\index{Scripture:word of
God:word of God ``f10!} written, are now contained all the books of the Old and New Testament\index{Scripture:New
Testament ``f10!}, which are these\index{Scripture:Old and New Testaments ``f10!}:


\bigskip

{\centering
\textit{Of the Old Testament}\index{Scripture:Old Testament ``f10!}\textit{:}
\par}

\begin{multicols}{3}
Genesis

Exodus

Leviticus

Numbers

Deuteronomy

Joshua

Judges

Ruth

1 Samuel

2 Samuel

1 Kings

2 Kings

1 Chronicles

2 Chronicles

Ezra

Nehemiah

Esther

Job

Psalms

Proverbs

Ecclesiastes

Song of Songs

Isaiah

Jeremiah

Lamentations

Ezekiel

Daniel

Hosea

Joel

Amos

Obadiah

Jonah

Micah

Nahum

Habakkuk

Zephaniah

Haggai

Zechariah

Malachi 
\end{multicols}

\bigskip

{\centering
\textit{Of the New Testament}\index{Scripture:New Testament ``f10!}\textit{:}
\par}
\begin{multicols}{3}
Matthew

Mark

Luke

John

Acts

Romans

1 Corinthians

2 Corinthians

Galatians

Ephesians

Philippians

Colossians

1 Thessalonians

2 Thessalonians

1 Timothy

2 Timothy

Titus

Philemon

Hebrews

James

1 Peter

2 Peter

1 John

2 John

3 John

Jude

Revelation
\end{multicols}

\bigskip

All which are given by inspiration of\index{Scripture:inspired by God ``f10!} God, to be the rule of
faith\index{Faith:rule of ``f10!} and life.\footnote{\ Luke 16:29, 31\index{42 Luke:ch.  16, v.  29, 31, pg. ``f30!};
Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!};
Revelation 22:18-19\index{66 Revelation:ch.  22, v.  18-19, pg. ``f30!}; 2 Timothy 3:16\index{55 2 Timothy:ch.   3, v. 
16, pg. ``f30!}.}


\bigskip

1.3\ \ The books commonly called Apocrypha\index{Apocrypha:not God breathed ``f10!}, not being of divine inspiration,
are no part of the canon\index{Scripture:canon ``f10!} of the Scripture; and\index{Scripture:Apocrypha not part of
canon ``f10!} therefore are of no authority in the Church of God\index{Church:God's ``f10!}, nor to be any otherwise
approved, or made use of, than other human writings\index{Apocrypha:human writings ``f10!}.\footnote{\ Luke 24:27,
44\index{42 Luke:ch.  24, v.  27, 44, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}; 2 Peter
1:2\index{61 2 Peter:ch.   1, v.   2, pg. ``f30!}1\index{61 2 Peter:ch.   1, v.  21, pg. ``f30!}.}


\bigskip

1.4\ \ The authority of the Holy Scripture\index{Scripture:holy ``f10!}, for which it ought to be believed
and\index{Scripture:to be:believed ``f10!} obeyed\index{Scripture:to be:obeyed:because word of God ``f10!}, depends not
upon the testimony\index{Scripture:authority not from testimony of any man ``f10!} of any man, or
Church\index{Church:institution ``f10!}; but wholly upon God (who is truth\index{God's attributes:truth ``f10!} itself)
the author thereof\index{Scripture:authored by God ``f10!}: and therefore it is to be received because it is the Word
of God\index{Scripture:word of God:word of God ``f10!}.\footnote{\ 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg.
``f30!}, 21\index{61 2 Peter:ch.   1, v.  19, 21, pg. ``f30!}; 2 Timothy 3:16\index{55 2 Timothy:ch.   3, v.  16, pg.
``f30!}; 1 John 5:9\index{62 1 John:ch.   5, v.   9, pg. ``f30!}; 1 Thessalonians 2:13\index{52 1 Thessalonians:ch.  
2, v.  13, pg. ``f30!}.}

1.5\ \ We may be moved and induced by the testimony\index{Scripture:testimony of the Church ``f10!} of the Church
to\index{Church:institution ``f10!} a high and reverent\index{Scripture:to be:treated with reverent esteem ``f10!}
esteem of the\index{Scripture:to be:esteemed ``f10!} Holy Scripture\index{Scripture:holy ``f10!}.\footnote{\ 1 Timothy
3:1\index{54 1 Timothy:ch.   3, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   3, v.  15, pg. ``f30!}.} And the
heavenliness\index{Scripture:heavenly matter ``f10!} of the matter, the efficacy\index{Scripture:efficacy of doctrine
``f10!} of the doctrine, the\index{Doctrine:of Scripture ``f10!} majesty of the\index{Scripture:majesty of the style
``f10!} style, the consent of all\index{Scripture:consent of all parts ``f10!} the parts, the scope of the whole (which
is, to give all glory\index{Scripture:gives glory to God ``f10!} to God), the full discovery\index{Scripture:makes full
discovery of the only way of salvation ``f10!} it makes of the only way of man's salvation\index{Salvation:belief in
Jesus by faith ``f10!}, the many other incomparable excellencies\index{Scripture:excellency ``f10!}, and the entire
perfection thereof\index{Scripture:perfection of ``f10!}, are arguments whereby it does abundantly evidence itself to
be the Word of God\index{Scripture:word of God:word of God ``f10!}: yet notwithstanding, our full
persuasion\index{Scripture:Holy Spirit persuades us that it is word of God ``f10!} and assurance of
the\index{Assurance:that God's word is true ``f10!} infallible truth\index{Scripture:infallible truth ``f10!} and
divine authority\index{Scripture:divine authority ``f10!} thereof, is from the inward\index{Scripture:witnessed
inwardly by Holy Spirit ``f10!} work of the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!} bearing witness
by\index{Scripture:witness to heart of man ``f10!} and with the Word in our hearts\index{Heart:new:through ministry of
Word ``f10!}.\footnote{\ 1 John 2:2\index{62 1 John:ch.   2, v.   2, pg. ``f30!}0, 27\index{62 1 John:ch.   2, v.  20,
27, pg. ``f30!}; John 16:13-14\index{43 John:ch.  16, v.  13-14, pg. ``f30!}; 1 Corinthians 2:10, 11-12\index{46 1
Corinthians:ch.   2, v.  10, 11-12, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}1\index{23
Isaiah:ch.  59, v.  21, pg. ``f30!}.} 


\bigskip

1.6\ \ The whole counsel of God\index{Causation:God's own will ``f10!} concerning all things necessary for His own
glory\index{God's attributes:glory:his own ``f10!}, man's salvation\index{Salvation:belief in Jesus by faith ``f10!},
faith, and life, is either expressly set\index{Scripture:interpretation:some things expressly set down ``f10!} down in
Scripture, or by good and necessary consequence\index{Scripture:interpretation:deduced by good and necessary
consequence ``f10!} may be deduced from Scripture: unto which nothing at any time is to be added, whether by new
revelations\index{Revelation:new, not added to Scripture ``f10!} of the Spirit, or\index{Revelation:Word and Spirit
``f10!} traditions\index{Scripture:supersedes traditions of men ``f10!} of men.\footnote{\ 2 Timothy 3:15\index{55 2
Timothy:ch.   3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  15-17, pg. ``f30!}; Galatians 1:8-9\index{48
Galatians:ch.   1, v.   8-9, pg. ``f30!}; 2 Thessalonians 2:2\index{53 2 Thessalonians:ch.   2, v.   2, pg. ``f30!}.}
Nevertheless we acknowledge the inward illumination\index{Conversion:Holy Spirit:provides inner illumination ``f10!} of
the Spirit of God\index{Holy Spirit:God's Spirit ``f10!} to be necessary for the saving
understanding\index{Salvation:by working of Holy Spirit ``f10!} of such things as are revealed in the
Word:\index{Scripture:word of God:standard for:life ``f10!}\footnote{\ John 6:45\index{43 John:ch.   6, v.  45, pg.
``f30!}; 1 Corinthians 2:9-12\index{46 1 Corinthians:ch.   2, v.   9-12, pg. ``f30!}.} and that there are some
circumstances\index{Worship:circumstances ``f10!} concerning the worship of God, and\index{Worship:correct:instituted
by God ``f10!} government of\index{Church government:government ``f10!} the Church, common\index{Church:institution
``f10!} to human actions and societies\index{Scripture:interpretation:circumstances of societies ``f10!}, which are to
be ordered\index{Scripture:interpretation:by light of nature according rules of Word ``f10!} by the light of nature and
Christian prudence, according\index{Scripture:interpretation:by prudence according rules of Word ``f10!} to the general
rules of the Word, which are always to be observed\index{Scripture:interpretation:general rules to be observed
``f10!}.\footnote{\ 1 Corinthians 11:13-14\index{46 1 Corinthians:ch.  11, v.  13-14, pg. ``f30!}; 1 Corinthians
14:26\index{46 1 Corinthians:ch.  14, v.  26, pg. ``f30!}, 40\index{46 1 Corinthians:ch.  14, v.  26, 40, pg. ``f30!}.}


\bigskip

1.7\ \ All things in Scripture are not alike plain in\index{Scripture:not all parts plain in themselves ``f10!}
themselves, nor alike clear unto all:\footnote{\ 2 Peter 3:16\index{61 2 Peter:ch.   3, v.  16, pg. ``f30!}.} yet those
things which are necessary to be known, believed, and\index{Scripture:to be:believed ``f10!} observed for
salvation\index{Salvation:belief in Jesus by faith ``f10!}, are so clearly propounded\index{Scripture:what necessary to
know, clearly propounded ``f10!} and opened in some place of Scripture or\index{Scripture:interpretation:scripture with
scripture ``f10!} other, that not only the learned\index{Scripture:useful for learned and unlearned ``f10!}, but the
unlearned, in a due use of the ordinary means, may\index{Scripture:understandable:through ordinary means ``f10!} attain
unto a sufficient\index{Scripture:understandable:by unlearned ``f10!} understanding of them.\footnote{\ Psalm
119:105\index{19 Psalms:ch. 119, v. 105, pg. ``f30!}, 130\index{19 Psalms:ch. 119, v. 105, 130, pg. ``f30!}.}


\bigskip

1.8\ \ The Old Testament\index{Scripture:Old Testament ``f10!} in Hebrew\index{Scripture:Old Testament Hebrew ``f10!}
(which was the native language of the people of God of old), and the New Testament\index{Scripture:New Testament
``f10!} in Greek\index{Scripture:New Testament Greek ``f10!} (which, at the time of the writing
of\index{Scripture:written ``f10!} it was most generally known to the nations\index{Scripture:written in languages
known to nations ``f10!}), being immediately inspired\index{Scripture:inspired by God ``f10!} by God, and, by His
singular care and providence\index{Providence:providence ``f10!} kept pure in\index{Scripture:kept pure in all ages
``f10!} all ages, are therefore authentical\index{Scripture:original language:versions authentic
``f10!};\footnote{\ Matthew 5:18\index{40 Matthew:ch.   5, v.  18, pg. ``f30!}.} so as, in all
controversies\index{Scripture:determine all controversies ``f10!} of religion, the Church is\index{Church:institution
``f10!} finally to appeal unto them.\footnote{\ Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg. ``f30!} ,Acts
15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}; John 5:39\index{43 John:ch.   5, v.  39, pg. ``f30!}, 46\index{43
John:ch.   5, v.  39, 46, pg. ``f30!}.} But, because these original tongues\index{Scripture:original language:not known
to all ``f10!} are not known to all the people of God, who have right unto, and interest in the\index{Scripture:all
people have interest (share) in ``f10!} Scriptures, and are commanded, in the\index{Scripture:to be:read and heard
``f10!} fear of\index{Fear:of God:reverence ``f10!} God, to read and search the\index{Heart:God alone able to search
``f10!}m\index{Scripture:to be:searched ``f10!},\footnote{\ John 5:39\index{43 John:ch.   5, v.  39, pg. ``f30!}.}
therefore they are to be translated\index{Scripture:to be:translated into common (vulgar) languages ``f10!} into the
vulgar language of every nation unto which they come,\footnote{\ 1 Corinthians 14:6, 9, 11-12, 24, 27-28\index{46 1
Corinthians:ch.  14, v.   6, 9, 11-12, 24, 27-28, pg. ``f30!}.} that the Word of God\index{Scripture:word of God:word
of God ``f10!} dwelling plentifully in all, they may worship Him\index{Worship:acceptable manner ``f10!} in an
acceptable manner;\footnote{\ Colossians 3:16\index{51 Colossians:ch.   3, v.  16, pg. ``f30!}.} and, through patience
and comfort of the Scriptures\index{Scripture:comforts and builds up believers ``f10!}, may have
hope\index{Scripture:provides hope ``f10!}.\footnote{\ Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}.}


\bigskip

1.9\ \ The infallible\index{Scripture:interpretation:scripture with scripture ``f10!} rule of interpretation of
Scripture is the Scripture itself: and therefore, when there is a question about the true and full sense of any
Scripture (which is not manifold, but\index{Scripture:interpretation:not manifold, but on sense ``f10!} one), it must
be searched\index{Scripture:to be:searched ``f10!} and known by other places that speak more clearly.\footnote{\ 2
Peter 1:2\index{61 2 Peter:ch.   1, v.   2, pg. ``f30!}0-21\index{61 2 Peter:ch.   1, v.  20-21, pg. ``f30!}; Acts
15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}{}-16\index{44 Acts:ch.  15, v.  15-16, pg. ``f30!}.}


\bigskip

1.10\ \ The supreme judge by\index{Scripture:supreme judge of controversies ``f10!} which all
controversies\index{Scripture:determine all controversies ``f10!} of religion are to be determined, and all
decrees\index{Church councils:decrees ``f10!} of councils\index{Church government:councils ``f10!}, opinions
of\index{Scripture:final authority, not opinions of ancient writers ``f10!} ancient writers,
doctrines\index{Doctrine:false:of men ``f10!} of men, and private spirits, are to be examined\index{Scripture:to
be:used to examine all doctrines ``f10!}; and in whose sentence we are to rest; can be no other but the Holy
Spirit\index{Holy Spirit:Holy Spirit ``f10!} speaking in the Scripture\index{Scripture:word of God:Holy Spirit speaking
``f10!}.\footnote{\ Matthew 22:29, 31\index{40 Matthew:ch.  22, v.  29, 31, pg. ``f30!}; Ephesians 2:2\index{49
Ephesians:ch.   2, v.   2, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!} with Acts 28:25\index{44
Acts:ch.  28, v.  25, pg. ``f30!}.}

\section{Chapter 2 -- Of God, and of the Holy Trinity}
\index{God's attributes:trinity:one substance ``f10!}2.1\ \ There is but one only,\footnote{\ Deuteronomy 6:4\index{05
Deuteronomy:ch.   6, v.   4, pg. ``f30!}; 1 Corinthians 8:4, 6\index{46 1 Corinthians:ch.   8, v.   4, 6, pg. ``f30!}.}
living, and true God:\footnote{\ 1 Thessalonians 1:9\index{52 1 Thessalonians:ch.   1, v.   9, pg. ``f30!}; Jeremiah
10:10\index{24 Jeremiah:ch.  10, v.  10, pg. ``f30!}.} who is infinite\index{God's attributes:infinite:knowledge
``f10!} in being and perfection\index{God's attributes:perfect blessedness ``f10!},\footnote{\ Job 11:7-9\index{18
Job:ch.  11, v.   7-9, pg. ``f30!}; Job 26:14\index{18 Job:ch.  26, v.  14, pg. ``f30!}.} a most pure\index{God's
attributes:a spirit, most pure ``f10!} spirit,\footnote{\ John 4:24\index{43 John:ch.   4, v.  24, pg. ``f30!}.}
invisible,\footnote{\ 1 Timothy 1:17\index{54 1 Timothy:ch.   1, v.  17, pg. ``f30!}.} without body,
parts,\footnote{\ Deuteronomy 4:15-16\index{05 Deuteronomy:ch.   4, v.  15-16, pg. ``f30!}; John 4:24\index{43 John:ch.
  4, v.  24, pg. ``f30!} with Luke 24:39\index{42 Luke:ch.  24, v.  39, pg. ``f30!}.} or passions\index{God's
attributes:no passions ``f10!},\footnote{\ Acts 14:11, 15\index{44 Acts:ch.  14, v.  11, 15, pg. ``f30!}.}
immutable\index{God's attributes:immutable ``f10!},\footnote{\ James 1:17\index{59 James:ch.   1, v.  17, pg. ``f30!};
Malachi 3:6\index{39 Malachi:ch.   3, v.   6, pg. ``f30!}.} immense\index{God's attributes:immense
``f10!},\footnote{\ 1 Kings 8:27\index{11 1 Kings:ch.   8, v.  27, pg. ``f30!}; Jeremiah 23:23-24\index{24 Jeremiah:ch.
 23, v.  23-24, pg. ``f30!}.} eternal,\index{God's attributes:eternal ``f10!}\footnote{\ Psalm 90:2\index{19 Psalms:ch.
 90, v.   2, pg. ``f30!}; 1 Timothy 1:17\index{54 1 Timothy:ch.   1, v.  17, pg. ``f30!}.} incomprehensible\index{God's
attributes:incomprehensible ``f10!},\footnote{\ Psalm 145:3\index{19 Psalms:ch. 145, v.   3, pg. ``f30!}.}
almighty\index{God's attributes:almighty power ``f10!},\footnote{\ Genesis 17:1\index{01 Genesis:ch.  17, v.   1, pg.
``f30!}; Revelation 4:8\index{66 Revelation:ch.   4, v.   8, pg. ``f30!}.} most wise\index{God's attributes:wisdom
``f10!},\footnote{\index{45 Romans:ch.  16, v.   2, pg.
``f30!} most
holy\index{God's attributes:holy:most ``f10!},\footnote{\index{23 Isaiah:ch.   6,
v.   3, pg. ``f30!}\index{66 Revelation:ch.   4, v.   8, pg.
``f30!}\index{19 Psalms:ch.
115, v.   3, pg. ``f30!} most absolute\index{God's attributes:absolute
``f10!},\footnote{\index{02 Exodus:ch.   3, v.  14, pg.
``f30!} things
according to the counsel of His\index{Causation:God's own will ``f10!} own immutable and most righteous
will\index{God's attributes:righteousness ``f10!},\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}.} for His own glory\index{God's attributes:glory:his own
``f10!};\footnote{\ Proverbs 16:4\index{20 Proverbs:ch.  16, v.   4, pg. ``f30!}; Romans 11:36\index{45 Romans:ch.  11,
v.  36, pg. ``f30!}.} most loving\index{God's attributes:love ``f10!},\footnote{\ 1 John 4:8, 16\index{62 1 John:ch.  
4, v.   8, 16, pg. ``f30!}.} gracious\index{Grace:God's free ``f10!}, merciful\index{God's attributes:merciful ``f10!},
long-suffering\index{God's attributes:long-suffering ``f10!}, abundant in goodness\index{God's attributes:goodness
``f10!} and truth\index{God's attributes:truth ``f10!}, forgiving iniquity\index{God's attributes:forgiving iniquity
``f10!}, transgression, and\index{God's law:commandments:transgressions ``f10!} sin;\footnote{\ Exodus 34:6\index{02
Exodus:ch.  34, v.   6, pg. ``f30!}{}-7\index{02 Exodus:ch.  34, v.   6-7, pg. ``f30!}.} the
rewarder\index{Believers:rewarded by God ``f10!} of them that diligently seek\index{Salvation:for those who diligently
seek God ``f10!} Him;\footnote{\ Hebrews 11:6\index{58 Hebrews:ch.  11, v.   6, pg. ``f30!}.} and withal, most just and
terrible\index{Judgement:just and terrible ``f10!} in His judgments\index{God's attributes:judge
``f10!},\footnote{\ Nehemiah 9:32-33\index{16 Nehemiah:ch.   9, v.  32-33, pg. ``f30!}.} hating\index{God's
attributes:hates sin ``f10!} all sin,\footnote{\ Psalm 5:5-6\index{19 Psalms:ch.   5, v.   5-6, pg. ``f30!}.} and who
will by no means clear the guilty\index{guilt ``f10!}.\footnote{\ Nahum 1:2-3\index{34 Nahum:ch.   1, v.   2-3, pg.
``f30!}; Exodus 34:7\index{02 Exodus:ch.  34, v.   7, pg. ``f30!}.} 


\bigskip

2.2\ \ God has all life\index{God's attributes:glory:his own ``f10!},\footnote{\ John 5:26\index{43 John:ch.   5, v. 
26, pg. ``f30!}.} glory,\footnote{\ Acts 7:2\index{44 Acts:ch.   7, v.   2, pg. ``f30!}.} goodness\index{God's
attributes:goodness ``f10!},\footnote{\ Psalm 119:68\index{19 Psalms:ch. 119, v.  68, pg. ``f30!}.}
blessedness,\index{God's attributes:perfect blessedness ``f10!}\footnote{\ 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.
  1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}; Romans 9:5\index{45 Romans:ch.   9, v.   5, pg.
``f30!}.} in and of Himself; and is alone in and unto Himself all-sufficient\index{God's attributes:all-sufficient
``f10!}, not standing in need of any creatures\index{creatures ``f10!} which He has made,\footnote{\ Acts
17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}{}-25\index{44 Acts:ch.  17, v.  24-25, pg. ``f30!}.} nor
deriving\index{God's attributes:glory:not derived from creatures ``f10!} any glory from them,\footnote{\ Job
22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}.} but only manifesting\index{God's attributes:glory:manifested
``f10!} His own glory in, by, unto, and upon them: He is the alone fountain\index{God's attributes:creator:fountain of
existence ``f10!} of all being, of whom, through whom, and to whom are all things;\footnote{\ Rom 11:36\index{45
Romans:ch.  11, v.  36, pg. ``f30!}.} and has most sovereign\index{God's attributes:sovereign lordship ``f10!} dominion
over the\index{Mankind:dominion over creation ``f10!}m\index{God's attributes:sovereign dominion ``f10!}, to do by
them, for them, or upon them whatsoever Himself pleases\index{Providence:at God's pleasure
``f10!}.\footnote{\ Revelation 4:11\index{66 Revelation:ch.   4, v.  11, pg. ``f30!}; 1 Timothy 6:1\index{54 1
Timothy:ch.   6, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}; Daniel 4:25, 35\index{27
Daniel:ch.   4, v.  25, 35, pg. ``f30!}.} In His sight all things are open and manifest;\footnote{\ Hebrews
4:13\index{58 Hebrews:ch.   4, v.  13, pg. ``f30!}.} His knowledge\index{God's attributes:omniscience ``f10!} is
infinite\index{God's attributes:infinite:knowledge ``f10!}, infallible, and\index{God's attributes:infallible
fore-knowledge ``f10!} independent\index{God's attributes:knowledge independent of creature ``f10!} upon the
creature,\footnote{\ Romans 11:33\index{45 Romans:ch.  11, v.  33, pg. ``f30!}{}-34\index{45 Romans:ch.  11, v.  33-34,
pg. ``f30!}; Psalm 147:5\index{19 Psalms:ch. 147, v.   5, pg. ``f30!}.} so as nothing is to Him
contingent\index{Causation:no contingency with God ``f10!}, or uncertain.\footnote{\ Acts 15:18\index{44 Acts:ch.  15,
v.  18, pg. ``f30!}; Ezekiel 11:5\index{26 Ezekiel:ch.  11, v.   5, pg. ``f30!}.} He is most holy\index{God's
attributes:holy:most ``f10!} in all His counsels\index{God's attributes:holy:counsels ``f10!}, in all His
works\index{Providence:God's works ``f10!}, and in all His commands\index{God's law:commandments:ten
``f10!}.\footnote{\ Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg. ``f30!}; Romans 7:12\index{45 Romans:ch.   7, v. 
12, pg. ``f30!}.} To Him is due from angels\index{Angels:angels ``f10!} and men, and every other creature, whatsoever
worship\index{Worship:of God ``f10!}, service, or obedience\index{God's law:commandments:obedience:required ``f10!} He
is pleased to require of them.\footnote{\ Revelation 5:12-14\index{66 Revelation:ch.   5, v.  12-14, pg. ``f30!}.} 


\bigskip

2.3\ \ In the unity\index{God's attributes:unity of persons ``f10!} of the Godhead\index{God's attributes:Godhead
``f10!} there be three persons\index{God's attributes:trinity:three persons ``f10!}, of one substance\index{God's
attributes:trinity:one substance ``f10!}, power, and eternity\index{God's attributes:eternal ``f10!}; God the
Father\index{Father:God the ``f10!}, God the Son\index{Jesus:Son of God ``f10!}, and God the Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!}.\footnote{\ 1 John 5:7\index{62 1 John:ch.   5, v.   7, pg. ``f30!}; Matthew 3:16-17\index{40
Matthew:ch.   3, v.  16-17, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 2 Corinthians
13:1\index{47 2 Corinthians:ch.  13, v.   1, pg. ``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}.} The
Father is of none, neither begotten\index{Holy Spirit:not begotten ``f10!}, nor proceeding: the\index{Father:God the,
no procession ``f10!} Son is eternally\index{Jesus:begotten eternally ``f10!} begotten of the\index{Jesus:begotten by
the Father ``f10!} Father:\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43 John:ch.   1, v. 
14, pg. ``f30!}, 18\index{43 John:ch.   1, v.  14, 18, pg. ``f30!}.} the Holy Ghost eternally proceeding
from\index{Holy Spirit:procession from Father and Son ``f10!} the Father and the Son.\footnote{\ John 15:26\index{43
John:ch.  15, v.  26, pg. ``f30!}; Galatians 4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.}


\bigskip



\section{Chapter 3 -- Of God's Eternal Decree}
\index{Predestination:eternal decrees ``f10!}3.1\ \ God from all eternity\index{God's attributes:eternal ``f10!} did, by
the most wise\index{God's attributes:wisdom ``f10!} and holy counsel of His\index{Causation:God's own will ``f10!} own
will, freely, and unchangeably ordain whatsoever\index{Predestination:all that happens ``f10!} comes to
pass:\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg.
``f30!}; Romans 11:33\index{45 Romans:ch.  11, v.  33, pg. ``f30!}; Hebrews 6:17\index{58 Hebrews:ch.   6, v.  17, pg.
``f30!}; Romans 9:15, 18\index{45 Romans:ch.   9, v.  15, 18, pg. ``f30!}.} yet so, as thereby neither is God the
author\index{Sin:God not author or approver of ``f10!} of sin,\footnote{\ James 1:13, 17\index{59 James:ch.   1, v. 
13, 17, pg. ``f30!}; 1 John 1:5\index{62 1 John:ch.   1, v.   5, pg. ``f30!}.} nor is violence\index{Causation:will of
creatures:not destroyed by predestination ``f10!} offered to the will of the creatures\index{creatures ``f10!}, nor is
the liberty or\index{Liberty:natural ``f10!} contingency\index{Causation:contingency ``f10!} of second
causes\index{Causation:second causes ``f10!} taken away, but rather established.\footnote{\ Acts 2:23\index{44 Acts:ch.
  2, v.  23, pg. ``f30!}; Matthew 17:12\index{40 Matthew:ch.  17, v.  12, pg. ``f30!}; Acts 4:27-28\index{44 Acts:ch.  
4, v.  27-28, pg. ``f30!}; John 19:11\index{43 John:ch.  19, v.  11, pg. ``f30!}; Proverbs 16:33\index{20 Proverbs:ch. 
16, v.  33, pg. ``f30!}.}


\bigskip

3.2\ \ Although God knows whatsoever\index{God's attributes:omniscience ``f10!} may or can come to pass upon all
supposed conditions\index{Predestination:not conditional ``f10!},\footnote{\ Acts 15:18\index{44 Acts:ch.  15, v.  18,
pg. ``f30!}; 1 Samuel 23:11-12\index{09 1 Samuel:ch.  23, v.  11-12, pg. ``f30!}; Matthew 11:21, 23\index{40
Matthew:ch.  11, v.  21, 23, pg. ``f30!}.} yet has He not decreed\index{Predestination:eternal decrees ``f10!} anything
because He foresaw\index{Predestination:not based on God's foresight:of events ``f10!} it as future, or as that which
would come to pass upon such conditions.\footnote{\ Romans 9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}, 13,
16\index{45 Romans:ch.   9, v.  11, 13, 16, pg. ``f30!}, 18\index{45 Romans:ch.   9, v.  11, 13, 16, 18, pg. ``f30!}.}


\bigskip

3.3\ \ By the decree\index{Predestination:eternal decrees ``f10!} of God, for the manifestation of His glory\index{God's
attributes:glory:his own ``f10!}, some men and angels\index{Angels:angels ``f10!}\footnote{\ 1 Timothy 5:21\index{54 1
Timothy:ch.   5, v.  21, pg. ``f30!}; Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}.} are
predestinated\index{Predestination:predestination ``f10!} unto everlasting life\index{everlasting life ``f10!}, and
others fore-ordained\index{Predestination:fore-ordination ``f10!} to everlasting death\index{Death:everlasting
``f10!}.\footnote{\ Romans 9:22, 23\index{45 Romans:ch.   9, v.  22, 23, pg. ``f30!}; Ephesians 1:5\index{49
Ephesians:ch.   1, v.   5, pg. ``f30!}{}-6\index{49 Ephesians:ch.   1, v.   5-6, pg. ``f30!}; Proverbs 16:4\index{20
Proverbs:ch.  16, v.   4, pg. ``f30!}.}


\bigskip

3.4\ \ These angels\index{Angels:angels ``f10!} and men, thus predestinated\index{Predestination:predestination ``f10!},
and fore-ordained\index{Predestination:fore-ordination ``f10!}, are particularly and unchangeably
designed\index{Angels:unchangeably designed ``f10!}, and their number so certain and definite\index{Angels:definite
number, no increase or decrease ``f10!}, that it cannot be either increased or diminished.\footnote{\ 2 Timothy
2:19\index{55 2 Timothy:ch.   2, v.  19, pg. ``f30!}; John 13:18\index{43 John:ch.  13, v.  18, pg. ``f30!}.}


\bigskip

3.5\ \ Those of mankind\index{Predestination:mankind:unto life ``f10!} that are
predestinated\index{Predestination:predestination ``f10!} unto life, God, before the
foundation\index{Predestination:before foundation of world. ``f10!} of the world was laid, according to His eternal and
immutable\index{God's attributes:immutable ``f10!} purpose, and the secret counsel and\index{Causation:God's own will
``f10!} good pleasure\index{Providence:at God's pleasure ``f10!} of His will, has chosen\index{Election:chosen by God
``f10!}, in Christ, unto everlasting glory,\index{Resurrection:believers:into glory:everlasting
``f10!}\footnote{\ Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}, 9\index{49 Ephesians:ch.   1, v.   4,
9, pg. ``f30!}, 11\index{49 Ephesians:ch.   1, v.   4, 9, 11, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3,
pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}; 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.   9, pg.
``f30!}; 1 Thessalonians 5:9\index{52 1 Thessalonians:ch.   5, v.   9, pg. ``f30!}.} out of His mere free
grace\index{Grace:God's free ``f10!} and love, without\index{God's attributes:love ``f10!} any
foresight\index{Predestination:not based on God's foresight:of faith ``f10!} of faith or good works\index{Good
works:adorn profession of faith ``f10!}, or perseverance\index{Good works:not foreseen by God as cause of election
``f10!} in either of them, or any other thing in the creature\index{creatures ``f10!}, as
conditions\index{Predestination:not conditional ``f10!}, or causes moving Him thereunto:\footnote{\ Romans
9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}, 13, 16\index{45 Romans:ch.   9, v.  11, 13, 16, pg. ``f30!};
Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}, 9\index{49 Ephesians:ch.   1, v.   4, 9, pg. ``f30!}.}
and all to the praise of His\index{Worship:praising God ``f10!} glorious grace\index{Grace:glorious
``f10!}.\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}, 12\index{49 Ephesians:ch.   1, v.  
6, 12, pg. ``f30!}.}


\bigskip

3.6.\ \ As God has appointed the elect\index{Election:elect of God ``f10!} unto glory\index{Believers:elected unto glory
``f10!}, so has He, by the eternal and most\index{Causation:God's own will ``f10!} free purpose of His will,
fore-ordained\index{Predestination:fore-ordination ``f10!} all the means thereunto.\footnote{\ 1 Peter 1:2\index{60 1
Peter:ch.   1, v.   2, pg. ``f30!}; Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}{}-5\index{49
Ephesians:ch.   1, v.   4-5, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}0\index{49
Ephesians:ch.   2, v.  10, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.}
Wherefore they who are elected\index{Election:chosen by God ``f10!}, being fallen in\index{Sin:original:first
transgression (of Adam) ``f10!} Adam, are\index{Adam:guilt of first sin on all mankind ``f10!}
redeemed\index{Redemption:purchased by Christ ``f10!} by Christ,\footnote{\ 1 Thessalonians 5:9\index{52 1
Thessalonians:ch.   5, v.   9, pg. ``f30!}{}-10\index{52 1 Thessalonians:ch.   5, v.   9-10, pg. ``f30!}; Titus
2:14\index{56 Titus:ch.   2, v.  14, pg. ``f30!}.} are effectually called\index{Effectual call:calling ``f10!} unto
faith\index{Faith:worked in believers by the Holy Spirit ``f10!} in Christ by His Spirit working in\index{Salvation:by
working of Holy Spirit ``f10!} due season, are justified\index{Justification:justified ``f10!},
adopted\index{Adoption:in salvation ``f10!}, sanctified\index{Sanctification:sanctification ``f10!},\footnote{\ Romans
8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}; Ephesians 1:5\index{49
Ephesians:ch.   1, v.   5, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.}
and kept by His power\index{Perseverance:kept by God's power ``f10!} through faith, unto
salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Peter 1:5\index{60 1 Peter:ch.   1, v.   5,
pg. ``f30!}.} Neither are any other redeemed by Christ, effectually called, justified, adopted, sanctified, and saved,
but\index{Salvation:only the elect ``f10!} the elect only.\footnote{\ John 17:9\index{43 John:ch.  17, v.   9, pg.
``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!} to the
end, John 6:64-65\index{43 John:ch.   6, v.  64-65, pg. ``f30!}; John 10:26\index{43 John:ch.  10, v.  26, pg. ``f30!};
John 8:47\index{43 John:ch.   8, v.  47, pg. ``f30!}; 1 John 2:19\index{62 1 John:ch.   2, v.  19, pg. ``f30!}.}


\bigskip

3.7\ \ The rest of mankind\index{Predestination:mankind:unto reprobation ``f10!} God was pleased\index{Providence:at
God's pleasure ``f10!}, according to the unsearchable\index{Providence:unsearchable wise counsel ``f10!} counsel of
His\index{Causation:God's own will ``f10!} own will, whereby He extends or withholds\index{Providence:God withholds
mercy, grace, favour as he wills ``f10!} mercy, as\index{God's attributes:merciful ``f10!} He pleases, for the glory of
His\index{God's attributes:glory:his own ``f10!} sovereign\index{God's attributes:sovereign lordship ``f10!} power over
His creatures\index{creatures ``f10!}, to pass by; and to ordain the\index{Predestination:Jesus as mediator
``f10!}m\index{Predestination:unto wrath ``f10!} to dishonour\index{Resurrection:wicked raised in dishonour ``f10!} and
wrath, for\index{Wrath:of God:for sin ``f10!} their sin, to the praise of His\index{Worship:praising God ``f10!}
glorious justice\index{God's attributes:administers justice ``f10!}.\footnote{\ Matthew 11:25-26\index{40 Matthew:ch. 
11, v.  25-26, pg. ``f30!}; Romans 9:17\index{45 Romans:ch.   9, v.  17, pg. ``f30!}, 18, 21-22\index{45 Romans:ch.  
9, v.  17, 18, 21-22, pg. ``f30!}; 2 Timothy 2:19\index{55 2 Timothy:ch.   2, v.  19, pg. ``f30!}{}-20\index{55 2
Timothy:ch.   2, v.  19-20, pg. ``f30!}; Jude 4\index{65 Jude:v.   4, pg. ``f30!}; 1 Peter 2:8\index{60 1 Peter:ch.  
2, v.   8, pg. ``f30!}.}


\bigskip

3.8\ \ The doctrine of this\index{Doctrine:of predestination ``f10!} high mystery\index{Predestination:mystery of
``f10!} of predestination\index{Predestination:predestination ``f10!} is to be handled with special prudence
and\index{Predestination:doctrine to be handled with prudence ``f10!} care,\footnote{{\ Romans
9:20}\index{45 Romans:ch.   9, v.  20, pg. ``f30!}\index{45 Romans:ch.  11, v. 
33, pg. ``f30!}\index{05 Deuteronomy:ch.  29, v.  29, pg.
``f30!} that men attending the will of God revealed in His Word, and\index{Scripture:word
of God:standard for:life ``f10!} yielding\index{God's law:commandments:obedience:required ``f10!} obedience thereunto,
may, from the certainty of their effectual vocation\index{Redemption:effectually:assured ``f10!}, be assured
of\index{Assurance:of election ``f10!} their eternal election.\index{Election:eternal
``f10!}\footnote{\index{61 2 Peter:ch.   1, v.  10, pg.
``f30!},
reverence, and admiration\index{Predestination:to elicit admiration of God ``f10!} of
God,\footnote{\index{49 Ephesians:ch.   1, v.   6, pg.
``f30!}\index{45 Romans:ch.  11, v.  33, pg.
``f30!}, diligence, and abundant
consolation to\index{Predestination:consolation for believers ``f10!} all that sincerely obey\index{Gospel:to be
sincerely obeyed ``f10!} the Gospel\index{Gospel:gospel ``f10!}.\footnote{\ Romans 11:5-6, 20\index{45 Romans:ch.  11,
v.   5-6, 20, pg. ``f30!}; 2 Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.  
8, v.   3, pg. ``f30!}3\index{45 Romans:ch.   8, v.  33, pg. ``f30!}; Luke 10:20\index{42 Luke:ch.  10, v.  20, pg.
``f30!}.}

\section{Chapter 4 -- Of Creation}
4.1\ \ It pleased\index{Providence:at God's pleasure ``f10!} God the Father\index{Father:God the ``f10!},
Son\index{Jesus:Son of God ``f10!}, and Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!},\footnote{\ Hebrews
1:2\index{58 Hebrews:ch.   1, v.   2, pg. ``f30!}; John 1:2-3\index{43 John:ch.   1, v.   2-3, pg. ``f30!}; Genesis
1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}; Job 26:13\index{18 Job:ch.  26, v.  13, pg. ``f30!}; Job
33:4\index{18 Job:ch.  33, v.   4, pg. ``f30!}.} for the manifestation of the glory of His\index{God's
attributes:glory:his own ``f10!} eternal power\index{God's attributes:eternal power ``f10!}, wisdom, and
goodness\index{God's attributes:goodness ``f10!},\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.  20, pg. ``f30!};
Jeremiah 10:12\index{24 Jeremiah:ch.  10, v.  12, pg. ``f30!}; Psalm 104:24\index{19 Psalms:ch. 104, v.  24, pg.
``f30!}; Psalm 33:5-6\index{19 Psalms:ch.  33, v.   5-6, pg. ``f30!}.} in the beginning, to create, or\index{God's
attributes:creator:from nothing ``f10!} make of nothing, the world, and all things therein whether visible
or\index{God's attributes:creator:of visible or invisible ``f10!} invisible, in\index{Creation:visible and invisible
``f10!} the space of six\index{Creation:in six days ``f10!} days; and all very good.\footnote{\ Genesis 1 chap. Hebrews
11:3\index{58 Hebrews:ch.  11, v.   3, pg. ``f30!}; Colossians 1:16\index{51 Colossians:ch.   1, v.  16, pg. ``f30!};
Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the
}\textit{\textcolor{black}{Statement Concerning Subscription
}}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Clarifications}}\textcolor{black}{).}}


\bigskip

4.2\ \ After God\index{Mankind:created:attributes ``f10!} had made all other creatures\index{creatures ``f10!}, He
created man\index{Mankind:created:by God ``f10!}, male\index{Mankind:male and female ``f10!} and
female\index{Mankind:created:male and female ``f10!},\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg.
``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}.} with reasonable\index{Mankind:created:reasonable soul ``f10!}
and immortal\index{Mankind:immortal soul ``f10!} souls,\footnote{\ Genesis 2:7\index{01 Genesis:ch.   2, v.   7, pg.
``f30!} with Ecclesiastes 12:7\index{21 Ecclesiastes:ch.  12, v.   7, pg. ``f30!} \& Luke 23:43\index{42 Luke:ch.  23,
v.  43, pg. ``f30!} and Matthew 10:28\index{40 Matthew:ch.  10, v.  28, pg. ``f30!}.} endued with
knowledge\index{Mankind:created:with knowledge, righteousness, and holiness ``f10!}, righteousness, and true
holiness\index{Mankind:created:holy ``f10!}, after His own image\index{Mankind:image of God ``f10!};\footnote{\ Genesis
1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}; Colossians
3:10\index{51 Colossians:ch.   3, v.  10, pg. ``f30!}; Ephesians 4:24\index{49 Ephesians:ch.   4, v.  24, pg. ``f30!}.}
having the law of God written in their hearts\index{God's law:written on man's heart ``f10!},\footnote{\ Romans
2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}.} and power to
fulfil it\index{Covenant:works:conditional on perfect obedience ``f10!}:\footnote{\ Ecclesiastes 7:2\index{21
Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.} and yet under a
possibility of transgressing\index{Mankind:created:with possibility of transgressing ``f10!}, being left to the liberty
of their\index{Liberty:of own will ``f10!} own will, which was subject unto change.\footnote{\ Genesis 3:6\index{01
Genesis:ch.   3, v.   6, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21
Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.} Beside this law written in their hearts, they received a command,
not\index{God's law:commandments:given to Adam ``f10!} to eat of the tree of the\index{Tree:of the knowledge of good
and evil ``f10!} knowledge of good and evil, which while they kept, they were happy\index{happiness ``f10!} in their
communion with God\index{Communion:with God ``f10!},\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg.
``f30!}; Genesis 3:8\index{01 Genesis:ch.   3, v.   8, pg. ``f30!}{}-11, 23\index{01 Genesis:ch.   3, v.   8-11, 23,
pg. ``f30!}.} and had dominion over the\index{Mankind:dominion over creation ``f10!} creatures.\footnote{\ Genesis
1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}, 28\index{01
Genesis:ch.   1, v.  26, 28, pg. ``f30!}.}

\section{Chapter 5 -- Of Providence}
\index{Providence:providence ``f10!}5.1\ \ God the great Creator\index{God's attributes:creator:of all things ``f10!} of
all things does uphold\index{Providence:upholds all creatures ``f10!},\footnote{\ Hebrews 1:3\index{58 Hebrews:ch.   1,
v.   3, pg. ``f30!}.} direct, dispose, and\index{Providence:God disposes all things for his glory ``f10!}
govern\index{Providence:governing all ``f10!} all creatures\index{creatures ``f10!}, actions, and
things,\footnote{\ Daniel 4:34-35\index{27 Daniel:ch.   4, v.  34-35, pg. ``f30!}; Psalm 135:6\index{19 Psalms:ch. 135,
v.   6, pg. ``f30!}; Acts 17:25\index{44 Acts:ch.  17, v.  25, pg. ``f30!}{}-26, 28\index{44 Acts:ch.  17, v.  25-26,
28, pg. ``f30!}; Job 38 to 41 chapters.} from the greatest even to the least,\footnote{\ Matthew 10:29, 30-31\index{40
Matthew:ch.  10, v.  29, 30-31, pg. ``f30!}.} by His most wise\index{God's attributes:wisdom ``f10!} and holy
providence\index{Providence:providence ``f10!},\footnote{\ Proverbs 15:3\index{20 Proverbs:ch.  15, v.   3, pg.
``f30!}; Psalm 104:24\index{19 Psalms:ch. 104, v.  24, pg. ``f30!}; Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg.
``f30!}.} according to His infallible\index{God's attributes:infallible fore-knowledge ``f10!}
fore-knowledge\index{God's attributes:foreknowledge ``f10!},\footnote{\ Acts 15:18\index{44 Acts:ch.  15, v.  18, pg.
``f30!}; Psalm 94:8, 11\index{19 Psalms:ch.  94, v.   8, 11, pg. ``f30!}.} and the free and immutable\index{God's
attributes:immutable ``f10!} counsel of His\index{Causation:God's own will ``f10!} own will,\footnote{\ Ephesians
1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}; Psalm
33:10-11\index{19 Psalms:ch.  33, v.  10-11, pg. ``f30!}.} to the praise of the\index{Worship:praising God ``f10!}
glory of His\index{God's attributes:glory:his own ``f10!} wisdom, power\index{God's attributes:listed (definition of
God) ``f10!}, justice, goodness\index{God's attributes:goodness ``f10!}, and mercy.\footnote{\ Isaiah 63:14\index{23
Isaiah:ch.  63, v.  14, pg. ``f30!}; Ephesians 3:10\index{49 Ephesians:ch.   3, v.  10, pg. ``f30!}; Romans
9:17\index{45 Romans:ch.   9, v.  17, pg. ``f30!}; Genesis 45:7\index{01 Genesis:ch.  45, v.   7, pg. ``f30!}; Psalm
145:7\index{19 Psalms:ch. 145, v.   7, pg. ``f30!}.}


\bigskip

5.2\ \ Although, in relation to the\index{Causation:relation to fore-knowledge ``f10!} fore-knowledge\index{God's
attributes:foreknowledge ``f10!} and decree\index{Predestination:eternal decrees ``f10!} of God, the first
Cause\index{Causation:first cause ``f10!}, all things come to pass immutably\index{God's attributes:immutable ``f10!},
and infallibly\index{Providence:infallible ``f10!}:\footnote{\ Acts 2:23\index{44 Acts:ch.   2, v.  23, pg. ``f30!}.}
yet, by the same providence\index{Providence:providence ``f10!}, He orders them to fall out, according to the nature of
second causes\index{Causation:second causes ``f10!}, either necessarily, freely, or\index{Causation:God's own will
``f10!} contingently\index{Causation:no contingency with God ``f10!}.\footnote{\ Genesis 8:22\index{01 Genesis:ch.   8,
v.  22, pg. ``f30!}; Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}5\index{24 Jeremiah:ch.  31, v.  35,
pg. ``f30!}; Exodus 21:13\index{02 Exodus:ch.  21, v.  13, pg. ``f30!} with Deuteronomy 19:5\index{05 Deuteronomy:ch. 
19, v.   5, pg. ``f30!}; 1 Kings 22:28, 34\index{11 1 Kings:ch.  22, v.  28, 34, pg. ``f30!}; Isaiah 10:6-7\index{23
Isaiah:ch.  10, v.   6-7, pg. ``f30!}.}


\bigskip

5.3\ \ God in His ordinary\index{Providence:ordinary means not binding on God ``f10!}
providence\index{Providence:providence ``f10!} makes use of means,\footnote{\ Acts 27:31, 44\index{44 Acts:ch.  27, v. 
31, 44, pg. ``f30!}; Isaiah 55:1\index{23 Isaiah:ch.  55, v.   1, pg. ``f30!}0-11\index{23 Isaiah:ch.  55, v.  10-11,
pg. ``f30!}; Hosea 2:21-22\index{28 Hosea:ch.   2, v.  21-22, pg. ``f30!}.} yet is free to work
without,\footnote{\ Hosea 1:7\index{28 Hosea:ch.   1, v.   7, pg. ``f30!}; Matthew 4:4\index{40 Matthew:ch.   4, v.  
4, pg. ``f30!}; Job 34:20\index{18 Job:ch.  34, v.  20, pg. ``f30!}.} above,\footnote{\ Romans 4:19-21\index{45
Romans:ch.   4, v.  19-21, pg. ``f30!}.} and against them at His pleasure\index{Providence:at God's pleasure
``f10!}.\footnote{\ 2 Kings 6:6\index{12 2 Kings:ch.   6, v.   6, pg. ``f30!}; Daniel 3:27\index{27 Daniel:ch.   3, v. 
27, pg. ``f30!}.}


\bigskip

5.4\ \ The almighty\index{God's attributes:almighty power ``f10!} power, unsearchable\index{Providence:unsearchable wise
counsel ``f10!} wisdom, and infinite\index{God's attributes:infinite:goodness ``f10!} goodness\index{God's
attributes:goodness ``f10!} of God so far manifest themselves in\index{Providence:manifests goodness of God ``f10!} His
providence\index{Providence:providence ``f10!}, that it extends itself\index{Providence:extends even to first sin
``f10!} even to the first fall\index{Sin:original:first transgression (of Adam) ``f10!}, and all other sins of
angels\index{Angels:angels ``f10!} and men;\footnote{\ Romans 11:32\index{45 Romans:ch.  11, v.  32, pg.
``f30!}{}-34\index{45 Romans:ch.  11, v.  32-34, pg. ``f30!}; 2 Samuel 24:1\index{10 2 Samuel:ch.  24, v.   1, pg.
``f30!} with 1 Chronicles 21:1\index{13 1 Chronicles:ch.  21, v.   1, pg. ``f30!}; 1 Kings 22:22-23\index{11 1
Kings:ch.  22, v.  22-23, pg. ``f30!}; 1 Chronicles 10:4, 13-14\index{13 1 Chronicles:ch.  10, v.   4, 13-14, pg.
``f30!}; 2 Samuel 16:10\index{10 2 Samuel:ch.  16, v.  10, pg. ``f30!}; Acts 2:23\index{44 Acts:ch.   2, v.  23, pg.
``f30!}; Acts 4:27-28\index{44 Acts:ch.   4, v.  27-28, pg. ``f30!}.} and that not by a bare permission\index{Sin:God
purposes and permits for his own glory ``f10!},\footnote{\ Acts 14:16\index{44 Acts:ch.  14, v.  16, pg. ``f30!}.} but
such as has joined with it a most wise\index{God's attributes:wisdom ``f10!} and powerful
bounding\index{Providence:governing all ``f10!},\footnote{\ Psalm 76:10\index{19 Psalms:ch.  76, v.  10, pg. ``f30!}; 2
Kings 19:28\index{12 2 Kings:ch.  19, v.  28, pg. ``f30!}.} and otherwise ordering and\index{Providence:ordering:all
things for his glory and believers' good ``f10!} governing of them, in a manifold dispensation\index{Providence:God's
dispensation ``f10!}, to His own holy\index{God's attributes:holy:ends ``f10!} ends;\footnote{\ Genesis 50:20\index{01
Genesis:ch.  50, v.  20, pg. ``f30!}; Isaiah 10:6-7\index{23 Isaiah:ch.  10, v.   6-7, pg. ``f30!}, 12\index{23
Isaiah:ch.  10, v.   6-7, 12, pg. ``f30!}.} yet so, as the sinfulness thereof proceeded only\index{Sin:proceeds only
from man, not God ``f10!} from the creature\index{creatures ``f10!}, and not from God, who, being most holy\index{God's
attributes:holy:most ``f10!} and righteous, neither\index{God's attributes:righteousness ``f10!} is, nor can be, the
author or approver\index{Sin:God not author or approver of ``f10!} of sin.\footnote{\ James 1:13-14, 17\index{59
James:ch.   1, v.  13-14, 17, pg. ``f30!}; 1 John 2:16\index{62 1 John:ch.   2, v.  16, pg. ``f30!}; Psalm
50:21\index{19 Psalms:ch.  50, v.  21, pg. ``f30!}.}

5.5\ \ The most wise\index{God's attributes:wisdom ``f10!}, righteous, and gracious\index{Grace:God's free ``f10!} God
does oftentimes leave for a season His\index{Temptation:left by God for a season ``f10!} own children to manifold
temptations\index{Temptation:temptation ``f10!}, and the corruption of the\index{Sin:original:corrupt nature:flesh
``f10!}ir\index{Sin:original:corrupt nature:hearts ``f10!} own hearts\index{Heart:corrupt ``f10!}, to
chastise\index{Believers:chastened by God ``f10!} them for their former sins, or to discover\index{Temptation:exposes
(discovers) corruption ``f10!} unto them the hidden strength of\index{Sin:original:hidden strength of corruption
``f10!} corruption, and\index{Sin:original:corrupt nature:hidden ``f10!} deceitfulness\index{Deceitfulness:of heart
``f10!} of their hearts, that\index{Heart:deceitful ``f10!} they may be humbled\index{Humility:before God
``f10!};\footnote{\ 2 Chronicles 32:25-26, 31\index{14 2 Chronicles:ch.  32, v.  25-26, 31, pg. ``f30!}; 2 Samuel
24:1\index{10 2 Samuel:ch.  24, v.   1, pg. ``f30!}.} and, to raise them to a more close and constant
dependence\index{Temptation:increase dependence on God ``f10!} for their support upon Himself, and to make them more
watchful against\index{Temptation:watchfulness required to avoid ``f10!} all future occasions of sin, and for sundry
other just and holy\index{God's attributes:holy:nature ``f10!} ends.\footnote{\ 2 Corinthians 12:7-9\index{47 2
Corinthians:ch.  12, v.   7-9, pg. ``f30!}; Psalm 73 throughout, Psalm 77:1-12\index{19 Psalms:ch.  77, v.   1-12, pg.
``f30!}; Mark 14:66 to the end, with John 21:15-17\index{43 John:ch.  21, v.  15-17, pg. ``f30!}.}


\bigskip

5.6\ \ As for those wicked\index{blindness, spiritual ``f10!} and ungodly\index{Unbelievers:ungodly ``f10!} men whom
God, as a righteous Judge, for\index{God's attributes:judge ``f10!} former sins, does blind and
harden\index{Heart:hardened ``f10!},\footnote{\ Romans 1:24, 26, 28\index{45 Romans:ch.   1, v.  24, 26, 28, pg.
``f30!}; Romans 11:7\index{45 Romans:ch.  11, v.   7, pg. ``f30!}{}-8\index{45 Romans:ch.  11, v.   7-8, pg. ``f30!}.}
from them He not only withholds\index{Providence:God withholds mercy, grace, favour as he wills ``f10!} His grace,
whereby they might have been enlightened\index{Providence:God withholds enlightening from wicked ``f10!} in their
understandings\index{Unbelievers:understandings darkened ``f10!}, and wrought upon\index{Salvation:by working of Holy
Spirit ``f10!} in their hearts;\index{Heart:new:enlightened ``f10!}\footnote{\ Deuteronomy 29:4\index{05
Deuteronomy:ch.  29, v.   4, pg. ``f30!}.} but sometimes also withdraws\index{Punishment:God withdraws gifts ``f10!}
the gifts which\index{Punishment:gifts from God withdrawn ``f10!} they had,\footnote{\ Matthew 13:12\index{40
Matthew:ch.  13, v.  12, pg. ``f30!}; Matthew 25:29\index{40 Matthew:ch.  25, v.  29, pg. ``f30!}.} and
exposes\index{Punishment:exposure to consequences of sin ``f10!} them to such objects as their corruption
makes\index{Sin:original:corrupt nature:lustful ``f10!} occasions of sin;\footnote{\ Deuteronomy 2:30\index{05
Deuteronomy:ch.   2, v.  30, pg. ``f30!}; 2 Kings 8:12-13\index{12 2 Kings:ch.   8, v.  12-13, pg. ``f30!}.} and,
withal, gives them over to their own lusts\index{Sexual immorality:lust ``f10!}, the
temptations\index{Temptation:temptation ``f10!} of the world, and the power of Satan\index{Satan:power of
``f10!}:\footnote{\ Psalm 81:11\index{19 Psalms:ch.  81, v.  11, pg. ``f30!}{}-12\index{19 Psalms:ch.  81, v.  11-12,
pg. ``f30!}; 2 Thessalonians 2:10\index{53 2 Thessalonians:ch.   2, v.  10, pg. ``f30!}{}-12\index{53 2
Thessalonians:ch.   2, v.  10-12, pg. ``f30!}.} whereby it comes to pass that they harden themselves, even under those
means which God uses for the softening\index{Providence:hardens some, softens others ``f10!} of
others.\footnote{\ Exodus 7:3\index{02 Exodus:ch.   7, v.   3, pg. ``f30!} with Exodus 8:15, 32\index{02 Exodus:ch.  
8, v.  15, 32, pg. ``f30!}; 2 Corinthians 2:15-16\index{47 2 Corinthians:ch.   2, v.  15-16, pg. ``f30!}; Isaiah
8:14\index{23 Isaiah:ch.   8, v.  14, pg. ``f30!}; 1 Peter 2:7-8\index{60 1 Peter:ch.   2, v.   7-8, pg. ``f30!};
Isaiah 6:9-10\index{23 Isaiah:ch.   6, v.   9-10, pg. ``f30!} with Acts 28:26-27\index{44 Acts:ch.  28, v.  26-27, pg.
``f30!}.}


\bigskip

5.7\ \ As the providence\index{Providence:providence ``f10!} of God does in general reach to all
creatures\index{creatures ``f10!}, so after a most special manner, it takes care of His Church\index{Church:Christ's
``f10!}, and disposes\index{Providence:God disposes all for good of his Church ``f10!} all things to the good
thereof.\footnote{\ 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   4, v.  10,
pg. ``f30!}; Amos 9:8-9\index{30 Amos:ch.   9, v.   8-9, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg.
``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!}; Isaiah 43:3-5, 14\index{23 Isaiah:ch.  43, v.   3-5, 14, pg.
``f30!}.}

\paragraph[Chapter 6 {}-- Of the Fall of Man, of Sin, and of the Punishment thereof]{Chapter 6 -- Of the Fall of Man, of
Sin, and of the Punishment thereof}
\index{Sin:original:first transgression (of Adam) ``f10!}\index{Punishment:of sin ``f10!}6.1\ \ Our first
parents\index{Parents:first, sinned against God ``f10!}, being seduced\index{Temptation:Satan's seduction of our first
parents ``f10!} by the subtilty\index{Satan:tempter ``f10!} and temptation\index{Temptation:temptation ``f10!} of
Satan, sinned, in eating the forbidden\index{Sin:original:first transgression (of Adam) ``f10!}
fruit.\footnote{\ Genesis 3:13\index{01 Genesis:ch.   3, v.  13, pg. ``f30!}; 2 Corinthians 11:3\index{47 2
Corinthians:ch.  11, v.   3, pg. ``f30!}.} This their sin God was pleased\index{Providence:at God's pleasure ``f10!},
according to His wise\index{God's attributes:wisdom ``f10!} and holy counsel, to\index{Causation:God's own will ``f10!}
permit\index{Sin:God purposes and permits for his own glory ``f10!}, having purposed to order it to His own
glory\index{God's attributes:glory:his own ``f10!}.\footnote{\ Romans 11:32\index{45 Romans:ch.  11, v.  32, pg.
``f30!}.}


\bigskip

6.2\ \ By this sin they fell\index{Sin:original:all mankind descending from Adam ``f10!} from their original
righteousness and\index{Mankind:created:original righteousness ``f10!} communion, with\index{Communion:with God ``f10!}
God,\footnote{\ Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}{}-8\index{01 Genesis:ch.   3, v.   6-8, pg.
``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29,
pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}3\index{45 Romans:ch.   3, v.  23, pg. ``f30!}.}
and so became dead\index{Death:spiritual, in sin ``f10!} in sin,\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v. 
17, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}.} and wholly defiled in
all\index{Sin:original:corrupt nature:defiled in all parts and faculties ``f10!} the parts and faculties of
soul\index{Sin:original:corrupt nature:faculties of the soul ``f10!} and body.\footnote{\ Titus 1:15\index{56 Titus:ch.
  1, v.  15, pg. ``f30!}; Genesis 6:5\index{01 Genesis:ch.   6, v.   5, pg. ``f30!}; Jeremiah 17:9\index{24
Jeremiah:ch.  17, v.   9, pg. ``f30!}; Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}{}-19\index{45
Romans:ch.   3, v.  10-19, pg. ``f30!}.}

6.3\ \ They being the root\index{Sin:original:first transgression (of Adam) ``f10!} of all mankind,
the\index{Sin:original:all mankind descending from Adam ``f10!} guilt\index{guilt ``f10!} of this sin was
imputed,\index{Imputation:of Adam's sin ``f10!}\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg.
``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}{}-28\index{01 Genesis:ch.   1, v.  27-28, pg. ``f30!} and
Genesis 2:16-17\index{01 Genesis:ch.   2, v.  16-17, pg. ``f30!} and Acts 17:26\index{44 Acts:ch.  17, v.  26, pg.
``f30!} with Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!},
15-19\index{45 Romans:ch.   5, v.  12, 15-19, pg. ``f30!} and 1 Corinthians 15:21-22\index{46 1 Corinthians:ch.  15, v.
 21-22, pg. ``f30!}, 49\index{46 1 Corinthians:ch.  15, v.  21-22, 49, pg. ``f30!}.} and the same
death\index{Death:spiritual, in sin ``f10!} in sin and corrupted\index{Sin:original:corrupt nature:conveyed from Adam
``f10!} nature conveyed, to all their posterity descending from them by ordinary\index{Sin:original:by ordinary
generation ``f10!} generation.\footnote{\ Psalm 51:5\index{19 Psalms:ch.  51, v.   5, pg. ``f30!}; Genesis 5:3\index{01
Genesis:ch.   5, v.   3, pg. ``f30!}; Job 14:4\index{18 Job:ch.  14, v.   4, pg. ``f30!}; Job 15:14\index{18 Job:ch. 
15, v.  14, pg. ``f30!}.}


\bigskip

6.4\ \ From this original corruption, whereby\index{Sin:original:corrupt nature:whole ``f10!} we are utterly
indisposed\index{Sin:original:utterly indisposed ``f10!}, disabled\index{Sin:original:utterly unable and unwilling to
do will of God ``f10!}, and made opposite to all good,\footnote{\ Romans 5:6\index{45 Romans:ch.   5, v.   6, pg.
``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7, pg. ``f30!}; Romans 7:18\index{45 Romans:ch.   7, v.  18, pg.
``f30!}; Colossians 1:21\index{51 Colossians:ch.   1, v.  21, pg. ``f30!}.} and wholly
inclined\index{Sin:original:inclined to do evil ``f10!} to all evil,\footnote{\ Genesis 6:5\index{01 Genesis:ch.   6,
v.   5, pg. ``f30!}; Genesis 8:21\index{01 Genesis:ch.   8, v.  21, pg. ``f30!}; Romans 3:10\index{45 Romans:ch.   3,
v.  10, pg. ``f30!}{}-12\index{45 Romans:ch.   3, v.  10-12, pg. ``f30!}.} do proceed all\index{Sin:original:proceeds
all transgressions ``f10!} actual transgressions\index{God's law:commandments:transgressions ``f10!}.\footnote{\ James
1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}{}-15\index{59 James:ch.   1, v.  14-15, pg. ``f30!}; Ephesians
2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   2, v.   2-3, pg. ``f30!}; Matthew
15:19\index{40 Matthew:ch.  15, v.  19, pg. ``f30!}.}


\bigskip

6.5\ \ This corruption\index{Sanctification:remnants of corruption remain ``f10!} of nature, during this life, does
remain in those that are regenerated\index{Regeneration:regeneration ``f10!};\footnote{\ 1 John 1:8, 10\index{62 1
John:ch.   1, v.   8, 10, pg. ``f30!}; Romans 7:14\index{45 Romans:ch.   7, v.  14, pg. ``f30!}, 17-18, 23\index{45
Romans:ch.   7, v.  14, 17-18, 23, pg. ``f30!}; James 3:2\index{59 James:ch.   3, v.   2, pg. ``f30!}; Proverbs
20:9\index{20 Proverbs:ch.  20, v.   9, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg.
``f30!}0\index{21 Ecclesiastes:ch.   7, v.  20, pg. ``f30!}.} and although it be, through Christ, pardoned
and\index{Regeneration:pardoned of sin ``f10!} mortified\index{Sin:mortified ``f10!}, yet both itself and all the
motions thereof are truly\index{Sin:original:corrupt nature:whole ``f10!} and properly sin.\footnote{\ Romans 7:5, 7-8,
25\index{45 Romans:ch.   7, v.   5, 7-8, 25, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg.
``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}.}


\bigskip

6.6\ \ Every sin, both original and actual, being a transgression of\index{God's law:commandments:transgressions ``f10!}
the righteous law\index{God's law:righteous requirements ``f10!} of God, and contrary thereunto,\footnote{\ 1 John
3:4\index{62 1 John:ch.   3, v.   4, pg. ``f30!}.} does, in its own nature, bring guilt\index{guilt ``f10!} upon the
sinner;\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}5\index{45 Romans:ch.   2, v.  15, pg.
``f30!}; Romans 3:9\index{45 Romans:ch.   3, v.   9, pg. ``f30!}, 19\index{45 Romans:ch.   3, v.   9, 19, pg. ``f30!}.}
whereby he is bound over to the wrath of God\index{Wrath:of God:for sin ``f10!},\footnote{\ Ephesians 2:3\index{49
Ephesians:ch.   2, v.   3, pg. ``f30!}.} and curse of the\index{Curse:of the law ``f10!} law,\footnote{\ Galatians
3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}.} and so made subject to death\index{Death:punishment for sin
``f10!},\footnote{\ Romans 6:23\index{45 Romans:ch.   6, v.  23, pg. ``f30!}.} with all
miseries\index{Punishment:misery ``f10!} spiritual,\footnote{\ Ephesians 4:18\index{49 Ephesians:ch.   4, v.  18, pg.
``f30!}.} temporal,\index{Judgement:temporal ``f10!}\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg.
``f30!}0\index{45 Romans:ch.   8, v.  20, pg. ``f30!}; Lamentations 3:39\index{25 Lamentations:ch.   3, v.  39, pg.
``f30!}.} and eternal.\footnote{\ Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}; 2 Thessalonians
1:9\index{53 2 Thessalonians:ch.   1, v.   9, pg. ``f30!}.}

\section{Chapter 7 -- Of God's Covenant with Man}
\index{Covenant:of God with man ``f10!}7.1\ \ The distance\index{Mankind:great distance between him and God ``f10!}
between God and the creature\index{creatures ``f10!} is so great, that although
reasonable\index{Mankind:created:reasonable soul ``f10!} creatures do owe obedience\index{God's
law:commandments:obedience:required ``f10!} unto Him as their Creator\index{God's attributes:creator:of all things
``f10!}, yet they could never have any fruition of Him as their blessedness and\index{Believers:blessed in/by God
``f10!} reward\index{Believers:rewarded by God ``f10!}, but by some voluntary condescension\index{God's
attributes:condescension, voluntary ``f10!} on God's part, which He has been pleased\index{Providence:at God's pleasure
``f10!} to express by way of covenant\index{Covenant:administrations ``f10!}.\footnote{\ Isaiah 40:13-17\index{23
Isaiah:ch.  40, v.  13-17, pg. ``f30!}; Job 9:32-33\index{18 Job:ch.   9, v.  32-33, pg. ``f30!}; 1 Samuel
2:25\index{09 1 Samuel:ch.   2, v.  25, pg. ``f30!}; Psalm 113:5-6\index{19 Psalms:ch. 113, v.   5-6, pg. ``f30!};
Psalm 100:2\index{19 Psalms:ch. 100, v.   2, pg. ``f30!}{}-3\index{19 Psalms:ch. 100, v.   2-3, pg. ``f30!}; Job
22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}; Job 35:7-8\index{18 Job:ch.  35, v.   7-8, pg. ``f30!}; Luke
17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}; Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}{}-25\index{44
Acts:ch.  17, v.  24-25, pg. ``f30!}.}


\bigskip


7.2\ \ The first covenant made\index{Covenant:first:made with man ``f10!} with man was a covenant of
works\index{Covenant:works:works ``f10!},\footnote{\ Galatians 3:12\index{48 Galatians:ch.   3, v.  12, pg. ``f30!}.}
wherein life was promised to\index{Covenant:first:promised life ``f10!} Adam, and in him to\index{Adam:covenant made
with him ``f10!} his posterity\index{Sin:original:all mankind descending from Adam ``f10!},\footnote{\ Romans
5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}{}-20\index{45 Romans:ch.
  5, v.  12-20, pg. ``f30!}; Romans 10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}.} upon
condition\index{Covenant:works:conditional on perfect obedience ``f10!} of perfect and personal
obedience\index{Covenant:works:perfect obedience required ``f10!}.\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.
 17, pg. ``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}.}


\bigskip

7.3\ \ Man by his fall\index{Sin:original:first transgression (of Adam) ``f10!} having made himself
incapable\index{Covenant:works:Adam:incapable of life after his sin ``f10!} of life by that covenant,
the\index{Covenant:first:broken ``f10!} Lord was pleased\index{Providence:at God's pleasure ``f10!} to make a
second\index{Covenant:second ``f10!},\footnote{\ Galatians 3:21\index{48 Galatians:ch.   3, v.  21, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}{}-21\index{45 Romans:ch.
  3, v.  20-21, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}; Genesis 3:15\index{01 Genesis:ch.
  3, v.  15, pg. ``f30!}; Isaiah 42:6\index{23 Isaiah:ch.  42, v.   6, pg. ``f30!}.} commonly called the covenant of
grace\index{Covenant:grace:grace ``f10!}; wherein He freely\index{Salvation:freely offered ``f10!} offers unto sinners
life and salvation\index{Salvation:belief in Jesus by faith ``f10!} by Jesus Christ, requiring\index{Salvation:faith
alone required ``f10!} of them faith\index{Faith:in Jesus Christ ``f10!} in Him, that they may be
saved,\footnote{\ Mark 16:15\index{41 Mark:ch.  16, v.  15, pg. ``f30!}{}-16\index{41 Mark:ch.  16, v.  15-16, pg.
``f30!}; John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}; Romans 10:6, 9\index{45 Romans:ch.  10, v.   6, 9, pg.
``f30!}; Galatians 3:11\index{48 Galatians:ch.   3, v.  11, pg. ``f30!}.} and promising to\index{Holy Spirit:promising
to believers ``f10!} give unto all those that are ordained unto\index{Predestination:unto life ``f10!} life His Holy
Spirit\index{Holy Spirit:Holy Spirit ``f10!}, to make them willing and able to\index{Conversion:renewal of the will
``f10!} believe.\footnote{\ Ezekiel 36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36,
v.  26-27, pg. ``f30!}; John 6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}{}-45\index{43 John:ch.   6, v.  44-45, pg.
``f30!}.}


\bigskip

7.4\ \ This covenant of grace\index{Covenant:grace:grace ``f10!} is frequently set forth in Scripture by the name of a
Testament\index{Covenant:testament ``f10!}, in reference to the death\index{Jesus:death:death ``f10!} of Jesus Christ
the Testator\index{Covenant:grace:Jesus testator ``f10!}, and to the everlasting
inheritance\index{Believers:inheritance:everlasting ``f10!}, with all things belonging to it, therein
bequeathed.\footnote{\ Hebrews 9:15\index{58 Hebrews:ch.   9, v.  15, pg. ``f30!}{}-17\index{58 Hebrews:ch.   9, v. 
15-17, pg. ``f30!}; Hebrews 7:22\index{58 Hebrews:ch.   7, v.  22, pg. ``f30!}; Luke 22:20\index{42 Luke:ch.  22, v. 
20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch. 
11, v.  25, pg. ``f30!}.}


\bigskip

7.5\ \ This covenant was\index{Covenant:administrations ``f10!} differently administered in the time of the law, and in
the time of the gospel\index{Gospel:gospel ``f10!}:\footnote{\ 2 Corinthians 3:6\index{47 2 Corinthians:ch.   3, v.  
6, pg. ``f30!}{}-9\index{47 2 Corinthians:ch.   3, v.   6-9, pg. ``f30!}.} under the law, it was administered by
promises\index{Covenant:promises ``f10!}, prophecies\index{Covenant:prophecies ``f10!}, sacrifices\index{Ceremonial
law:sacrifices ``f10!}, circumcision\index{circumcision ``f10!}, the paschal\index{paschal lamb ``f10!}
lamb\index{lamb, paschal ``f10!}, and other types\index{types, OT ``f10!} and ordinances delivered\index{Covenant:OT
ordinances:signified Christ ``f10!} to the people of the Jews, all\index{Jews:OT ordinances ``f10!}
fore-signifying\index{Ceremonial law:fore-signifying Christ ``f10!} Christ to come:\footnote{\ Hebrews 8-10\index{58
Hebrews:ch. 8-10, pg. ``f30!} chapters, Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Colossians
2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg. ``f30!}; 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7,
pg. ``f30!}.} which were, for that time, sufficient and\index{Covenant:OT ordinances:sufficient for that time ``f10!}
efficacious\index{Covenant:OT ordinances:efficacious ``f10!}, through the operation\index{Holy Spirit:operations:saving
``f10!} of the Spirit, to\index{Holy Spirit:instructs the elect ``f10!} instruct and build up the\index{Holy
Spirit:builds up the elect ``f10!} elect\index{Election:elect of God ``f10!} in faith\index{Faith:in Jesus Christ, as
Messiah ``f10!} in the promised\index{Jesus:Messiah promised ``f10!} Messiah,\footnote{\ 1 Corinthians 10:1-4\index{46
1 Corinthians:ch.  10, v.   1-4, pg. ``f30!}; Hebrews 11:13\index{58 Hebrews:ch.  11, v.  13, pg. ``f30!}; John
8:56\index{43 John:ch.   8, v.  56, pg. ``f30!}.} by whom they had full remission\index{Redemption:remission of sins
``f10!} of sins, and eternal salvation\index{Salvation:belief in Jesus by faith ``f10!}; and is called, the Old
Testament\index{Scripture:Old Testament ``f10!}.\footnote{\ Galatians 3:7-9, 14\index{48 Galatians:ch.   3, v.   7-9,
14, pg. ``f30!}.}


\bigskip

7.6\ \ Under the gospel\index{Gospel:gospel ``f10!}, when Christ, the substance,\footnote{\ Colossians 2:17\index{51
Colossians:ch.   2, v.  17, pg. ``f30!}.} was exhibited, the\index{Covenant:grace:exhibited ``f10!} ordinances
in\index{Church:furnished with:ordinances of God ``f10!} which this covenant is dispensed are\index{Covenant:new:ways
dispensed ``f10!} the preaching of the Word\index{Preaching:ministry of the Word ``f10!}, and the
administration\index{Church:true:sacraments administered ``f10!} of the sacraments\index{Sacraments:holy:ordinances
``f10!} of Baptism and\index{Baptism:sacrament ``f10!} the Lord's Supper\index{Lord's Supper:sacrament
``f10!}:\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v. 
19-20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch. 
11, v.  23, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  23-25, pg. ``f30!}.} which, though fewer in number,
and administered with\index{Sacraments:to be:administered with simplicity ``f10!} more simplicity, and less
outward\index{Sacraments:NT less outward glory than OT ``f10!} glory; yet, in them, it is held forth in more
fulness\index{Covenant:new:more fulness ``f10!}, evidence, and spiritual efficacy,\footnote{\ Hebrews 12:2\index{58
Hebrews:ch.  12, v.   2, pg. ``f30!}2\index{58 Hebrews:ch.  12, v.  22, pg. ``f30!} to 28; Jeremiah 31:3\index{24
Jeremiah:ch.  31, v.   3, pg. ``f30!}3\index{24 Jeremiah:ch.  31, v.  33, pg. ``f30!}{}-34\index{24 Jeremiah:ch.  31,
v.  33-34, pg. ``f30!}.} to all nations\index{Gospel:to all nations ``f10!}, both Jews and\index{Jews:called ``f10!}
Gentiles\index{Gentiles:called ``f10!};\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!};
Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}5-19\index{49 Ephesians:ch.   2, v.  15-19, pg. ``f30!}.}
and is called the New Testament\index{Scripture:New Testament ``f10!}.\footnote{\ Luke 22:20\index{42 Luke:ch.  22, v. 
20, pg. ``f30!}.} There are not therefore two covenants of\index{Covenant:grace:one only ``f10!} grace, differing in
substance\index{Covenant:grace:one, but differ in substance ``f10!}, but one and the same, under various
dispensations\index{Covenant:dispensations ``f10!}.\footnote{\ Galatians 3:14, 16\index{48 Galatians:ch.   3, v.  14,
16, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}1-23, 30\index{45 Romans:ch.   3, v.  21-23,
30, pg. ``f30!}; Psalm 32:1\index{19 Psalms:ch.  32, v.   1, pg. ``f30!} with Romans 4:3, 6, 16-17, 23-24\index{45
Romans:ch.   4, v.   3, 6, 16-17, 23-24, pg. ``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}; Acts
15:11\index{44 Acts:ch.  15, v.  11, pg. ``f30!}.}

\section{Chapter 8 -- Of Christ the Mediator}
\index{Jesus:mediator:of covenant ``f10!}8.1\ \ It pleased\index{Providence:at God's pleasure ``f10!} God, in His
eternal purpose\index{Predestination:eternal purpose ``f10!}, to choose and ordain the\index{Predestination:Jesus as
mediator ``f10!} Lord Jesus, His only begotten\index{Jesus:begotten by the Father ``f10!} Son\index{Jesus:Son of God
``f10!}, to be the Mediator\index{Jesus:mediator:of covenant ``f10!} between God and man;\footnote{\ Isaiah
42:1\index{23 Isaiah:ch.  42, v.   1, pg. ``f30!}; 1 Peter 1:19-20\index{60 1 Peter:ch.   1, v.  19-20, pg. ``f30!};
John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.}
the Prophet\index{Jesus:prophet:office ``f10!},\footnote{\ Acts 3:22\index{44 Acts:ch.   3, v.  22, pg. ``f30!}.}
Priest\index{Jesus:priest ``f10!},\footnote{\ Hebrews 5:5-6\index{58 Hebrews:ch.   5, v.   5-6, pg. ``f30!}.} and
King\index{Jesus:King ``f10!},\footnote{\ Psalm 2:6\index{19 Psalms:ch.   2, v.   6, pg. ``f30!}; Luke 1:33\index{42
Luke:ch.   1, v.  33, pg. ``f30!}.} the Head\index{Church:Christ as head ``f10!} and Saviour\index{Jesus:saviour of the
Church ``f10!} of His Church\index{Church:Christ's ``f10!},\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.  
2, pg. ``f30!}3\index{49 Ephesians:ch.   5, v.  23, pg. ``f30!}.} the Heir\index{Jesus:heir of all things ``f10!} of
all things,\footnote{\ Hebrews 1:2\index{58 Hebrews:ch.   1, v.   2, pg. ``f30!}.} and Judge of\index{Jesus:judge:of
world ``f10!} the world:\footnote{\ Acts 17:31\index{44 Acts:ch.  17, v.  31, pg. ``f30!}.} unto whom He did from all
eternity\index{God's attributes:eternal ``f10!} give a people, to be His seed\index{Election:God's seed
``f10!},\footnote{\ John 17:6\index{43 John:ch.  17, v.   6, pg. ``f30!}; Psalm 22:30\index{19 Psalms:ch.  22, v.  30,
pg. ``f30!}; Isaiah 53:10\index{23 Isaiah:ch.  53, v.  10, pg. ``f30!}.} and to be by Him in time
redeemed\index{Redemption:purchased by Christ ``f10!}, called, justified\index{Justification:justified ``f10!},
sanctified\index{Sanctification:sanctification ``f10!}, and glorified\index{Resurrection:believers:into
glory:glorification ``f10!}.\footnote{\ 1 Timothy 2:6\index{54 1 Timothy:ch.   2, v.   6, pg. ``f30!}; Isaiah
55:4-5\index{23 Isaiah:ch.  55, v.   4-5, pg. ``f30!}; 1 Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg.
``f30!}.}


\bigskip

8.2\ \ The Son\index{Jesus:Son of God ``f10!} of God, the second person in\index{Jesus:second person in Trinity ``f10!}
the Trinity\index{God's attributes:trinity:one substance ``f10!}, being very and eternal God\index{God's
attributes:eternal ``f10!}, of one substance and equal with the\index{God's attributes:trinity:persons equal ``f10!}
Father\index{Father:God the ``f10!}, did, when the fulness of time\index{Jesus:incarnation:in fulness of time ``f10!}
was come, take upon Him man's nature,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}, 14\index{43
John:ch.   1, v.   1, 14, pg. ``f30!}; 1 John 5:20\index{62 1 John:ch.   5, v.  20, pg. ``f30!}; Philippians
2:6\index{50 Philippians:ch.   2, v.   6, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.}
with all the essential properties and\index{Jesus:human nature:essential properties ``f10!} common infirmities
thereof\index{Jesus:humiliation:suffered:infirmities ``f10!}, yet without sin:\footnote{\ Hebrews 2:1\index{58
Hebrews:ch.   2, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   2, v.  14, pg. ``f30!}, 16-17\index{58 Hebrews:ch.   2, v.
 14, 16-17, pg. ``f30!}; Hebrews 4:15\index{58 Hebrews:ch.   4, v.  15, pg. ``f30!}.} being conceived
by\index{Jesus:conceived:by the power of the Holy Spirit ``f10!} the power of the Holy Ghost\index{Holy Spirit:Holy
Ghost ``f10!}, in the womb\index{Jesus:conceived:in womb of virgin Mary ``f10!} of the virgin\index{Jesus:born of
virgin Mary ``f10!} Mary\index{Mary, mother of Jesus ``f10!}, of her substance\index{Jesus:conceived:of Mary's
substance ``f10!}.\footnote{\ Luke 1:27, 31, 35\index{42 Luke:ch.   1, v.  27, 31, 35, pg. ``f30!}; Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} So that two whole, perfect, and distinct natures\index{Jesus:two
distinct natures ``f10!}, the Godhead\index{God's attributes:Godhead ``f10!} and the manhood\index{Jesus:human
nature:manhood ``f10!}, were inseparably\index{Jesus:human nature:inseparable from divine, one person ``f10!} joined
together in one person, without conversion, composition\index{God's attributes:being:no composition ``f10!}, or
confusion\index{God's attributes:being:no confusion ``f10!}.\footnote{\ Luke 1:35\index{42 Luke:ch.   1, v.  35, pg.
``f30!}; Colossians 2:9\index{51 Colossians:ch.   2, v.   9, pg. ``f30!}; Romans 9:5\index{45 Romans:ch.   9, v.   5,
pg. ``f30!}; 1 Peter 3:18\index{60 1 Peter:ch.   3, v.  18, pg. ``f30!}; 1 Timothy 3:1\index{54 1 Timothy:ch.   3, v.  
1, pg. ``f30!}6\index{54 1 Timothy:ch.   3, v.  16, pg. ``f30!}.} Which person is very God, and very man, yet one
Christ, the only Mediator\index{Jesus:mediator:of covenant ``f10!} between God and man.\footnote{\ Romans
1:3-4\index{45 Romans:ch.   1, v.   3-4, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.}


\bigskip

8.3\ \ The Lord Jesus, in His human nature thus\index{Jesus:human nature:united with divine ``f10!} united to the
divine, was sanctified\index{Sanctification:sanctification ``f10!} and anointed\index{Jesus:anointed with Holy Spirit
``f10!} with the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}, above measure,\footnote{\ Psalm 45:7\index{19
Psalms:ch.  45, v.   7, pg. ``f30!}; John 3:3\index{43 John:ch.   3, v.   3, pg. ``f30!}4\index{43 John:ch.   3, v. 
34, pg. ``f30!}.} having in Him all the treasures of wisdom and\index{Jesus:wisdom, treasures of ``f10!}
knowledge;\index{Jesus:human nature:all knowledge ``f10!}\footnote{\ Colossians 2:3\index{51 Colossians:ch.   2, v.  
3, pg. ``f30!}.} in whom it pleased\index{Providence:at God's pleasure ``f10!} the Father\index{Father:God the ``f10!}
that all fulness should\index{Jesus:fulness of Godhead dwells ``f10!} dwell;\footnote{\ Colossians 1:19\index{51
Colossians:ch.   1, v.  19, pg. ``f30!}.} to the end that, being holy\index{Jesus:holy ``f10!},
harmless\index{Jesus:harmless ``f10!}, undefiled\index{Jesus:undefiled ``f10!}, and full of grace\index{Jesus:full of
grace ``f10!} and truth\index{God's attributes:truth ``f10!},\footnote{\ Hebrews 7:26\index{58 Hebrews:ch.   7, v.  26,
pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}.} He might
be thoroughly furnished\index{Jesus:mediator:fully furnished to execute office ``f10!} to execute the
office\index{Jesus:offices ``f10!} of a mediator\index{Jesus:mediator:of covenant ``f10!} and surety\index{Jesus:surety
for the elect ``f10!}.\footnote{\ Acts 10:38\index{44 Acts:ch.  10, v.  38, pg. ``f30!}; Hebrews 12:2\index{58
Hebrews:ch.  12, v.   2, pg. ``f30!}4\index{58 Hebrews:ch.  12, v.  24, pg. ``f30!}; Hebrews 7:22\index{58 Hebrews:ch. 
 7, v.  22, pg. ``f30!}.} Which office He took not unto Himself, but was thereunto called by His
Father,\footnote{\ Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}{}-5\index{58 Hebrews:ch.   5, v.   4-5,
pg. ``f30!}.} who put all power\index{Jesus:almighty power ``f10!} and judgment into\index{Jesus:judge:assigned by
Father ``f10!} His hand, and gave Him commandment\index{Jesus:commanded by the Father ``f10!} to execute the
same\index{Jesus:obedience:to God's law ``f10!}.\footnote{\ John 5:22, 27\index{43 John:ch.   5, v.  22, 27, pg.
``f30!}; Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}; Acts 2:36\index{44 Acts:ch.   2, v.  36, pg.
``f30!}.}

8.4\ \ This office\index{Jesus:offices ``f10!} the Lord Jesus did most willingly\index{Jesus:humiliation:willingly
undertaken ``f10!} undertake;\footnote{\ Psalm 40:7-8\index{19 Psalms:ch.  40, v.   7-8, pg. ``f30!} with Hebrews
10:5-10\index{58 Hebrews:ch.  10, v.   5-10, pg. ``f30!}; John 10:18\index{43 John:ch.  10, v.  18, pg. ``f30!};
Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg. ``f30!}.} which that He might discharge,
He\index{Jesus:discharged:his duties ``f10!} was made under the law,\footnote{\ Galatians 4:4\index{48 Galatians:ch.  
4, v.   4, pg. ``f30!}.} and did perfectly fulfil\index{Jesus:obedience:to God's law ``f10!} it,\footnote{\ Matthew
3:15\index{40 Matthew:ch.   3, v.  15, pg. ``f30!}; Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}.}
endured most grievous torments immediately\index{Jesus:humiliation:suffered:torments ``f10!} in His
soul,\footnote{\ Matthew 26:37-38\index{40 Matthew:ch.  26, v.  37-38, pg. ``f30!}; Luke 22:44\index{42 Luke:ch.  22,
v.  44, pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v.   4, pg. ``f30!}6\index{40 Matthew:ch.  27, v.  46, pg.
``f30!}.} and most painful\index{Jesus:humiliation:suffered pain ``f10!} sufferings in\index{Jesus:death:sufferings and
death ``f10!} His body;\footnote{\ Matthew Chapters 26-27.} was crucified\index{Jesus:crucified ``f10!}, and
died\index{Jesus:died ``f10!};\footnote{\ Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg. ``f30!}.} was
buried\index{Jesus:buried ``f10!}, and remained under the power of death\index{Jesus:death:under power of ``f10!}; yet
saw no corruption\index{Jesus:death:no corruption ``f10!}.\footnote{\ Acts. 2:23-24, 27; Acts 13:3\index{44 Acts:ch. 
13, v.   3, pg. ``f30!}7\index{44 Acts:ch.  13, v.  37, pg. ``f30!}; Romans 6:9\index{45 Romans:ch.   6, v.   9, pg.
``f30!}.} On the third day\index{Jesus:resurrection:third day ``f10!} He arose\index{Jesus:resurrection:historical fact
``f10!} from the dead\index{Jesus:resurrection:from the dead ``f10!},\footnote{\ 1 Corinthians 15:3-4\index{46 1
Corinthians:ch.  15, v.   3-4, pg. ``f30!}.} with the same body in which He suffered\index{Jesus:resurrection:body
same:in which he suffered ``f10!},\footnote{\ John 20:25, 27\index{43 John:ch.  20, v.  25, 27, pg. ``f30!}.} with
which also he ascended\index{Jesus:ascension:into heaven ``f10!} into heaven, and there sits\index{Jesus:in
heaven:sitting at right hand of Fathe ``f10!} at the right hand of His Father\index{Father:God the
``f10!},\footnote{\ Mark 16:19\index{41 Mark:ch.  16, v.  19, pg. ``f30!}.} making intercession\index{Jesus:in
heaven:intercession made for all believers ``f10!},\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg.
``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}; Hebrews 9:24\index{58 Hebrews:ch.   9, v.  24, pg. ``f30!};
Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}.} and shall return to judge men\index{Jesus:judge:of mankind
``f10!} and angels\index{Angels:angels ``f10!} at the end of the world.\footnote{\ Romans 14:9\index{45 Romans:ch.  14,
v.   9, pg. ``f30!}{}-10\index{45 Romans:ch.  14, v.   9-10, pg. ``f30!}; Acts 1:11\index{44 Acts:ch.   1, v.  11, pg.
``f30!}; Acts 10:42\index{44 Acts:ch.  10, v.  42, pg. ``f30!}; Matthew 13:40, 41-42\index{40 Matthew:ch.  13, v.  40,
41-42, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg. ``f30!}; 2 Peter 2:4\index{61 2 Peter:ch.   2, v.   4, pg.
``f30!}.}


\bigskip

8.5\ \ The Lord Jesus, by His perfect obedience\index{Jesus:obedience:to God's law ``f10!}, and sacrifice of
Himself\index{Jesus:sacrifice:of himself ``f10!}, which He, through the eternal\index{Holy Spirit:eternal ``f10!}
Spirit, once offered up unto God, has fully satisfied the justice of His\index{Jesus:satisfied divine justice ``f10!}
Father\index{Father:God the ``f10!};\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}9\index{45
Romans:ch.   5, v.  19, pg. ``f30!}; Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}, 16\index{58
Hebrews:ch.   9, v.  14, 16, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}4\index{58
Hebrews:ch.  10, v.  14, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}5-26\index{45 Romans:ch.   3, v.  25-26, pg. ``f30!}.} and
purchased\index{Redemption:purchased by Christ ``f10!}, not only reconciliation\index{Jesus:purchased reconciliation
``f10!}, but an everlasting inheritance\index{Believers:inheritance:everlasting ``f10!} in the kingdom of
heaven\index{Kingdom:of heaven ``f10!}, for all those whom the Father has given unto Him.\footnote{\ Daniel 9:24,
26\index{27 Daniel:ch.   9, v.  24, 26, pg. ``f30!}; Colossians 1:19\index{51 Colossians:ch.   1, v.  19, pg.
``f30!}{}-20\index{51 Colossians:ch.   1, v.  19-20, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}, 14\index{49 Ephesians:ch.   1, v.  11, 14, pg. ``f30!}; John
17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}; Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!},
15\index{58 Hebrews:ch.   9, v.  12, 15, pg. ``f30!}.}


\bigskip

8.6\ \ Although the work of redemption\index{Redemption:work of by Christ ``f10!} was not actually wrought by
Christ\index{Jesus:redeemer of God's elect ``f10!} till after His incarnation\index{Jesus:incarnation:appearing in our
nature ``f10!}, yet the virtue\index{Jesus:vicarious atonement ``f10!}, efficacy, and\index{Redemption:efficacy of
Christ's incarnation ``f10!} benefits thereof were\index{Believers:benefits:of redemption ``f10!} communicated unto the
elect\index{Election:elect of God ``f10!} in all ages successively\index{Redemption:effectually:communicated ``f10!}
from the beginning of the world, in and by those promises\index{Salvation:promises made ``f10!}, types\index{types, OT
``f10!}, and sacrifices\index{Ceremonial law:sacrifices ``f10!}, wherein He was revealed, and\index{Revelation:God's
word ``f10!} signified to be the seed of the\index{Jesus:seed of the woman ``f10!} woman which should
bruise\index{Jesus:bruised serpent's head ``f10!} the serpent\index{Satan:serpent, head to be bruised ``f10!}{}'s head;
and the Lamb\index{Jesus:Lamb slain ``f10!} slain\index{Jesus:slain from beginning of world ``f10!} from the beginning
of the world: being yesterday\index{Jesus:unchanging ``f10!} and to-day the same, and forever.\footnote{\ Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}; Genesis
3:15\index{01 Genesis:ch.   3, v.  15, pg. ``f30!}; Revelation 13:8\index{66 Revelation:ch.  13, v.   8, pg. ``f30!};
Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}.}


\bigskip

8.7\ \ Christ, in the work of mediation\index{Jesus:mediator:of covenant ``f10!}, acts according to both
natures\index{Jesus:two distinct natures ``f10!}, by each nature doing that which is proper to
itself:\footnote{\ Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}; 1 Peter 3:18\index{60 1 Peter:ch.   3,
v.  18, pg. ``f30!}.} yet, by reason of the unity\index{Jesus:human nature:united with divine ``f10!} of the person,
that which is proper to one nature, is sometimes in Scripture attributed to the person\index{Jesus:one person, two
natures ``f10!} denominated by the other nature.\footnote{\ Acts 20:28\index{44 Acts:ch.  20, v.  28, pg. ``f30!}; John
3:13\index{43 John:ch.   3, v.  13, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}6\index{62 1
John:ch.   3, v.  16, pg. ``f30!}.}


\bigskip


\bigskip


8.8\ \ To all those for whom Christ has purchased\index{Redemption:purchased by Christ ``f10!} redemption, He does
certainly and effectually apply\index{Redemption:effectually:applied ``f10!} and communicate the same,\footnote{\ John
6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}, 39\index{43 John:ch.   6, v.  37, 39, pg. ``f30!}; John
10:15-16\index{43 John:ch.  10, v.  15-16, pg. ``f30!}.} making intercession\index{Jesus:in heaven:intercession made
for all believers ``f10!} for them,\footnote{\ 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}; Romans
8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}.} and revealing
unto\index{Revelation:God's word ``f10!} them, in and by the Word, the\index{Scripture:word of God:for salvation
``f10!} mysteries of\index{Salvation:mystery of ``f10!} salvation\index{Salvation:belief in Jesus by faith
``f10!},\footnote{\ John 15:13, 15\index{43 John:ch.  15, v.  13, 15, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.
  1, v.   7, pg. ``f30!}{}-9\index{49 Ephesians:ch.   1, v.   7-9, pg. ``f30!}; John 17:6\index{43 John:ch.  17, v.  
6, pg. ``f30!}.} effectually persuading\index{Effectual call:persuading to believe and obey ``f10!} them by His Spirit
to\index{Holy Spirit:applies:faith ``f10!} believe and obey, and\index{God's law:obligation to obey:in all things
``f10!} governing\index{Providence:governing all ``f10!} their hearts by\index{Heart:new:governed by Holy Spirit
``f10!} His Word and Spirit\index{Revelation:Word and Spirit ``f10!};\footnote{\ John 14:26\index{43 John:ch.  14, v. 
26, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}; 2 Corinthians 4:13\index{47 2
Corinthians:ch.   4, v.  13, pg. ``f30!}; Romans 8:9\index{45 Romans:ch.   8, v.   9, pg. ``f30!}, 14\index{45
Romans:ch.   8, v.   9, 14, pg. ``f30!}; Romans 15:18-19\index{45 Romans:ch.  15, v.  18-19, pg. ``f30!}; John
17:17\index{43 John:ch.  17, v.  17, pg. ``f30!}.} overcoming\index{Jesus:conqueror of his and our enemies ``f10!} all
their enemies\index{Enemies:of believers overcome ``f10!} by His almighty\index{God's attributes:almighty power ``f10!}
power and wisdom, in\index{God's attributes:wisdom ``f10!} such manner, and ways, as are most consonant to His
wonderful and unsearchable\index{Providence:unsearchable wise counsel ``f10!} dispensation\index{Providence:God's
dispensation ``f10!}.\footnote{\ Psalm 110:1\index{19 Psalms:ch. 110, v.   1, pg. ``f30!}; 1 Corinthians 15:25\index{46
1 Corinthians:ch.  15, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  15, v.  25-26, pg. ``f30!}; Malachi
4:2-3\index{39 Malachi:ch.   4, v.   2-3, pg. ``f30!}; Colossians 2:15\index{51 Colossians:ch.   2, v.  15, pg.
``f30!}.}

\section{Chapter 9 -- Of Free Will}
\index{Causation:will of creatures:free ``f10!}9.1\ \ God has endued the\index{Causation:will of creatures:natural
liberty ``f10!} will of man with that natural liberty, that\index{Liberty:natural ``f10!} is neither forced, nor by any
absolute\index{Mankind:human will, natural liberty ``f10!} necessity of\index{Liberty:not absolute necessity to good or
evil ``f10!} nature determined to good or evil.\footnote{\ Matthew 17:12\index{40 Matthew:ch.  17, v.  12, pg. ``f30!};
James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}; Deuteronomy 30:19\index{05 Deuteronomy:ch.  30, v.  19, pg.
``f30!}.}


\bigskip

9.2\ \ Man, in his state of innocency, had\index{Mankind:created:state of innocence ``f10!} freedom
and\index{Causation:will of creatures:free ``f10!} power to will and to do that which was good, and well
pleasing\index{Mankind:created:initially able to please God ``f10!} to God;\footnote{\ Ecclesiastes 7:2\index{21
Ecclesiastes:ch.   7, v.   2, pg. ``f30!}; Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01
Genesis:ch.   1, v.  26, pg. ``f30!}.} but yet, mutably\index{Mankind:created:mutable ``f10!}, so that he might
fall\index{Sin:original:first transgression (of Adam) ``f10!} from it.\footnote{\ Genesis 2:16-17\index{01 Genesis:ch. 
 2, v.  16-17, pg. ``f30!}; Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}.}


\bigskip

9.3\ \ Man, by his fall\index{Sin:original:first transgression (of Adam) ``f10!} into a state of sin, has wholly lost
all ability of\index{Sin:original:inability to:do anything good ``f10!} will to any spiritual
good\index{Sin:original:lost all spiritual good ``f10!} accompanying salvation\index{Salvation:belief in Jesus by faith
``f10!}:\footnote{\ Romans 5:6\index{45 Romans:ch.   5, v.   6, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.  
7, pg. ``f30!}; John 15:5\index{43 John:ch.  15, v.   5, pg. ``f30!}.} so as, a natural man, being altogether averse
from that good,\footnote{\ Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}, 12\index{45 Romans:ch.   3, v. 
10, 12, pg. ``f30!}.} and dead in sin\index{Death:spiritual, in sin ``f10!},\footnote{\ Ephesians 2:1\index{49
Ephesians:ch.   2, v.   1, pg. ``f30!}, 5\index{49 Ephesians:ch.   2, v.   1, 5, pg. ``f30!}; Colossians 2:13\index{51
Colossians:ch.   2, v.  13, pg. ``f30!}.} is not able, by his own strength, to\index{Conversion:not by own strength
``f10!} convert himself\index{Conversion:not of self ``f10!}, or to prepare himself thereunto.\footnote{\ John
6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}, 65\index{43 John:ch.   6, v.  44, 65, pg. ``f30!}; Ephesians
2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-5\index{49 Ephesians:ch.   2, v.   2-5, pg. ``f30!}; 1
Corinthians 2:14\index{46 1 Corinthians:ch.   2, v.  14, pg. ``f30!}; Titus 3:3-5\index{56 Titus:ch.   3, v.   3-5, pg.
``f30!}.} 


\bigskip

9.4\ \ When God converts\index{Conversion:conversion ``f10!} a sinner, and translates\index{Conversion:translated into
state of grace ``f10!} him into the state of grace\index{Grace:state:translation into at conversion ``f10!}, He
frees\index{Believers:freed from:guilt of sin ``f10!} him from his natural bondage under\index{Sin:bondage to ``f10!}
sin;\footnote{\ Colossians 1:13\index{51 Colossians:ch.   1, v.  13, pg. ``f30!}; John 8:34\index{43 John:ch.   8, v. 
34, pg. ``f30!}, 36\index{43 John:ch.   8, v.  34, 36, pg. ``f30!}.} and, by His grace\index{Grace:God's free ``f10!}
alone, enables\index{Sanctification:enabled by:grace ``f10!} him freely\index{Causation:will of creatures:free ``f10!}
to will and to do that which is spiritually good;\index{Good works:spiritual services ``f10!}\footnote{\ Philippians
2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Romans 6:18, 22\index{45 Romans:ch.   6, v.  18, 22, pg.
``f30!}.} yet so, as that by reason of his remaining corruption, he\index{Sanctification:remnants of corruption remain
``f10!} does not perfectly, nor only, will that which is good, but does also will that which is
evil.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg.
``f30!}; Romans 7:15, 18\index{45 Romans:ch.   7, v.  15, 18, pg. ``f30!}{}-19, 21, 23\index{45 Romans:ch.   7, v.  15,
18-19, 21, 23, pg. ``f30!}.} 


\bigskip

9.5\ \ The will of man is made perfectly and immutably\index{God's attributes:immutable ``f10!} free to do good alone,
in the state of glory\index{Resurrection:believers:into glory:light ``f10!} only.\footnote{\ Ephesians 4:13\index{49
Ephesians:ch.   4, v.  13, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}3\index{58
Hebrews:ch.  12, v.  23, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg. ``f30!}; Jude 24\index{65 Jude:v.
 24, pg. ``f30!}.}

\section{Chapter 10 -- Of Effectual Calling}
\index{Effectual call:calling ``f10!}10.1\ \ All those whom God has predestinated\index{Predestination:predestination
``f10!} unto life, and those only, He is pleased\index{Providence:at God's pleasure ``f10!} in His appointed and
accepted time\index{Salvation:at appointed time ``f10!} effectually to\index{Effectual call:calling ``f10!}
call,\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!};
Romans 11:7\index{45 Romans:ch.  11, v.   7, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}0-11\index{49 Ephesians:ch.   1, v.  10-11, pg. ``f30!}.} by His Word and Spirit\index{Revelation:Word and
Spirit ``f10!},\footnote{\ 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}{}-14\index{53 2
Thessalonians:ch.   2, v.  13-14, pg. ``f30!}; 2 Corinthians 3:3, 6\index{47 2 Corinthians:ch.   3, v.   3, 6, pg.
``f30!}.} out of that state of sin and death\index{Death:spiritual, in sin ``f10!}, in which they are by nature, to
grace\index{Grace:of salvation ``f10!} and salvation\index{Salvation:belief in Jesus by faith ``f10!} by Jesus
Christ;\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2,
v.   1, pg. ``f30!}{}-5\index{49 Ephesians:ch.   2, v.   1-5, pg. ``f30!}; 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.
  9, pg. ``f30!}{}-10\index{55 2 Timothy:ch.   1, v.   9-10, pg. ``f30!}.} enlightening\index{Effectual
call:enlightening minds spiritually ``f10!} their minds spiritually and savingly to understand the things of
God,\footnote{\ Acts 26:18\index{44 Acts:ch.  26, v.  18, pg. ``f30!}; 1 Corinthians 2:10, 12\index{46 1
Corinthians:ch.   2, v.  10, 12, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}7-18\index{49 Ephesians:ch.   1, v.  17-18, pg. ``f30!}.} taking away their heart of stone\index{Heart:stone
``f10!}, and giving unto them a heart of flesh\index{Heart:flesh, sinful nature ``f10!};\footnote{\ Ezekiel
36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}.} renewing the\index{Sanctification:renewed in image of God
``f10!}ir\index{Conversion:renewed by Holy Spirit ``f10!} wills, and, by\index{Conversion:renewal of the will ``f10!}
His almighty\index{God's attributes:almighty power ``f10!} power determining them\index{Predestination:determining
wills ``f10!} to that which is good,\footnote{\ Ezekiel 11:19\index{26 Ezekiel:ch.  11, v.  19, pg. ``f30!};
Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Deuteronomy 30:6\index{05 Deuteronomy:ch.  30, v.  
6, pg. ``f30!}; Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}.} and effectually drawing
them\index{Effectual call:drawing to Christ ``f10!} to Jesus Christ:\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.  
1, v.   1, pg. ``f30!}9\index{49 Ephesians:ch.   1, v.  19, pg. ``f30!}; John 6:44\index{43 John:ch.   6, v.  44, pg.
``f30!}{}-45\index{43 John:ch.   6, v.  44-45, pg. ``f30!}.} yet so, as they come most freely\index{Effectual
call:enables person to:freely answer call ``f10!}, being made willing by His grace\index{Grace:God's free
``f10!}.\footnote{\ Song of Solomon 1:4\index{22 Song of Solomon:ch.   1, v.   4, pg. ``f30!}; Psalm 110:3\index{19
Psalms:ch. 110, v.   3, pg. ``f30!}; John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}; Romans 6:16-18\index{45
Romans:ch.   6, v.  16-18, pg. ``f30!}.}


\bigskip

10.2\ \ This effectual call\index{Effectual call:calling ``f10!} is of God's free and special grace\index{Grace:God's
free ``f10!} alone, not from anything at all foreseen\index{Predestination:not based on God's foresight:of anything in
man ``f10!} in man,\footnote{\ 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.   9, pg. ``f30!}; Titus 3:4-5\index{56
Titus:ch.   3, v.   4-5, pg. ``f30!}; Ephesians 2:4-5, 8-9\index{49 Ephesians:ch.   2, v.   4-5, 8-9, pg. ``f30!};
Romans 9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}.} who is altogether passive\index{Effectual call:mankind
passive ``f10!} therein, until being quickened\index{Conversion:quickened to new life ``f10!} and renewed
by\index{Conversion:renewed by Holy Spirit ``f10!} the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!},\footnote{\ 1
Corinthians 2:14\index{46 1 Corinthians:ch.   2, v.  14, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7, pg.
``f30!}; Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5, pg. ``f30!}.} he is thereby enabled to
answer\index{Effectual call:enables person to:freely answer call ``f10!} this call, and to embrace the
grace\index{Grace:in believers:by Holy Spirit ``f10!} offered and\index{Effectual call:enables person to:embrace grace
``f10!} conveyed in it.\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}; John 5:25\index{43
John:ch.   5, v.  25, pg. ``f30!}; John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}.}


\bigskip

10.3\ \ Elect\index{Infants:elect, dying in infancy ``f10!} infants, dying in infancy, are
regenerated\index{Regeneration:regeneration ``f10!}, and saved by Christ\index{Salvation:by Christ ``f10!} through the
Spirit\index{Holy Spirit:applies:regeneration ``f10!},\footnote{\ Luke 18:1\index{42 Luke:ch.  18, v.   1, pg.
``f30!}5\index{42 Luke:ch.  18, v.  15, pg. ``f30!}{}-16\index{42 Luke:ch.  18, v.  15-16, pg. ``f30!} and Acts
2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!}; John 3:3\index{43
John:ch.   3, v.   3, pg. ``f30!}, 5\index{43 John:ch.   3, v.   3, 5, pg. ``f30!}; 1 John 5:12\index{62 1 John:ch.  
5, v.  12, pg. ``f30!}; and Romans 8:9\index{45 Romans:ch.   8, v.   9, pg. ``f30!} compared.} who
works\index{Providence:God's works ``f10!} when, and where, and how He pleases\index{Providence:at God's pleasure
``f10!}:\footnote{\ John 3:8\index{43 John:ch.   3, v.   8, pg. ``f30!}.} so also, are all other
elect\index{Election:elect of God ``f10!} persons who are uncapable\index{Infants:incapable of being outwardly called
``f10!} of being outwardly called\index{Preaching:outward call ``f10!} by the ministry of the
Word\index{Preaching:ministry of the Word ``f10!}.\footnote{\ 1 John 5:12\index{62 1 John:ch.   5, v.  12, pg. ``f30!};
Acts 4:12\index{44 Acts:ch.   4, v.  12, pg. ``f30!}.}


\bigskip

10.4\ \ Others, not elected\index{Election:some not chosen ``f10!}, although they may be called by the ministry of the
Word\index{Preaching:ministry of the Word ``f10!},\footnote{\ Matthew 22:14\index{40 Matthew:ch.  22, v.  14, pg.
``f30!}.} and may have some common operations\index{Holy Spirit:operations:common ``f10!} of the
Spirit,\footnote{\ Matthew 7:22\index{40 Matthew:ch.   7, v.  22, pg. ``f30!}; Matthew 13:20-21\index{40 Matthew:ch. 
13, v.  20-21, pg. ``f30!}; Hebrews 6:4-5\index{58 Hebrews:ch.   6, v.   4-5, pg. ``f30!}.} yet they never
truly\index{Unbelievers:non-elect never truly come to Jesus ``f10!} come unto Christ, and therefore cannot be
saved:\index{Salvation:not in other religions ``f10!}\footnote{\ John 6:64-66\index{43 John:ch.   6, v.  64-66, pg.
``f30!}; John 8:24\index{43 John:ch.   8, v.  24, pg. ``f30!}.} much less can men, not professing
the\index{Profession:absent ``f10!} Christian religion\index{Profession:of false religion ``f10!}, be saved in any
other way whatsoever, be they never so diligent to\index{God's law:diligence in application ``f10!} frame
their\index{Salvation:not by:obedience ``f10!} lives according to the light of nature, and the law of that religion
they do profess.\footnote{\ Acts 4:12\index{44 Acts:ch.   4, v.  12, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch. 
 2, v.   1, pg. ``f30!}2\index{49 Ephesians:ch.   2, v.  12, pg. ``f30!}; John 4:22\index{43 John:ch.   4, v.  22, pg.
``f30!}; John 14:6\index{43 John:ch.  14, v.   6, pg. ``f30!}; John 17:3\index{43 John:ch.  17, v.   3, pg. ``f30!}.}
And to assert and maintain that they may, is very pernicious\index{Salvation:outside of Christ, pernicious and
detestable claim ``f10!}, and to be detested.\footnote{\ 2 John 9-11\index{63 2 John:v. 9-11, pg. ``f30!}; 1
Corinthians 16:2\index{46 1 Corinthians:ch.  16, v.   2, pg. ``f30!}2\index{46 1 Corinthians:ch.  16, v.  22, pg.
``f30!}; Galatians 1:6-8\index{48 Galatians:ch.   1, v.   6-8, pg. ``f30!}.}

\section{Chapter 11 -- Of Justification}
\index{Justification:justification ``f10!}11.1\ \ Those whom God effectually calls\index{Effectual call:calling ``f10!},
He also freely justifies\index{Justification:freely by God ``f10!};\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.  
2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg.
``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}.} not by infusing righteousness into\index{Jesus:righteousness
imputed to believers ``f10!} them, but by pardoning\index{Justification:pardoned of sin ``f10!} their sins, and by
accounting and\index{Justification:accounted by God as righteous ``f10!} accepting their\index{Justification:accepted
by God as righteous ``f10!} persons as righteous, not for anything wrought in them\index{Salvation:not by:works
``f10!}, or done by them, but for Christ's sake alone; nor by imputing\index{Imputation:of Christ's righteousness by
faith ``f10!} faith itself, the act of believing, or\index{Effectual call:not based on act of believing ``f10!} any
other evangelical obedience\index{God's law:commandments:obedience:required ``f10!} to them, as their righteousness,
but by imputing the obedience and\index{Jesus:obedience:imputed to believers ``f10!} satisfaction of
Christ\index{Jesus:satisfied divine justice ``f10!} unto them,\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2,
pg. ``f30!}2, 24-25\index{45 Romans:ch.   3, v.  22, 24-25, pg. ``f30!}, 27-28\index{45 Romans:ch.   3, v.  22, 24-25,
27-28, pg. ``f30!}; Romans 4:5\index{45 Romans:ch.   4, v.   5, pg. ``f30!}{}-8\index{45 Romans:ch.   4, v.   5-8, pg.
``f30!}; 2 Corinthians 5:19\index{47 2 Corinthians:ch.   5, v.  19, pg. ``f30!}, 21\index{47 2 Corinthians:ch.   5, v. 
19, 21, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}, 7\index{56 Titus:ch.   3, v.   5, 7, pg.
``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}; Jeremiah 23:6\index{24 Jeremiah:ch.  23, v.  
6, pg. ``f30!}; 1 Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg. ``f30!}{}-31\index{46 1 Corinthians:ch. 
 1, v.  30-31, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}7-19\index{45 Romans:ch.   5, v. 
17-19, pg. ``f30!}.} they receiving and\index{Faith:instrument:for receiving Christ ``f10!} resting
on\index{Faith:resting on Christ ``f10!} Him and His righteousness\index{Jesus:righteous ``f10!} by faith;
which\index{Faith:in Jesus Christ ``f10!} faith they have not of themselves, it is the gift of God\index{Faith:gift of
God ``f10!}.\footnote{\ Acts 10:43\index{44 Acts:ch.  10, v.  43, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.  
2, v.  16, pg. ``f30!}; Philippians 3:19\index{50 Philippians:ch.   3, v.  19, pg. ``f30!}; Acts 13:3\index{44 Acts:ch.
 13, v.   3, pg. ``f30!}8-39\index{44 Acts:ch.  13, v.  38-39, pg. ``f30!}; Ephesians 2:7\index{49 Ephesians:ch.   2,
v.   7, pg. ``f30!}{}-8\index{49 Ephesians:ch.   2, v.   7-8, pg. ``f30!}.}


\bigskip

11.2\ \ Faith\index{Faith:saving ``f10!}, thus receiving and\index{Faith:instrument:for receiving Christ ``f10!} resting
on\index{Faith:resting on Christ ``f10!} Christ and His righteousness\index{Jesus:righteous ``f10!}, is the alone
instrument\index{Faith:instrument:of justification ``f10!} of justification\index{Justification:justification
``f10!};\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!};
Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}8\index{45 Romans:ch.   3, v.  28, pg. ``f30!}; Romans
5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}.} yet is it not alone in the person
justified\index{Justification:justified ``f10!}, but is ever accompanied with all other saving graces\index{Grace:of
salvation ``f10!}, and is no dead\index{Faith:not dead when resting on Christ ``f10!} faith, but
works\index{Faith:works by love ``f10!} by love.\footnote{\ James 2:17, 22, 26\index{59 James:ch.   2, v.  17, 22, 26,
pg. ``f30!}; Galatians 5:6\index{48 Galatians:ch.   5, v.   6, pg. ``f30!}.}


\bigskip

11.3\ \ Christ, by His obedience\index{Jesus:obedience:to God's law ``f10!} and death, did\index{Jesus:death:death
``f10!} fully discharge the\index{Jesus:discharged:debt of sin for his people ``f10!} debt\index{Forgiveness:of debtors
``f10!} of all those that are thus justified\index{Justification:justified ``f10!}, and did make a proper, real, and
full satisfaction to His\index{Jesus:satisfied divine justice ``f10!} Father\index{Father:God the ``f10!}{}'s justice
in their behalf.\footnote{\ Romans 5:8-10, 19\index{45 Romans:ch.   5, v.   8-10, 19, pg. ``f30!}; 1 Timothy
2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}{}-6\index{54 1 Timothy:ch.   2, v.   5-6, pg. ``f30!}; Hebrews
10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}0\index{58 Hebrews:ch.  10, v.  10, pg. ``f30!}, 14\index{58
Hebrews:ch.  10, v.  10, 14, pg. ``f30!}; Daniel 9:24, 26\index{27 Daniel:ch.   9, v.  24, 26, pg. ``f30!}; Isaiah
53:4-6, 10-12\index{23 Isaiah:ch.  53, v.   4-6, 10-12, pg. ``f30!}.} Yet, inasmuch as He was given by the Father for
them;\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}2\index{45 Romans:ch.   8, v.  32, pg.
``f30!}.} and His obedience and satisfaction accepted in their stead\index{Jesus:vicarious atonement
``f10!};\footnote{\ 2 Corinthians 5:21\index{47 2 Corinthians:ch.   5, v.  21, pg. ``f30!}; Matthew 3:17\index{40
Matthew:ch.   3, v.  17, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}.} and both
freely\index{Salvation:freely offered ``f10!}, not for anything in them; their
justification\index{Justification:justification ``f10!} is only of free grace\index{Grace:God's free
``f10!};\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg.
``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}.} that both the exact justice\index{God's
attributes:administers justice ``f10!}, and rich grace\index{Grace:God's rich ``f10!} of God, might be
glorified\index{God's attributes:glory:his own ``f10!} in the justification of sinners.\footnote{\ Romans 3:2\index{45
Romans:ch.   3, v.   2, pg. ``f30!}6\index{45 Romans:ch.   3, v.  26, pg. ``f30!}; Ephesians 2:7\index{49 Ephesians:ch.
  2, v.   7, pg. ``f30!}.}


\bigskip

11.4\ \ God did, from all eternity\index{God's attributes:eternal ``f10!}, decree\index{Predestination:eternal decrees
``f10!} to justify a\index{Justification:by faith ``f10!}ll\index{Justification:by God's decree ``f10!} the
elect\index{Election:elect of God ``f10!},\footnote{\ Galatians 3:8\index{48 Galatians:ch.   3, v.   8, pg. ``f30!}; 1
Peter 1:2\index{60 1 Peter:ch.   1, v.   2, pg. ``f30!}, 19-20\index{60 1 Peter:ch.   1, v.   2, 19-20, pg. ``f30!};
Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}.} and Christ did,
in the fulness of time\index{Jesus:incarnation:in fulness of time ``f10!}, die for\index{Death:punishment for sin
``f10!} their sins, and rise again\index{Jesus:resurrection:for justification ``f10!} for their
justification\index{Justification:justification ``f10!}:\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4,
pg. ``f30!}; 1 Timothy 2:6\index{54 1 Timothy:ch.   2, v.   6, pg. ``f30!}; Romans 4:25\index{45 Romans:ch.   4, v. 
25, pg. ``f30!}.} nevertheless, they are not justified\index{Justification:justified ``f10!}, until the Holy
Spirit\index{Holy Spirit:Holy Spirit ``f10!} does, in due time, actually apply Christ unto them.\footnote{\ Colossians
1:21\index{51 Colossians:ch.   1, v.  21, pg. ``f30!}{}-22\index{51 Colossians:ch.   1, v.  21-22, pg. ``f30!};
Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Titus 3:3-7\index{56 Titus:ch.   3, v.   3-7, pg.
``f30!}.}


\bigskip

11.5\ \ God does continue to forgive the\index{Forgiveness:God forgives sin ``f10!} sins of those that are
justified\index{Justification:justified ``f10!}:\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg.
``f30!}2\index{40 Matthew:ch.   6, v.  12, pg. ``f30!}; 1 John 1:7\index{62 1 John:ch.   1, v.   7, pg. ``f30!},
9\index{62 1 John:ch.   1, v.   7, 9, pg. ``f30!}; 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}.} and
although they can never fall\index{Assurance:never fall from being saved ``f10!} from the state of
justification\index{Justification:justification ``f10!};\footnote{\ Luke 22:32\index{42 Luke:ch.  22, v.  32, pg.
``f30!}; John 10:28\index{43 John:ch.  10, v.  28, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg.
``f30!}4\index{58 Hebrews:ch.  10, v.  14, pg. ``f30!}.} yet they may, by their sins, fall under God's
fatherly\index{Father:God the ``f10!} displeasure, and not\index{Believers:may experience God's displeasure for sin
``f10!} have the light of His countenance\index{Believers:chastened by God ``f10!} restored unto them, until they
humble themselves\index{Humility:before God ``f10!}, confess\index{Confession:of sin ``f10!} their sins, beg
pardon\index{Forgiveness:beg pardon for ``f10!}, and renew their faith\index{Faith:renewal of ``f10!} and
repentance\index{Repentance:for all sins ``f10!}.\footnote{\ Psalm 89:31-33\index{19 Psalms:ch.  89, v.  31-33, pg.
``f30!}; Psalm 51:7-12\index{19 Psalms:ch.  51, v.   7-12, pg. ``f30!}; Psalm 32:5\index{19 Psalms:ch.  32, v.   5, pg.
``f30!}; Matthew 26:75\index{40 Matthew:ch.  26, v.  75, pg. ``f30!}; 1 Corinthians 11:30, 32\index{46 1
Corinthians:ch.  11, v.  30, 32, pg. ``f30!}; Luke 1:20\index{42 Luke:ch.   1, v.  20, pg. ``f30!}.} 

11.6\ \ The justification\index{Justification:justification ``f10!} of believers under\index{Justification:of believers,
same under OT and NT ``f10!} the old testament\index{Scripture:Old Testament ``f10!} was, in all these respects, one
and the same with the justification of believers under the new testament\index{Scripture:New Testament
``f10!}.\footnote{\ Galatians 3:9, 13-14\index{48 Galatians:ch.   3, v.   9, 13-14, pg. ``f30!}; Romans
4:22-24\index{45 Romans:ch.   4, v.  22-24, pg. ``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}.}

\section{Chapter 12 -- Of Adoption}
\index{Adoption:in salvation ``f10!}12.1\ \ All those that are justified\index{Justification:justified ``f10!}, God
vouchsafes\index{Justification:God vouchsafes in Jesus ``f10!}, in and for His only Son\index{Jesus:Son of God ``f10!}
Jesus Christ, to make partakers of the grace of\index{Grace:of adoption ``f10!} adoption\index{Adoption:in salvation
``f10!}:\footnote{\ Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}.} by which they are taken into the
number, and enjoy the\index{Believers:enjoy:sense of God's love ``f10!} liberties\index{Christian liberty:privileges of
children of God ``f10!} and privileges\index{Believers:privileges ``f10!} of the children of God,\footnote{\ Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}; Romans
8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}7\index{45 Romans:ch.   8, v.  17, pg. ``f30!}; John 1:1\index{43
John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}.} have His name put upon
them,\footnote{\ Jeremiah 14:9\index{24 Jeremiah:ch.  14, v.   9, pg. ``f30!}; 2 Corinthians 6:18\index{47 2
Corinthians:ch.   6, v.  18, pg. ``f30!}; Revelation 3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}2\index{66
Revelation:ch.   3, v.  12, pg. ``f30!}.} receive the spirit\index{Justification:receive the spirit of adoption ``f10!}
of adoption,\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}5\index{45 Romans:ch.   8, v.  15, pg.
``f30!}.} have access to the\index{Believers:bold access to God's throne ``f10!} throne of grace\index{Grace:throne of
``f10!} with boldness\index{Prayer:with boldness ``f10!},\footnote{\ Ephesians 3:12\index{49 Ephesians:ch.   3, v.  12,
pg. ``f30!}; Romans 5:2\index{45 Romans:ch.   5, v.   2, pg. ``f30!}.} are enabled to cry\index{Prayer:cry Abba Father
``f10!}, Abba\index{God's attributes:Abba ``f10!}, Father\index{Father:God the ``f10!},\footnote{\ Galatians
4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.} are pitied\index{Justification:pitied, protected, provided for
``f10!},\footnote{\ Psalm 103:1\index{19 Psalms:ch. 103, v.   1, pg. ``f30!}3\index{19 Psalms:ch. 103, v.  13, pg.
``f30!}.} protected,\footnote{\ Proverbs 14:26\index{20 Proverbs:ch.  14, v.  26, pg. ``f30!}.} provided
for,\index{Believers:protected and provided for by God ``f10!}\footnote{\ Matthew 6:30\index{40 Matthew:ch.   6, v. 
30, pg. ``f30!}, 32\index{40 Matthew:ch.   6, v.  30, 32, pg. ``f30!}; 1 Peter 5:7\index{60 1 Peter:ch.   5, v.   7,
pg. ``f30!}.} and chastened\index{Believers:chastened by God ``f10!} by Him as by a Father;\footnote{\ Hebrews
12:6\index{58 Hebrews:ch.  12, v.   6, pg. ``f30!}.} yet never cast off\index{Believers:never cast off
``f10!},\footnote{\ Lamentations 3:31\index{25 Lamentations:ch.   3, v.  31, pg. ``f30!}.} but sealed to
the\index{Salvation:sealed to the day of redemption ``f10!} day of redemption\index{Redemption:day of
``f10!},\footnote{\ Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30,
pg. ``f30!}.} and inherit\index{Believers:inheritance:everlasting ``f10!} the promises,\footnote{\ Hebrews
6:12\index{58 Hebrews:ch.   6, v.  12, pg. ``f30!}.} as heirs\index{Believers:heirs (salvation, promises, glory)
``f10!} of everlasting\index{Salvation:everlasting ``f10!} salvation\index{Salvation:belief in Jesus by faith
``f10!}.\footnote{\ 1 Peter 1:3-4\index{60 1 Peter:ch.   1, v.   3-4, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.  
1, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   1, v.  14, pg. ``f30!}.}

\section{Chapter 13 -- Of Sanctification}
\index{Sanctification:sanctification ``f10!}13.1\ \ They who are once effectually called\index{Effectual call:calling
``f10!} and regenerated\index{Regeneration:regeneration ``f10!}, having a new heart\index{Heart:new:regenerate ``f10!}
and a new spirit created in them\index{Conversion:created in Christ ``f10!}, are further
sanctified\index{Sanctification:sanctification ``f10!}, really and personally\index{Sanctification:personal ``f10!},
through the virtue\index{Jesus:vicarious atonement ``f10!} of Christ's death\index{Jesus:death:death ``f10!} and
resurrection,\index{Jesus:resurrection:from the dead ``f10!}\footnote{\ 1 Corinthians 6:11\index{46 1 Corinthians:ch.  
6, v.  11, pg. ``f30!}; Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}; Philippians 3:10\index{50
Philippians:ch.   3, v.  10, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}{}-6\index{45
Romans:ch.   6, v.   5-6, pg. ``f30!}.} by His Word and Spirit\index{Revelation:Word and Spirit ``f10!} dwelling
in\index{Holy Spirit:abiding in believers ``f10!} them:\footnote{\ John 17:17\index{43 John:ch.  17, v.  17, pg.
``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}6\index{49 Ephesians:ch.   5, v.  26, pg.
``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.} the dominion of
the\index{Regeneration:dominion of sin destroyed ``f10!} whole body of sin is destroyed,\footnote{\ Romans 6:6,
14\index{45 Romans:ch.   6, v.   6, 14, pg. ``f30!}.} and the several lusts\index{Sexual immorality:lust ``f10!}
thereof are more and more weakened\index{Sanctification:lusts weakened and mortified ``f10!} and
mortified\index{Sin:mortified ``f10!};\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg.
``f30!}4\index{48 Galatians:ch.   5, v.  24, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg.
``f30!}3\index{45 Romans:ch.   8, v.  13, pg. ``f30!}.} and they more and more quickened\index{Conversion:quickened to
new life ``f10!} and strengthened in\index{Sanctification:supply of strength from Holy Spirit ``f10!} all saving
graces\index{Grace:of salvation ``f10!},\footnote{\ Colossians 1:11\index{51 Colossians:ch.   1, v.  11, pg. ``f30!};
Ephesians 3:16\index{49 Ephesians:ch.   3, v.  16, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  16-19, pg.
``f30!}.} to the practice of true holiness\index{Believers:holy:being perfected ``f10!}, without which no man shall see
the Lord.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; Hebrews 12:1\index{58
Hebrews:ch.  12, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.  12, v.  14, pg. ``f30!}.}


\bigskip

13.2\ \ This sanctification\index{Sanctification:sanctification ``f10!} is throughout, in the whole man;\footnote{\ 1
Thessalonians 5:23\index{52 1 Thessalonians:ch.   5, v.  23, pg. ``f30!}.} yet imperfect
in\index{Sanctification:imperfection in believers ``f10!} this life, there abiding still\index{Sin:remnants in
believers ``f10!} some remnants\index{Sanctification:remnants of corruption remain ``f10!} of corruption in every
part:\footnote{\ 1 John 1:10\index{62 1 John:ch.   1, v.  10, pg. ``f30!}; Romans 7:18\index{45 Romans:ch.   7, v.  18,
pg. ``f30!}, 23\index{45 Romans:ch.   7, v.  18, 23, pg. ``f30!}; Philippians 3:12\index{50 Philippians:ch.   3, v. 
12, pg. ``f30!}.} whence arises a continual and irreconcilable war; the flesh lusting\index{Sexual immorality:lust
``f10!} against the Spirit, and the Spirit against\index{Holy Spirit:at war with believers' lusts and flesh ``f10!} the
flesh.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17,
pg. ``f30!}; 1 Peter 2:11\index{60 1 Peter:ch.   2, v.  11, pg. ``f30!}.}


\bigskip

13.3\ \ In which war, although the remaining corruption, for\index{Sanctification:remnants of corruption remain ``f10!}
a time, may much prevail\index{Sanctification:imperfection in believers ``f10!};\footnote{\ Romans 7:23\index{45
Romans:ch.   7, v.  23, pg. ``f30!}.} yet through the continual supply\index{Sanctification:supply of strength from
Holy Spirit ``f10!} of strength from the sanctifying\index{Sanctification:sanctification ``f10!} Spirit of
Christ\index{Holy Spirit:Spirit of Christ ``f10!}, the regenerate\index{Regeneration:regeneration ``f10!} part does
overcome\index{Sin:overcoming ``f10!};\footnote{\ Romans 6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}; 1 John
5:4\index{62 1 John:ch.   5, v.   4, pg. ``f30!}; Ephesians 4:15-16\index{49 Ephesians:ch.   4, v.  15-16, pg.
``f30!}.} and so, the saints\index{Believers:saints ``f10!} grow in\index{Sanctification:spiritual growth:in grace
``f10!} grace,\footnote{\ 2 Peter 3:18\index{61 2 Peter:ch.   3, v.  18, pg. ``f30!}; 2 Corinthians 3:18\index{47 2
Corinthians:ch.   3, v.  18, pg. ``f30!}.} perfecting holiness\index{Believers:holy:being perfected ``f10!} in the fear
of\index{Fear:of God:reverence ``f10!} God.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg.
``f30!}.}

\section{Chapter 14 -- Of Saving Faith}
\index{Salvation:belief in Jesus by faith ``f10!}\index{Faith:saving ``f10!}14.1\ \ The grace of faith\index{Faith:given
by grace ``f10!}, whereby the elect\index{Election:elect of God ``f10!} are enabled to believe\index{Conversion:Holy
Spirit:enables belief ``f10!} to the saving of\index{Salvation:by working of Holy Spirit ``f10!} their
souls,\index{Holy Spirit:applies:faith ``f10!}\footnote{\ Hebrews 10:39\index{58 Hebrews:ch.  10, v.  39, pg. ``f30!}.}
is the work of the Spirit of Christ\index{Holy Spirit:Spirit of Christ ``f10!} in their
hearts;\index{Heart:new:enlightened ``f10!}\footnote{\ 2 Corinthians 4:13\index{47 2 Corinthians:ch.   4, v.  13, pg.
``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}7-19\index{49 Ephesians:ch.   1, v.  17-19, pg.
``f30!}; Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}.} and is ordinarily wrought by
the\index{Preaching:convincing and converting sinners ``f10!} ministry of the Word\index{Preaching:ministry of the Word
``f10!}:\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg.
``f30!}, 17\index{45 Romans:ch.  10, v.  14, 17, pg. ``f30!}.} by which also, and by the
administration\index{Sacraments:grace:administered ``f10!} of the sacraments\index{Sacraments:holy:ordinances ``f10!},
and prayer, it\index{Prayer:act of worship ``f10!} is increased and strengthened\index{Sacraments:strengthening
ordinances ``f10!}.\footnote{\ 1 Peter 2:2\index{60 1 Peter:ch.   2, v.   2, pg. ``f30!}; Acts 20:32\index{44 Acts:ch. 
20, v.  32, pg. ``f30!}; Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Luke 17:5\index{42 Luke:ch.  17, v. 
 5, pg. ``f30!}; Romans 1:16\index{45 Romans:ch.   1, v.  16, pg. ``f30!}{}-17\index{45 Romans:ch.   1, v.  16-17, pg.
``f30!}.}


\bigskip

14.2\ \ By this faith\index{Faith:believes Scripture to be word of God ``f10!}, a Christian believes
to\index{Scripture:to be:believed ``f10!} be true whatsoever is revealed in the Word, for\index{Scripture:word of
God:standard for:life ``f10!} the authority of God Himself speaking therein;\footnote{\ John 4:42\index{43 John:ch.  
4, v.  42, pg. ``f30!}; 1 Thessalonians 2:13\index{52 1 Thessalonians:ch.   2, v.  13, pg. ``f30!}; 1 John
5:10\index{62 1 John:ch.   5, v.  10, pg. ``f30!}; Acts 24:14\index{44 Acts:ch.  24, v.  14, pg. ``f30!}.} and acts
differently upon that which each particular passage thereof contains; yielding\index{God's
law:commandments:obedience:required ``f10!} obedience to the commands\index{God's law:commandments:ten
``f10!},\footnote{\ Romans 16:2\index{45 Romans:ch.  16, v.   2, pg. ``f30!}6\index{45 Romans:ch.  16, v.  26, pg.
``f30!}.} trembling at the threatenings\index{God's law:threatenings for disobedience ``f10!},\footnote{\ Isaiah
66:2\index{23 Isaiah:ch.  66, v.   2, pg. ``f30!}.} and embracing the promises\index{Salvation:promises made ``f10!} of
God for this life, and that which is to come.\footnote{\ Hebrews 11:13\index{58 Hebrews:ch.  11, v.  13, pg. ``f30!}; 1
Timothy 4:8\index{54 1 Timothy:ch.   4, v.   8, pg. ``f30!}.} But the principal\index{Faith:principal acts of ``f10!}
acts of saving faith\index{Faith:saving ``f10!} are accepting\index{Salvation:belief in Jesus by faith ``f10!},
receiving, and\index{Faith:instrument:for receiving Christ ``f10!} resting upon\index{Faith:resting on Christ ``f10!}
Christ alone for justification\index{Justification:justification ``f10!},
sanctification\index{Sanctification:sanctification ``f10!}, and eternal life\index{eternal life ``f10!}, by virtue of
the covenant of grace\index{Covenant:grace:grace ``f10!}.\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}; Acts 16:31\index{44 Acts:ch.  16, v.  31, pg. ``f30!}; Galatians
2:20\index{48 Galatians:ch.   2, v.  20, pg. ``f30!}; Acts 15:11\index{44 Acts:ch.  15, v.  11, pg. ``f30!}.}


\bigskip

14.3\ \ This faith is different in degrees\index{Faith:degrees of ``f10!}, weak or strong;\footnote{\ Hebrews
5:13-14\index{58 Hebrews:ch.   5, v.  13-14, pg. ``f30!}; Romans 4:19-20\index{45 Romans:ch.   4, v.  19-20, pg.
``f30!}; Matthew 6:30\index{40 Matthew:ch.   6, v.  30, pg. ``f30!}; Matthew 8:10\index{40 Matthew:ch.   8, v.  10, pg.
``f30!}.} may be often and many ways assailed\index{Faith:assailed ``f10!}, and weakened\index{Faith:weak ``f10!}, but
gets the victory\index{Faith:victorious ``f10!};\footnote{\ Luke 22:31-32\index{42 Luke:ch.  22, v.  31-32, pg.
``f30!}; Ephesians 6:16\index{49 Ephesians:ch.   6, v.  16, pg. ``f30!}; 1 John 5:4\index{62 1 John:ch.   5, v.   4,
pg. ``f30!}{}-5\index{62 1 John:ch.   5, v.   4-5, pg. ``f30!}.} growing up in\index{Assurance:growing ``f10!} many to
the attainment of a full assurance through\index{Assurance:of faith ``f10!} Christ,\footnote{\ Hebrews 6:11-12\index{58
Hebrews:ch.   6, v.  11-12, pg. ``f30!}; Hebrews 10:22\index{58 Hebrews:ch.  10, v.  22, pg. ``f30!}; Colossians
2:2\index{51 Colossians:ch.   2, v.   2, pg. ``f30!}.} who is both the author\index{Jesus:author of a believer's faith
``f10!} and finisher\index{Jesus:finisher of a believer's faith ``f10!} of our faith.\footnote{\ Hebrews 12:2\index{58
Hebrews:ch.  12, v.   2, pg. ``f30!}.}

\section{Chapter 15 -- Of Repentance unto Life}
\index{Repentance:unto life ``f10!}15.1\ \ Repentance\index{Repentance:unto life ``f10!} unto life is an evangelical
grace\index{Grace:evangelical ``f10!},\footnote{\ Zechariah 12:10\index{38 Zechariah:ch.  12, v.  10, pg. ``f30!}; Acts
11:18\index{44 Acts:ch.  11, v.  18, pg. ``f30!}.} the doctrine whereof\index{Doctrine:Gospel ``f10!} is to be preached
by every\index{Preaching:call to repentance ``f10!} minister\index{Church officers:minsters ``f10!} of the
Gospel\index{Gospel:gospel ``f10!}, as well as that of faith\index{Faith:in Jesus Christ ``f10!} in
Christ.\footnote{\ Luke 24:47\index{42 Luke:ch.  24, v.  47, pg. ``f30!}; Mark 1:15\index{41 Mark:ch.   1, v.  15, pg.
``f30!}; Acts 20:21\index{44 Acts:ch.  20, v.  21, pg. ``f30!}.}


\bigskip

15.2\ \ By it, a sinner, out of the sight and sense not only of the danger\index{Repentance:apprehension of danger of
sin ``f10!}, but also of the filthiness\index{Confession:of sin ``f10!} and odiousness\index{Sin:odious before God
``f10!} of his sins, as contrary to the holy\index{God's law:holy ``f10!} nature and righteous law\index{God's
law:righteous requirements ``f10!} of God; and upon the apprehension\index{Repentance:apprehension of God's mercy
``f10!} of his mercy in\index{God's attributes:merciful ``f10!} Christ to such as are
penitent\index{Repentance:penitence ``f10!}, so grieves for\index{Sin:grief and hatred of ``f10!}, and
hates\index{Sin:to be hated ``f10!} his sins, as to turn from them all unto God,\footnote{\ Ezekiel 18:30-31\index{26
Ezekiel:ch.  18, v.  30-31, pg. ``f30!}; Ezekiel 36:31\index{26 Ezekiel:ch.  36, v.  31, pg. ``f30!}; Isaiah
30:22\index{23 Isaiah:ch.  30, v.  22, pg. ``f30!}; Psalm 51:4\index{19 Psalms:ch.  51, v.   4, pg. ``f30!}; Jeremiah
31:18\index{24 Jeremiah:ch.  31, v.  18, pg. ``f30!}{}-19\index{24 Jeremiah:ch.  31, v.  18-19, pg. ``f30!}; Joel
2:12\index{29 Joel:ch.   2, v.  12, pg. ``f30!}{}-13\index{29 Joel:ch.   2, v.  12-13, pg. ``f30!}; Amos 5:15\index{30
Amos:ch.   5, v.  15, pg. ``f30!}; Psalm 119:128\index{19 Psalms:ch. 119, v. 128, pg. ``f30!}; 2 Corinthians
7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}1\index{47 2 Corinthians:ch.   7, v.  11, pg. ``f30!}.}
purposing\index{Repentance:purposing to walk with God ``f10!} and endeavouring to walk with\index{God's
law:commandments:obedience:endeavoring ``f10!} Him in all the ways of His commandments\index{God's law:commandments:ten
``f10!}.\footnote{\ Psalm 119:6, 59, 106\index{19 Psalms:ch. 119, v.   6, 59, 106, pg. ``f30!}; Luke 1:6\index{42
Luke:ch.   1, v.   6, pg. ``f30!}; 2 Kings 23:25\index{12 2 Kings:ch.  23, v.  25, pg. ``f30!}.}


\bigskip

15.3\ \ Although repentance be\index{Repentance:required for pardon ``f10!} not to be rested in, as any satisfaction for
sin\index{Jesus:satisfied divine justice ``f10!}, or any cause of the pardon the\index{Forgiveness:God alone pardons
``f10!}reof,\footnote{\ Ezekiel 36:31\index{26 Ezekiel:ch.  36, v.  31, pg. ``f30!}{}-32\index{26 Ezekiel:ch.  36, v. 
31-32, pg. ``f30!}; Ezekiel 16:61-63\index{26 Ezekiel:ch.  16, v.  61-63, pg. ``f30!}.} which is the act of God's free
grace\index{Grace:God's free ``f10!} in Christ;\footnote{\ Hosea 14:2\index{28 Hosea:ch.  14, v.   2, pg. ``f30!},
4\index{28 Hosea:ch.  14, v.   2, 4, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45
Romans:ch.   3, v.  24, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}.} yet is it of such
necessity to\index{Repentance:necessity ``f10!} all sinners, that none may expect pardon without it.\footnote{\ Luke
13:3, 5\index{42 Luke:ch.  13, v.   3, 5, pg. ``f30!}; Acts 17:30-31\index{44 Acts:ch.  17, v.  30-31, pg. ``f30!}.}


\bigskip

15.4\ \ As there is no sin so small, but it deserves\index{Sin:deserves damnation ``f10!}
damnation\index{Punishment:everlasting:damnation ``f10!},\footnote{\ Romans 6:23\index{45 Romans:ch.   6, v.  23, pg.
``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}; Matthew
12:36\index{40 Matthew:ch.  12, v.  36, pg. ``f30!}.} so there is no sin so great, that it can bring damnation upon
those who truly\index{Repentance:for all sins ``f10!} repent.\footnote{\ Isaiah 55:7\index{23 Isaiah:ch.  55, v.   7,
pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}; Isaiah 1:16, 18\index{23 Isaiah:ch.   1, v.  16,
18, pg. ``f30!}.}


\bigskip

15.5\ \ Men ought not to content themselves with a general repentance, but\index{Repentance:for all sins ``f10!} it is
every man's duty to endeavour\index{Repentance:duty to endeavour ``f10!} to repent of his particular sins,
particularly.\footnote{\ Psalm 19:13\index{19 Psalms:ch.  19, v.  13, pg. ``f30!}; Luke 19:8\index{42 Luke:ch.  19, v. 
 8, pg. ``f30!}; 1 Timothy 1:13, 15\index{54 1 Timothy:ch.   1, v.  13, 15, pg. ``f30!}.}


\bigskip

15.6\ \ As every man is bound to make private confession\index{Confession:of sin ``f10!} of his sins to God, praying
for\index{Prayer:for pardon of sin ``f10!} the pardon the\index{Forgiveness:God alone pardons
``f10!}reof;\index{Forgiveness:beg pardon for ``f10!}\footnote{\ Psalm 51:4\index{19 Psalms:ch.  51, v.   4, pg.
``f30!}{}-5, 7, 9, 14\index{19 Psalms:ch.  51, v.   4-5, 7, 9, 14, pg. ``f30!}; Psalm 32:5\index{19 Psalms:ch.  32, v. 
 5, pg. ``f30!}{}-6\index{19 Psalms:ch.  32, v.   5-6, pg. ``f30!}.} upon which, and the
forsaking\index{Repentance:forsaking sin ``f10!} of them, he shall find mercy:\index{Mercy:on confession
``f10!}\footnote{\ Proverbs 28:13\index{20 Proverbs:ch.  28, v.  13, pg. ``f30!}; 1 John 1:9\index{62 1 John:ch.   1,
v.   9, pg. ``f30!}.} so, he that scandalizes\index{Scandalous behaviour:to be confessed ``f10!} his brother, or the
Church of Christ\index{Church:Christ's ``f10!}, ought to be willing, by a private or public confession, and
sorrow\index{Confession:sorrow for sin ``f10!} for his sin, to declare his\index{Repentance:public declaration ``f10!}
repentance to\index{Repentance:to those offended ``f10!} those that are offended,\index{Confession:to those offended
``f10!}\footnote{\ James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}; Luke 17:3-4\index{42 Luke:ch.  17, v.   3-4,
pg. ``f30!}; Joshua 7:19\index{06 Joshua:ch.   7, v.  19, pg. ``f30!}; Psalm 51.} who are thereupon to be
reconciled\index{Confession:to be reconciled to God ``f10!} to him, and in love to\index{Love toward God:for God
``f10!} receive him.\footnote{\ 2 Corinthians 2:8\index{47 2 Corinthians:ch.   2, v.   8, pg. ``f30!}.}

\section{Chapter 16 -- Of Good Works}
\index{Good works:definition ``f10!}16.1\ \ Good works\index{Good works:definition ``f10!} are only such as God has
commanded in His\index{God's law:commandments:ten ``f10!} holy Word\index{Scripture:holy ``f10!},\footnote{\ Micah
6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; Romans 12:2\index{45 Romans:ch.  12, v.   2, pg. ``f30!}; Hebrews
13:21\index{58 Hebrews:ch.  13, v.  21, pg. ``f30!}.} and not such as, without the warrant\index{Good works:not devised
by men without warrant ``f10!} thereof, are devised\index{Good works:not devised by men ``f10!} by men, out of blind
zeal, or\index{Good works:not out of blind zeal ``f10!} upon any pretence\index{pretence ``f10!} of good
intention\index{Good works:not by good intentions ``f10!}.\footnote{\ Matthew 15:9\index{40 Matthew:ch.  15, v.   9,
pg. ``f30!}; Isaiah 29:13\index{23 Isaiah:ch.  29, v.  13, pg. ``f30!}; 1 Peter 1:18\index{60 1 Peter:ch.   1, v.  18,
pg. ``f30!}; Romans 10:2\index{45 Romans:ch.  10, v.   2, pg. ``f30!}; John 16:2\index{43 John:ch.  16, v.   2, pg.
``f30!}; 1 Samuel 15:21\index{09 1 Samuel:ch.  15, v.  21, pg. ``f30!}{}-23\index{09 1 Samuel:ch.  15, v.  21-23, pg.
``f30!}.} 


\bigskip

16.2\ \ These good works\index{Good works:adorn profession of faith ``f10!}, done in obedience\index{God's
law:commandments:obedience:required ``f10!} to God's commandments\index{God's law:commandments:ten ``f10!}, are the
fruits\index{Good works:fruit of salvation ``f10!} and evidences of a true and lively\index{Faith:lively ``f10!}
faith:\footnote{\ James 2:18, 22\index{59 James:ch.   2, v.  18, 22, pg. ``f30!}.} and by them believers
manifest\index{Good works:believers manifest thankfulness ``f10!} their thankfulness\index{Prayer:thanking God
``f10!},\footnote{\ Psalm 116:12-13\index{19 Psalms:ch. 116, v.  12-13, pg. ``f30!}; 1 Peter 2:9\index{60 1 Peter:ch.  
2, v.   9, pg. ``f30!}.} strengthen their assurance\index{Assurance:strengthened ``f10!},\footnote{\ 1 John
2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}, 5\index{62 1 John:ch.   2, v.   3, 5, pg. ``f30!}; 2 Peter
1:5-10\index{61 2 Peter:ch.   1, v.   5-10, pg. ``f30!}.} edify\index{Good works:edify other Christians ``f10!} their
brethren\index{Believers:to build up fellow believers ``f10!},\footnote{\ 2 Corinthians 9:2\index{47 2 Corinthians:ch. 
 9, v.   2, pg. ``f30!}; Matthew 5:16\index{40 Matthew:ch.   5, v.  16, pg. ``f30!}.} adorn the
profession\index{Faith:profession of ``f10!} of the Gospel\index{Gospel:gospel ``f10!},\footnote{\ Titus 2:5,
9-12\index{56 Titus:ch.   2, v.   5, 9-12, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}.}
stop the mouths\index{Good works:stop mouths of adversaries ``f10!} of the adversaries,\footnote{\ 1 Peter
2:15\index{60 1 Peter:ch.   2, v.  15, pg. ``f30!}.} and glorify\index{Worship:glorifying God ``f10!} God,\footnote{\ 1
Peter 2:12\index{60 1 Peter:ch.   2, v.  12, pg. ``f30!}; Philippians 1:1\index{50 Philippians:ch.   1, v.   1, pg.
``f30!}1\index{50 Philippians:ch.   1, v.  11, pg. ``f30!}; John 15:8\index{43 John:ch.  15, v.   8, pg. ``f30!}.}
whose workmanship\index{Believers:God's workmanship ``f10!} they are, created in Christ\index{Conversion:created in
Christ ``f10!} Jesus thereunto;\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}0\index{49
Ephesians:ch.   2, v.  10, pg. ``f30!}.} that, having their fruit unto holiness\index{Believers:holy:made perfect at
death ``f10!}, they may have the end, eternal life\index{eternal life ``f10!}.\footnote{\ Romans 6:22\index{45
Romans:ch.   6, v.  22, pg. ``f30!}.}


\bigskip

16.3\ \ Their ability to do\index{Good works:only in Holy Spirit ``f10!} good works\index{Good works:adorn profession of
faith ``f10!} is not at all of themselves, but wholly from the Spirit of Christ\index{Holy Spirit:Spirit of Christ
``f10!}.\footnote{\ John 15:4-5\index{43 John:ch.  15, v.   4-5, pg. ``f30!}; Ezekiel 36:26\index{26 Ezekiel:ch.  36,
v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36, v.  26-27, pg. ``f30!}.} And that they may be enabled
thereunto\index{Good works:enabled by Holy Spirit ``f10!}, besides the graces\index{Grace:in believers:by Holy Spirit
``f10!} they have already received, there is required an actual influence of the same Holy Spirit\index{Holy
Spirit:Holy Spirit ``f10!}, to work in them to will and to do of His good pleasure\index{Providence:at God's pleasure
``f10!}:\footnote{\ Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Philippians 4:13\index{50
Philippians:ch.   4, v.  13, pg. ``f30!}; 2 Corinthians 3:5\index{47 2 Corinthians:ch.   3, v.   5, pg. ``f30!}.} yet
are they not hereupon to grow negligent, as\index{Good works:not be negligent ``f10!} if they were not bound to perform
any duty, unless\index{God's law:commandments:duties:required by God ``f10!} upon a special motion of the Spirit;
but\index{Holy Spirit:helps believers do good works ``f10!} they ought to be diligent in\index{Grace:diligent use of
means ``f10!} stirring up the grace of\index{Grace:in believers:to be stirred up ``f10!} God that is in
them.\footnote{\ Philippians 2:12\index{50 Philippians:ch.   2, v.  12, pg. ``f30!}; Hebrews 6:11-12\index{58
Hebrews:ch.   6, v.  11-12, pg. ``f30!}; 2 Peter 1:3, 5, 10-11\index{61 2 Peter:ch.   1, v.   3, 5, 10-11, pg. ``f30!};
Isaiah 64:7\index{23 Isaiah:ch.  64, v.   7, pg. ``f30!}; 2 Timothy 1:6\index{55 2 Timothy:ch.   1, v.   6, pg.
``f30!}; Acts 26:6-7\index{44 Acts:ch.  26, v.   6-7, pg. ``f30!}; Jude 20-21\index{65 Jude:v. 20-21, pg. ``f30!}.}


\bigskip

16.4\ \ They, who in their obedience\index{Salvation:not by:obedience ``f10!} attain to the greatest height which is
possible in this life, are so far from being able to supererogate\index{Good works:believers cannot supererogate (do
more than required) ``f10!}, and to do more than God requires, as\index{God's law:commandments:duties:required by God
``f10!} that they fall short\index{Sanctification:imperfection in believers ``f10!} of much which in duty they are
bound to do.\footnote{\ Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}; Nehemiah 13:22\index{16 Nehemiah:ch. 
13, v.  22, pg. ``f30!}; Job 9:2-3\index{18 Job:ch.   9, v.   2-3, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.  
5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}.}


\bigskip

16.5\ \ We cannot, by our best works, merit\index{Salvation:not by:works ``f10!} pardon of sin, or\index{Good works:not
merit pardon of sin ``f10!} eternal life\index{eternal life ``f10!} at the hand of God, by reason of the great
disproportion\index{Good works:great disproportion compared with glory ``f10!} that is between them and the glory
to\index{Resurrection:believers:into glory:state of ``f10!} come; and the infinite\index{God's
attributes:infinite:separation from creatures ``f10!} distance\index{Mankind:great distance between him and God ``f10!}
that is between us and God, whom, by them, we can neither profit, nor satisfy for\index{Jesus:satisfied divine justice
``f10!} the debt\index{Forgiveness:of debtors ``f10!} of our former sins,\footnote{\ Romans 3:2\index{45 Romans:ch.  
3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}; Romans 4:2, 4, 6\index{45 Romans:ch.   4, v.   2,
4, 6, pg. ``f30!}; Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}{}-9\index{49 Ephesians:ch.   2, v.  
8-9, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}{}-7\index{56 Titus:ch.   3, v.   5-7, pg.
``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}8\index{45 Romans:ch.   8, v.  18, pg. ``f30!}; Psalm
16:2\index{19 Psalms:ch.  16, v.   2, pg. ``f30!}; Job 22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}; Job
35:7-8\index{18 Job:ch.  35, v.   7-8, pg. ``f30!}.} but when we have done all we can, we have done but our duty, and
are\index{God's law:commandments:duties:required by God ``f10!} unprofitable servants\index{Good works:duty of
unprofitable servants ``f10!};\footnote{\ Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}.} and because, as they
are good, they proceed from His\index{Good works:only in Holy Spirit ``f10!} Spirit;\footnote{\ Galatians 5:2\index{48
Galatians:ch.   5, v.   2, pg. ``f30!}2-23\index{48 Galatians:ch.   5, v.  22-23, pg. ``f30!}.} and as they are wrought
by us, they are defiled, and\index{Good works:defiled ``f10!} mixed\index{Good works:mixed with weakness and
imperfection ``f10!} with so much weakness and\index{Sanctification:weaknesses and imperfections remain ``f10!}
imperfection, that they cannot endure the severity\index{Good works:cannot endure severity of God's judgement ``f10!}
of God's judgment\index{Judgement:severe ``f10!}.\footnote{\ Isaiah 64:6\index{23 Isaiah:ch.  64, v.   6, pg. ``f30!};
Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}; Romans
7:15, 18\index{45 Romans:ch.   7, v.  15, 18, pg. ``f30!}; Psalm 143:2\index{19 Psalms:ch. 143, v.   2, pg. ``f30!};
Psalm 130:3\index{19 Psalms:ch. 130, v.   3, pg. ``f30!}.} 


\bigskip

16.6\ \ Yet notwithstanding, the persons of believers being accepted through\index{Believers:accepted in Christ ``f10!}
Christ, their good works\index{Good works:adorn profession of faith ``f10!} also are accepted in Him\index{Good
works:accepted in Christ ``f10!},\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}; 1 Peter
2:5\index{60 1 Peter:ch.   2, v.   5, pg. ``f30!}; Exodus 28:38\index{02 Exodus:ch.  28, v.  38, pg. ``f30!}; Genesis
4:4\index{01 Genesis:ch.   4, v.   4, pg. ``f30!} with Hebrews 11:4\index{58 Hebrews:ch.  11, v.   4, pg. ``f30!}.} not
as though they were in this life wholly unblameable and unreproveable\index{Good works:not unreproveable ``f10!} in
God's sight;\footnote{\ Job 9:20\index{18 Job:ch.   9, v.  20, pg. ``f30!}; Psalm 143:2\index{19 Psalms:ch. 143, v.  
2, pg. ``f30!}.} but that He, looking upon them in His Son\index{Jesus:Son of God ``f10!}, is
pleased\index{Providence:at God's pleasure ``f10!} to accept and reward\index{Believers:rewarded by God ``f10!} that
which is sincere\index{Good works:sincere accepted by God ``f10!}, although accompanied with many
weaknesses\index{Sanctification:weaknesses and imperfections remain ``f10!} and
imperfections\index{Sanctification:imperfection in believers ``f10!}.\footnote{\ Hebrews 13:20-21\index{58 Hebrews:ch. 
13, v.  20-21, pg. ``f30!}; 2 Corinthians 8:1\index{47 2 Corinthians:ch.   8, v.   1, pg. ``f30!}2\index{47 2
Corinthians:ch.   8, v.  12, pg. ``f30!}; Hebrews 6:10\index{58 Hebrews:ch.   6, v.  10, pg. ``f30!}; Matthew
25:21\index{40 Matthew:ch.  25, v.  21, pg. ``f30!}, 23\index{40 Matthew:ch.  25, v.  21, 23, pg. ``f30!}.}


\bigskip

16.7\ \ Works done\index{Good works:not merit pardon of sin ``f10!} by unregenerate\index{Unbelievers:unregenerate
``f10!} men, although, for the matter of them, they may be things which God commands, and of\index{God's
law:commandments:obedience:not salvific ``f10!} good use both to themselves and others:\footnote{\ 2 Kings
10:30-31\index{12 2 Kings:ch.  10, v.  30-31, pg. ``f30!}; 1 Kings 21:27, 29\index{11 1 Kings:ch.  21, v.  27, 29, pg.
``f30!}; Philippians 1:1\index{50 Philippians:ch.   1, v.   1, pg. ``f30!}5-16, 18\index{50 Philippians:ch.   1, v. 
15-16, 18, pg. ``f30!}.} yet, because they proceed not\index{Good works:not from unregenerate ``f10!} from a
heart\index{Heart:new:purified ``f10!} purified by faith\index{Faith:false ``f10!};\footnote{\ Genesis 4:5\index{01
Genesis:ch.   4, v.   5, pg. ``f30!} with Hebrews 11:4\index{58 Hebrews:ch.  11, v.   4, pg. ``f30!}, 6\index{58
Hebrews:ch.  11, v.   4, 6, pg. ``f30!}.} nor are done in a right manner according to the Word;\index{Scripture:word of
God:standard for:life ``f10!}\footnote{\ 1 Corinthians 13:3\index{46 1 Corinthians:ch.  13, v.   3, pg. ``f30!}; Isaiah
1:12\index{23 Isaiah:ch.   1, v.  12, pg. ``f30!}.} nor to a right end, the glory of God;\index{Good works:to be done
for the glory of God ``f10!}\footnote{\ Matthew 6:2, 5, 16\index{40 Matthew:ch.   6, v.   2, 5, 16, pg. ``f30!}.} they
are therefore sinful, and cannot please God\index{Good works:cannot please God ``f10!}, or make a man meet to receive
grace\index{Grace:received from God ``f10!} from God.\footnote{\ Haggai 2:14\index{37 Haggai:ch.   2, v.  14, pg.
``f30!}; Titus 1:15\index{56 Titus:ch.   1, v.  15, pg. ``f30!}; Amos 5:22-23\index{30 Amos:ch.   5, v.  22-23, pg.
``f30!}; Hosea 1:4\index{28 Hosea:ch.   1, v.   4, pg. ``f30!}; Romans 9:16\index{45 Romans:ch.   9, v.  16, pg.
``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.} And yet, their neglect of them\index{Good works:not be
negligent ``f10!} is more sinful, and displeasing unto God.\footnote{\ Psalm 14:4\index{19 Psalms:ch.  14, v.   4, pg.
``f30!}; Psalm 36:3\index{19 Psalms:ch.  36, v.   3, pg. ``f30!}; Job 21:14\index{18 Job:ch.  21, v.  14, pg.
``f30!}{}-15\index{18 Job:ch.  21, v.  14-15, pg. ``f30!}; Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg.
``f30!}{}-43\index{40 Matthew:ch.  25, v.  41-43, pg. ``f30!}, 45\index{40 Matthew:ch.  25, v.  41-43, 45, pg. ``f30!};
Matthew 23:23\index{40 Matthew:ch.  23, v.  23, pg. ``f30!}.}

\section{Chapter 17 -- Of the Perseverance of the Saints}
\index{Perseverance:of the saints ``f10!}\index{Believers:saints ``f10!}17.1\ \ They, whom God has accepted in
His\index{Believers:accepted in Christ ``f10!} Beloved\index{Jesus:beloved of the Father ``f10!}, effectually
called\index{Effectual call:calling ``f10!}, and sanctified\index{Sanctification:sanctification ``f10!} by His Spirit,
can\index{Holy Spirit:applies:sanctification ``f10!} neither totally\index{Assurance:never fall from being saved
``f10!}, nor finally, fall away from the state of grace\index{Grace:state:fall from ``f10!}: but shall certainly
persevere\index{Perseverance:of the saints ``f10!} therein to the end, and be eternally\index{Salvation:eternal for
those effectually called ``f10!} saved.\footnote{\ Philippians 1:6\index{50 Philippians:ch.   1, v.   6, pg. ``f30!}; 2
Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}; John 10:28\index{43 John:ch.  10, v.  28, pg.
``f30!}{}-29\index{43 John:ch.  10, v.  28-29, pg. ``f30!}; 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}; 1
Peter 1:5\index{60 1 Peter:ch.   1, v.   5, pg. ``f30!}, 9\index{60 1 Peter:ch.   1, v.   5, 9, pg. ``f30!}.}


\bigskip

17.2\ \ This perseverance\index{Perseverance:of the saints ``f10!} of the saints\index{Believers:saints ``f10!} depends
not upon their\index{Perseverance:depends not on person's will ``f10!} own free will, but upon the
immutability\index{God's attributes:immutable ``f10!} of the decree\index{Predestination:eternal decrees ``f10!} of
election\index{Election:decree of ``f10!}, flowing from the free and unchangeable\index{God's attributes:unchangeable
``f10!} love of God\index{Love toward God:for God ``f10!} the Father\index{Father:God the ``f10!};\footnote{\ 2 Timothy
2:18-19\index{55 2 Timothy:ch.   2, v.  18-19, pg. ``f30!}; Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg.
``f30!}.} upon the efficacy\index{Perseverance:enabled by Holy Spirit ``f10!} of the merit
and\index{Jesus:obedience:merit of ``f10!} intercession\index{Jesus:in heaven:intercession made for all believers
``f10!} of Jesus Christ;\footnote{\ Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}0\index{58 Hebrews:ch. 
10, v.  10, pg. ``f30!}, 14\index{58 Hebrews:ch.  10, v.  10, 14, pg. ``f30!}; Hebrews 13:20-21\index{58 Hebrews:ch. 
13, v.  20-21, pg. ``f30!}; Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!}{}-15\index{58 Hebrews:ch.   9,
v.  12-15, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}3\index{45 Romans:ch.   8, v.  33, pg.
``f30!} to the end, John 17:11, 24\index{43 John:ch.  17, v.  11, 24, pg. ``f30!}; Luke 22:32\index{42 Luke:ch.  22, v.
 32, pg. ``f30!}; Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}.} the abiding of\index{Holy Spirit:abiding
in believers ``f10!} the Spirit, and of the seed of God\index{Believers:seed of God planted within ``f10!} within
them;\footnote{\ John 14:16-17\index{43 John:ch.  14, v.  16-17, pg. ``f30!}; 1 John 2:2\index{62 1 John:ch.   2, v.  
2, pg. ``f30!}7\index{62 1 John:ch.   2, v.  27, pg. ``f30!}; 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}.}
and the nature of the covenant of grace\index{Covenant:grace:grace ``f10!}:\footnote{\ Jeremiah 32:40\index{24
Jeremiah:ch.  32, v.  40, pg. ``f30!}.} from all which arises also the certainty and
infallibility\index{Covenant:grace:infallible ``f10!} thereof.\footnote{\ John 10:28\index{43 John:ch.  10, v.  28, pg.
``f30!}; 2 Thessalonians 3:3\index{53 2 Thessalonians:ch.   3, v.   3, pg. ``f30!}; 1 John 2:19\index{62 1 John:ch.  
2, v.  19, pg. ``f30!}.}


\bigskip

17.3\ \ Nevertheless, they may, through the temptations\index{Temptation:temptation ``f10!} of Satan and
of\index{Satan:tempter ``f10!} the world, the prevalence of corruption remaining\index{Sanctification:remnants of
corruption remain ``f10!} in them, and the neglect of the means\index{Love for others:neighbour as self ``f10!} of
their preservation\index{Perseverance:means neglected ``f10!}, fall into grievous sins\index{Temptation:fall into
grievous sins ``f10!};\footnote{\ Matthew 26:70, 72, 74\index{40 Matthew:ch.  26, v.  70, 72, 74, pg. ``f30!}.} and,
for a time, continue therein:\footnote{\ Psalm 51 title and 14.} whereby they incur\index{Believers:may experience
God's displeasure for sin ``f10!} God's displeasure,\footnote{\ Isaiah 64:5, 7, 9\index{23 Isaiah:ch.  64, v.   5, 7,
9, pg. ``f30!}; 2 Samuel 11:27\index{10 2 Samuel:ch.  11, v.  27, pg. ``f30!}.} and grieve\index{Sin:grieves the Holy
Spirit ``f10!} His Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!},\footnote{\ Ephesians 4:3\index{49 Ephesians:ch.  
4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}.} come to be deprived of some measure of their
graces and\index{Grace:withheld ``f10!} comforts\index{Temptation:comforts removed ``f10!},\footnote{\ Psalm
51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}, 10, 12\index{19 Psalms:ch.  51, v.   8, 10, 12, pg. ``f30!};
Revelation 2:4\index{66 Revelation:ch.   2, v.   4, pg. ``f30!}; Song of Solomon 5:2-4, 6\index{22 Song of Solomon:ch. 
 5, v.   2-4, 6, pg. ``f30!}.} have their hearts hardened\index{Heart:hardened ``f10!},\footnote{\ Isaiah
63:17\index{23 Isaiah:ch.  63, v.  17, pg. ``f30!}; Mark 6:52\index{41 Mark:ch.   6, v.  52, pg. ``f30!}; Mark
16:14\index{41 Mark:ch.  16, v.  14, pg. ``f30!}.} and their consciences\index{Conscience:wounded ``f10!}
wounded,\footnote{\ Psalm 32:3-4\index{19 Psalms:ch.  32, v.   3-4, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.
  8, pg. ``f30!}.} hurt and scandalize\index{Scandalous behaviour:scandalize others ``f10!} others,\footnote{\ 2 Samuel
12:14\index{10 2 Samuel:ch.  12, v.  14, pg. ``f30!}.} and bring temporal judgments upon\index{Judgement:temporal
``f10!} themselves.\footnote{\ Psalm 89:31-32\index{19 Psalms:ch.  89, v.  31-32, pg. ``f30!}; 1 Corinthians
11:32\index{46 1 Corinthians:ch.  11, v.  32, pg. ``f30!}.}

\paragraph[Chapter 18 {}-- Of the Assurance of Grace and Salvation]{Chapter 18 -- Of the Assurance of Grace and
Salvation}
\index{Assurance:of state of grace ``f10!}\index{Salvation:assurance of ``f10!}18.1\ \ Although
hypocrites\index{hypocrisy ``f10!} and other unregenerate\index{Unbelievers:unregenerate ``f10!} men may vainly
deceive\index{Deceitfulness:of self ``f10!} themselves with false hopes\index{Hope:false ``f10!}, and carnal
presumptions\index{presumption ``f10!} of being in the favour\index{Unbelievers:presumptuous about favour with God
``f10!} of God, and estate of salvation\index{Salvation:belief in Jesus by faith ``f10!};\footnote{\ Job
8:13-14\index{18 Job:ch.   8, v.  13-14, pg. ``f30!}; Micah 3:11\index{33 Micah:ch.   3, v.  11, pg. ``f30!};
Deuteronomy 29:19\index{05 Deuteronomy:ch.  29, v.  19, pg. ``f30!}; John 8:41\index{43 John:ch.   8, v.  41, pg.
``f30!}.} which hope of theirs shall perish:\footnote{\ Matthew 7:22\index{40 Matthew:ch.   7, v.  22, pg.
``f30!}{}-23\index{40 Matthew:ch.   7, v.  22-23, pg. ``f30!}.} yet such as truly believe in the Lord Jesus, and love
Him\index{Love toward God:for God ``f10!} in sincerity\index{Love:of Christ ``f10!}, endeavouring to walk in
all\index{Believers:strive to walk in good conscience ``f10!} good conscience\index{Conscience:good, endeavour for
``f10!} before Him, may, in this life, be certainly assured that\index{Assurance:of state of grace ``f10!} they are in
the state of grace\index{Grace:state:believers in ``f10!},\footnote{\ 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg.
``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}4\index{62 1 John:ch.   3, v.  14, pg. ``f30!}, 18-19,
21, 24\index{62 1 John:ch.   3, v.  14, 18-19, 21, 24, pg. ``f30!}; John 5:13\index{43 John:ch.   5, v.  13, pg.
``f30!}.} and may rejoice in the hope of\index{Hope:of glory of God, rejoice in ``f10!} the glory of God,
which\index{Resurrection:believers:into glory:everlasting ``f10!} hope shall\index{Hope:certain ``f10!} never make them
ashamed\index{Believers:never made ashamed by God ``f10!}.\footnote{\ Romans 5:2\index{45 Romans:ch.   5, v.   2, pg.
``f30!}, 5\index{45 Romans:ch.   5, v.   2, 5, pg. ``f30!}.}


\bigskip

18.2\ \ This certainty is not a bare conjectural\index{Assurance:not bare conjecture ``f10!} and probable
persuasion\index{Assurance:not probable persuasion ``f10!}, grounded upon a fallible hope;\index{Hope:fallible
``f10!}\footnote{\ Hebrews 6:11, 19\index{58 Hebrews:ch.   6, v.  11, 19, pg. ``f30!}.} but an
infallible\index{Assurance:infallible ``f10!} assurance of faith\index{Assurance:of faith ``f10!},
founded\index{Assurance:founded on God's promises ``f10!} upon the divine\index{Salvation:founded on divine truth
``f10!} truth of the promises\index{Salvation:promises made ``f10!} of salvation\index{Salvation:belief in Jesus by
faith ``f10!},\footnote{\ Hebrews 6:17\index{58 Hebrews:ch.   6, v.  17, pg. ``f30!}{}-18\index{58 Hebrews:ch.   6, v. 
17-18, pg. ``f30!}.} the inward evidence\index{Assurance:inward evidence ``f10!} of those graces\index{Grace:in
believers:by Holy Spirit ``f10!} unto which these promises\index{Assurance:promises of God ``f10!} are
made,\footnote{\ 2 Peter 1:4-5, 10-11\index{61 2 Peter:ch.   1, v.   4-5, 10-11, pg. ``f30!}; 1 John 2:3\index{62 1
John:ch.   2, v.   3, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}4\index{62 1 John:ch.   3, v.
 14, pg. ``f30!}; 2 Corinthians 1:1\index{47 2 Corinthians:ch.   1, v.   1, pg. ``f30!}2\index{47 2 Corinthians:ch.  
1, v.  12, pg. ``f30!}.} the testimony of the Spirit\index{Assurance:Holy Spirit witnesses with our spirits ``f10!} of
adoption\index{Adoption:in salvation ``f10!} witnessing with our spirits\index{Conversion:Holy Spirit:bears witness to
our spirits ``f10!} that we are the children of God:\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg.
``f30!}5\index{45 Romans:ch.   8, v.  15, pg. ``f30!}{}-16\index{45 Romans:ch.   8, v.  15-16, pg. ``f30!}.} which
Spirit\index{Holy Spirit:earnest of believers' inheritance ``f10!} is the earnest
of\index{Believers:inheritance:earnest of ``f10!} our inheritance\index{Believers:inheritance:everlasting ``f10!},
whereby we are sealed to the\index{Salvation:sealed to the day of redemption ``f10!} day of
redemption\index{Redemption:day of ``f10!}.\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}3\index{49 Ephesians:ch.   1, v.  13, pg. ``f30!}{}-14\index{49 Ephesians:ch.   1, v.  13-14, pg. ``f30!};
Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}; 2
Corinthians 1:21-22\index{47 2 Corinthians:ch.   1, v.  21-22, pg. ``f30!}.}

18.3\ \ This infallible\index{Assurance:infallible ``f10!} assurance does\index{Assurance:believer may await ``f10!} not
so belong to the essence of faith\index{Faith:essence of ``f10!}, but that a true believer may wait long, and conflict
with many difficulties\index{Assurance:believer may experience difficulties ``f10!} before he be partaker of
it:\footnote{\ 1 John 5:13\index{62 1 John:ch.   5, v.  13, pg. ``f30!}; Isaiah 50:10\index{23 Isaiah:ch.  50, v.  10,
pg. ``f30!}; Mark 9:24\index{41 Mark:ch.   9, v.  24, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18,
pg. ``f30!}; Psalm 77:1-12\index{19 Psalms:ch.  77, v.   1-12, pg. ``f30!}.} yet, being enabled
by\index{Assurance:enabled by Holy Spirit ``f10!} the Spirit to\index{Revelation:Word and Spirit ``f10!} know the
things which are freely\index{Salvation:freely offered ``f10!} given him of God, he may without extraordinary
revelation\index{Revelation:extraordinary ``f10!}, in the right use of ordinary means, attain thereunto.\footnote{\ 1
Corinthians 2:12\index{46 1 Corinthians:ch.   2, v.  12, pg. ``f30!}; 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg.
``f30!}3\index{62 1 John:ch.   4, v.  13, pg. ``f30!}; Hebrews 6:11-12\index{58 Hebrews:ch.   6, v.  11-12, pg.
``f30!}; Ephesians 3:17\index{49 Ephesians:ch.   3, v.  17, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  17-19, pg.
``f30!}.} And therefore it is the duty of everyone\index{Election:duty to make calling sure ``f10!} to give all
diligence to\index{Election:diligence to make calling sure ``f10!} make his calling and\index{Effectual call:elect
``f10!} election sure;\footnote{\ 2 Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}.} that thereby his heart
may be enlarged\index{Heart:enlarged in peace and joy ``f10!} in peace\index{Assurance:peace and joy ``f10!} and joy
in\index{Joy:in Holy Spirit ``f10!} the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, in love and\index{Love toward
God:for God ``f10!} thankfulness\index{Prayer:thanking God ``f10!} to God, and in strength\index{Sanctification:supply
of strength from Holy Spirit ``f10!} and cheerfulness\index{cheerfulness ``f10!} in the duties of obedience,
the\index{God's law:commandments:obedience:required ``f10!} proper fruits\index{Good works:fruit of salvation ``f10!}
of this assurance:\index{Assurance:not to be used to justify sin ``f10!}\footnote{\ Romans 5:1\index{45 Romans:ch.   5,
v.   1, pg. ``f30!}{}-2\index{45 Romans:ch.   5, v.   1-2, pg. ``f30!}, 5\index{45 Romans:ch.   5, v.   1-2, 5, pg.
``f30!}; Romans 14:17\index{45 Romans:ch.  14, v.  17, pg. ``f30!}; Romans 15:13\index{45 Romans:ch.  15, v.  13, pg.
``f30!}; Ephesians 1:3-4\index{49 Ephesians:ch.   1, v.   3-4, pg. ``f30!}; Psalm 4:6-7\index{19 Psalms:ch.   4, v.  
6-7, pg. ``f30!}; Psalm 119:32\index{19 Psalms:ch. 119, v.  32, pg. ``f30!}.} so far is it from inclining men to
looseness.\footnote{\ 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}; Romans 6:1-2\index{45 Romans:ch.  
6, v.   1-2, pg. ``f30!}; Titus 2:11-12, 14\index{56 Titus:ch.   2, v.  11-12, 14, pg. ``f30!}; 2 Corinthians
7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!},
12\index{45 Romans:ch.   8, v.   1, 12, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg.
``f30!}{}-3\index{62 1 John:ch.   3, v.   2-3, pg. ``f30!}; Psalm 130:4\index{19 Psalms:ch. 130, v.   4, pg. ``f30!}; 1
John 1:6-7\index{62 1 John:ch.   1, v.   6-7, pg. ``f30!}.}


\bigskip

18.4\ \ True believers may\index{Assurance:believer may not have at all times ``f10!} have the assurance of their
salvation\index{Salvation:belief in Jesus by faith ``f10!} divers ways shaken\index{Assurance:shaken at times ``f10!},
diminished, and intermitted; as, by negligence\index{Assurance:negligence in preserving ``f10!} in
preserving\index{Perseverance:means neglected ``f10!} of it, by falling\index{God's law:broken daily by believers
``f10!} into some special sin, which wounds the conscience\index{Conscience:wounded ``f10!} and grieves
the\index{Sin:grieves the Holy Spirit ``f10!} Spirit; by some sudden\index{Temptation:sudden ``f10!} or
vehement\index{Temptation:vehement ``f10!} temptation\index{Temptation:temptation ``f10!}, by God's withdrawing the
light of His countenance\index{Believers:chastened by God ``f10!}, and suffering even such as fear Him\index{Fear:of
God:reverence ``f10!} to walk in darkness\index{darkness, spiritual ``f10!} and to have no light:\footnote{\ Song of
Solomon 5:2-3, 6\index{22 Song of Solomon:ch.   5, v.   2-3, 6, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.  
8, pg. ``f30!}, 12\index{19 Psalms:ch.  51, v.   8, 12, pg. ``f30!}, 14\index{19 Psalms:ch.  51, v.   8, 12, 14, pg.
``f30!}; Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg.
``f30!}{}-31\index{49 Ephesians:ch.   4, v.  30-31, pg. ``f30!}; Psalm 77:1-10\index{19 Psalms:ch.  77, v.   1-10, pg.
``f30!}; Matthew 26:69-72\index{40 Matthew:ch.  26, v.  69-72, pg. ``f30!}; Psalm 31:22\index{19 Psalms:ch.  31, v. 
22, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}; Isaiah 50:10\index{23 Isaiah:ch.  50,
v.  10, pg. ``f30!}.} yet are they never so utterly destitute\index{Assurance:never destitute ``f10!} of that seed of
God\index{Believers:seed of God planted within ``f10!}, and life of faith\index{Faith:living by ``f10!}, that love of
Christ\index{Love:of Christ ``f10!} and the brethren\index{Believers:to build up fellow believers ``f10!}, that
sincerity of\index{Heart:sincerity of ``f10!} heart, and conscience of\index{Believers:desire to obey God ``f10!} duty,
out\index{God's law:commandments:duties:required by God ``f10!} of which, by the operation\index{Holy
Spirit:operations:saving ``f10!} of the Spirit, this assurance may, in due time, be revived\index{Assurance:revived
``f10!};\footnote{\ 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}; Luke 22:32\index{42 Luke:ch.  22, v.  32,
pg. ``f30!}; Job 13:15\index{18 Job:ch.  13, v.  15, pg. ``f30!}; Psalm 73:15\index{19 Psalms:ch.  73, v.  15, pg.
``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}, 12\index{19 Psalms:ch.  51, v.   8, 12, pg. ``f30!};
Isaiah 50:10\index{23 Isaiah:ch.  50, v.  10, pg. ``f30!}.} and by the which, in the mean time, they are supported
from\index{Assurance:Holy Spirit keeps believer from utter despair ``f10!} utter despair\index{despair
``f10!}.\footnote{\ Micah 7:7\index{33 Micah:ch.   7, v.   7, pg. ``f30!}{}-9\index{33 Micah:ch.   7, v.   7-9, pg.
``f30!}; Jeremiah 32:40\index{24 Jeremiah:ch.  32, v.  40, pg. ``f30!}; Isaiah 54:7-10\index{23 Isaiah:ch.  54, v.  
7-10, pg. ``f30!}; Psalm 22:1\index{19 Psalms:ch.  22, v.   1, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.  
1-18, pg. ``f30!}.}

\section{Chapter 19 -- Of the Law of God}
19.1\ \ God gave to Adam a law\index{Adam:given a law ``f10!}, as a covenant of works\index{Covenant:works:works
``f10!}, by which He bound him and all his posterity\index{Sin:original:all mankind descending from Adam ``f10!} to
personal, entire, exact, and perpetual obedience\index{Covenant:works:perpetual obedience required ``f10!}; promised
life\index{Covenant:first:promised life ``f10!} upon the fulfilling, and\index{Covenant:works:conditional on perfect
obedience ``f10!} threatened death\index{Death:punishment for sin ``f10!} upon the breach of it\index{God's
law:breach:death threatened ``f10!}: and endued him\index{Mankind:created:attributes ``f10!} with power and ability to
keep\index{Covenant:works:Adam:original ability and power to keep ``f10!} it.\footnote{\ Genesis 1:2\index{01
Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}{}-27\index{01 Genesis:ch.   1, v. 
26-27, pg. ``f30!} with Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Romans 2:1\index{45 Romans:ch.   2,
v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}; Romans 10:5\index{45 Romans:ch.  10, v.   5,
pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!},
19\index{45 Romans:ch.   5, v.  12, 19, pg. ``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!},
12\index{48 Galatians:ch.   3, v.  10, 12, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg.
``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}; Job 28:28\index{18 Job:ch.  28, v.  28, pg. ``f30!}.}


\bigskip

19.2\ \ This law, after his fall\index{Sin:original:first transgression (of Adam) ``f10!}, continued to be a perfect
rule of righteousness, and, as\index{God's law:righteous requirements ``f10!} such, was delivered by\index{God's
law:commandments:delivered on Sinai ``f10!} God upon Mount Sinai\index{Sinai, mount ``f10!}, in ten\index{God's
law:commandments:ten ``f10!} commandments, and written in two tables\index{God's law:commandments:two tables
``f10!}:\footnote{\index{59 James:ch.   1, v.  25, pg.
``f30!}{,
10-12}\index{59 James:ch.   2, v.   8, 10-12, pg. ``f30!}\index{45 Romans:ch. 
13, v.   8, pg. ``f30!}\index{45 Romans:ch.  13, v.   8-9, pg.
``f30!}\index{05 Deuteronomy:ch.   5, v.  32, pg.
``f30!}\index{05 Deuteronomy:ch.  10, v.   4, pg.
``f30!}
the four first commandment\index{God's law:commandments:first ``f10!}s\index{God's law:commandments:first four, duty to
God ``f10!} containing our duty towards\index{God's law:commandments:duties:required by God ``f10!} God; and the other
six\index{Sabbath:six days:for own affairs ``f10!} our duty to man\index{God's law:commandments:duties:to others
``f10!}.\footnote{\ Matthew 22:37-40\index{40 Matthew:ch.  22, v.  37-40, pg. ``f30!}.}


\bigskip

19.3\ \ Beside this law, commonly called moral\index{God's law:commandments:moral ``f10!}, God was
pleased\index{Providence:at God's pleasure ``f10!} to give to the people of Israel\index{Israel ``f10!}, as a
church\index{Church:OT ``f10!} under age, ceremonial\index{Ceremonial law:fore-signifying Christ ``f10!} laws,
containing several typical\index{types, OT ``f10!} ordinances, partly\index{Covenant:OT ordinances:signified Christ
``f10!} of worship\index{Worship:of God ``f10!}, prefiguring\index{Jesus:prefigured in ceremonial ordinances ``f10!}
Christ, His graces\index{Jesus:full of grace ``f10!}, actions, sufferings, and\index{Jesus:death:sufferings and death
``f10!} benefits;\footnote{\ Hebrews chap. 9; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}; Galatians
4:1-3\index{48 Galatians:ch.   4, v.   1-3, pg. ``f30!}; Colossians 2:17\index{51 Colossians:ch.   2, v.  17, pg.
``f30!}.} and partly holding\index{God's law:benefits ``f10!} forth divers instructions of moral duties.\index{God's
law:commandments:duties:required by God ``f10!}\footnote{\ 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7,
pg. ``f30!}; 2 Corinthians 6:17\index{47 2 Corinthians:ch.   6, v.  17, pg. ``f30!}; Jude 23\index{65 Jude:v.  23, pg.
``f30!}.} All which ceremonial laws are now abrogated\index{Ceremonial law:abrogated ``f10!}, under the new
testament\index{Scripture:New Testament ``f10!}.\footnote{\ Colossians 2:14, 16-17\index{51 Colossians:ch.   2, v.  14,
16-17, pg. ``f30!}; Daniel 9:27\index{27 Daniel:ch.   9, v.  27, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2,
v.   1, pg. ``f30!}5-16\index{49 Ephesians:ch.   2, v.  15-16, pg. ``f30!}.}


\bigskip

19.4\ \ To them also, as a body politic, He gave sundry judicial laws, which expired\index{God's law:judicial:expired
``f10!} together with the State of that people; not obliging any other now, further than the general equity
thereof\index{God's law:judicial:apply equity today ``f10!} may require.\footnote{\ Exodus chap. 21; Exodus
22:1-29\index{02 Exodus:ch.  22, v.   1-29, pg. ``f30!}; Genesis 49:10\index{01 Genesis:ch.  49, v.  10, pg. ``f30!}
with 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!};
Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!} with Matthew 5:38-39\index{40 Matthew:ch.   5, v.  38-39,
pg. ``f30!}; 1 Corinthians 9:8-10\index{46 1 Corinthians:ch.   9, v.   8-10, pg. ``f30!}.}


\bigskip

19.5\ \ The moral\index{God's law:commandments:moral ``f10!} law does for ever bind all\index{God's law:binding on all
men ``f10!}, as well justified\index{Justification:justified ``f10!} persons as others, to the obedience
thereof\index{God's law:commandments:obedience:required ``f10!};\footnote{\ Romans 13:8\index{45 Romans:ch.  13, v.  
8, pg. ``f30!}{}-10\index{45 Romans:ch.  13, v.   8-10, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2,
pg. ``f30!}; 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}{}-4, 7-8\index{62 1 John:ch.   2, v.   3-4, 7-8,
pg. ``f30!}.} and that, not only in regard of the matter contained in it, but also in respect of the authority of God
the Creator\index{God's attributes:creator:of all things ``f10!}, who gave it:\footnote{\ James 2:10\index{59 James:ch.
  2, v.  10, pg. ``f30!}{}-11\index{59 James:ch.   2, v.  10-11, pg. ``f30!}.} neither does Christ, in the
Gospel\index{Gospel:gospel ``f10!}, any way dissolve, but much strengthen this obligation\index{God's law:obligation to
obey:strengthened by Gospel ``f10!}.\footnote{\ Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg.
``f30!}{}-19\index{40 Matthew:ch.   5, v.  17-19, pg. ``f30!}; James 2:8\index{59 James:ch.   2, v.   8, pg. ``f30!};
Romans 3:31\index{45 Romans:ch.   3, v.  31, pg. ``f30!}.}


\bigskip

19.6\ \ Although true believers\index{Covenant:works:believers not under ``f10!} be not under the law, as a covenant of
works\index{Covenant:works:works ``f10!}, to be thereby justified\index{Justification:justified ``f10!}, or
condemned\index{Covenant:works:believers delivered from ``f10!};\footnote{\ Romans 6:14\index{45 Romans:ch.   6, v. 
14, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Galatians 3:13\index{48 Galatians:ch. 
 3, v.  13, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4,
v.   4-5, pg. ``f30!}; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}9\index{44 Acts:ch.  13, v.  39, pg.
``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}.} yet is it of great use to them, as well as to
others; in that, as a rule of life\index{God's law:moral:rule for life ``f10!} informing them of the will of God, and
their duty, it\index{God's law:commandments:duties:required by God ``f10!} directs, and binds them\index{God's
law:binding on all men ``f10!} to walk accordingly;\footnote{\ Romans 7:12\index{45 Romans:ch.   7, v.  12, pg.
``f30!}, 22, 25\index{45 Romans:ch.   7, v.  12, 22, 25, pg. ``f30!}; Psalm 119:4-6\index{19 Psalms:ch. 119, v.   4-6,
pg. ``f30!}; 1 Corinthians 7:19\index{46 1 Corinthians:ch.   7, v.  19, pg. ``f30!}; Galatians 5:1\index{48
Galatians:ch.   5, v.   1, pg. ``f30!}4, 16, 18-23\index{48 Galatians:ch.   5, v.  14, 16, 18-23, pg. ``f30!}.}
discovering also the sinful pollutions\index{Sin:original:corrupt nature:polluted ``f10!} of their nature, hearts, and
lives\index{Heart:polluted ``f10!};\footnote{\ Romans 7:7\index{45 Romans:ch.   7, v.   7, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}.} so as,
examining\index{God's law:examine self against ``f10!} themselves thereby, they may come to further
conviction\index{God's law:conviction of sin furthered ``f10!} of, humiliation for\index{Humility:before God ``f10!},
and hatred against\index{Sin:to be hated ``f10!} sin;\footnote{\ James 1:23-25\index{59 James:ch.   1, v.  23-25, pg.
``f30!}; Romans 7:9, 14, 24\index{45 Romans:ch.   7, v.   9, 14, 24, pg. ``f30!}.} together with a clearer sight of the
need they have of Christ, and the perfection of His obedience\index{Jesus:obedience:to God's law
``f10!}.\footnote{\index{48 Galatians:ch.   3, v.  24, pg.
``f30!}\index{45 Romans:ch.   7, v.  24-25, pg.
``f30!}\index{45 Romans:ch.   8, v.   3, pg.
``f30!} It is
likewise of use to the regenerate\index{Regeneration:regeneration ``f10!}, to restrain their corruptions,
in\index{God's law:restrains corruptions ``f10!} that it forbids\index{God's law:forbids sin ``f10!}
sin:\footnote{\index{59 James:ch.   2, v.  11, pg.
``f30!}\index{19 Psalms:ch. 119, v. 101, 104, 128, pg.
``f30!} of it
serve to show what even their sins deserve\index{Sin:deserves damnation ``f10!}; and what afflictions, in\index{God's
law:breach:define afflictions imposed on sin ``f10!} this life, they may expect for them, although
freed\index{Believers:freed from:curse of the law ``f10!} from the curse thereof\index{Curse:of the law ``f10!}
threatened in\index{God's law:commandments:threatenings for disobedience ``f10!} the law.\footnote{\ Ezra
9:13-14\index{15 Ezra:ch.   9, v.  13-14, pg. ``f30!}; Psalm 89:30-34\index{19 Psalms:ch.  89, v.  30-34, pg. ``f30!}.}
The promises\index{God's law:promises ``f10!} of it, in like manner, show them God's approbation\index{God's
law:approbation for obedience ``f10!} of obedience, and\index{Believers:obedience:rewarded ``f10!} what blessings
they\index{Believers:blessed in/by God ``f10!} may expect upon the performance thereof;\footnote{\ Leviticus
26:1-14\index{03 Leviticus:ch.  26, v.   1-14, pg. ``f30!} with 2 Corinthians 6:16\index{47 2 Corinthians:ch.   6, v. 
16, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   6, v.  
2-3, pg. ``f30!}; Psalm 37:11\index{19 Psalms:ch.  37, v.  11, pg. ``f30!} with Matthew 5:5\index{40 Matthew:ch.   5,
v.   5, pg. ``f30!}; Psalm 19:11\index{19 Psalms:ch.  19, v.  11, pg. ``f30!}.} although not as due to them by the law,
as a covenant of works.\footnote{\ Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Luke 17:10\index{42
Luke:ch.  17, v.  10, pg. ``f30!}.} So as, a man's doing good, and refraining from evil\index{God's
law:commandments:obedience:refraining from evil ``f10!}, because the law encourages to the one and
deterrers\index{God's law:deterrence from sin ``f10!} from the other, is no evidence of his being under the law; and
not under grace\index{Grace:in believers:by Holy Spirit ``f10!}.\footnote{\ Romans 6:12, 14\index{45 Romans:ch.   6, v.
 12, 14, pg. ``f30!}; 1 Peter 3:8-12\index{60 1 Peter:ch.   3, v.   8-12, pg. ``f30!} with Psalm 34:1\index{19
Psalms:ch.  34, v.   1, pg. ``f30!}2-16\index{19 Psalms:ch.  34, v.  12-16, pg. ``f30!}; Hebrews 12:2\index{58
Hebrews:ch.  12, v.   2, pg. ``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}{}-29\index{58 Hebrews:ch.  12, v. 
28-29, pg. ``f30!}.}


\bigskip

19.7\ \ Neither are the forementioned uses of the law contrary to the grace of\index{Grace:of Gospel ``f10!} the
Gospel\index{Gospel:gospel ``f10!}, but do sweetly comply\index{God's law:compliance with help of Holy Spirit ``f10!}
with it;\footnote{\ Galatians 3:21\index{48 Galatians:ch.   3, v.  21, pg. ``f30!}.} the Spirit of Christ\index{Holy
Spirit:Spirit of Christ ``f10!} subduing and\index{Holy Spirit:subdues will of man ``f10!} enabling the\index{God's
law:commandments:obedience:enabled by Holy Spirit ``f10!} will of man to do that, freely\index{Causation:will of
creatures:free ``f10!} and cheerfully\index{cheerfulness ``f10!}, which the will of God, revealed in the
law\index{Revelation:God's will ``f10!}, requires to be\index{God's law:commandments:duties:required by God ``f10!}
done.\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}; Hebrews 8:10\index{58 Hebrews:ch.   8, v.
 10, pg. ``f30!} with Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}3\index{24 Jeremiah:ch.  31, v.  33,
pg. ``f30!}.}

\paragraph[Chapter 20 {}-- Of Christian Liberty, and Liberty of Conscience]{Chapter 20 -- Of Christian Liberty, and
Liberty of Conscience}
\index{Christian liberty:defined ``f10!}\index{Conscience:liberty of:defined ``f10!}20.1\ \ The liberty
which\index{Christian liberty:purchased by Christ ``f10!} Christ has purchased\index{Redemption:purchased by Christ
``f10!} for believers under\index{Christian liberty:believers under the Gospel ``f10!} the Gospel\index{Gospel:gospel
``f10!} consists in their freedom\index{Believers:freed from:guilt of sin ``f10!} from the guilt\index{guilt ``f10!} of
sin, and condemning\index{Punishment:condemnation ``f10!} wrath of God\index{Wrath:of God:for sin ``f10!}, the curse of
the\index{Curse:of the law ``f10!} moral\index{God's law:commandments:moral ``f10!} law;\footnote{\ Titus 2:14\index{56
Titus:ch.   2, v.  14, pg. ``f30!}; 1 Thessalonians 1:10\index{52 1 Thessalonians:ch.   1, v.  10, pg. ``f30!};
Galatians 3:13\index{48 Galatians:ch.   3, v.  13, pg. ``f30!}.} and, in their being delivered from
this\index{Believers:delivered from:this evil world ``f10!} present evil\index{Christian liberty:deliverance from evil
world ``f10!} world, bondage to Satan\index{Satan:bondage to ``f10!}, and dominion of sin\index{Christian
liberty:dominion of sin destroyed ``f10!};\footnote{\ Galatians 1:4\index{48 Galatians:ch.   1, v.   4, pg. ``f30!};
Colossians 1:13\index{51 Colossians:ch.   1, v.  13, pg. ``f30!}; Acts 26:18\index{44 Acts:ch.  26, v.  18, pg.
``f30!}; Romans 6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}.} from the evil of afflictions,
the\index{Salvation:release from evil afflictions ``f10!} sting of death\index{Death:believers:delivered from sting of
``f10!}, the victory of\index{Death:defeated by Christ ``f10!} the grave, and
everlasting\index{Punishment:everlasting:damnation ``f10!} damnation;\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.
  2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!}; Psalm 119:71\index{19 Psalms:ch. 119, v.  71, pg.
``f30!}; 1 Corinthians 15:54-57\index{46 1 Corinthians:ch.  15, v.  54-57, pg. ``f30!}; Romans 8:1\index{45 Romans:ch. 
 8, v.   1, pg. ``f30!}.} as also, in their free access to God\index{Believers:access to God only through Christ
``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}{}-2\index{45 Romans:ch.   5, v.   1-2, pg.
``f30!}.} and their yielding\index{God's law:commandments:obedience:required ``f10!} obedience unto Him, not out of
slavish\index{God's law:commandments:obedience:not of slavish fear ``f10!} fear, but\index{Fear:slavish ``f10!} a
child-like\index{Love toward God:childlike ``f10!} love and willing mind.\index{God's
law:commandments:obedience:willing mind ``f10!}\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg.
``f30!}4-15\index{45 Romans:ch.   8, v.  14-15, pg. ``f30!}; 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg.
``f30!}8\index{62 1 John:ch.   4, v.  18, pg. ``f30!}.} All which were common also to believers under\index{Christian
liberty:believers under the law ``f10!} the law.\footnote{\ Galatians 3:9, 14\index{48 Galatians:ch.   3, v.   9, 14,
pg. ``f30!}.} But, under the new testament\index{Scripture:New Testament ``f10!}, the liberty of
Christians\index{Christian liberty:enlarged under NT economy ``f10!} is further enlarged, in
their\index{Believers:freed from:yoke of the law ``f10!} freedom from the yoke\index{Ceremonial law:yoke ``f10!} of the
ceremonial\index{Ceremonial law:fore-signifying Christ ``f10!} law, to which the Jewish\index{Church:Jewish ``f10!}
Church was subjected\index{Ceremonial law:subjected Jewish Church ``f10!};\footnote{\ Galatians 4:1-3\index{48
Galatians:ch.   4, v.   1-3, pg. ``f30!}, 6-7\index{48 Galatians:ch.   4, v.   1-3, 6-7, pg. ``f30!}; Galatians
5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}; Acts 15:10-11\index{44 Acts:ch.  15, v.  10-11, pg. ``f30!}.} and
in greater boldness\index{Prayer:with boldness ``f10!} of access to the\index{Believers:bold access to God's throne
``f10!} throne of grace\index{Grace:throne of ``f10!},\footnote{\ Hebrews 4:14, 16\index{58 Hebrews:ch.   4, v.  14,
16, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}9-22\index{58 Hebrews:ch.  10, v.  19-22,
pg. ``f30!}.} and in fuller communications of the free Spirit\index{Holy Spirit:instructs the elect ``f10!} of God,
than believers under the law did ordinarily partake\index{Believers:benefits:in this life ``f10!} of.\footnote{\ John
7:38-39\index{43 John:ch.   7, v.  38-39, pg. ``f30!}; 2 Corinthians 3:13, 17-18\index{47 2 Corinthians:ch.   3, v. 
13, 17-18, pg. ``f30!}.}


\bigskip

20.2\ \ God alone is Lord of the conscience\index{Conscience:God alone lord of ``f10!},\footnote{\ James 4:12\index{59
James:ch.   4, v.  12, pg. ``f30!}; Romans 14:4\index{45 Romans:ch.  14, v.   4, pg. ``f30!}.} and has left it free
from the doctrines\index{Doctrine:false:of men ``f10!} and commandments\index{Commandments of men:false ``f10!} of men,
which are in any thing contrary to His Word; or beside it, in matters of faith\index{Faith:contrary opinions to the
``f10!} or worship\index{Worship:of God ``f10!}.\footnote{\ Acts 4:19\index{44 Acts:ch.   4, v.  19, pg. ``f30!}; Acts
5:2\index{44 Acts:ch.   5, v.   2, pg. ``f30!}9\index{44 Acts:ch.   5, v.  29, pg. ``f30!}; 1 Corinthians 7:23\index{46
1 Corinthians:ch.   7, v.  23, pg. ``f30!}; Matthew 23:8-10\index{40 Matthew:ch.  23, v.   8-10, pg. ``f30!}; 2
Corinthians 1:24\index{47 2 Corinthians:ch.   1, v.  24, pg. ``f30!}; Matthew 15:9\index{40 Matthew:ch.  15, v.   9,
pg. ``f30!}.} So that, to believe such doctrines, or to obey such commands, out of
conscience\index{Conscience:influenced falsely by others ``f10!},\footnote{\ Colossians 2:2\index{51 Colossians:ch.  
2, v.   2, pg. ``f30!}0, 22-23\index{51 Colossians:ch.   2, v.  20, 22-23, pg. ``f30!}; Galatians 1:10\index{48
Galatians:ch.   1, v.  10, pg. ``f30!}; Galatians 2:4-5\index{48 Galatians:ch.   2, v.   4-5, pg. ``f30!}; Galatians
5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}.} is to betray true liberty of conscience\index{Conscience:liberty
of:defined ``f10!}: and the requiring\index{Salvation:faith alone required ``f10!} of an implicit\index{Faith:implicit
``f10!} faith, and an absolute\index{Conscience:liberty of:not absolute obedience when contrary to God's Word ``f10!}
and blind obedience is to destroy\index{Conscience:liberty of:destroyed ``f10!} liberty of conscience, and reason
also.\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}7\index{45 Romans:ch.  10, v.  17, pg.
``f30!}; Romans 14:23\index{45 Romans:ch.  14, v.  23, pg. ``f30!}; Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg.
``f30!}; Acts 17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}; John 4:22\index{43 John:ch.   4, v.  22, pg. ``f30!};
Hosea 5:11\index{28 Hosea:ch.   5, v.  11, pg. ``f30!}; Revelation 13:12, 16-17\index{66 Revelation:ch.  13, v.  12,
16-17, pg. ``f30!}; Jeremiah 8:9\index{24 Jeremiah:ch.   8, v.   9, pg. ``f30!}.}


\bigskip

20.3\ \ They who, upon pretence\index{pretence ``f10!} of Christian liberty\index{Christian liberty:defined ``f10!}, do
practice any sin, or cherish any lust\index{Sexual immorality:lust ``f10!}, do thereby destroy
the\index{Conscience:liberty of:destroyed ``f10!} end of Christian liberty, which is, that being delivered
out\index{Believers:delivered from:hands of enemies ``f10!} of the hands of our enemies\index{Enemies:believers
delivered from ``f10!}, we might serve the\index{Love toward God:in service ``f10!} Lord, without fear\index{Fear:none
``f10!}, in holiness and\index{Believers:holy:being perfected ``f10!} righteousness before\index{Believers:to serve God
in righteousness ``f10!} Him, all the days of our life.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1,
pg. ``f30!}3\index{48 Galatians:ch.   5, v.  13, pg. ``f30!}; 1 Peter 2:16\index{60 1 Peter:ch.   2, v.  16, pg.
``f30!}; 2 Peter 2:19\index{61 2 Peter:ch.   2, v.  19, pg. ``f30!}; John 8:34\index{43 John:ch.   8, v.  34, pg.
``f30!}; Luke 1:74-75\index{42 Luke:ch.   1, v.  74-75, pg. ``f30!}.}


\bigskip

20.4\ \ And because the powers which God has ordained, and\index{Civil magistrate:ordained by God ``f10!} the liberty
which\index{Christian liberty:purchased by Christ ``f10!} Christ has purchased\index{Redemption:purchased by Christ
``f10!}, are not intended by God to destroy\index{Conscience:liberty of:destroyed ``f10!}, but mutually to uphold
and\index{Christian liberty:upholds just laws of men ``f10!} preserve one\index{Human life:preservation ``f10!}
another; they who, upon pretence\index{pretence ``f10!} of Christian liberty\index{Christian liberty:defined ``f10!},
shall oppose\index{Liberty:not oppose lawful governance ``f10!} any lawful power, or the lawful exercise of it, whether
it be civil or ecclesiastical\index{Church government:ecclesiastical powers not to be resisted ``f10!}, resist the
ordinance of\index{Worship:ordinances:to be:esteemed ``f10!} God.\footnote{\ Matthew 12:25\index{40 Matthew:ch.  12, v.
 25, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14,
pg. ``f30!}, 16\index{60 1 Peter:ch.   2, v.  13-14, 16, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg.
``f30!}{}-8\index{45 Romans:ch.  13, v.   1-8, pg. ``f30!}; Hebrews 13:17\index{58 Hebrews:ch.  13, v.  17, pg.
``f30!}.} And, for their publishing\index{Church discipline:for opinions contrary to truth ``f10!} of such opinions, or
maintaining of such practices, as are contrary to the light of nature, or to the known principles of
Christianity\index{Christianity ``f10!}, whether concerning faith\index{Faith:contrary opinions to the ``f10!},
worship, or\index{Worship:of God ``f10!} conversation\index{Believers:to walk faithfully in obedience (conversation)
``f10!}; or, to the power of godliness\index{God's attributes:power of godliness ``f10!}; or, such
erroneous\index{Falsehood:false opinions ``f10!} opinions or practices, as either in their own nature, or in the manner
of publishing or maintaining them, are destructive\index{Heresy:heresies ``f10!} to the
external\index{Heresy:destructive or external peace of Church ``f10!} peace and order\index{Church:to maintain peace
and order ``f10!} which Christ has established\index{Church:order established by Christ ``f10!} in the Church,
they\index{Church:institution ``f10!} may lawfully be\index{Church discipline:heresies lawfully censured ``f10!} called
to account, and proceeded against by the censures of\index{Church discipline:censures ``f10!} the
Church,\footnote{\ Romans 1:32\index{45 Romans:ch.   1, v.  32, pg. ``f30!} with 1 Corinthians 5:1\index{46 1
Corinthians:ch.   5, v.   1, pg. ``f30!}, 5, 11, 13\index{46 1 Corinthians:ch.   5, v.   1, 5, 11, 13, pg. ``f30!}; 2
John 10-11\index{63 2 John:v. 10-11, pg. ``f30!} and 2 Thessalonians 3:1\index{53 2 Thessalonians:ch.   3, v.   1, pg.
``f30!}4\index{53 2 Thessalonians:ch.   3, v.  14, pg. ``f30!}; 1 Timothy 6:3\index{54 1 Timothy:ch.   6, v.   3, pg.
``f30!}{}-5\index{54 1 Timothy:ch.   6, v.   3-5, pg. ``f30!}; Titus 1:10-11, 13\index{56 Titus:ch.   1, v.  10-11, 13,
pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!} with
Matthew 18:15\index{40 Matthew:ch.  18, v.  15, pg. ``f30!}{}-17\index{40 Matthew:ch.  18, v.  15-17, pg. ``f30!}; 1
Timothy 1:19-20\index{54 1 Timothy:ch.   1, v.  19-20, pg. ``f30!}; Revelation 2:2, 14-15, 20\index{66 Revelation:ch.  
2, v.   2, 14-15, 20, pg. ``f30!}; Revelation 3:9\index{66 Revelation:ch.   3, v.   9, pg. ``f30!}.} and by the power
of the civil magistrate\index{Civil magistrate:to punish evil doers ``f10!}.\footnote{\ Deuteronomy 13:6-12\index{05
Deuteronomy:ch.  13, v.   6-12, pg. ``f30!}; Romans 13:3\index{45 Romans:ch.  13, v.   3, pg. ``f30!}{}-4\index{45
Romans:ch.  13, v.   3-4, pg. ``f30!} with 2 John 10-11\index{63 2 John:v. 10-11, pg. ``f30!}; Ezra 7:23,
25-28\index{15 Ezra:ch.   7, v.  23, 25-28, pg. ``f30!}; Revelation 17:12, 16-17\index{66 Revelation:ch.  17, v.  12,
16-17, pg. ``f30!}; Nehemiah 13:15, 17\index{16 Nehemiah:ch.  13, v.  15, 17, pg. ``f30!}, 21-22, 25, 30\index{16
Nehemiah:ch.  13, v.  15, 17, 21-22, 25, 30, pg. ``f30!}; 2 Kings 23:5-6, 9, 20-21\index{12 2 Kings:ch.  23, v.   5-6,
9, 20-21, pg. ``f30!}; 2 Chronicles 34:33\index{14 2 Chronicles:ch.  34, v.  33, pg. ``f30!}; 2 Chronicles
15:12-13\index{14 2 Chronicles:ch.  15, v.  12-13, pg. ``f30!}, 16\index{14 2 Chronicles:ch.  15, v.  12-13, 16, pg.
``f30!}; Daniel 3:29\index{27 Daniel:ch.   3, v.  29, pg. ``f30!}; 1 Timothy 2:2\index{54 1 Timothy:ch.   2, v.   2,
pg. ``f30!}; Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; Zechariah 13:2, 3\index{38 Zechariah:ch.  13,
v.   2, 3, pg. ``f30!}.}

\paragraph[Chapter 21 {}-- Of Religious Worship and the Sabbath Day]{Chapter 21 -- Of Religious Worship and the Sabbath
Day}
\index{Worship:glorifying God ``f10!}\index{Worship:on the Sabbath ``f10!}\index{Sabbath:for worship ``f10!}21.1\ \ The
light of nature shows\index{Revelation:natural:declares there is a God ``f10!} that there is a God, who has
lordship\index{God's attributes:sovereign lordship ``f10!} and sovereignty over all, is good, and does good unto all,
and is therefore to be feared\index{Fear:of God:reverence ``f10!}, loved\index{Love toward God:for God ``f10!},
praised\index{Worship:praising God ``f10!}, called upon, trusted\index{trust in God ``f10!} in, and served\index{Love
toward God:in service ``f10!}, with all the heart, and with all the soul\index{Love toward God:with all our:soul
``f10!}, and with all the might.\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.  20, pg. ``f30!}; Acts
17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}; Psalm 119:68\index{19 Psalms:ch. 119, v.  68, pg. ``f30!}; Jeremiah
10:7\index{24 Jeremiah:ch.  10, v.   7, pg. ``f30!}; Psalm 31:23\index{19 Psalms:ch.  31, v.  23, pg. ``f30!}; Psalm
18:3\index{19 Psalms:ch.  18, v.   3, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}2\index{45
Romans:ch.  10, v.  12, pg. ``f30!}; Psalm 62:8\index{19 Psalms:ch.  62, v.   8, pg. ``f30!}; Joshua 24:14\index{06
Joshua:ch.  24, v.  14, pg. ``f30!}; Mark 12:33\index{41 Mark:ch.  12, v.  33, pg. ``f30!}.} But the acceptable way of
worship\index{Worship:of God ``f10!}ping the true God is instituted by Himself\index{Worship:correct:instituted by God
``f10!}, and so limited\index{Worship:only acts approved by God ``f10!} by His own revealed will\index{Revelation:God's
will ``f10!}, that He may not be worshipped according to the imaginations and\index{Worship:false:imaginations and
devices of men ``f10!} devices of men, or the suggestions of Satan\index{Satan:tempter ``f10!}, under any
visible\index{Idolatry:worship of likeness of any created thing ``f10!} representation\index{Idolatry:any visible
representation of God or gods ``f10!}, or any other way not prescribed in the holy\index{Scripture:holy ``f10!}
Scripture.\footnote{\ Deuteronomy 12:32\index{05 Deuteronomy:ch.  12, v.  32, pg. ``f30!}; Matthew 15:9\index{40
Matthew:ch.  15, v.   9, pg. ``f30!}; Acts 17:25\index{44 Acts:ch.  17, v.  25, pg. ``f30!}; Matthew 4:9-10\index{40
Matthew:ch.   4, v.   9-10, pg. ``f30!}; Deuteronomy 4:15-20\index{05 Deuteronomy:ch.   4, v.  15-20, pg. ``f30!};
Exodus 20:4-6\index{02 Exodus:ch.  20, v.   4-6, pg. ``f30!}; Colossians 2:2\index{51 Colossians:ch.   2, v.   2, pg.
``f30!}3\index{51 Colossians:ch.   2, v.  23, pg. ``f30!}.}


\bigskip

21.2\ \ Religious\index{Worship:glorifying God ``f10!} worship is to\index{Worship:of Trinity ``f10!} be given to God,
the Father\index{Father:God the ``f10!}, Son\index{Jesus:Son of God ``f10!}, and Holy Ghost\index{Holy Spirit:Holy
Ghost ``f10!}; and to Him alone;\footnote{\ Matthew 4:10\index{40 Matthew:ch.   4, v.  10, pg. ``f30!} with John
5:23\index{43 John:ch.   5, v.  23, pg. ``f30!} and 2 Corinthians 13:1\index{47 2 Corinthians:ch.  13, v.   1, pg.
``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}.} not to angels\index{Angels:angels ``f10!},
saints\index{Believers:saints ``f10!}, or any other creature\index{creatures ``f10!}:\footnote{\ Colossians
2:18\index{51 Colossians:ch.   2, v.  18, pg. ``f30!}; Revelation 19:10\index{66 Revelation:ch.  19, v.  10, pg.
``f30!}; Romans 1:25\index{45 Romans:ch.   1, v.  25, pg. ``f30!}.} and since the fall\index{Sin:original:first
transgression (of Adam) ``f10!}, not without a Mediator\index{Jesus:mediator:of covenant ``f10!}; nor in the mediation
of any other but of Christ alone.\footnote{\ John 14:6\index{43 John:ch.  14, v.   6, pg. ``f30!}; 1 Timothy
2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg.
``f30!}8\index{49 Ephesians:ch.   2, v.  18, pg. ``f30!}; Colossians 3:17\index{51 Colossians:ch.   3, v.  17, pg.
``f30!}.}


\bigskip

21.3\ \ Prayer, with\index{Prayer:with thanksgiving ``f10!} thanksgiving\index{Prayer:thanking God ``f10!}, being one
special part of religious worship,\index{Worship:elements ``f10!}\footnote{\ Philippians 4:6\index{50 Philippians:ch.  
4, v.   6, pg. ``f30!}.} is by God required of all men:\footnote{\ Psalm 65:2\index{19 Psalms:ch.  65, v.   2, pg.
``f30!}.} and that it may be accepted, it is to be made in\index{Prayer:accepted in Christ ``f10!} the name of the
Son\index{Jesus:Son of God ``f10!},\footnote{\ John 14:13-14\index{43 John:ch.  14, v.  13-14, pg. ``f30!}; 1 Peter
2:5\index{60 1 Peter:ch.   2, v.   5, pg. ``f30!}.} by the help of His\index{Prayer:helped by Holy Spirit ``f10!}
Spirit,\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}6\index{45 Romans:ch.   8, v.  26, pg.
``f30!}.} according to His will,\footnote{\ 1 John 5:14\index{62 1 John:ch.   5, v.  14, pg. ``f30!}.} with
understanding, reverence, humility\index{Humility:before God ``f10!}, fervency\index{Prayer:with fervency ``f10!},
faith, love\index{Scripture:to be:loved in faith ``f10!}, and perseverance;\index{Prayer:with perseverance
``f10!}\footnote{\ Psalm 47:7\index{19 Psalms:ch.  47, v.   7, pg. ``f30!}; Ecclesiastes 5:1\index{21 Ecclesiastes:ch. 
 5, v.   1, pg. ``f30!}{}-2\index{21 Ecclesiastes:ch.   5, v.   1-2, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch. 
12, v.   2, pg. ``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}; Genesis 18:27\index{01 Genesis:ch.  18, v.  27,
pg. ``f30!}; James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}; James 1:6\index{59 James:ch.   1, v.   6, pg.
``f30!}{}-7\index{59 James:ch.   1, v.   6-7, pg. ``f30!}; Mark 11:24\index{41 Mark:ch.  11, v.  24, pg. ``f30!};
Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}2\index{40 Matthew:ch.   6, v.  12, pg. ``f30!},
14-15\index{40 Matthew:ch.   6, v.  12, 14-15, pg. ``f30!}; Colossians 4:2\index{51 Colossians:ch.   4, v.   2, pg.
``f30!}; Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}.} and, if vocal\index{Prayer:vocal in a known
language ``f10!}, in a known tongue.\footnote{\ 1 Corinthians 14:14\index{46 1 Corinthians:ch.  14, v.  14, pg.
``f30!}.}


\bigskip

21.4\ \ Prayer is to be made for things lawful\index{Prayer:for lawful things ``f10!};\footnote{\ 1 John 5:14\index{62 1
John:ch.   5, v.  14, pg. ``f30!}.} and for all sorts of men living, or that shall live hereafter:\footnote{\ 1 Timothy
2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}; John
17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}0\index{43 John:ch.  17, v.  20, pg. ``f30!}; 2 Samuel 7:29\index{10 2
Samuel:ch.   7, v.  29, pg. ``f30!}; Ruth 4:12\index{08 Ruth:ch.   4, v.  12, pg. ``f30!}.} but not for the
dead\index{Prayer:not to be offered for the dead ``f10!},\footnote{\ 2 Samuel 12:21-23\index{10 2 Samuel:ch.  12, v. 
21-23, pg. ``f30!} with Luke 16:25-26\index{42 Luke:ch.  16, v.  25-26, pg. ``f30!}; Revelation 14:13\index{66
Revelation:ch.  14, v.  13, pg. ``f30!}.} nor for those of whom it may be known that they have sinned the
sin\index{Sin:sinned the sin unto death ``f10!} unto death\index{Death:committed sin unto ``f10!}.\footnote{\ 1 John
5:16\index{62 1 John:ch.   5, v.  16, pg. ``f30!}.}


\bigskip

21.5\ \ The reading\index{Scripture:to be:read and heard ``f10!} of the Scriptures with godly fear,\footnote{\ Acts
15:2\index{44 Acts:ch.  15, v.   2, pg. ``f30!}1\index{44 Acts:ch.  15, v.  21, pg. ``f30!}; Revelation 1:3\index{66
Revelation:ch.   1, v.   3, pg. ``f30!}.} the sound preaching\index{Preaching:sound doctrine ``f10!}\footnote{\ 2
Timothy 4:2\index{55 2 Timothy:ch.   4, v.   2, pg. ``f30!}.} and conscionable\index{Preaching:hearing:conscionable
``f10!} hearing of the Word, in obedience\index{God's law:commandments:obedience:required ``f10!} unto God, with
understanding\index{Scripture:to be:read with understanding ``f10!}, faith and reverence;\footnote{\ James
1:22\index{59 James:ch.   1, v.  22, pg. ``f30!}; Acts 10:33\index{44 Acts:ch.  10, v.  33, pg. ``f30!}; Matthew
13:19\index{40 Matthew:ch.  13, v.  19, pg. ``f30!}; Hebrews 4:2\index{58 Hebrews:ch.   4, v.   2, pg. ``f30!}; Isaiah
66:2\index{23 Isaiah:ch.  66, v.   2, pg. ``f30!}.} singing of psalms\index{Psalms:singing ``f10!} with
grace\index{Psalms:to be sung with grace ``f10!} in the heart;\index{Worship:grace in heart, as singing Psalms
``f10!}\footnote{\ Colossians 3:16\index{51 Colossians:ch.   3, v.  16, pg. ``f30!}; Ephesians 5:19\index{49
Ephesians:ch.   5, v.  19, pg. ``f30!}; James 5:13\index{59 James:ch.   5, v.  13, pg. ``f30!}.} as also, the due
administration\index{Worship:administration of sacraments ``f10!} and worthy receiving of the
sacraments\index{Sacraments:holy:ordinances ``f10!} instituted by Christ\index{Sacraments:instituted by:Christ ``f10!};
are all parts of the ordinary\index{Sacraments:part of ordinary worship ``f10!} religious worship of
God\index{Worship:of God ``f10!}:\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg.
``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  23-29, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg.
``f30!}.} beside religious\index{Oaths and Vows:religious worship ``f10!} oaths\index{Oaths and Vows:oaths
``f10!},\footnote{\ Deuteronomy 6:13\index{05 Deuteronomy:ch.   6, v.  13, pg. ``f30!} with Nehemiah 10:29\index{16
Nehemiah:ch.  10, v.  29, pg. ``f30!}.} vows,\footnote{\ Isaiah 19:21\index{23 Isaiah:ch.  19, v.  21, pg. ``f30!} with
Ecclesiastes 5:4-5\index{21 Ecclesiastes:ch.   5, v.   4-5, pg. ``f30!}.} solemn fastings\index{fasting
``f10!},\footnote{\ Joel 2:12\index{29 Joel:ch.   2, v.  12, pg. ``f30!}; Esther 4:16\index{17 Esther:ch.   4, v.  16,
pg. ``f30!}; Matthew 9:15\index{40 Matthew:ch.   9, v.  15, pg. ``f30!}; 1 Corinthians 7:5\index{46 1 Corinthians:ch.  
7, v.   5, pg. ``f30!}.} and thanksgivings\index{Prayer:thanking God ``f10!}, upon special occasions,\footnote{\ Psalm
107; Esther 9:22\index{17 Esther:ch.   9, v.  22, pg. ``f30!}.} which are, in their several times and
seasons\index{Worship:elements in times and seasons ``f10!}, to be used in a holy\index{Scripture:holy ``f10!} and
religious manner\index{Worship:acceptable manner ``f10!}.\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg.
``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}.}


\bigskip

21.6\ \ Neither prayer\index{Prayer:act of worship ``f10!}, nor any other part of religious
worship,\index{Worship:elements ``f10!} is now under the Gospel\index{Gospel:gospel ``f10!} either tied unto, or made
more acceptable by any place in which it is performed, or towards which it is directed:\footnote{\ John 4:21\index{43
John:ch.   4, v.  21, pg. ``f30!}.} but God is to be worshipped everywhere\index{Worship:in spirit and truth
``f10!},\footnote{\ Malachi 1:11\index{39 Malachi:ch.   1, v.  11, pg. ``f30!}; 1 Timothy 2:8\index{54 1 Timothy:ch.  
2, v.   8, pg. ``f30!}.} in spirit and truth\index{God's attributes:truth ``f10!};\footnote{\ John 4:23-24\index{43
John:ch.   4, v.  23-24, pg. ``f30!}.} as in private families\index{family, natural ``f10!}\footnote{\ Jeremiah
10:25\index{24 Jeremiah:ch.  10, v.  25, pg. ``f30!}; Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg.
``f30!}{}-7\index{05 Deuteronomy:ch.   6, v.   6-7, pg. ``f30!}; Job 1:5\index{18 Job:ch.   1, v.   5, pg. ``f30!}; 2
Samuel 6:18, 20\index{10 2 Samuel:ch.   6, v.  18, 20, pg. ``f30!}; 1 Peter 3:7\index{60 1 Peter:ch.   3, v.   7, pg.
``f30!}; Acts 10:2\index{44 Acts:ch.  10, v.   2, pg. ``f30!}.} daily,\index{Prayer:daily in private families,
``f10!}\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}.} and in secret each\index{Prayer:secret
``f10!} one by himself;\footnote{\ Matthew 6:6\index{40 Matthew:ch.   6, v.   6, pg. ``f30!}; Ephesians 6:18\index{49
Ephesians:ch.   6, v.  18, pg. ``f30!}.} so, more solemnly, in\index{Prayer:solemn ``f10!} the public assemblies,
which\index{Worship:assemblies not to be neglected ``f10!} are not carelessly or wilfully to be neglected, or forsaken,
when\index{Worship:not to be:forsaken ``f10!} God, by His Word or providence\index{Providence:providence ``f10!}, calls
thereunto\index{Worship:assembly called together ``f10!}.\footnote{\ Isaiah 56:6-7\index{23 Isaiah:ch.  56, v.   6-7,
pg. ``f30!}; Hebrews 10:25\index{58 Hebrews:ch.  10, v.  25, pg. ``f30!}; Proverbs 1:20-21, 24\index{20 Proverbs:ch.  
1, v.  20-21, 24, pg. ``f30!}; Proverbs 8:34\index{20 Proverbs:ch.   8, v.  34, pg. ``f30!}; Acts 13:42\index{44
Acts:ch.  13, v.  42, pg. ``f30!}; Luke 4:16\index{42 Luke:ch.   4, v.  16, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.  
2, v.  42, pg. ``f30!}.}


\bigskip

21.7\ \ As it is the law of nature, that, in general, a due proportion\index{Sabbath:proportion of time set apart for
worship ``f10!} of time be set apart for the worship of God; so\index{Worship:times of ``f10!}, in His Word, by a
positive\index{Sabbath:positive, moral, perpetual command ``f10!}, moral\index{God's law:commandments:moral ``f10!},
and perpetual commandment\index{Sabbath:perpetual commandment ``f10!}, binding all\index{Sabbath:binding on all men
``f10!} men, in all ages, He has particularly appointed one day in seven\index{Sabbath:one whole day in seven ``f10!},
for a Sabbath, to\index{Sabbath:seventh day ``f10!} be kept holy\index{Sabbath:to be:kept holy ``f10!} unto
Him:\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}, 10\index{02 Exodus:ch.  20, v.   8, 10, pg.
``f30!}{}-11\index{02 Exodus:ch.  20, v.   8, 10-11, pg. ``f30!}; Isaiah 56:2, 4, 6-7\index{23 Isaiah:ch.  56, v.   2,
4, 6-7, pg. ``f30!}.} which, from the beginning of the world to the resurrection of
Christ\index{Jesus:resurrection:from the dead ``f10!}, was the last day of\index{Sabbath:last day of week ``f10!} the
week; and, from the resurrection of Christ, was changed into the first day\index{Sabbath:first day of week ``f10!} of
the week,\footnote{\ Genesis 2:2-3\index{01 Genesis:ch.   2, v.   2-3, pg. ``f30!}; 1 Corinthians 16:1-2\index{46 1
Corinthians:ch.  16, v.   1-2, pg. ``f30!}; Acts 20:7\index{44 Acts:ch.  20, v.   7, pg. ``f30!}.} which, in Scripture,
is\index{Sabbath:Lord's Day ``f10!} called the Lord's Day,\footnote{\ Revelation 1:10\index{66 Revelation:ch.   1, v. 
10, pg. ``f30!}.} and is to be continued to the end of the world, as the Christian Sabbath\index{Sabbath:Christian
``f10!}.\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}, 10\index{02 Exodus:ch.  20, v.   8, 10,
pg. ``f30!} with Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.   5, v.  17-18,
pg. ``f30!}.}


\bigskip

21.8\ \ This Sabbath is\index{Sabbath:to be:kept holy ``f10!} then kept holy unto the Lord, when men, after a due
preparing\index{Heart:prepared for Sabbath ``f10!} of their hearts, and ordering of the\index{Church
councils:directions for ordering public worship ``f10!}ir common affairs beforehand\index{Sabbath:order common affairs
before ``f10!}, do not only observe an holy rest, all the day, from their own works\index{Sabbath:cease:own works
``f10!}, words, and thoughts about\index{Sabbath:avoid unnecessary, worldly thoughts ``f10!} their worldly employments
and recreations\index{Sabbath:cease:worldly employments and recreations ``f10!},\footnote{\ Exodus 20:8\index{02
Exodus:ch.  20, v.   8, pg. ``f30!}; Exodus 16:23\index{02 Exodus:ch.  16, v.  23, pg. ``f30!}, 25-26, 29-30\index{02
Exodus:ch.  16, v.  23, 25-26, 29-30, pg. ``f30!}; Exodus 31:15\index{02 Exodus:ch.  31, v.  15, pg.
``f30!}{}-17\index{02 Exodus:ch.  31, v.  15-17, pg. ``f30!}; Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg.
``f30!}; Nehemiah 13:15-19, 21-22\index{16 Nehemiah:ch.  13, v.  15-19, 21-22, pg. ``f30!}.} but also are taken up the
whole time in the public and private exercises\index{Worship:private exercises ``f10!} of His worship, and
in\index{Worship:of God ``f10!} the duties of necessity and\index{Sabbath:works of necessity and mercy ``f10!}
mercy.\footnote{\ Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}; Matthew 12:1-13\index{40 Matthew:ch.  12,
v.   1-13, pg. ``f30!}.}

\section{Chapter 22 -- Of Lawful Oaths and Vows}
\index{Oaths and Vows:lawful ``f10!}\index{Oaths and Vows:vows ``f10!}22.1\ \ A lawful\index{Oaths and Vows:lawful
``f10!} oath\index{Oaths and Vows:oaths ``f10!} is a part of religious worship,\index{Worship:elements
``f10!}\footnote{\ Deuteronomy 10:20\index{05 Deuteronomy:ch.  10, v.  20, pg. ``f30!}.} wherein, upon just occasion,
the person swearing\index{Oaths and Vows:swear only by God's name ``f10!} solemnly calls God\index{Oaths and
Vows:solemnly calls God to witness ``f10!} to witness what he asserts\index{Oaths and Vows:assertion with God as
witness ``f10!}, or promises\index{Oaths and Vows:promises made ``f10!}, and to judge him according to the truth
or\index{Oaths and Vows:assertion of truth ``f10!} falsehood\index{Oaths and Vows:consistent with truth or falsehood
``f10!} of what he swears.\footnote{\index{02 Exodus:ch.  20, v.   7, pg.
``f30!}\index{03 Leviticus:ch.  19, v.  12, pg.
``f30!}\index{47 2 Corinthians:ch.   1, v.  23, pg.
``f30!}\index{14 2 Chronicles:ch.   6, v.  22-23, pg.
``f30!}


\bigskip

22.2\ \ The name of God only is that by which men ought to swear\index{Oaths and Vows:swear only by God's name ``f10!};
and therein it is to be used with all holy\index{God's name:holy ``f10!} fear and\index{Fear:of God:reverence ``f10!}
reverence.\footnote{\index{05 Deuteronomy:ch.   6, v.  13, pg.
``f30!}
rashly\index{rashness ``f10!}, by that glorious and\index{God's name:glorious and dreadful ``f10!} dreadful Name; or,
to swear at all by any other thing, is sinful, and to be abhorred.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.  
7, pg. ``f30!}; Jeremiah 5:7\index{24 Jeremiah:ch.   5, v.   7, pg. ``f30!}; Matthew 5:34, 37\index{40 Matthew:ch.   5,
v.  34, 37, pg. ``f30!}; James 5:12\index{59 James:ch.   5, v.  12, pg. ``f30!}.} Yet, as in matters of weight and
moment, an oath\index{Oaths and Vows:oaths ``f10!} is warranted\index{Oaths and Vows:warranted by the Word of God
``f10!} by the Word of God\index{Scripture:word of God:word of God ``f10!}, under the New Testament\index{Scripture:New
Testament ``f10!}, as well as under the Old;\footnote{\ Hebrews 6:16\index{58 Hebrews:ch.   6, v.  16, pg. ``f30!}; 2
Corinthians 1:23\index{47 2 Corinthians:ch.   1, v.  23, pg. ``f30!}; Isaiah 65:16\index{23 Isaiah:ch.  65, v.  16, pg.
``f30!}.} so a lawful\index{Oaths and Vows:lawful ``f10!} oath, being imposed by lawful authority, in such matters
ought to be taken.\footnote{\ I Kings 8:31; Nehemiah 13:25\index{16 Nehemiah:ch.  13, v.  25, pg. ``f30!}; Ezra
10:5\index{15 Ezra:ch.  10, v.   5, pg. ``f30!}.}


\bigskip

22.3\ \ Whosoever takes an oath\index{Oaths and Vows:oaths ``f10!} ought duly to consider the weightiness\index{Oaths
and Vows:consider the weightiness ``f10!} of so solemn an\index{Oaths and Vows:solemn act ``f10!} act; and therein to
avouch\index{Oaths and Vows:avouch only truth ``f10!} nothing, but what he is fully persuaded is the truth\index{Oaths
and Vows:assertion of truth ``f10!}.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.   7, pg. ``f30!}; Jeremiah
4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}.} Neither may any man bind himself\index{Oaths and Vows:bind:only to
good and just ``f10!} by oath to anything but what is good and just, and what he believes so to be, and what he is able
and resolved to\index{Oaths and Vows:resolve to perform ``f10!} perform.\footnote{\ Genesis 24:2-3, 5-6, 8-9\index{01
Genesis:ch.  24, v.   2-3, 5-6, 8-9, pg. ``f30!}.} Yet is it a sin to refuse\index{Oaths and Vows:not to:refuse lawful
``f10!} an oath touching anything that is good and just, being imposed by lawful authority\index{Oaths and Vows:lawful
``f10!}.\footnote{\ Numbers 5:19, 21\index{04 Numbers:ch.   5, v.  19, 21, pg. ``f30!}; Nehemiah 5:12\index{16
Nehemiah:ch.   5, v.  12, pg. ``f30!}; Exodus 22:7-11\index{02 Exodus:ch.  22, v.   7-11, pg. ``f30!}.}


\bigskip

22.4\ \ An oath\index{Oaths and Vows:oaths ``f10!} is to be taken in the plain and\index{Oaths and Vows:take in plain
and common sense of words ``f10!} common sense of the words, without equivocation\index{Oaths and Vows:not with
equivocation ``f10!}, or mental reservation\index{Oaths and Vows:no mental reservations ``f10!}.\footnote{\ Jeremiah
4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}; Psalm 24:4\index{19 Psalms:ch.  24, v.   4, pg. ``f30!}.} It cannot
oblige to sin: but in anything not sinful, being taken, it binds to\index{Oaths and Vows:bind:performance ``f10!}
performance, although to a man's own hurt.\footnote{\ 1 Samuel 25:22, 32-34\index{09 1 Samuel:ch.  25, v.  22, 32-34,
pg. ``f30!}; Psalm 15:4\index{19 Psalms:ch.  15, v.   4, pg. ``f30!}.} Not is it to be violated\index{Oaths and
Vows:not to:be violated ``f10!}, although made to heretics\index{heretics ``f10!}, or infidels\index{Oaths and Vows:to
be kept, even if made to infidels ``f10!}.\footnote{\ Ezekiel 17:16, 18-19\index{26 Ezekiel:ch.  17, v.  16, 18-19, pg.
``f30!}; Joshua 9:18-19\index{06 Joshua:ch.   9, v.  18-19, pg. ``f30!} with 2 Samuel 21:1\index{10 2 Samuel:ch.  21,
v.   1, pg. ``f30!}.}


\bigskip

22.5\ \ A vow is\index{Oaths and Vows:vow defined ``f10!} of the like nature with a promissory\index{Oaths and
Vows:promissory ``f10!} oath\index{Oaths and Vows:oaths ``f10!}, and ought to be made with the like religious
care\index{Oaths and Vows:religious worship ``f10!}, and to be performed with the like faithfulness\index{faithfulness
``f10!}.\footnote{\ Isaiah 19:21\index{23 Isaiah:ch.  19, v.  21, pg. ``f30!}; Ecclesiastes 5:4-6\index{21
Ecclesiastes:ch.   5, v.   4-6, pg. ``f30!}; Psalm 61:8\index{19 Psalms:ch.  61, v.   8, pg. ``f30!}; Psalm
66:13-14\index{19 Psalms:ch.  66, v.  13-14, pg. ``f30!}.}


\bigskip

22.6\ \ It is not to be made to any creature\index{creatures ``f10!}, but to God alone:\footnote{\ Psalm 76:11\index{19
Psalms:ch.  76, v.  11, pg. ``f30!}; Jeremiah 44:25-26\index{24 Jeremiah:ch.  44, v.  25-26, pg. ``f30!}.} and that it
may be accepted, it is to be made voluntarily\index{Oaths and Vows:voluntary to be accepted ``f10!}, out of
faith\index{Faith:when making oath or vow ``f10!}, and conscience of\index{Believers:desire to obey God ``f10!} duty,
in way of thankfulness\index{Prayer:thanking God ``f10!} for mercy received\index{Mercy:thankfulness for ``f10!}, or
for the obtaining\index{Prayer:faith for obtaining wants ``f10!} of what we want; whereby we more strictly bind
ourselves\index{Oaths and Vows:bind:only to good and just ``f10!} to necessary duties\index{Oaths and Vows:bind:only to
necessary duties ``f10!}; or to other things, so far and so long as they may fitly conduce
thereunto.\footnote{\ Deuteronomy 23:21-23\index{05 Deuteronomy:ch.  23, v.  21-23, pg. ``f30!}; Psalm 50:14\index{19
Psalms:ch.  50, v.  14, pg. ``f30!}; Genesis 28:20\index{01 Genesis:ch.  28, v.  20, pg. ``f30!}{}-22\index{01
Genesis:ch.  28, v.  20-22, pg. ``f30!}; 1 Samuel 1:11\index{09 1 Samuel:ch.   1, v.  11, pg. ``f30!}; Psalm
66:13-14\index{19 Psalms:ch.  66, v.  13-14, pg. ``f30!}; Psalm 132:2-5\index{19 Psalms:ch. 132, v.   2-5, pg.
``f30!}.}


\bigskip

22.7\ \ No man may vow to do anything forbidden\index{Oaths and Vows:not for things:forbidden by Word of God ``f10!} in
the Word of God\index{Scripture:word of God:word of God ``f10!}, or what would hinder any\index{Oaths and
Vows:bind:only to good and just ``f10!} duty therein\index{God's law:commandments:duties:required by God ``f10!}
commanded, or\index{God's law:commandments:duties:not to hinder ``f10!} which is not in his own power\index{Oaths and
Vows:not for things:not in person's power ``f10!}, and for the performance whereof he has no promise of
ability\index{Oaths and Vows:not for things:unable to perform ``f10!} from God.\footnote{\ Acts 23:12, 14\index{44
Acts:ch.  23, v.  12, 14, pg. ``f30!}; Mark 6:26\index{41 Mark:ch.   6, v.  26, pg. ``f30!}; Numbers 30:5, 8,
12-13\index{04 Numbers:ch.  30, v.   5, 8, 12-13, pg. ``f30!}.} In which respects, Popish\index{Popish rituals ``f10!}
monastical\index{Oaths and Vows:not to:be Popish monastical ``f10!} vows of perpetual single\index{Oaths and Vows:not
to:perpetual single life ``f10!} life, professed poverty\index{Oaths and Vows:not to:professed poverty ``f10!}, and
regular obedience\index{Salvation:not by:obedience ``f10!}, are so far from being degrees\index{Oaths and Vows:none of
higher degrees ``f10!} of higher perfection, that\index{perfectionism ``f10!} they are
superstitious\index{Superstitions:superstitions ``f10!} and sinful snares, in which no Christian may entangle
himself.\footnote{\ Matthew 19:11-12\index{40 Matthew:ch.  19, v.  11-12, pg. ``f30!}; 1 Corinthians 7:2, 9\index{46 1
Corinthians:ch.   7, v.   2, 9, pg. ``f30!}; Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}; 1 Peter
4:2\index{60 1 Peter:ch.   4, v.   2, pg. ``f30!}; 1 Corinthians 7:23\index{46 1 Corinthians:ch.   7, v.  23, pg.
``f30!}.}

\section{Chapter 23 -- Of the Civil Magistrate}
\index{Civil magistrate:office of ``f10!}23.1\ \ God, the supreme\index{God's attributes:sovereign lordship ``f10!} Lord
and King\index{Jesus:King ``f10!} of all the world, has ordained\index{Civil magistrate:ordained by God ``f10!} civil
magistrates, to be, under Him, over the people, for His own glory\index{God's attributes:glory:his own ``f10!}, and the
public good: and, to this end, has armed them with the power of the sword\index{Civil magistrate:power of sword
``f10!}, for the defence\index{Civil magistrate:to defend citizens ``f10!} and encouragement of\index{Civil
magistrate:to encourage those doing good ``f10!} them that are good, and for the punishment\index{Punishment:of sin
``f10!} of evil doers\index{Civil magistrate:to punish evil doers ``f10!}.\footnote{\ Romans 13:1\index{45 Romans:ch. 
13, v.   1, pg. ``f30!}{}-4\index{45 Romans:ch.  13, v.   1-4, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v. 
13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}.}


\bigskip

23.2\ \ It is lawful for Christians\index{Civil magistrate:lawful for Christians to execute office ``f10!} to accept and
execute the office of a magistrate, when called thereunto;\footnote{\ Proverbs 8:15-16\index{20 Proverbs:ch.   8, v. 
15-16, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}, 2, 4\index{45 Romans:ch.  13, v.   1, 2,
4, pg. ``f30!}.} in the managing whereof, as they ought especially to maintain piety, justice\index{Civil magistrate:to
maintain justice ``f10!}, and peace, according\index{Civil magistrate:to maintain peace:in society ``f10!} to the
wholesome\index{Commandments of men:wholesome ``f10!} laws of each commonwealth\index{commonwealth
``f10!};\footnote{\ Psalm 2:10-12\index{19 Psalms:ch.   2, v.  10-12, pg. ``f30!}; 1 Timothy 2:2\index{54 1 Timothy:ch.
  2, v.   2, pg. ``f30!}; Psalm 82:3-4\index{19 Psalms:ch.  82, v.   3-4, pg. ``f30!}; 2 Samuel 23:3\index{10 2
Samuel:ch.  23, v.   3, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}.} so for that end, they
may lawfully now\index{war, just, lawful ``f10!}, under the New Testament\index{Scripture:New Testament ``f10!}, wage
war, upon just and necessary occasion.\footnote{\ Luke 3:14\index{42 Luke:ch.   3, v.  14, pg. ``f30!}; Romans
13:4\index{45 Romans:ch.  13, v.   4, pg. ``f30!}; Matthew 8:9-10\index{40 Matthew:ch.   8, v.   9-10, pg. ``f30!};
Acts 10:1-2\index{44 Acts:ch.  10, v.   1-2, pg. ``f30!}; Revelation 17:14, 16\index{66 Revelation:ch.  17, v.  14, 16,
pg. ``f30!}.}

23.3\ \ The civil magistrate may not assume to himself the administration\index{Civil magistrate:not to administer word
or sacraments ``f10!} of the Word and sacraments\index{Sacraments:holy:ordinances ``f10!}, or the
power\index{Church:power delegated by Christ ``f10!} of the keys\index{Kingdom:keys of ``f10!} of the kingdom of
heaven\index{Kingdom:of heaven ``f10!}:\footnote{\ 2 Chronicles 26:18\index{14 2 Chronicles:ch.  26, v.  18, pg.
``f30!} with Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!} and Matthew 16:19\index{40 Matthew:ch.  16, v.
 19, pg. ``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!}{}-29\index{46 1
Corinthians:ch.  12, v.  28-29, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg.
``f30!}{}-12\index{49 Ephesians:ch.   4, v.  11-12, pg. ``f30!}; 1 Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.  
1, pg. ``f30!}{}-2\index{46 1 Corinthians:ch.   4, v.   1-2, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1,
pg. ``f30!}5\index{45 Romans:ch.  10, v.  15, pg. ``f30!}; Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}.}
yet he has authority, and it is his duty, to take order, that unity\index{Civil magistrate:to ensure unity and peace in
Church ``f10!} and peace be\index{Civil magistrate:to maintain peace:in Church ``f10!} preserved in the Church,
that\index{Church:institution ``f10!} the truth of God be\index{Truth:to be kept pure by civil magistrate ``f10!} kept
pure and entire; that\index{Civil magistrate:to keep truth of God pure ``f10!} all blasphemies\index{blasphemy ``f10!}
and heresies\index{Heresy:heresies ``f10!} be suppressed\index{Civil magistrate:to suppress blasphemies and heresies
``f10!}; all corruptions and\index{Worship:false:corruptions in ``f10!} abuses\index{Worship:not to be:abused ``f10!}
in worship and\index{Worship:abuses ``f10!} discipline\index{Church discipline:discipline ``f10!} prevented\index{Civil
magistrate:to keep worship of God pure ``f10!} or reformed; and all the ordinances of God duly\index{Civil
magistrate:to ensure ordinances of God administrated ``f10!} settled, administrated, and observed.\footnote{\ Isaiah
49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; Psalm 122:9\index{19 Psalms:ch. 122, v.   9, pg. ``f30!}; Ezra
7:23, 25-28\index{15 Ezra:ch.   7, v.  23, 25-28, pg. ``f30!}; Leviticus 24:16\index{03 Leviticus:ch.  24, v.  16, pg.
``f30!}; Deuteronomy 13:5-6, 12\index{05 Deuteronomy:ch.  13, v.   5-6, 12, pg. ``f30!}; 1 Kings 18:4\index{11 1
Kings:ch.  18, v.   4, pg. ``f30!}; 1 Chronicles 13:1-9\index{13 1 Chronicles:ch.  13, v.   1-9, pg. ``f30!}; 2 Kings
23:1-26\index{12 2 Kings:ch.  23, v.   1-26, pg. ``f30!}; 2 Chronicles 34:33\index{14 2 Chronicles:ch.  34, v.  33, pg.
``f30!}; 2 Chronicles 15:12-13\index{14 2 Chronicles:ch.  15, v.  12-13, pg. ``f30!}.} For the better effecting
whereof, he has power\index{Civil magistrate:to call synods to consult and advise ``f10!} to call synods\index{Church
government:synods ``f10!}, to be present at them, and to provide that whatsoever is transacted in them be according to
the mind of God.\footnote{\ 2 Chronicles 19:8-11\index{14 2 Chronicles:ch.  19, v.   8-11, pg. ``f30!}; 2 Chronicles
29; 2 Chronicles 30; Matthew 2:4-5\index{40 Matthew:ch.   2, v.   4-5, pg. ``f30!}.}\textsuperscript{,}
\footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription
}}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Clarifications}}\textcolor{black}{).}}


\bigskip

23.4\ \ It is the duty of people\index{Prayer:for magistrates ``f10!} to pray for magistrates\index{Civil
magistrate:pray for ``f10!},\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1
Timothy:ch.   2, v.   1-2, pg. ``f30!}.} to honour their\index{Civil magistrate:to be honoured ``f10!}
persons,\footnote{\ 1 Peter 2:17\index{60 1 Peter:ch.   2, v.  17, pg. ``f30!}.} to pay them tribute\index{Civil
magistrate:to be paid tribute and dues ``f10!} or other dues,\footnote{\ Romans 13:6-7\index{45 Romans:ch.  13, v.  
6-7, pg. ``f30!}.} to obey their\index{Commandments of men:lawful to be obeyed ``f10!} lawful commands, and to be
subject to their authority, for conscience\index{Conscience:for sake of ``f10!}{}' sake.\footnote{\ Romans
13:5\index{45 Romans:ch.  13, v.   5, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}.}
Infidelity\index{Civil magistrate:legal authority even if:infidels ``f10!}, or difference in religion, does\index{Civil
magistrate:legal authority even if:different religion ``f10!} not make void the magistrates' just and legal\index{Civil
magistrate:legal authority even if:unbelievers ``f10!} authority, nor free the people from their due obedience to
them:\footnote{\ 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14,
pg. ``f30!}, 16\index{60 1 Peter:ch.   2, v.  13-14, 16, pg. ``f30!}.} from which ecclesiastical\index{Church
officers:to obey legal laws of civil authorities ``f10!} persons are not exempted,\footnote{\ Romans 13:1\index{45
Romans:ch.  13, v.   1, pg. ``f30!}; 1 Kings 2:35\index{11 1 Kings:ch.   2, v.  35, pg. ``f30!}; Acts 25:9-11\index{44
Acts:ch.  25, v.   9-11, pg. ``f30!}; 2 Peter 2:1, 10-11\index{61 2 Peter:ch.   2, v.   1, 10-11, pg. ``f30!}; Jude
8-11\index{65 Jude:v. 8-11, pg. ``f30!}.} much less has the Pope\index{Pope:of Rome ``f10!} any power and
jurisdiction\index{Civil magistrate:pope has no jurisdiction over ``f10!} over them in their dominions\index{Pope:no
jurisdiction over people in their dominions ``f10!}, or over any of their people; and, least of all, to deprive them of
their dominions, or lives, if he shall judge the\index{Jesus:judge:world ``f10!}m\index{Pope:not to judge magistrates
as heretics ``f10!} to be heretics\index{heretics ``f10!}, or upon any other pretence\index{pretence ``f10!}
whatsoever.\footnote{\ 2 Thessalonians 2:4\index{53 2 Thessalonians:ch.   2, v.   4, pg. ``f30!}; Revelation
13:15-17\index{66 Revelation:ch.  13, v.  15-17, pg. ``f30!}.}

\section{Chapter 24 -- Of Marriage and Divorce}
\index{Marriage:Biblical definition ``f10!}\index{Divorce:divorce ``f10!}24.1\ \ Marriage\index{Marriage:Biblical
definition ``f10!} is to be between one man and one woman: neither is it lawful for any\index{Polygamy:unlawful ``f10!}
man to have more than one wife\index{Marriage:one spouse ``f10!}, nor for any woman to have more than one husband; at
the same time.\footnote{\ Genesis 2:24\index{01 Genesis:ch.   2, v.  24, pg. ``f30!}; Matthew 19:5\index{40 Matthew:ch.
 19, v.   5, pg. ``f30!}{}-6\index{40 Matthew:ch.  19, v.   5-6, pg. ``f30!}; Proverbs 2:1\index{20 Proverbs:ch.   2,
v.   1, pg. ``f30!}7\index{20 Proverbs:ch.   2, v.  17, pg. ``f30!}.}


\bigskip

24.2\ \ Marriage\index{Marriage:Biblical definition ``f10!} was ordained for the mutual help\index{Marriage:mutual help
``f10!} of husband and wife,\footnote{\ Genesis 2:18\index{01 Genesis:ch.   2, v.  18, pg. ``f30!}.} for the increase
of mankind with\index{Marriage:to increase mankind ``f10!} a legitimate\index{Marriage:to produce legitimate issue
(offspring) ``f10!} issue, and of the Church with\index{Church:home of believers ``f10!} an holy
seed\index{Marriage:produce holy seed for Church ``f10!};\footnote{\ Malachi 2:15\index{39 Malachi:ch.   2, v.  15, pg.
``f30!}.} and for preventing of uncleanness\index{Marriage:to prevent uncleanness ``f10!}.\footnote{\ 1 Corinthians
7:2, 9\index{46 1 Corinthians:ch.   7, v.   2, 9, pg. ``f30!}.}


\bigskip

24.3\ \ It is lawful for all\index{Marriage:lawful:for all ``f10!} sorts of people to marry, who are able with judgment
to\index{Judgement:temporal ``f10!} give their consent\index{Marriage:consent of parties ``f10!}.\footnote{\ Hebrews
13:4\index{58 Hebrews:ch.  13, v.   4, pg. ``f30!}; 1 Timothy 4:3\index{54 1 Timothy:ch.   4, v.   3, pg. ``f30!}; 1
Corinthians 7:36-38\index{46 1 Corinthians:ch.   7, v.  36-38, pg. ``f30!}; Genesis 24:57-58\index{01 Genesis:ch.  24,
v.  57-58, pg. ``f30!}.} Yet is it the duty of Christians to marry only\index{Marriage:Christians:only in the Lord
``f10!} in the Lord:\footnote{\ 1 Corinthians 7:39\index{46 1 Corinthians:ch.   7, v.  39, pg. ``f30!}.} and therefore
such as profess the true\index{Profession:of true religion ``f10!} reformed religion should not marry
with\index{Marriage:Christians:not with infidels or papists ``f10!} infidels, papists\index{papists ``f10!}, or other
idolaters\index{Idolatry:idolatry ``f10!}: neither should such as are godly be unequally yoked\index{Marriage:not
unequally yoked ``f10!}, by marrying with such as are notoriously\index{Marriage:not with notoriously wicked ``f10!}
wicked in their life, or maintain damnable heresies\index{Heresy:heresies ``f10!}.\footnote{\ Genesis 34:14\index{01
Genesis:ch.  34, v.  14, pg. ``f30!}; Exodus 34:1\index{02 Exodus:ch.  34, v.   1, pg. ``f30!}6\index{02 Exodus:ch. 
34, v.  16, pg. ``f30!}; Deuteronomy 7:3-4\index{05 Deuteronomy:ch.   7, v.   3-4, pg. ``f30!}; I Kings 11:4; Nehemiah
13:25\index{16 Nehemiah:ch.  13, v.  25, pg. ``f30!}{}-26, 27\index{16 Nehemiah:ch.  13, v.  25-26, 27, pg. ``f30!};
Malachi 2:11-12\index{39 Malachi:ch.   2, v.  11-12, pg. ``f30!}; 2 Corinthians 6:14\index{47 2 Corinthians:ch.   6, v.
 14, pg. ``f30!}.} 


\bigskip

24.4\ \ Marriage\index{Marriage:Biblical definition ``f10!} ought not to be within the degrees of
consanguinity\index{Consanguinity:consanguinity ``f10!} or affinity forbidden\index{Consanguinity:forbidden affinity
``f10!} by the Word;\footnote{\ Leviticus 18; 1 Corinthians 5:1\index{46 1 Corinthians:ch.   5, v.   1, pg. ``f30!};
Amos 2:7\index{30 Amos:ch.   2, v.   7, pg. ``f30!}.} nor can such incestuous marriages\index{Marriage:unlawful ``f10!}
ever be made lawful by any\index{Consanguinity:incestuous marriage not lawful ``f10!} law of man or consent of
parties\index{Marriage:consent of parties ``f10!}, so as those persons may live together as man and
wife\index{Marriage:live together as man and wife ``f10!}.\footnote{\ Mark 6:18\index{41 Mark:ch.   6, v.  18, pg.
``f30!}; Leviticus 18:24-28\index{03 Leviticus:ch.  18, v.  24-28, pg. ``f30!}.} The man may not marry
any\index{Marriage:not near kindred of wife ``f10!} of his wife's kindred\index{Consanguinity:near kindred ``f10!}
nearer in blood\index{Consanguinity:nearness of blood ``f10!} than he may of his own; nor the woman of her husband's
kindred nearer in blood than of her own.\footnote{\ Leviticus 20:19-21\index{03 Leviticus:ch.  20, v.  19-21, pg.
``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning
Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).} }


\bigskip

24.5\ \ Adultery\index{Sexual immorality:adultery ``f10!} or fornication\index{Sexual immorality:fornication ``f10!}
committed after a contract\index{Contracts:marriage ``f10!}, being detected\index{Marriage:valid reasons for ending
``f10!} before marriage, gives just occasion to the innocent party to dissolve that contract.\footnote{\ Matthew
1:18-20\index{40 Matthew:ch.   1, v.  18-20, pg. ``f30!}.} In the case of adultery after marriage, it is lawful for the
innocent party to sue out a divorce\index{Divorce:divorce ``f10!}:\footnote{\ Matthew 5:31-32\index{40 Matthew:ch.   5,
v.  31-32, pg. ``f30!}.} and, after the divorce, to marry another\index{Marriage:remarriage permitted for innocent
party after divorce ``f10!}, as if the offending\index{Divorce:offending treated as dead ``f10!} party were
dead\index{Death:divorced marriage partner viewed as dead ``f10!}.\footnote{\ Matthew 19:9\index{40 Matthew:ch.  19, v.
  9, pg. ``f30!}; Romans 7:2-3\index{45 Romans:ch.   7, v.   2-3, pg. ``f30!}.}


\bigskip

24.6\ \ Although the corruption of man\index{Sin:original:corrupt nature:whole ``f10!} be such as is apt to study
arguments unduly to put asunder\index{Marriage:not to be broken, except if adultery ``f10!} those whom God has joined
together in marriage: yet nothing but adultery\index{Sexual immorality:adultery ``f10!}, or such wilful
desertion\index{Marriage:desertion ``f10!} as can no way be remedied by the Church or\index{Church:institution ``f10!}
civil magistrate, is\index{Civil magistrate:role in divorce cases ``f10!} cause sufficient of
dissolving\index{Marriage:valid reasons for ending ``f10!} the bond of marriage:\footnote{\ Matthew 19:8-9\index{40
Matthew:ch.  19, v.   8-9, pg. ``f30!}; 1 Corinthians 7:15\index{46 1 Corinthians:ch.   7, v.  15, pg. ``f30!}; Matthew
19:6\index{40 Matthew:ch.  19, v.   6, pg. ``f30!}.} wherein, a public and orderly course of proceeding is to be
observed; and the persons concerned in it not left to their own wills and\index{Causation:will of creatures:not to be
exercised to dissolve marriage ``f10!} discretion, in their own case.\footnote{\ Deuteronomy 24:1-4\index{05
Deuteronomy:ch.  24, v.   1-4, pg. ``f30!}.}

\section{Chapter 25 -- Of the Church}
\index{Church:institution ``f10!}25.1\ \ The catholic\index{Church:catholic ``f10!} or universal
Church\index{Church:universal ``f10!} which is invisible, consists\index{Church:invisible ``f10!} of the whole number
of the elect\index{Election:elect of God ``f10!}, that have been, are, or shall be gathered into one, under Christ the
Head\index{Church:Christ as head ``f10!} thereof; and is the spouse\index{Church:Christ's spouse ``f10!}, the body, the
fulness of Him\index{Jesus:fulness of all things ``f10!} that fills\index{Jesus:fills all in all ``f10!} all in
all.\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}0, 22-23\index{49 Ephesians:ch.   1, v. 
10, 22-23, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}3\index{49 Ephesians:ch.   5, v. 
23, pg. ``f30!}, 27, 32\index{49 Ephesians:ch.   5, v.  23, 27, 32, pg. ``f30!}; Colossians 1:18\index{51
Colossians:ch.   1, v.  18, pg. ``f30!}.}


\bigskip

25.2\ \ The visible Church\index{Church:visible ``f10!}, which is also catholic\index{Church:catholic ``f10!} or
universal under\index{Church:universal ``f10!} the Gospel\index{Gospel:gospel ``f10!} (not confined to one nation
as\index{Gospel:to all nations ``f10!} before under the law), consists of all those throughout the world that profess
the true\index{Profession:of true religion ``f10!} religion;\footnote{\ 1 Corinthians 1:2\index{46 1 Corinthians:ch.  
1, v.   2, pg. ``f30!}; 1 Corinthians 12:12-13\index{46 1 Corinthians:ch.  12, v.  12-13, pg. ``f30!}; Psalm
2:8\index{19 Psalms:ch.   2, v.   8, pg. ``f30!}; Revelation 7:9\index{66 Revelation:ch.   7, v.   9, pg. ``f30!};
Romans 15:9, 10-12\index{45 Romans:ch.  15, v.   9, 10-12, pg. ``f30!}.} and of their children:\footnote{\ 1
Corinthians 7:14\index{46 1 Corinthians:ch.   7, v.  14, pg. ``f30!}; Acts 2:39\index{44 Acts:ch.   2, v.  39, pg.
``f30!}; Ezekiel 16:20-21\index{26 Ezekiel:ch.  16, v.  20-21, pg. ``f30!}; Romans 11:16\index{45 Romans:ch.  11, v. 
16, pg. ``f30!}; Genesis 3:15\index{01 Genesis:ch.   3, v.  15, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v. 
 7, pg. ``f30!}.} and is the kingdom of the\index{Kingdom:of Christ ``f10!} Lord Jesus Christ,\footnote{\ Matthew
13:47\index{40 Matthew:ch.  13, v.  47, pg. ``f30!}; Isaiah 9:7\index{23 Isaiah:ch.   9, v.   7, pg. ``f30!}.} the
house and\index{Church:home of believers ``f10!} family of\index{Church:family of God ``f10!} God,\footnote{\ Ephesians
2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}9\index{49 Ephesians:ch.   2, v.  19, pg. ``f30!}; Ephesians
3:15\index{49 Ephesians:ch.   3, v.  15, pg. ``f30!}.} out of which there is no ordinary\index{Church:ordinary means of
salvation ``f10!} possibility of salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Acts
2:47\index{44 Acts:ch.   2, v.  47, pg. ``f30!}.}


\bigskip

25.3\ \ Unto this catholic\index{Church:catholic ``f10!} visible Church\index{Church:visible ``f10!} Christ has given
the ministry\index{Church:given the ministry of God ``f10!}, oracles\index{Church:endowed with oracles of God ``f10!},
and ordinances of God, for\index{Church:furnished with:ordinances of God ``f10!} the gathering\index{Church:universal
``f10!} and perfecting of\index{Church:perfecting of saints ``f10!} the saints\index{Believers:saints ``f10!}, in this
life, to the end of the world: and does by His own presence and Spirit\index{Holy Spirit:works in the Church ``f10!},
according to His promise\index{Church:promised presence of God ``f10!}, make them effectual
thereunto\index{Worship:ordinances:made effectual for salvation ``f10!}.\footnote{\ 1 Corinthians 12:28\index{46 1
Corinthians:ch.  12, v.  28, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-13\index{49
Ephesians:ch.   4, v.  11-13, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40
Matthew:ch.  28, v.  19-20, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}1\index{23 Isaiah:ch. 
59, v.  21, pg. ``f30!}.}


\bigskip

25.4\ \ This catholic\index{Church:catholic ``f10!} Church has been sometimes more, sometimes less
visible.\footnote{\ Romans 11:3-4\index{45 Romans:ch.  11, v.   3-4, pg. ``f30!}; Revelation 12:6, 14\index{66
Revelation:ch.  12, v.   6, 14, pg. ``f30!}.} And particular Churches, which\index{Church:mix of truth and error
``f10!} are members thereof\index{Church members ``f10!}, are more or less pure\index{Church:true:marks ``f10!},
according as the doctrine of the\index{Doctrine:Gospel ``f10!} Gospel\index{Gospel:gospel ``f10!} is taught
and\index{Gospel:to be taught and embraced ``f10!} embraced, ordinances administered, and\index{Church:true:ordinances
administered ``f10!} public worship\index{Worship:public ``f10!} performed more or less purely in
them.\footnote{\ Revelation 2; Revelation 3; 1 Corinthians 5:6-7\index{46 1 Corinthians:ch.   5, v.   6-7, pg.
``f30!}.}


\bigskip

25.5\ \ The purest Churches under\index{Church:mix of truth and error ``f10!} heaven are subject both to mixture and
error:\footnote{\ 1 Corinthians 13:12\index{46 1 Corinthians:ch.  13, v.  12, pg. ``f30!}; Revelation 2; Revelation 3;
Matthew 13:24-30, 47\index{40 Matthew:ch.  13, v.  24-30, 47, pg. ``f30!}.} and some have so
degenerated\index{Church:degenerated churches ``f10!}, as to become no Churches of\index{Church:false, synagogues of
Satan ``f10!} Christ, but synagogues of Satan.\footnote{\ Revelation 18:2\index{66 Revelation:ch.  18, v.   2, pg.
``f30!}; Romans 11:18-22\index{45 Romans:ch.  11, v.  18-22, pg. ``f30!}.} Nevertheless, there shall be always a
Church\index{Church:indestructible ``f10!} on earth, to worship God\index{Worship:always be a Church to worship ``f10!}
according to His will.\footnote{\ Matthew 16:18\index{40 Matthew:ch.  16, v.  18, pg. ``f30!}; Psalm 72:17\index{19
Psalms:ch.  72, v.  17, pg. ``f30!}; Psalm 102:28\index{19 Psalms:ch. 102, v.  28, pg. ``f30!}; Matthew 28:19\index{40
Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}


\bigskip

25.6\ \ There is no other head of the\index{Church:Christ as head ``f10!} Church, but the Lord Jesus
Christ;\footnote{\ Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg. ``f30!}; Ephesians 1:22\index{49
Ephesians:ch.   1, v.  22, pg. ``f30!}.} nor can the Pope\index{Pope:of Rome ``f10!} of Rome, in any sense, be head
thereof\index{Pope:not head of the Church ``f10!}; but is that Antichrist\index{Antichrist (Pope) ``f10!}, that man of
sin, and son of perdition\index{Pope:son of perdition ``f10!}, that exalts\index{Pope:exalts self ``f10!} himself, in
the Church, against\index{Church:institution ``f10!} Christ and all that is called God.\footnote{\ Matthew
23:8-10\index{40 Matthew:ch.  23, v.   8-10, pg. ``f30!}; 2 Thessalonians 2:3-4, 8-9\index{53 2 Thessalonians:ch.   2,
v.   3-4, 8-9, pg. ``f30!}; Revelation 13:6\index{66 Revelation:ch.  13, v.   6, pg. ``f30!}.}\textsuperscript{,}
\footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription
}}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).}}

\section{Chapter 26 -- Of the Communion of the Saints}
\index{Communion:of saints ``f10!}\index{Believers:saints ``f10!}26.1\ \ All saints\index{Believers:saints ``f10!}, that
are united to Jesus\index{Believers:union with Christ:mystical ``f10!} Christ their Head\index{Church:Christ as head
``f10!} by His Spirit and by faith\index{Faith:worked in believers by the Holy Spirit ``f10!}, have fellowship
with\index{Fellowship:with Christ ``f10!} Him in His grace\index{Grace:fellowship with Jesus in ``f10!}, sufferings,
death\index{Jesus:death:sufferings and death ``f10!}, resurrection, and glory:\index{Resurrection:believers:into
glory:with Christ ``f10!}\footnote{\ John 1:3\index{43 John:ch.   1, v.   3, pg. ``f30!}; Ephesians 3:16\index{49
Ephesians:ch.   3, v.  16, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  16-19, pg. ``f30!}; John 1:1\index{43
John:ch.   1, v.   1, pg. ``f30!}6\index{43 John:ch.   1, v.  16, pg. ``f30!}; Ephesians 2:5\index{49 Ephesians:ch.  
2, v.   5, pg. ``f30!}{}-6\index{49 Ephesians:ch.   2, v.   5-6, pg. ``f30!}; Philippians 3:10\index{50 Philippians:ch.
  3, v.  10, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}{}-6\index{45 Romans:ch.   6, v.  
5-6, pg. ``f30!}; 2 Timothy 2:12\index{55 2 Timothy:ch.   2, v.  12, pg. ``f30!}.} and, being united to
one\index{Believers:united to one another ``f10!} another in love, they\index{Love for others:united ``f10!} have
communion\index{Communion:of saints ``f10!} in each other's gifts and graces,\index{Love for others:acknowledge gifts
and graces of others ``f10!}\footnote{\ Ephesians 4:15-16\index{49 Ephesians:ch.   4, v.  15-16, pg. ``f30!}; 1
Corinthians 12:7\index{46 1 Corinthians:ch.  12, v.   7, pg. ``f30!}; 1 Corinthians 3:2\index{46 1 Corinthians:ch.   3,
v.   2, pg. ``f30!}1-23\index{46 1 Corinthians:ch.   3, v.  21-23, pg. ``f30!}; Colossians 2:19\index{51 Colossians:ch.
  2, v.  19, pg. ``f30!}.} and are obliged to the performance of such duties\index{God's law:commandments:duties:to
others ``f10!}, public and private, as do conduce to their mutual good\index{Good works:for mutual good ``f10!}, both
in the inward\index{Love for others:inward thoughts further others' good ``f10!} and outward man\index{Love for
others:for saints ``f10!}.\footnote{\ 1 Thessalonians 5:11\index{52 1 Thessalonians:ch.   5, v.  11, pg. ``f30!},
14\index{52 1 Thessalonians:ch.   5, v.  11, 14, pg. ``f30!}; Romans 1:11-12, 14\index{45 Romans:ch.   1, v.  11-12,
14, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}6\index{62 1 John:ch.   3, v.  16, pg.
``f30!}{}-18\index{62 1 John:ch.   3, v.  16-18, pg. ``f30!}; Galatians 6:10\index{48 Galatians:ch.   6, v.  10, pg.
``f30!}.}


\bigskip

26.2\ \ Saints\index{Believers:saints ``f10!} by profession are\index{Faith:profession of ``f10!} bound to maintain a
holy\index{Worship:holy fellowship to be maintained ``f10!} fellowship and\index{Fellowship:with fellow believers
``f10!} communion in the\index{Communion:in worship ``f10!} worship of God; and in performing such other spiritual
services\index{Good works:spiritual services ``f10!} as tend to their mutual edification\index{Love for
others:mutual:edification ``f10!};\footnote{\ Hebrews 10:24-25\index{58 Hebrews:ch.  10, v.  24-25, pg. ``f30!}; Acts
2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}, 46\index{44 Acts:ch.   2, v.  42, 46, pg. ``f30!}; Isaiah 2:3\index{23
Isaiah:ch.   2, v.   3, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1
Corinthians:ch.  11, v.  20, pg. ``f30!}.} as also in relieving\index{Love for others:mutual:relieving in outward
things ``f10!} each other in outward things\index{Love for others:for saints ``f10!}, according to their several
abilities, and necessities. Which communion, as\index{Communion:with God ``f10!} God offers\index{Church:communion with
all who call upon Jesus ``f10!} opportunity\index{Communion:of saints ``f10!}, is to be extended unto all those who, in
every place, call upon the name of the Lord Jesus.\footnote{\ Acts 2:44-45\index{44 Acts:ch.   2, v.  44-45, pg.
``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}7\index{62 1 John:ch.   3, v.  17, pg. ``f30!}; 2
Corinthians 8; 2 Corinthians 9; Acts 11:29-30\index{44 Acts:ch.  11, v.  29-30, pg. ``f30!}.}


\bigskip

26.3\ \ This communion which\index{Communion:with Christ ``f10!} the saints\index{Believers:saints ``f10!} have with
Christ, does not make them, in any wise, partakers\index{Believers:union with Christ:not partakers of Godhead ``f10!}
of the substance of His Godhead\index{God's attributes:Godhead ``f10!}; or to be equal with Christ\index{Communion:with
Christ, does not make believers equal to him ``f10!}, in any respect: either of which to affirm is impious and
blasphemous\index{blasphemy ``f10!}.\footnote{\ Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg.
``f30!}{}-19\index{51 Colossians:ch.   1, v.  18-19, pg. ``f30!}; 1 Corinthians 8:6\index{46 1 Corinthians:ch.   8, v. 
 6, pg. ``f30!}; Isaiah 42:8\index{23 Isaiah:ch.  42, v.   8, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.
  1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}{}-16\index{54 1 Timothy:ch.   6, v.  15-16, pg.
``f30!}; Psalm 45:7\index{19 Psalms:ch.  45, v.   7, pg. ``f30!} with Hebrews 1:8\index{58 Hebrews:ch.   1, v.   8, pg.
``f30!}{}-9\index{58 Hebrews:ch.   1, v.   8-9, pg. ``f30!}.} Nor does their communion one\index{Communion:of saints
``f10!} with another, as saints, take away, or infringe\index{property, rights not to be infringed ``f10!} the title or
propriety which each man has in his goods\index{Theft:infringe on property of others ``f10!} and
possessions.\footnote{\ Exodus 20:15\index{02 Exodus:ch.  20, v.  15, pg. ``f30!}; Ephesians 4:28\index{49
Ephesians:ch.   4, v.  28, pg. ``f30!}; Acts 5:4\index{44 Acts:ch.   5, v.   4, pg. ``f30!}.}

\section{Chapter 27 -- Of the Sacraments}
\index{Sacraments:holy:ordinances ``f10!}27.1\ \ Sacraments\index{Sacraments:holy:ordinances ``f10!} are
holy\index{Sacraments:signs and seals:holy ``f10!} signs\index{Sacraments:holy:signs ``f10!} and
seals\index{Sacraments:covenant seals ``f10!} of the covenant of grace\index{Covenant:grace:grace
``f10!},\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v.
  7, pg. ``f30!}, 10\index{01 Genesis:ch.  17, v.   7, 10, pg. ``f30!}.} immediately instituted by
God\index{Sacraments:instituted by:God ``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!};
1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg.
``f30!}.} to represent Christ\index{Sacraments:represent Christ ``f10!} and His benefits; and
to\index{Sacraments:benefits:worthy believers receive ``f10!} confirm\index{Sacraments:confirm union with Christ
``f10!} our interest in Him\index{Believers:interest (share) in Christ ``f10!};\footnote{\ 1 Corinthians 10:16\index{46
1 Corinthians:ch.  10, v.  16, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}5\index{46 1 Corinthians:ch.  11, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  25-26, pg.
``f30!}; Galatians 3:17\index{48 Galatians:ch.   3, v.  17, pg. ``f30!}.} as also, to put a
visible\index{Sacraments:signs and seals:visible difference ``f10!} difference between those that belong unto the
Church, and\index{Church:institution ``f10!} the rest of the world;\footnote{\ Romans 15:8\index{45 Romans:ch.  15, v. 
 8, pg. ``f30!}; Exodus 12:48\index{02 Exodus:ch.  12, v.  48, pg. ``f30!}; Genesis 34:14\index{01 Genesis:ch.  34, v. 
14, pg. ``f30!}.} and solemnly to\index{Sacraments:solemn engagement to God's service ``f10!}
engage\index{Sacraments:engage participants to service of God ``f10!} them to the service of\index{Love toward God:in
service ``f10!} God in Christ, according to His Word.\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6, v.   3-4, pg.
``f30!}; 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}, 21\index{46 1 Corinthians:ch.  10, v.
 16, 21, pg. ``f30!}.}


\bigskip

27.2\ \ There is in every sacrament\index{Sacraments:holy:ordinances ``f10!} a spiritual
relation\index{Sacraments:spiritual relation between sign and signified ``f10!}, or sacramental\index{Lord's
Supper:elements:union between sign and signified ``f10!} union, between\index{Sacraments:union between sign and
signified ``f10!} the sign and the thing signified: whence it comes to pass, that the names and effects of the one are
attributed to the other.\footnote{\ Genesis 17:1\index{01 Genesis:ch.  17, v.   1, pg. ``f30!}0\index{01 Genesis:ch. 
17, v.  10, pg. ``f30!}; Matthew 26:27-28\index{40 Matthew:ch.  26, v.  27-28, pg. ``f30!}; Titus 3:5\index{56
Titus:ch.   3, v.   5, pg. ``f30!}.}


\bigskip

27.3\ \ The grace which\index{Grace:exhibited in sacraments ``f10!} is exhibited in\index{Sacraments:grace:exhibited
``f10!} or by the sacraments\index{Sacraments:holy:ordinances ``f10!} rightly used\index{Sacraments:grace:conferred
``f10!}, is not conferred by any\index{Sacraments:grace:not conferred by elements ``f10!} power in them; neither does
the efficacy of a\index{Sacraments:spiritual efficacy ``f10!} sacrament depend upon the piety or\index{Lord's
Supper:efficacy, not dependent on piety or intention of person administering ``f10!} intention of him that does
administer it:\footnote{\ Romans 2:28-29\index{45 Romans:ch.   2, v.  28-29, pg. ``f30!}; 1 Peter 3:21\index{60 1
Peter:ch.   3, v.  21, pg. ``f30!}.} but upon the work of the Spirit,\index{Holy Spirit:Spirit and the word
``f10!}\footnote{\ Matthew 3:11\index{40 Matthew:ch.   3, v.  11, pg. ``f30!}; 1 Corinthians 12:13\index{46 1
Corinthians:ch.  12, v.  13, pg. ``f30!}.} and the word of institution\index{Sacraments:instituted by:Christ ``f10!},
which contains, together with a precept authorizing the use thereof, a promise of benefit
to\index{Sacraments:benefits:worthy believers receive ``f10!} worthy receivers.\footnote{\ Matthew 26:27-28\index{40
Matthew:ch.  26, v.  27-28, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40
Matthew:ch.  28, v.  19-20, pg. ``f30!}.}


\bigskip

27.4\ \ There are only two sacraments\index{Sacraments:holy:ordinances ``f10!} ordained by\index{Sacraments:ordained by
Jesus ``f10!} Christ our Lord in the Gospel\index{Gospel:gospel ``f10!}; that is to say, Baptism
and\index{Baptism:sacrament ``f10!} the Supper\index{Lord's Supper:sacrament ``f10!} of the Lord: neither of which may
be dispensed\index{Sacraments:to be:dispensed by minister of the Word ``f10!} by any but by a minister\index{Church
officers:minsters ``f10!} of the Word lawfully\index{Preachers:to be:lawfully ordained ``f10!}
ordained.\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}, 23\index{46 1
Corinthians:ch.  11, v.  20, 23, pg. ``f30!}; 1 Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.   1, pg. ``f30!};
Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}.}


\bigskip

27.5\ \ The sacraments\index{Sacraments:holy:ordinances ``f10!} of the Old Testament\index{Scripture:Old Testament
``f10!}, in regard to the spiritual things\index{Sacraments:spiritual things signified, same in both Testaments ``f10!}
thereby signified and exhibited, were, for substance\index{Sacraments:OT and NT same substance ``f10!}, the same with
those of the New.\footnote{\ 1 Corinthians 10:1-4\index{46 1 Corinthians:ch.  10, v.   1-4, pg. ``f30!}.}

\section{Chapter 28 -- Of Baptism}
\index{Baptism:sacrament ``f10!}28.1\ \ Baptism is a\index{Baptism:sacrament ``f10!}
sacrament\index{Sacraments:holy:ordinances ``f10!} of the New Testament\index{Scripture:New Testament ``f10!}, ordained
by\index{Sacraments:ordained by Jesus ``f10!} Jesus Christ,\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19,
pg. ``f30!}.} not only for the solemn admission\index{Baptism:admission into visible Church ``f10!} of the party
baptized into the visible Church\index{Church:visible ``f10!};\footnote{\ 1 Corinthians 12:13\index{46 1
Corinthians:ch.  12, v.  13, pg. ``f30!}.} but also, to be unto him a sign\index{Baptism:sign of salvation ``f10!} and
seal of\index{Baptism:seal of salvationof ``f10!} the covenant of grace\index{Covenant:grace:grace
``f10!},\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!} with Colossians 2:11-12\index{51
Colossians:ch.   2, v.  11-12, pg. ``f30!}.} of his ingrafting\index{Believers:ingrafted into Christ ``f10!} into
Christ,\footnote{\ Galatians 3:27\index{48 Galatians:ch.   3, v.  27, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6,
v.   5, pg. ``f30!}.} of regeneration\index{Regeneration:regeneration ``f10!},\footnote{\ Titus 3:5\index{56 Titus:ch. 
 3, v.   5, pg. ``f30!}.} of remission\index{Redemption:remission of sins ``f10!} of sins,\footnote{\ Mark 1:4\index{41
Mark:ch.   1, v.   4, pg. ``f30!}.} and of his giving up unto God through Jesus Christ, to walk in the
newness\index{Baptism:walk in newness of life ``f10!} of life.\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6, v.  
3-4, pg. ``f30!}.} Which sacrament is, by Christ's own appointment, to be\index{Baptism:appointed by Christ for Church
``f10!} continued in His Church\index{Church:Christ's ``f10!} until the end of the world.\footnote{\ Matthew
28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}


\bigskip

28.2\ \ The outward element to be used in this sacrament\index{Sacraments:holy:ordinances ``f10!} is
water\index{Baptism:to use water ``f10!}, wherewith the party is to be baptized\index{Baptism:in name of Trinity
``f10!}, in the name of the Father\index{Father:God the ``f10!}, and of the Son\index{Jesus:Son of God ``f10!}, and of
the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, by a minister\index{Church officers:minsters ``f10!} of the
Gospel\index{Gospel:gospel ``f10!}, lawfully call\index{Civil magistrate:to call synods to consult and advise
``f10!}ed\index{Preachers:to be:lawfully called ``f10!} thereunto.\footnote{\ Matthew 3:11\index{40 Matthew:ch.   3, v.
 11, pg. ``f30!}; John 1:3\index{43 John:ch.   1, v.   3, pg. ``f30!}3\index{43 John:ch.   1, v.  33, pg. ``f30!};
Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}


\bigskip

28.3\ \ Dipping\index{Baptism:modes:dipping ``f10!} of the person into the water is\index{Baptism:to use water ``f10!}
not necessary; but Baptism is rightly administered\index{Baptism:modes:valid ``f10!} by
pouring\index{Baptism:modes:pouring ``f10!} or sprinkling\index{Baptism:modes:sprinkling ``f10!} water upon the
person.\footnote{\ Hebrews 9:10, 19-22\index{58 Hebrews:ch.   9, v.  10, 19-22, pg. ``f30!}; Acts 2:41\index{44
Acts:ch.   2, v.  41, pg. ``f30!}; Acts 16:33\index{44 Acts:ch.  16, v.  33, pg. ``f30!}; Mark 7:4\index{41 Mark:ch.  
7, v.   4, pg. ``f30!}.}

28.4\ \ Not only those that do actually profess faith\index{Faith:profession of ``f10!} in and obedience\index{God's
law:commandments:obedience:required ``f10!} unto Christ,\footnote{\ Mark 16:15\index{41 Mark:ch.  16, v.  15, pg.
``f30!}{}-16\index{41 Mark:ch.  16, v.  15-16, pg. ``f30!}; Acts 8:37-38\index{44 Acts:ch.   8, v.  37-38, pg.
``f30!}.} but also the infants of one\index{Infants:of believing parents, to be baptized ``f10!} or both
believing\index{Baptism:infants of believing parents ``f10!} parents, are to be baptized\index{Baptism:infants
``f10!}.\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 9\index{01 Genesis:ch.  17, v.   7, 9,
pg. ``f30!}{}-10\index{01 Genesis:ch.  17, v.   7, 9-10, pg. ``f30!} with Galatians 3:9, 14\index{48 Galatians:ch.   3,
v.   9, 14, pg. ``f30!}; Colossians 2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg. ``f30!}; Acts 2:38\index{44
Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!} and Romans 4:11\index{45
Romans:ch.   4, v.  11, pg. ``f30!}{}-12\index{45 Romans:ch.   4, v.  11-12, pg. ``f30!}; 1 Corinthians 7:14\index{46 1
Corinthians:ch.   7, v.  14, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; Mark
10:13-16\index{41 Mark:ch.  10, v.  13-16, pg. ``f30!}; Luke 18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}5\index{42
Luke:ch.  18, v.  15, pg. ``f30!}.}


\bigskip

28.5\ \ Although it be a great sin to contemn\index{Baptism:not to be treated with contempt ``f10!} or neglect
this\index{Baptism:not to neglect ``f10!} ordinance,\footnote{\ Luke 7:30\index{42 Luke:ch.   7, v.  30, pg. ``f30!}
with Exodus 4:24-26\index{02 Exodus:ch.   4, v.  24-26, pg. ``f30!}.} yet grace\index{Grace:exhibited in sacraments
``f10!} and salvation\index{Salvation:belief in Jesus by faith ``f10!} are not so
inseparably\index{Sanctification:inseparable from:grace ``f10!} annexed unto it, as that no person can be
regenerated\index{Regeneration:regeneration ``f10!} or saved without\index{Salvation:not without grace ``f10!}
it;\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Acts 10:2\index{44 Acts:ch.  10, v.   2, pg.
``f30!}, 4, 22, 31, 45, 47\index{44 Acts:ch.  10, v.   2, 4, 22, 31, 45, 47, pg. ``f30!}.} or, that all that are
baptized\index{Baptism:not all baptized are regenerated ``f10!} are undoubtedly regenerated.\footnote{\ Acts 8:13,
23\index{44 Acts:ch.   8, v.  13, 23, pg. ``f30!}.}


\bigskip

28.6\ \ The efficacy\index{Baptism:spiritual efficacy ``f10!} of Baptism is not tied to that moment
of\index{Baptism:efficacy not time dependent ``f10!} time wherein it is administered;\footnote{\ John 3:5\index{43
John:ch.   3, v.   5, pg. ``f30!}, 8\index{43 John:ch.   3, v.   5, 8, pg. ``f30!}.} yet notwithstanding, by the right
use of this ordinance\index{Baptism:ordinance, to be used rightly ``f10!}, the grace promised is\index{Baptism:means of
grace ``f10!} not only offered, but really exhibited and\index{Baptism:grace:exhibited ``f10!} conferred,
by\index{Baptism:grace:conferred by Holy Spirit ``f10!} the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, to such
(whether of age or infants) as\index{Infants:of believing parents, to be baptized ``f10!} that grace\index{Grace:in
believers:by Holy Spirit ``f10!} belongs unto, according to the counsel of God's\index{Causation:God's own will ``f10!}
own will, in His appointed time.\footnote{\ Galatians 3:27\index{48 Galatians:ch.   3, v.  27, pg. ``f30!}; Titus
3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg.
``f30!}5-26\index{49 Ephesians:ch.   5, v.  25-26, pg. ``f30!}; Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!},
41\index{44 Acts:ch.   2, v.  38, 41, pg. ``f30!}.}


\bigskip

28.7\ \ The sacrament\index{Sacraments:holy:ordinances ``f10!} of Baptism is but\index{Baptism:to be administered once
``f10!} once to be administered unto any person.\footnote{\ Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.}

\section{Chapter 29 -- Of the Lord's Supper}
\index{Lord's Supper:sacrament ``f10!}29.1\ \ Our Lord Jesus, in the night wherein He was betrayed\index{Jesus:betrayed
``f10!}, instituted the\index{Sacraments:instituted by:Christ ``f10!} sacrament\index{Sacraments:holy:ordinances
``f10!} of His body and blood\index{Lord's Supper:body and blood of Jesus ``f10!}, called the Lord's
Supper\index{Lord's Supper:sacrament ``f10!}, to be observed in\index{Lord's Supper:to be observed to end of world
``f10!} His Church\index{Church:Christ's ``f10!}, unto the end of the world, for the perpetual remembrance\index{Lord's
Supper:remembrance of Christ's sacrifice ``f10!} of the sacrifice of Himself\index{Jesus:sacrifice:of himself ``f10!}
in His death\index{Jesus:death:death ``f10!}; the sealing\index{Lord's Supper:benefits sealed ``f10!} all benefits
thereof unto\index{Lord's Supper:benefits believers receive ``f10!} true believers, their spiritual
nourishment\index{Lord's Supper:spiritual feeding:nourishment ``f10!} and growth in Him\index{Sanctification:spiritual
growth:in Christ ``f10!}, their further engagement in and\index{God's law:commandments:obedience:required ``f10!} to
all duties which\index{God's law:commandments:duties:required by God ``f10!} they owe unto Him; and to be a bond and
pledge\index{Lord's Supper:pledge of communion with Christ ``f10!} of their communion with Him,
and\index{Communion:with Christ ``f10!} with each other, as members of His\index{Church members ``f10!}
mystical\index{Believers:union with Christ:mystical ``f10!} body.\footnote{\ 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1
Corinthians:ch.  11, v.  23-26, pg. ``f30!}; 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg.
``f30!}{}-17\index{46 1 Corinthians:ch.  10, v.  16-17, pg. ``f30!}, 21\index{46 1 Corinthians:ch.  10, v.  16-17, 21,
pg. ``f30!}; 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.}


\bigskip

29.2\ \ In this sacrament\index{Sacraments:holy:ordinances ``f10!}, Christ is not offered up to\index{Lord's Supper:not
a sacrifice:of Jesus ``f10!} His Father\index{Father:God the ``f10!}; nor any real sacrifice made at all for
remission\index{Redemption:remission of sins ``f10!} of sins of the quick or dead\index{Lord's Supper:not a
sacrifice:for quick or dead ``f10!};\footnote{\ Hebrews 9:22\index{58 Hebrews:ch.   9, v.  22, pg. ``f30!}, 25-26,
28\index{58 Hebrews:ch.   9, v.  22, 25-26, 28, pg. ``f30!}.} but only a commemoration\index{Lord's
Supper:commemoration ``f10!} of that one offering up of Himself\index{Jesus:sacrifice:of himself ``f10!}, by Himself,
upon the cross\index{Jesus:death:on cross ``f10!}, once for all: and a spiritual oblation\index{Lord's
Supper:remembrance of Christ's sacrifice ``f10!} of all possible praise unto\index{Worship:praising God ``f10!} God for
the same:\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.
 11, v.  24, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  24-26, pg. ``f30!}; Matthew 26:26\index{40
Matthew:ch.  26, v.  26, pg. ``f30!}{}-27\index{40 Matthew:ch.  26, v.  26-27, pg. ``f30!}.} so that the
Popish\index{Popish rituals ``f10!} sacrifice of the\index{Lord's Supper:not a sacrifice:of the mass ``f10!}
mass\index{Mass:Roman Catholic ``f10!} (as they call it) is most abominably injurious to Christ's one, only sacrifice,
the\index{Jesus:sacrifice:once only ``f10!} alone propitiation\index{Jesus:sacrifice:propitiation ``f10!} for all the
sins of His elect\index{Election:elect of God ``f10!}.\footnote{\ Hebrews 7:23-24, 27\index{58 Hebrews:ch.   7, v. 
23-24, 27, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}1-12, 14, 18\index{58 Hebrews:ch. 
10, v.  11-12, 14, 18, pg. ``f30!}.}


\bigskip

29.3\ \ The Lord Jesus has, in this ordinance, appointed His ministers\index{Church officers:minsters ``f10!} to declare
His\index{Preachers:declare word of institution at Lord's Supper ``f10!} word of
institution\index{Sacraments:instituted by:Christ ``f10!} to the people; to pray, and bless the\index{Lord's
Supper:elements:Jesus blessed ``f10!} elements\index{Lord's Supper:elements:bread and wine ``f10!} of bread
and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}, and thereby to set them apart from a
common to a holy\index{Lord's Supper:elements:holy use ``f10!} use; and to take and break the\index{Lord's Supper:bread
broken ``f10!} bread, to take the cup\index{Lord's Supper:cup ``f10!}, and (they communicating\index{Lord's
Supper:communion ``f10!} also themselves) to give both to the communicants\index{Lord's Supper:both bread and wine to
communicants ``f10!};\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch. 
26, v.  26-28, pg. ``f30!} and Mark 14:22-24\index{41 Mark:ch.  14, v.  22-24, pg. ``f30!} and Luke 22:19\index{42
Luke:ch.  22, v.  19, pg. ``f30!}{}-20\index{42 Luke:ch.  22, v.  19-20, pg. ``f30!} with 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1
Corinthians:ch.  11, v.  23-26, pg. ``f30!}.} but to none who are not then present in the congregation\index{Lord's
Supper:only to those present in congregation ``f10!}.\footnote{\ Acts. 20:7; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}.}

29.4\ \ Private\index{Mass:private ``f10!} masses\index{Mass:Roman Catholic ``f10!}, or receiving this\index{Lord's
Supper:not private ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} by a priest or\index{Lord's Supper:not
consecrated by a priest ``f10!} any other alone;\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16,
pg. ``f30!}.} as likewise, the denial\index{Lord's Supper:both bread and wine to communicants ``f10!} of the
cup\index{Lord's Supper:cup ``f10!} to the people,\footnote{\ Mark 14:23\index{41 Mark:ch.  14, v.  23, pg. ``f30!}; 1
Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch.  11, v.  25, pg.
``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  25-29, pg. ``f30!}.} worshipping the elements\index{Lord's
Supper:elements:bread and wine ``f10!}, the lifting\index{Lord's Supper:elements:not to be lifted for adoration ``f10!}
them up or carrying them about for adoration\index{Idolatry:adoration of eucharistic elements ``f10!}, and the
reserving them for any pretended religious use\index{Lord's Supper:elements:not to be used incorrectly ``f10!}; are all
contrary to the nature of this sacrament, and to the institution\index{Sacraments:instituted by:Christ ``f10!} of
Christ.\footnote{\ Matthew 15:9\index{40 Matthew:ch.  15, v.   9, pg. ``f30!}.}


\bigskip

29.5\ \ The outward elements\index{Lord's Supper:elements:bread and wine ``f10!} in this
sacrament\index{Sacraments:holy:ordinances ``f10!}, duly set apart to the uses ordained by\index{Sacraments:ordained by
Jesus ``f10!} Christ, have such relation to Him\index{Lord's Supper:elements:relation to Christ crucified ``f10!}
crucified\index{Jesus:crucified ``f10!}, as that, truly, yet sacramentally\index{Lord's Supper:elements:union between
sign and signified ``f10!} only, they are sometimes called by the name of the things they represent, to\index{Lord's
Supper:elements:represent Christ ``f10!} wit, the body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of
Christ;\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v. 
26-28, pg. ``f30!}.} albeit in substance and nature they still remain truly\index{Lord's Supper:elements:remain truly
and only natural ``f10!} and only bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}, as
they were before.\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}6\index{46 1
Corinthians:ch.  11, v.  26, pg. ``f30!}{}-28\index{46 1 Corinthians:ch.  11, v.  26-28, pg. ``f30!}; Matthew
26:29\index{40 Matthew:ch.  26, v.  29, pg. ``f30!}.}


\bigskip

29.6\ \ That doctrine\index{Doctrine:false:transubstantiation ``f10!} which maintains\index{Lord's
Supper:elements:transubstantiation:repugnant ``f10!} a change of the substance of bread and\index{Lord's Supper:bread
``f10!} wine\index{Lord's Supper:wine ``f10!} into the substance of Christ's body and blood\index{Lord's Supper:body
and blood of Jesus ``f10!} (commonly called transubstantiation\index{Lord's
Supper:elements:transubstantiation:superstition ``f10!}) by consecration\index{Lord's Supper:elements:not changed by
consecration of priest ``f10!} of a priest, or by\index{Lord's Supper:not consecrated by a priest ``f10!} any other
way, is repugnant, not to Scripture alone, but even to common sense and reason; overthrows the nature of the
sacrament\index{Sacraments:holy:ordinances ``f10!}, and has been, and is the cause of manifold
superstitions\index{Superstitions:superstitions ``f10!}; yea, of gross idolatries\index{Idolatry:idolatry
``f10!}.\footnote{\ Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!} with 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-26\index{46 1
Corinthians:ch.  11, v.  24-26, pg. ``f30!}; Luke 24:6, 39\index{42 Luke:ch.  24, v.   6, 39, pg. ``f30!}.}


\bigskip

29.7\ \ Worthy\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!} receivers outwardly partaking\index{Lord's
Supper:elements:outward partaking ``f10!} of the visible elements\index{Lord's Supper:elements:bread and wine ``f10!}
in this sacrament\index{Sacraments:holy:ordinances ``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch. 
11, v.   2, pg. ``f30!}8\index{46 1 Corinthians:ch.  11, v.  28, pg. ``f30!}.} do then also, inwardly by faith, really
and indeed, yet not carnally\index{Lord's Supper:spiritual feeding:not carnal ``f10!} and corporally\index{Lord's
Supper:spiritual feeding:not corporal ``f10!}, but spiritually, receive and feed\index{Lord's Supper:spiritual
feeding:in faith ``f10!} upon Christ crucified\index{Jesus:crucified ``f10!}, and all benefits of\index{Lord's
Supper:benefits believers receive ``f10!} His death\index{Jesus:death:death ``f10!}: the body and blood\index{Lord's
Supper:body and blood of Jesus ``f10!} of Christ being then, not corporally or carnally, in, with, or under the bread
and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}; yet, as really, but spiritually, present
to the faith\index{Faith:profession of ``f10!} of believers in that ordinance, as the elements themselves are to their
outward\index{Lord's Supper:elements:outward senses ``f10!} senses.\footnote{\ 1 Corinthians 10:16\index{46 1
Corinthians:ch.  10, v.  16, pg. ``f30!}.}


\bigskip

29.8\ \ Although ignorant\index{ignorance ``f10!} and wicked men receive\index{Lord's Supper:wicked do not receive thing
signified ``f10!} the outward elements\index{Lord's Supper:elements:bread and wine ``f10!} in this
sacrament\index{Sacraments:holy:ordinances ``f10!}: yet they receive not the thing signified thereby, but by their
unworthy coming\index{Lord's Supper:unworthy participation ``f10!} thereunto are guilty\index{guilt ``f10!} of the body
and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of the Lord to their own
damnation\index{Punishment:everlasting:damnation ``f10!}. Wherefore, all ignorant and ungodly\index{Unbelievers:ungodly
``f10!} persons, as they are unfit\index{Lord's Supper:ungodly not to be admitted ``f10!} to enjoy communion with Him,
so are they unworthy of the Lord's table; and cannot, without great sin against Christ while they remain such, partake
of these holy\index{Lord's Supper:elements:holy mysteries ``f10!} mysteries,\index{Lord's Supper:mystery of
``f10!}\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch. 
11, v.  27, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  27-29, pg. ``f30!}; 2 Corinthians 6:14\index{47 2
Corinthians:ch.   6, v.  14, pg. ``f30!}{}-16\index{47 2 Corinthians:ch.   6, v.  14-16, pg. ``f30!}.} or be admitted
thereunto.\footnote{\ 1 Corinthians 5:6-7\index{46 1 Corinthians:ch.   5, v.   6-7, pg. ``f30!}, 13\index{46 1
Corinthians:ch.   5, v.   6-7, 13, pg. ``f30!}; 2 Thessalonians 3:6, 14-15\index{53 2 Thessalonians:ch.   3, v.   6,
14-15, pg. ``f30!}; Matthew 7:6\index{40 Matthew:ch.   7, v.   6, pg. ``f30!}.}

\section{Chapter 30 -- Of Church Censures}
\index{Church discipline:censures ``f10!}30.1\ \ The Lord Jesus, as King\index{Jesus:King ``f10!} and
Head\index{Church:Christ as head ``f10!} of His Church\index{Church:Christ's ``f10!}, has therein appointed a
government, in\index{Church government:government ``f10!} the hand of Church\index{Church:institution ``f10!}
officers\index{Church officers:officers ``f10!}, distinct from\index{Church government:distinct from civil ``f10!} the
civil magistrate.\footnote{\ Isaiah 9:6\index{23 Isaiah:ch.   9, v.   6, pg. ``f30!}{}-7\index{23 Isaiah:ch.   9, v.  
6-7, pg. ``f30!}; 1 Timothy 5:1\index{54 1 Timothy:ch.   5, v.   1, pg. ``f30!}7\index{54 1 Timothy:ch.   5, v.  17,
pg. ``f30!}; 1 Thessalonians 5:12\index{52 1 Thessalonians:ch.   5, v.  12, pg. ``f30!}; Acts 20:17, 28\index{44
Acts:ch.  20, v.  17, 28, pg. ``f30!}; Hebrews 13:7\index{58 Hebrews:ch.  13, v.   7, pg. ``f30!}, 17, 24\index{58
Hebrews:ch.  13, v.   7, 17, 24, pg. ``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!};
Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  18-20, pg. ``f30!}.}


\bigskip

30.2\ \ To these officers\index{Church officers:officers ``f10!} the keys\index{Kingdom:keys of ``f10!} of the kingdom
of heaven\index{Kingdom:of heaven ``f10!} are committed: by virtue whereof, they have power\index{Church:power
delegated by Christ ``f10!} respectively to retain, and remit\index{Church officers:power to retain or remit sins
``f10!} sins; to shut that kingdom against the impenitent\index{Punishment:of impenitent ``f10!}, both by the Word
and\index{Scripture:word of God:standard for:life ``f10!} censures; and\index{Church discipline:censures ``f10!} to
open it unto penitent\index{Repentance:penitence ``f10!} sinners, by the ministry of the Gospel\index{Gospel:gospel
``f10!}, and by absolution\index{Church discipline:absolution ``f10!} from censures, as occasion shall
require.\footnote{\ Matthew 16:19\index{40 Matthew:ch.  16, v.  19, pg. ``f30!}; Matthew 18:17\index{40 Matthew:ch. 
18, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.  18, v.  17-18, pg. ``f30!}; John 20:21-23\index{43 John:ch.  20, v. 
21-23, pg. ``f30!}; 2 Corinthians 2:6-8\index{47 2 Corinthians:ch.   2, v.   6-8, pg. ``f30!}.}


\bigskip

30.3\ \ Church censures are\index{Church discipline:censures ``f10!} necessary, for the reclaiming\index{Church
discipline:for reclaiming offending brothers ``f10!} and gaining of offending brethren, for\index{Church discipline:for
gaining offending brothers ``f10!} deterring\index{Church discipline:deterrence from sin ``f10!} of others from the
like offences, for purging\index{Church discipline:purging leaven ``f10!} out of that leaven\index{leaven, spiritual
``f10!} which might infect the whole lump, for vindicating the honour of\index{Church discipline:vindicate honour of
Christ ``f10!} Christ, and the holy\index{Preaching:holy profession of Gospel ``f10!} profession of the
Gospel\index{Gospel:gospel ``f10!}, and for preventing the wrath of God\index{Wrath:of God:for sin ``f10!}, which might
justly fall\index{Punishment:just:on Church for failure of discipline ``f10!} upon the Church,
if\index{Church:institution ``f10!} they should suffer His covenant and\index{Covenant:new:seals ``f10!} the
seals\index{Sacraments:covenant seals ``f10!} thereof to be profaned by\index{Covenant:not to be profaned ``f10!}
notorious\index{Church discipline:notorious offenders ``f10!} and obstinate\index{Church discipline:obstinate offenders
``f10!} offenders.\footnote{\ 1 Corinthians chap. 5; 1 Timothy 5:20\index{54 1 Timothy:ch.   5, v.  20, pg. ``f30!};
Matthew 7:6\index{40 Matthew:ch.   7, v.   6, pg. ``f30!}; 1 Timothy 1:20\index{54 1 Timothy:ch.   1, v.  20, pg.
``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v. 
27, pg. ``f30!} to the end, with Jude 23\index{65 Jude:v.  23, pg. ``f30!}.}


\bigskip

30.4\ \ For the better attaining of these ends, the officers\index{Church officers:officers ``f10!} of the Church
are\index{Church:institution ``f10!} to proceed by admonition\index{Church discipline:admonition ``f10!};
suspension\index{Church discipline:suspension from Lord's Supper ``f10!} from the
sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's Supper\index{Lord's Supper:sacrament ``f10!} for a
season; and\index{Lord's Supper:suspension for a season ``f10!} by excommunication\index{Church
discipline:excommunication ``f10!} from the Church; according to the nature of the crime\index{Church
discipline:consistent with:nature of crime ``f10!}, and demerit\index{Church discipline:consistent with:demerit of
person ``f10!} of the person.\footnote{\ 1 Thessalonians 5:12\index{52 1 Thessalonians:ch.   5, v.  12, pg. ``f30!}; 2
Thessalonians 3:6, 14-15\index{53 2 Thessalonians:ch.   3, v.   6, 14-15, pg. ``f30!}; 1 Corinthians 5:4-5\index{46 1
Corinthians:ch.   5, v.   4-5, pg. ``f30!}, 13\index{46 1 Corinthians:ch.   5, v.   4-5, 13, pg. ``f30!}; Matthew
18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}0\index{56
Titus:ch.   3, v.  10, pg. ``f30!}.}

\section{Chapter 31 -- Of Synods and Councils}
\index{Church government:synods ``f10!}\index{Church government:councils ``f10!}31.1\ \ For the better government,
and\index{Church government:government ``f10!} further edification of\index{Church councils:edification of Church
``f10!} the Church, there\index{Church:institution ``f10!} ought to be such assemblies\index{Church councils:assemblies
to be held ``f10!} as are commonly called synods\index{Church government:synods ``f10!} or councils\index{Church
government:councils ``f10!}.\footnote{\ Acts 15:2\index{44 Acts:ch.  15, v.   2, pg. ``f30!}, 4, 6\index{44 Acts:ch. 
15, v.   2, 4, 6, pg. ``f30!}.}


\bigskip

31.2\ \ As magistrates may lawfully call\index{Civil magistrate:to call synods to consult and advise ``f10!} a synod of
ministers\index{Church officers:minsters ``f10!}, and other fit persons, to consult and advise with, about matters of
religion\index{Civil magistrate:to call synods to consult about matters of religion ``f10!};\footnote{\ Isaiah
49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg.
``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}; 2 Chronicles 19:8-11\index{14 2 Chronicles:ch.  19, v.  
8-11, pg. ``f30!}; 2 Chronicles 29; 2 Chronicles 30; Matthew 2:4-5\index{40 Matthew:ch.   2, v.   4-5, pg. ``f30!};
Proverbs 11:1\index{20 Proverbs:ch.  11, v.   1, pg. ``f30!}4\index{20 Proverbs:ch.  11, v.  14, pg. ``f30!}.} so, if
magistrates be\index{Civil magistrate:legal authority even if:unbelievers ``f10!} open enemies\index{Enemies:of the
Church ``f10!} to the Church, the\index{Church:institution ``f10!} ministers of Christ of themselves, by virtue of
their office\index{Church officers:officers ``f10!}, or they, with other fit persons upon delegation\index{Church
government:delegate representatives to meet with government ``f10!} from their Churches, may meet together in such
assemblies\index{Church councils:assemblies to be held ``f10!}.\footnote{\ Acts 15:2\index{44 Acts:ch.  15, v.   2, pg.
``f30!}, 4, 22-23, 25\index{44 Acts:ch.  15, v.   2, 4, 22-23, 25, pg. ``f30!}.}\textsuperscript{,}
\footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription
}}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).}}

31.3\ \ It belongs to synods\index{Church government:synods ``f10!} and councils\index{Church government:councils
``f10!}, ministerially to determine controversies\index{Church councils:determine controversies of faith ``f10!} of
faith and cases of conscience\index{Conscience:cases of, in controversies ``f10!}; to set down rules and\index{Church
councils:set rules for public worship ``f10!} directions for the better ordering of the\index{Church
councils:directions for ordering public worship ``f10!} public worship\index{Worship:public ``f10!} of God, and
government of\index{Church government:government ``f10!} his Church\index{Church:Christ's ``f10!}; to receive
complaints\index{Church discipline:synods to receive complaints ``f10!} in cases of maladministration\index{Church
discipline:synods to address maladministration ``f10!}, and authoritatively to determine the same: which
decrees\index{Church councils:decrees ``f10!} and determinations\index{Church councils:Biblical determinations to be
received ``f10!}, if consonant to the Word of God\index{Scripture:word of God:word of God ``f10!}, are to be received
with reverence and\index{Church councils:submission to ``f10!} submission; not only for their agreement with the Word,
but\index{Scripture:word of God:standard:for Church ``f10!} also for the power\index{Church:power delegated by Christ
``f10!} whereby they are made, as being an ordinance of\index{Church:furnished with:ordinances of God ``f10!} God
appointed thereunto in His Word.\footnote{\ Acts 15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}, 19, 24,
27-31\index{44 Acts:ch.  15, v.  15, 19, 24, 27-31, pg. ``f30!}; Acts 16:4\index{44 Acts:ch.  16, v.   4, pg. ``f30!};
Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}{}-20\index{40 Matthew:ch.  18, v.  17-20, pg. ``f30!}.}


\bigskip

31.4\ \ All synods\index{Church government:synods ``f10!} or councils\index{Church government:councils ``f10!}, since
the Apostles\index{Apostles:apostles ``f10!}{}' times, whether general or particular, may err\index{Church councils:may
err ``f10!}; and many have erred\index{Church councils:many have erred ``f10!}. Therefore they are not to be made the
rule of faith\index{Faith:rule of ``f10!}, or practice; but to be used as a help in\index{Church councils:not the rule
of faith or practice, but help ``f10!} both.\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg.
``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!}; Acts 17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}; 1
Corinthians 2:5\index{46 1 Corinthians:ch.   2, v.   5, pg. ``f30!}; 2 Corinthians 1:24\index{47 2 Corinthians:ch.   1,
v.  24, pg. ``f30!}.}


\bigskip

31.5\ \ Synods\index{Church government:synods ``f10!} and councils\index{Church government:councils ``f10!} are to
handle, or conclude, nothing, but that which is ecclesiastical\index{Church councils:to address only ecclesiastical
matters ``f10!}: and are not to intermeddle with civil affairs which\index{Church councils:not to meddle with civil
affairs ``f10!} concern the commonwealth\index{commonwealth ``f10!}; unless by way of humble\index{Church councils:to
receive humble petitions ``f10!} petition, in cases extraordinary; or by way of advice, for satisfaction of conscience,
if\index{Conscience:satisfaction of ``f10!} they be thereunto required by the civil\index{Civil magistrate:to call
synods to consult and advise ``f10!} magistrate.\footnote{\ Luke 12:13-14\index{42 Luke:ch.  12, v.  13-14, pg.
``f30!}; John 18:36\index{43 John:ch.  18, v.  36, pg. ``f30!}.}

\paragraph[Chapter 32 {}-- Of the State of Man After Death, and of the Resurrection of the Dead]{Chapter 32 -- Of the
State of Man After Death, and of the Resurrection of the Dead}
\index{Death:following:state of man ``f10!}\index{Resurrection:general, of dead ``f10!}32.1\ \ The bodies of men, after
death\index{Death:bodies:return to dust ``f10!}, return to dust\index{Adam:formed of dust ``f10!} and see
corruption:\index{Death:corruption of body ``f10!}\footnote{\ Genesis 3:19\index{01 Genesis:ch.   3, v.  19, pg.
``f30!}; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}6\index{44 Acts:ch.  13, v.  36, pg. ``f30!}.} but their
souls (which neither die nor\index{Mankind:immortal soul ``f10!} sleep) having an immortal subsistence, immediately
return to God who gave them:\index{Believers:souls made perfect at death ``f10!}\footnote{\ Luke 23:43\index{42
Luke:ch.  23, v.  43, pg. ``f30!}; Ecclesiastes 12:7\index{21 Ecclesiastes:ch.  12, v.   7, pg. ``f30!}.} the souls of
the righteous, being then\index{Resurrection:believers:perfected ``f10!} made perfect in
holiness\index{Believers:holy:made perfect at death ``f10!}, are received into the highest heavens,
where\index{Heaven:God's abode ``f10!} they behold\index{Believers:behold face of God in glory ``f10!} the face
of\index{Resurrection:believers:behold face of God ``f10!} God, in light and glory\index{Resurrection:believers:into
glory:light ``f10!}, waiting for\index{Resurrection:believers:waiting expectantly ``f10!} the full
redemption\index{Redemption:full ``f10!} of their bodies.\index{Bodies in graves:believers ``f10!}\footnote{\ Hebrews
12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}3\index{58 Hebrews:ch.  12, v.  23, pg. ``f30!}; 2 Corinthians 5:1,
6, 8\index{47 2 Corinthians:ch.   5, v.   1, 6, 8, pg. ``f30!}; Philippians 1:23\index{50 Philippians:ch.   1, v.  23,
pg. ``f30!} with Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!} and Ephesians 4:10\index{49 Ephesians:ch.   4, v.
 10, pg. ``f30!}.} And the souls of the wicked are\index{Judgement day:wicked:cast into:hell ``f10!} cast
into\index{Punishment:hell:wicked cast into ``f10!} hell\index{Hell:hell ``f10!}, where they remain in torments
and\index{Punishment:everlasting:torments (body and soul) ``f10!} utter darkness\index{darkness, spiritual ``f10!},
reserved to the judgment of the\index{Judgement:final ``f10!} great day.\footnote{\ Luke 16:23-24\index{42 Luke:ch. 
16, v.  23-24, pg. ``f30!}; Acts 1:25\index{44 Acts:ch.   1, v.  25, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg.
``f30!}{}-7\index{65 Jude:v. 6-7, pg. ``f30!}; Peter 3:19.} Beside these two places, for
soul\index{Superiors:to:provide inferiors with things necessary for soul and body ``f10!}s separated\index{Death:soul
separated from body ``f10!} from their bodies, the\index{Bodies in graves:believers and unbelievers ``f10!} Scripture
acknowledges\index{Last day:heaven or hell, no other option ``f10!} none.


\bigskip

32.2\ \ At the last day\index{Last day:last day ``f10!}, such as are found alive\index{Last day:those alive on earth
``f10!} shall not die, but\index{Last day:not all shall die ``f10!} be changed:\footnote{\ 1 Thessalonians
4:17\index{52 1 Thessalonians:ch.   4, v.  17, pg. ``f30!}; 1 Corinthians 15:51-52\index{46 1 Corinthians:ch.  15, v. 
51-52, pg. ``f30!}.} and all the dead\index{Death:following:all men raised ``f10!} shall be raised up, with the
selfsame bodies, and none other, although with different qualities\index{Resurrection:bodies:different qualities
``f10!}, which shall be united again\index{Resurrection:bodies:united to souls ``f10!} to their souls for
ever.\footnote{\ Job 19:26-27\index{18 Job:ch.  19, v.  26-27, pg. ``f30!}; 1 Corinthians 15:4\index{46 1
Corinthians:ch.  15, v.   4, pg. ``f30!}2-44\index{46 1 Corinthians:ch.  15, v.  42-44, pg. ``f30!}.}


\bigskip

32.3\ \ The bodies of the\index{Resurrection:bodies:same ones as before death, for just and unjust ``f10!} unjust shall,
by the power\index{Resurrection:believers:by the power of Christ ``f10!} of Christ, be raised to
dishonour\index{Resurrection:wicked raised in dishonour ``f10!}: the bodies of the just, by His Spirit, unto\index{Holy
Spirit:raises bodies of just to honour ``f10!} honour; and\index{Resurrection:believers:raised in honour ``f10!} be
made conformable\index{Resurrection:bodies:conformable to Christ's ``f10!} to His own glorious
body\index{Jesus:resurrection:body:glorious ``f10!}.\footnote{\ Acts 24:15\index{44 Acts:ch.  24, v.  15, pg. ``f30!};
John 5:28-29\index{43 John:ch.   5, v.  28-29, pg. ``f30!}; 1 Corinthians 15:4\index{46 1 Corinthians:ch.  15, v.   4,
pg. ``f30!}3\index{46 1 Corinthians:ch.  15, v.  43, pg. ``f30!}; Philippians 3:21\index{50 Philippians:ch.   3, v. 
21, pg. ``f30!}.}

\section{Chapter 33 -- Of the Last Judgment}
\index{Judgement:final ``f10!}33.1\ \ God has appointed a day, wherein He will judge the\index{Jesus:judge:world ``f10!}
world in righteousness\index{Judgement day:in righteousness ``f10!}, by Jesus Christ,\footnote{\ Acts 17:31\index{44
Acts:ch.  17, v.  31, pg. ``f30!}.} to whom all power\index{Jesus:almighty power ``f10!} and judgment is
given\index{Jesus:judge:assigned by Father ``f10!} of the Father\index{Father:God the ``f10!}.\footnote{\ John 5:22,
27\index{43 John:ch.   5, v.  22, 27, pg. ``f30!}.} In which day, not only the apostate angels\index{Angels:angels
``f10!} shall be judged\index{Angels:apostate, to be judged ``f10!},\footnote{\ 1 Corinthians 6:3\index{46 1
Corinthians:ch.   6, v.   3, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg. ``f30!}; 2 Peter 2:4\index{61 2 Peter:ch.  
2, v.   4, pg. ``f30!}.} but likewise all persons that have lived upon earth shall appear before the
tribunal\index{Judgement day:tribunal of Christ ``f10!} of Christ, to give an account of their thoughts, words, and
deeds\index{Judgement day:account for thoughts, words, and deeds ``f10!}; and to receive according to what they have
done in the body, whether good or evil\index{Judgement day:give account of deeds, whether good or evil
``f10!}.\footnote{\ 2 Corinthians 5:10\index{47 2 Corinthians:ch.   5, v.  10, pg. ``f30!}; Ecclesiastes 12:1\index{21
Ecclesiastes:ch.  12, v.   1, pg. ``f30!}4\index{21 Ecclesiastes:ch.  12, v.  14, pg. ``f30!}; Romans 2:1\index{45
Romans:ch.   2, v.   1, pg. ``f30!}6\index{45 Romans:ch.   2, v.  16, pg. ``f30!}; Romans 14:10, 12\index{45 Romans:ch.
 14, v.  10, 12, pg. ``f30!}; Matthew 12:36\index{40 Matthew:ch.  12, v.  36, pg. ``f30!}{}-37\index{40 Matthew:ch. 
12, v.  36-37, pg. ``f30!}.}


\bigskip

33.2\ \ The end of God's appointing this day is for the manifestation of the glory of His\index{God's
attributes:glory:his own ``f10!} mercy, in\index{God's attributes:merciful ``f10!} the eternal
salvation\index{Salvation:belief in Jesus by faith ``f10!} of the elect\index{Election:elect of God ``f10!}; and of His
justice\index{God's attributes:administers justice ``f10!}, in the damnation\index{Punishment:everlasting:damnation
``f10!} of the reprobate\index{Punishment:of reprobate ``f10!} who are wicked and\index{Judgement day:wicked:damned
``f10!} disobedient\index{Judgement day:disobedient cast into hell ``f10!}. For then shall the righteous
go\index{Resurrection:believers:to everlasting life ``f10!} into everlasting life\index{everlasting life ``f10!}, and
receive that fulness of joy\index{Joy:everlasting ``f10!} and refreshing\index{Judgement day:believers:joy and
refreshing ``f10!}, which shall come from the presence of the Lord: but the wicked who\index{Judgement day:wicked:cast
into:eternal torments ``f10!} know not God, and obey not\index{Punishment:those who do not obey the Gospel ``f10!} the
Gospel\index{Gospel:gospel ``f10!} of Jesus Christ, shall be cast into\index{Punishment:hell:wicked cast into ``f10!}
eternal torments, and\index{Punishment:everlasting:torments (body and soul) ``f10!} be punished with everlasting
destruction from\index{Punishment:everlasting:destruction ``f10!} the presence of the Lord, and from the glory of His
power.\footnote{\ Matthew 25:31\index{40 Matthew:ch.  25, v.  31, pg. ``f30!} to the end, Romans 2:5\index{45
Romans:ch.   2, v.   5, pg. ``f30!}{}-6\index{45 Romans:ch.   2, v.   5-6, pg. ``f30!}; Romans 9:22-23\index{45
Romans:ch.   9, v.  22-23, pg. ``f30!}; Matthew 25:21\index{40 Matthew:ch.  25, v.  21, pg. ``f30!}; Acts 3:19\index{44
Acts:ch.   3, v.  19, pg. ``f30!}; 2 Thessalonians 1:7\index{53 2 Thessalonians:ch.   1, v.   7, pg.
``f30!}{}-10\index{53 2 Thessalonians:ch.   1, v.   7-10, pg. ``f30!}.}


\bigskip

33.3\ \ As Christ would have us to be certainly persuaded that\index{Judgement day:be persuaded that there is ``f10!}
there shall be a day of judgment\index{Judgement day:believers:perfected ``f10!}, both to deter\index{Judgement
day:knowledge as deterrence from sin ``f10!} all men from sin, and for the greater consolation of the godly in their
adversity\index{Judgement day:encourage godly in adversity ``f10!};\footnote{\ 2 Peter 3:11, 14\index{61 2 Peter:ch.  
3, v.  11, 14, pg. ``f30!}; 2 Corinthians 5:10\index{47 2 Corinthians:ch.   5, v.  10, pg. ``f30!}{}-11\index{47 2
Corinthians:ch.   5, v.  10-11, pg. ``f30!}; 2 Thessalonians 1:5-7\index{53 2 Thessalonians:ch.   1, v.   5-7, pg.
``f30!}; Luke 21:27-28\index{42 Luke:ch.  21, v.  27-28, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg.
``f30!}3\index{45 Romans:ch.   8, v.  23, pg. ``f30!}{}-25\index{45 Romans:ch.   8, v.  23-25, pg. ``f30!}.} so will He
have that day unknown\index{Judgement day:day unknown to men ``f10!} to men, that they may shake off all carnal
security\index{Idolatry:carnal:security ``f10!}, and be always watchful, because\index{Jesus:return:watchfulness
``f10!} they know not at what hour the\index{Jesus:return:no one knows hour ``f10!} Lord will come; and may be ever
prepared\index{Last day:be prepared to say, 'Come, Lord Jesus' ``f10!} to say, Come, Lord Jesus, come quickly,
Amen\index{Amen, testimony of our true desire for God's glory ``f10!}.\footnote{\ Matthew 24:36\index{40 Matthew:ch. 
24, v.  36, pg. ``f30!}, 42-44\index{40 Matthew:ch.  24, v.  36, 42-44, pg. ``f30!}; Mark 13:35-37\index{41 Mark:ch. 
13, v.  35-37, pg. ``f30!}; Luke 12:35-36\index{42 Luke:ch.  12, v.  35-36, pg. ``f30!}; Revelation 22:20\index{66
Revelation:ch.  22, v.  20, pg. ``f30!}.}


\subsubsection[The Westminster Larger Catechism of 1648]{{The Westminster Larger Catechism of
1648}}
Q. 1. \textit{What is the }chief\index{Mankind:chief end/purpose ``f10!}\textit{ and highest end of man?}

A. Man's chief\index{Mankind:chief end/purpose ``f10!} and highest end is to glorify\index{Worship:glorifying God
``f10!} God,\footnote{\ Romans 11:36\index{45 Romans:ch.  11, v.  36, pg. ``f30!}; 1 Corinthians 10:31\index{46 1
Corinthians:ch.  10, v.  31, pg. ``f30!}.} and fully to enjoy him forever.\footnote{\ Psalm 73:24-28\index{19
Psalms:ch.  73, v.  24-28, pg. ``f30!}.}


\bigskip

Q. 2 \textit{How does it appear that there is a God?}

A. The very light of nature in man, and the works of God\index{Providence:God's works ``f10!}, declare
plainly\index{Revelation:natural:declares there is a God ``f10!} that there is a God;\footnote{\ Romans
1:19-20\index{45 Romans:ch.   1, v.  19-20, pg. ``f30!}; Psalm 19:1-3\index{19 Psalms:ch.  19, v.   1-3, pg. ``f30!};
Acts 17:28\index{44 Acts:ch.  17, v.  28, pg. ``f30!}.} but his word and Spirit\index{Revelation:Word and Spirit
``f10!} only do sufficiently and\index{Revelation:natural:not sufficient for salvation ``f10!} effectually
reveal\index{God's law:commandments:first revelation ``f10!} him unto men for their salvation\index{Salvation:belief in
Jesus by faith ``f10!}.\footnote{\ 1 Corinthians 2:9-10\index{46 1 Corinthians:ch.   2, v.   9-10, pg. ``f30!}; 2
Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  15-17, pg. ``f30!};
Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}1\index{23 Isaiah:ch.  59, v.  21, pg. ``f30!}.}


\bigskip

Q. 3 \textit{What is the word of God?}

A. The holy\index{Scripture:holy ``f10!} scriptures of\index{Scripture:word of God:revelation ``f10!} the Old and New
Testament\index{Scripture:New Testament ``f10!}s are the word of God,\footnote{\ 2 Timothy 3:16\index{55 2 Timothy:ch. 
 3, v.  16, pg. ``f30!}; 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg. ``f30!}{}-21\index{61 2 Peter:ch.   1, v. 
19-21, pg. ``f30!}.} the only rule of faith\index{Faith:rule of ``f10!} and obedience\index{God's
law:commandments:obedience:required ``f10!}.\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg.
``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!}; Revelation 22:18-19\index{66 Revelation:ch.  22, v.  18-19,
pg. ``f30!}; Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg. ``f30!}; Luke 16:29, 31\index{42 Luke:ch.  16, v.  29,
31, pg. ``f30!}; Galatians 1:8-9\index{48 Galatians:ch.   1, v.   8-9, pg. ``f30!}; 2 Timothy 3:15\index{55 2
Timothy:ch.   3, v.  15, pg. ``f30!}{}-16\index{55 2 Timothy:ch.   3, v.  15-16, pg. ``f30!}.}


\bigskip

Q. 4 \textit{How does it appear that the scriptures are}\index{Scripture:word of God:revelation ``f10!}\textit{ the word
of God?}

A. The scriptures manifest the\index{Providence:manifests goodness of God ``f10!}mselves to\index{Scripture:manifested
as word of God ``f10!} be the word of God, by their majesty\footnote{\ Hosea 8:12\index{28 Hosea:ch.   8, v.  12, pg.
``f30!}; 1 Corinthians 2:6-7, 13\index{46 1 Corinthians:ch.   2, v.   6-7, 13, pg. ``f30!}; Psalm 119:18\index{19
Psalms:ch. 119, v.  18, pg. ``f30!}, 129\index{19 Psalms:ch. 119, v.  18, 129, pg. ``f30!}.} and
purity\index{Scripture:purity of ``f10!};\footnote{\ Psalm 12:6\index{19 Psalms:ch.  12, v.   6, pg. ``f30!}; Psalm
119:140\index{19 Psalms:ch. 119, v. 140, pg. ``f30!}.} by the consent of all\index{Scripture:consent of all parts
``f10!} the parts,\footnote{\ Acts 10:43\index{44 Acts:ch.  10, v.  43, pg. ``f30!}; Acts 26:22\index{44 Acts:ch.  26,
v.  22, pg. ``f30!}.} and the scope of the whole, which is to give all glory\index{Scripture:gives glory to God ``f10!}
to God;\footnote{\ Romans 3:19, 27\index{45 Romans:ch.   3, v.  19, 27, pg. ``f30!}.} by their light and
power\index{Scripture:light and power to convict sinners ``f10!} to convince and\index{Scripture:power to convince and
convert ``f10!} convert sinners\index{Conversion:conversion ``f10!}, to comfort\index{Scripture:comforts and builds up
believers ``f10!} and build up believers\index{Holy Spirit:builds up the elect ``f10!} unto
salvation\index{Salvation:belief in Jesus by faith ``f10!}:\footnote{\ Acts 23:28\index{44 Acts:ch.  23, v.  28, pg.
``f30!}; Hebrews 4:12\index{58 Hebrews:ch.   4, v.  12, pg. ``f30!}; James 1:18\index{59 James:ch.   1, v.  18, pg.
``f30!}; Psalm 19:7\index{19 Psalms:ch.  19, v.   7, pg. ``f30!}{}-9\index{19 Psalms:ch.  19, v.   7-9, pg. ``f30!};
Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}; Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}.} but
the Spirit of God\index{Holy Spirit:God's Spirit ``f10!} bearing witness by\index{Scripture:witness to heart of man
``f10!} and with the scriptures in the heart of man, is alone able fully to persuade it\index{Scripture:Holy Spirit
persuades us that it is word of God ``f10!} that they are the very word of God.\footnote{\ John 16:13-14\index{43
John:ch.  16, v.  13-14, pg. ``f30!}; John 20:31\index{43 John:ch.  20, v.  31, pg. ``f30!}; 1 John 2:2\index{62 1
John:ch.   2, v.   2, pg. ``f30!}0, 27\index{62 1 John:ch.   2, v.  20, 27, pg. ``f30!}.}


\bigskip

Q. 5 \textit{What do the scriptures principally}\index{Scripture:principal teaching ``f10!}\textit{ teach?}

A. The scriptures principally\index{Scripture:principal teaching ``f10!} teach, what man is to believe concerning God,
and what duty God\index{God's law:commandments:duties:required by God ``f10!} requires of man.\footnote{\ 2 Timothy
1:13\index{55 2 Timothy:ch.   1, v.  13, pg. ``f30!}.}


\bigskip

Q. 6 \textit{What do the scriptures make}\index{Scripture:makes God known ``f10!}\textit{ known of God?}

A. The scriptures make\index{Scripture:makes God known ``f10!} known what God is,\footnote{\ Hebrews 11:6\index{58
Hebrews:ch.  11, v.   6, pg. ``f30!}.} the persons in the Godhead\index{God's attributes:Godhead ``f10!},\footnote{\ 1
John 5:7\index{62 1 John:ch.   5, v.   7, pg. ``f30!}.} his decrees\index{Predestination:eternal decrees
``f10!},\footnote{\ Acts 15:14-15, 18\index{44 Acts:ch.  15, v.  14-15, 18, pg. ``f30!}.} and the execution of his
decrees.\footnote{\ Acts 4:27-28\index{44 Acts:ch.   4, v.  27-28, pg. ``f30!}.}

Q. 7 \textit{What is God?}

A. God is a\index{God's attributes:listed (definition of God) ``f10!} Spirit,\footnote{\ John 4:24\index{43 John:ch.  
4, v.  24, pg. ``f30!}.} in and of himself infinite in\index{God's attributes:infinite:in being ``f10!}
being,\footnote{\ Exodus 3:14\index{02 Exodus:ch.   3, v.  14, pg. ``f30!}; Job 11:7-9\index{18 Job:ch.  11, v.   7-9,
pg. ``f30!}.} glory,\footnote{\ Acts 7:2\index{44 Acts:ch.   7, v.   2, pg. ``f30!}.} blessedness,\index{God's
attributes:perfect blessedness ``f10!}\footnote{\ 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg.
``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}.} and perfection;\footnote{\ Matthew 5:48\index{40 Matthew:ch.
  5, v.  48, pg. ``f30!}.} all-sufficient\index{God's attributes:all-sufficient ``f10!},\footnote{\ Genesis
17:1\index{01 Genesis:ch.  17, v.   1, pg. ``f30!}.} eternal,\index{God's attributes:eternal ``f10!}\footnote{\ Psalm
90:2\index{19 Psalms:ch.  90, v.   2, pg. ``f30!}.} unchangeable\index{God's attributes:unchangeable
``f10!},\footnote{\ Malachi 3:6\index{39 Malachi:ch.   3, v.   6, pg. ``f30!}; James 1:17\index{59 James:ch.   1, v. 
17, pg. ``f30!}.} incomprehensible\index{God's attributes:incomprehensible ``f10!},\footnote{\ 1 Kings 8:27\index{11 1
Kings:ch.   8, v.  27, pg. ``f30!}.} everywhere present\index{God's attributes:everywhere present
``f10!},\footnote{\ Psalm 139:1-13\index{19 Psalms:ch. 139, v.   1-13, pg. ``f30!}.} almighty\index{God's
attributes:almighty power ``f10!},\footnote{\ Revelation 4:8\index{66 Revelation:ch.   4, v.   8, pg. ``f30!}.} knowing
all\index{God's attributes:omniscience ``f10!} things,\footnote{\ Hebrews 4:13\index{58 Hebrews:ch.   4, v.  13, pg.
``f30!}; Psalm 147:5\index{19 Psalms:ch. 147, v.   5, pg. ``f30!}.} most wise\index{God's attributes:wisdom
``f10!},\footnote{\index{45 Romans:ch.  16, v.   2, pg.
``f30!} most
holy\index{God's attributes:holy:most ``f10!},\footnote{\index{23 Isaiah:ch.   6,
v.   3, pg. ``f30!}\index{66 Revelation:ch.  15, v.   4, pg.
``f30!}\index{05
Deuteronomy:ch.  32, v.   4, pg. ``f30!} most merciful\index{God's attributes:merciful
``f10!} and gracious\index{Grace:God's free ``f10!}, long-suffering\index{God's attributes:long-suffering ``f10!}, and
abundant in goodness\index{God's attributes:goodness ``f10!} and truth\index{God's attributes:truth
``f10!}.\footnote{\index{02 Exodus:ch.  34, v.   6, pg.
``f30!}


\bigskip

Q. 8 \textit{Are there more Gods than one?}

A. There is but one only, the living and true God.\footnote{\ Deuteronomy 6:4\index{05 Deuteronomy:ch.   6, v.   4, pg.
``f30!}; 1 Corinthians 8:4, 6\index{46 1 Corinthians:ch.   8, v.   4, 6, pg. ``f30!}; Jeremiah 10:10\index{24
Jeremiah:ch.  10, v.  10, pg. ``f30!}.}


\bigskip

Q. 9 \textit{How many persons}\index{God's attributes:trinity:three persons ``f10!}\textit{ are there in the
Godhead}\index{God's attributes:Godhead ``f10!}\textit{?}

A. There be three persons\index{God's attributes:trinity:three persons ``f10!} in the Godhead\index{God's
attributes:Godhead ``f10!}, the Father\index{Father:God the ``f10!}, the Son\index{Jesus:Son of God ``f10!}, and the
Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}; and these three are one true, eternal God\index{God's
attributes:eternal ``f10!}, the same in substance\index{God's attributes:trinity:one substance ``f10!}, equal in power
and glory\index{God's attributes:trinity:persons equal ``f10!}; although distinguished\index{God's
attributes:trinity:distinguished by personal properties ``f10!} by their personal properties.\footnote{\ 1 John
5:7\index{62 1 John:ch.   5, v.   7, pg. ``f30!}; Matthew 3:16-17\index{40 Matthew:ch.   3, v.  16-17, pg. ``f30!};
Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 2 Corinthians 13:1\index{47 2 Corinthians:ch.  13, v.   1,
pg. ``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}; John 10:30\index{43 John:ch.  10, v.  30, pg.
``f30!}.} 


\bigskip

Q. 10 \textit{What are the personal properties of}\index{God's attributes:trinity:distinguished by personal properties
``f10!}\textit{ the three persons}\index{God's attributes:trinity:three persons ``f10!}\textit{ in the
Godhead}\index{God's attributes:Godhead ``f10!}\textit{?}

A. It is proper to the Father\index{Father:God the ``f10!} to beget\index{Jesus:begotten by the Father ``f10!} the
Son\index{Jesus:Son of God ``f10!},\footnote{\ Hebrews 1:5-6, 8\index{58 Hebrews:ch.   1, v.   5-6, 8, pg. ``f30!}.}
and to the Son to be begotten of the Father,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43
John:ch.   1, v.  14, pg. ``f30!}, 18\index{43 John:ch.   1, v.  14, 18, pg. ``f30!}.} and to the Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!} to proceed from the\index{Holy Spirit:procession from Father and Son ``f10!} Father and the
Son from all eternity\index{God's attributes:eternal ``f10!}.\footnote{\ John 15:26\index{43 John:ch.  15, v.  26, pg.
``f30!}; Galatians 4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.} 


\bigskip


\bigskip


Q. 11 \textit{How does it appear that the Son}\index{Jesus:Son of God ``f10!}\textit{ and the Holy Ghost}\index{Holy
Spirit:Holy Ghost ``f10!}\textit{ are God equal with the}\index{God's attributes:trinity:persons equal ``f10!}\textit{
Father}\index{Father:God the ``f10!}\textit{?}

A. The scriptures manifest that\index{God's attributes:trinity:persons equal ``f10!} the Son\index{Jesus:Son of God
``f10!} and the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!} are God equal with the Father\index{Father:God the
``f10!}, ascribing unto\index{Worship:praising God's names ``f10!} them such names,\footnote{\ Isaiah 6:3\index{23
Isaiah:ch.   6, v.   3, pg. ``f30!}, 5, 8\index{23 Isaiah:ch.   6, v.   3, 5, 8, pg. ``f30!}; John 12:41\index{43
John:ch.  12, v.  41, pg. ``f30!}; Acts 5:3\index{44 Acts:ch.   5, v.   3, pg. ``f30!}{}-4\index{44 Acts:ch.   5, v.  
3-4, pg. ``f30!}; Acts 28:25\index{44 Acts:ch.  28, v.  25, pg. ``f30!}; 1 John 5:20\index{62 1 John:ch.   5, v.  20,
pg. ``f30!}.} attributes\index{God's attributes:to be:revered ``f10!},\footnote{\ John 1:1\index{43 John:ch.   1, v.  
1, pg. ``f30!}; John 2:24-25\index{43 John:ch.   2, v.  24-25, pg. ``f30!}; Isaiah 9:6\index{23 Isaiah:ch.   9, v.   6,
pg. ``f30!}; 1 Corinthians 2:10-11\index{46 1 Corinthians:ch.   2, v.  10-11, pg. ``f30!}.}
works,\footnote{\ Colossians 1:16\index{51 Colossians:ch.   1, v.  16, pg. ``f30!}; Genesis 1:2\index{01 Genesis:ch.  
1, v.   2, pg. ``f30!}.} and worship,\index{Worship:acceptable manner ``f10!}\footnote{\ Matthew 28:19\index{40
Matthew:ch.  28, v.  19, pg. ``f30!}; 2 Corinthians 13:1\index{47 2 Corinthians:ch.  13, v.   1, pg. ``f30!}4\index{47
2 Corinthians:ch.  13, v.  14, pg. ``f30!}.} as are proper to God only.


\bigskip

Q. 12 \textit{What are the decrees}\index{Predestination:eternal decrees ``f10!}\textit{ of God?}

A. God's decrees are\index{Predestination:eternal decrees ``f10!} the wise\index{God's attributes:wisdom ``f10!}, free,
and holy acts\index{God's attributes:holy:acts ``f10!} of the counsel of his\index{Causation:God's own will ``f10!}
will,\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg.
``f30!}; Romans 9:14-15, 18\index{45 Romans:ch.   9, v.  14-15, 18, pg. ``f30!}; Romans 11:33\index{45 Romans:ch.  11,
v.  33, pg. ``f30!}.} whereby, from all eternity\index{God's attributes:eternal ``f10!}, he has, for his own
glory\index{God's attributes:glory:his own ``f10!}, unchangeably foreordained\index{Predestination:fore-ordination
``f10!}: Whatsoever comes to pass in time,\footnote{\ Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!},
11\index{49 Ephesians:ch.   1, v.   4, 11, pg. ``f30!}; Romans 9:22-23\index{45 Romans:ch.   9, v.  22-23, pg. ``f30!};
Psalm 33:11\index{19 Psalms:ch.  33, v.  11, pg. ``f30!}.} especially concerning angels\index{Angels:angels ``f10!} and
men.


\bigskip

Q. 13 \textit{What has God especially decreed}\index{Predestination:eternal decrees ``f10!}\textit{ concerning
angels}\index{Angels:angels ``f10!}\textit{ and men?}

A. God, by an eternal and immutable\index{God's attributes:immutable ``f10!} decree\index{Predestination:eternal decrees
``f10!}, out of his mere love, for\index{God's attributes:love ``f10!} the praise of his glorious
grace\index{Grace:glorious ``f10!}, to be manifested\index{Grace:manifested ``f10!} in due time, has
elected\index{Election:some angels ``f10!} some angels\index{Angels:angels ``f10!} to glory\index{Angels:elected to
glory ``f10!};\footnote{\ 1 Timothy 5:21\index{54 1 Timothy:ch.   5, v.  21, pg. ``f30!}.} and in Christ has
chosen\index{Election:chosen by God ``f10!} some men to eternal life\index{eternal life ``f10!}, and the means
thereof:\footnote{\ Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}{}-6\index{49 Ephesians:ch.   1, v.  
4-6, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}{}-14\index{53 2
Thessalonians:ch.   2, v.  13-14, pg. ``f30!}.} and also, according to his sovereign\index{God's attributes:sovereign
lordship ``f10!} power, and the unsearchable\index{Providence:unsearchable wise counsel ``f10!} counsel of
his\index{Causation:God's own will ``f10!} own will (whereby he extends or withholds\index{Providence:God withholds
mercy, grace, favour as he wills ``f10!} favor as\index{Predestination:God's favour as he pleases ``f10!} he
pleases\index{Providence:at God's pleasure ``f10!}), has passed by and
foreordained\index{Predestination:fore-ordination ``f10!} the rest to dishonor and\index{Resurrection:wicked raised in
dishonour ``f10!} wrath, to be\index{Wrath:of God:for sin ``f10!} for their sin
inflicted\index{Predestination:mankind:unto reprobation ``f10!}, to the praise of the\index{Worship:praising God
``f10!} glory of his justice\index{God's attributes:administers justice ``f10!}.\footnote{\ Romans 9:17\index{45
Romans:ch.   9, v.  17, pg. ``f30!}{}-18, 21-22\index{45 Romans:ch.   9, v.  17-18, 21-22, pg. ``f30!}; Matthew
11:25-26\index{40 Matthew:ch.  11, v.  25-26, pg. ``f30!}; 2 Timothy 2:2\index{55 2 Timothy:ch.   2, v.   2, pg.
``f30!}0\index{55 2 Timothy:ch.   2, v.  20, pg. ``f30!}; Jude 1:4\index{65 Jude:v.   1, v.   4, pg. ``f30!}; 1 Peter
2:8\index{60 1 Peter:ch.   2, v.   8, pg. ``f30!}.} 


\bigskip

Q. 14 \textit{How does God execute his decrees}\index{Predestination:eternal decrees ``f10!}\textit{?}

A. God executes his decrees\index{Predestination:eternal decrees ``f10!} in the works of creation\index{God's
attributes:creator:works of ``f10!} and providence\index{Providence:providence ``f10!}, according to his
infallible\index{God's attributes:infallible fore-knowledge ``f10!} foreknowledge\index{God's attributes:foreknowledge
``f10!}, and the free and immutable\index{God's attributes:immutable ``f10!} counsel of his\index{Causation:God's own
will ``f10!} own will.\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.
  1, v.  11, pg. ``f30!}.} 


\bigskip

Q. 15 \textit{What is the work of creation}\index{God's attributes:creator:works of ``f10!}\textit{?}

A. The work of creation\index{God's attributes:creator:works of ``f10!} is that wherein God did in the beginning, by the
word of his power\index{Creation:by word of God's power ``f10!}, make of nothing the world, and all things therein, for
himself, within the space of six\index{Creation:in six days ``f10!} days, and all very good.\footnote{\ Genesis 1;
Hebrews 11:3\index{58 Hebrews:ch.  11, v.   3, pg. ``f30!}; Proverbs 16:4\index{20 Proverbs:ch.  16, v.   4, pg.
``f30!}.} 


\bigskip


\bigskip


Q. 16 \textit{How did God create angels}\index{Angels:angels ``f10!}\textit{?}

A. God created all the angels\index{Angels:angels ``f10!}\footnote{\ Colossians 1:16\index{51 Colossians:ch.   1, v. 
16, pg. ``f30!}.} spirits,\footnote{\ Psalm 104:4\index{19 Psalms:ch. 104, v.   4, pg. ``f30!}.}
immortal,\footnote{\ Matthew 22:30\index{40 Matthew:ch.  22, v.  30, pg. ``f30!}.} holy,\footnote{\ Matthew
25:31\index{40 Matthew:ch.  25, v.  31, pg. ``f30!}.} excelling in knowledge\index{Angels:attributes
``f10!},\footnote{\ 2 Samuel 14:17\index{10 2 Samuel:ch.  14, v.  17, pg. ``f30!}; Matthew 24:36\index{40 Matthew:ch. 
24, v.  36, pg. ``f30!}.} mighty in power,\footnote{\ 2 Thessalonians 1:7\index{53 2 Thessalonians:ch.   1, v.   7, pg.
``f30!}.} to execute his\index{Angels:created to execute God's commands ``f10!} commandments, and to praise
his\index{Worship:praising God ``f10!} name,\footnote{\ Psalm 103:20-21\index{19 Psalms:ch. 103, v.  20-21, pg.
``f30!}.} yet subject to change.\footnote{\ 2 Peter 2:4\index{61 2 Peter:ch.   2, v.   4, pg. ``f30!}.}


\bigskip

Q. 17 \textit{How did God create man}\index{Mankind:created:by God ``f10!}\textit{?}

A. After God\index{Mankind:created:attributes ``f10!} had made all other creatures\index{creatures ``f10!}, he created
man\index{Mankind:created:by God ``f10!} male\index{Mankind:male and female ``f10!} and
female\index{Mankind:created:male and female ``f10!};\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg.
``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}.} formed the body of the man of the dust\index{Adam:formed of
dust ``f10!} of the ground,\footnote{\ Genesis 2:7\index{01 Genesis:ch.   2, v.   7, pg. ``f30!}.} and the woman of the
rib\index{Eve, from Adam's rib ``f10!} of the man,\footnote{\ Genesis 2:22\index{01 Genesis:ch.   2, v.  22, pg.
``f30!}.} endued the\index{Causation:will of creatures:natural liberty ``f10!}m with living,
reasonable\index{Mankind:created:reasonable soul ``f10!}, and immortal\index{Mankind:immortal soul ``f10!}
souls;\footnote{\ Genesis 2:7\index{01 Genesis:ch.   2, v.   7, pg. ``f30!}; Job 35:11\index{18 Job:ch.  35, v.  11,
pg. ``f30!}; Ecclesiastes 12:7\index{21 Ecclesiastes:ch.  12, v.   7, pg. ``f30!}; Matthew 10:28\index{40 Matthew:ch. 
10, v.  28, pg. ``f30!}; Luke 23:43\index{42 Luke:ch.  23, v.  43, pg. ``f30!}.} made them after his own
image\index{Mankind:image of God ``f10!},\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg.
``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}.} in knowledge,\footnote{\ Colossians 3:10\index{51
Colossians:ch.   3, v.  10, pg. ``f30!}.} righteousness, and holiness;\index{Mankind:created:holy
``f10!}\footnote{\ Ephesians 4:24\index{49 Ephesians:ch.   4, v.  24, pg. ``f30!}.} having the law of God written in
their hearts\index{God's law:written on man's heart ``f10!},\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg.
``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}.} and power to fulfil it\index{Covenant:works:conditional
on perfect obedience ``f10!},\footnote{\ Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21
Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.} and dominion over the\index{Mankind:dominion over creation ``f10!}
creatures;\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}8\index{01 Genesis:ch.   1, v.  28, pg.
``f30!}.} yet subject to fall.\footnote{\ Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}; Ecclesiastes
7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.}


\bigskip

Q. 18 \textit{What are God's works}\index{Providence:God's works ``f10!}\textit{ of
providence}\index{Providence:providence ``f10!}\textit{?}

A. God's works of providence\index{Providence:providence ``f10!} are his most holy\index{God's attributes:holy:most
``f10!},\footnote{\ Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg. ``f30!}.} wise,\footnote{\ Psalm 104:24\index{19
Psalms:ch. 104, v.  24, pg. ``f30!}; Isaiah 28:29\index{23 Isaiah:ch.  28, v.  29, pg. ``f30!}.} and powerful
preserving\index{Providence:powerful preserving ``f10!}\footnote{\ Hebrews 1:8\index{58 Hebrews:ch.   1, v.   8, pg.
``f30!}.} and governing\index{Providence:governing all ``f10!}\footnote{\ Psalm 103:1\index{19 Psalms:ch. 103, v.   1,
pg. ``f30!}9\index{19 Psalms:ch. 103, v.  19, pg. ``f30!}.} all his creatures\index{creatures ``f10!}; ordering
them\index{Providence:ordering:all things for his glory and believers' good ``f10!}, and all their
actions,\footnote{\ Matthew 10:29-31\index{40 Matthew:ch.  10, v.  29-31, pg. ``f30!}; Genesis 45:7\index{01
Genesis:ch.  45, v.   7, pg. ``f30!}.} to his own glory\index{God's attributes:glory:his own ``f10!}.\footnote{\ Romans
11:36\index{45 Romans:ch.  11, v.  36, pg. ``f30!}; Isaiah 43:14\index{23 Isaiah:ch.  43, v.  14, pg. ``f30!}.}


\bigskip

Q. 19 \textit{What is God's providence}\index{Providence:providence ``f10!}\textit{ towards the
angels}\index{Angels:angels ``f10!}\textit{?}

A. God by his providence\index{Providence:providence ``f10!} permitted some\index{Angels:some were permitted to fall
into sin ``f10!} of the angels\index{Angels:angels ``f10!}, wilfully and\index{Angels:some wilfully and irrecoverably
fell into damnation ``f10!} irrecoverably, to fall into sin and damnation\index{Punishment:everlasting:damnation
``f10!},\footnote{\ Jude 1:6\index{65 Jude:v.   1, v.   6, pg. ``f30!}; 2 Peter 2:4\index{61 2 Peter:ch.   2, v.   4,
pg. ``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}6\index{58 Hebrews:ch.   2, v.  16, pg. ``f30!};
John 8:44\index{43 John:ch.   8, v.  44, pg. ``f30!}.} limiting\index{Angels:limited by God's providence ``f10!} and
ordering that\index{Providence:ordering:all aspects of fallen angels' existence ``f10!}, and all their sins, to his own
glory\index{God's attributes:glory:his own ``f10!};\footnote{\ Job 1:12\index{18 Job:ch.   1, v.  12, pg. ``f30!};
Matthew 8:31\index{40 Matthew:ch.   8, v.  31, pg. ``f30!}.} and established the\index{Angels:established in holiness
and happiness ``f10!} rest in holiness and\index{Believers:holy:being perfected ``f10!} happiness\index{happiness
``f10!};\footnote{\ 1 Timothy 5:21\index{54 1 Timothy:ch.   5, v.  21, pg. ``f30!}; Mark 8:38\index{41 Mark:ch.   8, v.
 38, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}2\index{58 Hebrews:ch.  12, v.  22, pg.
``f30!}.} employing them all,\footnote{\ Psalm 104:4\index{19 Psalms:ch. 104, v.   4, pg. ``f30!}.} at his
pleasure\index{Providence:at God's pleasure ``f10!}, in the administrations of his\index{Angels:administration of God's
providences ``f10!} power, mercy, and\index{God's attributes:merciful ``f10!} justice.\footnote{\ 2 Kings
19:35\index{12 2 Kings:ch.  19, v.  35, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg.
``f30!}4\index{58 Hebrews:ch.   1, v.  14, pg. ``f30!}.}


\bigskip

Q. 20 \textit{What was the providence}\index{Providence:providence ``f10!}\textit{ of God toward man in the estate
in}\index{Mankind:created:state of innocence ``f10!}\textit{ which he was created}\index{Mankind:created:state
``f10!}\textit{?}

A. The providence\index{Providence:providence ``f10!} of God toward man in the estate in\index{Mankind:created:state of
innocence ``f10!} which he was created\index{Mankind:created:state ``f10!}, was the placing\index{Adam:placed in
paradise ``f10!} him in paradise\index{paradise ``f10!}, appointing him to dress\index{Adam:assigned work of dressing
garden ``f10!} it, giving him liberty to eat\index{Liberty:original, abused ``f10!} of the fruit of the
earth;\footnote{\ Genesis 2:8, 15-16\index{01 Genesis:ch.   2, v.   8, 15-16, pg. ``f30!}.}
putting\index{Mankind:dominion over creation ``f10!} the creatures\index{creatures ``f10!} under his
dominion,\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}8\index{01 Genesis:ch.   1, v.  28, pg.
``f30!}.} and ordaining\index{Marriage:ordained by God ``f10!} marriage for his help\index{Marriage:mutual help
``f10!};\footnote{\ Genesis 2:18\index{01 Genesis:ch.   2, v.  18, pg. ``f30!}.} affording him communion with
him\index{Communion:with Christ ``f10!}self\index{Communion:with God ``f10!};\footnote{\ Genesis 1:2\index{01
Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}{}-29\index{01 Genesis:ch.   1, v. 
26-29, pg. ``f30!}; Genesis 3:8\index{01 Genesis:ch.   3, v.   8, pg. ``f30!}.} instituting\index{Sabbath:instituted by
God ``f10!} the sabbath;\footnote{\ Genesis 2:3\index{01 Genesis:ch.   2, v.   3, pg. ``f30!}.} entering into a
covenant of life\index{Covenant:life ``f10!} with him, upon condition\index{Covenant:works:conditional on perfect
obedience ``f10!} of personal, perfect, and perpetual obedience\index{Covenant:works:perpetual obedience required
``f10!},\footnote{\ Galatians 3:12\index{48 Galatians:ch.   3, v.  12, pg. ``f30!}; Romans 10:5\index{45 Romans:ch. 
10, v.   5, pg. ``f30!}.} of which the tree of life\index{Tree:of life, pledge to first covenant ``f10!} was a
pledge\index{Covenant:first:pledge in tree of life ``f10!};\footnote{\ Genesis 2:9\index{01 Genesis:ch.   2, v.   9,
pg. ``f30!}.} and forbidding\index{God's law:commandments:given to Adam ``f10!} to eat of the tree of the\index{Tree:of
the knowledge of good and evil ``f10!} knowledge of good and evil, upon the pain of death\index{Death:punishment for
sin ``f10!}.\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}.}


\bigskip

Q. 21 \textit{Did man continue in that estate}\index{Mankind:created:state of innocence ``f10!}\textit{ wherein God at
first created him}\index{Mankind:created:state ``f10!}\textit{?}

A. Our first parents\index{Parents:first, sinned against God ``f10!} being left to the freedom of\index{Causation:will
of creatures:free ``f10!} their own will, through the temptation\index{Temptation:temptation ``f10!} of Satan,
transgressed\index{Satan:tempter ``f10!} the commandment\index{God's law:commandments:given to Adam ``f10!} of God in
eating the forbidden\index{Sin:original:first transgression (of Adam) ``f10!} fruit; and thereby
fell\index{Sin:original:all mankind descending from Adam ``f10!} from the estate of innocency
wherein\index{Mankind:created:state of innocence ``f10!} they were created\index{Mankind:created:state
``f10!}.\footnote{\ Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}{}-8\index{01 Genesis:ch.   3, v.   6-8,
pg. ``f30!}, 13\index{01 Genesis:ch.   3, v.   6-8, 13, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.
  2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}; 2 Corinthians 11:3\index{47 2 Corinthians:ch. 
11, v.   3, pg. ``f30!}.}


\bigskip

Q. 22 \textit{Did all mankind fall}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{ in that first
transgression}\index{Sin:original:first transgression (of Adam) ``f10!}\textit{?}

A. The covenant being\index{Covenant:first:made with Adam ``f10!} made with Adam\index{Adam:covenant made with him
``f10!} as a public person, not for himself only, but for his posterity\index{Sin:original:all mankind descending from
Adam ``f10!}, all mankind descending from him by ordinary\index{Sin:original:by ordinary generation ``f10!}
generation,\footnote{\ Acts 17:26\index{44 Acts:ch.  17, v.  26, pg. ``f30!}.} sinned in him, and fell with him in that
first transgression\index{Sin:original:first transgression (of Adam) ``f10!}.\footnote{\ Genesis 2:16-17\index{01
Genesis:ch.   2, v.  16-17, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.  
5, v.  12, pg. ``f30!}{}-20\index{45 Romans:ch.   5, v.  12-20, pg. ``f30!}; 1 Corinthians 15:21-22\index{46 1
Corinthians:ch.  15, v.  21-22, pg. ``f30!}.} 


\bigskip

Q. 23 \textit{Into what estate}\index{Sin:original:an estate of sin and misery ``f10!}\textit{ did the
fall}\index{Sin:original:first transgression (of Adam) ``f10!}\textit{ bring mankind}\index{Sin:original:all mankind
descending from Adam ``f10!}\textit{?}

A. The fall\index{Sin:original:first transgression (of Adam) ``f10!} brought mankind\index{Sin:original:all mankind
descending from Adam ``f10!} into an estate of sin\index{Sin:original:an estate of sin and misery ``f10!} and
misery\index{Punishment:misery ``f10!}.\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}3\index{45
Romans:ch.   3, v.  23, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5,
v.  12, pg. ``f30!}}

Q. 24 \textit{What is sin?}

A. Sin is\index{Sin:defined ``f10!} any want of conformity unto, or transgression of\index{God's
law:commandments:transgressions ``f10!}, any law of God, given as a rule\index{God's law:moral:rule for obedience
``f10!} to the reasonable\index{Mankind:created:reasonable soul ``f10!} creature\index{creatures ``f10!}.\footnote{\ 1
John 3:4\index{62 1 John:ch.   3, v.   4, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}0,
12\index{62 1 John:ch.   3, v.  10, 12, pg. ``f30!}.} 


\bigskip

Q. 25 \textit{Wherein consists the sinfulness of that estate whereinto}\index{Sin:original:an estate of sin and misery
``f10!}\textit{ man fell}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{?}

A. The sinfulness of that estate whereinto\index{Sin:original:an estate of sin and misery ``f10!} man
fell\index{Sin:original:all mankind descending from Adam ``f10!}, consists in the guilt\index{guilt ``f10!} of
Adam\index{Adam:guilt of first sin on all mankind ``f10!}{}'s first sin\index{Sin:original:first transgression (of
Adam) ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg.
``f30!}, 19\index{45 Romans:ch.   5, v.  12, 19, pg. ``f30!}.} the want of that righteousness
wherein\index{Sin:original:original righteousness corrupted ``f10!} he was created\index{Mankind:created:state ``f10!},
and the corruption of his\index{Sin:original:corrupt nature:whole ``f10!} nature, whereby he is utterly
indisposed\index{Sin:original:utterly indisposed ``f10!}, disabled\index{Sin:original:utterly unable and unwilling to
do will of God ``f10!}, and made opposite unto all that is spiritually good, and\index{Sin:original:lost all spiritual
good ``f10!} wholly inclined\index{Sin:original:inclined to do evil ``f10!} to all evil, and that
continually;\footnote{\ Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}{}-19\index{45 Romans:ch.   3, v. 
10-19, pg. ``f30!}; Romans 5:6\index{45 Romans:ch.   5, v.   6, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.  
7, pg. ``f30!}{}-8\index{45 Romans:ch.   8, v.   7-8, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1,
pg. ``f30!}{}-3\index{49 Ephesians:ch.   2, v.   1-3, pg. ``f30!}; Genesis 6:5\index{01 Genesis:ch.   6, v.   5, pg.
``f30!}.} which is commonly called Original\index{Sin:original:source of all transgressions ``f10!} Sin, and from which
do proceed all\index{Sin:original:proceeds all transgressions ``f10!} actual transgressions\index{God's
law:commandments:transgressions ``f10!}.\footnote{\ James 1:14\index{59 James:ch.   1, v.  14, pg.
``f30!}{}-15\index{59 James:ch.   1, v.  14-15, pg. ``f30!}; Matthew 15:19\index{40 Matthew:ch.  15, v.  19, pg.
``f30!}.} 


\bigskip

Q. 26 \textit{How is original sin conveyed from our first parents}\index{Parents:first, sinned against God
``f10!}\textit{ unto their posterity}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{?}

A. Original\index{Sin:original:source of all transgressions ``f10!} sin is conveyed from our first
parents\index{Parents:first, sinned against God ``f10!} unto their posterity\index{Sin:original:all mankind descending
from Adam ``f10!} by natural generation, so\index{Sin:original:by ordinary generation ``f10!} as all that proceed from
the\index{Holy Spirit:procession from Father and Son ``f10!}m in that way are conceived
and\index{Sin:original:conceived and born in ``f10!} born in sin.\footnote{\ Psalm 51:5\index{19 Psalms:ch.  51, v.  
5, pg. ``f30!}; Job 14:4\index{18 Job:ch.  14, v.   4, pg. ``f30!}; Job 15:14\index{18 Job:ch.  15, v.  14, pg.
``f30!}; John 3:6\index{43 John:ch.   3, v.   6, pg. ``f30!}.} 


\bigskip

Q. 27 \textit{What misery}\index{Punishment:misery ``f10!}\textit{ did the fall}\index{Sin:original:first transgression
(of Adam) ``f10!}\textit{ bring upon mankind}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{?}

A. The fall\index{Sin:original:first transgression (of Adam) ``f10!} brought upon mankind\index{Sin:original:all mankind
descending from Adam ``f10!} the loss\index{Communion:with God, lost ``f10!} of communion with God\index{Communion:with
God ``f10!},\footnote{\ Genesis 3:8\index{01 Genesis:ch.   3, v.   8, pg. ``f30!}, 10, 24\index{01 Genesis:ch.   3, v. 
 8, 10, 24, pg. ``f30!}.} his displeasure and\index{Sin:original:under God's displeasure and curse ``f10!} curse;
so\index{Curse:of God ``f10!} as we are by nature children of wrath\index{Wrath:of God:mankind by nature under
``f10!},\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   2, v.  
2-3, pg. ``f30!}.} bond slaves to Satan\index{Satan:bondage to ``f10!},\footnote{\ 2 Timothy 2:2\index{55 2 Timothy:ch.
  2, v.   2, pg. ``f30!}6\index{55 2 Timothy:ch.   2, v.  26, pg. ``f30!}.} and justly
liable\index{Punishment:just:result of Adam's sin ``f10!} to all punishments\index{Punishment:temporal:evils that
befall us ``f10!} in this world, and that which is to come.\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17,
pg. ``f30!}; Lamentations 3:39\index{25 Lamentations:ch.   3, v.  39, pg. ``f30!}; Romans 6:23\index{45 Romans:ch.   6,
v.  23, pg. ``f30!}; Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}, 46\index{40 Matthew:ch.  25, v.  41,
46, pg. ``f30!}; Jude 1:7\index{65 Jude:v.   1, v.   7, pg. ``f30!}.}


\bigskip


\bigskip


Q. 28 \textit{What are the punishment}\index{Punishment:of sin ``f10!}\textit{s}\index{Punishment:temporal:evils that
befall us ``f10!}\textit{ of sin in this world?}

A. The punishments\index{Punishment:temporal:evils that befall us ``f10!} of sin in this world are either inward,
as\index{Punishment:temporal:inward instance classes ``f10!} blindness\index{blindness, spiritual ``f10!} of
mind,\footnote{\ Ephesians 4:18\index{49 Ephesians:ch.   4, v.  18, pg. ``f30!}.} a
reprobate\index{Punishment:temporal:reprobate sense ``f10!} sense,\footnote{\ Romans 1:28\index{45 Romans:ch.   1, v. 
28, pg. ``f30!}.} strong delusions\index{Punishment:temporal:delusions ``f10!},\footnote{\ 2 Thessalonians
2:11\index{53 2 Thessalonians:ch.   2, v.  11, pg. ``f30!}.} hardness of heart\index{Heart:hardened
``f10!},\footnote{\ Romans 2:5\index{45 Romans:ch.   2, v.   5, pg. ``f30!}.} horror of
conscience\index{Conscience:horror of ``f10!},\footnote{\ Isaiah 33:14\index{23 Isaiah:ch.  33, v.  14, pg. ``f30!};
Genesis 4:13\index{01 Genesis:ch.   4, v.  13, pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v.   4, pg.
``f30!}.} and vile affections\index{Sexual immorality:vile ``f10!};\footnote{\ Romans 1:26\index{45 Romans:ch.   1, v. 
26, pg. ``f30!}.} or outward, as\index{Punishment:temporal:outward instance classes ``f10!} the curse of
God\index{Curse:of God ``f10!} upon the creatures\index{creatures ``f10!} for our sakes,\footnote{\ Genesis
3:17\index{01 Genesis:ch.   3, v.  17, pg. ``f30!}.} and all other evils that befall us in our bodies, names, estates,
relations, and employments\index{Punishment:temporal:aspects of life ``f10!};\footnote{\ Deuteronomy 28:15-18\index{05
Deuteronomy:ch.  28, v.  15-18, pg. ``f30!}.} together with death\index{Death:punishment for sin ``f10!}
itself.\footnote{\ Romans 6:21, 23\index{45 Romans:ch.   6, v.  21, 23, pg. ``f30!}.} 


\bigskip

Q. 29 \textit{What are the punishment}\index{Punishment:of sin ``f10!}\textit{s}\index{Punishment:temporal:evils that
befall us ``f10!}\textit{ of sin in the world to come?}

A. The punishments\index{Punishment:temporal:evils that befall us ``f10!} of sin in the world to come, are everlasting
separation\index{Punishment:everlasting:separation from God ``f10!} from the comfortable presence of God, and most
grievous torments in\index{Punishment:everlasting:torments (body and soul) ``f10!} soul and body, without
intermission\index{Punishment:hell:no intermission ``f10!}, in hell\index{Hell:hell ``f10!} fire\index{Hell:fire
``f10!} forever.\footnote{\ 2 Thessalonians 1:9\index{53 2 Thessalonians:ch.   1, v.   9, pg. ``f30!}; Mark 9:43-44,
46, 48\index{41 Mark:ch.   9, v.  43-44, 46, 48, pg. ``f30!}; Luke 16:24\index{42 Luke:ch.  16, v.  24, pg. ``f30!}.}


\bigskip

Q. 30 \textit{Does God leave all mankind}\index{Mankind:not all left to perish in sin and misery ``f10!}\textit{ to
perish}\index{Redemption:not all perish ``f10!}\textit{ in the estate of sin}\index{Sin:original:an estate of sin and
misery ``f10!}\textit{ and misery}\index{Punishment:misery ``f10!}\textit{?}

A. God does not leave all men to perish\index{Redemption:not all perish ``f10!} in the estate of
sin\index{Sin:original:an estate of sin and misery ``f10!} and misery\index{Punishment:misery ``f10!},\footnote{\ 1
Thessalonians 5:9\index{52 1 Thessalonians:ch.   5, v.   9, pg. ``f30!}.} into which they fell\index{Sin:original:all
mankind descending from Adam ``f10!} by the breach of the\index{God's law:breach:first covenant ``f10!} first covenant,
commonly called the Covenant of Works\index{Covenant:works:works ``f10!};\footnote{\ Galatians 3:10\index{48
Galatians:ch.   3, v.  10, pg. ``f30!}, 12\index{48 Galatians:ch.   3, v.  10, 12, pg. ``f30!}.} but of his mere love
and\index{God's attributes:love ``f10!} mercy delivers\index{God's attributes:merciful ``f10!} his
elect\index{Election:elect of God ``f10!} out of it, and brings them into an estate of salvation\index{Salvation:belief
in Jesus by faith ``f10!} by the second covenant\index{Covenant:second ``f10!}, commonly called the Covenant of
Grace\index{Covenant:grace:grace ``f10!}.\footnote{\ Titus 3:4-7\index{56 Titus:ch.   3, v.   4-7, pg. ``f30!};
Galatians 3:21\index{48 Galatians:ch.   3, v.  21, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg.
``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}{}-22\index{45 Romans:ch.   3, v.  20-22, pg. ``f30!}.}


\bigskip

Q. 31 \textit{With whom was the covenant of grace}\index{Covenant:grace:grace ``f10!}\textit{ made?}

A. The covenant of grace\index{Covenant:grace:grace ``f10!} was made with Christ as the second Adam\index{Jesus:second
Adam ``f10!}, and in him with all the elect\index{Election:elect of God ``f10!} as his seed\index{Election:God's seed
``f10!}.\footnote{\ Galatians 3:16\index{48 Galatians:ch.   3, v.  16, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5,
v.   1, pg. ``f30!}5-21\index{45 Romans:ch.   5, v.  15-21, pg. ``f30!}; Isaiah 53:10\index{23 Isaiah:ch.  53, v.  10,
pg. ``f30!}{}-11\index{23 Isaiah:ch.  53, v.  10-11, pg. ``f30!}.} 


\bigskip

Q. 32 \textit{How is the grace}\index{Covenant:grace:manifested ``f10!}\textit{ of God
manifested}\index{Grace:manifested ``f10!}\textit{ in the second covenant}\index{Covenant:second ``f10!}\textit{?}

A. The grace of God\index{Covenant:grace:manifested ``f10!} is manifested\index{Grace:manifested ``f10!} in the second
covenant\index{Covenant:second ``f10!}, in that he freely\index{Salvation:freely offered ``f10!} provides and offers to
sinners a Mediator\index{Jesus:mediator:of covenant ``f10!},\footnote{\ Genesis 3:15\index{01 Genesis:ch.   3, v.  15,
pg. ``f30!}; Isaiah 42:6\index{23 Isaiah:ch.  42, v.   6, pg. ``f30!}; John 6:27\index{43 John:ch.   6, v.  27, pg.
``f30!}.} and life and salvation\index{Salvation:belief in Jesus by faith ``f10!} by him;\footnote{\ 1 John
5:11-12\index{62 1 John:ch.   5, v.  11-12, pg. ``f30!}.} and requiring\index{Salvation:faith alone required ``f10!}
faith as the condition\index{Faith:necessary for salvation ``f10!} to interest them\index{Faith:required for interest
(share) in Christ ``f10!}\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v. 
12, pg. ``f30!}; John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}.} in him, promises\index{Covenant:promises
``f10!} and gives his Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}\footnote{\ Proverbs 1:23\index{20 Proverbs:ch. 
 1, v.  23, pg. ``f30!}.} to all his elect\index{Election:elect of God ``f10!}, to work in them that
faith\index{Faith:worked in believers by the Holy Spirit ``f10!},\footnote{\ 2 Corinthians 4:13\index{47 2
Corinthians:ch.   4, v.  13, pg. ``f30!}.} with all other saving graces\index{Grace:of salvation
``f10!};\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}2-23\index{48 Galatians:ch.   5, v. 
22-23, pg. ``f30!}.} and to enable\index{God's law:commandments:obedience:enabled by Holy Spirit ``f10!} them unto all
holy\index{Believers:holy:being perfected ``f10!} obedience,\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27,
pg. ``f30!}.} as the evidence of the truth of their faith\index{Faith:evidenced by obedience ``f10!}\footnote{\ James
2:18, 22\index{59 James:ch.   2, v.  18, 22, pg. ``f30!}.} and thankfulness\index{Prayer:thanking God ``f10!} to
God,\footnote{\ 2 Corinthians 5:14-15\index{47 2 Corinthians:ch.   5, v.  14-15, pg. ``f30!}.} and as the way which he
has appointed them to salvation.\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}8\index{49
Ephesians:ch.   2, v.  18, pg. ``f30!}.} 


\bigskip

Q. 33 \textit{Was the covenant of grace}\index{Covenant:grace:grace ``f10!}\textit{ always administered after one and
the same manner?}

A. The covenant of grace\index{Covenant:grace:grace ``f10!} was not always administered after the same manner, but the
administrations of it\index{Covenant:administrations ``f10!} under the Old Testament\index{Scripture:Old Testament
``f10!} were different from those under the New.\footnote{\ 2 Corinthians 3:6\index{47 2 Corinthians:ch.   3, v.   6,
pg. ``f30!}{}-9\index{47 2 Corinthians:ch.   3, v.   6-9, pg. ``f30!}.}


\bigskip

Q. 34 \textit{How was the covenant of grace}\index{Covenant:grace:grace ``f10!}\textit{ administered under the Old
Testament}\index{Scripture:Old Testament ``f10!}\textit{?}

A. The covenant of grace\index{Covenant:grace:grace ``f10!} was administered under the Old Testament\index{Scripture:Old
Testament ``f10!}, by promises\index{Covenant:promises ``f10!},\footnote{\ Romans 15:8\index{45 Romans:ch.  15, v.   8,
pg. ``f30!}.} prophecies\index{Covenant:prophecies ``f10!},\footnote{\ Acts 3:20, 24\index{44 Acts:ch.   3, v.  20, 24,
pg. ``f30!}.} sacrifices\index{Ceremonial law:sacrifices ``f10!},\footnote{\ Hebrews 10:1\index{58 Hebrews:ch.  10, v. 
 1, pg. ``f30!}.} circumcision\index{circumcision ``f10!},\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg.
``f30!}.} the passover\index{passover ``f10!},\footnote{\ 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7, pg.
``f30!}.} and other types\index{types, OT ``f10!} and ordinances, which\index{Covenant:OT ordinances:signified Christ
``f10!} did all fore signify Christ then to come, and were for that time sufficient to\index{Covenant:OT
ordinances:sufficient for that time ``f10!} build up the\index{Holy Spirit:builds up the elect ``f10!}
elect\index{Election:elect of God ``f10!} in faith\index{Faith:in Jesus Christ, as Messiah ``f10!} in the
promised\index{Jesus:Messiah promised ``f10!} Messiah,\footnote{\ Hebrews 8-10\index{58 Hebrews:ch. 8-10, pg. ``f30!};
Hebrews 11:13\index{58 Hebrews:ch.  11, v.  13, pg. ``f30!}.} by whom they then had full
remission\index{Redemption:remission of sins ``f10!} of sin, and eternal salvation\index{Salvation:belief in Jesus by
faith ``f10!}.\footnote{\ Galatians 3:7-9, 14\index{48 Galatians:ch.   3, v.   7-9, 14, pg. ``f30!}.} 


\bigskip

Q. 35 \textit{How is the covenant of grace}\index{Covenant:grace:grace ``f10!}\textit{ administered under the New
Testament}\index{Scripture:New Testament ``f10!}\textit{?}

A. Under the New Testament\index{Scripture:New Testament ``f10!}, when Christ the substance was exhibited,
the\index{Covenant:grace:exhibited ``f10!} same covenant of grace\index{Covenant:grace:grace ``f10!} was and still is
to be administered\index{Covenant:grace:administered through preaching of word ``f10!} in the preaching of the
word,\footnote{\ Mark 16:15\index{41 Mark:ch.  16, v.  15, pg. ``f30!}.} and the
administration\index{Church:true:sacraments administered ``f10!} of the sacraments\index{Sacraments:holy:ordinances
``f10!} of Baptism\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch. 
28, v.  19-20, pg. ``f30!}.} and the Lord's Supper\index{Lord's Supper:sacrament ``f10!};\footnote{\ 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg.
``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  23-25, pg. ``f30!}.} in which grace\index{Grace:exhibited in
sacraments ``f10!} and salvation\index{Salvation:belief in Jesus by faith ``f10!} are held forth in more
fulness\index{Covenant:new:more fulness ``f10!}, evidence, and efficacy, to\index{Sacraments:spiritual efficacy ``f10!}
all nations\index{Gospel:to all nations ``f10!}.\footnote{\ 2 Corinthians 3:6\index{47 2 Corinthians:ch.   3, v.   6,
pg. ``f30!}{}-9\index{47 2 Corinthians:ch.   3, v.   6-9, pg. ``f30!}; Hebrews 8:6, 10-11\index{58 Hebrews:ch.   8, v. 
 6, 10-11, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.}


\bigskip


\bigskip

Q. 36 \textit{Who is the Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ of the covenant of
grace}\index{Covenant:grace:grace ``f10!}\textit{?}

A. The only Mediator\index{Jesus:mediator:of covenant ``f10!} of the covenant of grace\index{Covenant:grace:grace
``f10!} is the Lord Jesus Christ,\footnote{\ 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.} who, being
the eternal\index{Jesus:eternal Son ``f10!} Son\index{Jesus:Son of God ``f10!} of God, of one substance\index{God's
attributes:trinity:one substance ``f10!} and equal with the\index{God's attributes:trinity:persons equal ``f10!}
Father\index{Father:God the ``f10!},\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}, 14\index{43
John:ch.   1, v.   1, 14, pg. ``f30!}; John 10:30\index{43 John:ch.  10, v.  30, pg. ``f30!}; Philippians 2:6\index{50
Philippians:ch.   2, v.   6, pg. ``f30!}.} in the fulness of time\index{Jesus:incarnation:in fulness of time ``f10!}
became man,\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} and so was and
continues\index{Jesus:God-man ``f10!} to be God and man, in two entire distinct natures\index{Jesus:two distinct
natures ``f10!}, and one person, forever.\footnote{\ Luke 1:35\index{42 Luke:ch.   1, v.  35, pg. ``f30!}; Romans
9:5\index{45 Romans:ch.   9, v.   5, pg. ``f30!}; Colossians 2:9\index{51 Colossians:ch.   2, v.   9, pg. ``f30!};
Hebrews 7:24-25\index{58 Hebrews:ch.   7, v.  24-25, pg. ``f30!}.} 


\bigskip

Q. 37 \textit{How did Christ, being the Son}\index{Jesus:Son of God ``f10!}\textit{ of God, become man?}

A. Christ the Son\index{Jesus:Son of God ``f10!} of God became man, by taking to himself a true body, and a
reasonable\index{Jesus:human nature:reasonable soul ``f10!} soul,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1,
pg. ``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}; Matthew 26:38\index{40 Matthew:ch.  26, v.  38, pg. ``f30!}.}
being conceived by\index{Jesus:conceived:by the power of the Holy Spirit ``f10!} the power of the Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!} in the womb\index{Jesus:conceived:in womb of virgin Mary ``f10!} of the
virgin\index{Jesus:born of virgin Mary ``f10!} Mary\index{Mary, mother of Jesus ``f10!}, of her
substance\index{Jesus:conceived:of Mary's substance ``f10!}, and born of her,\footnote{\ Luke 1:27, 31, 35\index{42
Luke:ch.   1, v.  27, 31, 35, pg. ``f30!}, 42\index{42 Luke:ch.   1, v.  27, 31, 35, 42, pg. ``f30!}; Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} yet without sin.\footnote{\ Hebrews 4:15\index{58 Hebrews:ch.  
4, v.  15, pg. ``f30!}; Hebrews 7:26\index{58 Hebrews:ch.   7, v.  26, pg. ``f30!}.}


\bigskip

Q. 38 \textit{Why was it requisite that}\index{Jesus:mediator:requisite that he be both God and man ``f10!}\textit{ the
Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ should be God?}

A. It was requisite that\index{Jesus:mediator:requisite that he be both God and man ``f10!} the
Mediator\index{Jesus:mediator:of covenant ``f10!} should be God, that he might sustain and keep the human nature
from\index{Jesus:human nature:sustained by God ``f10!} sinking under the infinite\index{God's attributes:infinite:wrath
``f10!} wrath of God\index{Wrath:of God:for sin ``f10!}, and the power of death\index{Jesus:death:under power of
``f10!};\footnote{\ Acts. 2:24-25; Romans 1:4\index{45 Romans:ch.   1, v.   4, pg. ``f30!}; Romans 4:25\index{45
Romans:ch.   4, v.  25, pg. ``f30!}; Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}.} give worth
and\index{Jesus:worth of his sufferings ``f10!} efficacy to his\index{Redemption:efficacy of Christ's incarnation
``f10!} sufferings, obedience\index{Jesus:obedience:to God's law ``f10!}, and intercession\index{Jesus:in
heaven:intercession made for all believers ``f10!};\footnote{\ Acts 20:28\index{44 Acts:ch.  20, v.  28, pg. ``f30!};
Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}{}-28\index{58 Hebrews:ch.   7, v.  25-28, pg. ``f30!};
Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}.} and to satisfy God's justice\index{Jesus:satisfied divine
justice ``f10!},\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24,
pg. ``f30!}{}-26\index{45 Romans:ch.   3, v.  24-26, pg. ``f30!}.} procure his favor\index{Jesus:favour with the Father
``f10!},\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}; Matthew 3:17\index{40 Matthew:ch.  
3, v.  17, pg. ``f30!}.} purchase\index{Redemption:purchased by Christ ``f10!} a peculiar people,\footnote{\ Titus.
2:13-14.} give his Spirit to them,\footnote{\ Galatians 4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.}
conquer\index{Jesus:conqueror of his and our enemies ``f10!} all their enemies\index{Enemies:of believers overcome
``f10!},\footnote{\ Luke 1:6\index{42 Luke:ch.   1, v.   6, pg. ``f30!}8-69, 71, 74\index{42 Luke:ch.   1, v.  68-69,
71, 74, pg. ``f30!}.} and bring them to everlasting\index{Salvation:everlasting ``f10!}
salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Hebrews 5:8-9\index{58 Hebrews:ch.   5, v.  
8-9, pg. ``f30!}; Hebrews 9:11-15\index{58 Hebrews:ch.   9, v.  11-15, pg. ``f30!}.}


\bigskip

Q. 39 \textit{Why was it requisite that}\index{Jesus:mediator:requisite that he be both God and man ``f10!}\textit{ the
Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ should be man?}

A. It was requisite that\index{Jesus:mediator:requisite that he be both God and man ``f10!} the
Mediator\index{Jesus:mediator:of covenant ``f10!} should be man, that he might advance\index{Jesus:mediator:advances
human nature ``f10!} our nature,\footnote{\ Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}6\index{58
Hebrews:ch.   2, v.  16, pg. ``f30!}.} perform obedience\index{Jesus:obedience:to God's law ``f10!} to the
law,\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} suffer and\index{Jesus:death:sufferings
and death ``f10!} make intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} for us in our
nature,\footnote{\ Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   2, v.  14, pg.
``f30!}; Hebrews 7:24-25\index{58 Hebrews:ch.   7, v.  24-25, pg. ``f30!}.} have a fellow feeling\index{Jesus:fellow
feelings with our infirmities ``f10!} of our infirmities;\index{Jesus:humiliation:suffered:infirmities
``f10!}\footnote{\ Hebrews 4:15\index{58 Hebrews:ch.   4, v.  15, pg. ``f30!}.} that we might receive the
adoption\index{Adoption:in salvation ``f10!} of sons,\footnote{\ Galatians 4:5\index{48 Galatians:ch.   4, v.   5, pg.
``f30!}.} and have comfort\index{Faith:source of comfort ``f10!} and access with\index{Believers:bold access to God's
throne ``f10!} boldness\index{Prayer:with boldness ``f10!} unto the throne of grace\index{Grace:throne of
``f10!}.\footnote{\ Hebrews 4:16\index{58 Hebrews:ch.   4, v.  16, pg. ``f30!}.}

Q. 40 \textit{Why was it requisite that}\index{Jesus:mediator:requisite that he be both God and man ``f10!}\textit{ the
Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ should be God and man in one person?}

A. It was requisite that\index{Jesus:mediator:requisite that he be both God and man ``f10!} the
Mediator\index{Jesus:mediator:of covenant ``f10!}, who was to reconcile\index{Jesus:purchased reconciliation ``f10!}
God and man, should himself be both God and man, and this in one person, that the proper works\index{Jesus:proper works
of each nature ``f10!} of each nature might be accepted of\index{Jesus:accepted of God ``f10!} God for us, and
relied\index{Jesus:mediator:relied on by us for salvation ``f10!} on by us,\footnote{\ Matthew 1:21\index{40
Matthew:ch.   1, v.  21, pg. ``f30!}, 23\index{40 Matthew:ch.   1, v.  21, 23, pg. ``f30!}; Matthew 3:17\index{40
Matthew:ch.   3, v.  17, pg. ``f30!}; Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}.} as the works of the
whole person.\footnote{\ 1 Peter 2:6\index{60 1 Peter:ch.   2, v.   6, pg. ``f30!}.}


\bigskip

Q. 41 \textit{Why was our Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ called
Jesus}\index{Jesus:mediator:named Jesus ``f10!}\textit{?}

A. Our Mediator\index{Jesus:mediator:of covenant ``f10!} was called Jesus\index{Jesus:mediator:named Jesus ``f10!},
because he saves\index{Jesus:saviour of the Church ``f10!} his people from their sins.\footnote{\ Matthew 1:21\index{40
Matthew:ch.   1, v.  21, pg. ``f30!}.} 


\bigskip

Q. 42 \textit{Why was our Mediator}\index{Jesus:mediator:of covenant ``f10!}\textit{ called Christ?}

A. Our Mediator\index{Jesus:mediator:of covenant ``f10!} was called Christ, because he was anointed\index{Jesus:anointed
with Holy Spirit ``f10!} with the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!} above measure;\footnote{\ John
3:3\index{43 John:ch.   3, v.   3, pg. ``f30!}4\index{43 John:ch.   3, v.  34, pg. ``f30!}; Psalm 45:7\index{19
Psalms:ch.  45, v.   7, pg. ``f30!}.} and so set apart, and fully furnished\index{Jesus:mediator:fully furnished to
execute office ``f10!} with all authority and ability,\footnote{\ John 6:27\index{43 John:ch.   6, v.  27, pg. ``f30!};
Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  18-20, pg. ``f30!}.} to
execute the office\index{Jesus:offices ``f10!}s of prophet\index{Jesus:prophet:office ``f10!},\footnote{\ Acts
3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!}{}-22\index{44 Acts:ch.   3, v.  21-22, pg. ``f30!}; Luke 4:18,
21\index{42 Luke:ch.   4, v.  18, 21, pg. ``f30!}.} priest,\footnote{\ Hebrews 4:14-15\index{58 Hebrews:ch.   4, v. 
14-15, pg. ``f30!}; Hebrews 5:5-7\index{58 Hebrews:ch.   5, v.   5-7, pg. ``f30!}.} and king\index{Jesus:King ``f10!}
of his church\index{Church:Christ's ``f10!},\footnote{\ Psalm 2:6\index{19 Psalms:ch.   2, v.   6, pg. ``f30!}; Matthew
21:5\index{40 Matthew:ch.  21, v.   5, pg. ``f30!}; Isaiah 9:6\index{23 Isaiah:ch.   9, v.   6, pg.
``f30!}{}-7\index{23 Isaiah:ch.   9, v.   6-7, pg. ``f30!}; Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg.
``f30!}{}-11\index{50 Philippians:ch.   2, v.   8-11, pg. ``f30!}.} in the estate both of his humiliation
and\index{Jesus:humiliation:aspects of ``f10!} exaltation\index{Jesus:exaltation ``f10!}.


\bigskip

Q. 43 \textit{How does Christ execute the office}\index{Jesus:offices ``f10!}\textit{ of a
prophet}\index{Jesus:prophet:office ``f10!}\textit{?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a prophet\index{Jesus:prophet:office ``f10!}, in his
revealing\index{Revelation:God's word ``f10!} to the church,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}8\index{43 John:ch.   1, v.  18, pg. ``f30!}.} in all ages, by\index{Church:universal ``f10!} his Spirit and
word,\footnote{\ 1 Peter 1:10-12\index{60 1 Peter:ch.   1, v.  10-12, pg. ``f30!}.} in divers ways of
administration\index{Jesus:prophet:administration ``f10!},\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg.
``f30!}{}-2\index{58 Hebrews:ch.   1, v.   1-2, pg. ``f30!}.} the whole will of God,\footnote{\ John 15:15\index{43
John:ch.  15, v.  15, pg. ``f30!}.} in all things concerning their edification and\index{Holy Spirit:edification of
believers ``f10!} salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Acts 20:32\index{44 Acts:ch. 
20, v.  32, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-13\index{49 Ephesians:ch.  
4, v.  11-13, pg. ``f30!}; John 20:31\index{43 John:ch.  20, v.  31, pg. ``f30!}.}


\bigskip

Q. 44 \textit{How does Christ execute the office}\index{Jesus:offices ``f10!}\textit{ of a priest?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a priest, in his\index{Jesus:priest ``f10!} once offering
himself\index{Jesus:sacrifice:of himself ``f10!} a sacrifice without spot\index{Jesus:sacrifice:without spot ``f10!} to
God,\footnote{\ Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}, 28\index{58 Hebrews:ch.   9, v.  14, 28,
pg. ``f30!}.} to be a reconciliation\index{Jesus:purchased reconciliation ``f10!} for the sins of his
people;\footnote{\ Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}7\index{58 Hebrews:ch.   2, v.  17, pg.
``f30!}.} and in making continual intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} for
them.\footnote{\ Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}.} 


\bigskip

Q. 45 \textit{How does Christ execute the office}\index{Jesus:offices ``f10!}\textit{ of a king}\index{Jesus:King
``f10!}\textit{?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a king\index{Jesus:King ``f10!}, in calling
out\index{Effectual call:elect ``f10!} of the world a people to himself,\footnote{\ Acts 15:14-16\index{44 Acts:ch. 
15, v.  14-16, pg. ``f30!}; Isaiah 4:4-5\index{23 Isaiah:ch.   4, v.   4-5, pg. ``f30!}; Genesis 49:10\index{01
Genesis:ch.  49, v.  10, pg. ``f30!}; Psalm 110:3\index{19 Psalms:ch. 110, v.   3, pg. ``f30!}.} and giving them
officers\index{Church officers:officers ``f10!},\footnote{\ Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg.
``f30!}{}-12\index{49 Ephesians:ch.   4, v.  11-12, pg. ``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.
 28, pg. ``f30!}.} laws,\footnote{\ Isaiah 33:22\index{23 Isaiah:ch.  33, v.  22, pg. ``f30!}.} and censures,
by\index{Church discipline:censures ``f10!} which he visibly governs\index{Providence:governing all ``f10!}
them;\footnote{\ Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.  18, v.  17-18,
pg. ``f30!}; 1 Corinthians 5:4-5\index{46 1 Corinthians:ch.   5, v.   4-5, pg. ``f30!}.} in bestowing saving
grace\index{Grace:of salvation ``f10!} upon his elect\index{Election:elect of God ``f10!},\footnote{\ Acts 5:3\index{44
Acts:ch.   5, v.   3, pg. ``f30!}1\index{44 Acts:ch.   5, v.  31, pg. ``f30!}.} rewarding their
obedience,\index{Believers:obedience:rewarded ``f10!}\footnote{\ Revelation 2:1\index{66 Revelation:ch.   2, v.   1,
pg. ``f30!}0\index{66 Revelation:ch.   2, v.  10, pg. ``f30!}; Revelation 22:12\index{66 Revelation:ch.  22, v.  12,
pg. ``f30!}.} and correcting them for\index{Jesus:corrects believers ``f10!} their sins,\footnote{\ Revelation
3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}9\index{66 Revelation:ch.   3, v.  19, pg. ``f30!}.} preserving and
supporting\index{Temptation:believers supported by Christ during ``f10!} them under all their
temptations\index{Temptation:temptation ``f10!} and sufferings,\footnote{\ Isaiah 63:9\index{23 Isaiah:ch.  63, v.   9,
pg. ``f30!}.} restraining\index{Enemies:of believers restrained ``f10!} and overcoming\index{Jesus:conqueror of his and
our enemies ``f10!} all their enemies\index{Enemies:of believers overcome ``f10!},\footnote{\ 1 Corinthians
15:25\index{46 1 Corinthians:ch.  15, v.  25, pg. ``f30!}; Psalm 110:1\index{19 Psalms:ch. 110, v.   1, pg.
``f30!}{}-2\index{19 Psalms:ch. 110, v.   1-2, pg. ``f30!}.} and powerfully ordering all\index{Providence:ordering:all
things for his glory and believers' good ``f10!} things for his own glory\index{God's attributes:glory:his own
``f10!},\footnote{\ Romans 14:10-11\index{45 Romans:ch.  14, v.  10-11, pg. ``f30!}.} and their good;\footnote{\ Romans
8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!}.} and also in taking
vengeance\index{Jesus:judge:taking vengeance ``f10!} on the rest, who know not God, and obey not\index{Punishment:those
who do not obey the Gospel ``f10!} the gospel\index{Gospel:gospel ``f10!}.\footnote{\ 2 Thessalonians 1:8-9\index{53 2
Thessalonians:ch.   1, v.   8-9, pg. ``f30!}; Psalm 2:8\index{19 Psalms:ch.   2, v.   8, pg. ``f30!}{}-9\index{19
Psalms:ch.   2, v.   8-9, pg. ``f30!}.} 


\bigskip

Q. 46 \textit{What was the estate of Christ's humiliation}\index{Jesus:humiliation:estate ``f10!}\textit{?}

A. The estate of Christ's humiliation\index{Jesus:humiliation:estate ``f10!} was that low
condition\index{Jesus:humiliation:low condition ``f10!}, wherein he for our sakes,
emptying\index{Jesus:humiliation:emptying himself of glory ``f10!} himself of his glory, took\index{Jesus:glory ``f10!}
upon him the form of a servant\index{Jesus:humiliation:took on form of a servant ``f10!}, in his
conception\index{Jesus:humiliation:conception ``f10!} and birth\index{Jesus:birth ``f10!}, life, death, and
after\index{Jesus:death:death ``f10!} his death, until his resurrection\index{Jesus:resurrection:from the dead
``f10!}.\footnote{\ Philippians 2:6\index{50 Philippians:ch.   2, v.   6, pg. ``f30!}{}-8\index{50 Philippians:ch.   2,
v.   6-8, pg. ``f30!}; Luke 1:31\index{42 Luke:ch.   1, v.  31, pg. ``f30!}; 2 Corinthians 8:9\index{47 2
Corinthians:ch.   8, v.   9, pg. ``f30!}; Acts 2:24\index{44 Acts:ch.   2, v.  24, pg. ``f30!}.}


\bigskip

Q. 47 \textit{How did Christ humble himself}\index{Jesus:humiliation:aspects of ``f10!}\textit{ in his
conception}\index{Jesus:humiliation:conception ``f10!}\textit{ and birth}\index{Jesus:birth ``f10!}\textit{?}

A. Christ humbled himself\index{Jesus:humiliation:aspects of ``f10!} in his
conception\index{Jesus:humiliation:conception ``f10!} and birth\index{Jesus:birth ``f10!}, in that, being from all
eternity\index{God's attributes:eternal ``f10!} the Son\index{Jesus:Son of God ``f10!} of God, in the bosom of the
Father\index{Father:God the ``f10!}, he was pleased\index{Providence:at God's pleasure ``f10!} in the fulness of
time\index{Jesus:incarnation:in fulness of time ``f10!} to become the son of man, made of a woman of low
estate\index{Jesus:born of woman of low estate ``f10!}, and to be born of\index{Jesus:born of virgin Mary ``f10!} her;
with divers circumstances of more than ordinary abasement\index{Jesus:humiliation:abasement ``f10!}.\footnote{\ John
1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}, 18\index{43 John:ch.   1,
v.  14, 18, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}; Luke 2:7\index{42 Luke:ch.   2,
v.   7, pg. ``f30!}.}


\bigskip

Q. 48 \textit{How did Christ humble himself}\index{Jesus:humiliation:aspects of ``f10!}\textit{ in his life?}

A. Christ humbled himself\index{Jesus:humiliation:aspects of ``f10!} in his life, by
subjecting\index{Jesus:humiliation:subjected himself to the law ``f10!} himself to the law,\footnote{\ Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} which he perfectly fulfil\index{Jesus:obedience:to God's law
``f10!}led;\footnote{\ Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5,
v.   1, pg. ``f30!}9\index{45 Romans:ch.   5, v.  19, pg. ``f30!}.} and by conflicting\index{Jesus:conflicted
with:indignities of world ``f10!} with the indignities\index{Jesus:humiliation:suffered:indignities ``f10!} of the
world,\footnote{\ Psalm 22:6\index{19 Psalms:ch.  22, v.   6, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.  
2, pg. ``f30!}{}-3\index{58 Hebrews:ch.  12, v.   2-3, pg. ``f30!}.} temptations\index{Temptation:temptation ``f10!} of
Satan,\footnote{\ Matthew 4:1-12\index{40 Matthew:ch.   4, v.   1-12, pg. ``f30!}; Luke 4:13\index{42 Luke:ch.   4, v. 
13, pg. ``f30!}.} and infirmities in\index{Jesus:humiliation:suffered:infirmities ``f10!} his flesh, whether common to
the nature of man, or particularly accompanying that his low condition\index{Jesus:humiliation:low condition
``f10!}.\footnote{\ Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}7\index{58 Hebrews:ch.   2, v.  17, pg.
``f30!}{}-18\index{58 Hebrews:ch.   2, v.  17-18, pg. ``f30!}; Hebrews 4:15\index{58 Hebrews:ch.   4, v.  15, pg.
``f30!}; Isaiah 52:13-14\index{23 Isaiah:ch.  52, v.  13-14, pg. ``f30!}.}


\bigskip


\bigskip


Q. 49 \textit{How did Christ humble himself}\index{Jesus:humiliation:aspects of ``f10!}\textit{ in his
death}\index{Jesus:death:death ``f10!}\textit{?}

A. Christ humbled himself\index{Jesus:humiliation:aspects of ``f10!} in his death\index{Jesus:death:death ``f10!}, in
that having been betrayed\index{Jesus:betrayed ``f10!} by Judas\index{Judas ``f10!},\footnote{\ Matthew 27:4\index{40
Matthew:ch.  27, v.   4, pg. ``f30!}.} forsaken by\index{Jesus:forsaken by disciples ``f10!} his
disciples\index{disciples of Jesus ``f10!},\footnote{\ Matthew 26:56\index{40 Matthew:ch.  26, v.  56, pg. ``f30!}.}
scorned\index{Jesus:humiliation:scorned ``f10!} and rejected\index{Jesus:humiliation:rejected by world ``f10!} by the
world,\footnote{\ Isaiah 53:2-3\index{23 Isaiah:ch.  53, v.   2-3, pg. ``f30!}.} condemned by\index{Jesus:condemned by
wicked men ``f10!} Pilate\index{Pilate ``f10!}, and tormented\index{Jesus:humiliation:tormented ``f10!} by his
persecutors\index{Jesus:humiliation:persecuted ``f10!};\footnote{\ Matthew 27:26-50\index{40 Matthew:ch.  27, v. 
26-50, pg. ``f30!}; John 19:34\index{43 John:ch.  19, v.  34, pg. ``f30!}.} having also
conflicted\index{Jesus:conflicted with:terror of death ``f10!} with the terrors of death\index{Jesus:death:terrors of
``f10!}, and the powers of darkness\index{darkness, spiritual ``f10!}, felt and borne the weight of God's
wrath,\index{Wrath:of God:borne by Jesus as sin offering ``f10!}\footnote{\ Luke 22:44\index{42 Luke:ch.  22, v.  44,
pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v.   4, pg. ``f30!}6\index{40 Matthew:ch.  27, v.  46, pg.
``f30!}.} he laid down\index{Jesus:laid down life as sin offering ``f10!} his life an offering
for\index{Jesus:sacrifice:offering for sin ``f10!} sin,\footnote{\ Isaiah 53:10\index{23 Isaiah:ch.  53, v.  10, pg.
``f30!}.} enduring the painful\index{Jesus:humiliation:suffered pain ``f10!},
shameful\index{Jesus:humiliation:suffered:shameful death of cross ``f10!}, and cursed\index{Jesus:death:cursed ``f10!}
death of the cross\index{Jesus:death:on cross ``f10!}.\footnote{\ Philippians 2:8\index{50 Philippians:ch.   2, v.   8,
pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}; Galatians 3:13\index{48 Galatians:ch.   3, v. 
13, pg. ``f30!}.} 


\bigskip

Q. 50 \textit{Wherein consisted}\index{Jesus:humiliation:aspects of ``f10!}\textit{ Christ's humiliation after his
death}\index{Jesus:death:death ``f10!}\textit{?}

A. Christ's humiliation after\index{Jesus:humiliation:aspects of ``f10!} his death\index{Jesus:death:death ``f10!}
consisted in his being buried\index{Jesus:buried ``f10!},\footnote{\ 1 Corinthians 15:3-4\index{46 1 Corinthians:ch. 
15, v.   3-4, pg. ``f30!}.} and continuing\index{Jesus:death:continuing under power for time ``f10!} in the state of
the dead, and under the power of death\index{Jesus:death:under power of ``f10!} till the third
day\index{Jesus:resurrection:third day ``f10!};\footnote{\ Psalm 16:10\index{19 Psalms:ch.  16, v.  10, pg. ``f30!};
Acts 2:24\index{44 Acts:ch.   2, v.  24, pg. ``f30!}{}-27, 31\index{44 Acts:ch.   2, v.  24-27, 31, pg. ``f30!}; Romans
6:9\index{45 Romans:ch.   6, v.   9, pg. ``f30!}; Matthew 12:40\index{40 Matthew:ch.  12, v.  40, pg. ``f30!}.} which
has been otherwise expressed in these words, he descended\index{Jesus:descended into hell ``f10!} into
hell\index{Hell:hell ``f10!}. 


\bigskip

Q. 51 \textit{What was the estate of Christ's exaltation}\index{Jesus:exaltation ``f10!}\textit{?}

A. The estate of Christ's exaltation\index{Jesus:exaltation ``f10!} comprehends his
resurrection\index{Jesus:resurrection:from the dead ``f10!},\footnote{\ 1 Corinthians 15:4\index{46 1 Corinthians:ch. 
15, v.   4, pg. ``f30!}.} ascension\index{Jesus:ascension:into heaven ``f10!},\footnote{\ Mark 16:19\index{41 Mark:ch. 
16, v.  19, pg. ``f30!}.} sitting\index{Jesus:in heaven:sitting at right hand of Father ``f10!} at the right hand of
the Father\index{Father:God the ``f10!},\footnote{\ Ephesians 1:20\index{49 Ephesians:ch.   1, v.  20, pg. ``f30!}.}
and his coming again to judge the\index{Jesus:judge:world ``f10!} world.\footnote{\ Acts 1:11\index{44 Acts:ch.   1, v.
 11, pg. ``f30!}; Acts 17:31\index{44 Acts:ch.  17, v.  31, pg. ``f30!}.}


\bigskip

Q. 52 \textit{How was Christ exalted}\index{Jesus:exaltation ``f10!}\textit{ in his
resurrection}\index{Jesus:resurrection:from the dead ``f10!}\textit{?}

A. Christ was exalted\index{Jesus:exaltation ``f10!} in his resurrection\index{Jesus:resurrection:from the dead ``f10!},
in that, not having seen corruption\index{Jesus:death:no corruption ``f10!} in death (of which it was not possible for
him to be held),\footnote{\ Acts 2:24\index{44 Acts:ch.   2, v.  24, pg. ``f30!}, 27\index{44 Acts:ch.   2, v.  24, 27,
pg. ``f30!}.} and having the very same body in which he suffered\index{Jesus:resurrection:body same:in which he
suffered ``f10!}, with the essential properties thereof\index{Jesus:resurrection:body same:essential properties
``f10!}\footnote{\ Luke 24:39\index{42 Luke:ch.  24, v.  39, pg. ``f30!}.} (but without
mortality\index{Jesus:resurrection:body:immortal ``f10!}, and other common infirmities
belonging\index{Jesus:resurrection:freed from infirmities ``f10!} to this life), really united to
his\index{Jesus:resurrection:body same:united to his soul ``f10!} soul,\footnote{\ Romans 6:9\index{45 Romans:ch.   6,
v.   9, pg. ``f30!}; Revelation 1:18\index{66 Revelation:ch.   1, v.  18, pg. ``f30!}.} he
rose\index{Jesus:resurrection:third day ``f10!} again from the dead the third day by his own
power\index{Jesus:resurrection:by his own power ``f10!};\footnote{\ John 10:18\index{43 John:ch.  10, v.  18, pg.
``f30!}.} whereby he declared\index{Jesus:resurrection:declared Son of God ``f10!} himself to be the
Son\index{Jesus:Son of God ``f10!} of God,\footnote{\ Romans 1:4\index{45 Romans:ch.   1, v.   4, pg. ``f30!}.} to have
satisfied divine justice\index{God's attributes:administers justice ``f10!},\footnote{\ Romans 8:3\index{45 Romans:ch. 
 8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}.} to have vanquished\index{Jesus:death:vanquished
``f10!} death, and him that had the power\index{Satan:vanquished by Christ ``f10!} of it,\footnote{\ Hebrews
2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   2, v.  14, pg. ``f30!}.} and to be Lord of
quick and dead\index{Jesus:lord of quick and dead ``f10!}:\footnote{\ Romans 14:9\index{45 Romans:ch.  14, v.   9, pg.
``f30!}.} all which he did as a public person,\footnote{\ 1 Corinthians 15:21-22\index{46 1 Corinthians:ch.  15, v. 
21-22, pg. ``f30!}.} the head\index{Church:Christ as head ``f10!} of his church\index{Church:Christ's
``f10!},\footnote{\ Ephesians 1:20\index{49 Ephesians:ch.   1, v.  20, pg. ``f30!}, 22-23\index{49 Ephesians:ch.   1,
v.  20, 22-23, pg. ``f30!}; Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg. ``f30!}.} for their
justification\index{Justification:justification ``f10!},\footnote{\ Romans 4:25\index{45 Romans:ch.   4, v.  25, pg.
``f30!}.} quickening in grace,\index{Grace:in believers:by Holy Spirit ``f10!}\footnote{\ Ephesians 2:1\index{49
Ephesians:ch.   2, v.   1, pg. ``f30!}, 5\index{49 Ephesians:ch.   2, v.   1, 5, pg. ``f30!}{}-6\index{49 Ephesians:ch.
  2, v.   1, 5-6, pg. ``f30!}; Colossians 2:12\index{51 Colossians:ch.   2, v.  12, pg. ``f30!}.} support against
enemies\index{Enemies:believers delivered from ``f10!},\footnote{\ 1 Corinthians 15:25\index{46 1 Corinthians:ch.  15,
v.  25, pg. ``f30!}{}-27\index{46 1 Corinthians:ch.  15, v.  25-27, pg. ``f30!}.} and to
assure\index{Jesus:resurrection:assures of believers' resurrection ``f10!} them of their
resurrection\index{Resurrection:believers:assurance of ``f10!} from the dead at the last day\index{Last day:last day
``f10!}.\footnote{\ 1 Corinthians 15:20\index{46 1 Corinthians:ch.  15, v.  20, pg. ``f30!}.} 


\bigskip

Q. 53 \textit{How was Christ exalted}\index{Jesus:exaltation ``f10!}\textit{ in his
ascension}\index{Jesus:ascension:into heaven ``f10!}\textit{?}

A. Christ was exalted\index{Jesus:exaltation ``f10!} in his ascension\index{Jesus:ascension:into heaven ``f10!}, in that
having after his resurrection\index{Jesus:resurrection:from the dead ``f10!} often
appeared\index{Jesus:post-resurrection appearances ``f10!} unto and conversed\index{Believers:conversed with apostles
``f10!} with his apostles\index{Apostles:apostles ``f10!}, speaking to them of the things pertaining to the kingdom of
God\index{Kingdom of God:Kingdom of God ``f10!},\footnote{\ Acts 1:2-3\index{44 Acts:ch.   1, v.   2-3, pg. ``f30!}.}
and giving them commission\index{Apostles:commissioned to preach the Gospel ``f10!} to preach
the\index{Preaching:Gospel ``f10!} gospel\index{Gospel:gospel ``f10!} to all nations\index{Gospel:to all nations
``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v. 
19-20, pg. ``f30!}.} forty\index{Jesus:ascension:forty days after resurrection ``f10!} days after his resurrection, he,
in our nature, and as our head\index{Church:Christ as head ``f10!},\footnote{\ Hebrews 6:20\index{58 Hebrews:ch.   6,
v.  20, pg. ``f30!}.} triumphing\index{Jesus:ascension:triumph over enemies ``f10!} over
enemies\index{Enemies:triumphed over by Christ ``f10!},\footnote{\ Ephesians 4:8\index{49 Ephesians:ch.   4, v.   8,
pg. ``f30!}.} visibly went\index{Jesus:resurrection:visible ``f10!} up into the highest heavens, there to receive gifts
for\index{Jesus:ascension:receive gifts for men ``f10!} men,\footnote{\ Acts 1:9-11\index{44 Acts:ch.   1, v.   9-11,
pg. ``f30!}; Ephesians 4:10\index{49 Ephesians:ch.   4, v.  10, pg. ``f30!}; Psalm 68:18\index{19 Psalms:ch.  68, v. 
18, pg. ``f30!}.} to raise up our affections thither\index{Worship:affections directed to God
``f10!},\footnote{\ Colossians 3:1-2\index{51 Colossians:ch.   3, v.   1-2, pg. ``f30!}.} and to prepare
a\index{Jesus:preparing a place in heaven for believers ``f10!} place for us,\footnote{\ John 14:3\index{43 John:ch. 
14, v.   3, pg. ``f30!}.} where himself is, and shall continue till his second coming\index{Jesus:second coming ``f10!}
at the end of the world.\footnote{\ Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!}.}


\bigskip

Q. 54 \textit{How is Christ exalted}\index{Jesus:exaltation ``f10!}\textit{ in his sitting}\index{Jesus:in
heaven:sitting at right hand of Father ``f10!}\textit{ at the right hand of God?}

A. Christ is exalted\index{Jesus:exaltation ``f10!} in his sitting\index{Jesus:in heaven:sitting at right hand of Father
``f10!} at the right hand of God, in that as God-man\index{Jesus:God-man ``f10!} he is advanced to the highest favor
with\index{Jesus:favour with the Father ``f10!} God the Father\index{Father:God the ``f10!},\footnote{\ Philippians
2:9\index{50 Philippians:ch.   2, v.   9, pg. ``f30!}.} with all fulness of joy\index{Joy:everlasting
``f10!},\footnote{\ Acts 2:28\index{44 Acts:ch.   2, v.  28, pg. ``f30!}; Psalm 16:11\index{19 Psalms:ch.  16, v.  11,
pg. ``f30!}.} glory,\footnote{\ John 17:5\index{43 John:ch.  17, v.   5, pg. ``f30!}.} and power over\index{God's
attributes:almighty power ``f10!} all things in heaven\index{Jesus:ruler over heaven and earth ``f10!} and
earth;\footnote{\ Ephesians 1:22\index{49 Ephesians:ch.   1, v.  22, pg. ``f30!}; 1 Peter 3:22\index{60 1 Peter:ch.  
3, v.  22, pg. ``f30!}.} and does gather and defend\index{Jesus:defends his Church ``f10!} his
church\index{Church:Christ's ``f10!}, and subdue the\index{Prayer:for God to subdue the flesh (sin in the world)
``f10!}ir enemies; furnishes his ministers\index{Church officers:minsters ``f10!} and people with
gifts\index{Church:furnished with:gifts and graces ``f10!} and graces,\footnote{\ Ephesians 4:10\index{49 Ephesians:ch.
  4, v.  10, pg. ``f30!}{}-12\index{49 Ephesians:ch.   4, v.  10-12, pg. ``f30!}; Psalm 110:1\index{19 Psalms:ch. 110,
v.   1, pg. ``f30!}.} and makes intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} for
them.\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg.
``f30!}.}


\bigskip


\bigskip


\bigskip

Q. 55 \textit{How does Christ make intercession}\index{Jesus:in heaven:intercession made for all believers
``f10!}\textit{?}

A. Christ makes intercession\index{Jesus:in heaven:intercession made for all believers ``f10!}, by his appearing
in\index{Jesus:incarnation:appearing in our nature ``f10!} our nature continually before the Father\index{Father:God
the ``f10!} in heaven,\footnote{\ Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!}, 24\index{58 Hebrews:ch.  
9, v.  12, 24, pg. ``f30!}.} in the merit of\index{Jesus:obedience:merit of ``f10!} his obedience and sacrifice
on\index{Jesus:sacrifice:of himself ``f10!} earth,\footnote{\ Hebrews 1:3\index{58 Hebrews:ch.   1, v.   3, pg.
``f30!}.} declaring\index{Revelation:special declaration ``f10!} his will to have it applied to all
believers;\footnote{\ John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}; John 17:9\index{43 John:ch.  17, v.   9,
pg. ``f30!}, 20\index{43 John:ch.  17, v.   9, 20, pg. ``f30!}, 24\index{43 John:ch.  17, v.   9, 20, 24, pg. ``f30!}.}
answering all accusations against them,\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}3\index{45
Romans:ch.   8, v.  33, pg. ``f30!}{}-34\index{45 Romans:ch.   8, v.  33-34, pg. ``f30!}.} and procuring for them quiet
of conscience\index{Conscience:quiet ``f10!}, notwithstanding daily failings\index{God's law:broken daily by all men
(thought, word, deed) ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}{}-2\index{45
Romans:ch.   5, v.   1-2, pg. ``f30!}; 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}.} access
with\index{Believers:bold access to God's throne ``f10!} boldness\index{Prayer:with boldness ``f10!} to the throne of
grace\index{Grace:throne of ``f10!},\footnote{\ Hebrews 4:16\index{58 Hebrews:ch.   4, v.  16, pg. ``f30!}.} and
acceptance of\index{Believers:accepted in Christ ``f10!} their persons\footnote{\ Ephesians 1:6\index{49 Ephesians:ch. 
 1, v.   6, pg. ``f30!}.} and services.\footnote{\ 1 Peter 2:5\index{60 1 Peter:ch.   2, v.   5, pg. ``f30!}.}


\bigskip

Q. 56 \textit{How is Christ to be exalted}\index{Jesus:exaltation ``f10!}\textit{ in his coming again to judge
the}\index{Jesus:judge:world ``f10!}\textit{ world?}

A. Christ is to be exalted\index{Jesus:exaltation ``f10!} in his coming again to judge the\index{Jesus:judge:world
``f10!} world, in that he, who was unjustly judged and\index{Jesus:judged and condemned unjustly ``f10!} condemned
by\index{Jesus:condemned by wicked men ``f10!} wicked men,\footnote{\ Acts 3:14-15\index{44 Acts:ch.   3, v.  14-15,
pg. ``f30!}.} shall come again at the last day\index{Last day:last day ``f10!} in great power\index{Jesus:return:in
great power ``f10!},\footnote{\ Matthew 24:30\index{40 Matthew:ch.  24, v.  30, pg. ``f30!}.} and in the full
manifestation of his own glory\index{God's attributes:glory:his own ``f10!}, and of his Father\index{Father:God the
``f10!}{}'s, with all his holy\index{Angels:holy ``f10!} angels\index{Angels:angels ``f10!},\footnote{\ Luke
9:26\index{42 Luke:ch.   9, v.  26, pg. ``f30!}; Matthew 25:31\index{40 Matthew:ch.  25, v.  31, pg. ``f30!}.} with a
shout\index{Jesus:return:with shout and voice of archangel ``f10!}, with the voice of the\index{Angels:voice of, at
judgement day ``f10!} archangel\index{Angels:archangel ``f10!}, and with the trumpet\index{Last day:trumpet ``f10!} of
God,\footnote{\ 1 Thessalonians 4:16\index{52 1 Thessalonians:ch.   4, v.  16, pg. ``f30!}.} to judge the world in
righteousness\index{Judgement day:in righteousness ``f10!}.\footnote{\ Acts 17:31\index{44 Acts:ch.  17, v.  31, pg.
``f30!}.}


\bigskip

Q. 57 \textit{What benefits has}\index{Believers:benefits:of Christ's mediation ``f10!}\textit{ Christ procured by his
mediation}\index{Jesus:mediator:of covenant ``f10!}\textit{?}

A. Christ, by his mediation\index{Jesus:mediator:of covenant ``f10!}, has procured\index{Believers:benefits:of
redemption ``f10!} redemption,\footnote{\ Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!}.} with all other
benefits of the\index{Covenant:grace:benefits ``f10!} covenant of grace\index{Covenant:grace:grace
``f10!}.\footnote{\ 2 Corinthians 1:20\index{47 2 Corinthians:ch.   1, v.  20, pg. ``f30!}.}


\bigskip

Q. 58 \textit{How do we come to be made partakers of the benefits which Christ}\index{Believers:benefits:Christ procured
``f10!}\textit{ has procured}\index{Believers:benefits:of redemption ``f10!}\textit{?}

A. We are made partakers\index{Redemption:partakers through Christ ``f10!} of the benefits which
Christ\index{Believers:benefits:Christ procured ``f10!} has procured\index{Believers:benefits:of redemption
``f10!},\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}1-12\index{43 John:ch.   1, v.  11-12, pg.
``f30!}.} by the application of them unto us, which is the work especially of God the Holy Ghost\index{Holy Spirit:Holy
Ghost ``f10!}.\footnote{\ Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}{}-6\index{56 Titus:ch.   3, v.   5-6,
pg. ``f30!}.}


\bigskip

Q. 59 \textit{Who are made partakers}\index{Redemption:partakers through Christ ``f10!}\textit{ of redemption through
Christ?}

A. Redemption is certainly applied, and effectually communicated\index{Redemption:effectually:communicated ``f10!}, to
all those for whom Christ has purchased\index{Redemption:purchased by Christ ``f10!} it;\footnote{\ Ephesians
1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}3\index{49 Ephesians:ch.   1, v.  13, pg. ``f30!}{}-14\index{49
Ephesians:ch.   1, v.  13-14, pg. ``f30!}; John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}, 39\index{43 John:ch.  
6, v.  37, 39, pg. ``f30!}; John 10:15-16\index{43 John:ch.  10, v.  15-16, pg. ``f30!}.} who are in time by the Holy
Ghost\index{Holy Spirit:Holy Ghost ``f10!} enabled to believe\index{Conversion:Holy Spirit:enables belief ``f10!} in
Christ according to the gospel\index{Gospel:gospel ``f10!}.\footnote{\ Ephesians 2:8\index{49 Ephesians:ch.   2, v.  
8, pg. ``f30!}; 2 Corinthians 4:13\index{47 2 Corinthians:ch.   4, v.  13, pg. ``f30!}.}


\bigskip

Q. 60 \textit{Can they who have never heard the}\index{Salvation:none where Gospel not heard ``f10!}\textit{
gospel}\index{Gospel:gospel ``f10!}\textit{, and so know not Jesus Christ, nor believe in him, be saved by
their}\index{Salvation:not by:living according to light of nature ``f10!}\textit{ living according to the light of
nature?}

A. They who, having never heard the\index{Salvation:none where Gospel not heard ``f10!} gospel\index{Gospel:gospel
``f10!},\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg.
``f30!}.} know not Jesus Christ,\footnote{\ 2 Thessalonians 1:8-9\index{53 2 Thessalonians:ch.   1, v.   8-9, pg.
``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}2\index{49 Ephesians:ch.   2, v.  12, pg.
``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}0-12\index{43 John:ch.   1, v.  10-12, pg. ``f30!}.} and
believe not in him, cannot be saved,\footnote{\ John 8:24\index{43 John:ch.   8, v.  24, pg. ``f30!}; Mark
16:16\index{41 Mark:ch.  16, v.  16, pg. ``f30!}.} be they never so diligent to\index{God's law:diligence in
application ``f10!} frame their\index{Salvation:not by:obedience ``f10!} lives according to the light of
nature,\footnote{\ 1 Corinthians 1:2\index{46 1 Corinthians:ch.   1, v.   2, pg. ``f30!}0-24\index{46 1 Corinthians:ch.
  1, v.  20-24, pg. ``f30!}.} or the laws of that religion which they profess\index{Profession:of false religion
``f10!};\footnote{\ John 4:22\index{43 John:ch.   4, v.  22, pg. ``f30!}; Romans 9:31-32\index{45 Romans:ch.   9, v. 
31-32, pg. ``f30!}; Philippians 3:4-9\index{50 Philippians:ch.   3, v.   4-9, pg. ``f30!}.} neither is there
salvation\index{Salvation:belief in Jesus by faith ``f10!} in any other, but in Christ alone,\footnote{\ Acts
4:12\index{44 Acts:ch.   4, v.  12, pg. ``f30!}.} who is the Savior\index{Jesus:saviour of the Church ``f10!} only of
his body the church\index{Church:Christ's body ``f10!}.\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2,
pg. ``f30!}3\index{49 Ephesians:ch.   5, v.  23, pg. ``f30!}.}


\bigskip

Q. 61 \textit{Are all they saved who}\index{Salvation:not for all who hear the gospel ``f10!}\textit{ hear the
gospel}\index{Gospel:gospel ``f10!}\textit{, and live in the church}\index{Church:home of believers ``f10!}\textit{?}

A. All that hear the gospel\index{Gospel:gospel ``f10!}, and live in the visible church\index{Church:visible ``f10!},
are not saved; but\index{Salvation:not for all who hear the gospel ``f10!} they only who are true members of
the\index{Church members ``f10!} church invisible\index{Church:invisible ``f10!}.\footnote{\ John 12:38-40\index{43
John:ch.  12, v.  38-40, pg. ``f30!}; Romans 9:6\index{45 Romans:ch.   9, v.   6, pg. ``f30!}; Romans 11:7\index{45
Romans:ch.  11, v.   7, pg. ``f30!}; Matthew 7:21\index{40 Matthew:ch.   7, v.  21, pg. ``f30!}; Matthew 22:14\index{40
Matthew:ch.  22, v.  14, pg. ``f30!}.}


\bigskip

Q. 62 \textit{What is the visible church}\index{Church:visible ``f10!}\textit{?}

A. The visible church\index{Church:visible ``f10!} is a society made up of all such as in all ages and places of the
world do profess the true\index{Profession:of true religion ``f10!} religion,\footnote{\ 1 Corinthians 1:2\index{46 1
Corinthians:ch.   1, v.   2, pg. ``f30!}; 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!};
Romans 15:9-12\index{45 Romans:ch.  15, v.   9-12, pg. ``f30!}; Revelation 7:9\index{66 Revelation:ch.   7, v.   9, pg.
``f30!}; Psalm 2:8\index{19 Psalms:ch.   2, v.   8, pg. ``f30!}; Psalm 22:27-31\index{19 Psalms:ch.  22, v.  27-31, pg.
``f30!}; Psalm 45:17\index{19 Psalms:ch.  45, v.  17, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg.
``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg.
``f30!}1\index{23 Isaiah:ch.  59, v.  21, pg. ``f30!}.} and of their children.\footnote{\ 1 Corinthians 7:14\index{46 1
Corinthians:ch.   7, v.  14, pg. ``f30!}; Acts 2:39\index{44 Acts:ch.   2, v.  39, pg. ``f30!}; Romans 11:16\index{45
Romans:ch.  11, v.  16, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}.}


\bigskip

Q. 63 \textit{What are the special privileges}\index{Believers:privileges ``f10!}\textit{ of the visible
church}\index{Church:visible ``f10!}\textit{?}

A. The visible church\index{Church:visible ``f10!} has the privilege\index{Believers:privileges ``f10!} of being under
God's special care and government;\index{Church:Christ's ``f10!}\footnote{\ Isaiah 4:5-6\index{23 Isaiah:ch.   4, v.  
5-6, pg. ``f30!}; 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   4, v.  10,
pg. ``f30!}.} of being protected and\index{Church:protected in all ages by God ``f10!} preserved in
all\index{Church:preserved in all ages by God ``f10!} ages, not withstanding the opposition\index{Church:opposition
from enemies ``f10!} of all enemies\index{Enemies:of the Church ``f10!};\footnote{\ Psalm 115:1-2, 9\index{19
Psalms:ch. 115, v.   1-2, 9, pg. ``f30!}; Isaiah 31:4-5\index{23 Isaiah:ch.  31, v.   4-5, pg. ``f30!}; Zechariah
12:2-4, 8-9\index{38 Zechariah:ch.  12, v.   2-4, 8-9, pg. ``f30!}.} and of enjoying the\index{Communion:enjoying
``f10!} communion of\index{Communion:of saints ``f10!} saints\index{Believers:saints ``f10!}, the ordinary means of
salvation\index{Salvation:belief in Jesus by faith ``f10!},\footnote{\ Acts 2:39\index{44 Acts:ch.   2, v.  39, pg.
``f30!}, 42\index{44 Acts:ch.   2, v.  39, 42, pg. ``f30!}.} and offers of\index{Church:ordinary means of salvation
``f10!} grace by Christ\index{Church:furnished with:gifts and graces ``f10!} to all the members of\index{Church members
``f10!} it in the ministry\index{Gospel:ministry of ``f10!} of the gospel\index{Gospel:gospel ``f10!},
testifying\index{Preachers:testify that all who believe will be saved ``f10!}, that whosoever believes in him shall be
saved,\footnote{\ Psalm 147:19-20\index{19 Psalms:ch. 147, v.  19-20, pg. ``f30!}; Romans 9:4\index{45 Romans:ch.   9,
v.   4, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-12\index{49 Ephesians:ch.   4, v.
 11-12, pg. ``f30!}; Mark 16:15\index{41 Mark:ch.  16, v.  15, pg. ``f30!}{}-16\index{41 Mark:ch.  16, v.  15-16, pg.
``f30!}.} and excluding\index{Salvation:no one excluded who comes to God ``f10!} none that will come unto
him.\footnote{\ John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}.}

Q. 64 \textit{What is the invisible church}\index{Church:invisible ``f10!}\textit{?}

A. The invisible church\index{Church:invisible ``f10!} is the whole number of the elect\index{Election:elect of God
``f10!}, that have been, are, or shall be gathered\index{Church:universal ``f10!} into one under Christ the
head\index{Church:Christ as head ``f10!}.\footnote{\ Ephesians 1:20\index{49 Ephesians:ch.   1, v.  20, pg. ``f30!},
22-23\index{49 Ephesians:ch.   1, v.  20, 22-23, pg. ``f30!}; John 10:16\index{43 John:ch.  10, v.  16, pg. ``f30!};
John 11:52\index{43 John:ch.  11, v.  52, pg. ``f30!}.}


\bigskip

Q. 65 \textit{What special benefits do the}\index{Believers:benefits:as members of the Church ``f10!}\textit{ members of
the}\index{Church members ``f10!}\textit{ invisible church}\index{Church:invisible ``f10!}\textit{ enjoy by Christ?}

A. The members of the\index{Church members ``f10!} invisible church\index{Church:invisible ``f10!} by Christ enjoy union
and\index{Believers:union with Christ:mystical ``f10!} communion with him\index{Communion:with Christ ``f10!} in grace
and glory\index{Resurrection:believers:into glory:with Christ ``f10!}.\footnote{\ John 17:2\index{43 John:ch.  17, v.  
2, pg. ``f30!}1, 24\index{43 John:ch.  17, v.  21, 24, pg. ``f30!}; Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5,
pg. ``f30!}{}-6\index{49 Ephesians:ch.   2, v.   5-6, pg. ``f30!}.}


\bigskip

Q. 66 \textit{What is that union which}\index{Believers:union with Christ:mystical ``f10!}\textit{ the
elect}\index{Election:elect of God ``f10!}\textit{ have with Christ?}

A. The union which\index{Believers:union with Christ:mystical ``f10!} the elect\index{Election:elect of God ``f10!} have
with Christ is the work of God's grace\index{Grace:of salvation ``f10!},\footnote{\ Ephesians 1:22\index{49
Ephesians:ch.   1, v.  22, pg. ``f30!}; Ephesians 2:6-8\index{49 Ephesians:ch.   2, v.   6-8, pg. ``f30!}.} whereby
they are spiritually\index{Believers:union with Christ:spiritually ``f10!} and mystically, yet really and
inseparably\index{Believers:union with Christ:inseparable ``f10!}, joined to Christ as their head\index{Church:Christ
as head ``f10!} and husband;\footnote{\ 1 Corinthians 6:17\index{46 1 Corinthians:ch.   6, v.  17, pg. ``f30!}; John
10:28\index{43 John:ch.  10, v.  28, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg.
``f30!}3\index{49 Ephesians:ch.   5, v.  23, pg. ``f30!}, 30\index{49 Ephesians:ch.   5, v.  23, 30, pg. ``f30!}.}
which is done in their effectual calling\index{Effectual call:calling ``f10!}.\footnote{\ 1 Peter 5:1\index{60 1
Peter:ch.   5, v.   1, pg. ``f30!}0\index{60 1 Peter:ch.   5, v.  10, pg. ``f30!}; 1 Corinthians 1:9\index{46 1
Corinthians:ch.   1, v.   9, pg. ``f30!}.}


\bigskip

Q. 67 \textit{What is effectual calling}\index{Effectual call:calling ``f10!}\textit{?}

A. Effectual\index{Effectual call:calling ``f10!} calling is the work of God's almighty\index{God's attributes:almighty
power ``f10!} power and grace,\footnote{\ John 5:25\index{43 John:ch.   5, v.  25, pg. ``f30!}; Ephesians 1:1\index{49
Ephesians:ch.   1, v.   1, pg. ``f30!}8-20\index{49 Ephesians:ch.   1, v.  18-20, pg. ``f30!}; 2 Timothy 1:8-9\index{55
2 Timothy:ch.   1, v.   8-9, pg. ``f30!}.} whereby (out\index{Grace:God's free ``f10!} of his free and special love to
his\index{God's attributes:love ``f10!} elect\index{Election:elect of God ``f10!}, and from nothing in them moving him
thereunto)\footnote{\ Titus. 3:4-5; Ephesians 2:4-5, 7-9\index{49 Ephesians:ch.   2, v.   4-5, 7-9, pg. ``f30!}; Romans
9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}.} he does, in his accepted time\index{Salvation:at appointed time
``f10!}, invite\index{Effectual call:invitation to draw near to Christ ``f10!} and draw them\index{Effectual
call:drawing to Christ ``f10!} to Jesus Christ, by his word and Spirit\index{Revelation:Word and Spirit
``f10!};\footnote{\ 2 Corinthians 5:20\index{47 2 Corinthians:ch.   5, v.  20, pg. ``f30!}; 2 Corinthians
6:1-2\index{47 2 Corinthians:ch.   6, v.   1-2, pg. ``f30!}; John 6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}; 2
Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}{}-14\index{53 2 Thessalonians:ch.   2, v. 
13-14, pg. ``f30!}.} savingly enlightening\index{Effectual call:enlightening minds spiritually ``f10!} their
minds,\footnote{\ Acts 26:18\index{44 Acts:ch.  26, v.  18, pg. ``f30!}; 1 Corinthians 2:10, 12\index{46 1
Corinthians:ch.   2, v.  10, 12, pg. ``f30!}.} renewing and\index{Conversion:renewed by Holy Spirit ``f10!} powerfully
determining their\index{Effectual call:determining wills ``f10!} wills,\footnote{\ Ezekiel 11:19\index{26 Ezekiel:ch. 
11, v.  19, pg. ``f30!}; Ezekiel 36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36, v. 
26-27, pg. ``f30!}; John 6:45\index{43 John:ch.   6, v.  45, pg. ``f30!}.} so as they (although in themselves
dead\index{Death:spiritual, in sin ``f10!} in sin) are hereby made willing and able freely\index{Conversion:renewal of
the will ``f10!} to answer\index{Effectual call:enables person to:freely answer call ``f10!} his call, and to accept
and embrace the grace\index{Grace:in believers:by Holy Spirit ``f10!} offered and\index{Effectual call:enables person
to:embrace grace ``f10!} conveyed therein.\footnote{\ Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5, pg. ``f30!};
Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Deuteronomy 30:6\index{05 Deuteronomy:ch.  30, v.  
6, pg. ``f30!}.}


\bigskip

Q. 68 \textit{Are the elect}\index{Election:elect of God ``f10!}\textit{ only effectually called}\index{Effectual
call:calling ``f10!}\textit{?}

A. All the elect\index{Election:elect of God ``f10!}, and they only, are effectually called\index{Effectual call:calling
``f10!};\footnote{\ Acts 13:48\index{44 Acts:ch.  13, v.  48, pg. ``f30!}.} although others may be, and often are,
outwardly called\index{Preaching:outward call ``f10!} by the ministry of the word\index{Preaching:ministry of the Word
``f10!},\footnote{\ Matthew 22:14\index{40 Matthew:ch.  22, v.  14, pg. ``f30!}.} and have some common
operations\index{Holy Spirit:operations:common ``f10!} of the Spirit;\footnote{\ Matthew 7:22\index{40 Matthew:ch.   7,
v.  22, pg. ``f30!}; Matthew 13:20-21\index{40 Matthew:ch.  13, v.  20-21, pg. ``f30!}; Hebrews 6:4-6\index{58
Hebrews:ch.   6, v.   4-6, pg. ``f30!}.} who, for their wilful neglect\index{Unbelievers:wilful neglect and contempt of
grace ``f10!} and contempt of the\index{Contempt:for grace of Gospel ``f10!} grace offered to\index{Grace:opposed
``f10!} them, being justly left\index{Punishment:just:contempt of grace left in unbelief ``f10!} in their
unbelief\index{unbelief (misbelief) ``f10!}, do never truly\index{Unbelievers:non-elect never truly come to Jesus
``f10!} come to Jesus Christ.\footnote{\ John 6:64-65\index{43 John:ch.   6, v.  64-65, pg. ``f30!}; John
12:38-40\index{43 John:ch.  12, v.  38-40, pg. ``f30!}; Acts 18:25\index{44 Acts:ch.  18, v.  25, pg.
``f30!}{}-27\index{44 Acts:ch.  18, v.  25-27, pg. ``f30!}; Psalm 81:11\index{19 Psalms:ch.  81, v.  11, pg.
``f30!}{}-12\index{19 Psalms:ch.  81, v.  11-12, pg. ``f30!}.}

Q. 69 \textit{What is the communion in grace}\index{Communion:in grace ``f10!}\textit{ which the members of
the}\index{Church members ``f10!}\textit{ invisible church}\index{Church:invisible ``f10!}\textit{ have with Christ?}

A. The communion in grace\index{Communion:in grace ``f10!} which the members of the\index{Church members ``f10!}
invisible church\index{Church:invisible ``f10!} have with Christ, is their partaking of the
virtue\index{Jesus:vicarious atonement ``f10!} of his mediation\index{Jesus:mediator:of covenant ``f10!}, in their
justification\index{Justification:justification ``f10!},\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg.
``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}.} adoption\index{Adoption:in salvation
``f10!},\footnote{\ Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}.}
sanctification\index{Sanctification:sanctification ``f10!}, and whatever else, in this life, manifests their union
with\index{Believers:union with Christ:mystical ``f10!} him.\footnote{\ 1 Corinthians 1:30\index{46 1 Corinthians:ch.  
1, v.  30, pg. ``f30!}.}


\bigskip

Q. 70 \textit{What is justification}\index{Justification:justification ``f10!}\textit{?}

A. Justification\index{Justification:justification ``f10!} is an act of God's free grace\index{Grace:God's free ``f10!}
unto sinners,\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}2, 24-25\index{45 Romans:ch.   3, v. 
22, 24-25, pg. ``f30!}; Romans 4:5\index{45 Romans:ch.   4, v.   5, pg. ``f30!}.} in which he
pardons\index{Justification:pardoned of sin ``f10!} all their sins, accepts and\index{Justification:accepted by God as
righteous ``f10!} accounts\index{Justification:accounted by God as righteous ``f10!} their persons righteous
in\index{Jesus:righteousness imputed to believers ``f10!} his sight;\footnote{\ 2 Corinthians 5:19\index{47 2
Corinthians:ch.   5, v.  19, pg. ``f30!}, 21\index{47 2 Corinthians:ch.   5, v.  19, 21, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}2-25, 27-28\index{45 Romans:ch.   3, v.  22-25, 27-28, pg. ``f30!}.}
not for any thing wrought in them\index{Salvation:not by:works ``f10!}, or done by them,\footnote{\ Titus 3:5\index{56
Titus:ch.   3, v.   5, pg. ``f30!}, 7\index{56 Titus:ch.   3, v.   5, 7, pg. ``f30!}; Ephesians 1:7\index{49
Ephesians:ch.   1, v.   7, pg. ``f30!}.} but only for the perfect obedience\index{Jesus:obedience:to God's law ``f10!}
and full satisfaction of Christ\index{Jesus:satisfied divine justice ``f10!}, by God imputed to\index{Imputation:of
Christ's righteousness by faith ``f10!} them,\footnote{\ Romans 4:6-8\index{45 Romans:ch.   4, v.   6-8, pg. ``f30!};
Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}7-19\index{45 Romans:ch.   5, v.  17-19, pg. ``f30!}.} and
received by faith\index{Faith:instrument:of justification ``f10!} alone.\footnote{\ Acts 10:43\index{44 Acts:ch.  10,
v.  43, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Philippians 3:9\index{50
Philippians:ch.   3, v.   9, pg. ``f30!}.}


\bigskip

Q. 71 \textit{How is justification}\index{Justification:justification ``f10!}\textit{ an act of God's free
grace}\index{Grace:God's free ``f10!}\textit{?}

A. Although Christ, by his obedience and death, did\index{Jesus:death:death ``f10!} make a proper, real, and full
satisfaction to God's\index{Jesus:satisfied divine justice ``f10!} justice in the behalf of them that are
justified\index{Justification:justified ``f10!};\footnote{\ Romans 5:8-10, 19\index{45 Romans:ch.   5, v.   8-10, 19,
pg. ``f30!}.} yet inasmuch as God accepts the\index{Believers:accepted in Christ ``f10!} satisfaction from a
surety\index{Jesus:surety for the elect ``f10!}, which he might have demanded of them, and did provide this surety, his
own only Son\index{Jesus:Son of God ``f10!},\footnote{\ 2 Timothy 2:5-6\index{55 2 Timothy:ch.   2, v.   5-6, pg.
``f30!}; Hebrews 7:22\index{58 Hebrews:ch.   7, v.  22, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg.
``f30!}0\index{58 Hebrews:ch.  10, v.  10, pg. ``f30!}; Matthew 20:28\index{40 Matthew:ch.  20, v.  28, pg. ``f30!};
Daniel 9:24, 26\index{27 Daniel:ch.   9, v.  24, 26, pg. ``f30!}; Isaiah 53:4-6, 10-12\index{23 Isaiah:ch.  53, v.  
4-6, 10-12, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}2\index{45 Romans:ch.   8, v.  32, pg.
``f30!}; 1 Peter 1:18\index{60 1 Peter:ch.   1, v.  18, pg. ``f30!}{}-19\index{60 1 Peter:ch.   1, v.  18-19, pg.
``f30!}.} imputing\index{Imputation:of Christ's righteousness by faith ``f10!} his
righteousness\index{Jesus:righteousness imputed to believers ``f10!} to them,\footnote{\ 2 Corinthians 5:21\index{47 2
Corinthians:ch.   5, v.  21, pg. ``f30!}.} and requiring\index{Salvation:faith alone required ``f10!} nothing of them
for their justification\index{Justification:justification ``f10!} but faith\index{Faith:instrument:of justification
``f10!},\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg.
``f30!}{}-25\index{45 Romans:ch.   3, v.  24-25, pg. ``f30!}.} which also is his gift,\index{Justification:free gift
``f10!}\footnote{\ Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}.} their justification is to them of
free grace\index{Grace:God's free ``f10!}.\footnote{\ Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}.}


\bigskip

Q. 72 \textit{What is justifying}\index{Justification:by faith ``f10!}\textit{ faith?}

A. Justifying\index{Justification:by faith ``f10!} faith is a saving grace\index{Grace:of salvation
``f10!},\footnote{\ Hebrews 10:39\index{58 Hebrews:ch.  10, v.  39, pg. ``f30!}.} wrought in the heart of
a\index{Heart:new:grace applied ``f10!} sinner by the Spirit\index{Revelation:Word and Spirit ``f10!}\footnote{\ 2
Corinthians 4:13\index{47 2 Corinthians:ch.   4, v.  13, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1,
pg. ``f30!}7-19\index{49 Ephesians:ch.   1, v.  17-19, pg. ``f30!}.} and word of God,\footnote{\ Romans 10:1\index{45
Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg. ``f30!}, 17\index{45 Romans:ch.  10, v.  14,
17, pg. ``f30!}.} whereby he, being convinced\index{Repentance:conviction of sin, misery, and disability ``f10!} of his
sin and misery\index{Punishment:misery ``f10!}, and of the disability in\index{Sin:original:utterly unable and
unwilling to do will of God ``f10!} himself and all other creatures\index{creatures ``f10!} to recover
him\index{Justification:recovery from lost condition ``f10!} out of his lost condition\index{Mankind:lost condition
``f10!},\footnote{\ Acts 2:37\index{44 Acts:ch.   2, v.  37, pg. ``f30!}; Acts 4:12\index{44 Acts:ch.   4, v.  12, pg.
``f30!}; Acts 16:30\index{44 Acts:ch.  16, v.  30, pg. ``f30!}; John 16:8-9\index{43 John:ch.  16, v.   8-9, pg.
``f30!}; Romans 5:6\index{45 Romans:ch.   5, v.   6, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg.
``f30!}.} not only assents\index{Justification:faith assents to truth of Gospel ``f10!} to the truth of\index{God's
attributes:truth ``f10!} the promise of the\index{Gospel:promise ``f10!} gospel\index{Gospel:gospel
``f10!},\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}3\index{49 Ephesians:ch.   1, v.  13,
pg. ``f30!}.} but receives\index{Faith:instrument:for receiving Christ ``f10!} and rests\index{Faith:resting on Christ
``f10!} upon Christ and his righteousness\index{Jesus:righteousness imputed to believers ``f10!}, therein held forth,
for pardon of sin\index{Justification:pardoned of sin ``f10!},\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}; Acts 10:43\index{44 Acts:ch.  10, v.  43, pg. ``f30!}; Acts
16:31\index{44 Acts:ch.  16, v.  31, pg. ``f30!}.} and for the accepting and\index{Justification:accepted by God as
righteous ``f10!} accounting of\index{Justification:accounted by God as righteous ``f10!} his person righteous in the
sight of God for salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Philippians 3:9\index{50
Philippians:ch.   3, v.   9, pg. ``f30!}; Acts 15:11\index{44 Acts:ch.  15, v.  11, pg. ``f30!}.}


\bigskip

Q. 73 \textit{How does faith}\index{Faith:instrument:of justification ``f10!}\textit{ justify a}\index{Justification:by
faith ``f10!}\textit{ sinner in the sight of God?}

A. Faith\index{Faith:saving ``f10!} justifies a\index{Justification:by faith ``f10!} sinner in the sight of God, not
because of those other graces\index{Faith:accompanied by graces ``f10!} which do always accompany it, or of good
works\index{Good works:adorn profession of faith ``f10!} that are the fruits\index{Good works:fruit of salvation
``f10!} of it,\footnote{\ Galatians 3:11\index{48 Galatians:ch.   3, v.  11, pg. ``f30!}; Romans 3:2\index{45
Romans:ch.   3, v.   2, pg. ``f30!}8\index{45 Romans:ch.   3, v.  28, pg. ``f30!}.} nor as if the grace of
faith\index{Faith:given by grace ``f10!}, or any act thereof, were imputed to\index{Imputation:of Christ's
righteousness by faith ``f10!} him for his justification\index{Justification:justification ``f10!};\footnote{\ Romans
4:5\index{45 Romans:ch.   4, v.   5, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}0\index{45
Romans:ch.  10, v.  10, pg. ``f30!}.} but only as it is an instrument\index{Faith:instrument:of justification ``f10!}
by which he receives\index{Faith:instrument:for receiving Christ ``f10!} and applies\index{Faith:instrument:which
applies Christ to believers ``f10!} Christ and his righteousness\index{Jesus:righteousness imputed to believers
``f10!}.\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!};
Philippians 3:9\index{50 Philippians:ch.   3, v.   9, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16,
pg. ``f30!}.}


\bigskip

Q. 74 \textit{What is adoption}\index{Adoption:in salvation ``f10!}\textit{?}

A. Adoption\index{Adoption:in salvation ``f10!} is an act of the free grace\index{Grace:God's free ``f10!} of
God,\footnote{\ 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}.} in and for his only Son\index{Jesus:Son of
God ``f10!} Jesus Christ,\footnote{\ Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}; Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}.} whereby all
those that are justified\index{Justification:justified ``f10!} are received into the number of his
children,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}.}
have his name put upon them,\footnote{\ 2 Corinthians 4:18\index{47 2 Corinthians:ch.   4, v.  18, pg. ``f30!};
Revelation 3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}2\index{66 Revelation:ch.   3, v.  12, pg. ``f30!}.} the
Spirit of his\index{Holy Spirit:Spirit of Christ ``f10!} Son given to them,\footnote{\ Galatians 4:6\index{48
Galatians:ch.   4, v.   6, pg. ``f30!}.} are under his fatherly\index{Father:God the ``f10!} care and
dispensations\index{Adoption:fatherly dispensations ``f10!},\footnote{\ Psalm 103:1\index{19 Psalms:ch. 103, v.   1,
pg. ``f30!}3\index{19 Psalms:ch. 103, v.  13, pg. ``f30!}; Proverbs 14:26\index{20 Proverbs:ch.  14, v.  26, pg.
``f30!}; Matthew 6:32\index{40 Matthew:ch.   6, v.  32, pg. ``f30!}.} admitted to all the liberties\index{Christian
liberty:privileges of children of God ``f10!} and privileges\index{Believers:privileges ``f10!} of the sons of God,
made heirs\index{Believers:heirs (salvation, promises, glory) ``f10!} of all the promises, and fellow heirs with Christ
in glory\index{Resurrection:believers:into glory:with Christ ``f10!}.\footnote{\ Hebrews 6:12\index{58 Hebrews:ch.   6,
v.  12, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}7\index{45 Romans:ch.   8, v.  17, pg.
``f30!}.}


\bigskip

Q. 75 \textit{What is sanctification}\index{Sanctification:sanctification ``f10!}\textit{?}

A. Sanctification\index{Sanctification:sanctification ``f10!} is a work of God's grace\index{Grace:of salvation ``f10!},
whereby they whom God has, before the foundation\index{Predestination:before foundation of world. ``f10!} of the world,
chosen\index{Election:chosen by God ``f10!} to be holy\index{Believers:holy:chosen to be ``f10!}, are in time, through
the powerful operation\index{Holy Spirit:operations:saving ``f10!} of his Spirit\footnote{\ Ephesians 1:4\index{49
Ephesians:ch.   1, v.   4, pg. ``f30!}; 1 Corinthians 6:11\index{46 1 Corinthians:ch.   6, v.  11, pg. ``f30!}; 2
Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.} applying the death\index{Jesus:death:death
``f10!} and resurrection of Christ\index{Jesus:resurrection:from the dead ``f10!} unto them,\footnote{\ Romans
6:4\index{45 Romans:ch.   6, v.   4, pg. ``f30!}{}-6\index{45 Romans:ch.   6, v.   4-6, pg. ``f30!}.} renewed
in\index{Sanctification:renewed in image of God ``f10!} their whole man after the image\index{Mankind:image of God
``f10!} of God;\footnote{\ Ephesians 4:23-24\index{49 Ephesians:ch.   4, v.  23-24, pg. ``f30!}.} having the seeds of
repentance unto\index{Repentance:unto life ``f10!} life, and all other saving graces, put into their
hearts\index{Heart:new:grace applied ``f10!},\footnote{\ Acts 11:18\index{44 Acts:ch.  11, v.  18, pg. ``f30!}; 1 John
3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}.} and those graces\index{Grace:in believers:by Holy Spirit ``f10!} so
stirred up, increased, and strengthened,\index{Sanctification:supply of strength from Holy Spirit
``f10!}\footnote{\ Jude 1:20\index{65 Jude:v.   1, v.  20, pg. ``f30!}; Hebrews 6:11-12\index{58 Hebrews:ch.   6, v. 
11-12, pg. ``f30!}; Ephesians 3:16\index{49 Ephesians:ch.   3, v.  16, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v. 
16-19, pg. ``f30!}; Colossians 1:10-11\index{51 Colossians:ch.   1, v.  10-11, pg. ``f30!}.} as that they more and more
die unto\index{Sanctification:die unto sin ``f10!} sin, and rise unto\index{Sanctification:spiritual growth:in newness
of life ``f10!} newness of life.\footnote{\ Romans 6:4\index{45 Romans:ch.   6, v.   4, pg. ``f30!}; Romans
6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}; Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}.}

Q. 76 \textit{What is repentance unto}\index{Repentance:unto life ``f10!}\textit{ life?}

A. Repentance\index{Repentance:unto life ``f10!} unto life is a saving grace\index{Grace:of salvation
``f10!},\footnote{\ 2 Timothy 2:2\index{55 2 Timothy:ch.   2, v.   2, pg. ``f30!}5\index{55 2 Timothy:ch.   2, v.  25,
pg. ``f30!}.} wrought in the heart of a\index{Heart:new:grace applied ``f10!} sinner by the
Spirit\index{Revelation:Word and Spirit ``f10!}\footnote{\ Zechariah 12:10\index{38 Zechariah:ch.  12, v.  10, pg.
``f30!}.} and word of God,\footnote{\ Acts 11:18\index{44 Acts:ch.  11, v.  18, pg. ``f30!}, 20-21\index{44 Acts:ch. 
11, v.  18, 20-21, pg. ``f30!}.} whereby, out of the sight and sense, not only of the
danger\index{Repentance:apprehension of danger of sin ``f10!},\footnote{\ Ezekiel 18:28, 30, 32\index{26 Ezekiel:ch. 
18, v.  28, 30, 32, pg. ``f30!}; Luke 15:17-18\index{42 Luke:ch.  15, v.  17-18, pg. ``f30!}; Hosea 2:6-7\index{28
Hosea:ch.   2, v.   6-7, pg. ``f30!}.} but also of the filthiness\index{Confession:of sin ``f10!} and
odiousness\index{Sin:odious before God ``f10!} of his sins,\footnote{\ Ezekiel 36:31\index{26 Ezekiel:ch.  36, v.  31,
pg. ``f30!}; Isaiah 30:22\index{23 Isaiah:ch.  30, v.  22, pg. ``f30!}.} and upon the
apprehension\index{Repentance:apprehension of God's mercy ``f10!} of God's mercy in\index{God's attributes:merciful
``f10!} Christ to such as are penitent\index{Repentance:penitence ``f10!},\footnote{\ Joel 2:12\index{29 Joel:ch.   2,
v.  12, pg. ``f30!}{}-13\index{29 Joel:ch.   2, v.  12-13, pg. ``f30!}.} he so grieves for\index{Sin:grief and hatred
of ``f10!}\footnote{\ Jeremiah 31:18\index{24 Jeremiah:ch.  31, v.  18, pg. ``f30!}{}-19\index{24 Jeremiah:ch.  31, v. 
18-19, pg. ``f30!}.} and hates\index{Sin:to be hated ``f10!} his sins,\footnote{\ 2 Corinthians 7:1\index{47 2
Corinthians:ch.   7, v.   1, pg. ``f30!}1\index{47 2 Corinthians:ch.   7, v.  11, pg. ``f30!}.} as that he
turns\index{Repentance:forsaking sin ``f10!} from them all to God,\footnote{\ Acts 26:18\index{44 Acts:ch.  26, v.  18,
pg. ``f30!}; Ezekiel 14:6\index{26 Ezekiel:ch.  14, v.   6, pg. ``f30!}; 1 Kings 8:47-48\index{11 1 Kings:ch.   8, v. 
47-48, pg. ``f30!}.} purposing\index{Repentance:purposing to walk with God ``f10!} and endeavoring\index{God's
law:commandments:obedience:endeavoring ``f10!} constantly\index{Believers:desire to obey God ``f10!} to walk with
him\index{Believers:obedience:strive for new ``f10!} in all the ways of new obedience.\footnote{\ Psalm 119:6, 59,
128\index{19 Psalms:ch. 119, v.   6, 59, 128, pg. ``f30!}; Luke 1:6\index{42 Luke:ch.   1, v.   6, pg. ``f30!}; 2 Kings
23:25\index{12 2 Kings:ch.  23, v.  25, pg. ``f30!}.} 


\bigskip

Q. 77 \textit{Wherein do justification}\index{Justification:justification ``f10!}\textit{ and
sanctification}\index{Sanctification:sanctification ``f10!}\textit{ differ?}

A. Although sanctification\index{Sanctification:sanctification ``f10!} be inseparably\index{Sanctification:inseparable
from:justification ``f10!} joined with justification\index{Justification:justification ``f10!},\footnote{\ 1
Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg. ``f30!}; 1 Corinthians 6:11\index{46 1 Corinthians:ch.  
6, v.  11, pg. ``f30!}.} yet they differ, in that God in justification imputes\index{Imputation:of Christ's
righteousness by faith ``f10!} the righteousness of\index{Jesus:righteous ``f10!} Christ;\footnote{\ Romans 4:6,
8\index{45 Romans:ch.   4, v.   6, 8, pg. ``f30!}.} in sanctification his Spirit infuses grace\index{Grace:in
believers:by Holy Spirit ``f10!}, and enables\index{Sanctification:enabled by:grace ``f10!} to the exercise
thereof;\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}.} in the former, sin is
pardoned;\index{Sanctification:pardoned of sin ``f10!}\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg.
``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}{}-25\index{45 Romans:ch.   3, v.  24-25, pg. ``f30!}.} in the
other, it is subdued\index{Sanctification:sin subdued ``f10!}:\footnote{\ Romans 6:6, 14\index{45 Romans:ch.   6, v.  
6, 14, pg. ``f30!}.} the one does equally free all believers\index{Justification:frees all believers from wrath of God
``f10!} from the revenging\index{Revenge:God's ``f10!} wrath of God\index{Wrath:of God:for sin ``f10!}, and that
perfectly in this life, that they never fall\index{Assurance:never fall from being saved ``f10!} into
condemnation\index{Believers:never fall into condemnation ``f10!};\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.  
3, pg. ``f30!}3\index{45 Romans:ch.   8, v.  33, pg. ``f30!}{}-34\index{45 Romans:ch.   8, v.  33-34, pg. ``f30!}.} the
other is neither equal in all\index{Sanctification:not equal in all ``f10!},\footnote{\ 1 John 2:12-14\index{62 1
John:ch.   2, v.  12-14, pg. ``f30!}; Hebrews 5:12-14\index{58 Hebrews:ch.   5, v.  12-14, pg. ``f30!}.} nor in this
life perfect in any,\footnote{\ 1 John 1:8, 10\index{62 1 John:ch.   1, v.   8, 10, pg. ``f30!}.} but growing
up\index{Sanctification:spiritual growth:to perfection ``f10!} to perfection\index{Sanctification:growing to perfection
``f10!}.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; Philippians 3:12\index{50
Philippians:ch.   3, v.  12, pg. ``f30!}{}-14\index{50 Philippians:ch.   3, v.  12-14, pg. ``f30!}.}


\bigskip

Q. 78 \textit{Whence arises the imperfection of}\index{Sanctification:imperfection in believers ``f10!}\textit{
sanctification}\index{Sanctification:sanctification ``f10!}\textit{ in believers?}

A. The imperfection of\index{Sanctification:imperfection in believers ``f10!}
sanctification\index{Sanctification:sanctification ``f10!} in believers arises from the
remnants\index{Sanctification:remnants of corruption remain ``f10!} of sin abiding in every\index{Sin:remnants in
believers ``f10!} part of them, and the perpetual lustings\index{Sexual immorality:lust ``f10!} of the flesh against
the spirit\index{Sanctification:ongoing lusting of flesh against the spirit ``f10!}; whereby they are often foiled with
temptations\index{Temptation:temptation ``f10!}, and fall into many sins,\footnote{\ Romans 7:18\index{45 Romans:ch.  
7, v.  18, pg. ``f30!}, 23\index{45 Romans:ch.   7, v.  18, 23, pg. ``f30!}; Mark 14:66-72\index{41 Mark:ch.  14, v. 
66-72, pg. ``f30!}; Galatians 2:11-12\index{48 Galatians:ch.   2, v.  11-12, pg. ``f30!}.} are hindered in all their
spiritual services\index{Good works:spiritual services ``f10!},\footnote{\ Hebrews 12:1\index{58 Hebrews:ch.  12, v.  
1, pg. ``f30!}.} and their best works are imperfect and\index{Good works:mixed with weakness and imperfection ``f10!}
defiled in the sight of God.\footnote{\ Isaiah 64:6\index{23 Isaiah:ch.  64, v.   6, pg. ``f30!}; Exodus 28:38\index{02
Exodus:ch.  28, v.  38, pg. ``f30!}.}

Q. 79 \textit{May not true believers}\index{Believers:cannot fall from state of grace ``f10!}\textit{, by reason of
their imperfections}\index{Sanctification:imperfection in believers ``f10!}\textit{, and the many
temptations}\index{Temptation:temptation ``f10!}\textit{ and sins they are overtaken}\index{Perseverance:not overtaken
by temptation and sin ``f10!}\textit{ with, fall away from the state of grace}\index{Grace:state:fall from
``f10!}\textit{?}

A. True believers, by\index{Assurance:of salvation ``f10!} reason of the unchangeable\index{God's
attributes:unchangeable ``f10!} love of God\index{Love toward God:for God ``f10!},\footnote{\ Jeremiah 31:3\index{24
Jeremiah:ch.  31, v.   3, pg. ``f30!}.} and his decree\index{Predestination:eternal decrees ``f10!} and covenant
to\index{Covenant:promises ``f10!} give them perseverance\index{Perseverance:of the saints ``f10!},\footnote{\ 2
Timothy 2:19\index{55 2 Timothy:ch.   2, v.  19, pg. ``f30!}{}-21\index{55 2 Timothy:ch.   2, v.  19-21, pg. ``f30!}; 2
Samuel 23:5\index{10 2 Samuel:ch.  23, v.   5, pg. ``f30!}.} their inseparable\index{Believers:union with
Christ:inseparable ``f10!} union with\index{Believers:union with Christ:mystical ``f10!} Christ,\footnote{\ 1
Corinthians 1:8-9\index{46 1 Corinthians:ch.   1, v.   8-9, pg. ``f30!}.} his continual intercession\index{Jesus:in
heaven:intercession made for all believers ``f10!} for them,\footnote{\ Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25,
pg. ``f30!}; Luke 22:32\index{42 Luke:ch.  22, v.  32, pg. ``f30!}.} and the Spirit and seed of
God\index{Believers:seed of God planted within ``f10!} abiding in them\index{Holy Spirit:abiding in believers
``f10!},\footnote{\ 1 John 2:2\index{62 1 John:ch.   2, v.   2, pg. ``f30!}7\index{62 1 John:ch.   2, v.  27, pg.
``f30!}; 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}.} can neither totally\index{Assurance:never fall from
being saved ``f10!} nor finally fall away from the state of grace\index{Grace:state:fall from
``f10!},\footnote{\ Jeremiah 32:40\index{24 Jeremiah:ch.  32, v.  40, pg. ``f30!}; John 10:28\index{43 John:ch.  10, v.
 28, pg. ``f30!}.} but are kept by the power\index{Perseverance:kept by God's power ``f10!} of God through faith
unto\index{Faith:saving ``f10!} salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Peter
1:5\index{60 1 Peter:ch.   1, v.   5, pg. ``f30!}.}


\bigskip

Q. 80 \textit{Can true believers}\index{Assurance:believer may not have at all times ``f10!}\textit{ be infallibly
assured that}\index{Assurance:of state of grace ``f10!}\textit{ they are in the estate of grace}\index{Grace:estate of
``f10!}\textit{, and that they shall persevere}\index{Perseverance:of the saints ``f10!}\textit{ therein unto
salvation}\index{Salvation:belief in Jesus by faith ``f10!}\textit{?}

A. Such as truly\index{Salvation:belief in Jesus by faith ``f10!} believe in Christ, and endeavor to walk in
all\index{Believers:strive to walk in good conscience ``f10!} good conscience\index{Conscience:good, endeavour for
``f10!} before him,\footnote{\ 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}.} may, without extraordinary
revelation\index{Revelation:extraordinary ``f10!}, by faith grounded\index{Faith:grounded on God's promises ``f10!}
upon the truth\index{God's attributes:truth ``f10!} of God's promises\index{Assurance:promises of God ``f10!}, and by
the Spirit enabling the\index{God's law:commandments:obedience:enabled by Holy Spirit ``f10!}m\index{Assurance:enabled
by Holy Spirit ``f10!} to discern in\index{Grace:discern in self ``f10!} themselves those graces\index{Grace:in
believers:by Holy Spirit ``f10!} to which the promises\index{Salvation:promises made ``f10!} of life are
made,\footnote{\ 1 Corinthians 2:12\index{46 1 Corinthians:ch.   2, v.  12, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.
  3, v.   1, pg. ``f30!}4\index{62 1 John:ch.   3, v.  14, pg. ``f30!}, 18-19, 21, 24\index{62 1 John:ch.   3, v.  14,
18-19, 21, 24, pg. ``f30!}; 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg. ``f30!}3\index{62 1 John:ch.   4, v.  13,
pg. ``f30!}, 16\index{62 1 John:ch.   4, v.  13, 16, pg. ``f30!}; Hebrews 6:11-12\index{58 Hebrews:ch.   6, v.  11-12,
pg. ``f30!}.} and bearing witness with\index{Assurance:Holy Spirit witnesses with our spirits ``f10!} their
spirits\index{Conversion:Holy Spirit:bears witness to our spirits ``f10!} that they are the children of
God,\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}6\index{45 Romans:ch.   8, v.  16, pg. ``f30!}.}
be infallibly assured that\index{Assurance:of state of grace ``f10!} they are in the estate of grace\index{Grace:estate
of ``f10!}, and shall persevere\index{Perseverance:of the saints ``f10!} therein unto salvation.\footnote{\ 1 John
5:13\index{62 1 John:ch.   5, v.  13, pg. ``f30!}.}


\bigskip

Q. 81 \textit{Are all true believers}\index{Assurance:believer may not have at all times ``f10!}\textit{ at all times
assured of}\index{Assurance:of state of grace ``f10!}\textit{ their present being in the estate of
grace}\index{Grace:estate of ``f10!}\textit{, and that they shall be saved}\index{Salvation:belief in Jesus by faith
``f10!}\textit{?}

A. Assurance\index{Assurance:of state of grace ``f10!} of grace and salvation\index{Salvation:belief in Jesus by faith
``f10!} not being of the essence of faith\index{Faith:essence of ``f10!},\footnote{\ Ephesians 1:1\index{49
Ephesians:ch.   1, v.   1, pg. ``f30!}3\index{49 Ephesians:ch.   1, v.  13, pg. ``f30!}.} true believers may wait
long\index{Assurance:believer may await ``f10!} before they obtain it;\footnote{\ Isaiah 50:10\index{23 Isaiah:ch.  50,
v.  10, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}.} and, after the
enjoyment\index{Assurance:enjoyment of ``f10!} thereof, may have it weakened\index{Assurance:weak ``f10!} and
intermitted\index{Assurance:believer may not have at all times ``f10!}, through manifold distempers, sins,
temptations\index{Temptation:temptation ``f10!}, and desertions;\footnote{\ Psalm 22:1\index{19 Psalms:ch.  22, v.   1,
pg. ``f30!}; Psalm 31:22\index{19 Psalms:ch.  31, v.  22, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg.
``f30!}, 12\index{19 Psalms:ch.  51, v.   8, 12, pg. ``f30!}; Psalm 77:1-12\index{19 Psalms:ch.  77, v.   1-12, pg.
``f30!}; Song of Solomon 5:2-3, 6\index{22 Song of Solomon:ch.   5, v.   2-3, 6, pg. ``f30!}.} yet are they never left
without such a presence and support of the Spirit of God\index{Holy Spirit:God's Spirit ``f10!} as keeps them from
sinking into\index{Assurance:Holy Spirit keeps believer from utter despair ``f10!} utter despair\index{despair
``f10!}.\footnote{\ 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}; Job 13:15\index{18 Job:ch.  13, v.  15,
pg. ``f30!}; Psalm 73:15\index{19 Psalms:ch.  73, v.  15, pg. ``f30!}, 23\index{19 Psalms:ch.  73, v.  15, 23, pg.
``f30!}; Isaiah 54:7-10\index{23 Isaiah:ch.  54, v.   7-10, pg. ``f30!}.}


\bigskip

Q. 82 \textit{What is the communion in glory}\index{Communion:in glory ``f10!}\textit{ which the members of
the}\index{Church members ``f10!}\textit{ invisible church}\index{Church:invisible ``f10!}\textit{ have with Christ?}

A. The communion in glory\index{Communion:in glory ``f10!} which the members of the\index{Church members ``f10!}
invisible church\index{Church:invisible ``f10!} have with Christ, is in this life,\footnote{\ 2 Corinthians
3:18\index{47 2 Corinthians:ch.   3, v.  18, pg. ``f30!}.} immediately after death\index{Death:believers:immediate
perfection ``f10!},\footnote{\ Luke 23:43\index{42 Luke:ch.  23, v.  43, pg. ``f30!}.} and at last perfected
at\index{Resurrection:believers:perfected ``f10!} the resurrection and day of judgment\index{Judgement
day:believers:perfected ``f10!}.\footnote{\ 1 Thessalonians 4:17\index{52 1 Thessalonians:ch.   4, v.  17, pg.
``f30!}.}

Q. 83 \textit{What is the communion in glory}\index{Communion:in glory ``f10!}\textit{ with Christ which the members of
the}\index{Church members ``f10!}\textit{ invisible church}\index{Church:invisible ``f10!}\textit{ enjoy
in}\index{Believers:enjoy:communion with God ``f10!}\textit{ this life?}

A. The members of the\index{Church members ``f10!} invisible church\index{Church:invisible ``f10!} have communicated to
them in this life the firstfruits of glory\index{Church:communicates firstfruits of glory with Christ ``f10!} with
Christ, as they are members of him their head\index{Church:Christ as head ``f10!}, and so in him are
interested\index{Believers:interest (share) in Christ ``f10!} in that glory\index{Jesus:glory ``f10!} which he is fully
possessed of;\footnote{\ Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5, pg. ``f30!}.} and, as an
earnest\index{Believers:inheritance:earnest of ``f10!} thereof, enjoy the\index{Believers:enjoy:sense of God's love
``f10!} sense of God's love\index{God's attributes:love ``f10!},\footnote{\ Romans 5:5\index{45 Romans:ch.   5, v.   5,
pg. ``f30!}; 2 Corinthians 1:22\index{47 2 Corinthians:ch.   1, v.  22, pg. ``f30!}.} peace of
conscience\index{Conscience:peace of ``f10!}, joy in\index{Joy:in Holy Spirit ``f10!} the Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!}, and hope of glory\index{Hope:of glory ``f10!};\footnote{\ Romans 5:1\index{45 Romans:ch.  
5, v.   1, pg. ``f30!}{}-2\index{45 Romans:ch.   5, v.   1-2, pg. ``f30!}; Romans 14:17\index{45 Romans:ch.  14, v. 
17, pg. ``f30!}.} as, on the contrary, sense of God's revenging\index{Revenge:God's ``f10!} wrath, horror of
conscience\index{Conscience:horror of ``f10!}, and a fearful\index{Fear:of God:abject ``f10!}
expectation\index{Judgement day:fearful expectation ``f10!} of judgment, are\index{Judgement day:expectation ``f10!} to
the wicked the\index{Judgement day:wicked:fearful expectation ``f10!} beginning of their torments
which\index{Punishment:everlasting:torments (body and soul) ``f10!} they shall endure after
death\index{Death:following:wicked endure torments ``f10!}.\footnote{\ Genesis 4:13\index{01 Genesis:ch.   4, v.  13,
pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v.   4, pg. ``f30!}; Hebrews 10:27\index{58 Hebrews:ch.  10, v. 
27, pg. ``f30!}; Romans 2:9\index{45 Romans:ch.   2, v.   9, pg. ``f30!}; Mark 9:44\index{41 Mark:ch.   9, v.  44, pg.
``f30!}.}


\bigskip

Q. 84 \textit{Shall all men die}\index{Death:punishment for sin ``f10!}\textit{?}

A. Death being\index{Death:wages of sin ``f10!} threatened as\index{God's law:commandments:threatenings for disobedience
``f10!} the wages\index{Death:punishment for sin ``f10!} of sin,\footnote{\ Romans 6:23\index{45 Romans:ch.   6, v. 
23, pg. ``f30!}.} it is appointed unto all men once to die;\footnote{\ Hebrews 9:27\index{58 Hebrews:ch.   9, v.  27,
pg. ``f30!}.} for that all have sinned\index{Sin:all have sinned ``f10!}.\footnote{\ Romans 5:1\index{45 Romans:ch.  
5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}.}


\bigskip

Q. 85 \textit{Death, being}\index{Death:wages of sin ``f10!}\textit{ the wages}\index{Death:punishment for sin
``f10!}\textit{ of sin, why are not the righteous delivered from death, seeing}\index{Death:righteous not delivered
from temporal death ``f10!}\textit{ all their sins are forgiven}\index{Salvation:sins forgiven in Christ
``f10!}\textit{ in Christ?}

A. The righteous shall\index{Death:believers:delivered from sting of ``f10!} be delivered from death itself at the last
day\index{Last day:last day ``f10!}, and even in death are delivered from the sting and curse of it\index{Curse:of
death, delivered from ``f10!};\footnote{\ 1 Corinthians 15:26, 55-57\index{46 1 Corinthians:ch.  15, v.  26, 55-57, pg.
``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}5\index{58 Hebrews:ch.   2, v.  15, pg. ``f30!}.} so
that, although they die, yet\index{Death:righteous not delivered from temporal death ``f10!} it is out of God's
love\index{God's attributes:love ``f10!},\footnote{\ Isaiah 57:1-2\index{23 Isaiah:ch.  57, v.   1-2, pg. ``f30!}; 2
Kings 22:20\index{12 2 Kings:ch.  22, v.  20, pg. ``f30!}.} to free them perfectly from sin and
misery\index{Punishment:misery ``f10!},\footnote{\ Revelation 14:13\index{66 Revelation:ch.  14, v.  13, pg. ``f30!};
Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}7\index{49 Ephesians:ch.   5, v.  27, pg. ``f30!}.} and to
make them capable of further communion\index{Communion:with Christ ``f10!} with Christ in
glory\index{Resurrection:believers:into glory:with Christ ``f10!}, which they then enter upon.\footnote{\ Luke
23:43\index{42 Luke:ch.  23, v.  43, pg. ``f30!}; Philippians 1:23\index{50 Philippians:ch.   1, v.  23, pg. ``f30!}.}


\bigskip

Q. 86 \textit{What is the communion in glory}\index{Communion:in glory ``f10!}\textit{ with Christ, which the members of
the}\index{Church members ``f10!}\textit{ invisible church}\index{Church:invisible ``f10!}\textit{ enjoy immediately
after death}\index{Death:believers:immediate perfection ``f10!}\textit{?}

A. The communion in glory\index{Communion:in glory ``f10!} with Christ, which the members of the\index{Church members
``f10!} invisible church\index{Church:invisible ``f10!} enjoy immediately after death\index{Death:believers:immediate
perfection ``f10!}, is, in that their souls are\index{Believers:souls made perfect at death ``f10!} then made perfect
in holiness\index{Believers:holy:made perfect at death ``f10!},\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.  
2, pg. ``f30!}3\index{58 Hebrews:ch.  12, v.  23, pg. ``f30!}.} and received into the highest heavens,\footnote{\ 2
Corinthians 5:1, 6, 8\index{47 2 Corinthians:ch.   5, v.   1, 6, 8, pg. ``f30!}; Philippians 1:23\index{50
Philippians:ch.   1, v.  23, pg. ``f30!}; Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!}; Ephesians 4:10\index{49
Ephesians:ch.   4, v.  10, pg. ``f30!}.} where they behold\index{Believers:behold face of God in glory ``f10!} the face
of\index{Resurrection:believers:behold face of God ``f10!} God in light and glory\index{Resurrection:believers:into
glory:light ``f10!},\footnote{\ 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg. ``f30!}; 1 Corinthians 13:12\index{46 1
Corinthians:ch.  13, v.  12, pg. ``f30!}.} waiting for\index{Resurrection:believers:waiting expectantly ``f10!} the
full redemption\index{Redemption:full ``f10!} of their bodies,\index{Bodies in graves:believers
``f10!}\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}3\index{45 Romans:ch.   8, v.  23, pg.
``f30!}; Psalm 16:9\index{19 Psalms:ch.  16, v.   9, pg. ``f30!}.} which even in death
continue\index{Death:believers:continue united to Christ ``f10!} united to Christ\index{Believers:union with
Christ:mystical ``f10!},\footnote{\ 1 Thessalonians 4:14\index{52 1 Thessalonians:ch.   4, v.  14, pg. ``f30!}.} and
rest in their graves\index{Resurrection:bodies:await in graves ``f10!} as in their beds,\footnote{\ Isaiah
57:2\index{23 Isaiah:ch.  57, v.   2, pg. ``f30!}.} till at the last day\index{Last day:last day ``f10!} they be again
united to the\index{Jesus:human nature:united with divine ``f10!}ir\index{Resurrection:bodies:united to souls ``f10!}
souls.\footnote{\ Job 19:26-27\index{18 Job:ch.  19, v.  26-27, pg. ``f30!}.} Whereas the
souls\index{Punishment:hell:wicked cast into ``f10!} of the wicked are\index{Judgement day:wicked:cast into:hell
``f10!} at their death cast into hell\index{Hell:hell ``f10!}, where they remain in torments
and\index{Punishment:everlasting:torments (body and soul) ``f10!} utter darkness\index{darkness, spiritual ``f10!}, and
their bodies kept\index{Bodies in graves:unbelievers ``f10!} in their graves, as in their prisons, till the
resurrection\index{Resurrection:general, of dead ``f10!} and judgment of the\index{Judgement:final ``f10!} great
day.\footnote{\ Luke 16:23-24\index{42 Luke:ch.  16, v.  23-24, pg. ``f30!}; Acts 1:25\index{44 Acts:ch.   1, v.  25,
pg. ``f30!}; Jude 1:6\index{65 Jude:v.   1, v.   6, pg. ``f30!}{}-7\index{65 Jude:v.   1, v.   6-7, pg. ``f30!}.}


\bigskip

Q. 87 \textit{What are we to believe concerning the resurrection}\index{Resurrection:general, of dead ``f10!}\textit{?}

A. We are to believe, that at the last day\index{Last day:last day ``f10!} there shall be a general
resurrection\index{Resurrection:general, of dead ``f10!} of the dead, both\index{Death:following:all men raised ``f10!}
of the just and unjust\index{Resurrection:of just and unjust ``f10!}:\footnote{\ Acts 24:15\index{44 Acts:ch.  24, v. 
15, pg. ``f30!}.} when they that are then found alive\index{Last day:those alive on earth ``f10!} shall in a moment
be\index{Resurrection:bodies:changed in a moment ``f10!} changed; and the selfsame bodies of the dead
which\index{Death:bodies:laid in graves ``f10!} were laid in the grave\index{Resurrection:bodies:await in graves
``f10!}, being then again united to the\index{Jesus:human nature:united with divine
``f10!}ir\index{Resurrection:bodies:united to souls ``f10!} souls forever, shall be raised up by the
power\index{Resurrection:believers:by the power of Christ ``f10!} of Christ.\footnote{\ 1 Corinthians 15:51-53\index{46
1 Corinthians:ch.  15, v.  51-53, pg. ``f30!}; 1 Thessalonians 4:15-17\index{52 1 Thessalonians:ch.   4, v.  15-17, pg.
``f30!}; John 5:28-29\index{43 John:ch.   5, v.  28-29, pg. ``f30!}.} The bodies of the\index{Resurrection:bodies:same
ones as before death, for just and unjust ``f10!} just, by the Spirit of Christ\index{Holy Spirit:Spirit of Christ
``f10!}, and by virtue\index{Jesus:vicarious atonement ``f10!} of his resurrection\index{Jesus:resurrection:from the
dead ``f10!} as their head\index{Church:Christ as head ``f10!}, shall be raised in
power\index{Resurrection:believers:in power ``f10!}, spiritual,
incorruptible\index{Resurrection:believers:incorruptible ``f10!}, and made like to his glorious
body\index{Jesus:resurrection:body:glorious ``f10!};\footnote{\ 1 Corinthians 15:21-23, 42-44\index{46 1
Corinthians:ch.  15, v.  21-23, 42-44, pg. ``f30!}; Philippians 3:21\index{50 Philippians:ch.   3, v.  21, pg.
``f30!}.} and the bodies of the wicked shall\index{Resurrection:wicked raised in dishonour ``f10!} be raised up in
dishonor by him, as an offended judge\index{Judgement:final ``f10!}.\footnote{\ John 5:27-29\index{43 John:ch.   5, v. 
27-29, pg. ``f30!}; Matthew 25:33\index{40 Matthew:ch.  25, v.  33, pg. ``f30!}.} 


\bigskip

Q. 88 \textit{What shall immediately follow after the resurrection}\index{Resurrection:general, of dead
``f10!}\textit{?}

A. Immediately after the resurrection\index{Resurrection:general, of dead ``f10!} shall follow the general and final
judgment\index{Judgement:final ``f10!} of angels\index{Angels:angels ``f10!} and men;\footnote{\ 2 Peter 2:4\index{61 2
Peter:ch.   2, v.   4, pg. ``f30!}; Jude 1:6\index{65 Jude:v.   1, v.   6, pg. ``f30!}{}-7\index{65 Jude:v.   1, v.  
6-7, pg. ``f30!}, 14-15\index{65 Jude:v.   1, v.   6-7, 14-15, pg. ``f30!}; Matthew 25:46\index{40 Matthew:ch.  25, v. 
46, pg. ``f30!}.} the day and hour whereof no man knows, that\index{Jesus:return:no one knows hour ``f10!} all may
watch and\index{Jesus:return:watchfulness ``f10!} pray, and be\index{Prayer:in watchfulness for coming of the Lord
``f10!} ever ready for the coming of the Lord.\footnote{\ Matthew 24:36\index{40 Matthew:ch.  24, v.  36, pg. ``f30!},
42, 44\index{40 Matthew:ch.  24, v.  36, 42, 44, pg. ``f30!}; Luke 21:35-36\index{42 Luke:ch.  21, v.  35-36, pg.
``f30!}.}


\bigskip

Q. 89 \textit{What shall be done to the wicked at}\index{Judgement day:wicked:cast into:hell ``f10!}\textit{ the day of
judgment}\index{Judgement day:believers:perfected ``f10!}\textit{?}

A. At the day of judgment\index{Judgement day:believers:perfected ``f10!}, the wicked shall be set\index{Judgement
day:wicked:set on left ``f10!} on Christ's left hand,\footnote{\ Matthew 25:33\index{40 Matthew:ch.  25, v.  33, pg.
``f30!}.} and, upon clear evidence, and full conviction\index{Judgement day:wicked:full conviction of consciences
``f10!} of their own consciences,\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}5\index{45
Romans:ch.   2, v.  15, pg. ``f30!}{}-16\index{45 Romans:ch.   2, v.  15-16, pg. ``f30!}.} shall have the
fearful\index{Fear:of God:abject ``f10!} but just sentence\index{Punishment:just:sentence ``f10!} of
condemnation\index{Punishment:condemnation ``f10!} pronounced against them;\footnote{\ Matthew 25:41\index{40
Matthew:ch.  25, v.  41, pg. ``f30!}{}-43\index{40 Matthew:ch.  25, v.  41-43, pg. ``f30!}.} and thereupon shall be
cast out\index{Punishment:hell:wicked cast into ``f10!} from the favorable presence of God, and the
glorious\index{Resurrection:believers:into glory:fellowship ``f10!} fellowship with\index{Fellowship:with Christ
``f10!} Christ, his saints\index{Believers:saints ``f10!}, and all his holy\index{Angels:holy ``f10!}
angels\index{Angels:angels ``f10!}, into hell\index{Hell:hell ``f10!}, to be
punished\index{Punishment:everlasting:torments (body and soul) ``f10!} with unspeakable torments, both of body and
soul, with the devil\index{Devil:devil ``f10!} and his angels forever.\footnote{\ Luke 16:26\index{42 Luke:ch.  16, v. 
26, pg. ``f30!}; 2 Thessalonians 1:8-9\index{53 2 Thessalonians:ch.   1, v.   8-9, pg. ``f30!}.}

Q. 90 \textit{What shall be done to the righteous at}\index{Resurrection:believers:into glory:with Christ
``f10!}\textit{ the day of judgment}\index{Judgement day:believers:perfected ``f10!}\textit{?}

A. At the day of judgment\index{Judgement day:believers:perfected ``f10!}, the righteous, being
caught\index{Resurrection:believers:into glory:with Christ ``f10!} up to Christ in the clouds,\footnote{\ 1
Thessalonians 4:17\index{52 1 Thessalonians:ch.   4, v.  17, pg. ``f30!}.} shall be set on his right hand, and there
openly acknowledged\index{Judgement day:believers:acknowledged by God ``f10!} and acquitted\index{Judgement
day:believers:acquitted by God ``f10!},\footnote{\ Matthew 10:32\index{40 Matthew:ch.  10, v.  32, pg. ``f30!}; Matthew
25:33\index{40 Matthew:ch.  25, v.  33, pg. ``f30!}.} shall join with him in the judging of
reprobate\index{Judgement:of reprobate angels and men ``f10!} angels\index{Angels:angels ``f10!} and men,\footnote{\ 1
Corinthians 6:2-3\index{46 1 Corinthians:ch.   6, v.   2-3, pg. ``f30!}.} and shall be received into
heaven\index{Believers:received into heaven ``f10!},\footnote{\ Matthew 25:34, 46\index{40 Matthew:ch.  25, v.  34, 46,
pg. ``f30!}.} where they shall be fully and forever freed from all\index{Resurrection:believers:freed from sin ``f10!}
sin and misery\index{Punishment:misery ``f10!};\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg.
``f30!}7\index{49 Ephesians:ch.   5, v.  27, pg. ``f30!}; Revelation 14:13\index{66 Revelation:ch.  14, v.  13, pg.
``f30!}.} filled with inconceivable joys\index{Joy:inconceivable ``f10!},\footnote{\ Psalm 16:11\index{19 Psalms:ch. 
16, v.  11, pg. ``f30!}.} made perfectly holy\index{Believers:holy:made perfect at death ``f10!} and
happy\index{happiness ``f10!} both in body and soul, in\index{Believers:souls made perfect at death ``f10!} the company
of innumerable\index{Heaven:innumerable saints and angels ``f10!} saints\index{Believers:saints ``f10!} and holy
angels,\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}2\index{58 Hebrews:ch.  12, v.  22, pg.
``f30!}{}-23\index{58 Hebrews:ch.  12, v.  22-23, pg. ``f30!}.} but especially in the immediate
vision\index{Resurrection:believers:immediate vision of God ``f10!} and fruition of God the Father\index{Father:God the
``f10!}, of our Lord Jesus Christ, and of the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}, to all
eternity\index{Believers:blessed to all eternity ``f10!}.\footnote{\ 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg.
``f30!}; 1 Corinthians 13:12\index{46 1 Corinthians:ch.  13, v.  12, pg. ``f30!}; 1 Thessalonians 4:17\index{52 1
Thessalonians:ch.   4, v.  17, pg. ``f30!}{}-18\index{52 1 Thessalonians:ch.   4, v.  17-18, pg. ``f30!}.} And this is
the perfect and full communion, which\index{Communion:with God ``f10!} the members of the\index{Church members ``f10!}
invisible church\index{Church:invisible ``f10!} shall enjoy with Christ in glory, at the resurrection and day of
judgment.


\bigskip

Q. 91 \textit{What is the duty which God}\index{God's law:commandments:duties:required by God ``f10!}\textit{ requires
of man?}

A. The duty which God\index{God's law:commandments:duties:required by God ``f10!} requires of man, is obedience to his
revealed will\index{Revelation:God's will ``f10!}.\footnote{\ Romans 12:1-2\index{45 Romans:ch.  12, v.   1-2, pg.
``f30!}; Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; 1 Samuel 15:22\index{09 1 Samuel:ch.  15, v.  22, pg.
``f30!}.}


\bigskip

Q. 92 \textit{What did God at first reveal unto}\index{God's law:commandments:first revelation ``f10!}\textit{ man as
the rule of his}\index{God's law:moral:rule for obedience ``f10!}\textit{ obedience?}

A. The rule of obedience\index{Covenant:works:perfect obedience required ``f10!} revealed to\index{Revelation:God's law
``f10!} Adam in the\index{Adam:given a law ``f10!} estate of innocence\index{Mankind:created:state of innocence
``f10!}, and to all mankind in\index{God's law:God's will for mankind ``f10!} him, besides a special command not to eat
of the fruit of the tree of the\index{Tree:of the knowledge of good and evil ``f10!} knowledge of good and evil, was
the moral\index{God's law:commandments:moral ``f10!} law.\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg.
``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}{}-27\index{01 Genesis:ch.   1, v.  26-27, pg. ``f30!}; Genesis
2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Romans 2:1\index{45 Romans:ch.   2, v.   1, pg.
``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}; Romans 10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}.}


\bigskip

Q. 93 \textit{What is the moral}\index{God's law:commandments:moral ``f10!}\textit{ law?}

A. The moral\index{God's law:commandments:moral ``f10!} law is the declaration\index{God's law:moral:declaration of
God's will ``f10!} of the will of God to mankind\index{God's law:God's will for mankind ``f10!}, directing and binding
everyone\index{God's law:binding on all men ``f10!} to personal, perfect, and perpetual conformity\index{God's
law:moral:perpetual conformity required ``f10!} and obedience\index{God's law:commandments:obedience:required ``f10!}
thereunto, in the frame and\index{God's law:commandments:address words, works, gestures ``f10!} disposition of the
whole man, soul\index{God's law:binding on whole man, soul and body ``f10!} and body,\footnote{\ Deuteronomy 5:1-3, 31,
33\index{05 Deuteronomy:ch.   5, v.   1-3, 31, 33, pg. ``f30!}; Luke 10:26-27\index{42 Luke:ch.  10, v.  26-27, pg.
``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}; 1 Thessalonians 5:23\index{52 1
Thessalonians:ch.   5, v.  23, pg. ``f30!}.} and in performance of all those duties\index{God's
law:commandments:duties:required by God ``f10!} of holiness\index{Believers:holy:being perfected ``f10!} and
righteousness which\index{God's law:righteous requirements ``f10!} he owes to God and man:\footnote{\ Luke
1:75\index{42 Luke:ch.   1, v.  75, pg. ``f30!}; Acts 14:16\index{44 Acts:ch.  14, v.  16, pg. ``f30!}.} promising
life\index{God's law:commandments:promises included ``f10!} upon the fulfilling, and\index{Covenant:works:conditional
on perfect obedience ``f10!} threatening death\index{Death:punishment for sin ``f10!} upon the breach of it\index{God's
law:breach:death threatened ``f10!}.\footnote{\ Romans 10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}; Galatians
3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}, 12\index{48 Galatians:ch.   3, v.  10, 12, pg. ``f30!}.}


\bigskip


\bigskip


\bigskip

Q. 94 \textit{Is there any use of the moral}\index{God's law:commandments:moral ``f10!}\textit{ law to man since the
fall}\index{Sin:original:first transgression (of Adam) ``f10!}\textit{?}

A. Although no man, since the fall\index{Sin:original:first transgression (of Adam) ``f10!}, can attain to
righteousness\index{God's law:moral:no person attain righteousness ``f10!} and life by the moral\index{God's
law:commandments:moral ``f10!} law;\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}; Galatians
2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}.} yet there is great use thereof, as well common to all men, as
peculiar either to the unregenerate\index{Unbelievers:unregenerate ``f10!}, or the
regenerate\index{Regeneration:regeneration ``f10!}.\footnote{\ 1 Timothy 1:8\index{54 1 Timothy:ch.   1, v.   8, pg.
``f30!}.}


\bigskip

Q. 95 \textit{Of what use is the moral}\index{God's law:commandments:moral ``f10!}\textit{ law to all men?}

A. The moral\index{God's law:commandments:moral ``f10!} law is of use to all men, to inform them of the holy\index{God's
attributes:holy:nature ``f10!} nature and will of God,\footnote{\ Leviticus 11:44-45\index{03 Leviticus:ch.  11, v. 
44-45, pg. ``f30!}; Leviticus 20:7-8\index{03 Leviticus:ch.  20, v.   7-8, pg. ``f30!}; Romans 7:12\index{45 Romans:ch.
  7, v.  12, pg. ``f30!}.} and of their duty, binding them\index{God's law:binding on all men ``f10!} to walk
accordingly;\footnote{\ Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; James 2:10\index{59 James:ch.   2, v. 
10, pg. ``f30!}{}-11\index{59 James:ch.   2, v.  10-11, pg. ``f30!}.} to convince them\index{God's law:convince men of
disability to keep ``f10!} of their disability to keep it, and of the sinful pollution\index{Sin:original:corrupt
nature:polluted ``f10!} of their nature, hearts, and lives\index{Heart:polluted ``f10!};\footnote{\ Psalm
19:11\index{19 Psalms:ch.  19, v.  11, pg. ``f30!}{}-12\index{19 Psalms:ch.  19, v.  11-12, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}; Romans 7:7\index{45
Romans:ch.   7, v.   7, pg. ``f30!}.} to humble them\index{Humility:before God ``f10!} in the sense of their sin and
misery\index{Punishment:misery ``f10!},\footnote{\ Romans 3:9\index{45 Romans:ch.   3, v.   9, pg. ``f30!}, 23\index{45
Romans:ch.   3, v.   9, 23, pg. ``f30!}.} and thereby help them\index{God's law:provides sight of need for Christ
``f10!} to a clearer sight of the need they have of Christ,\footnote{\ Galatians 3:21\index{48 Galatians:ch.   3, v. 
21, pg. ``f30!}{}-22\index{48 Galatians:ch.   3, v.  21-22, pg. ``f30!}.} and of the perfection of
his\index{Jesus:obedience:to God's law ``f10!} obedience.\footnote{\ Romans 10:4\index{45 Romans:ch.  10, v.   4, pg.
``f30!}.}


\bigskip

Q. 96 \textit{What particular use is there of the moral}\index{God's law:commandments:moral ``f10!}\textit{ law to
unregenerate}\index{Unbelievers:unregenerate ``f10!}\textit{ men?}

A. The moral\index{God's law:commandments:moral ``f10!} law is of use to unregenerate\index{Unbelievers:unregenerate
``f10!} men, to awaken\index{God's law:awaken conscience ``f10!} their consciences to\index{Conscience:awakened ``f10!}
flee\index{Wrath:of God:flee from ``f10!} from wrath to\index{Wrath:of God:pending ``f10!} come,\footnote{\ 1 Timothy
1:9-10\index{54 1 Timothy:ch.   1, v.   9-10, pg. ``f30!}.} and to drive\index{God's law:drive men to Christ ``f10!}
them to Christ;\footnote{\ Galatians 3:24\index{48 Galatians:ch.   3, v.  24, pg. ``f30!}.} or, upon their continuance
in the estate and\index{Unbelievers:in estate and way of sin ``f10!} way of sin, to leave them
inexcusable\index{Unbelievers:inexcusable ``f10!},\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.  20, pg. ``f30!};
Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}5\index{45 Romans:ch.   2, v.  15, pg. ``f30!}.} and under the
curse thereof\index{Curse:of the law ``f10!}.\footnote{\ Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg.
``f30!}.}


\bigskip

Q. 97 \textit{What special use is there of the moral}\index{God's law:commandments:moral ``f10!}\textit{ law to the
regenerate}\index{Regeneration:regeneration ``f10!}\textit{?}

A. Although they that are regenerate\index{Regeneration:regeneration ``f10!}, and believe in Christ, be
delivered\index{Covenant:works:believers delivered from ``f10!} from the moral\index{God's law:commandments:moral
``f10!} law as a covenant of works\index{Covenant:works:works ``f10!},\footnote{\ Romans 6:14\index{45 Romans:ch.   6,
v.  14, pg. ``f30!}; Romans 7:4, 6\index{45 Romans:ch.   7, v.   4, 6, pg. ``f30!}; Galatians 4:4\index{48
Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}.} so as thereby they are
neither justified\index{Justification:justified ``f10!}\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg.
``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}.} nor condemned;\footnote{\ Galatians 5:2\index{48 Galatians:ch. 
 5, v.   2, pg. ``f30!}3\index{48 Galatians:ch.   5, v.  23, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1,
pg. ``f30!}.} yet, besides the general uses thereof common to them with all men, it is of special use, to show them how
much they are bound to Christ for his fulfilling it\index{Jesus:obedience:to God's law ``f10!}, and enduring the curse
thereof in\index{Jesus:curse in believers' stead ``f10!} their stead\index{Jesus:vicarious atonement ``f10!}, and for
their good;\footnote{\ Romans 7:24-25\index{45 Romans:ch.   7, v.  24-25, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.  
8, v.   3, pg. ``f30!}{}-4\index{45 Romans:ch.   8, v.   3-4, pg. ``f30!}; Galatians 3:13\index{48 Galatians:ch.   3,
v.  13, pg. ``f30!}{}-14\index{48 Galatians:ch.   3, v.  13-14, pg. ``f30!}.} and thereby to
provoke\index{Prayer:thanking God ``f10!} them to more thankfulness,\footnote{\ Luke 1:6\index{42 Luke:ch.   1, v.   6,
pg. ``f30!}8-69, 74-75\index{42 Luke:ch.   1, v.  68-69, 74-75, pg. ``f30!}; Colossians 1:12-14\index{51 Colossians:ch.
  1, v.  12-14, pg. ``f30!}.} and to express the same in their greater care to conform\index{God's
law:commandments:obedience:required ``f10!} themselves thereunto as the rule of their obedience.\footnote{\ Romans
7:22\index{45 Romans:ch.   7, v.  22, pg. ``f30!}; Romans 12:2\index{45 Romans:ch.  12, v.   2, pg. ``f30!}; Titus
2:11-14\index{56 Titus:ch.   2, v.  11-14, pg. ``f30!}.}

Q. 98 \textit{Where is the moral}\index{God's law:commandments:moral ``f10!}\textit{ law summarily
comprehended}\index{God's law:moral:summarily comprehended in ten commandments ``f10!}\textit{?}

A. The moral\index{God's law:commandments:moral ``f10!} law is summarily comprehended\index{God's law:moral:summarily
comprehended in ten commandments ``f10!} in the ten\index{God's law:commandments:ten ``f10!} commandments, which were
delivered by\index{God's law:commandments:delivered on Sinai ``f10!} the voice of God\index{God's
law:commandments:delivered by voice of God ``f10!} upon mount Sinai\index{Sinai, mount ``f10!}, and written by him in
two tables\index{God's law:commandments:two tables ``f10!} of stone;\index{God's law:commandments:tablets of stone
``f10!}\footnote{\ Deuteronomy 10:4\index{05 Deuteronomy:ch.  10, v.   4, pg. ``f30!}; Exodus 34:1\index{02 Exodus:ch. 
34, v.   1, pg. ``f30!}{}-4\index{02 Exodus:ch.  34, v.   1-4, pg. ``f30!}.} and are recorded in the twentieth chapter
of Exodus. The four first commandment\index{God's law:commandments:first ``f10!}s\index{God's law:commandments:first
four, duty to God ``f10!} containing our duty to God\index{God's law:commandments:duties:required by God ``f10!}, and
the other six\index{Sabbath:six days:for own affairs ``f10!} our duty to man\index{God's law:commandments:duties:to
others ``f10!}.\footnote{\ Matthew 22:37-38, 40\index{40 Matthew:ch.  22, v.  37-38, 40, pg. ``f30!}.}


\bigskip

Q. 99 \textit{What rules are to be observed for the}\index{God's law:commandments:rules for right understanding
``f10!}\textit{ right understanding of the ten}\index{God's law:commandments:ten ``f10!}\textit{ commandments?}

A. For the right understanding\index{God's law:commandments:rules for right understanding ``f10!} of the ten\index{God's
law:commandments:ten ``f10!} commandments, these rules are to be observed: 1. That the law is perfect, and binds
everyone\index{God's law:binding on all men ``f10!} to full conformity in the whole man unto the righteousness
thereof\index{God's law:righteous requirements ``f10!}, and unto entire obedience\index{God's
law:commandments:obedience:required ``f10!} forever; so as to require the\index{God's law:commandments:duties:required
by God ``f10!} utmost perfection of every duty, and to\index{God's law:commandments:duties:perfection required ``f10!}
forbid the least degree\index{God's law:breach:least degree forbidden ``f10!} of every sin.\footnote{\ Psalm
19:7\index{19 Psalms:ch.  19, v.   7, pg. ``f30!}; James 2:10\index{59 James:ch.   2, v.  10, pg. ``f30!}; Matthew
5:21-22\index{40 Matthew:ch.   5, v.  21-22, pg. ``f30!}.} 2. That it is spiritual, and\index{God's law:spiritual
``f10!} so reaches the understanding, will, affections, and all\index{God's law:addresses affections ``f10!} other
powers of the soul; as\index{God's law:commandments:obedience:whole being ``f10!} well as words, works\index{God's
law:commandments:address words, works, gestures ``f10!}, and gestures.\footnote{\ Romans 7:14\index{45 Romans:ch.   7,
v.  14, pg. ``f30!}; Deuteronomy 6:5\index{05 Deuteronomy:ch.   6, v.   5, pg. ``f30!}; Matthew 5:21-22\index{40
Matthew:ch.   5, v.  21-22, pg. ``f30!}, 27-28\index{40 Matthew:ch.   5, v.  21-22, 27-28, pg. ``f30!}, 33-34, 37-39,
43-44\index{40 Matthew:ch.   5, v.  21-22, 27-28, 33-34, 37-39, 43-44, pg. ``f30!}; Matthew 22:37-39\index{40
Matthew:ch.  22, v.  37-39, pg. ``f30!}.} 3. That one and the same thing, in divers respects, is required or
forbidden\index{God's law:commandments:sins forbidden ``f10!} in several commandments\index{God's
law:commandments:multiple may apply ``f10!}.\footnote{\ Colossians 3:5\index{51 Colossians:ch.   3, v.   5, pg.
``f30!}; Amos 8:5\index{30 Amos:ch.   8, v.   5, pg. ``f30!}; Proverbs 1:19\index{20 Proverbs:ch.   1, v.  19, pg.
``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   6, v.  10, pg.
``f30!}.} 4. That as, where a duty is commanded\index{God's law:commandments:prohibitions and duties ``f10!}, the
contrary sin is forbidden;\footnote{\ Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}; Deuteronomy
6:13\index{05 Deuteronomy:ch.   6, v.  13, pg. ``f30!}; Matthew 4:9-10\index{40 Matthew:ch.   4, v.   9-10, pg.
``f30!}; Matthew 15:4-6\index{40 Matthew:ch.  15, v.   4-6, pg. ``f30!}.} and, where a sin is forbidden, the contrary
duty is commanded:\footnote{\ Matthew 5:21-25\index{40 Matthew:ch.   5, v.  21-25, pg. ``f30!}; Ephesians 4:28\index{49
Ephesians:ch.   4, v.  28, pg. ``f30!}.} so, where a promise is\index{God's law:commandments:promises included ``f10!}
annexed, the contrary threatening is\index{God's law:commandments:threat and promise ``f10!}
included;\footnote{\ Exodus 20:12\index{02 Exodus:ch.  20, v.  12, pg. ``f30!}; Proverbs 30:17\index{20 Proverbs:ch. 
30, v.  17, pg. ``f30!}.} and, where a threatening is annexed, the contrary promise is included.\footnote{\ Jeremiah
18:7-8\index{24 Jeremiah:ch.  18, v.   7-8, pg. ``f30!}; Exodus 20:7\index{02 Exodus:ch.  20, v.   7, pg. ``f30!};
Psalm 15:1, 4-5\index{19 Psalms:ch.  15, v.   1, 4-5, pg. ``f30!}; Psalm 24:4\index{19 Psalms:ch.  24, v.   4, pg.
``f30!}{}-5\index{19 Psalms:ch.  24, v.   4-5, pg. ``f30!}.} 5. That what God forbids, is at no time to be
done;\footnote{\ Job 13:7\index{18 Job:ch.  13, v.   7, pg. ``f30!}; Job 36:21\index{18 Job:ch.  36, v.  21, pg.
``f30!}; Romans 3:8\index{45 Romans:ch.   3, v.   8, pg. ``f30!}; Hebrews 11:25\index{58 Hebrews:ch.  11, v.  25, pg.
``f30!}.}[8] What he commands, is always our duty;\footnote{\ Deuteronomy 4:8-9\index{05 Deuteronomy:ch.   4, v.   8-9,
pg. ``f30!}.} and yet every particular duty is not\index{God's law:commandments:duties:not at all times ``f10!} to be
done at all times.\footnote{\ Matthew 12:7\index{40 Matthew:ch.  12, v.   7, pg. ``f30!}.} 6. That under one sin or
duty, all of the same kind are forbidden or commanded; together with all the causes, means, occasions, and appearances
thereof, and provocations thereunto.\footnote{\ Matthew 5:21-22\index{40 Matthew:ch.   5, v.  21-22, pg. ``f30!},
27-28\index{40 Matthew:ch.   5, v.  21-22, 27-28, pg. ``f30!}; Matthew 15:4-6\index{40 Matthew:ch.  15, v.   4-6, pg.
``f30!}; Hebrews 10:24-25\index{58 Hebrews:ch.  10, v.  24-25, pg. ``f30!}; 1 Thessalonians 5:22-23\index{52 1
Thessalonians:ch.   5, v.  22-23, pg. ``f30!}; Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}6\index{48
Galatians:ch.   5, v.  26, pg. ``f30!}; Colossians 3:2\index{51 Colossians:ch.   3, v.   2, pg. ``f30!}1\index{51
Colossians:ch.   3, v.  21, pg. ``f30!}.} 7. That what is forbidden or commanded to ourselves\index{God's
law:commandments:duties:to others ``f10!}, we are bound, according to our places, to endeavor that\index{God's
law:commandments:obedience:endeavoring ``f10!} it may be avoided or performed by others, according to the duty of their
places.\footnote{\ Exodus 20:10\index{02 Exodus:ch.  20, v.  10, pg. ``f30!}; Leviticus 19:17\index{03 Leviticus:ch. 
19, v.  17, pg. ``f30!}; Genesis 18:19\index{01 Genesis:ch.  18, v.  19, pg. ``f30!}; Joshua 24:15\index{06 Joshua:ch. 
24, v.  15, pg. ``f30!}; Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg. ``f30!}{}-7\index{05 Deuteronomy:ch.
  6, v.   6-7, pg. ``f30!}.} 8. That in what is commanded to others, we are bound, according to our places and
calling\index{Calling:in life ``f10!}s, to be helpful to them;\footnote{\ 2 Corinthians 1:24\index{47 2 Corinthians:ch.
  1, v.  24, pg. ``f30!}.} and to take heed\index{God's law:commandments:duties:not what is forbidden ``f10!} of
partaking with others in: What is forbidden them.\footnote{\ 1 Timothy 5:22\index{54 1 Timothy:ch.   5, v.  22, pg.
``f30!}.}


\bigskip

Q. 100 \textit{What special things are we to consider in the ten}\index{God's law:commandments:ten ``f10!}\textit{
commandments?}

A. We are to consider, in the ten\index{God's law:commandments:ten ``f10!} commandments, the preface, the\index{God's
law:commandments:preface ``f10!} substance of the commandment\index{God's law:commandments:given to Adam ``f10!}s
themselves, and several reasons annexed to some of them, the more to enforce them\index{God's law:commandments:words of
enforcement ``f10!}.


\bigskip

Q. 101 \textit{What is the preface to}\index{God's law:commandments:preface ``f10!}\textit{ the ten}\index{God's
law:commandments:ten ``f10!}\textit{ commandments?}

A. The preface to\index{God's law:commandments:preface ``f10!} the ten\index{God's law:commandments:ten ``f10!}
commandments is contained in these words, I am the Lord thy God, which have brought you out of the land of
Egypt\index{Egypt ``f10!}, out of the house of bondage\index{bondage, slavery in Egypt ``f10!}.\footnote{\ Exodus
20:2\index{02 Exodus:ch.  20, v.   2, pg. ``f30!}.} Wherein God manifests his sovereignty\index{God's
attributes:sovereign lordship ``f10!}, as being JEHOVAH\index{God's name:JEHOVAH ``f10!}, the eternal,\index{God's
attributes:eternal ``f10!} immutable\index{God's attributes:immutable ``f10!}, and almighty\index{God's
attributes:almighty power ``f10!} God;\footnote{\ Isaiah 44:6\index{23 Isaiah:ch.  44, v.   6, pg. ``f30!}.} having his
being in and of himself,\footnote{\ Exodus 3:14\index{02 Exodus:ch.   3, v.  14, pg. ``f30!}.} and giving being to all
his words\footnote{\ Exodus 6:3\index{02 Exodus:ch.   6, v.   3, pg. ``f30!}.} and works\index{Providence:God's works
``f10!}:\footnote{\ Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}, 28\index{44 Acts:ch.  17, v.  24, 28, pg.
``f30!}.} and that he is a God in covenant, as\index{Covenant:administrations ``f10!} with Israel\index{Israel ``f10!}
of old, so with all his people;\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}; Romans
3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}9\index{45 Romans:ch.   3, v.  29, pg. ``f30!}.} who, as he brought
them out of their bondage in Egypt, so he delivers us from our spiritual thraldom\index{Salvation:delivered from
thraldom to sin ``f10!};\footnote{\ Luke 1:74-75\index{42 Luke:ch.   1, v.  74-75, pg. ``f30!}.} and that therefore we
are bound to take him for our God alone, and to keep all his commandments\index{God's
law:commandments:obedience:required ``f10!}.\footnote{\ 1 Peter 1:15-18\index{60 1 Peter:ch.   1, v.  15-18, pg.
``f30!}; Leviticus 18:30\index{03 Leviticus:ch.  18, v.  30, pg. ``f30!}; Leviticus 19:3\index{03 Leviticus:ch.  19, v.
  3, pg. ``f30!}7\index{03 Leviticus:ch.  19, v.  37, pg. ``f30!}.}


\bigskip

Q. 102 \textit{What is the sum of the four commandments}\index{God's law:commandments:first four, duty to God
``f10!}\textit{ which contain our duty to God}\index{God's law:commandments:duties:required by God ``f10!}\textit{?}

A. The sum of the four commandments\index{God's law:commandments:first four, duty to God ``f10!} containing our duty to
God\index{God's law:commandments:duties:required by God ``f10!} is, to love the Lord\index{Love toward God:for God
``f10!} our God with all our heart, and with all our soul\index{Love toward God:with all our:soul ``f10!}, and with all
our strength\index{Love toward God:with all our:strength ``f10!}, and with all our mind.\footnote{\ Luke 10:27\index{42
Luke:ch.  10, v.  27, pg. ``f30!}.}


\bigskip

Q. 103 \textit{Which is the first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The first commandment\index{God's law:commandments:first ``f10!} is, You shall have no other
gods\index{Idolatry:worshiping gods ``f10!} before me.\footnote{\ Exodus 20:3\index{02 Exodus:ch.  20, v.   3, pg.
``f30!}.}


\bigskip

Q. 104 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the first commandment\index{God's
law:commandments:first ``f10!} are, the knowing and acknowledging of God\index{Worship:acknowledge God ``f10!} to be
the only true God, and our God;\footnote{\ 1 Chronicles 28:9\index{13 1 Chronicles:ch.  28, v.   9, pg. ``f30!};
Deuteronomy 26:17\index{05 Deuteronomy:ch.  26, v.  17, pg. ``f30!}; Isaiah 43:10\index{23 Isaiah:ch.  43, v.  10, pg.
``f30!}; Jeremiah 14:22\index{24 Jeremiah:ch.  14, v.  22, pg. ``f30!}.} and to worship and\index{Worship:glorifying
God ``f10!} glorify him accordingly,\footnote{\ Psalm 29:2\index{19 Psalms:ch.  29, v.   2, pg. ``f30!}; Psalm
95:6\index{19 Psalms:ch.  95, v.   6, pg. ``f30!}{}-7\index{19 Psalms:ch.  95, v.   6-7, pg. ``f30!}; Matthew
4:10\index{40 Matthew:ch.   4, v.  10, pg. ``f30!}.} by thinking\index{Worship:affections directed to God
``f10!},\footnote{\ Malachi 3:16\index{39 Malachi:ch.   3, v.  16, pg. ``f30!}.} meditating\index{meditation
``f10!},\footnote{\ Psalm 63:6\index{19 Psalms:ch.  63, v.   6, pg. ``f30!}.} remembering,\footnote{\ Ecclesiastes
12:1\index{21 Ecclesiastes:ch.  12, v.   1, pg. ``f30!}.} highly esteeming\index{Worship:esteem God
``f10!},\footnote{\ Psalm 71:19\index{19 Psalms:ch.  71, v.  19, pg. ``f30!}.} honoring\index{God's attributes:to
be:honoured ``f10!},\footnote{\ Malachi 1:6\index{39 Malachi:ch.   1, v.   6, pg. ``f30!}.}
adoring\index{Worship:adoration of God ``f10!},\footnote{\ Isaiah 45:23\index{23 Isaiah:ch.  45, v.  23, pg. ``f30!}.}
choosing,\index{Love toward God:for God ``f10!}\footnote{\ Joshua 24:15\index{06 Joshua:ch.  24, v.  15, pg. ``f30!},
22\index{06 Joshua:ch.  24, v.  15, 22, pg. ``f30!}.} loving,\footnote{\ Deuteronomy 6:5\index{05 Deuteronomy:ch.   6,
v.   5, pg. ``f30!}.} desiring,\footnote{\ Psalm 73:25\index{19 Psalms:ch.  73, v.  25, pg. ``f30!}.}
fearing\index{Fear:of God:reverence ``f10!} of him;\footnote{\ Isaiah 8:13\index{23 Isaiah:ch.   8, v.  13, pg.
``f30!}.} believing him\index{Faith:belief to be in God alone ``f10!};\footnote{\ Exodus 14:31\index{02 Exodus:ch.  14,
v.  31, pg. ``f30!}.} trusting,\footnote{\ Isaiah 26:4\index{23 Isaiah:ch.  26, v.   4, pg. ``f30!}.}
hoping\index{Hope:in God ``f10!},\footnote{\ Psalm 130:7\index{19 Psalms:ch. 130, v.   7, pg. ``f30!}.}
delighting\index{Worship:delighting in God ``f10!},\footnote{\ Psalm 37:4\index{19 Psalms:ch.  37, v.   4, pg.
``f30!}.} rejoicing in him\index{Worship:rejoice in Christ ``f10!};\footnote{\ Psalm 32:1\index{19 Psalms:ch.  32, v.  
1, pg. ``f30!}1\index{19 Psalms:ch.  32, v.  11, pg. ``f30!}.} being zealous\index{Worship:zeal for God ``f10!} for
him;\footnote{\ Romans 12:11\index{45 Romans:ch.  12, v.  11, pg. ``f30!}; Numbers 25:11\index{04 Numbers:ch.  25, v. 
11, pg. ``f30!}.} calling upon\index{Worship:calling upon God ``f10!} him, giving all praise\index{Worship:praising God
``f10!} and thanks\index{Prayer:thanking God ``f10!},\footnote{\ Philippians 4:6\index{50 Philippians:ch.   4, v.   6,
pg. ``f30!}.} and yielding\index{God's law:commandments:obedience:required ``f10!} all obedience and submission to
him\index{Humility:submission before God ``f10!} with the whole man;\footnote{\ Jeremiah 7:28\index{24 Jeremiah:ch.  
7, v.  28, pg. ``f30!}; James 4:7\index{59 James:ch.   4, v.   7, pg. ``f30!}.} being careful in all things to please
him\index{God's law:commandments:obedience:to please God ``f10!},\footnote{\ 1 John 3:2\index{62 1 John:ch.   3, v.  
2, pg. ``f30!}2\index{62 1 John:ch.   3, v.  22, pg. ``f30!}.} and sorrowful\index{Confession:sorrow for offending God
``f10!} when in anything he is offended;\index{Confession:of sin ``f10!}\footnote{\ Jeremiah 31:18\index{24
Jeremiah:ch.  31, v.  18, pg. ``f30!}; Psalm 119:136\index{19 Psalms:ch. 119, v. 136, pg. ``f30!}.} and walking
humbly\index{Humility:before God ``f10!} with him.\footnote{\ Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}.} 


\bigskip

Q. 105 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the first
commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the first commandment\index{God's
law:commandments:first ``f10!} are, atheism\index{atheism ``f10!}, in denying or\index{Idolatry:denying the true God
``f10!} not having a God;\footnote{\ Psalm 14:1\index{19 Psalms:ch.  14, v.   1, pg. ``f30!}; Ephesians 2:1\index{49
Ephesians:ch.   2, v.   1, pg. ``f30!}2\index{49 Ephesians:ch.   2, v.  12, pg. ``f30!}.}
Idolatry\index{Idolatry:idolatry ``f10!}, in having or worship\index{Worship:of God ``f10!}ing more
gods\index{Idolatry:worshiping gods ``f10!} than one, or any with or instead of the true God;\footnote{\ Jeremiah
2:27-28\index{24 Jeremiah:ch.   2, v.  27-28, pg. ``f30!}; 1 Thessalonians 1:9\index{52 1 Thessalonians:ch.   1, v.  
9, pg. ``f30!}.} the not having and avouching\index{Idolatry:not avouching God as God ``f10!} him for God, and our
God;\footnote{\ Psalm 81:11\index{19 Psalms:ch.  81, v.  11, pg. ``f30!}.} the omission\index{Sin:omission of duties
``f10!} or neglect of anything\index{Worship:not to be:neglected ``f10!} due to him, required in this
commandment;\footnote{\ Isaiah 43:22-24\index{23 Isaiah:ch.  43, v.  22-24, pg. ``f30!}.} ignorance\index{ignorance
``f10!},\footnote{\ Jeremiah 4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}2\index{24 Jeremiah:ch.   4, v.  22, pg.
``f30!}; Hosea 4:1, 6\index{28 Hosea:ch.   4, v.   1, 6, pg. ``f30!}.} forgetfulness\index{Idolatry:forgetting God
``f10!},\footnote{\ Jeremiah 2:32\index{24 Jeremiah:ch.   2, v.  32, pg. ``f30!}} misapprehensions,\footnote{\ Acts
17:23\index{44 Acts:ch.  17, v.  23, pg. ``f30!}, 29\index{44 Acts:ch.  17, v.  23, 29, pg. ``f30!}.} false
opinions\index{Falsehood:false opinions ``f10!},\footnote{\ Isaiah 40:18\index{23 Isaiah:ch.  40, v.  18, pg. ``f30!}.}
unworthy and\index{Idolatry:unworthy and wicked thoughts about God ``f10!} wicked thoughts of\index{Idolatry:wicked
thoughts about God ``f10!} him;\footnote{\ Psalm 50:21\index{19 Psalms:ch.  50, v.  21, pg. ``f30!}.} bold and curious
searching\index{Providence:not to search into God's secrets ``f10!} into his secrets;\footnote{\ Deuteronomy
29:29\index{05 Deuteronomy:ch.  29, v.  29, pg. ``f30!}.} all profaneness\index{God's name:not to be:profaned
``f10!},\footnote{\ Titus 1:16\index{56 Titus:ch.   1, v.  16, pg. ``f30!}; Hebrews 12:1\index{58 Hebrews:ch.  12, v.  
1, pg. ``f30!}6\index{58 Hebrews:ch.  12, v.  16, pg. ``f30!}.} hatred of God\index{Sin:hatred of God
``f10!};\footnote{\ Romans 1:30\index{45 Romans:ch.   1, v.  30, pg. ``f30!}.} self-love\index{self-love
``f10!},\footnote{\ 2 Timothy 3:2\index{55 2 Timothy:ch.   3, v.   2, pg. ``f30!}.} self-seeking\index{self-seeking
``f10!},\footnote{\ Philippians 2:21\index{50 Philippians:ch.   2, v.  21, pg. ``f30!}.} and all other inordinate and
immoderate setting\index{Idolatry:immoderate setting of mind on other than God ``f10!} of our
mind\index{Idolatry:inwardly in the mind ``f10!}, will, or affections upon\index{Worship:false:affections on other
things ``f10!} other things, and taking them off from him in whole or in part;\footnote{\ 1 John 2:15-16\index{62 1
John:ch.   2, v.  15-16, pg. ``f30!}; 1 Samuel 2:29\index{09 1 Samuel:ch.   2, v.  29, pg. ``f30!}; Colossians
3:2\index{51 Colossians:ch.   3, v.   2, pg. ``f30!}, 5\index{51 Colossians:ch.   3, v.   2, 5, pg. ``f30!}.} vain
credulity\index{credulity, vain ``f10!},\footnote{\ 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg. ``f30!}.}
unbelief\index{unbelief (misbelief) ``f10!},\footnote{\ Hebrews 3:12\index{58 Hebrews:ch.   3, v.  12, pg. ``f30!}.}
heresy\index{Heresy:heresies ``f10!},\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg.
``f30!}0\index{48 Galatians:ch.   5, v.  20, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg.
``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!}.} misbelief,\footnote{\ Acts 26:9\index{44 Acts:ch.  26, v.   9,
pg. ``f30!}.} distrust\index{Idolatry:distrust ``f10!},\footnote{\ Psalm 78:22\index{19 Psalms:ch.  78, v.  22, pg.
``f30!}.} despair\index{despair ``f10!},\footnote{\ Genesis 4:13\index{01 Genesis:ch.   4, v.  13, pg. ``f30!}.}
incorrigibleness\index{Idolatry:incorrigibleness ``f10!},\footnote{\ Jeremiah 5:3\index{24 Jeremiah:ch.   5, v.   3,
pg. ``f30!}.} and insensibleness\index{Unbelievers:insensible under judgement ``f10!} under
judgments\index{Judgement:temporal ``f10!},\footnote{\ Isaiah 42:25\index{23 Isaiah:ch.  42, v.  25, pg. ``f30!}.}
hardness of heart\index{Heart:hardened ``f10!},\footnote{\ Romans 2:5\index{45 Romans:ch.   2, v.   5, pg. ``f30!}.}
pride\index{pride ``f10!},\footnote{\ Jeremiah 13:15\index{24 Jeremiah:ch.  13, v.  15, pg. ``f30!}.}
presumption\index{presumption ``f10!},\footnote{\ Psalm 19:13\index{19 Psalms:ch.  19, v.  13, pg. ``f30!}.} carnal
security\index{Idolatry:carnal:security ``f10!},\footnote{\ Zephaniah 1:12\index{36 Zephaniah:ch.   1, v.  12, pg.
``f30!}.} tempting\index{tempting God ``f10!} of God;\footnote{\ Matthew 4:7\index{40 Matthew:ch.   4, v.   7, pg.
``f30!}.} using unlawful\index{Idolatry:using unlawful means ``f10!} means,\footnote{\ Romans 3:8\index{45 Romans:ch.  
3, v.   8, pg. ``f30!}.} and trusting\index{Idolatry:trust in lawful means ``f10!} in lawful
means\index{Worship:false:corruptions in ``f10!};\footnote{\ Jeremiah 17:5\index{24 Jeremiah:ch.  17, v.   5, pg.
``f30!}.} carnal delights\index{Idolatry:carnal:delights ``f10!} and joys;\index{Joy:corrupt ``f10!}\footnote{\ 2
Timothy 3:4\index{55 2 Timothy:ch.   3, v.   4, pg. ``f30!}.} corrupt, blind, and\index{Good works:not out of blind
zeal ``f10!} indiscreet zeal\index{Idolatry:indiscreet zeal ``f10!};\footnote{\ Galatians 4:17\index{48 Galatians:ch.  
4, v.  17, pg. ``f30!}; John 16:2\index{43 John:ch.  16, v.   2, pg. ``f30!}; Romans 10:2\index{45 Romans:ch.  10, v.  
2, pg. ``f30!}; Luke 9:54-55\index{42 Luke:ch.   9, v.  54-55, pg. ``f30!}.} lukewarmness\index{lukewarmness
``f10!},\footnote{\ Revelation 3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}6\index{66 Revelation:ch.   3, v. 
16, pg. ``f30!}.} and deadness\index{Death:spiritual, in sin ``f10!} in the things of God;\footnote{\ Revelation
3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}.} estranging\index{Idolatry:estranging self from God ``f10!}
ourselves, and apostatizing\index{apostasy ``f10!} from God;\footnote{\ Ezekiel 14:5\index{26 Ezekiel:ch.  14, v.   5,
pg. ``f30!}; Isaiah 1:4-5\index{23 Isaiah:ch.   1, v.   4-5, pg. ``f30!}.} praying, or\index{Prayer:not to saints
``f10!} giving any religious worship, to\index{Worship:false:of saints ``f10!} saints\index{Believers:saints ``f10!},
angels\index{Angels:angels ``f10!}, or any other creatures\index{creatures ``f10!};\footnote{\ Romans 1:25\index{45
Romans:ch.   1, v.  25, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}3-14\index{45 Romans:ch. 
10, v.  13-14, pg. ``f30!}; Hosea 4:12\index{28 Hosea:ch.   4, v.  12, pg. ``f30!}; Acts 10:2\index{44 Acts:ch.  10, v.
  2, pg. ``f30!}5-26\index{44 Acts:ch.  10, v.  25-26, pg. ``f30!}; Revelation 19:10\index{66 Revelation:ch.  19, v. 
10, pg. ``f30!}; Matthew 4:10\index{40 Matthew:ch.   4, v.  10, pg. ``f30!}; Colossians 2:18\index{51 Colossians:ch.  
2, v.  18, pg. ``f30!}.} all compacts\index{Devil:compacts with ``f10!} and consulting\index{Devil:consulting with
``f10!} with the devil\index{Devil:devil ``f10!},\footnote{\ Leviticus 20:6\index{03 Leviticus:ch.  20, v.   6, pg.
``f30!}; 1 Samuel 28:7, 11\index{09 1 Samuel:ch.  28, v.   7, 11, pg. ``f30!}; 1 Chronicles 10:13-14\index{13 1
Chronicles:ch.  10, v.  13-14, pg. ``f30!}.} and hearkening\index{Temptation:hearkening to devil's suggestions ``f10!}
to his suggestions;\footnote{\ Acts 5:3\index{44 Acts:ch.   5, v.   3, pg. ``f30!}.} making men the
lords\index{Idolatry:men lords or our faith ``f10!} of our faith and conscience\index{Conscience:influenced falsely by
others ``f10!};\footnote{\ 2 Corinthians 1:24\index{47 2 Corinthians:ch.   1, v.  24, pg. ``f30!}; Matthew
23:9\index{40 Matthew:ch.  23, v.   9, pg. ``f30!}.} slighting\index{Sin:despising and slighting God ``f10!} and
despising\index{Idolatry:despising and slighting God ``f10!} God and his commands\index{Sin:despising God's commands
``f10!};\footnote{\ Deuteronomy 32:15\index{05 Deuteronomy:ch.  32, v.  15, pg. ``f30!}; 2 Samuel 12:9\index{10 2
Samuel:ch.  12, v.   9, pg. ``f30!}; Proverbs 13:13\index{20 Proverbs:ch.  13, v.  13, pg. ``f30!}.} resisting
and\index{Idolatry:resisting Holy Spirit ``f10!} grieving of\index{Sin:grieves the Holy Spirit ``f10!} his
Spirit,\footnote{\ Acts 7:51\index{44 Acts:ch.   7, v.  51, pg. ``f30!}; Ephesians 4:3\index{49 Ephesians:ch.   4, v.  
3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}.} discontent\index{discontentment ``f10!} and
impatience\index{impatience, with God's dispositions ``f10!} at his dispensations\index{Providence:impatience with
God's dispensations ``f10!}, charging him foolishly for the evil\index{Christian liberty:deliverance from evil world
``f10!}s he inflicts\index{Idolatry:charging God for evils he inflicts on us ``f10!} on us;\footnote{\ Psalm 73:2-3,
13-15, 22\index{19 Psalms:ch.  73, v.   2-3, 13-15, 22, pg. ``f30!}; Job 1:22\index{18 Job:ch.   1, v.  22, pg.
``f30!}.} and ascribing the\index{Worship:false:ascribing praise to other than God ``f10!} praise of
any\index{Idolatry:praise of anything created ``f10!} good we either are, have, or can do, to
fortune\index{Idolatry:fortune, praise of ``f10!},\footnote{\ 1 Samuel 6:7-9\index{09 1 Samuel:ch.   6, v.   7-9, pg.
``f30!}.} idols,\footnote{\ Daniel 5:23\index{27 Daniel:ch.   5, v.  23, pg. ``f30!}.}
ourselves,\footnote{\ Deuteronomy 8:17\index{05 Deuteronomy:ch.   8, v.  17, pg. ``f30!}; Daniel 4:30\index{27
Daniel:ch.   4, v.  30, pg. ``f30!}.} or any other creature.\footnote{\ Habakkuk 1:16\index{35 Habakkuk:ch.   1, v. 
16, pg. ``f30!}.}


\bigskip

Q. 106 \textit{What are we specially taught by}\index{God's law:commandments:first, words 'before me' ``f10!}\textit{
these words before me in the first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. These words before me, or before my face, in\index{God's attributes:omniscience ``f10!} the first
commandment\index{God's law:commandments:first ``f10!}, teach us, that\index{God's law:commandments:first, words
'before me' ``f10!} God, who sees all things, takes special notice\index{Idolatry:God takes special notice of this sin
``f10!} of, and is much displeased\index{Idolatry:displeases God greatly ``f10!} with, the sin of having any other God:
that so it may be an argument to dissuade from it, and to aggravate it as a most impudent\index{impudence ``f10!}
provocation\index{Idolatry:impudent provocation of God ``f10!}:\footnote{\ Ezekiel 8:5-18\index{26 Ezekiel:ch.   8, v. 
 5-18, pg. ``f30!}; Psalm 44:20-21\index{19 Psalms:ch.  44, v.  20-21, pg. ``f30!}.} as also to persuade us to do as in
his sight, whatever we do in his service\index{Love toward God:in service ``f10!}.\footnote{\ 1 Chronicles
28:9\index{13 1 Chronicles:ch.  28, v.   9, pg. ``f30!}.}


\bigskip

Q. 107 \textit{Which is the second commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The second commandment\index{God's law:commandments:second ``f10!} is, You shalt not make unto you any graven image,
or\index{Idolatry:graven image ``f10!} any likeness\index{Idolatry:worship of likeness of any created thing ``f10!} of
anything that is in heaven above, or that is in the earth beneath, or that is in the water under\index{Idolatry:bowing
to created entity ``f10!} the earth: You shalt not bow down thyself to them, nor serve the\index{Love toward God:in
service ``f10!}m\index{Idolatry:serve images ``f10!}: for I the Lord thy God am a jealous\index{God's
attributes:jealous ``f10!} God, visiting\index{Punishment:visiting iniquity ``f10!} the iniquity of the fathers upon
the children unto the third and\index{Punishment:to third and fourth generation ``f10!} fourth generation of them that
hate\index{Sin:hatred of God ``f10!} me; and shewing mercy unto\index{God's attributes:merciful ``f10!} thousands of
them that love me\index{Love toward God:for God ``f10!}, and keep my commandments\index{God's law:commandments:ten
``f10!}.\footnote{\ Exodus 20:4-6\index{02 Exodus:ch.  20, v.   4-6, pg. ``f30!}.}


\bigskip

Q. 108 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
second commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the second commandment\index{God's
law:commandments:second ``f10!} are, the receiving, observing, and keeping pure\index{Worship:to be observed and kept
pure ``f10!} and entire, all such religious worship and ordinances as\index{Worship:ordinances:to be:kept pure ``f10!}
God has instituted in his word\index{Worship:correct:instituted by God ``f10!};\footnote{\ Deuteronomy 32:4\index{05
Deuteronomy:ch.  32, v.   4, pg. ``f30!}6-47\index{05 Deuteronomy:ch.  32, v.  46-47, pg. ``f30!}; Matthew
28:20\index{40 Matthew:ch.  28, v.  20, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}; 1 Timothy
6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}3-14\index{54 1 Timothy:ch.   6, v.  13-14, pg. ``f30!}.}
particularly prayer\index{Prayer:thanking God ``f10!} and thanksgiving in the name of Christ;\footnote{\ Philippians
4:6\index{50 Philippians:ch.   4, v.   6, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg.
``f30!}0\index{49 Ephesians:ch.   5, v.  20, pg. ``f30!}.} the reading\index{Scripture:to be:read and heard ``f10!},
preaching, and\index{Preaching:ministry of the Word ``f10!} hearing of the word;\footnote{\ Deuteronomy
17:18-19\index{05 Deuteronomy:ch.  17, v.  18-19, pg. ``f30!}; Acts 15:2\index{44 Acts:ch.  15, v.   2, pg.
``f30!}1\index{44 Acts:ch.  15, v.  21, pg. ``f30!}; 2 Timothy 4:2\index{55 2 Timothy:ch.   4, v.   2, pg. ``f30!};
James 1:21\index{59 James:ch.   1, v.  21, pg. ``f30!}{}-22\index{59 James:ch.   1, v.  21-22, pg. ``f30!}.} the
administration and\index{Worship:administration of sacraments ``f10!} receiving of the
sacraments\index{Sacraments:holy:ordinances ``f10!};\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg.
``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v. 
23, pg. ``f30!}{}-30\index{46 1 Corinthians:ch.  11, v.  23-30, pg. ``f30!}.} church government\index{Church
government:government ``f10!} and discipline\index{Church discipline:discipline ``f10!};\footnote{\ Matthew
16:19\index{40 Matthew:ch.  16, v.  19, pg. ``f30!}; Matthew 18:15\index{40 Matthew:ch.  18, v.  15, pg.
``f30!}{}-17\index{40 Matthew:ch.  18, v.  15-17, pg. ``f30!}; 1 Corinthians 5; 1 Corinthians 12:28\index{46 1
Corinthians:ch.  12, v.  28, pg. ``f30!}.} the ministry and maintenance thereof\index{Preachers:to be:maintained
``f10!};\footnote{\ Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-12\index{49 Ephesians:ch.   4, v. 
11-12, pg. ``f30!}; 1 Timothy 5:1\index{54 1 Timothy:ch.   5, v.   1, pg. ``f30!}7\index{54 1 Timothy:ch.   5, v.  17,
pg. ``f30!}{}-18\index{54 1 Timothy:ch.   5, v.  17-18, pg. ``f30!}; 1 Corinthians 9:1-15\index{46 1 Corinthians:ch.  
9, v.   1-15, pg. ``f30!}.} religious fasting\index{fasting ``f10!};\footnote{\ Joel 2:12\index{29 Joel:ch.   2, v. 
12, pg. ``f30!}{}-13\index{29 Joel:ch.   2, v.  12-13, pg. ``f30!}; 1 Corinthians 7:5\index{46 1 Corinthians:ch.   7,
v.   5, pg. ``f30!}.} swearing\index{Oaths and Vows:swear only by God's name ``f10!} by the name of
God,\footnote{\ Deuteronomy 6:13\index{05 Deuteronomy:ch.   6, v.  13, pg. ``f30!}.} and vowing\index{Oaths and
Vows:vowing only by God's name ``f10!} unto him:\footnote{\ Isaiah 19:21\index{23 Isaiah:ch.  19, v.  21, pg. ``f30!};
Psalm 76:11\index{19 Psalms:ch.  76, v.  11, pg. ``f30!}.} as also the disapproving\index{Worship:false:to
be:disapproved ``f10!}, detesting\index{Worship:false:to be:detested ``f10!}, opposing, all\index{Worship:false:to
be:opposed ``f10!} false worship;\footnote{\ Acts 17:16-17\index{44 Acts:ch.  17, v.  16-17, pg. ``f30!}; Psalm
16:4\index{19 Psalms:ch.  16, v.   4, pg. ``f30!}.} and, according to each one's place and calling\index{Calling:in
life ``f10!}, removing it\index{Civil magistrate:to remove false worship and idolatry ``f10!}, and all
monuments\index{Idolatry:monuments to be removed ``f10!} of idolatry\index{Idolatry:idolatry
``f10!}.\footnote{\ Deuteronomy 7:5\index{05 Deuteronomy:ch.   7, v.   5, pg. ``f30!}; Isaiah 30:22\index{23 Isaiah:ch.
 30, v.  22, pg. ``f30!}.} 

Q. 109 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the second
commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the second commandment\index{God's
law:commandments:second ``f10!} are, all devising\index{Worship:false:corruptions in ``f10!},\footnote{\ Numbers
15:39\index{04 Numbers:ch.  15, v.  39, pg. ``f30!}.} counseling,\footnote{\ Deuteronomy 13:6-8\index{05
Deuteronomy:ch.  13, v.   6-8, pg. ``f30!}.} commanding\index{Sin:commanding things unlawful ``f10!},\footnote{\ Hosea
5:11\index{28 Hosea:ch.   5, v.  11, pg. ``f30!}; Micah 6:16\index{33 Micah:ch.   6, v.  16, pg. ``f30!}.}
using,\footnote{\ 1 Kings 11:33\index{11 1 Kings:ch.  11, v.  33, pg. ``f30!}; 1 Kings 12:33\index{11 1 Kings:ch.  12,
v.  33, pg. ``f30!}.} and anywise approving\index{Worship:only acts approved by God ``f10!}, any religious worship
not\index{Worship:correct:instituted by God ``f10!} instituted by God\index{Sacraments:instituted by:God ``f10!}
himself;\footnote{\ Deuteronomy 12:30-32\index{05 Deuteronomy:ch.  12, v.  30-32, pg. ``f30!}.} tolerating
a\index{Idolatry:tolerating a false religion ``f10!} false religion\index{Profession:of true religion
``f10!};\footnote{\ Deuteronomy 13:6-12\index{05 Deuteronomy:ch.  13, v.   6-12, pg. ``f30!}; Zechariah 13:2-3\index{38
Zechariah:ch.  13, v.   2-3, pg. ``f30!}; Revelation 2:2, 14-15, 20\index{66 Revelation:ch.   2, v.   2, 14-15, 20, pg.
``f30!}; Revelation 17:12, 16-17\index{66 Revelation:ch.  17, v.  12, 16-17, pg. ``f30!}.} the making any
representation\index{Idolatry:any visible representation of God or gods ``f10!} of God, of all or of any of the three
persons\index{God's attributes:trinity:three persons ``f10!}, either inwardly in\index{Idolatry:inwardly in the mind
``f10!} our mind, or outwardly in\index{Idolatry:outwardly with images or likenesses ``f10!} any kind of image
or\index{Idolatry:graven image ``f10!} likeness\index{Idolatry:worship of likeness of any created thing ``f10!} of any
creature\index{creatures ``f10!} whatsoever;\footnote{\ Deuteronomy 4:15-19\index{05 Deuteronomy:ch.   4, v.  15-19,
pg. ``f30!}; Acts 17:29\index{44 Acts:ch.  17, v.  29, pg. ``f30!}; Romans 1:21-23, 25\index{45 Romans:ch.   1, v. 
21-23, 25, pg. ``f30!}.} all worshiping of it,\footnote{\ Daniel 3:18\index{27 Daniel:ch.   3, v.  18, pg. ``f30!};
Galatians 4:8\index{48 Galatians:ch.   4, v.   8, pg. ``f30!}.} or God in it or by it;\footnote{\ Exodus 32:5\index{02
Exodus:ch.  32, v.   5, pg. ``f30!}.} the making of any representation of feigned deities\index{Idolatry:deities
``f10!},\footnote{\ Exodus 32:8\index{02 Exodus:ch.  32, v.   8, pg. ``f30!}.} and all worship of them, or service
belonging\index{Idolatry:serve images ``f10!} to them;\footnote{\ 1 Kings 18:26, 28\index{11 1 Kings:ch.  18, v.  26,
28, pg. ``f30!}; Isaiah 65:11\index{23 Isaiah:ch.  65, v.  11, pg. ``f30!}.} all
superstitious\index{Superstitions:superstitions ``f10!} devices\index{Worship:false:imaginations and devices of men
``f10!},\footnote{\ Acts 17:22\index{44 Acts:ch.  17, v.  22, pg. ``f30!}; Colossians 2:2\index{51 Colossians:ch.   2,
v.   2, pg. ``f30!}1-23\index{51 Colossians:ch.   2, v.  21-23, pg. ``f30!}.} corrupting the worship of
God,\footnote{\ Malachi 1:7-8, 14\index{39 Malachi:ch.   1, v.   7-8, 14, pg. ``f30!}.} adding\index{Worship:not to
be:corrupted with additions ``f10!} to it, or taking from it,\footnote{\ Deuteronomy 4:2\index{05 Deuteronomy:ch.   4,
v.   2, pg. ``f30!}.} whether invented\index{Worship:false:inventions of men ``f10!} and taken up of
ourselves,\footnote{\ Psalm 106:39\index{19 Psalms:ch. 106, v.  39, pg. ``f30!}.} or received by
tradition\index{Worship:false:traditions of men ``f10!} from others,\footnote{\ Matthew 15:9\index{40 Matthew:ch.  15,
v.   9, pg. ``f30!}.} though under the title of antiquity,\footnote{\ 1 Peter 1:18\index{60 1 Peter:ch.   1, v.  18,
pg. ``f30!}.} custom\index{Worship:false:customs ``f10!},\footnote{\ Jeremiah 44:17\index{24 Jeremiah:ch.  44, v.  17,
pg. ``f30!}.} devotion\index{Worship:false:mere devotion or good intent ``f10!},\footnote{\ Isaiah 65:3-5\index{23
Isaiah:ch.  65, v.   3-5, pg. ``f30!}; Galatians 1:13-14\index{48 Galatians:ch.   1, v.  13-14, pg. ``f30!}.} good
intent\index{Worship:not defined by good intentions ``f10!}, or any other pretense\index{Idolatry:human invention
``f10!} whatsoever;\footnote{\ 1 Samuel 13:11-12\index{09 1 Samuel:ch.  13, v.  11-12, pg. ``f30!}; 1 Samuel
15:21\index{09 1 Samuel:ch.  15, v.  21, pg. ``f30!}.} simony\index{simony ``f10!};\footnote{\ Acts 8:18\index{44
Acts:ch.   8, v.  18, pg. ``f30!}.} sacrilege\index{Worship:false:sacrilege ``f10!};\footnote{\ Romans 2:22\index{45
Romans:ch.   2, v.  22, pg. ``f30!}; Malachi 3:8\index{39 Malachi:ch.   3, v.   8, pg. ``f30!}.} all
neglect,\footnote{\ Exodus 4:24-26\index{02 Exodus:ch.   4, v.  24-26, pg. ``f30!}.} contempt,\footnote{\ Matthew
22:5\index{40 Matthew:ch.  22, v.   5, pg. ``f30!}; Malachi 1:7, 13\index{39 Malachi:ch.   1, v.   7, 13, pg. ``f30!}.}
hindering,\index{Contempt:for proper worship ``f10!}\footnote{\ Matthew 23:13\index{40 Matthew:ch.  23, v.  13, pg.
``f30!}.} and opposing the\index{Worship:not to oppose what God has appointed ``f10!} worship and ordinances which
God\index{Worship:ordinances:reverent use of ``f10!} has appointed.\footnote{\ Acts 13:44-45\index{44 Acts:ch.  13, v. 
44-45, pg. ``f30!}; 1 Thessalonians 2:15-16\index{52 1 Thessalonians:ch.   2, v.  15-16, pg. ``f30!}.}


\bigskip

Q. 110 \textit{What are the reasons annexed to the second commandment}\index{God's law:commandments:second
``f10!}\textit{, the more to enforce it}\index{God's law:commandments:words of enforcement ``f10!}\textit{?}

A. The reasons annexed to the second commandment\index{God's law:commandments:second ``f10!}, the more to enforce
it\index{God's law:commandments:words of enforcement ``f10!}, contained in these words, For I the Lord thy God am a
jealous\index{God's attributes:jealous ``f10!} God, visiting\index{Punishment:visiting iniquity ``f10!} the iniquity of
the fathers upon the children unto the third and\index{Punishment:to third and fourth generation ``f10!} fourth
generation of them that hate\index{Sin:hatred of God ``f10!} me; and showing mercy unto\index{God's attributes:merciful
``f10!} thousands of them that love me\index{Love toward God:for God ``f10!}, and keep my commandments\index{God's
law:commandments:ten ``f10!};\footnote{\ Exodus 20:5-6\index{02 Exodus:ch.  20, v.   5-6, pg. ``f30!}.} are, besides
God's sovereignty\index{God's attributes:sovereign lordship ``f10!} over us, and propriety in us\index{Providence:God's
propriety over mankind ``f10!},\footnote{\ Psalm 45:11\index{19 Psalms:ch.  45, v.  11, pg. ``f30!}; Revelation
20:3-4\index{66 Revelation:ch.  20, v.   3-4, pg. ``f30!}.} his fervent zeal for\index{Worship:God's zeal for proper
``f10!} his own worship\index{Worship:of God ``f10!},\footnote{\ Exodus 34:1\index{02 Exodus:ch.  34, v.   1, pg.
``f30!}3-14\index{02 Exodus:ch.  34, v.  13-14, pg. ``f30!}.} and his revengeful\index{Revenge:God's ``f10!}
indignation\index{Worship:false:indignation of God against ``f10!} against all false worship, as
being\index{Worship:false:spiritual whoredom ``f10!} a spiritual whoredom\index{Idolatry:spiritual whoredom
``f10!};\footnote{\ 1 Corinthians 10:20-22\index{46 1 Corinthians:ch.  10, v.  20-22, pg. ``f30!}; Jeremiah
7:18-20\index{24 Jeremiah:ch.   7, v.  18-20, pg. ``f30!}; Ezekiel 16:26-27\index{26 Ezekiel:ch.  16, v.  26-27, pg.
``f30!}; Deuteronomy 32:16-20\index{05 Deuteronomy:ch.  32, v.  16-20, pg. ``f30!}.} accounting the breakers of this
commandment such as hate him, and threatening to\index{God's law:commandments:threatenings for disobedience ``f10!}
punish\index{Punishment:for idolatry ``f10!} them unto divers generations;\footnote{\ Hosea 2:2-4\index{28 Hosea:ch.  
2, v.   2-4, pg. ``f30!}.} and esteeming\index{Love for others:mutual:esteem ``f10!} the observers of it such as love
him and keep his commandments\index{God's law:commandments:obedience:endeavoring ``f10!}, and promising mercy to them
unto many generations.\footnote{\ Deuteronomy 5:29\index{05 Deuteronomy:ch.   5, v.  29, pg. ``f30!}.}


\bigskip

Q. 111 \textit{Which is the third commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The third commandment\index{God's law:commandments:third ``f10!} is, You shalt not take the name of the Lord thy God
in vain\index{God's name:not to be:taken in vain ``f10!}: for the Lord will not hold him guiltless\index{God's name:not
guiltless if take in vain ``f10!} that takes his name in vain.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.   7,
pg. ``f30!}.}


\bigskip

Q. 112 \textit{What is required in the third commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The third commandment\index{God's law:commandments:third ``f10!} requires, That the name of God, his
titles\index{God's name:titles to be honoured ``f10!}, attributes\index{God's attributes:to be:revered
``f10!},\footnote{\ Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}; Deuteronomy 28:58\index{05
Deuteronomy:ch.  28, v.  58, pg. ``f30!}; Psalm 29:2\index{19 Psalms:ch.  29, v.   2, pg. ``f30!}; Psalm 68:4\index{19
Psalms:ch.  68, v.   4, pg. ``f30!}; Revelation 15:3-4\index{66 Revelation:ch.  15, v.   3-4, pg. ``f30!}.}
ordinances,\footnote{\ Malachi 1:14\index{39 Malachi:ch.   1, v.  14, pg. ``f30!}; Ecclesiastes 5:1\index{21
Ecclesiastes:ch.   5, v.   1, pg. ``f30!}.} the word,\footnote{\ Psalm 138:2\index{19 Psalms:ch. 138, v.   2, pg.
``f30!}.} sacraments\index{Sacraments:holy:ordinances ``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.
 11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v. 
24-25, pg. ``f30!}, 28-29\index{46 1 Corinthians:ch.  11, v.  24-25, 28-29, pg. ``f30!}.} prayer,\footnote{\ 1 Timothy
2:8\index{54 1 Timothy:ch.   2, v.   8, pg. ``f30!}.} oaths\index{Oaths and Vows:oaths ``f10!},\footnote{\ Jeremiah
4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}.} vows,\footnote{\ Ecclesiastes 5:2, 4-6\index{21 Ecclesiastes:ch.  
5, v.   2, 4-6, pg. ``f30!}.} lots,\footnote{\ Acts 1:24\index{44 Acts:ch.   1, v.  24, pg. ``f30!}, 26\index{44
Acts:ch.   1, v.  24, 26, pg. ``f30!}.} his works,\footnote{\ Job 36:24\index{18 Job:ch.  36, v.  24, pg. ``f30!}.} and
whatsoever else there is whereby he makes himself known, be holily\index{God's name:to be used:in holy and reverent way
``f10!} and reverently\index{God's name:to be used:reverently in thoughts and words ``f10!} used in
thought,\footnote{\ Malachi 3:16\index{39 Malachi:ch.   3, v.  16, pg. ``f30!}.} meditation\index{meditation
``f10!},\footnote{\ Psalm 8:1, 3-4, 9\index{19 Psalms:ch.   8, v.   1, 3-4, 9, pg. ``f30!}.}
word,\footnote{\ Colossians 3:17\index{51 Colossians:ch.   3, v.  17, pg. ``f30!}; Psalm 105:2, 5\index{19 Psalms:ch.
105, v.   2, 5, pg. ``f30!}.} and writing;\footnote{\ Psalm 102:18\index{19 Psalms:ch. 102, v.  18, pg. ``f30!}.} by an
holy profession\index{God's name:holy ``f10!},\footnote{\ 1 Peter 3:15\index{60 1 Peter:ch.   3, v.  15, pg. ``f30!};
Micah 4:5\index{33 Micah:ch.   4, v.   5, pg. ``f30!}.} and Answerable conversation\index{Believers:to walk faithfully
in obedience (conversation) ``f10!},\footnote{\ Philippians 1:27\index{50 Philippians:ch.   1, v.  27, pg. ``f30!}.} to
the glory of God,\index{Mankind:to glorify God in thought, word, deed ``f10!}\footnote{\ 1 Corinthians 10:31\index{46 1
Corinthians:ch.  10, v.  31, pg. ``f30!}.} and the good of ourselves,\footnote{\ Jeremiah 32:39\index{24 Jeremiah:ch. 
32, v.  39, pg. ``f30!}.} and others.\footnote{\ 1 Peter 2:12\index{60 1 Peter:ch.   2, v.  12, pg. ``f30!}.}


\bigskip


\bigskip

Q. 113 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the third
commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the third commandment\index{God's
law:commandments:third ``f10!} are, the not using of God's name as is required;\footnote{\ Malachi 2:2\index{39
Malachi:ch.   2, v.   2, pg. ``f30!}.} and the abuse\index{God's name:not to be:abused ``f10!} of it in an
ignorant\index{ignorance ``f10!},\footnote{\ Acts 17:23\index{44 Acts:ch.  17, v.  23, pg. ``f30!}.}
vain,\footnote{\ Proverbs 30:9\index{20 Proverbs:ch.  30, v.   9, pg. ``f30!}.} irreverent, profane\index{God's
name:not to be:profaned ``f10!},\footnote{\ Malachi 1:6\index{39 Malachi:ch.   1, v.   6, pg. ``f30!}{}-7, 12\index{39
Malachi:ch.   1, v.   6-7, 12, pg. ``f30!}; Malachi 3:14\index{39 Malachi:ch.   3, v.  14, pg. ``f30!}.}
superstitious\index{Superstitions:superstitions ``f10!},\footnote{\ 1 Samuel 4:3-5\index{09 1 Samuel:ch.   4, v.   3-5,
pg. ``f30!}; Jeremiah 7:4, 9-10, 14, 31\index{24 Jeremiah:ch.   7, v.   4, 9-10, 14, 31, pg. ``f30!}; Colossians
2:2\index{51 Colossians:ch.   2, v.   2, pg. ``f30!}0-22\index{51 Colossians:ch.   2, v.  20-22, pg. ``f30!}.} or
wicked\index{God's name:not to be:mentioned wickedly ``f10!} mentioning, or otherwise using his titles\index{God's
name:titles to be honoured ``f10!}, attributes\index{God's attributes:to be:revered ``f10!},\footnote{\ 2 Kings 18:30,
35\index{12 2 Kings:ch.  18, v.  30, 35, pg. ``f30!}; Exodus 5:2\index{02 Exodus:ch.   5, v.   2, pg. ``f30!}; Psalm
139:20\index{19 Psalms:ch. 139, v.  20, pg. ``f30!}.} ordinances,\footnote{\ Psalm 50:16-17\index{19 Psalms:ch.  50, v.
 16-17, pg. ``f30!}.} or works,\footnote{\ Psalm 50:16-17\index{19 Psalms:ch.  50, v.  16-17, pg. ``f30!}.} by
blasphemy\index{blasphemy ``f10!},\footnote{\ Isaiah 5:12\index{23 Isaiah:ch.   5, v.  12, pg. ``f30!}.}
perjury\index{Falsehood:perjury ``f10!};\footnote{\ 2 Kings 19:22\index{12 2 Kings:ch.  19, v.  22, pg. ``f30!};
Leviticus 24:11\index{03 Leviticus:ch.  24, v.  11, pg. ``f30!}.} all sinful cursings\index{cursing
``f10!},\footnote{\ Zechariah 5:4\index{38 Zechariah:ch.   5, v.   4, pg. ``f30!}; Zechariah 8:17\index{38
Zechariah:ch.   8, v.  17, pg. ``f30!}.} oaths\index{Oaths and Vows:oaths ``f10!},\footnote{\ 1 Samuel 17:43\index{09 1
Samuel:ch.  17, v.  43, pg. ``f30!}; 2 Samuel 16:5\index{10 2 Samuel:ch.  16, v.   5, pg. ``f30!}.}
vows,\footnote{\ Jeremiah 5:7\index{24 Jeremiah:ch.   5, v.   7, pg. ``f30!}; Jeremiah 23:10\index{24 Jeremiah:ch.  23,
v.  10, pg. ``f30!}.} and lots;\footnote{\ Deuteronomy 23:18\index{05 Deuteronomy:ch.  23, v.  18, pg. ``f30!}; Acts
23:12, 14\index{44 Acts:ch.  23, v.  12, 14, pg. ``f30!}.} violating\index{Oaths and Vows:not to:be violated ``f10!} of
our oaths and vows, if lawful\index{Oaths and Vows:lawful ``f10!};\footnote{\ Esther 3:7\index{17 Esther:ch.   3, v.  
7, pg. ``f30!}; Esther 9:24\index{17 Esther:ch.   9, v.  24, pg. ``f30!}; Psalm 22:1\index{19 Psalms:ch.  22, v.   1,
pg. ``f30!}8\index{19 Psalms:ch.  22, v.  18, pg. ``f30!}.} and fulfilling them\index{Oaths and Vows:duty to fulfill
``f10!}, if of things unlawful\index{Oaths and Vows:not for things:unlawful ``f10!};\footnote{\ Psalm 24:4\index{19
Psalms:ch.  24, v.   4, pg. ``f30!}; Ezekiel 17:16, 18-19\index{26 Ezekiel:ch.  17, v.  16, 18-19, pg. ``f30!}.}
murmuring\index{murmuring ``f10!} and quarreling\index{quarreling ``f10!} at,\footnote{\ Mark 6:26\index{41 Mark:ch.  
6, v.  26, pg. ``f30!}; 1 Samuel 25:22, 32-34\index{09 1 Samuel:ch.  25, v.  22, 32-34, pg. ``f30!}} curious
prying\index{Providence:not to pry into with curiosity ``f10!} into,\footnote{\ Romans 9:14, 19-20\index{45 Romans:ch. 
 9, v.  14, 19-20, pg. ``f30!}.} and misapplying\index{God's name:not to be:misapplied ``f10!} of God's
decrees\index{Predestination:eternal decrees ``f10!}\footnote{\ Deuteronomy 29:29\index{05 Deuteronomy:ch.  29, v.  29,
pg. ``f30!}.} and providences\index{Providence:providence ``f10!};\footnote{\ Romans 3:5, 7\index{45 Romans:ch.   3, v.
  5, 7, pg. ``f30!}; Romans 6:1-2\index{45 Romans:ch.   6, v.   1-2, pg. ``f30!}.} misinterpreting\index{God's name:not
to be:misinterpreted ``f10!},\footnote{\ Ecclesiastes 8:11\index{21 Ecclesiastes:ch.   8, v.  11, pg. ``f30!};
Ecclesiastes 9:3\index{21 Ecclesiastes:ch.   9, v.   3, pg. ``f30!}; Psalm 39:1-13\index{19 Psalms:ch.  39, v.   1-13,
pg. ``f30!}.} misapplying,\footnote{\ Matthew 5:21-48\index{40 Matthew:ch.   5, v.  21-48, pg. ``f30!}.} or any way
perverting the\index{Falsehood:perverting Biblical truth ``f10!} word, or any part of it,\footnote{\ Ezekiel
13:22\index{26 Ezekiel:ch.  13, v.  22, pg. ``f30!}.} to profane jests\index{jests, profane ``f10!},\footnote{\ 2 Peter
3:16\index{61 2 Peter:ch.   3, v.  16, pg. ``f30!}; Matthew 22:24-31\index{40 Matthew:ch.  22, v.  24-31, pg. ``f30!}.}
curious or unprofitable Questions\index{questions, unprofitable ``f10!},\footnote{\ Isaiah 22:18\index{23 Isaiah:ch. 
22, v.  18, pg. ``f30!}; Jeremiah 23:34, 36, 38\index{24 Jeremiah:ch.  23, v.  34, 36, 38, pg. ``f30!}.} vain
janglings, or the maintaining of false\index{God's name:not to be:abused by false doctrines ``f10!}
doctrines\index{Doctrine:false:of men ``f10!};\footnote{\ 1 Timothy 1:4, 6-7\index{54 1 Timothy:ch.   1, v.   4, 6-7,
pg. ``f30!}; 1 Timothy 6:4\index{54 1 Timothy:ch.   6, v.   4, pg. ``f30!}{}-5, 20\index{54 1 Timothy:ch.   6, v.  
4-5, 20, pg. ``f30!}; 2 Timothy 2:14\index{55 2 Timothy:ch.   2, v.  14, pg. ``f30!}; Titus. 3:9.} abusing it, the
creatures\index{creatures ``f10!}, or anything contained under the name of God, to charms\index{Superstitions:charms
``f10!},\footnote{\ Deuteronomy 18:10-14\index{05 Deuteronomy:ch.  18, v.  10-14, pg. ``f30!}; Acts 19:1\index{44
Acts:ch.  19, v.   1, pg. ``f30!}3\index{44 Acts:ch.  19, v.  13, pg. ``f30!}.} or sinful lusts\index{Sexual
immorality:lust ``f10!} and practices;\footnote{\ 2 Timothy 4:3\index{55 2 Timothy:ch.   4, v.   3, pg.
``f30!}{}-4\index{55 2 Timothy:ch.   4, v.   3-4, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg.
``f30!}3\index{45 Romans:ch.  13, v.  13, pg. ``f30!}{}-14\index{45 Romans:ch.  13, v.  13-14, pg. ``f30!}; 1 Kings
21:9-10\index{11 1 Kings:ch.  21, v.   9-10, pg. ``f30!}; Jude 1:4\index{65 Jude:v.   1, v.   4, pg. ``f30!}.} the
maligning\index{God's name:not to be:maligned ``f10!},\footnote{\ Acts 13:45\index{44 Acts:ch.  13, v.  45, pg.
``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}2\index{62 1 John:ch.   3, v.  12, pg. ``f30!}.}
scorning\index{God's name:not to be:scorned ``f10!},\footnote{\ Psalm 1:1\index{19 Psalms:ch.   1, v.   1, pg. ``f30!};
2 Peter 3:3\index{61 2 Peter:ch.   3, v.   3, pg. ``f30!}.} reviling\index{reviling God or man ``f10!},\footnote{\ 1
Peter 4:4\index{60 1 Peter:ch.   4, v.   4, pg. ``f30!}.} or any wise opposing of\index{Falsehood:opposing truth
``f10!} God's truth, grace, and ways\index{Grace:opposed ``f10!};\footnote{\ Acts 4:18\index{44 Acts:ch.   4, v.  18,
pg. ``f30!}; Acts 13:45\index{44 Acts:ch.  13, v.  45, pg. ``f30!}{}-46, 50\index{44 Acts:ch.  13, v.  45-46, 50, pg.
``f30!}; Acts 19:9\index{44 Acts:ch.  19, v.   9, pg. ``f30!}; 1 Thessalonians 2:16\index{52 1 Thessalonians:ch.   2,
v.  16, pg. ``f30!}; Hebrews 10:29\index{58 Hebrews:ch.  10, v.  29, pg. ``f30!}.} making
profession\index{Profession:false:hypocrisy ``f10!} of religion in\index{Profession:of false religion ``f10!}
hypocrisy\index{hypocrisy ``f10!}, or for sinister\index{Profession:false:sinister ends ``f10!} ends;\footnote{\ 2
Timothy 3:5\index{55 2 Timothy:ch.   3, v.   5, pg. ``f30!}; Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg.
``f30!}{}-2, 5, 16\index{40 Matthew:ch.   6, v.   1-2, 5, 16, pg. ``f30!}; Matthew 23:14\index{40 Matthew:ch.  23, v. 
14, pg. ``f30!}.} being ashamed\index{Sin:ashamed of God's truth ``f10!} of it,\footnote{\ Mark 8:38\index{41 Mark:ch. 
 8, v.  38, pg. ``f30!}.} or a shame to\index{God's name:not to be:shamed ``f10!} it, by
unconformable,\footnote{\ Psalm 73:14-15\index{19 Psalms:ch.  73, v.  14-15, pg. ``f30!}.} unwise,\footnote{\ 1
Corinthians 6:5-6\index{46 1 Corinthians:ch.   6, v.   5-6, pg. ``f30!}; Ephesians 5:15-17\index{49 Ephesians:ch.   5,
v.  15-17, pg. ``f30!}.} unfruitful,\footnote{\ Isaiah 5:4\index{23 Isaiah:ch.   5, v.   4, pg. ``f30!}; 2 Peter
1:8-9\index{61 2 Peter:ch.   1, v.   8-9, pg. ``f30!}.} and offensive walking,\footnote{\ Romans 2:23-24\index{45
Romans:ch.   2, v.  23-24, pg. ``f30!}.} or backsliding\index{backsliding ``f10!} from it.\footnote{\ Galatians 3:1,
3\index{48 Galatians:ch.   3, v.   1, 3, pg. ``f30!}; Hebrews 6:6\index{58 Hebrews:ch.   6, v.   6, pg. ``f30!}.}


\bigskip

Q. 114 \textit{What reasons are annexed to the third commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The reasons annexed to the third commandment\index{God's law:commandments:third ``f10!}, in these words, The Lord thy
God, and, For the Lord will not hold him guiltless\index{God's name:not guiltless if take in vain ``f10!} that takes
his name in vain\index{God's name:not to be:taken in vain ``f10!},\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.  
7, pg. ``f30!}.} are, because he is the Lord and our God, therefore his name is not to be profaned, or\index{God's
name:not to be:profaned ``f10!} any way abused\index{God's name:not to be:abused ``f10!} by us;\footnote{\ Leviticus
19:12\index{03 Leviticus:ch.  19, v.  12, pg. ``f30!}.} especially because he will be so far from
acquitting\index{Judgement day:blasphemers:not acquitted ``f10!} and sparing\index{Judgement day:blasphemers:not spared
``f10!} the transgressors\index{God's law:commandments:transgressions ``f10!} of this commandment, as\index{God's
law:commandments:second ``f10!} that he will not suffer them to escape\index{Punishment:no escape ``f10!} his righteous
judgment\index{Judgement:righteous ``f10!},\footnote{\ Ezekiel 36:21-23\index{26 Ezekiel:ch.  36, v.  21-23, pg.
``f30!}; Deuteronomy 28:58\index{05 Deuteronomy:ch.  28, v.  58, pg. ``f30!}{}-59\index{05 Deuteronomy:ch.  28, v. 
58-59, pg. ``f30!}; Zechariah 5:2-4\index{38 Zechariah:ch.   5, v.   2-4, pg. ``f30!}.} albeit many such escape the
censures and\index{Punishment:escape from man's ``f10!} punishments of men.\footnote{\ 1 Samuel 2:1\index{09 1
Samuel:ch.   2, v.   1, pg. ``f30!}2, 17, 22, 24\index{09 1 Samuel:ch.   2, v.  12, 17, 22, 24, pg. ``f30!}; 1 Samuel
3:18\index{09 1 Samuel:ch.   3, v.  18, pg. ``f30!}.}


\bigskip

Q. 115 \textit{Which is the fourth commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The fourth commandment\index{God's law:commandments:fourth ``f10!} is, Remember\index{Sabbath:to be:remembered
``f10!} the sabbath day, to keep it holy\index{Sabbath:to be:kept holy ``f10!}. Six\index{Sabbath:six days:for labour
``f10!} days shalt you labor, and do\index{Sabbath:labour six days, rest on Sabbath ``f10!} all thy work: but the
seven\index{Sabbath:one whole day in seven ``f10!}th day is the sabbath of\index{Sabbath:instituted by God ``f10!} the
Lord thy God: in it you shalt not do any work, thou, nor thy son, nor thy daughter, thy manservant, nor thy
maidservant\index{Sabbath:servants not to be made to work ``f10!}, nor thy cattle, nor thy stranger that is within thy
gates: For in six\index{Creation:in six days ``f10!} days the Lord made heaven\index{God's attributes:creator:of heaven
and earth ``f10!} and earth, the sea\index{God's attributes:creator:of sea ``f10!}, and all that in them is, and rested
the\index{Sabbath:God rested the seventh day ``f10!} seventh day\index{Sabbath:seventh day ``f10!}: wherefore the Lord
blessed the\index{Sabbath:blessed ``f10!} sabbath day, and hallowed\index{God's name:hallowed ``f10!}
it.\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}{}-11\index{02 Exodus:ch.  20, v.   8-11, pg.
``f30!}.}


\bigskip

Q. 116 \textit{What is required in the fourth commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The fourth commandment\index{God's law:commandments:fourth ``f10!} requires of all\index{Sabbath:to be:kept holy
``f10!} men the sanctifying or\index{Sabbath:sanctified ``f10!} keeping holy to God such set times as he has appointed
in his word, expressly one\index{Sabbath:one whole day in seven ``f10!} whole day in seven; which was the seventh
from\index{Sabbath:seventh day ``f10!} the beginning of the world to the resurrection of
Christ\index{Jesus:resurrection:from the dead ``f10!}, and the first day\index{Sabbath:first day of week ``f10!} of the
week ever since, and so to continue to the end of the world; which is the Christian sabbath,\footnote{\ Deuteronomy
5:12, 14, 18\index{05 Deuteronomy:ch.   5, v.  12, 14, 18, pg. ``f30!}; Genesis 2:2-3\index{01 Genesis:ch.   2, v.  
2-3, pg. ``f30!}; 1 Corinthians 16:1-2\index{46 1 Corinthians:ch.  16, v.   1-2, pg. ``f30!}; Acts 20:7\index{44
Acts:ch.  20, v.   7, pg. ``f30!}; Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.
  5, v.  17-18, pg. ``f30!}; Isaiah 56:2, 4, 6-7\index{23 Isaiah:ch.  56, v.   2, 4, 6-7, pg. ``f30!}.} and in the New
Testament\index{Scripture:New Testament ``f10!} called The Lord's day\index{Sabbath:Lord's Day
``f10!}.\footnote{\ Revelation 1:10\index{66 Revelation:ch.   1, v.  10, pg. ``f30!}.}


\bigskip

Q. 117 \textit{How is the sabbath or}\index{Sabbath:Lord's Day ``f10!}\textit{ the Lord's day to be
sanctified}\index{Sanctification:sanctification ``f10!}\textit{?}

A. The sabbath or\index{Sabbath:Lord's Day ``f10!} Lord's day is to be sanctified\index{Sanctification:sanctification
``f10!} by an holy resting\index{Sabbath:to be:kept holy ``f10!} all the day,\footnote{\ Exodus 20:8\index{02
Exodus:ch.  20, v.   8, pg. ``f30!}, 10\index{02 Exodus:ch.  20, v.   8, 10, pg. ``f30!}.} not only from such works as
are at all times sinful, but even from such worldly employments and recreations\index{Sabbath:cease:worldly employments
and recreations ``f10!} as are on other days lawful\index{Sabbath:cease:activities lawful on other days
``f10!};\footnote{\ Exodus 16:25-28\index{02 Exodus:ch.  16, v.  25-28, pg. ``f30!}; Nehemiah 13:15-22\index{16
Nehemiah:ch.  13, v.  15-22, pg. ``f30!}; Jeremiah 17:21-22\index{24 Jeremiah:ch.  17, v.  21-22, pg. ``f30!}.} and
making it our delight\index{Worship:Sabbath delight ``f10!} to spend\index{Sabbath:spend bulk of time in public and
private worship ``f10!} the whole time (except so much of it as is to be taken up in works of necessity
and\index{Sabbath:works of necessity and mercy ``f10!} mercy)\footnote{\ Matthew 12:1-13\index{40 Matthew:ch.  12, v.  
1-13, pg. ``f30!}.} in the public and private exercises\index{Worship:private exercises ``f10!} of God's
worship\index{Worship:of God ``f10!}:\footnote{\ Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg.
``f30!}{}-14\index{23 Isaiah:ch.  58, v.  13-14, pg. ``f30!}; Isaiah 66:2\index{23 Isaiah:ch.  66, v.   2, pg.
``f30!}3\index{23 Isaiah:ch.  66, v.  23, pg. ``f30!}; Luke 4:16\index{42 Luke:ch.   4, v.  16, pg. ``f30!}; Acts
20:7\index{44 Acts:ch.  20, v.   7, pg. ``f30!}; 1 Corinthians 16:1-2\index{46 1 Corinthians:ch.  16, v.   1-2, pg.
``f30!}; Psalm 92; Leviticus 23:3\index{03 Leviticus:ch.  23, v.   3, pg. ``f30!}.} and, to that end, we are to prepare
our hearts\index{Worship:prepare hearts ``f10!}, and with such foresight\index{Lord's Supper:preparation:foresight
``f10!}, diligence, and moderation\index{Material things:moderation:in use ``f10!}, to dispose
and\index{Sabbath:seasonably dispatch and dispose worldly business ``f10!} seasonably dispatch our worldly
business\index{Sabbath:order common affairs before ``f10!}, that we may be the more free and fit for the
duties\index{Sabbath:duties ``f10!} of that day.\footnote{\ Exodus 16:22, 25-26, 29\index{02 Exodus:ch.  16, v.  22,
25-26, 29, pg. ``f30!}; Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}; Luke 23:54, 56\index{42 Luke:ch.  23,
v.  54, 56, pg. ``f30!}; Nehemiah 13:19\index{16 Nehemiah:ch.  13, v.  19, pg. ``f30!}.}


\bigskip

Q. 118 \textit{Why is the charge of}\index{Sabbath:charge of keeping directed to:governors ``f10!}\textit{ keeping the
sabbath more specially directed to governors of families}\index{family, natural ``f10!}\textit{, and other
superiors}\index{Sabbath:charge of keeping directed to:other superiors ``f10!}\textit{?}

A. The charge of\index{Sabbath:charge of keeping directed to:governors ``f10!} keeping the sabbath is more specially
directed to governors of families\index{family, natural ``f10!}, and other superiors\index{Sabbath:charge of keeping
directed to:other superiors ``f10!}, because they are bound not only to keep it themselves, but to see that it be
observed by\index{Sabbath:superiors to ensure inferiors keep ``f10!} all those that are under their charge; and because
they are prone\index{Sabbath:minds distracted by worldly business ``f10!} ofttimes to hinder them by
employments\index{Sabbath:cease:worldly employments and recreations ``f10!} of their own.\footnote{\ Exodus
20:10\index{02 Exodus:ch.  20, v.  10, pg. ``f30!}; Exodus 23:1\index{02 Exodus:ch.  23, v.   1, pg. ``f30!}2\index{02
Exodus:ch.  23, v.  12, pg. ``f30!}; Joshua 24:15\index{06 Joshua:ch.  24, v.  15, pg. ``f30!}; Nehemiah 13:15,
17\index{16 Nehemiah:ch.  13, v.  15, 17, pg. ``f30!}; Jeremiah 17:20-22\index{24 Jeremiah:ch.  17, v.  20-22, pg.
``f30!}.}


\bigskip

Q. 119 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the fourth
commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the fourth commandment\index{God's
law:commandments:fourth ``f10!} are, all omissions\index{Sin:omission of duties ``f10!} of the duties
required\index{God's law:commandments:duties:required by God ``f10!},\footnote{\ Ezekiel 22:26\index{26 Ezekiel:ch. 
22, v.  26, pg. ``f30!}.} all careless, negligent, and\index{Sabbath:negligent of duties ``f10!} unprofitable
performing of them, and being weary of them;\footnote{\ Acts 15:7, 9\index{44 Acts:ch.  15, v.   7, 9, pg. ``f30!};
Ezekiel 33:30-32\index{26 Ezekiel:ch.  33, v.  30-32, pg. ``f30!}; Amos 8:5\index{30 Amos:ch.   8, v.   5, pg. ``f30!};
Malachi 1:13\index{39 Malachi:ch.   1, v.  13, pg. ``f30!}.} all profaning the\index{Sabbath:not to be profaned ``f10!}
day by idleness\index{idleness ``f10!}, and doing that which is in itself sinful;\footnote{\ Ezekiel 23:38\index{26
Ezekiel:ch.  23, v.  38, pg. ``f30!}.} and by all needless\index{Sabbath:cease:activities lawful on other days ``f10!}
works, words, and thoughts, about\index{Sabbath:avoid unnecessary, worldly thoughts ``f10!} our worldly employments and
recreations\index{Sabbath:cease:worldly employments and recreations ``f10!}.\footnote{\ Jeremiah 17:24, 27\index{24
Jeremiah:ch.  17, v.  24, 27, pg. ``f30!}; Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}.}


\bigskip

Q. 120 \textit{What are the reasons annexed to the fourth commandment}\index{God's law:commandments:fourth
``f10!}\textit{, the more to enforce it}\index{God's law:commandments:words of enforcement ``f10!}\textit{?}

A. The reasons annexed to the fourth commandment\index{God's law:commandments:fourth ``f10!}, the more to enforce
it\index{God's law:commandments:words of enforcement ``f10!}, are taken from the equity of it, God allowing us
six\index{Sabbath:six days:for own affairs ``f10!} days of seven for our own affairs, and reserving
but\index{Sabbath:one whole day in seven ``f10!} one for himself, in these words, Six\index{Sabbath:six days:for labour
``f10!} days shalt you labor, and do\index{Sabbath:labour six days, rest on Sabbath ``f10!} all thy
work:\footnote{\ Exodus 20:9\index{02 Exodus:ch.  20, v.   9, pg. ``f30!}.} from God's challenging a special propriety
in that\index{Sabbath:God's propriety over ``f10!} day, The seventh day\index{Sabbath:seventh day ``f10!} is the
sabbath of\index{Sabbath:instituted by God ``f10!} the Lord thy God:\footnote{\ Exodus 20:10\index{02 Exodus:ch.  20,
v.  10, pg. ``f30!}.} from the example of God, who in six\index{Creation:in six days ``f10!} days made
heaven\index{God's attributes:creator:of heaven and earth ``f10!} and earth, the sea\index{God's attributes:creator:of
sea ``f10!}, and all that in them is, and rested the\index{Sabbath:God rested the seventh day ``f10!} seventh day: and
from that blessing which God put upon that day, not only in sanctifying it\index{Sabbath:sanctified ``f10!} to be a day
for his service\index{Sabbath:day for service of God ``f10!}, but in ordaining it\index{Sabbath:ordained as means of
blessing ``f10!} to be a means of blessing to us in our sanctifying it; Wherefore the Lord blessed
the\index{Sabbath:blessed ``f10!} sabbath day, and hallowed\index{God's name:hallowed ``f10!} it.\footnote{\ Exodus
20:11\index{02 Exodus:ch.  20, v.  11, pg. ``f30!}.}


\bigskip

Q. 121 \textit{Why is the word Remember}\index{Sabbath:to be:remembered ``f10!}\textit{ set in the beginning of the
fourth commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The word Remember\index{Sabbath:to be:remembered ``f10!} is set in the beginning of the fourth
commandment\index{God's law:commandments:fourth ``f10!},\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg.
``f30!}.} partly, because of the great benefit of\index{Sabbath:benefit of ``f10!} remembering it, we being thereby
helped in our preparation to keep\index{Sabbath:preparation for ``f10!} it,\footnote{\ Exodus 16:23\index{02 Exodus:ch.
 16, v.  23, pg. ``f30!}; Luke 23:54, 56\index{42 Luke:ch.  23, v.  54, 56, pg. ``f30!}; Mark 15:42\index{41 Mark:ch. 
15, v.  42, pg. ``f30!}; Nehemiah 13:19\index{16 Nehemiah:ch.  13, v.  19, pg. ``f30!}.} and, in keeping it, better to
keep all the rest of the commandment\index{God's law:commandments:given to Adam ``f10!}s\index{God's
law:commandments:ten ``f10!},\footnote{\ Psalm 92:13-14\index{19 Psalms:ch.  92, v.  13-14, pg. ``f30!}; Ezekiel 20:12,
19-20\index{26 Ezekiel:ch.  20, v.  12, 19-20, pg. ``f30!}.} and to continue a thankful\index{Prayer:thanking God
``f10!} remembrance of the two\index{Sabbath:thankful remembrance of benefits ``f10!} great benefits of
creation\index{Believers:benefits:of creation ``f10!} and redemption, which\index{Redemption:purchased by Christ
``f10!} contain a short abridgment of religion;\footnote{\ Genesis 2:2-3\index{01 Genesis:ch.   2, v.   2-3, pg.
``f30!}; Psalm 118:22, 24\index{19 Psalms:ch. 118, v.  22, 24, pg. ``f30!}; Acts 4:10, 11\index{44 Acts:ch.   4, v. 
10, 11, pg. ``f30!}; Revelation 1:10\index{66 Revelation:ch.   1, v.  10, pg. ``f30!}.} and partly, because we are very
ready to forget\index{Sabbath:mankind prone to forget ``f10!} it,\footnote{\ Ezekiel 22:26\index{26 Ezekiel:ch.  22, v.
 26, pg. ``f30!}.} for that there is less light of nature for it,\footnote{\ Nehemiah 9:1\index{16 Nehemiah:ch.   9, v.
  1, pg. ``f30!}4\index{16 Nehemiah:ch.   9, v.  14, pg. ``f30!}.} and yet it restrains\index{Sabbath:cease:activities
lawful on other days ``f10!} our natural liberty in things\index{Liberty:natural ``f10!} at other times
lawful;\footnote{\ Exodus 34:21\index{02 Exodus:ch.  34, v.  21, pg. ``f30!}.} that it comes but once in
seven\index{Sabbath:one whole day in seven ``f10!} days, and many worldly business\index{Sabbath:order common affairs
before ``f10!}es\index{Sabbath:cease:worldly employments and recreations ``f10!} come between, and too often take off
our minds from\index{Sabbath:minds distracted by worldly business ``f10!} thinking of\index{Sabbath:avoid unnecessary,
worldly thoughts ``f10!} it, either to prepare for it, or to sanctify\index{Sabbath:sanctified ``f10!}
it;\footnote{\ Deuteronomy 5:14-15\index{05 Deuteronomy:ch.   5, v.  14-15, pg. ``f30!}; Amos 8:5\index{30 Amos:ch.  
8, v.   5, pg. ``f30!}.} and that Satan with\index{Satan:enemy of God ``f10!} his
instruments\index{Unbelievers:instruments of Satan ``f10!} much labor to blot\index{Sabbath:Satan labours to blot out
``f10!} out the glory, and even the memory of it, to bring in all irreligion\index{Satan:labours to bring irreligion
and impiety ``f10!} and impiety.\footnote{\ Lamentations 1:7\index{25 Lamentations:ch.   1, v.   7, pg. ``f30!};
Jeremiah 17:21-23\index{24 Jeremiah:ch.  17, v.  21-23, pg. ``f30!}; Nehemiah 13:15-23\index{16 Nehemiah:ch.  13, v. 
15-23, pg. ``f30!}.}


\bigskip

Q. 122 \textit{What is the sum of the six commandments}\index{God's law:commandments:last six, duty to man
``f10!}\textit{ which contain our duty to man}\index{God's law:commandments:duties:to others ``f10!}\textit{?}

A. The sum of the six commandments\index{God's law:commandments:last six, duty to man ``f10!} which contain our duty to
man\index{God's law:commandments:duties:to others ``f10!} is, to love our neighbor\index{Love for others:neighbour as
self ``f10!} as ourselves,\footnote{\ Matthew 22:39\index{40 Matthew:ch.  22, v.  39, pg. ``f30!}.} and to do to others
what we would have them to do to us.\footnote{\ Matthew 7:1\index{40 Matthew:ch.   7, v.   1, pg. ``f30!}2\index{40
Matthew:ch.   7, v.  12, pg. ``f30!}.}


\bigskip

Q. 123 \textit{Which is the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. The fifth commandment\index{God's law:commandments:fifth ``f10!} is, Honor\index{Parents:to be honoured ``f10!} thy
father and thy mother\index{Superiors:styled as fathers and mothers ``f10!}; that thy days may be long upon the land
which the Lord thy God gives you.\footnote{\ Exodus 20:12\index{02 Exodus:ch.  20, v.  12, pg. ``f30!}.} 


\bigskip

Q. 124 \textit{Who are meant by father and}\index{Parents:to be honoured ``f10!}\textit{ mother}\index{Superiors:styled
as fathers and mothers ``f10!}\textit{ in the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. By father and\index{Parents:to be honoured ``f10!} mother\index{Superiors:styled as fathers and mothers ``f10!}, in
the fifth commandment\index{God's law:commandments:fifth ``f10!}, are meant, not only natural
parents\index{Parents:natural ``f10!},\footnote{\ Proverbs 23:22\index{20 Proverbs:ch.  23, v.  22, pg.
``f30!}{}-25\index{20 Proverbs:ch.  23, v.  22-25, pg. ``f30!}; Ephesians 6:1-2\index{49 Ephesians:ch.   6, v.   1-2,
pg. ``f30!}.} but all superiors in age\footnote{\ 1 Timothy 5:1\index{54 1 Timothy:ch.   5, v.   1, pg.
``f30!}{}-2\index{54 1 Timothy:ch.   5, v.   1-2, pg. ``f30!}.} and gifts;\index{Church:furnished with:gifts and graces
``f10!}\footnote{\ Genesis 4:20-22\index{01 Genesis:ch.   4, v.  20-22, pg. ``f30!}; Genesis 45:8\index{01 Genesis:ch. 
45, v.   8, pg. ``f30!}.} and especially such as, by God's ordinance, are\index{Superiors:authority by God's ordinance
``f10!} over us in place of authority, whether in family\index{family, natural ``f10!},\footnote{\ 2 Kings
5:13\index{12 2 Kings:ch.   5, v.  13, pg. ``f30!}.} church,\footnote{\ 2 Kings 2:12\index{12 2 Kings:ch.   2, v.  12,
pg. ``f30!}; 2 Kings 13:14\index{12 2 Kings:ch.  13, v.  14, pg. ``f30!}; Galatians 4:19\index{48 Galatians:ch.   4, v.
 19, pg. ``f30!}.} or commonwealth\index{commonwealth ``f10!}.\footnote{\ Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23,
pg. ``f30!}.}


\bigskip

Q. 125 \textit{Why are superiors styled Father}\index{Father:God the ``f10!}\textit{ and Mother}\index{Superiors:styled
as fathers and mothers ``f10!}\textit{?}

A. Superiors are styled Father\index{Father:God the ``f10!} and Mother\index{Superiors:styled as fathers and mothers
``f10!}, both to teach them in all duties toward\index{Superiors:duties toward inferiors ``f10!} their inferiors, like
natural parents\index{Parents:natural ``f10!}, to express love\index{Love for others:parents ``f10!} and
tenderness\index{Superiors:to:express love and tenderness ``f10!} to them, according to their several
relations;\footnote{\ Ephesians 6:4\index{49 Ephesians:ch.   6, v.   4, pg. ``f30!}; 2 Corinthians 12:14\index{47 2
Corinthians:ch.  12, v.  14, pg. ``f30!}; 1 Thessalonians 2:7-8, 11\index{52 1 Thessalonians:ch.   2, v.   7-8, 11, pg.
``f30!}; Numbers 11:11-12\index{04 Numbers:ch.  11, v.  11-12, pg. ``f30!}.} and to work inferiors to
a\index{Superiors:to:encourage inferiors ``f10!} greater willingness\index{Inferiors:to:willingly perform duties
``f10!} and cheerfulness\index{cheerfulness ``f10!} in performing their duties\index{Inferiors:to:perform duties
cheerfully ``f10!} to their superiors, as to their parents\index{Parents:superiors, equivalent ``f10!}.\footnote{\ 1
Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.   1, pg. ``f30!}4-16\index{46 1 Corinthians:ch.   4, v.  14-16, pg.
``f30!}; 2 Kings 5:13\index{12 2 Kings:ch.   5, v.  13, pg. ``f30!}.}


\bigskip

Q. 126 \textit{What is the general scope of the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. The general scope of the fifth commandment\index{God's law:commandments:fifth ``f10!} is, the performance of those
duties\index{God's law:commandments:duties:to others ``f10!} which we mutually owe in\index{Love for
others:mutual:duties ``f10!} our several relations, as inferiors, superiors, or equals\index{Equals:duties of
``f10!}.\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}1\index{49 Ephesians:ch.   5, v.  21,
pg. ``f30!}; 1 Peter 2:17\index{60 1 Peter:ch.   2, v.  17, pg. ``f30!}; Romans 12:10\index{45 Romans:ch.  12, v.  10,
pg. ``f30!}.}


\bigskip

Q. 127 \textit{What is the honor that}\index{Inferiors:to:honour superiors ``f10!}\textit{ inferiors owe to their
superiors?}

A. The honor which\index{Inferiors:to:honour superiors ``f10!} inferiors owe to their superiors is, all due
reverence\index{Inferiors:due reverence to superiors ``f10!} in heart\index{Love for others:from heart
``f10!},\footnote{\ Malachi 1:6\index{39 Malachi:ch.   1, v.   6, pg. ``f30!}; Leviticus 19:3\index{03 Leviticus:ch. 
19, v.   3, pg. ``f30!}.} word,\footnote{\ Proverbs 31:28\index{20 Proverbs:ch.  31, v.  28, pg. ``f30!}; 1 Peter
3:6\index{60 1 Peter:ch.   3, v.   6, pg. ``f30!}.} and behavior\index{Good works:behavior
``f10!};\footnote{\ Leviticus 19:3\index{03 Leviticus:ch.  19, v.   3, pg. ``f30!}2\index{03 Leviticus:ch.  19, v.  32,
pg. ``f30!}; 1 Kings 2:19\index{11 1 Kings:ch.   2, v.  19, pg. ``f30!}.} prayer and thanksgiving\index{Prayer:thanking
God ``f10!} for them;\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1
Timothy:ch.   2, v.   1-2, pg. ``f30!}.} imitation of their virtues\index{Inferiors:to:imitate superiors' virtues
``f10!} and graces;\footnote{\ Hebrews 13:7\index{58 Hebrews:ch.  13, v.   7, pg. ``f30!}; Philippians 3:17\index{50
Philippians:ch.   3, v.  17, pg. ``f30!}.} willing obedience\index{Commandments of men:lawful to be obeyed ``f10!} to
their lawful commands and counsels;\footnote{\ Ephesians 6:1-2\index{49 Ephesians:ch.   6, v.   1-2, pg. ``f30!},
5-7\index{49 Ephesians:ch.   6, v.   1-2, 5-7, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg.
``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg.
``f30!}{}-5\index{45 Romans:ch.  13, v.   1-5, pg. ``f30!}; Hebrews 13:17\index{58 Hebrews:ch.  13, v.  17, pg.
``f30!}; Proverbs 4:3-4\index{20 Proverbs:ch.   4, v.   3-4, pg. ``f30!}; Proverbs 23:22\index{20 Proverbs:ch.  23, v. 
22, pg. ``f30!}; Exodus 18:19, 24\index{02 Exodus:ch.  18, v.  19, 24, pg. ``f30!}.} due
submission\index{Inferiors:submission to superiors ``f10!} to their corrections\index{Inferiors:to:accept correction
``f10!};\footnote{\ Hebrews 12:9\index{58 Hebrews:ch.  12, v.   9, pg. ``f30!}; 1 Peter 2:18-20\index{60 1 Peter:ch.  
2, v.  18-20, pg. ``f30!}.} fidelity\index{Inferiors:fidelity to superiors ``f10!} to,\footnote{\ Titus 2:9-10\index{56
Titus:ch.   2, v.   9-10, pg. ``f30!}.} defense\footnote{\ 1 Samuel 26:15-16\index{09 1 Samuel:ch.  26, v.  15-16, pg.
``f30!}; 2 Samuel 18:3\index{10 2 Samuel:ch.  18, v.   3, pg. ``f30!}; Esther 6:2\index{17 Esther:ch.   6, v.   2, pg.
``f30!}.} and maintenance of their persons and authority, according to their several ranks\index{station in life, ranks
``f10!}, and the nature of their places;\footnote{\ Matthew 22:21\index{40 Matthew:ch.  22, v.  21, pg. ``f30!}; Romans
13:6-7\index{45 Romans:ch.  13, v.   6-7, pg. ``f30!}; 1 Timothy 5:1\index{54 1 Timothy:ch.   5, v.   1, pg.
``f30!}7\index{54 1 Timothy:ch.   5, v.  17, pg. ``f30!}{}-18\index{54 1 Timothy:ch.   5, v.  17-18, pg. ``f30!};
Galatians 6:6\index{48 Galatians:ch.   6, v.   6, pg. ``f30!}; Genesis 45:11\index{01 Genesis:ch.  45, v.  11, pg.
``f30!}; Genesis 47:12\index{01 Genesis:ch.  47, v.  12, pg. ``f30!}.} bearing with their infirmities,
and\index{Inferiors:to:bear with infirmities of superiors ``f10!} covering them in love, that\index{Inferiors:to:love
superiors ``f10!} so they may be an honor to them and to their government\index{Inferiors:to:honour superiors'
government ``f10!}.\footnote{\ Psalm 127:3-5\index{19 Psalms:ch. 127, v.   3-5, pg. ``f30!}; Proverbs 31:23\index{20
Proverbs:ch.  31, v.  23, pg. ``f30!}.}


\bigskip


\bigskip

Q. 128 \textit{What are the sins of inferiors}\index{Inferiors:sins against superiors ``f10!}\textit{ against their
superiors}\index{Superiors:sins against by inferiors ``f10!}\textit{?}

A. The sins of inferiors\index{Inferiors:sins against superiors ``f10!} against their superiors\index{Superiors:sins
against by inferiors ``f10!} are, all neglect of the duties required\index{God's law:commandments:duties:required by
God ``f10!} toward them;\footnote{\ Matthew 15:4-6\index{40 Matthew:ch.  15, v.   4-6, pg. ``f30!}.}
envying\index{Envy:envy ``f10!} at,\footnote{\ Numbers 11:28-29\index{04 Numbers:ch.  11, v.  28-29, pg. ``f30!}.}
contempt of,\index{Contempt:for superiors ``f10!}\footnote{\ 1 Samuel 8:7\index{09 1 Samuel:ch.   8, v.   7, pg.
``f30!}; Isaiah 3:5\index{23 Isaiah:ch.   3, v.   5, pg. ``f30!}.} and rebellion\index{Inferiors:not to:rebel against
superiors ``f10!}\footnote{\ 2 Samuel 15:1-12\index{10 2 Samuel:ch.  15, v.   1-12, pg. ``f30!}.} against, their
persons\footnote{\ Exodus 21:15\index{02 Exodus:ch.  21, v.  15, pg. ``f30!}.} and places,\index{Commandments of
men:lawful to be obeyed ``f10!}\footnote{\ 1 Samuel 10:27\index{09 1 Samuel:ch.  10, v.  27, pg. ``f30!}.} in their
lawful counsels,\footnote{\ 1 Samuel 2:25\index{09 1 Samuel:ch.   2, v.  25, pg. ``f30!}.} commands, and
corrections\index{Inferiors:to:accept correction ``f10!};\footnote{\ Deuteronomy 21:18-21\index{05 Deuteronomy:ch.  21,
v.  18-21, pg. ``f30!}.} cursing\index{cursing ``f10!}, mocking\index{Inferiors:not to:mock superiors
``f10!},\footnote{\ Proverbs 30:11, 17\index{20 Proverbs:ch.  30, v.  11, 17, pg. ``f30!}.} and all such
refractory\index{Inferiors:not to:dishonor superiors ``f10!} and scandalous\index{Scandalous behaviour:behaviour
``f10!} carriage, as proves a shame and dishonor to them and their government\index{Inferiors:to:honour superiors'
government ``f10!}.\footnote{\ Proverbs 19:26\index{20 Proverbs:ch.  19, v.  26, pg. ``f30!}.}


\bigskip

Q. 129 \textit{What is required of superiors}\index{Superiors:duties toward inferiors ``f10!}\textit{ towards their
inferiors?}

A. It is required of superiors\index{Superiors:duties toward inferiors ``f10!}, according to that
power\index{Superiors:power delegated by Christ ``f10!} they receive from God, and that relation wherein they stand, to
love\index{Superiors:to:love inferiors ``f10!},\footnote{\ Colossians 3:19\index{51 Colossians:ch.   3, v.  19, pg.
``f30!}; Titus 2:4\index{56 Titus:ch.   2, v.   4, pg. ``f30!}.} pray for,\index{Prayer:for others ``f10!}\footnote{\ 1
Samuel 12:23\index{09 1 Samuel:ch.  12, v.  23, pg. ``f30!}; Job 1:5\index{18 Job:ch.   1, v.   5, pg. ``f30!}.} and
bless the\index{Lord's Supper:elements:Jesus blessed ``f10!}ir\index{Superiors:to:bless inferiors ``f10!}
inferiors;\footnote{\ 1 Kings 8:55-56\index{11 1 Kings:ch.   8, v.  55-56, pg. ``f30!}; Hebrews 7:7\index{58
Hebrews:ch.   7, v.   7, pg. ``f30!}; Genesis 49:28\index{01 Genesis:ch.  49, v.  28, pg. ``f30!}.} to
instruct,\footnote{\ Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg. ``f30!}{}-7\index{05 Deuteronomy:ch.  
6, v.   6-7, pg. ``f30!}.} counsel, and\index{Superiors:to:instruct and counsel inferiors ``f10!}
admonish\index{Superiors:to:admonish inferiors ``f10!} them;\footnote{\ Ephesians 6:4\index{49 Ephesians:ch.   6, v.  
4, pg. ``f30!}.} countenancing,\footnote{\ 1 Peter 3:7\index{60 1 Peter:ch.   3, v.   7, pg. ``f30!}.}
commending\index{Superiors:to:commend inferiors ``f10!},\footnote{\ 1 Peter 2:14\index{60 1 Peter:ch.   2, v.  14, pg.
``f30!}; Romans 13:3\index{45 Romans:ch.  13, v.   3, pg. ``f30!}.} and rewarding such\index{Superiors:to:reward justly
``f10!} as do well;\footnote{\ Esther 6:3\index{17 Esther:ch.   6, v.   3, pg. ``f30!}.} and
discountenancing\index{Superiors:to:use discountenancing wisely ``f10!},\footnote{\ Romans 13:3\index{45 Romans:ch. 
13, v.   3, pg. ``f30!}{}-4\index{45 Romans:ch.  13, v.   3-4, pg. ``f30!}.} reproving\index{Superiors:to:reprove those
who do ill ``f10!}, and chastising\index{Superiors:to:chastise inferiors ``f10!} such as do ill;\footnote{\ Proverbs
29:1\index{20 Proverbs:ch.  29, v.   1, pg. ``f30!}5\index{20 Proverbs:ch.  29, v.  15, pg. ``f30!}; 1 Peter
2:14\index{60 1 Peter:ch.   2, v.  14, pg. ``f30!}.} protecting,\footnote{\ Job 29:12-17\index{18 Job:ch.  29, v. 
12-17, pg. ``f30!}; Isaiah 1:10\index{23 Isaiah:ch.   1, v.  10, pg. ``f30!}, 17\index{23 Isaiah:ch.   1, v.  10, 17,
pg. ``f30!}.} and providing\index{Superiors:to:proved for and protect inferiors ``f10!} for them all things necessary
for soul\index{Superiors:to:provide inferiors with things necessary for soul and body ``f10!}\footnote{\ Ephesians
6:4\index{49 Ephesians:ch.   6, v.   4, pg. ``f30!}.} and body:\footnote{\ 1 Timothy 5:8\index{54 1 Timothy:ch.   5, v.
  8, pg. ``f30!}.} and by grave, wise\index{Superiors:to:display wisdom ``f10!}, holy, and\index{Superiors:to:display
holy lives ``f10!} exemplary\index{Superiors:to:set example ``f10!} carriage, to procure glory\index{Superiors:to:give
glory to God ``f10!} to God,\footnote{\ 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.   1, pg. ``f30!}2\index{54 1
Timothy:ch.   4, v.  12, pg. ``f30!}; Titus 2:3-5\index{56 Titus:ch.   2, v.   3-5, pg. ``f30!}.} honor to
them\index{Inferiors:to:honour superiors ``f10!}selves\index{Superiors:procure honour through exemplary behaviour
``f10!},\footnote{\ 1 Kings 3:28\index{11 1 Kings:ch.   3, v.  28, pg. ``f30!}.} and so to preserve
that\index{Superiors:to:preserve authority ``f10!} authority which God has put upon them.\footnote{\ Titus
2:15\index{56 Titus:ch.   2, v.  15, pg. ``f30!}.}


\bigskip


\bigskip


Q. 130 \textit{What are the sins of superiors}\index{Superiors:sins of ``f10!}\textit{?}

A. The sins of superiors\index{Superiors:sins of ``f10!} are, besides the neglect of the duties required\index{God's
law:commandments:duties:required by God ``f10!} of them,\footnote{\ Ezekiel 34:2-4\index{26 Ezekiel:ch.  34, v.   2-4,
pg. ``f30!}.} an inordinate seeking\index{Superiors:not to:seek selfish ends ``f10!} of
themselves,\footnote{\ Philippians 2:21\index{50 Philippians:ch.   2, v.  21, pg. ``f30!}.} their own glory\index{God's
attributes:glory:his own ``f10!},\footnote{\ John 5:44\index{43 John:ch.   5, v.  44, pg. ``f30!}; John 7:18\index{43
John:ch.   7, v.  18, pg. ``f30!}.} ease, profit, or pleasure;\footnote{\ Isaiah 56:10-11\index{23 Isaiah:ch.  56, v. 
10-11, pg. ``f30!}; Deuteronomy 17:17\index{05 Deuteronomy:ch.  17, v.  17, pg. ``f30!}.}
commanding\index{Sin:commanding things unlawful ``f10!} things unlawful,\index{Superiors:not to:command things unlawful
``f10!}\footnote{\ Daniel 3:4-6\index{27 Daniel:ch.   3, v.   4-6, pg. ``f30!}; Acts 4:17-18\index{44 Acts:ch.   4, v. 
17-18, pg. ``f30!}.} or not in the power\index{Superiors:not to:command things not in inferiors' power ``f10!} of
inferiors to perform;\index{Superiors:not to:counsel or encourage evil ``f10!}\footnote{\ Exodus 5:10-18\index{02
Exodus:ch.   5, v.  10-18, pg. ``f30!}; Matthew 23:2, 4\index{40 Matthew:ch.  23, v.   2, 4, pg. ``f30!}.}
counseling,\footnote{\ Matthew 14:8\index{40 Matthew:ch.  14, v.   8, pg. ``f30!}; Mark 6:24\index{41 Mark:ch.   6, v. 
24, pg. ``f30!}.} encouraging,\footnote{\ 2 Samuel 13:28\index{10 2 Samuel:ch.  13, v.  28, pg. ``f30!}.} or favoring
them in that which is evil;\footnote{\ 1 Samuel 3:13\index{09 1 Samuel:ch.   3, v.  13, pg. ``f30!}.}
dissuading\index{Superiors:not to:dissuade inferiors from doing good ``f10!}, discouraging, or\index{Superiors:not
to:discourage inferiors ``f10!} discountenancing\index{Superiors:to:use discountenancing wisely ``f10!} them in that
which is good;\footnote{\ John 7:46-49\index{43 John:ch.   7, v.  46-49, pg. ``f30!}; Colossians 3:2\index{51
Colossians:ch.   3, v.   2, pg. ``f30!}1\index{51 Colossians:ch.   3, v.  21, pg. ``f30!}; Exodus 5:17\index{02
Exodus:ch.   5, v.  17, pg. ``f30!}.} correcting them unduly\index{Superiors:not to:correct inferiors unduly
``f10!};\footnote{\ 1 Peter 2:18-20\index{60 1 Peter:ch.   2, v.  18-20, pg. ``f30!}; Hebrews 12:1\index{58 Hebrews:ch.
 12, v.   1, pg. ``f30!}0\index{58 Hebrews:ch.  12, v.  10, pg. ``f30!}; Deuteronomy 25:3\index{05 Deuteronomy:ch.  25,
v.   3, pg. ``f30!}.} careless exposing\index{Superiors:not to:expose inferiors to temptation ``f10!}, or
leaving\index{Superiors:not to:leave inferiors in wrong ``f10!} them to wrong, temptation\index{Temptation:temptation
``f10!}, and danger\index{Superiors:not to:leave inferiors in danger ``f10!};\footnote{\ Genesis 38:11, 26\index{01
Genesis:ch.  38, v.  11, 26, pg. ``f30!}; Acts 18:17\index{44 Acts:ch.  18, v.  17, pg. ``f30!}.}
provoking\index{Sin:provoking others to commit ``f10!} them to wrath;\index{Wrath:provoking others to
``f10!}\footnote{\ Ephesians 6:4\index{49 Ephesians:ch.   6, v.   4, pg. ``f30!}.} or any way
dishonoring\index{Superiors:not to:dishonor themselves through sin ``f10!} themselves, or lessening\index{Superiors:not
to:lessen authority by bad behaviour ``f10!} their authority, by an unjust\index{unjust behaviour ``f10!}, indiscreet,
rigorous\index{Superiors:not to:be rigorous ``f10!}, or remiss behavior\index{Sin:remiss behavior by superiors
``f10!}.\footnote{\ Genesis 9:21\index{01 Genesis:ch.   9, v.  21, pg. ``f30!}; 1 Kings 1:6\index{11 1 Kings:ch.   1,
v.   6, pg. ``f30!}; 1 Kings 12:13-16\index{11 1 Kings:ch.  12, v.  13-16, pg. ``f30!}; Samuel 2:29-31.}


\bigskip

Q. 131 \textit{What are the duties of equals}\index{Equals:duties of ``f10!}\textit{?}

A. The duties of equals\index{Equals:duties of ``f10!} are, to regard the dignity\index{Equals:regard dignity of others
``f10!} and worth of\index{Equals:regard worth of others ``f10!} each other,\footnote{\ 1 Peter 2:17\index{60 1
Peter:ch.   2, v.  17, pg. ``f30!}.} in giving honor to go\index{Love for others:seek neighbours' honour ``f10!} one
before another;\footnote{\ Romans 12:10\index{45 Romans:ch.  12, v.  10, pg. ``f30!}.} and to rejoice in
each\index{Love for others:rejoice in:gifts of others ``f10!} other's gifts and advancement\index{Love for
others:acknowledge gifts and graces of others ``f10!}, as their own.\footnote{\ Romans 12:15\index{45 Romans:ch.  12,
v.  15, pg. ``f30!}{}-16\index{45 Romans:ch.  12, v.  15-16, pg. ``f30!}; Philippians 2:3-4\index{50 Philippians:ch.  
2, v.   3-4, pg. ``f30!}.}


\bigskip

Q. 132 \textit{What are the sins of equals}\index{Equals:sins of ``f10!}\textit{?}

A. The sins of equals\index{Equals:sins of ``f10!} are, besides the neglect of the duties required\index{God's
law:commandments:duties:required by God ``f10!},\footnote{\ Romans 13:8\index{45 Romans:ch.  13, v.   8, pg. ``f30!}.}
the undervaluing\index{Equals:not to undervalue worth of another ``f10!} of the worth,\footnote{\ 2 Timothy
3:3\index{55 2 Timothy:ch.   3, v.   3, pg. ``f30!}.} envying\index{Envy:envy ``f10!} the gifts,\footnote{\ Acts
7:9\index{44 Acts:ch.   7, v.   9, pg. ``f30!}; Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}6\index{48
Galatians:ch.   5, v.  26, pg. ``f30!}.} grieving at\index{Envy:grieving at success of others ``f10!} the advancement
of\index{Envy:advancement of others ``f10!} prosperity one\index{Envy:prosperity of another person ``f10!} of
another;\footnote{\ Numbers 12:2\index{04 Numbers:ch.  12, v.   2, pg. ``f30!}; Esther 6:12-13\index{17 Esther:ch.   6,
v.  12-13, pg. ``f30!}.} and usurping\index{Equals:not usurp pre-eminence over one another ``f10!} pre-eminence one
over another.\footnote{\ 3 John 1:9\index{64 3 John:v.   1, v.   9, pg. ``f30!}; Luke 22:24\index{42 Luke:ch.  22, v. 
24, pg. ``f30!}.}


\bigskip


\bigskip


Q. 133 \textit{What is the reason annexed to the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{,
the more to enforce it}\index{God's law:commandments:words of enforcement ``f10!}\textit{?}

A. The reason annexed to the fifth commandment\index{God's law:commandments:fifth ``f10!}, in these words, That thy days
may be long upon the land which the Lord thy God gives you,\footnote{\ Exodus 20:12\index{02 Exodus:ch.  20, v.  12,
pg. ``f30!}.} is an express promise of long\index{God's law:commandments:promises included ``f10!} life and prosperity,
as far as it shall serve for God's glory\index{God's law:commandments:obedience:for God's glory ``f10!} and their own
good, to all such as keep this commandment.\footnote{\ Deuteronomy 5:16\index{05 Deuteronomy:ch.   5, v.  16, pg.
``f30!}; 1 Kings 8:25\index{11 1 Kings:ch.   8, v.  25, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2,
pg. ``f30!}{}-3\index{49 Ephesians:ch.   6, v.   2-3, pg. ``f30!}.}


\bigskip

Q. 134 \textit{Which is the sixth commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The sixth commandment\index{God's law:commandments:sixth ``f10!} is, You shalt not kill\index{murder, unjust killing
``f10!}.\footnote{\ Exodus 20:13\index{02 Exodus:ch.  20, v.  13, pg. ``f30!}.}


\bigskip

Q. 135 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
sixth commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the sixth commandment\index{God's
law:commandments:sixth ``f10!} are, all careful studies, and lawful endeavors\index{Human life:protect with:lawful
endeavors ``f10!}, to preserve the\index{Human life:preservation ``f10!} life of ourselves\footnote{\ Ephesians
5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}8-29\index{49 Ephesians:ch.   5, v.  28-29, pg. ``f30!}.} and
others\footnote{\ 1 Kings 18:4\index{11 1 Kings:ch.  18, v.   4, pg. ``f30!}.} by resisting\index{Temptation:resisting
``f10!} all thoughts and\index{Love for others:managing thoughts toward others ``f10!} purposes,\footnote{\ Jeremiah
26:15-16\index{24 Jeremiah:ch.  26, v.  15-16, pg. ``f30!}; Acts 23:12, 16-17, 21, 27\index{44 Acts:ch.  23, v.  12,
16-17, 21, 27, pg. ``f30!}.} subduing all passions\index{Sanctification:subduing all passions
``f10!},\footnote{\ Ephesians 4:26-27\index{49 Ephesians:ch.   4, v.  26-27, pg. ``f30!}.} and avoiding all
occasions,\footnote{\ 2 Samuel 2:22\index{10 2 Samuel:ch.   2, v.  22, pg. ``f30!}; Deuteronomy 22:8\index{05
Deuteronomy:ch.  22, v.   8, pg. ``f30!}.} temptations\index{Temptation:temptation ``f10!},\footnote{\ Matthew
4:6-7\index{40 Matthew:ch.   4, v.   6-7, pg. ``f30!}; Proverbs 1:10-11, 15-16\index{20 Proverbs:ch.   1, v.  10-11,
15-16, pg. ``f30!}.} and practices, which tend to the unjust\index{murder, unjust killing ``f10!} taking away the life
of any;\footnote{\ 1 Samuel 24:2\index{09 1 Samuel:ch.  24, v.   2, pg. ``f30!}; 1 Samuel 26:9-11\index{09 1 Samuel:ch.
 26, v.   9-11, pg. ``f30!}; Genesis 37:21-22\index{01 Genesis:ch.  37, v.  21-22, pg. ``f30!}.} by just
defense\index{Human life:defend ``f10!} thereof against violence\index{Civil magistrate:to defend citizens
``f10!},\footnote{\ Psalm 82:4\index{19 Psalms:ch.  82, v.   4, pg. ``f30!}; Proverbs 24:11-12\index{20 Proverbs:ch. 
24, v.  11-12, pg. ``f30!}; 1 Samuel 14:45\index{09 1 Samuel:ch.  14, v.  45, pg. ``f30!}.}
patient\index{Patience:patient bearing ``f10!} bearing of the hand of God,\footnote{\ James 5:7-11\index{59 James:ch.  
5, v.   7-11, pg. ``f30!}; Hebrews 12:9\index{58 Hebrews:ch.  12, v.   9, pg. ``f30!}.}
quietness\index{Conscience:quietness of mind ``f10!} of mind,\footnote{\ 1 Thessalonians 4:11\index{52 1
Thessalonians:ch.   4, v.  11, pg. ``f30!}; 1 Peter 3:3-4\index{60 1 Peter:ch.   3, v.   3-4, pg. ``f30!}; Psalm
37:8-11\index{19 Psalms:ch.  37, v.   8-11, pg. ``f30!}.} cheerfulness\index{cheerfulness ``f10!} of
spirit;\footnote{\ Proverbs 17:22\index{20 Proverbs:ch.  17, v.  22, pg. ``f30!}.} a sober use of
meat\index{Temperance:use of food, sleep, medicines, recreation ``f10!},\footnote{\ Proverbs 25:16, 27\index{20
Proverbs:ch.  25, v.  16, 27, pg. ``f30!}.} drink,\footnote{\ 1 Timothy 5:23\index{54 1 Timothy:ch.   5, v.  23, pg.
``f30!}.} physic,\footnote{\ Isaiah 38:21\index{23 Isaiah:ch.  38, v.  21, pg. ``f30!}.} sleep,\footnote{\ Psalm
127:2\index{19 Psalms:ch. 127, v.   2, pg. ``f30!}.} labor,\footnote{\ Ecclesiastes 5:1\index{21 Ecclesiastes:ch.   5,
v.   1, pg. ``f30!}2\index{21 Ecclesiastes:ch.   5, v.  12, pg. ``f30!}; 2 Thessalonians 3:1\index{53 2
Thessalonians:ch.   3, v.   1, pg. ``f30!}0, 12\index{53 2 Thessalonians:ch.   3, v.  10, 12, pg. ``f30!}; Proverbs
16:26\index{20 Proverbs:ch.  16, v.  26, pg. ``f30!}.} and recreations;\index{Human life:protect with:balanced labour
and recreation ``f10!}\footnote{\ Ecclesiastes 3:4, 11\index{21 Ecclesiastes:ch.   3, v.   4, 11, pg. ``f30!}.} by
charitable\index{Love for others:charitable thoughts ``f10!} thoughts,\footnote{\ 1 Samuel 19:4-5\index{09 1 Samuel:ch.
 19, v.   4-5, pg. ``f30!}; 1 Samuel 22:13-14\index{09 1 Samuel:ch.  22, v.  13-14, pg. ``f30!}.}
love,\footnote{\ Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}0\index{45 Romans:ch.  13, v.  10, pg.
``f30!}.} compassion\index{Love for others:compassion ``f10!},\footnote{\ Luke 10:33-34\index{42 Luke:ch.  10, v. 
33-34, pg. ``f30!}.} meekness\index{Love for others:meekness ``f10!}, gentleness\index{Love for others:gentleness
``f10!}, kindness\index{Love for others:kindness ``f10!};\footnote{\ Colossians 3:12-13\index{51 Colossians:ch.   3, v.
 12-13, pg. ``f30!}.} peaceable\index{Love for others:peaceable ``f10!},\footnote{\ James 3:1\index{59 James:ch.   3,
v.   1, pg. ``f30!}7\index{59 James:ch.   3, v.  17, pg. ``f30!}.} mild\index{Love for others:mildness ``f10!} and
courteous\index{Love for others:courteous speech ``f10!} speeches and behavior\index{Good works:behavior
``f10!};\footnote{\ 1 Peter 3:8-11\index{60 1 Peter:ch.   3, v.   8-11, pg. ``f30!}; Proverbs 15:1\index{20
Proverbs:ch.  15, v.   1, pg. ``f30!}; Judges 8:1-3\index{07 Judges:ch.   8, v.   1-3, pg. ``f30!}.}
forbearance\index{Patience:forbearance ``f10!}, readiness to be reconciled\index{Forgiveness:readiness to be reconciled
``f10!}, patient bearing and forgiving of\index{Forgiveness:forgiving ``f10!} injuries\index{Forgiveness:of injuries
``f10!}, and requiting good for evil\index{Love for others:repaying good for evil ``f10!};\footnote{\ Matthew
5:24\index{40 Matthew:ch.   5, v.  24, pg. ``f30!}; Ephesians 4:2, 32\index{49 Ephesians:ch.   4, v.   2, 32, pg.
``f30!}; Romans 12:17, 20-21\index{45 Romans:ch.  12, v.  17, 20-21, pg. ``f30!}.} comforting\index{Love for
others:comforting others ``f10!} and succoring the distressed\index{Human life:succor distressed ``f10!}, and
protecting and\index{Human life:protect innocent ``f10!} defending the innocent.\footnote{\ 1 Thessalonians
5:14\index{52 1 Thessalonians:ch.   5, v.  14, pg. ``f30!}; Job 31:1\index{18 Job:ch.  31, v.   1, pg.
``f30!}9-20\index{18 Job:ch.  31, v.  19-20, pg. ``f30!}; Matthew 25:35-36\index{40 Matthew:ch.  25, v.  35-36, pg.
``f30!}; Proverbs 31:8-9\index{20 Proverbs:ch.  31, v.   8-9, pg. ``f30!}.}


\bigskip

Q. 136 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the sixth
commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the sixth commandment\index{God's
law:commandments:sixth ``f10!} are, all taking away the life of ourselves,\footnote{\ Acts 16:28\index{44 Acts:ch.  16,
v.  28, pg. ``f30!}.} or of others,\footnote{\ Genesis 9:6\index{01 Genesis:ch.   9, v.   6, pg. ``f30!}.} except in
case of public justice\index{Civil magistrate:to maintain justice ``f10!},\footnote{\ Numbers 35:31, 33\index{04
Numbers:ch.  35, v.  31, 33, pg. ``f30!}.} lawful war,\footnote{\ Jeremiah 48:10\index{24 Jeremiah:ch.  48, v.  10, pg.
``f30!}; Deuteronomy 20:1-20\index{05 Deuteronomy:ch.  20, v.   1-20, pg. ``f30!}.} or necessary defense\index{Human
life:defend ``f10!};\footnote{\ Exodus 22:2-3\index{02 Exodus:ch.  22, v.   2-3, pg. ``f30!}.} the neglecting
or\index{Human life:not neglect or withdraw preservation ``f10!} withdrawing the lawful and necessary\index{Human
life:protect with:lawful endeavors ``f10!} means of preservation\index{Human life:preservation ``f10!} of
life;\footnote{\ Matthew 25:42-43\index{40 Matthew:ch.  25, v.  42-43, pg. ``f30!}; James 2:15-16\index{59 James:ch.  
2, v.  15-16, pg. ``f30!}; Ecclesiastes 6:1-2\index{21 Ecclesiastes:ch.   6, v.   1-2, pg. ``f30!}.} sinful
anger\index{anger ``f10!},\footnote{\ Matthew 5:22\index{40 Matthew:ch.   5, v.  22, pg. ``f30!}.} hatred,\footnote{\ 1
John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}5\index{62 1 John:ch.   3, v.  15, pg. ``f30!}; Leviticus
19:17\index{03 Leviticus:ch.  19, v.  17, pg. ``f30!}.} envy\index{Human life:no anger, hatred, envy, revenge
``f10!},\footnote{\ Proverbs 14:30\index{20 Proverbs:ch.  14, v.  30, pg. ``f30!}.} desire of
revenge\index{Revenge:man's desire for ``f10!};\footnote{\ Romans 12:19\index{45 Romans:ch.  12, v.  19, pg. ``f30!}.}
all excessive passions\index{Love for others:passions, subduing ``f10!},\footnote{\ Ephesians 4:3\index{49
Ephesians:ch.   4, v.   3, pg. ``f30!}1\index{49 Ephesians:ch.   4, v.  31, pg. ``f30!}.} distracting
cares\index{Worldly:cares ``f10!};\footnote{\ Matthew 6:31, 34\index{40 Matthew:ch.   6, v.  31, 34, pg. ``f30!}.}
immoderate use of meat\index{Temperance:use of food, sleep, medicines, recreation ``f10!}, drink,\footnote{\ Luke
21:34\index{42 Luke:ch.  21, v.  34, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}3\index{45
Romans:ch.  13, v.  13, pg. ``f30!}.} labor,\footnote{\ Ecclesiastes 2:22-23\index{21 Ecclesiastes:ch.   2, v.  22-23,
pg. ``f30!}; Ecclesiastes 12:1\index{21 Ecclesiastes:ch.  12, v.   1, pg. ``f30!}2\index{21 Ecclesiastes:ch.  12, v. 
12, pg. ``f30!}.} and recreations;\index{Human life:protect with:balanced labour and recreation
``f10!}\footnote{\ Isaiah 5:12\index{23 Isaiah:ch.   5, v.  12, pg. ``f30!}.} provoking\index{Sin:provoking others to
commit ``f10!} words,\footnote{\ Proverbs 12:18\index{20 Proverbs:ch.  12, v.  18, pg. ``f30!}; Proverbs 15:1\index{20
Proverbs:ch.  15, v.   1, pg. ``f30!}.} oppression\index{Human life:no oppression ``f10!},\footnote{\ Ezekiel
18:18\index{26 Ezekiel:ch.  18, v.  18, pg. ``f30!}; Exodus 1:14\index{02 Exodus:ch.   1, v.  14, pg. ``f30!}.}
quarreling\index{quarreling ``f10!},\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}5\index{48
Galatians:ch.   5, v.  15, pg. ``f30!}; Proverbs 23:29\index{20 Proverbs:ch.  23, v.  29, pg. ``f30!}.}
striking\index{Human life:no striking ``f10!}, wounding\index{Human life:no wounding ``f10!},\footnote{\ Numbers
35:16-18, 21\index{04 Numbers:ch.  35, v.  16-18, 21, pg. ``f30!}.} and whatsoever else tends to the destruction of the
life of any.\footnote{\ Exodus 21:18-36\index{02 Exodus:ch.  21, v.  18-36, pg. ``f30!}.}


\bigskip

Q. 137 \textit{Which is the seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The seventh commandment\index{God's law:commandments:seventh ``f10!} is, You shalt not commit adultery\index{Sexual
immorality:adultery ``f10!}.\footnote{\ Exodus 20:14\index{02 Exodus:ch.  20, v.  14, pg. ``f30!}.}


\bigskip

Q. 138 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the seven\index{Sabbath:one whole
day in seven ``f10!}th commandment are, chastity\index{Chastity:heart, speech, behaviour ``f10!} in body, mind,
affections\index{Sexual purity:affections, words, and behaviour ``f10!},\footnote{\ 1 Thessalonians 4:4\index{52 1
Thessalonians:ch.   4, v.   4, pg. ``f30!}; Job 31:1\index{18 Job:ch.  31, v.   1, pg. ``f30!}; 1 Corinthians
7:34\index{46 1 Corinthians:ch.   7, v.  34, pg. ``f30!}.} words,\footnote{\ Colossians 4:6\index{51 Colossians:ch.  
4, v.   6, pg. ``f30!}.} and behavior\index{Good works:behavior ``f10!};\footnote{\ 1 Peter 2:3\index{60 1 Peter:ch.  
2, v.   3, pg. ``f30!}.} and the preservation of it\index{Sexual purity:preservation ``f10!} in ourselves and
others;\footnote{\ 1 Corinthians 7:2, 35-36\index{46 1 Corinthians:ch.   7, v.   2, 35-36, pg. ``f30!}.}
watchfulness\index{Temptation:watchfulness required to avoid ``f10!} over the eyes and all the senses;\footnote{\ Job
31:1\index{18 Job:ch.  31, v.   1, pg. ``f30!}.} temperance\index{Temperance:sexual ``f10!},\footnote{\ Acts
24:24\index{44 Acts:ch.  24, v.  24, pg. ``f30!}.} keeping of chaste\index{Chastity:chaste company ``f10!}
company,\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.   2, v.   1, pg. ``f30!}6-20\index{20 Proverbs:ch.   2, v. 
16-20, pg. ``f30!}.} modesty\index{Apparel:modest ``f10!} in apparel;\footnote{\ 1 Timothy 2:9\index{54 1 Timothy:ch.  
2, v.   9, pg. ``f30!}.} marriage by\index{Marriage:reasons for ``f10!} those that have not the gift of
continency\index{Sexual purity:continency (sexual) ``f10!},\footnote{\ 1 Corinthians 7:2, 9\index{46 1 Corinthians:ch. 
 7, v.   2, 9, pg. ``f30!}.} conjugal\index{Sexual purity:conjugal love ``f10!} love,\footnote{\ Proverbs
5:19-20\index{20 Proverbs:ch.   5, v.  19-20, pg. ``f30!}.} and cohabitation\index{cohabitation ``f10!};\footnote{\ 1
Peter 3:7\index{60 1 Peter:ch.   3, v.   7, pg. ``f30!}.} diligent labor\index{Calling:diligence in ``f10!} in our
callings\index{Calling:in life ``f10!};\footnote{\ Proverbs 31:11, 27-28\index{20 Proverbs:ch.  31, v.  11, 27-28, pg.
``f30!}.} shunning\index{Sexual purity:shunning all uncleanness ``f10!} all occasions of uncleanness\index{Sexual
immorality:uncleanness ``f10!}, and resisting\index{Temptation:resisting ``f10!}
temptations\index{Temptation:temptation ``f10!} thereunto.\footnote{\ Proverbs 5:8\index{20 Proverbs:ch.   5, v.   8,
pg. ``f30!}; Genesis 39:8-10\index{01 Genesis:ch.  39, v.   8-10, pg. ``f30!}.}


\bigskip

Q. 139 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the seven\index{Sabbath:one whole day in
seven ``f10!}th commandment, besides the neglect of the duties required\index{God's law:commandments:duties:required by
God ``f10!},\footnote{\ Proverbs 5:7\index{20 Proverbs:ch.   5, v.   7, pg. ``f30!}.} are, adultery\index{Sexual
immorality:adultery ``f10!}, fornication\index{Sexual immorality:fornication ``f10!},\footnote{\ Hebrews 13:4\index{58
Hebrews:ch.  13, v.   4, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}9\index{48
Galatians:ch.   5, v.  19, pg. ``f30!}.} rape\index{rape ``f10!}, incest\index{Consanguinity:incest
``f10!},\footnote{\ 2 Samuel 13:14\index{10 2 Samuel:ch.  13, v.  14, pg. ``f30!}; 1 Corinthians 5:1\index{46 1
Corinthians:ch.   5, v.   1, pg. ``f30!}.} sodomy\index{Sexual immorality:sodomy ``f10!}, and all unnatural
lusts\index{Sexual immorality:lust ``f10!};\footnote{\ Romans 1:24, 26-27\index{45 Romans:ch.   1, v.  24, 26-27, pg.
``f30!}; Leviticus 20:15-16\index{03 Leviticus:ch.  20, v.  15-16, pg. ``f30!}.} all unclean\index{Sexual
immorality:unclean imaginations and thoughts ``f10!} imaginations, thoughts, purposes, and affections\index{Sexual
immorality:vile affections ``f10!};\footnote{\ Matthew 5:28\index{40 Matthew:ch.   5, v.  28, pg. ``f30!}; Matthew
15:19\index{40 Matthew:ch.  15, v.  19, pg. ``f30!}; Colossians 3:5\index{51 Colossians:ch.   3, v.   5, pg. ``f30!}.}
all corrupt or\index{Sexual immorality:corrupt communications ``f10!} filthy\index{Sexual immorality:filthy
communication (speaking or listening) ``f10!} communications, or listening thereunto;\footnote{\ Ephesians
5:3-4\index{49 Ephesians:ch.   5, v.   3-4, pg. ``f30!}; Proverbs 7:5, 21-22\index{20 Proverbs:ch.   7, v.   5, 21-22,
pg. ``f30!}.} wanton\index{Sexual immorality:wanton looks ``f10!} looks,\footnote{\ Isaiah 3:16\index{23 Isaiah:ch.  
3, v.  16, pg. ``f30!}; 2 Peter 2:14\index{61 2 Peter:ch.   2, v.  14, pg. ``f30!}.} impudent\index{impudence ``f10!}
or light behavior\index{Sin:light behavior ``f10!}, immodest\index{Apparel:immodest ``f10!}
apparel;\footnote{\ Proverbs 7:10, 13\index{20 Proverbs:ch.   7, v.  10, 13, pg. ``f30!}.}
prohibiting\index{Marriage:lawful:not to be prohibited ``f10!} of lawful,\index{Marriage:lawful:for all
``f10!}\footnote{\ 1 Timothy 4:3\index{54 1 Timothy:ch.   4, v.   3, pg. ``f30!}.} and
dispensing\index{Marriage:unlawful, to be dispensed ``f10!} with unlawful marriages\index{Marriage:unlawful
``f10!};\footnote{\ Leviticus 18:1-21\index{03 Leviticus:ch.  18, v.   1-21, pg. ``f30!}; Mark 6:18\index{41 Mark:ch.  
6, v.  18, pg. ``f30!}; Malachi 2:11-12\index{39 Malachi:ch.   2, v.  11-12, pg. ``f30!}.} allowing, tolerating,
keeping of stews\index{brothel (stew) ``f10!}, and resorting\index{Sexual immorality:resorting to prostitutes ``f10!}
to them;\footnote{\ 1 Kings 15:12\index{11 1 Kings:ch.  15, v.  12, pg. ``f30!}; 2 Kings 23:7\index{12 2 Kings:ch.  23,
v.   7, pg. ``f30!}; Deuteronomy 23:17-18\index{05 Deuteronomy:ch.  23, v.  17-18, pg. ``f30!}; Leviticus
19:29\index{03 Leviticus:ch.  19, v.  29, pg. ``f30!}; Jeremiah 5:7\index{24 Jeremiah:ch.   5, v.   7, pg. ``f30!};
Proverbs 7:24-27\index{20 Proverbs:ch.   7, v.  24-27, pg. ``f30!}.} entangling vows of single\index{Oaths and Vows:not
to:perpetual single life ``f10!} life,\footnote{\ Matthew 19:10-11\index{40 Matthew:ch.  19, v.  10-11, pg. ``f30!}.}
undue delay\index{Marriage:no undue delay ``f10!} of marriage;\footnote{\ 1 Corinthians 7:7-9\index{46 1
Corinthians:ch.   7, v.   7-9, pg. ``f30!}; Genesis 38:26\index{01 Genesis:ch.  38, v.  26, pg. ``f30!}.} having more
wives or husbands\index{Polygamy:multiple wives or husbands ``f10!} than one at the same time;\index{Divorce:unjust
``f10!}\footnote{\ Malachi 2:14-15\index{39 Malachi:ch.   2, v.  14-15, pg. ``f30!}; Matthew 19:5\index{40 Matthew:ch. 
19, v.   5, pg. ``f30!}.} unjust divorce\index{Divorce:divorce ``f10!},\footnote{\ Malachi 2:16\index{39 Malachi:ch.  
2, v.  16, pg. ``f30!}; Matthew 5:32\index{40 Matthew:ch.   5, v.  32, pg. ``f30!}.} or
desertion\index{Marriage:desertion ``f10!};\footnote{\ 1 Corinthians 7:12-13\index{46 1 Corinthians:ch.   7, v.  12-13,
pg. ``f30!}.} idleness\index{idleness ``f10!}, gluttony\index{gluttony ``f10!}, drunkenness\index{drunkenness
``f10!},\footnote{\ Ezekiel 16:49\index{26 Ezekiel:ch.  16, v.  49, pg. ``f30!}; Proverbs 23:30-33\index{20
Proverbs:ch.  23, v.  30-33, pg. ``f30!}.} unchaste\index{Sexual immorality:unchaste company and thoughts ``f10!}
company;\footnote{\ Genesis 39:10\index{01 Genesis:ch.  39, v.  10, pg. ``f30!}; Proverbs 5:8\index{20 Proverbs:ch.  
5, v.   8, pg. ``f30!}.} lascivious\index{lascivious ``f10!} songs, books, pictures\index{Sexual
immorality:provocative, songs, books and pictures ``f10!}, dancings\index{dancing ``f10!}, stage\index{stage plays
``f10!} plays\index{plays, stage ``f10!};\footnote{\ Ephesians 5:4\index{49 Ephesians:ch.   5, v.   4, pg. ``f30!};
Ezekiel 23:14-16\index{26 Ezekiel:ch.  23, v.  14-16, pg. ``f30!}; Isaiah 3:16\index{23 Isaiah:ch.   3, v.  16, pg.
``f30!}; Isaiah 23:15-17\index{23 Isaiah:ch.  23, v.  15-17, pg. ``f30!}; Mark 6:22\index{41 Mark:ch.   6, v.  22, pg.
``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}3\index{45 Romans:ch.  13, v.  13, pg. ``f30!}; 1
Peter 4:3\index{60 1 Peter:ch.   4, v.   3, pg. ``f30!}.} and all other provocations to, or acts of uncleanness,
either\index{Sexual immorality:uncleanness ``f10!} in ourselves or others.\footnote{\ 2 Kings 9:30\index{12 2 Kings:ch.
  9, v.  30, pg. ``f30!}; Jeremiah 4:30\index{24 Jeremiah:ch.   4, v.  30, pg. ``f30!}; Ezekiel 23:40\index{26
Ezekiel:ch.  23, v.  40, pg. ``f30!}.}


\bigskip

Q. 140 \textit{Which is the eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The eighth\index{God's law:commandments:eighth ``f10!} commandment is, You shalt not steal\index{Theft:stealing
``f10!}.\footnote{\ Exodus 20:15\index{02 Exodus:ch.  20, v.  15, pg. ``f30!}.}


\bigskip

Q. 141 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the eighth\index{God's
law:commandments:eighth ``f10!} commandment are, truth\index{Truth:speak, and only ``f10!},
faithfulness\index{faithfulness ``f10!}, and justice in contracts and\index{Contracts:justice in ``f10!}
commerce\index{commerce ``f10!} between man and man;\footnote{\ Psalm 15:2\index{19 Psalms:ch.  15, v.   2, pg.
``f30!}, 4\index{19 Psalms:ch.  15, v.   2, 4, pg. ``f30!}; Zechariah 7:4, 10\index{38 Zechariah:ch.   7, v.   4, 10,
pg. ``f30!}; Zechariah 8:16\index{38 Zechariah:ch.   8, v.  16, pg. ``f30!}{}-17\index{38 Zechariah:ch.   8, v.  16-17,
pg. ``f30!}.} rendering\index{Contracts:rendering what is due ``f10!} to everyone his due;\footnote{\ Romans
13:7\index{45 Romans:ch.  13, v.   7, pg. ``f30!}.} restitution\index{Punishment:restitution of goods unlawfully
detained ``f10!} of goods unlawfully detained from the right owners\index{Theft:unlawfully detaining goods ``f10!}
thereof;\footnote{\ Leviticus 6:2-5\index{03 Leviticus:ch.   6, v.   2-5, pg. ``f30!}; Luke 19:8\index{42 Luke:ch.  19,
v.   8, pg. ``f30!}.} giving and lending\index{lending freely ``f10!} freely, according to our abilities, and the
necessities of others;\footnote{\ Luke 6:30, 38\index{42 Luke:ch.   6, v.  30, 38, pg. ``f30!}; 1 John 3:1\index{62 1
John:ch.   3, v.   1, pg. ``f30!}7\index{62 1 John:ch.   3, v.  17, pg. ``f30!}; Ephesians 4:28\index{49 Ephesians:ch. 
 4, v.  28, pg. ``f30!}; Galatians 6:10\index{48 Galatians:ch.   6, v.  10, pg. ``f30!}.} moderation\index{Material
things:moderation:in use ``f10!} of our judgments\index{Judgement:personal ``f10!}, wills, and affections
concerning\index{Material things:moderation:of affections towards ``f10!} worldly goods\index{Worldly:cares
``f10!};\footnote{\ 1 Timothy 6:6\index{54 1 Timothy:ch.   6, v.   6, pg. ``f30!}{}-9\index{54 1 Timothy:ch.   6, v.  
6-9, pg. ``f30!}; Galatians 6:14\index{48 Galatians:ch.   6, v.  14, pg. ``f30!}.} a provident\index{Material
things:provident care over ``f10!} care and study to get,\footnote{\ 1 Timothy 5:8\index{54 1 Timothy:ch.   5, v.   8,
pg. ``f30!}.} keep, use, and dispose these things which are necessary and convenient for the
sustentation\index{Temperance:convenient for sustentation of our nature ``f10!} of our nature, and suitable to our
condition;\footnote{\ Proverbs 27:23-27\index{20 Proverbs:ch.  27, v.  23-27, pg. ``f30!}; Ecclesiastes 2:24\index{21
Ecclesiastes:ch.   2, v.  24, pg. ``f30!}; Ecclesiastes 3:12-13\index{21 Ecclesiastes:ch.   3, v.  12-13, pg. ``f30!};
1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}7-18\index{54 1 Timothy:ch.   6, v.  17-18, pg. ``f30!};
Isaiah 38:1\index{23 Isaiah:ch.  38, v.   1, pg. ``f30!}; Matthew 11:8\index{40 Matthew:ch.  11, v.   8, pg. ``f30!}.}
a lawful calling\index{Calling:lawful ``f10!},\footnote{\ 1 Corinthians 7:20\index{46 1 Corinthians:ch.   7, v.  20,
pg. ``f30!}; Genesis 2:15\index{01 Genesis:ch.   2, v.  15, pg. ``f30!}; Genesis 3:19\index{01 Genesis:ch.   3, v.  19,
pg. ``f30!}.} and diligence in\index{Truth:diligence in ``f10!} it;\footnote{\ Ephesians 4:28\index{49 Ephesians:ch.  
4, v.  28, pg. ``f30!}; Proverbs 10:4\index{20 Proverbs:ch.  10, v.   4, pg. ``f30!}.} frugality\index{frugality
``f10!};\footnote{\ John 6:12\index{43 John:ch.   6, v.  12, pg. ``f30!}; Proverbs 21:20\index{20 Proverbs:ch.  21, v. 
20, pg. ``f30!}.} avoiding unnecessary lawsuits\index{lawsuits, unnecessary or vexatious ``f10!}\footnote{\ 1
Corinthians 6:1-9\index{46 1 Corinthians:ch.   6, v.   1-9, pg. ``f30!}.} and suretyship\index{suretyship, unnecessary,
to be avoided ``f10!}, or other like engagements\index{Truth:in all engagements ``f10!};\footnote{\ Proverbs
6:1-6\index{20 Proverbs:ch.   6, v.   1-6, pg. ``f30!}; Proverbs 11:1\index{20 Proverbs:ch.  11, v.   1, pg.
``f30!}5\index{20 Proverbs:ch.  11, v.  15, pg. ``f30!}.} and an endeavor, by\index{God's
law:commandments:obedience:endeavoring ``f10!} all just and lawful means\index{Human life:protect with:lawful endeavors
``f10!}, to procure, preserve, and\index{Truth:preservation of ``f10!} further the wealth and outward estate\index{Love
for others:advancement of others ``f10!} of others, as well as our own.\footnote{\ Leviticus 25:35\index{03
Leviticus:ch.  25, v.  35, pg. ``f30!}; Deuteronomy 22:1-4\index{05 Deuteronomy:ch.  22, v.   1-4, pg. ``f30!}; Exodus
23:4-5\index{02 Exodus:ch.  23, v.   4-5, pg. ``f30!}; Genesis 47:14, 20\index{01 Genesis:ch.  47, v.  14, 20, pg.
``f30!}; Philippians 2:4\index{50 Philippians:ch.   2, v.   4, pg. ``f30!}; Matthew 22:39\index{40 Matthew:ch.  22, v. 
39, pg. ``f30!}}


\bigskip

Q. 142 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the eighth\index{God's
law:commandments:eighth ``f10!} commandment, besides the neglect of the duties required\index{God's
law:commandments:duties:required by God ``f10!},\footnote{\ James 2:15-16\index{59 James:ch.   2, v.  15-16, pg.
``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}7\index{62 1 John:ch.   3, v.  17, pg. ``f30!}.} are,
theft\index{Theft:theft ``f10!},\footnote{\ Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}; Psalm
42:10\index{19 Psalms:ch.  42, v.  10, pg. ``f30!}.} robbery\index{Theft:robbery ``f10!},\footnote{\ Psalm
62:10\index{19 Psalms:ch.  62, v.  10, pg. ``f30!}.} man-stealing\index{kidnapping (man-stealing) ``f10!},\footnote{\ 1
Timothy 1:10\index{54 1 Timothy:ch.   1, v.  10, pg. ``f30!}.} and receiving anything\index{Theft:receiving stolen
property ``f10!} that is stolen;\footnote{\ Proverbs 29:2\index{20 Proverbs:ch.  29, v.   2, pg. ``f30!}4\index{20
Proverbs:ch.  29, v.  24, pg. ``f30!}; Psalm 50:18\index{19 Psalms:ch.  50, v.  18, pg. ``f30!}.}
fraudulent\index{Theft:fraud ``f10!} dealing,\footnote{\ 1 Thessalonians 4:6\index{52 1 Thessalonians:ch.   4, v.   6,
pg. ``f30!}.} false weights\index{Theft:weights and measures, false ``f10!} and measures,\footnote{\ Proverbs
11:1\index{20 Proverbs:ch.  11, v.   1, pg. ``f30!}; Proverbs 20:10\index{20 Proverbs:ch.  20, v.  10, pg. ``f30!}.}
removing land marks\index{Theft:removing land marks ``f10!},\footnote{\ Deuteronomy 19:14\index{05 Deuteronomy:ch.  19,
v.  14, pg. ``f30!}; Proverbs 23:10\index{20 Proverbs:ch.  23, v.  10, pg. ``f30!}.} injustice\index{unjust behaviour
``f10!} and unfaithfulness in contracts between\index{Contracts:unfaithfulness in ``f10!} man and man,\footnote{\ Amos
8:5\index{30 Amos:ch.   8, v.   5, pg. ``f30!}; Psalm 37:21\index{19 Psalms:ch.  37, v.  21, pg. ``f30!}.} or in
matters of trust\index{Theft:oppression ``f10!};\footnote{\ Luke 16:10-12\index{42 Luke:ch.  16, v.  10-12, pg.
``f30!}.} oppression\index{Human life:no oppression ``f10!},\footnote{\ Ezekiel 22:29\index{26 Ezekiel:ch.  22, v.  29,
pg. ``f30!}; Leviticus 25:17\index{03 Leviticus:ch.  25, v.  17, pg. ``f30!}.} extortion\index{Theft:extortion
``f10!},\footnote{\ Matthew 23:25\index{40 Matthew:ch.  23, v.  25, pg. ``f30!}; Ezekiel 22:12\index{26 Ezekiel:ch. 
22, v.  12, pg. ``f30!}.} usury\index{Theft:usury ``f10!},\footnote{\ Psalm 15:5\index{19 Psalms:ch.  15, v.   5, pg.
``f30!}.} bribery\index{bribery ``f10!},\footnote{\ Job 15:34\index{18 Job:ch.  15, v.  34, pg. ``f30!}.} vexatious
lawsuits\index{lawsuits, unnecessary or vexatious ``f10!},\footnote{\ 1 Corinthians 6:6-8\index{46 1 Corinthians:ch.  
6, v.   6-8, pg. ``f30!}; Proverbs 3:29-30\index{20 Proverbs:ch.   3, v.  29-30, pg. ``f30!}.} unjust
enclosures\index{enclosures ``f10!} and depopulations\index{depopulations ``f10!};\footnote{\ Isaiah 5:8\index{23
Isaiah:ch.   5, v.   8, pg. ``f30!}; Micah 2:2\index{33 Micah:ch.   2, v.   2, pg. ``f30!}.} engrossing
commodities\index{engrossing commodities ``f10!} to enhance the price\index{Theft:enhancing prices
``f10!};\footnote{\ Proverbs 11:26\index{20 Proverbs:ch.  11, v.  26, pg. ``f30!}.} unlawful
callings\index{Calling:unlawful ``f10!},\footnote{\ Acts 19:1\index{44 Acts:ch.  19, v.   1, pg. ``f30!}9,
24-25\index{44 Acts:ch.  19, v.  19, 24-25, pg. ``f30!}.} and all other unjust or sinful ways of taking or
withholding\index{Theft:withholding what belongs to another ``f10!} from our neighbor\index{Love for others:neighbour
as self ``f10!} what belongs to him, or of enriching\index{Theft:enriching self ``f10!} ourselves;\footnote{\ Job
20:19\index{18 Job:ch.  20, v.  19, pg. ``f30!}; James 5:4\index{59 James:ch.   5, v.   4, pg. ``f30!}; Proverbs
21:6\index{20 Proverbs:ch.  21, v.   6, pg. ``f30!}.} covetousness\index{Covetousness:covetousness
``f10!};\footnote{\ Luke 12:15\index{42 Luke:ch.  12, v.  15, pg. ``f30!}.} inordinate
prizing\index{Covetousness:inordinate prizing ``f10!} and affecting worldly goods\index{Worldly:cares
``f10!};\footnote{\ 1 Timothy 6:5\index{54 1 Timothy:ch.   6, v.   5, pg. ``f30!}; Colossians 3:2\index{51
Colossians:ch.   3, v.   2, pg. ``f30!}; Proverbs 23:5\index{20 Proverbs:ch.  23, v.   5, pg. ``f30!}; Psalm
42:10\index{19 Psalms:ch.  42, v.  10, pg. ``f30!}.} distrustful\index{Idolatry:distrust ``f10!} and distracting cares
and studies in getting, keeping, and using them;\footnote{\ Matthew 6:25, 31, 34\index{40 Matthew:ch.   6, v.  25, 31,
34, pg. ``f30!}; Ecclesiastes 5:1\index{21 Ecclesiastes:ch.   5, v.   1, pg. ``f30!}2\index{21 Ecclesiastes:ch.   5, v.
 12, pg. ``f30!}.} envying\index{Envy:envy ``f10!} at the prosperity of\index{Envy:prosperity of another person ``f10!}
others;\footnote{\ Psalm 37:1, 7\index{19 Psalms:ch.  37, v.   1, 7, pg. ``f30!}; Psalm 73:3\index{19 Psalms:ch.  73,
v.   3, pg. ``f30!}.} as likewise idleness\index{idleness ``f10!},\footnote{\ 2 Thessalonians 3:1\index{53 2
Thessalonians:ch.   3, v.   1, pg. ``f30!}1\index{53 2 Thessalonians:ch.   3, v.  11, pg. ``f30!}; Proverbs
18:9\index{20 Proverbs:ch.  18, v.   9, pg. ``f30!}.} prodigality\index{Theft:prodigality ``f10!}, wasteful
gaming\index{gambling, wasteful gaming ``f10!}; and all other ways whereby we do unduly prejudice
our\index{Theft:prejudice our outward estate ``f10!} own outward estate\index{Love for others:advancement of others
``f10!},\footnote{\ Proverbs 21:17\index{20 Proverbs:ch.  21, v.  17, pg. ``f30!}; Proverbs 23:20-21\index{20
Proverbs:ch.  23, v.  20-21, pg. ``f30!}; Proverbs 28:19\index{20 Proverbs:ch.  28, v.  19, pg. ``f30!}.} and
defrauding\index{Material things:defrauding ourselves of the use ``f10!} ourselves of the due use and comfort of
that\index{Material things:temporal comfort ``f10!} estate which God has given us.\footnote{\ Ecclesiastes 4:8\index{21
Ecclesiastes:ch.   4, v.   8, pg. ``f30!}; Ecclesiastes 6:2\index{21 Ecclesiastes:ch.   6, v.   2, pg. ``f30!}; 1
Timothy 5:8\index{54 1 Timothy:ch.   5, v.   8, pg. ``f30!}.}


\bigskip

Q. 143 \textit{Which is the ninth}\index{God's law:commandments:ninth ``f10!}\textit{ commandment?}

A. The ninth\index{God's law:commandments:ninth ``f10!} commandment is, You shalt not bear false witness
against\index{Falsehood:false witness against neighbour ``f10!} thy neighbor\index{Love for others:neighbour as self
``f10!}.\footnote{\ Exodus 20:16\index{02 Exodus:ch.  20, v.  16, pg. ``f30!}.}


\bigskip

Q. 144 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
ninth}\index{God's law:commandments:ninth ``f10!}\textit{ commandment?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the ninth\index{God's
law:commandments:ninth ``f10!} commandment are, the preserving\index{Truth:preservation of ``f10!} and
promoting\index{Truth:maintain and promote ``f10!} of truth between man and man,\footnote{\ Zechariah 8:16\index{38
Zechariah:ch.   8, v.  16, pg. ``f30!}.} and the good name of our neighbor\index{Love for others:neighbour as self
``f10!}, as well as our own;\footnote{\ 3 John 1:12\index{64 3 John:v.   1, v.  12, pg. ``f30!}.} appearing and
standing\index{Truth:appear and stand for ``f10!} for the truth\index{Truth:defend ``f10!};\footnote{\ Proverbs
31:8-9\index{20 Proverbs:ch.  31, v.   8-9, pg. ``f30!}.} and from the heart\index{Love for others:from heart
``f10!},\footnote{\ Psalm 15:2\index{19 Psalms:ch.  15, v.   2, pg. ``f30!}.} sincerely,\footnote{\ 2 Chronicles
19:9\index{14 2 Chronicles:ch.  19, v.   9, pg. ``f30!}.} freely,\footnote{\ 1 Samuel 19:4-5\index{09 1 Samuel:ch.  19,
v.   4-5, pg. ``f30!}.} clearly,\index{Truth:speak freely and clearly ``f10!}\footnote{\ Joshua 7:19\index{06
Joshua:ch.   7, v.  19, pg. ``f30!}.} and fully,\footnote{\ 2 Samuel 14:18-20\index{10 2 Samuel:ch.  14, v.  18-20, pg.
``f30!}.} speaking the truth\index{Truth:speak, and only ``f10!}, and only the truth, in matters of
judgment\index{Truth:in matters of judgment and justice ``f10!} and justice,\footnote{\ Leviticus 19:15\index{03
Leviticus:ch.  19, v.  15, pg. ``f30!}; Proverbs 14:5, 25\index{20 Proverbs:ch.  14, v.   5, 25, pg. ``f30!}.} and in
all other things whatsoever;\footnote{\ 2 Corinthians 1:1\index{47 2 Corinthians:ch.   1, v.   1, pg.
``f30!}7-18\index{47 2 Corinthians:ch.   1, v.  17-18, pg. ``f30!}; Ephesians 4:25\index{49 Ephesians:ch.   4, v.  25,
pg. ``f30!}.} a charitable\index{Love for others:charitable thoughts ``f10!} esteem of our
neighbors;\footnote{\ Hebrews 6:9\index{58 Hebrews:ch.   6, v.   9, pg. ``f30!}; 1 Corinthians 13:7\index{46 1
Corinthians:ch.  13, v.   7, pg. ``f30!}.} loving, desiring, and rejoicing\index{Love for others:rejoice in:good name
of others ``f10!} in their good name;\footnote{\ Romans 1:8\index{45 Romans:ch.   1, v.   8, pg. ``f30!}; 2 John
1:4\index{63 2 John:v.   1, v.   4, pg. ``f30!}; 3 John 3-4\index{64 3 John:v. 3-4, pg. ``f30!}.} sorrowing
for,\index{Love for others:sorrowing over infirmities of others ``f10!}\footnote{\ 2 Corinthians 2:4\index{47 2
Corinthians:ch.   2, v.   4, pg. ``f30!}; 2 Corinthians 12:21\index{47 2 Corinthians:ch.  12, v.  21, pg. ``f30!}.} and
covering of their infirmities;\index{Love for others:covering infirmities of others ``f10!}\footnote{\ Proverbs
17:9\index{20 Proverbs:ch.  17, v.   9, pg. ``f30!}; 1 Peter 4:8\index{60 1 Peter:ch.   4, v.   8, pg. ``f30!}.} freely
acknowledging of their\index{Love for others:acknowledge gifts and graces of others ``f10!} gifts and
graces,\footnote{\ 1 Corinthians 1:4\index{46 1 Corinthians:ch.   1, v.   4, pg. ``f30!}{}-5, 7\index{46 1
Corinthians:ch.   1, v.   4-5, 7, pg. ``f30!}; 2 Timothy 1:4-5\index{55 2 Timothy:ch.   1, v.   4-5, pg. ``f30!}.}
defending their\index{Human life:defend ``f10!} innocency;\index{Love for others:defend innocence ``f10!}\footnote{\ 1
Samuel 22:14\index{09 1 Samuel:ch.  22, v.  14, pg. ``f30!}.} a ready receiving of a\index{Truth:receiving a good
report ``f10!} good report,\footnote{\ 1 Corinthians 13:6-7\index{46 1 Corinthians:ch.  13, v.   6-7, pg. ``f30!}.} and
unwillingness\index{Truth:unwillingness to admit evil report ``f10!} to admit of an evil\index{Truth:not admit evil
report ``f10!} report,\footnote{\ Psalm 15:3\index{19 Psalms:ch.  15, v.   3, pg. ``f30!}.} concerning them;
discouraging talebearers\index{Falsehood:talebearing ``f10!},\footnote{\ Proverbs 25:23\index{20 Proverbs:ch.  25, v. 
23, pg. ``f30!}.} flatterers\index{Falsehood:flattering ``f10!},\footnote{\ Proverbs 26:24-25\index{20 Proverbs:ch. 
26, v.  24-25, pg. ``f30!}.} and slanderers\index{slander ``f10!};\footnote{\ Psalm 101:5\index{19 Psalms:ch. 101, v.  
5, pg. ``f30!}.} love and care\index{Love:for own good name ``f10!} of our own good name, and defending it when need
requires;\footnote{\ Proverbs 22:1\index{20 Proverbs:ch.  22, v.   1, pg. ``f30!}; John 8:49\index{43 John:ch.   8, v. 
49, pg. ``f30!}.} keeping of lawful\index{Truth:keeping lawful promises ``f10!} promises;\footnote{\ Psalm
15:4\index{19 Psalms:ch.  15, v.   4, pg. ``f30!}.} studying and practicing of\index{Truth:practicing of things honest
``f10!} whatsoever things are true, honest\index{Truth:honesty ``f10!}, lovely\index{Love:for what is good ``f10!}, and
of good report.\footnote{\ Philippians 4:8\index{50 Philippians:ch.   4, v.   8, pg. ``f30!}.}


\bigskip

Q. 145 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
ninth}\index{God's law:commandments:ninth ``f10!}\textit{ commandment?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the ninth\index{God's
law:commandments:ninth ``f10!} commandment are, all prejudicing\index{Falsehood:prejudicing truth or justice ``f10!}
the truth, and the good name of our neighbor\index{Love for others:neighbour as self ``f10!}s, as well as our
own,\footnote{\ 1 Samuel 17:28\index{09 1 Samuel:ch.  17, v.  28, pg. ``f30!}; 2 Samuel 1:9-10, 15-16\index{10 2
Samuel:ch.   1, v.   9-10, 15-16, pg. ``f30!}; 2 Samuel 16:3\index{10 2 Samuel:ch.  16, v.   3, pg. ``f30!}.}
especially in public judicature\index{Truth:in public judicature ``f10!};\footnote{\ Leviticus 19:15\index{03
Leviticus:ch.  19, v.  15, pg. ``f30!}; Habakkuk 1:4\index{35 Habakkuk:ch.   1, v.   4, pg. ``f30!}.} giving false
evidence,\footnote{\ Proverbs 6:16, 19\index{20 Proverbs:ch.   6, v.  16, 19, pg. ``f30!}; Proverbs 19:5\index{20
Proverbs:ch.  19, v.   5, pg. ``f30!}.} suborning false witnesses\index{Falsehood:suborning false witnesses
``f10!},\footnote{\ Acts 6:1\index{44 Acts:ch.   6, v.   1, pg. ``f30!}3\index{44 Acts:ch.   6, v.  13, pg. ``f30!}.}
wittingly appearing and pleading for an evil\index{Falsehood:pleading for evil cause ``f10!} cause, outfacing and
overbearing\index{Truth:not overbear ``f10!} the truth;\footnote{\ Jeremiah 9:3, 5\index{24 Jeremiah:ch.   9, v.   3,
5, pg. ``f30!}; Acts 24:2, 5\index{44 Acts:ch.  24, v.   2, 5, pg. ``f30!}; Psalm 3:1-4\index{19 Psalms:ch.   3, v.  
1-4, pg. ``f30!}; Psalm 12:3-4\index{19 Psalms:ch.  12, v.   3-4, pg. ``f30!}.} passing unjust\index{Falsehood:unjust
sentence ``f10!} sentence,\footnote{\ Proverbs 17:15\index{20 Proverbs:ch.  17, v.  15, pg. ``f30!}; 1 Kings
21:9-14\index{11 1 Kings:ch.  21, v.   9-14, pg. ``f30!}.} calling evil\index{Falsehood:calling evil good ``f10!} good,
and good evil\index{Falsehood:calling good evil ``f10!}; rewarding the wicked according\index{Falsehood:unjust rewards
``f10!} to the work of the righteous, and the\index{Civil magistrate:to judge justly ``f10!} righteous according to the
work of the wicked;\footnote{\ Isaiah 5:23\index{23 Isaiah:ch.   5, v.  23, pg. ``f30!}.} forgery\index{Theft:forgery
``f10!},\footnote{\ Psalm 119:69\index{19 Psalms:ch. 119, v.  69, pg. ``f30!}; Luke 16:5-7\index{42 Luke:ch.  16, v.  
5-7, pg. ``f30!}; Luke 19:8\index{42 Luke:ch.  19, v.   8, pg. ``f30!}.} concealing the truth\index{Truth:not to be
concealed ``f10!}, undue silence\index{Falsehood:undue silence in just cause ``f10!} in a just
cause,\footnote{\ Leviticus 5:1\index{03 Leviticus:ch.   5, v.   1, pg. ``f30!}; Acts 5:3\index{44 Acts:ch.   5, v.  
3, pg. ``f30!}, 8-9\index{44 Acts:ch.   5, v.   3, 8-9, pg. ``f30!}; 2 Timothy 4:6\index{55 2 Timothy:ch.   4, v.   6,
pg. ``f30!}.} and holding our peace when iniquity\index{Falsehood:withholding reproof when iniquity requires ``f10!}
calls for either a reproof from ourselves,\footnote{\ 1 Kings 1:6\index{11 1 Kings:ch.   1, v.   6, pg. ``f30!};
Leviticus 19:17\index{03 Leviticus:ch.  19, v.  17, pg. ``f30!}.} or complaint\index{Sin:to be rebuked via complaint
``f10!} to others;\footnote{\ Isaiah 59:4\index{23 Isaiah:ch.  59, v.   4, pg. ``f30!}.} speaking the
truth\index{Truth:speak, and only ``f10!} unseasonably\index{Falsehood:unseasonably speaking truth
``f10!},\footnote{\ Proverbs 29:1\index{20 Proverbs:ch.  29, v.   1, pg. ``f30!}1\index{20 Proverbs:ch.  29, v.  11,
pg. ``f30!}.} or maliciously to\index{Falsehood:malicious statements for wrong purpose ``f10!} a wrong
end,\footnote{\ 1 Samuel 22:9-10\index{09 1 Samuel:ch.  22, v.   9-10, pg. ``f30!}; Psalm 52:1\index{19 Psalms:ch.  52,
v.   1, pg. ``f30!}.} or perverting it\index{Falsehood:perverting truth ``f10!} to a wrong meaning,\footnote{\ Psalm
56:5\index{19 Psalms:ch.  56, v.   5, pg. ``f30!}; John 2:19\index{43 John:ch.   2, v.  19, pg. ``f30!}; Matthew
26:60-61\index{40 Matthew:ch.  26, v.  60-61, pg. ``f30!}.} or in doubtful\index{Falsehood:doubtful and equivocal
expressions ``f10!} and equivocal expressions, to the prejudice of truth or justice\index{Civil magistrate:to maintain
justice ``f10!};\footnote{\ Genesis 3:5\index{01 Genesis:ch.   3, v.   5, pg. ``f30!}; Genesis 26:7, 9\index{01
Genesis:ch.  26, v.   7, 9, pg. ``f30!}.} speaking untruth\index{Falsehood:speaking untruth ``f10!},\footnote{\ Isaiah
59:13\index{23 Isaiah:ch.  59, v.  13, pg. ``f30!}.} lying\index{Falsehood:lying ``f10!},\footnote{\ Leviticus
19:11\index{03 Leviticus:ch.  19, v.  11, pg. ``f30!}; Colossians 3:9\index{51 Colossians:ch.   3, v.   9, pg.
``f30!}.} slandering\index{slander ``f10!},\footnote{\ Psalm 50:20\index{19 Psalms:ch.  50, v.  20, pg. ``f30!}.}
backbiting\index{backbiting ``f10!},\footnote{\ Psalm 15:3\index{19 Psalms:ch.  15, v.   3, pg. ``f30!}.}
detracting,\footnote{\ James 4:11\index{59 James:ch.   4, v.  11, pg. ``f30!}; Jeremiah 38:4\index{24 Jeremiah:ch.  38,
v.   4, pg. ``f30!}.} tale\index{Falsehood:talebearing ``f10!} bearing,\footnote{\ Leviticus 19:16\index{03
Leviticus:ch.  19, v.  16, pg. ``f30!}.} whispering\index{Gossip:whispering ``f10!},\footnote{\ Romans 1:29-30\index{45
Romans:ch.   1, v.  29-30, pg. ``f30!}.} scoffing\index{scoffing ``f10!},\footnote{\ Genesis 21:9\index{01 Genesis:ch. 
21, v.   9, pg. ``f30!}; Galatians 4:29\index{48 Galatians:ch.   4, v.  29, pg. ``f30!}.} reviling\index{reviling God
or man ``f10!},\footnote{\ 1 Corinthians 6:10\index{46 1 Corinthians:ch.   6, v.  10, pg. ``f30!}.} rash\index{rashness
``f10!},\footnote{\ Matthew 7:1\index{40 Matthew:ch.   7, v.   1, pg. ``f30!}.} harsh\index{harshness
``f10!},\footnote{\ Acts 28:4\index{44 Acts:ch.  28, v.   4, pg. ``f30!}.} and partial censuring\index{Sin:censuring
others ``f10!};\footnote{\ Genesis 38:24\index{01 Genesis:ch.  38, v.  24, pg. ``f30!}; Romans 2:1\index{45 Romans:ch. 
 2, v.   1, pg. ``f30!}.} misconstructing\index{Falsehood:misconstructing intentions ``f10!} intentions, words, and
actions;\footnote{\ Nehemiah 6:6-8\index{16 Nehemiah:ch.   6, v.   6-8, pg. ``f30!}; Romans 3:8\index{45 Romans:ch.  
3, v.   8, pg. ``f30!}; Psalm 69:10\index{19 Psalms:ch.  69, v.  10, pg. ``f30!}; 1 Samuel 1:13-15\index{09 1
Samuel:ch.   1, v.  13-15, pg. ``f30!}; 2 Samuel 10:3\index{10 2 Samuel:ch.  10, v.   3, pg. ``f30!}.}
flattering\index{Falsehood:flattering ``f10!},\footnote{\ Psalm 12:2-3\index{19 Psalms:ch.  12, v.   2-3, pg. ``f30!}.}
vainglorious\index{vanity ``f10!} boasting\index{boasting ``f10!},\footnote{\ 2 Timothy 3:2\index{55 2 Timothy:ch.   3,
v.   2, pg. ``f30!}.} thinking or speaking too highly or too meanly of ourselves or others;\footnote{\ Luke 18:9,
11\index{42 Luke:ch.  18, v.   9, 11, pg. ``f30!}; Romans 12:16\index{45 Romans:ch.  12, v.  16, pg. ``f30!}; 1
Corinthians 4:6\index{46 1 Corinthians:ch.   4, v.   6, pg. ``f30!}; Acts 12:22\index{44 Acts:ch.  12, v.  22, pg.
``f30!}; Exodus 4:10-14\index{02 Exodus:ch.   4, v.  10-14, pg. ``f30!}.} denying the\index{Sin:denying the gifts and
graces of God ``f10!} gifts and graces of\index{Sin:aggravations:against God's grace ``f10!} God;\footnote{\ Job
4:6\index{18 Job:ch.   4, v.   6, pg. ``f30!}; Job 27:5-6\index{18 Job:ch.  27, v.   5-6, pg. ``f30!}.}
aggravating\index{Falsehood:aggravating small faults of others ``f10!} smaller faults;\footnote{\ Matthew
7:3-5\index{40 Matthew:ch.   7, v.   3-5, pg. ``f30!}.} hiding\index{Sin:aggravations:hiding ``f10!},
excusing\index{Sin:aggravations:excusing ``f10!}, or extenuating\index{Sin:extenuating ``f10!} of sins, when called to
a free confession\index{Confession:of sin ``f10!};\footnote{\ Proverbs 28:13\index{20 Proverbs:ch.  28, v.  13, pg.
``f30!}; Proverbs 30:20\index{20 Proverbs:ch.  30, v.  20, pg. ``f30!}; Genesis 3:12-13\index{01 Genesis:ch.   3, v. 
12-13, pg. ``f30!}; Genesis 4:9\index{01 Genesis:ch.   4, v.   9, pg. ``f30!}; Jeremiah 2:35\index{24 Jeremiah:ch.   2,
v.  35, pg. ``f30!}; 2 Kings 5:25\index{12 2 Kings:ch.   5, v.  25, pg. ``f30!}.} unnecessary discovering of
infirmities;\index{Falsehood:discovering infirmities ``f10!}\footnote{\ Genesis 9:22\index{01 Genesis:ch.   9, v.  22,
pg. ``f30!}; Proverbs 25:9-10\index{20 Proverbs:ch.  25, v.   9-10, pg. ``f30!}.} raising false
rumors\index{Falsehood:raising rumors ``f10!},\footnote{\ Exodus 23:1\index{02 Exodus:ch.  23, v.   1, pg. ``f30!}.}
receiving and\index{Faith:instrument:for receiving Christ ``f10!} countenancing evil\index{Falsehood:countenancing evil
reports ``f10!} reports\index{Gossip:evil reports ``f10!},\footnote{\ Proverbs 29:1\index{20 Proverbs:ch.  29, v.   1,
pg. ``f30!}2\index{20 Proverbs:ch.  29, v.  12, pg. ``f30!}.} and stopping our ears against just\index{Falsehood:not
listening to just defense ``f10!} defense;\footnote{\ Acts 7:56-57\index{44 Acts:ch.   7, v.  56-57, pg. ``f30!}; Job
31:1\index{18 Job:ch.  31, v.   1, pg. ``f30!}3-14\index{18 Job:ch.  31, v.  13-14, pg. ``f30!}.} evil
suspicion\index{Falsehood:evil suspicion ``f10!};\footnote{\ 1 Corinthians 13:5\index{46 1 Corinthians:ch.  13, v.   5,
pg. ``f30!}; 1 Timothy 6:4\index{54 1 Timothy:ch.   6, v.   4, pg. ``f30!}.} envying\index{Envy:envy ``f10!} or
grieving at\index{Envy:grieving at success of others ``f10!} the deserved credit\index{Envy:credit of others ``f10!} of
any,\footnote{\ Numbers 11:29\index{04 Numbers:ch.  11, v.  29, pg. ``f30!}; Matthew 21:15\index{40 Matthew:ch.  21, v.
 15, pg. ``f30!}.} endeavoring or desiring to impair\index{Truth:not to impair ``f10!} it,\footnote{\ Ezra
4:12-13\index{15 Ezra:ch.   4, v.  12-13, pg. ``f30!}.} rejoicing in their disgrace\index{Falsehood:rejoicing in
disgrace and infamy of others ``f10!} and infamy\index{Sin:rejoicing in disgrace and infamy of others
``f10!};\footnote{\ Jeremiah 48:27\index{24 Jeremiah:ch.  48, v.  27, pg. ``f30!}.} scornful\index{Contempt:for truth
``f10!} contempt,\footnote{\ Psalm 35:15-16, 21\index{19 Psalms:ch.  35, v.  15-16, 21, pg. ``f30!}; Matthew
27:28-29\index{40 Matthew:ch.  27, v.  28-29, pg. ``f30!}.} fond admiration\index{Falsehood:fond admiration
``f10!};\footnote{\ Jude 1:16\index{65 Jude:v.   1, v.  16, pg. ``f30!}; Acts 12:22\index{44 Acts:ch.  12, v.  22, pg.
``f30!}.} breach of lawful\index{Oaths and Vows:not to:breach lawful promises ``f10!} promises;\footnote{\ Romans
1:31\index{45 Romans:ch.   1, v.  31, pg. ``f30!}; 2 Timothy 3:3\index{55 2 Timothy:ch.   3, v.   3, pg. ``f30!}.}
neglecting such\index{Truth:not neglect good report ``f10!} things as are of good report\index{Falsehood:neglecting a
good report ``f10!},\footnote{\ 1 Samuel 2:24\index{09 1 Samuel:ch.   2, v.  24, pg. ``f30!}.} and practicing, or not
avoiding ourselves, or not hindering: What we can in others, such things as procure an\index{Falsehood:procure ill name
``f10!} ill name\index{Falsehood:causing ill name for others ``f10!}.\footnote{\ 2 Samuel 13:12-13\index{10 2
Samuel:ch.  13, v.  12-13, pg. ``f30!}; Proverbs 5:8\index{20 Proverbs:ch.   5, v.   8, pg. ``f30!}{}-9\index{20
Proverbs:ch.   5, v.   8-9, pg. ``f30!}; Proverbs 6:33\index{20 Proverbs:ch.   6, v.  33, pg. ``f30!}.}


\bigskip

Q. 146 \textit{Which is the tenth}\index{God's law:commandments:tenth ``f10!}\textit{ commandment?}

A. The tenth\index{God's law:commandments:tenth ``f10!} commandment is, You shalt not
covet\index{Covetousness:covetousness ``f10!} thy neighbor\index{Love for others:neighbour as self ``f10!}{}'s house,
you\index{Covetousness:nothing of neighbour ``f10!} shalt not covet thy neighbor's wife, nor his manservant, nor his
maidservant\index{Sabbath:servants not to be made to work ``f10!}, nor his ox, nor his ass, nor any thing that is thy
neighbor's.\footnote{\ Exodus 20:17\index{02 Exodus:ch.  20, v.  17, pg. ``f30!}.}


\bigskip

Q. 147 \textit{What are the duties required}\index{God's law:commandments:duties:required by God ``f10!}\textit{ in the
tenth}\index{God's law:commandments:tenth ``f10!}\textit{ commandment?}

A. The duties required\index{God's law:commandments:duties:required by God ``f10!} in the tenth\index{God's
law:commandments:tenth ``f10!} commandment are, such a full contentment\index{contentment ``f10!} with our own
condition,\footnote{\ Hebrews 13:5\index{58 Hebrews:ch.  13, v.   5, pg. ``f30!}; 1 Timothy 6:6\index{54 1 Timothy:ch. 
 6, v.   6, pg. ``f30!}.} and such a charitable\index{Love for others:charitable thoughts ``f10!} frame of the whole
soul\index{Love for others:with whole soul ``f10!} toward our neighbor\index{Love for others:neighbour as self ``f10!},
as that all our inward motions\index{Love for others:inward thoughts further others' good ``f10!} and affections
to\index{Covetousness:inordinate affections to others' goods ``f10!}uching\index{Love for others:advancement of others
``f10!} him, tend unto, and further all that good which is his.\footnote{\ Job 31:29\index{18 Job:ch.  31, v.  29, pg.
``f30!}; Psalm 122:7-9\index{19 Psalms:ch. 122, v.   7-9, pg. ``f30!}; 1 Timothy 1:5\index{54 1 Timothy:ch.   1, v.  
5, pg. ``f30!}; Esther 10:3\index{17 Esther:ch.  10, v.   3, pg. ``f30!}; 1 Corinthians 13:4-7\index{46 1
Corinthians:ch.  13, v.   4-7, pg. ``f30!}.}


\bigskip

Q. 148 \textit{What are the sins forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
tenth}\index{God's law:commandments:tenth ``f10!}\textit{ commandment?}

A. The sins forbidden\index{God's law:commandments:sins forbidden ``f10!} in the tenth\index{God's
law:commandments:tenth ``f10!} commandment are, discontentment\index{discontentment ``f10!} with our own
estate;\footnote{\ 1 Kings 21:4\index{11 1 Kings:ch.  21, v.   4, pg. ``f30!}; Esther 5:13\index{17 Esther:ch.   5, v. 
13, pg. ``f30!}; 1 Corinthians 10:10\index{46 1 Corinthians:ch.  10, v.  10, pg. ``f30!}.} envying\index{Envy:envy
``f10!}\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}6\index{48 Galatians:ch.   5, v.  26,
pg. ``f30!}; James 3:1\index{59 James:ch.   3, v.   1, pg. ``f30!}4, 16\index{59 James:ch.   3, v.  14, 16, pg.
``f30!}.} and grieving at\index{Envy:grieving at success of others ``f10!} the good of our neighbor\index{Love for
others:neighbour as self ``f10!},\footnote{\ Psalm 112:9-10\index{19 Psalms:ch. 112, v.   9-10, pg. ``f30!}; Nehemiah
2:10\index{16 Nehemiah:ch.   2, v.  10, pg. ``f30!}.} together with all inordinate
motions\index{Covetousness:inordinate affections to others' goods ``f10!} and affections to anything that is
his.\footnote{\ Romans 7:7\index{45 Romans:ch.   7, v.   7, pg. ``f30!}{}-8\index{45 Romans:ch.   7, v.   7-8, pg.
``f30!}; Romans 13:9\index{45 Romans:ch.  13, v.   9, pg. ``f30!}; Colossians 3:5\index{51 Colossians:ch.   3, v.   5,
pg. ``f30!}; Deuteronomy 5:21\index{05 Deuteronomy:ch.   5, v.  21, pg. ``f30!}.}


\bigskip


\bigskip


Q. 149 \textit{Is any man able perfectly to keep the commandment}\index{God's law:commandments:given to Adam
``f10!}\textit{s}\index{God's law:commandments:ten ``f10!}\textit{ of God?}

A. No man is able, either of himself,\footnote{\ James 3:2\index{59 James:ch.   3, v.   2, pg. ``f30!}; John
15:5\index{43 John:ch.  15, v.   5, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}.} or by any
grace\index{Grace:empowerment ``f10!} received in this life, perfectly to keep the commandment\index{God's
law:commandments:given to Adam ``f10!}s\index{God's law:commandments:ten ``f10!} of God;\footnote{\ Ecclesiastes
7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}0\index{21 Ecclesiastes:ch.   7, v.  20, pg. ``f30!}; 1 John 1:8,
10\index{62 1 John:ch.   1, v.   8, 10, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg.
``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}; Romans 7:18\index{45 Romans:ch.   7, v.  18, pg.
``f30!}{}-19\index{45 Romans:ch.   7, v.  18-19, pg. ``f30!}.} but does daily break the\index{Lord's Supper:bread
broken ``f10!}m\index{God's law:broken daily by all men (thought, word, deed) ``f10!} in thought,\footnote{\ Genesis
6:5\index{01 Genesis:ch.   6, v.   5, pg. ``f30!}; Genesis 8:21\index{01 Genesis:ch.   8, v.  21, pg. ``f30!}.} word,
and deed.\footnote{\ Romans 3:9\index{45 Romans:ch.   3, v.   9, pg. ``f30!}{}-19\index{45 Romans:ch.   3, v.   9-19,
pg. ``f30!}; James 3:2\index{59 James:ch.   3, v.   2, pg. ``f30!}{}-13\index{59 James:ch.   3, v.   2-13, pg.
``f30!}.}


\bigskip

Q. 150 \textit{Are all transgressions}\index{God's law:commandments:transgressions ``f10!}\textit{ of the law of God
equally heinous}\index{God's law:commandments:all transgression equally heinous ``f10!}\textit{ in themselves, and in
the sight of God?}

A. All transgressions\index{God's law:commandments:transgressions ``f10!} of the law of God are not equally
heinous\index{God's law:commandments:all transgression equally heinous ``f10!}; but some sins in themselves, and by
reason of several aggravations\index{Sin:aggravations:which compound ``f10!}, are more
heinous\index{Sin:aggravations:make more heinous ``f10!} in the sight of God than others.\footnote{\ John
19:11\index{43 John:ch.  19, v.  11, pg. ``f30!}; Ezekiel 8:6, 13, 15\index{26 Ezekiel:ch.   8, v.   6, 13, 15, pg.
``f30!}; 1 John 5:16\index{62 1 John:ch.   5, v.  16, pg. ``f30!}; Psalm 78:17, 32, 56\index{19 Psalms:ch.  78, v.  17,
32, 56, pg. ``f30!}.}


\bigskip

Q. 151 \textit{What are those aggravations}\index{Sin:aggravations:which compound ``f10!}\textit{ that make some sins
more heinous}\index{Sin:aggravations:make more heinous ``f10!}\textit{ than others?}

A. Sins receive their aggravations\index{Sin:aggravations:which compound ``f10!}, 1. From the persons
offending:\index{Sin:aggravations:maturity of person ``f10!}\footnote{\ Jeremiah 2:8\index{24 Jeremiah:ch.   2, v.   8,
pg. ``f30!}.} if they be of riper\index{Sin:aggravations:performed by person of riper age ``f10!} age,\footnote{\ Job
32:7, 9\index{18 Job:ch.  32, v.   7, 9, pg. ``f30!}; Ecclesiastes 4:13\index{21 Ecclesiastes:ch.   4, v.  13, pg.
``f30!}.} greater experience\index{Sin:aggravations:performed by person with greater experience of grace ``f10!} or
grace\index{Sin:aggravations:those with greater experience or grace ``f10!},\footnote{\ 1 Kings 11:4, 9\index{11 1
Kings:ch.  11, v.   4, 9, pg. ``f30!}.} eminent\index{Sin:aggravations:person's place or office ``f10!} for
profession\index{Faith:profession of ``f10!},\footnote{\ 2 Samuel 12:14\index{10 2 Samuel:ch.  12, v.  14, pg. ``f30!};
1 Corinthians 5:1\index{46 1 Corinthians:ch.   5, v.   1, pg. ``f30!}.} gifts,\footnote{\ James 4:17\index{59 James:ch.
  4, v.  17, pg. ``f30!}; Luke 12:47-48\index{42 Luke:ch.  12, v.  47-48, pg. ``f30!}.} place,\footnote{\ Jeremiah
5:4-5\index{24 Jeremiah:ch.   5, v.   4-5, pg. ``f30!}.} office,\footnote{\ 2 Samuel 12:7-9\index{10 2 Samuel:ch.  12,
v.   7-9, pg. ``f30!}; Ezekiel 8:11-12\index{26 Ezekiel:ch.   8, v.  11-12, pg. ``f30!}.} guides to
others,\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}7-24\index{45 Romans:ch.   2, v.  17-24, pg.
``f30!}.} and whose example is likely to be followed\index{Sin:aggravations:setting bad example ``f10!} by
others.\footnote{\ Galatians 2:11-14\index{48 Galatians:ch.   2, v.  11-14, pg. ``f30!}.} 2. From the parties
offended:\footnote{\ Matthew 21:38-39\index{40 Matthew:ch.  21, v.  38-39, pg. ``f30!}.} if immediately against
God,\footnote{\ 1 Samuel 2:25\index{09 1 Samuel:ch.   2, v.  25, pg. ``f30!}; Acts 5:4\index{44 Acts:ch.   5, v.   4,
pg. ``f30!}; Psalm 51:4\index{19 Psalms:ch.  51, v.   4, pg. ``f30!}.} his attributes\index{God's attributes:to
be:revered ``f10!},\footnote{\ Romans 2:4\index{45 Romans:ch.   2, v.   4, pg. ``f30!}.} and
worship;\index{Worship:abuses ``f10!}\footnote{\ Malachi 1:8, 14\index{39 Malachi:ch.   1, v.   8, 14, pg. ``f30!}.}
against Christ, and his grace\index{Sin:aggravations:against God's grace ``f10!};\footnote{\ Hebrews 2:2-3\index{58
Hebrews:ch.   2, v.   2-3, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}5\index{58
Hebrews:ch.  12, v.  25, pg. ``f30!}.} the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!},\footnote{\ Hebrews
10:29\index{58 Hebrews:ch.  10, v.  29, pg. ``f30!}; Matthew 12:31-32\index{40 Matthew:ch.  12, v.  31-32, pg.
``f30!}.} his witness,\footnote{\ Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.
  4, v.  30, pg. ``f30!}.} and workings\index{Sin:aggravations:against Holy Spirit's working
``f10!};\footnote{\ Hebrews 6:4-6\index{58 Hebrews:ch.   6, v.   4-6, pg. ``f30!}.} against
superiors\index{Inferiors:sins against superiors ``f10!}, men of eminency,\footnote{\ Jude 1:8\index{65 Jude:v.   1, v.
  8, pg. ``f30!}; Numbers 12:8-9\index{04 Numbers:ch.  12, v.   8-9, pg. ``f30!}; Isaiah 3:5\index{23 Isaiah:ch.   3,
v.   5, pg. ``f30!}.} and such as we stand especially related and engaged unto;\footnote{\ Proverbs 30:17\index{20
Proverbs:ch.  30, v.  17, pg. ``f30!}; 2 Corinthians 12:15\index{47 2 Corinthians:ch.  12, v.  15, pg. ``f30!}; Psalm
55:12-15\index{19 Psalms:ch.  55, v.  12-15, pg. ``f30!}.} against any of the saints\index{Believers:saints
``f10!},\footnote{\ Zephaniah 2:8, 10-11\index{36 Zephaniah:ch.   2, v.   8, 10-11, pg. ``f30!}; Matthew 18:6\index{40
Matthew:ch.  18, v.   6, pg. ``f30!}; 1 Corinthians 6:8\index{46 1 Corinthians:ch.   6, v.   8, pg. ``f30!}; Revelation
17:6\index{66 Revelation:ch.  17, v.   6, pg. ``f30!}.} particularly weak brethren\index{Sin:aggravations:cause weak to
stumble ``f10!},\footnote{\ 1 Corinthians 8:11-12\index{46 1 Corinthians:ch.   8, v.  11-12, pg. ``f30!}; Romans 14:13,
15, 21\index{45 Romans:ch.  14, v.  13, 15, 21, pg. ``f30!}.} the souls of them, or any other,\footnote{\ Ezekiel
13:19\index{26 Ezekiel:ch.  13, v.  19, pg. ``f30!}; 1 Corinthians 8:12\index{46 1 Corinthians:ch.   8, v.  12, pg.
``f30!}; Revelation 18:12-13\index{66 Revelation:ch.  18, v.  12-13, pg. ``f30!}; Matthew 23:15\index{40 Matthew:ch. 
23, v.  15, pg. ``f30!}.} and the common good of all or many.\footnote{\ 1 Thessalonians 2:15-16\index{52 1
Thessalonians:ch.   2, v.  15-16, pg. ``f30!}; Joshua 22:20\index{06 Joshua:ch.  22, v.  20, pg. ``f30!}.} 3. From the
nature and quality\index{Sin:aggravations:nature and quality of the offense ``f10!} of the offense:\footnote{\ Proverbs
6:30-35\index{20 Proverbs:ch.   6, v.  30-35, pg. ``f30!}.} if it be against the express letter of the
law,\footnote{\ Ezra 9:10-12\index{15 Ezra:ch.   9, v.  10-12, pg. ``f30!}; 1 Kings 11:9-10\index{11 1 Kings:ch.  11,
v.   9-10, pg. ``f30!}.} break many\index{Sin:aggravations:break many commands ``f10!} commandments, contain in it many
sins:\footnote{\ Colossians 3:5\index{51 Colossians:ch.   3, v.   5, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch. 
 6, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   6, v.  10, pg. ``f30!}; Proverbs 5:8\index{20 Proverbs:ch.   5, v.  
8, pg. ``f30!}{}-12\index{20 Proverbs:ch.   5, v.   8-12, pg. ``f30!}; Proverbs 6:32-33\index{20 Proverbs:ch.   6, v. 
32-33, pg. ``f30!}; Joshua 7:21\index{06 Joshua:ch.   7, v.  21, pg. ``f30!}.} if not only conceived
in\index{Sin:conceived in heart ``f10!} the heart, but breaks\index{Heart:source of sin ``f10!} forth in words and
actions,\footnote{\ James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}{}-15\index{59 James:ch.   1, v.  14-15, pg.
``f30!}; Matthew 5:22\index{40 Matthew:ch.   5, v.  22, pg. ``f30!}; Micah 2:1\index{33 Micah:ch.   2, v.   1, pg.
``f30!}.} scandalize\index{Scandalous behaviour:scandalize others ``f10!} others,\footnote{\ Matthew 18:7\index{40
Matthew:ch.  18, v.   7, pg. ``f30!}; Romans 2:23-24\index{45 Romans:ch.   2, v.  23-24, pg. ``f30!}.} and admit of no
reparation\index{Sin:aggravations:admit of no reparation ``f10!}:\footnote{\ Deuteronomy 22:22, 28-29\index{05
Deuteronomy:ch.  22, v.  22, 28-29, pg. ``f30!}; Proverbs 6:32-35\index{20 Proverbs:ch.   6, v.  32-35, pg. ``f30!}.}
if against means\index{Judgement:temporal ``f10!},\footnote{\ Matthew 11:21-24\index{40 Matthew:ch.  11, v.  21-24, pg.
``f30!}; John 15:22\index{43 John:ch.  15, v.  22, pg. ``f30!}.} mercies,\footnote{\ Isaiah 1:3\index{23 Isaiah:ch.  
1, v.   3, pg. ``f30!}; Deuteronomy 32:6\index{05 Deuteronomy:ch.  32, v.   6, pg. ``f30!}.} judgments,\footnote{\ Amos
4:8-11\index{30 Amos:ch.   4, v.   8-11, pg. ``f30!}; Jeremiah 5:8\index{24 Jeremiah:ch.   5, v.   8, pg. ``f30!}.}
light of nature,\footnote{\ Romans 1:26\index{45 Romans:ch.   1, v.  26, pg. ``f30!}{}-27\index{45 Romans:ch.   1, v. 
26-27, pg. ``f30!}.} conviction of conscience\index{Conscience:conviction of ``f10!},\footnote{\ Romans 1:32\index{45
Romans:ch.   1, v.  32, pg. ``f30!}; Daniel 5:22\index{27 Daniel:ch.   5, v.  22, pg. ``f30!}; Titus 3:1\index{56
Titus:ch.   3, v.   1, pg. ``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!}{}-11\index{56 Titus:ch.   3, v.  10-11,
pg. ``f30!}.} public or private admonition\index{Church discipline:admonition ``f10!},\footnote{\ Proverbs
29:1\index{20 Proverbs:ch.  29, v.   1, pg. ``f30!}.} censures of\index{Church discipline:censures ``f10!} the
church,\footnote{\ Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!};
Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}.} civil punishments\index{Punishment:civil
``f10!};\footnote{\ Proverbs 23:35\index{20 Proverbs:ch.  23, v.  35, pg. ``f30!}; Proverbs 27:22\index{20 Proverbs:ch.
 27, v.  22, pg. ``f30!}.} and our prayers\index{Prayer:presumptuous ``f10!}, purposes, promises\index{Oaths and
Vows:not to:be violated ``f10!},\footnote{\ Psalm 78:34-37\index{19 Psalms:ch.  78, v.  34-37, pg. ``f30!}; Jeremiah
2:20\index{24 Jeremiah:ch.   2, v.  20, pg. ``f30!}; Jeremiah 13:5-6, 20-21\index{24 Jeremiah:ch.  13, v.   5-6, 20-21,
pg. ``f30!}.} vows,\footnote{\ Ecclesiastes 5:4-6\index{21 Ecclesiastes:ch.   5, v.   4-6, pg. ``f30!}; Proverbs
20:25\index{20 Proverbs:ch.  20, v.  25, pg. ``f30!}.} covenants,\index{Covenant:human contracts and engagements
``f10!}\footnote{\ Leviticus 26:25\index{03 Leviticus:ch.  26, v.  25, pg. ``f30!}.} and engagements to God or
men:\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.   2, v.   1, pg. ``f30!}7\index{20 Proverbs:ch.   2, v.  17, pg.
``f30!}; Ezekiel 17:18-19\index{26 Ezekiel:ch.  17, v.  18-19, pg. ``f30!}.} if done
deliberately\index{Sin:aggravations:deliberate ``f10!},\footnote{\ Psalm 36:4\index{19 Psalms:ch.  36, v.   4, pg.
``f30!}.} wilfully,\footnote{\ Jeremiah 6:16\index{24 Jeremiah:ch.   6, v.  16, pg. ``f30!}.}
presumptuously\index{presumption ``f10!},\footnote{\ Numbers 15:30\index{04 Numbers:ch.  15, v.  30, pg. ``f30!};
Exodus 21:14\index{02 Exodus:ch.  21, v.  14, pg. ``f30!}.} impudently\index{impudence ``f10!},\footnote{\ Jeremiah
3:3\index{24 Jeremiah:ch.   3, v.   3, pg. ``f30!}; Proverbs 7:13\index{20 Proverbs:ch.   7, v.  13, pg. ``f30!}.}
boastingly\index{boasting ``f10!},\footnote{\ Psalm 52:1\index{19 Psalms:ch.  52, v.   1, pg. ``f30!}.}
maliciously,\index{Sin:aggravations:maliciousness and frequently ``f10!}\footnote{\ 3 John 1:10\index{64 3 John:v.   1,
v.  10, pg. ``f30!}.} frequently,\footnote{\ Numbers 14:2\index{04 Numbers:ch.  14, v.   2, pg. ``f30!}2\index{04
Numbers:ch.  14, v.  22, pg. ``f30!}.} obstinately\index{Sin:aggravations:obstinance ``f10!},\footnote{\ Zechariah
7:11-12\index{38 Zechariah:ch.   7, v.  11-12, pg. ``f30!}.} with delight\index{Sin:aggravations:delighting in
``f10!},\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.   2, v.   1, pg. ``f30!}4\index{20 Proverbs:ch.   2, v.  14, pg.
``f30!}.} continuance,\footnote{\ Isaiah 57:17\index{23 Isaiah:ch.  57, v.  17, pg. ``f30!}.} or
relapsing\index{Sin:aggravations:relapsing after repentance ``f10!} after repentance\index{Repentance:relapsing after
``f10!}.\footnote{\ Jeremiah 34:8-11\index{24 Jeremiah:ch.  34, v.   8-11, pg. ``f30!}; 2 Peter 2:20-22\index{61 2
Peter:ch.   2, v.  20-22, pg. ``f30!}.} 4. From circumstances\index{Worship:circumstances ``f10!} of time\footnote{\ 2
Kings 5:26\index{12 2 Kings:ch.   5, v.  26, pg. ``f30!}.} and place:\footnote{\ Jeremiah 7:10\index{24 Jeremiah:ch.  
7, v.  10, pg. ``f30!}; Isaiah 26:10\index{23 Isaiah:ch.  26, v.  10, pg. ``f30!}.} if on the Lord's
day,\footnote{\ Ezekiel 23:37-39\index{26 Ezekiel:ch.  23, v.  37-39, pg. ``f30!}.} or other times of divine
worship\index{Worship:times of ``f10!};\footnote{\ Isaiah 58:3-5\index{23 Isaiah:ch.  58, v.   3-5, pg. ``f30!};
Numbers 25:6-7\index{04 Numbers:ch.  25, v.   6-7, pg. ``f30!}.} or immediately before\footnote{\ 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg.
``f30!}{}-21\index{46 1 Corinthians:ch.  11, v.  20-21, pg. ``f30!}.} or after these,\footnote{\ Jeremiah 7:8-10,
14-15\index{24 Jeremiah:ch.   7, v.   8-10, 14-15, pg. ``f30!}; John 13:27, 30\index{43 John:ch.  13, v.  27, 30, pg.
``f30!}.} or other helps to prevent or remedy such miscarriages\index{Sabbath:charge of keeping directed to:governors
``f10!}:\footnote{\ Ezra 9:13-14\index{15 Ezra:ch.   9, v.  13-14, pg. ``f30!}.} if in public, or in the presence of
others, who are thereby likely to be provoked\index{Sin:provoking others to commit ``f10!} or
defiled\index{Sin:aggravations:performed on Lord's Day or during worship ``f10!}.\footnote{\ 2 Samuel 16:22\index{10 2
Samuel:ch.  16, v.  22, pg. ``f30!}; 1 Samuel 2:22-24\index{09 1 Samuel:ch.   2, v.  22-24, pg. ``f30!}.}


\bigskip

Q. 152 \textit{What does every sin deserve}\index{Sin:deserves damnation ``f10!}\textit{ at the hands of God?}

A. Every sin, even the least, being against the sovereignty\index{God's attributes:sovereign lordship
``f10!},\footnote{\ James 2:10\index{59 James:ch.   2, v.  10, pg. ``f30!}{}-11\index{59 James:ch.   2, v.  10-11, pg.
``f30!}.} goodness\index{God's attributes:goodness ``f10!},\footnote{\ Exodus 20:1-2\index{02 Exodus:ch.  20, v.   1-2,
pg. ``f30!}.} and holiness of God\index{God's attributes:holy:nature ``f10!},\footnote{\ Habakkuk 1:13\index{35
Habakkuk:ch.   1, v.  13, pg. ``f30!}; Leviticus 10:3\index{03 Leviticus:ch.  10, v.   3, pg. ``f30!}; Leviticus
11:44-45\index{03 Leviticus:ch.  11, v.  44-45, pg. ``f30!}.} and against his righteous law\index{God's law:righteous
requirements ``f10!},\footnote{\ 1 John 3:4\index{62 1 John:ch.   3, v.   4, pg. ``f30!}; Romans 7:12\index{45
Romans:ch.   7, v.  12, pg. ``f30!}.} deserves\index{Sin:deserves damnation ``f10!} his wrath and\index{Wrath:of
God:for sin ``f10!} curse,\index{Curse:of God ``f10!}\footnote{\ Ephesians 5:6\index{49 Ephesians:ch.   5, v.   6, pg.
``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}.} both in this life,\footnote{\ Lamentations
3:39\index{25 Lamentations:ch.   3, v.  39, pg. ``f30!}; Deuteronomy 28:15-68\index{05 Deuteronomy:ch.  28, v.  15-68,
pg. ``f30!}.} and that which is to come;\footnote{\ Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}.} and
cannot be expiated\index{Sin:expiated only by blood of Christ ``f10!} but by the blood\index{Jesus:shed blood for
remission of sins ``f10!} of Christ.\footnote{\ Hebrews 9:22\index{58 Hebrews:ch.   9, v.  22, pg. ``f30!}; 1 Peter
1:18\index{60 1 Peter:ch.   1, v.  18, pg. ``f30!}{}-19\index{60 1 Peter:ch.   1, v.  18-19, pg. ``f30!}.}


\bigskip

Q. 153 \textit{What does God require of us}\index{God's law:commandments:duties:required by God ``f10!}\textit{, that we
may escape}\index{Wrath:of God:requirements to escape ``f10!}\textit{ his wrath and}\index{Wrath:of God:for sin
``f10!}\textit{ curse due}\index{Curse:deserved for sin ``f10!}\textit{ to us by reason of the transgression
of}\index{God's law:commandments:transgressions ``f10!}\textit{ the law?}

A. That we may escape\index{Wrath:of God:requirements to escape ``f10!} the wrath and\index{Wrath:of God:for sin ``f10!}
curse of God\index{Curse:of God ``f10!} due to us by reason of the transgression of\index{God's
law:commandments:transgressions ``f10!} the law, he requires of us\index{Repentance:required for pardon ``f10!}
repentance to\index{Repentance:to those offended ``f10!}ward God, and faith\index{Faith:in Jesus Christ ``f10!} toward
our Lord Jesus Christ,\footnote{\ Acts 16:30\index{44 Acts:ch.  16, v.  30, pg. ``f30!}{}-31\index{44 Acts:ch.  16, v. 
30-31, pg. ``f30!}; Acts 20:21\index{44 Acts:ch.  20, v.  21, pg. ``f30!}; Matthew 3:7-8\index{40 Matthew:ch.   3, v.  
7-8, pg. ``f30!}; Luke 13:3, 5\index{42 Luke:ch.  13, v.   3, 5, pg. ``f30!}; John 3:16\index{43 John:ch.   3, v.  16,
pg. ``f30!}, 18\index{43 John:ch.   3, v.  16, 18, pg. ``f30!}.} and the diligent use\index{Grace:diligent use of means
``f10!} of the outward means whereby Christ communicates to us the benefits of his\index{Believers:benefits:of Christ's
mediation ``f10!} mediation\index{Jesus:mediator:of covenant ``f10!}.\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.  
2, v.   1, pg. ``f30!}{}-5\index{20 Proverbs:ch.   2, v.   1-5, pg. ``f30!}; Proverbs 8:33-36\index{20 Proverbs:ch.  
8, v.  33-36, pg. ``f30!}.}


\bigskip

Q. 154 \textit{What are the outward means whereby Christ communicates to us the benefits of
his}\index{Believers:benefits:of Christ's mediation ``f10!}\textit{ mediation}\index{Jesus:mediator:of covenant
``f10!}\textit{?}

A. The outward and ordinary means whereby Christ communicates to his church\index{Church:Christ's ``f10!} the benefits
of his mediation\index{Jesus:mediator:of covenant ``f10!}, are all his ordinances;
especially\index{Worship:ordinances:means of Christ's mediation ``f10!} the word,
sacraments\index{Sacraments:holy:ordinances ``f10!}, and prayer; all\index{Prayer:act of worship ``f10!} which are made
effectual\index{Worship:ordinances:made effectual for salvation ``f10!} to the elect\index{Election:elect of God
``f10!} for their salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Matthew 28:19\index{40
Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.
  2, v.  42, pg. ``f30!}, 46\index{44 Acts:ch.   2, v.  42, 46, pg. ``f30!}{}-47\index{44 Acts:ch.   2, v.  42, 46-47,
pg. ``f30!}.}


\bigskip

Q. 155 \textit{How is the word made effectual}\index{Scripture:word of God:for salvation ``f10!}\textit{ to
salvation}\index{Salvation:belief in Jesus by faith ``f10!}\textit{?}

A. The Spirit\index{Holy Spirit:Holy Spirit ``f10!} of God makes the reading\index{Scripture:to be:read and heard
``f10!}, but especially the preaching of the word, an effectual\index{Preaching:effectual means of convincing and
enlightening ``f10!} means of enlightening\index{Effectual call:enlightening minds spiritually
``f10!},\footnote{\ Nehemiah 8:8\index{16 Nehemiah:ch.   8, v.   8, pg. ``f30!}; Acts 26:18\index{44 Acts:ch.  26, v. 
18, pg. ``f30!}; Psalm 19:8\index{19 Psalms:ch.  19, v.   8, pg. ``f30!}.} convincing, and\index{Preaching:convincing
and converting sinners ``f10!} humbling\index{Humility:sinners humbled ``f10!} sinners;\footnote{\ 1 Corinthians
14:24-25\index{46 1 Corinthians:ch.  14, v.  24-25, pg. ``f30!}; 2 Chronicles 34:18-19, 26-28\index{14 2 Chronicles:ch.
 34, v.  18-19, 26-28, pg. ``f30!}.} of driving\index{Preaching:drive men to Christ ``f10!} them out of themselves, and
drawing them\index{Effectual call:drawing to Christ ``f10!} unto Christ;\footnote{\ Acts 2:37\index{44 Acts:ch.   2, v.
 37, pg. ``f30!}, 41\index{44 Acts:ch.   2, v.  37, 41, pg. ``f30!}; Acts 8:27-39\index{44 Acts:ch.   8, v.  27-39, pg.
``f30!}.} of conforming\index{Sanctification:conforming to image of Christ ``f10!} them to his
image\index{Mankind:image of God ``f10!},\footnote{\ 2 Corinthians 3:18\index{47 2 Corinthians:ch.   3, v.  18, pg.
``f30!}.} and subduing them\index{Holy Spirit:subdues will of man ``f10!} to his will;\footnote{\ 2 Corinthians
10:4-6\index{47 2 Corinthians:ch.  10, v.   4-6, pg. ``f30!}; Romans 6:17\index{45 Romans:ch.   6, v.  17, pg.
``f30!}.} of strength\index{Sanctification:supply of strength from Holy Spirit ``f10!}ening them\index{Holy
Spirit:strengthens believers against temptation ``f10!} against temptations\index{Temptation:temptation ``f10!} and
corruptions\index{Sanctification:Holy Spirit strengthens against corruptions ``f10!};\footnote{\ Matthew 4:4\index{40
Matthew:ch.   4, v.   4, pg. ``f30!}, 7, 10\index{40 Matthew:ch.   4, v.   4, 7, 10, pg. ``f30!}; Ephesians
6:16\index{49 Ephesians:ch.   6, v.  16, pg. ``f30!}{}-17\index{49 Ephesians:ch.   6, v.  16-17, pg. ``f30!}; Psalm
19:11\index{19 Psalms:ch.  19, v.  11, pg. ``f30!}; 1 Corinthians 10:11\index{46 1 Corinthians:ch.  10, v.  11, pg.
``f30!}.} of building them\index{Holy Spirit:builds up the elect ``f10!} up in grace\index{Grace:in believers:by Holy
Spirit ``f10!},\footnote{\ Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}; 2 Timothy 3:15\index{55 2 Timothy:ch.
  3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  15-17, pg. ``f30!}.} and establishing their hearts in
holiness and\index{Believers:holy:being perfected ``f10!} comfort through\index{Faith:source of comfort ``f10!} faith
unto salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Romans 1:16\index{45 Romans:ch.   1, v. 
16, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}3-17\index{45 Romans:ch.  10, v.  13-17, pg.
``f30!}; Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}; Romans 16:2\index{45 Romans:ch.  16, v.   2, pg.
``f30!}5\index{45 Romans:ch.  16, v.  25, pg. ``f30!}; 1 Thessalonians 3:2, 10-11, 13\index{52 1 Thessalonians:ch.   3,
v.   2, 10-11, 13, pg. ``f30!}.}


\bigskip

Q. 156 \textit{Is the word of God to be read}\index{Scripture:to be:read and heard ``f10!}\textit{ by all?}

A. Although all are not to be permitted to read\index{Scripture:to be:read and heard ``f10!} the word publicly to the
congregation\index{Scripture:not all permitted to read to congregation ``f10!},\footnote{\ Deuteronomy 31:9,
11-13\index{05 Deuteronomy:ch.  31, v.   9, 11-13, pg. ``f30!}; Nehemiah 8:2-3\index{16 Nehemiah:ch.   8, v.   2-3, pg.
``f30!}; Nehemiah 9:3-5\index{16 Nehemiah:ch.   9, v.   3-5, pg. ``f30!}.} yet all sorts of people are bound to read it
apart by themselves,\footnote{\ Deuteronomy 17:19\index{05 Deuteronomy:ch.  17, v.  19, pg. ``f30!}; Revelation
1:3\index{66 Revelation:ch.   1, v.   3, pg. ``f30!}; John 5:39\index{43 John:ch.   5, v.  39, pg. ``f30!}; Isaiah
34:16\index{23 Isaiah:ch.  34, v.  16, pg. ``f30!}.} and with their families\index{family, natural
``f10!}:\footnote{\ Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg. ``f30!}{}-9\index{05 Deuteronomy:ch.   6,
v.   6-9, pg. ``f30!}; Genesis 18:17, 19\index{01 Genesis:ch.  18, v.  17, 19, pg. ``f30!}; Psalm 78:5\index{19
Psalms:ch.  78, v.   5, pg. ``f30!}{}-7\index{19 Psalms:ch.  78, v.   5-7, pg. ``f30!}.} to which end, the
holy\index{Scripture:holy ``f10!} scriptures are\index{Scripture:word of God:revelation ``f10!} to be
translated\index{Scripture:to be:translated into common (vulgar) languages ``f10!} out of the original into vulgar
languages.\footnote{\ 1 Corinthians 14:6, 9, 11-12, 15-16, 24, 27-28\index{46 1 Corinthians:ch.  14, v.   6, 9, 11-12,
15-16, 24, 27-28, pg. ``f30!}.}


\bigskip

Q. 157 \textit{How is the word of God to be read}\index{Scripture:to be:read and heard ``f10!}\textit{?}

A. The holy\index{Scripture:holy ``f10!} scriptures are\index{Scripture:word of God:revelation ``f10!} to be
read\index{Scripture:to be:read and heard ``f10!} with an high and reverent\index{Scripture:to be:treated with reverent
esteem ``f10!} esteem of them\index{Scripture:to be:esteemed ``f10!};\footnote{\ Psalm 19:10\index{19 Psalms:ch.  19,
v.  10, pg. ``f30!}; Nehemiah 8:3-10\index{16 Nehemiah:ch.   8, v.   3-10, pg. ``f30!}; Exodus 24:7\index{02 Exodus:ch.
 24, v.   7, pg. ``f30!}; 2 Chronicles 34:27\index{14 2 Chronicles:ch.  34, v.  27, pg. ``f30!}; Isaiah 66:2\index{23
Isaiah:ch.  66, v.   2, pg. ``f30!}.} with a firm persuasion\index{Scripture:authored by God ``f10!} that they are the
very word of God,\footnote{\ 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg. ``f30!}{}-21\index{61 2 Peter:ch.   1,
v.  19-21, pg. ``f30!}.} and that he only can enable\index{Scripture:understanding enabled by Holy Spirit ``f10!} us to
understand them;\footnote{\ Luke 24:45\index{42 Luke:ch.  24, v.  45, pg. ``f30!}; 2 Corinthians 3:13-16\index{47 2
Corinthians:ch.   3, v.  13-16, pg. ``f30!}.} with desire to know\index{God's law:desire to know and obey ``f10!},
believe, and obey the will\index{Scripture:to be:obeyed:will of God ``f10!} of God revealed in
them;\footnote{\ Deuteronomy 17:10, 20\index{05 Deuteronomy:ch.  17, v.  10, 20, pg. ``f30!}.} with
diligence,\footnote{\ Acts 17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}.} and attention
to\index{Scripture:diligence in application ``f10!} the matter and scope of them;\footnote{\ Acts 8:30, 34\index{44
Acts:ch.   8, v.  30, 34, pg. ``f30!}; Luke 10:26-28\index{42 Luke:ch.  10, v.  26-28, pg. ``f30!}.} with
meditation\index{meditation ``f10!},\footnote{\ Psalm 1:2\index{19 Psalms:ch.   1, v.   2, pg. ``f30!}; Psalm
119:97\index{19 Psalms:ch. 119, v.  97, pg. ``f30!}.} application,\footnote{\ 2 Chronicles 24:21\index{14 2
Chronicles:ch.  24, v.  21, pg. ``f30!}.} self-denial\index{Prayer:with self-denial ``f10!},\footnote{\ Proverbs
3:5\index{20 Proverbs:ch.   3, v.   5, pg. ``f30!}; Deuteronomy 33:3\index{05 Deuteronomy:ch.  33, v.   3, pg.
``f30!}.} and prayer.\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.   2, v.   1, pg. ``f30!}{}-6\index{20 Proverbs:ch. 
 2, v.   1-6, pg. ``f30!}; Psalm 119:18\index{19 Psalms:ch. 119, v.  18, pg. ``f30!}; Nehemiah 7:6, 8\index{16
Nehemiah:ch.   7, v.   6, 8, pg. ``f30!}.}


\bigskip

Q. 158 \textit{By whom is the word of God to be preached?}

A. The word of God is to be preached only by such as are sufficiently gifted\index{Preachers:to be:sufficiently gifted
``f10!},\footnote{\ 1 Timothy 3:2\index{54 1 Timothy:ch.   3, v.   2, pg. ``f30!}, 6\index{54 1 Timothy:ch.   3, v.  
2, 6, pg. ``f30!}; Ephesians 4:8\index{49 Ephesians:ch.   4, v.   8, pg. ``f30!}{}-11\index{49 Ephesians:ch.   4, v.  
8-11, pg. ``f30!}; Hosea 4:6\index{28 Hosea:ch.   4, v.   6, pg. ``f30!}; Malachi 2:7\index{39 Malachi:ch.   2, v.   7,
pg. ``f30!}; 2 Corinthians 3:6\index{47 2 Corinthians:ch.   3, v.   6, pg. ``f30!}.} and also duly approved
and\index{Preachers:only gifted, approved, and called ``f10!} called to that office\index{Preachers:called to office
``f10!}.\footnote{\ Jeremiah 14:15\index{24 Jeremiah:ch.  14, v.  15, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10,
v.   1, pg. ``f30!}5\index{45 Romans:ch.  10, v.  15, pg. ``f30!}; Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg.
``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  12,
v.  28-29, pg. ``f30!}; 1 Timothy 3:1\index{54 1 Timothy:ch.   3, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   3, v. 
10, pg. ``f30!}; 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.   1, pg. ``f30!}4\index{54 1 Timothy:ch.   4, v.  14, pg.
``f30!}; 1 Timothy 5:22\index{54 1 Timothy:ch.   5, v.  22, pg. ``f30!}.}


\bigskip

Q. 159 \textit{How is the word of God to be preached by those}\index{Preaching:sound doctrine ``f10!}\textit{ that are
called thereunto?}

A. They that are called to labor in the\index{Preachers:labor in ministry of the word ``f10!} ministry of the
word\index{Preaching:ministry of the Word ``f10!}, are to preach sound\index{Preaching:sound doctrine ``f10!}
doctrine,\footnote{\ Titus 2:1, 8\index{56 Titus:ch.   2, v.   1, 8, pg. ``f30!}.}
diligently,\index{Preaching:diligently ``f10!}\footnote{\ Acts 18:25\index{44 Acts:ch.  18, v.  25, pg. ``f30!}.} in
season and\index{Preaching:in all seasons ``f10!} out of season\index{Preaching:plainly ``f10!};\footnote{\ 2 Timothy
4:2\index{55 2 Timothy:ch.   4, v.   2, pg. ``f30!}.} plainly,\index{Preaching:plain, not enticing words
``f10!}\footnote{\ 1 Corinthians 14:19\index{46 1 Corinthians:ch.  14, v.  19, pg. ``f30!}.} not in the
enticing\index{Preaching:not enticing words of man's wisdom ``f10!} words of man's wisdom\index{wisdom, man's ``f10!},
but in demonstration\index{Preaching:demonstration of the Spirit ``f10!} of the Spirit, and of power\index{Preachers:to
demonstrate power ``f10!};\footnote{\ 1 Corinthians 2:4\index{46 1 Corinthians:ch.   2, v.   4, pg. ``f30!}.}
faithfully\index{Preaching:faithfully ``f10!},\footnote{\ Jeremiah 23:28\index{24 Jeremiah:ch.  23, v.  28, pg.
``f30!}; 1 Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.   1, pg. ``f30!}{}-2\index{46 1 Corinthians:ch.   4, v.  
1-2, pg. ``f30!}.} making known the whole counsel of God\index{Causation:God's own will ``f10!};\footnote{\ Acts
20:27\index{44 Acts:ch.  20, v.  27, pg. ``f30!}.} wisely,\footnote{\ Colossians 1:28\index{51 Colossians:ch.   1, v. 
28, pg. ``f30!}; 2 Timothy 2:15\index{55 2 Timothy:ch.   2, v.  15, pg. ``f30!}.} applying themselves to the
necessities and capacities\index{Preachers:wisely target capacities of hearers ``f10!} of the
hearers\index{Preaching:target capacities of hearers ``f10!};\footnote{\ 1 Corinthians 3:2\index{46 1 Corinthians:ch.  
3, v.   2, pg. ``f30!}; Hebrews 5:12-14\index{58 Hebrews:ch.   5, v.  12-14, pg. ``f30!}; Luke 12:42\index{42 Luke:ch. 
12, v.  42, pg. ``f30!}.} zealously\index{Preachers:to have zeal for souls of men ``f10!},\footnote{\ Acts
18:25\index{44 Acts:ch.  18, v.  25, pg. ``f30!}.} with fervent love to God\index{Love toward God:for God
``f10!}\footnote{\ 2 Corinthians 5:13-14\index{47 2 Corinthians:ch.   5, v.  13-14, pg. ``f30!}; Philippians
1:1\index{50 Philippians:ch.   1, v.   1, pg. ``f30!}5-17\index{50 Philippians:ch.   1, v.  15-17, pg. ``f30!}.} and
the souls of his people;\footnote{\ Colossians 4:12\index{51 Colossians:ch.   4, v.  12, pg. ``f30!}; 2 Corinthians
12:15\index{47 2 Corinthians:ch.  12, v.  15, pg. ``f30!}.} sincerely,\footnote{\ 2 Corinthians 2:17\index{47 2
Corinthians:ch.   2, v.  17, pg. ``f30!}; 2 Corinthians 4:2\index{47 2 Corinthians:ch.   4, v.   2, pg. ``f30!}.}
aiming at his glory\index{Preaching:for God's glory ``f10!},\footnote{\ 1 Thessalonians 2:4\index{52 1
Thessalonians:ch.   2, v.   4, pg. ``f30!}{}-6\index{52 1 Thessalonians:ch.   2, v.   4-6, pg. ``f30!}; John
7:18\index{43 John:ch.   7, v.  18, pg. ``f30!}.} and their conversion\index{Conversion:conversion
``f10!},\footnote{\ 1 Corinthians 9:19-22\index{46 1 Corinthians:ch.   9, v.  19-22, pg. ``f30!}.}
edification,\index{Holy Spirit:edification of believers ``f10!}\footnote{\ 2 Corinthians 12:19\index{47 2
Corinthians:ch.  12, v.  19, pg. ``f30!}; Ephesians 4:12\index{49 Ephesians:ch.   4, v.  12, pg. ``f30!}.} and
salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.  
1, pg. ``f30!}6\index{54 1 Timothy:ch.   4, v.  16, pg. ``f30!}; Acts 26:16-18\index{44 Acts:ch.  26, v.  16-18, pg.
``f30!}.}


\bigskip

Q. 160 \textit{What is required of those that hear the word preached}\index{Preaching:preparation for ``f10!}\textit{?}

A. It is required of those that hear the word preached\index{Preaching:preparation for ``f10!}, that they attend upon it
with diligence\index{Preaching:diligence in application ``f10!},\footnote{\ Proverbs 8:34\index{20 Proverbs:ch.   8, v.
 34, pg. ``f30!}.} preparation,\footnote{\ 1 Peter 2:1-2\index{60 1 Peter:ch.   2, v.   1-2, pg. ``f30!}; Luke
8:18\index{42 Luke:ch.   8, v.  18, pg. ``f30!}.} and prayer;\footnote{\ Psalm 119:18\index{19 Psalms:ch. 119, v.  18,
pg. ``f30!}; Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}{}-19\index{49 Ephesians:ch.   6, v.  18-19,
pg. ``f30!}.} examine: What\index{Scripture:to be:used to examine all doctrines ``f10!} they hear by the
scriptures\index{Scripture:to be:read and heard ``f10!};\footnote{\ Acts 17:11\index{44 Acts:ch.  17, v.  11, pg.
``f30!}.} receive the truth\index{Scripture:truth to be received with faith ``f10!} with faith,\footnote{\ Hebrews
4:2\index{58 Hebrews:ch.   4, v.   2, pg. ``f30!}.} love,\footnote{\ 2 Thessalonians 2:10\index{53 2 Thessalonians:ch. 
 2, v.  10, pg. ``f30!}.} meekness\index{Love for others:meekness ``f10!},\footnote{\ James 1:21\index{59 James:ch.  
1, v.  21, pg. ``f30!}.} and readiness of mind,\index{Preaching:hearing:readiness of mind ``f10!}\footnote{\ Acts
17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}.} as the word of God;\footnote{\ 1 Thessalonians 2:13\index{52 1
Thessalonians:ch.   2, v.  13, pg. ``f30!}.} meditate\index{meditation ``f10!},\footnote{\ Luke 9:44\index{42 Luke:ch. 
 9, v.  44, pg. ``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}.} and confer of it;\footnote{\ Luke
24:14\index{42 Luke:ch.  24, v.  14, pg. ``f30!}; Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg.
``f30!}{}-7\index{05 Deuteronomy:ch.   6, v.   6-7, pg. ``f30!}.} hide\index{Preaching:hearing:hiding in heart ``f10!}
it in their hearts\index{Heart:word preached hidden in ``f10!},\footnote{\ Proverbs 2:1\index{20 Proverbs:ch.   2, v.  
1, pg. ``f30!}; Psalm 119:11\index{19 Psalms:ch. 119, v.  11, pg. ``f30!}.} and bring forth the fruit of it in their
lives.\footnote{\ Luke 8:15\index{42 Luke:ch.   8, v.  15, pg. ``f30!}; James 1:25\index{59 James:ch.   1, v.  25, pg.
``f30!}.}

Q. 161 \textit{How do the sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ become
effectual}\index{Sacraments:effectual means of salvation ``f10!}\textit{ means of salvation}\index{Salvation:belief in
Jesus by faith ``f10!}\textit{?}

A. The sacraments\index{Sacraments:holy:ordinances ``f10!} become effectual\index{Sacraments:effectual means of
salvation ``f10!} means of salvation\index{Salvation:belief in Jesus by faith ``f10!}, not by any
power\index{Sacraments:no power in themselves ``f10!} in themselves, or any virtue\index{Sacraments:grace:not conferred
by elements ``f10!} derived from the piety or\index{Lord's Supper:efficacy, not dependent on piety or intention of
person administering ``f10!} intention of him by whom they are administered, but only\index{Sacraments:no virtue from
person administering ``f10!} by the working of the\index{Sacraments:Holy Spirit working ``f10!} Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!}, and the blessing of Christ, by whom they are instituted\index{Sacraments:instituted
by:Christ ``f10!}.\footnote{\ 1 Peter 3:21\index{60 1 Peter:ch.   3, v.  21, pg. ``f30!}; Acts 8:13, 23\index{44
Acts:ch.   8, v.  13, 23, pg. ``f30!}; 1 Corinthians 3:6-7\index{46 1 Corinthians:ch.   3, v.   6-7, pg. ``f30!}; 1
Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.}


\bigskip

Q. 162 \textit{What is a sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{?}

A. A sacrament\index{Sacraments:holy:ordinances ``f10!} is a holy ordinance instituted by
Christ\index{Sacraments:instituted by:Christ ``f10!} in his church\index{Church:Christ's ``f10!},\footnote{\ Genesis
17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 10\index{01 Genesis:ch.  17, v.   7, 10, pg. ``f30!}; Exodus 12;
Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg. ``f30!};
Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.} to signify\index{Sacraments:holy:signs ``f10!}, seal,
and\index{Sacraments:seals of covenant of grace ``f10!} exhibit\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11,
pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.  11, v.
 24, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  24-25, pg. ``f30!}.} unto those that are within the covenant
of grace\index{Covenant:grace:grace ``f10!},\footnote{\ Romans 15:8\index{45 Romans:ch.  15, v.   8, pg. ``f30!};
Exodus 12:48\index{02 Exodus:ch.  12, v.  48, pg. ``f30!}.} the benefits of his mediation\index{Jesus:mediator:of
covenant ``f10!};\footnote{\ Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}; 1 Corinthians 10:16\index{46 1
Corinthians:ch.  10, v.  16, pg. ``f30!}.} to strengthen and\index{Believers:benefits:of Christ's mediation ``f10!}
increase their faith\index{Faith:increase of ``f10!}, and all other graces\index{Grace:exhibited in sacraments
``f10!};\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Galatians 3:27\index{48 Galatians:ch.  
3, v.  27, pg. ``f30!}.} to oblige them to obedience\index{God's law:commandments:obedience:required
``f10!};\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6, v.   3-4, pg. ``f30!}; 1 Corinthians 10:21\index{46 1
Corinthians:ch.  10, v.  21, pg. ``f30!}.} to testify\index{Sacraments:testimony through participation ``f10!} and
cherish their love\index{Love for others:for saints ``f10!} and communion one\index{Communion:of saints ``f10!} with
another;\footnote{\ Ephesians 4:2-5\index{49 Ephesians:ch.   4, v.   2-5, pg. ``f30!}; 1 Corinthians 12:13\index{46 1
Corinthians:ch.  12, v.  13, pg. ``f30!}.} and to distinguish\index{Sacraments:distinguish from those who are outside
``f10!} them from those that are without.\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg.
``f30!}1-12\index{49 Ephesians:ch.   2, v.  11-12, pg. ``f30!}; Genesis 34:14\index{01 Genesis:ch.  34, v.  14, pg.
``f30!}.}


\bigskip

Q. 163 \textit{What are the parts of a sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{?}

A. The parts of a sacrament\index{Sacraments:holy:ordinances ``f10!} are two; the one an outward and sensible sign,
used\index{Sacraments:sensible signs ``f10!} according to Christ's own appointment; the\index{Lord's Supper:proper
administration ``f10!} other an inward\index{Sacraments:outward sign, inward grace ``f10!} and spiritual
grace\index{Grace:exhibited in sacraments ``f10!} thereby signified.\footnote{\ Matthew 3:11\index{40 Matthew:ch.   3,
v.  11, pg. ``f30!}; 1 Peter 3:21\index{60 1 Peter:ch.   3, v.  21, pg. ``f30!}; Romans 2:28-29\index{45 Romans:ch.  
2, v.  28-29, pg. ``f30!}.}


\bigskip

Q. 164 \textit{How many sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ has Christ instituted
in}\index{Sacraments:instituted by:Christ ``f10!}\textit{ his church}\index{Church:Christ's ``f10!}\textit{ under the
New Testament}\index{Scripture:New Testament ``f10!}\textit{?}

A. Under the New Testament\index{Scripture:New Testament ``f10!} Christ has instituted in\index{Sacraments:instituted
by:Christ ``f10!} his church\index{Church:Christ's ``f10!} only two sacraments\index{Sacraments:holy:ordinances
``f10!}, Baptism and\index{Baptism:sacrament ``f10!} the Lord's supper\index{Lord's Supper:sacrament
``f10!}.\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v. 
26-28, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}, 23\index{46 1
Corinthians:ch.  11, v.  20, 23, pg. ``f30!}.}


\bigskip

Q. 165 \textit{What is Baptism}\index{Baptism:definition ``f10!}\textit{?}

A. Baptism is a\index{Baptism:sacrament ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} of the New
Testament\index{Scripture:New Testament ``f10!}, wherein Christ has ordained the\index{Baptism:ordained by Jesus
``f10!} washing\index{Baptism:washing with water ``f10!} with water in\index{Baptism:to use water ``f10!} the name of
the Father\index{Father:God the ``f10!}, and of the Son\index{Jesus:Son of God ``f10!}, and of the Holy
Ghost\index{Holy Spirit:Holy Ghost ``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.} to
be a sign\index{Baptism:sign of salvation ``f10!} and seal of\index{Baptism:seal of salvationof ``f10!}
ingrafting\index{Believers:ingrafted into Christ ``f10!} into himself,\footnote{\ Galatians 3:27\index{48 Galatians:ch.
  3, v.  27, pg. ``f30!}.} of remission\index{Redemption:remission of sins ``f10!} of sins by his
blood\index{Jesus:shed blood for remission of sins ``f10!},\footnote{\ Mark 1:4\index{41 Mark:ch.   1, v.   4, pg.
``f30!}; Revelation 1:5\index{66 Revelation:ch.   1, v.   5, pg. ``f30!}.} and
regeneration\index{Regeneration:regeneration ``f10!} by his Spirit;\footnote{\ Titus 3:5\index{56 Titus:ch.   3, v.  
5, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}6\index{49 Ephesians:ch.   5, v.  26, pg.
``f30!}.} of adoption\index{Adoption:in salvation ``f10!},\footnote{\ Galatians 3:26-27\index{48 Galatians:ch.   3, v. 
26-27, pg. ``f30!}.} and resurrection unto\index{Resurrection:believers:to everlasting life ``f10!} everlasting
life\index{everlasting life ``f10!};\footnote{\ 1 Corinthians 15:29\index{46 1 Corinthians:ch.  15, v.  29, pg.
``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}.} and whereby the parties baptized are solemnly
admitted into\index{Baptism:admission into visible Church ``f10!} the visible church\index{Church:visible
``f10!},\footnote{\ 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.} and enter into an open
and professed engagement to be\index{Oaths and Vows:engagements to God ``f10!} wholly and only the
Lord's.\footnote{\ Romans 6:4\index{45 Romans:ch.   6, v.   4, pg. ``f30!}.}


\bigskip

Q. 166 \textit{Unto whom is}\index{Baptism:recipients ``f10!}\textit{ Baptism to be administered?}

A. Baptism is not to be administered to any\index{Baptism:not to be administered outside those in visible church ``f10!}
that are out of the visible church\index{Church:visible ``f10!}, and so strangers\index{Baptism:not for strangers from
covenant of promise ``f10!} from the covenant of promise\index{Covenant:promises ``f10!}, till they profess their
faith\index{Faith:profession of ``f10!} in Christ, and obedience\index{God's law:commandments:obedience:required
``f10!} to him,\footnote{\ Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}; Acts 8:36-37\index{44 Acts:ch.   8, v.
 36-37, pg. ``f30!}.} but infants\index{Infants:of believing parents, to be baptized ``f10!} descending from
parents\index{Baptism:infants of believing parents ``f10!}, either both, or but one of them, professing
faith\index{Faith:professed in baptism ``f10!} in Christ, and obedience to him, are in that respect within the
covenant\index{Covenant:new:ways dispensed ``f10!}, and to be baptized\index{Baptism:required as obedience
``f10!}.\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 9\index{01 Genesis:ch.  17, v.   7, 9,
pg. ``f30!}; Galatians 3:9, 14\index{48 Galatians:ch.   3, v.   9, 14, pg. ``f30!}; Colossians 2:11-12\index{51
Colossians:ch.   2, v.  11-12, pg. ``f30!}; Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.
  2, v.  38-39, pg. ``f30!}; Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}{}-12\index{45 Romans:ch.   4, v. 
11-12, pg. ``f30!}; Romans 11:16\index{45 Romans:ch.  11, v.  16, pg. ``f30!}; 1 Corinthians 7:14\index{46 1
Corinthians:ch.   7, v.  14, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; Luke
18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}5\index{42 Luke:ch.  18, v.  15, pg. ``f30!}{}-16\index{42 Luke:ch.  18,
v.  15-16, pg. ``f30!}.}


\bigskip

Q. 167 \textit{How is our Baptism to be improved by us?}

A. The needful but much neglected\index{Baptism:improvement of ``f10!} duty of improving\index{Baptism:duty of improving
``f10!} our Baptism, is to be performed by us all our life long, especially in the time of
temptation\index{Temptation:temptation ``f10!}, and when we are present at the administration\index{Baptism:efficacy
not time dependent ``f10!} of it to others;\footnote{\ Colossians 2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg.
``f30!}; Romans 6:4\index{45 Romans:ch.   6, v.   4, pg. ``f30!}, 6, 11\index{45 Romans:ch.   6, v.   4, 6, 11, pg.
``f30!}.}[1] by serious\index{Sacraments:improving by serious reflection ``f10!} and thankful\index{Prayer:thanking God
``f10!} consideration of the nature of it, and of the ends for which Christ instituted it\index{Baptism:instituted by
Christ ``f10!}, the privileges\index{Believers:privileges ``f10!} and benefits conferred\index{Baptism:benefits
conferred ``f10!} and sealed thereby\index{Baptism:benefits sealed ``f10!}, and our solemn vow
made\index{Baptism:solemn vow made ``f10!} therein;\footnote{\ Romans 6:3-5\index{45 Romans:ch.   6, v.   3-5, pg.
``f30!}.} by being humbled\index{Humility:before God ``f10!} for our sinful defilement\index{Sin:defilement ``f10!},
our falling\index{God's law:broken daily by believers ``f10!} short of, and walking contrary\index{Baptism:not to
walking contrary ``f10!} to, the grace\index{Grace:exhibited in sacraments ``f10!} of baptism, and\index{Baptism:means
of grace ``f10!} our engagements\index{Oaths and Vows:engagements to God ``f10!};\footnote{\ 1 Corinthians
1:11-13\index{46 1 Corinthians:ch.   1, v.  11-13, pg. ``f30!}; Romans 6:2-3\index{45 Romans:ch.   6, v.   2-3, pg.
``f30!}.} by growing up\index{Assurance:growing ``f10!} to assurance\index{Assurance:of pardon of sin ``f10!} of pardon
of sin, and of all other blessings\index{Baptism:blessings sealed ``f10!} sealed to us in that
sacrament\index{Sacraments:holy:ordinances ``f10!};\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg.
``f30!}{}-12\index{45 Romans:ch.   4, v.  11-12, pg. ``f30!}; 1 Peter 3:21\index{60 1 Peter:ch.   3, v.  21, pg.
``f30!}.} by drawing strength\index{Baptism:drawing strength from ``f10!} from the death\index{Jesus:resurrection:from
the dead ``f10!} and resurrection of Christ, into whom we are baptized\index{Baptism:into Christ ``f10!}, for the
mortifying\index{Sin:mortified ``f10!} of sin, and quickening of grace;\index{Grace:in believers:by Holy Spirit
``f10!}\footnote{\ Romans 6:3-5\index{45 Romans:ch.   6, v.   3-5, pg. ``f30!}.} and by endeavoring to live by
faith\index{Faith:living by ``f10!},\footnote{\ Galatians 3:26-27\index{48 Galatians:ch.   3, v.  26-27, pg. ``f30!}.}
to have our conversation\index{Believers:to walk faithfully in obedience (conversation) ``f10!} in holiness
and\index{Believers:holy:being perfected ``f10!} righteousness,\footnote{\ Romans 6:22\index{45 Romans:ch.   6, v.  22,
pg. ``f30!}.} as those that have therein given up their names to Christ;\footnote{\ Acts 2:38\index{44 Acts:ch.   2, v.
 38, pg. ``f30!}.} and to walk in brotherly\index{Love for others:for saints ``f10!} love, as being
baptized\index{Baptism:by Spirit ``f10!} by the same Spirit\index{Holy Spirit:believers baptized by, into one body
``f10!} into one body.\footnote{\ 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!},
25-27\index{46 1 Corinthians:ch.  12, v.  13, 25-27, pg. ``f30!}.}


\bigskip


\bigskip


Q. 168 \textit{What is the Lord's supper}\index{Lord's Supper:sacrament ``f10!}\textit{?}

A. The Lord's supper\index{Lord's Supper:sacrament ``f10!} is a sacrament\index{Sacraments:holy:ordinances ``f10!} of
the New Testament\index{Scripture:New Testament ``f10!},\footnote{\ Luke 22:20\index{42 Luke:ch.  22, v.  20, pg.
``f30!}.} wherein, by giving and receiving\index{Lord's Supper:defined ``f10!} bread and\index{Lord's Supper:bread
``f10!} wine\index{Lord's Supper:wine ``f10!} according to the appointment of\index{Lord's Supper:proper administration
``f10!} Jesus Christ, his death\index{Jesus:death:death ``f10!} is showed forth; and they that worthily\index{Lord's
Supper:spiritual feeding:worthy receivers ``f10!} communicate feed\index{Lord's Supper:spiritual feeding:in faith
``f10!} upon his body and blood\index{Lord's Supper:body and blood of Jesus ``f10!}, to their spiritual
nourishment\index{Lord's Supper:spiritual feeding:nourishment ``f10!} and growth in
grace\index{Sanctification:spiritual growth:in grace ``f10!};\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v. 
26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg. ``f30!}; 1 Corinthians 11:13-26\index{46 1
Corinthians:ch.  11, v.  13-26, pg. ``f30!}.} have their union and\index{Believers:union with Christ:mystical ``f10!}
communion with him\index{Communion:with Christ ``f10!} confirmed\index{Sacraments:confirm union with Christ
``f10!};\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}.}
testify\index{Sacraments:testimony through participation ``f10!} and renew their thankfulness\index{Prayer:thanking God
``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch. 
11, v.  24, pg. ``f30!}.} and engagement to God\index{Oaths and Vows:engagements to God ``f10!},\footnote{\ 1
Corinthians 10:14-16, 21\index{46 1 Corinthians:ch.  10, v.  14-16, 21, pg. ``f30!}.} and their mutual love\index{Love
for others:mutual:love ``f10!} and fellowship each\index{Fellowship:with fellow believers ``f10!} with other, as
members of the\index{Church members ``f10!} same mystical body.\footnote{\ 1 Corinthians 10:17\index{46 1
Corinthians:ch.  10, v.  17, pg. ``f30!}.}


\bigskip

Q. 169 \textit{How has Christ appointed bread and}\index{Lord's Supper:bread ``f10!}\textit{ wine}\index{Lord's
Supper:wine ``f10!}\textit{ to be given and received in the sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{
of the Lord's supper}\index{Lord's Supper:sacrament ``f10!}\textit{?}

A. Christ has appointed the ministers\index{Church officers:minsters ``f10!} of his word, in the administration of
this\index{Lord's Supper:ministers to administer ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} of the
Lord's Supper\index{Lord's Supper:sacrament ``f10!}, to set apart the bread and\index{Lord's Supper:bread ``f10!}
wine\index{Lord's Supper:wine ``f10!} from common use, by the word of institution\index{Sacraments:instituted by:Christ
``f10!}, thanksgiving\index{Prayer:thanking God ``f10!}, and prayer; to\index{Prayer:Jesus institution of Lord's Supper
``f10!} take and break the\index{Lord's Supper:bread broken ``f10!} bread, and to give both the bread and the wine to
the communicants\index{Lord's Supper:both bread and wine to communicants ``f10!}: who are, by the same appointment, to
take\index{Lord's Supper:proper administration ``f10!} and eat the bread, and to drink the wine, in thankful
remembrance that\index{Lord's Supper:remembrance of Christ's sacrifice ``f10!} the body of Christ was broken and given,
and his blood\index{Jesus:shed blood for remission of sins ``f10!} shed, for them.\footnote{\ 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg.
``f30!}{}-24\index{46 1 Corinthians:ch.  11, v.  23-24, pg. ``f30!}; Matthew 26:26\index{40 Matthew:ch.  26, v.  26,
pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg. ``f30!}; Mark 14:22-24\index{41 Mark:ch.  14, v.  22-24, pg.
``f30!}; Luke 22:19\index{42 Luke:ch.  22, v.  19, pg. ``f30!}{}-20\index{42 Luke:ch.  22, v.  19-20, pg. ``f30!}.}


\bigskip

Q. 170 \textit{How do they that worthily}\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!}\textit{
communicate in}\index{Lord's Supper:communion ``f10!}\textit{ the Lord's supper}\index{Lord's Supper:sacrament
``f10!}\textit{ feed}\index{Lord's Supper:spiritual feeding:in faith ``f10!}\textit{ upon the body and
blood}\index{Lord's Supper:body and blood of Jesus ``f10!}\textit{ of Christ therein?}

A. As the body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of Christ are not corporally\index{Lord's
Supper:spiritual feeding:not corporal ``f10!} or carnally\index{Lord's Supper:spiritual feeding:not carnal ``f10!}
present in, with, or under the bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!} in the
Lord's supper\index{Lord's Supper:sacrament ``f10!},\footnote{\ Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!}.}
and yet are spiritually\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!} present to the
faith\index{Faith:profession of ``f10!} of the receiver, no less truly and really than the elements\index{Lord's
Supper:elements:bread and wine ``f10!} themselves are to their outward\index{Lord's Supper:elements:outward senses
``f10!} senses;\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}, 28\index{40 Matthew:ch.  26, v.
 26, 28, pg. ``f30!}.} so they that worthily communicate in\index{Lord's Supper:communion ``f10!} the
sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's supper, do therein feed\index{Lord's Supper:spiritual
feeding:in faith ``f10!} upon the body and blood of Christ, not after a corporal and carnal, but in a spiritual manner;
yet truly and really,\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1
Corinthians:ch.  11, v.  24, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  24-29, pg. ``f30!}.} while by
faith\index{Faith:in Jesus Christ ``f10!} they receive and apply unto themselves Christ crucified\index{Jesus:crucified
``f10!}, and all the benefits of his death\index{Jesus:death:death ``f10!}.\footnote{\ 1 Corinthians 10:16\index{46 1
Corinthians:ch.  10, v.  16, pg. ``f30!}.}


\bigskip

Q. 171 \textit{How are they that receive the sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{ of the Lord's
supper}\index{Lord's Supper:sacrament ``f10!}\textit{ to prepare themselves}\index{Lord's Supper:preparation:actions
``f10!}\textit{ before they come unto it?}

A. They that receive the sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's supper\index{Lord's
Supper:sacrament ``f10!} are, before they come, to prepare themselves\index{Lord's Supper:preparation:actions ``f10!}
thereunto, by examining\index{Lord's Supper:preparation:self-examination ``f10!} themselves\footnote{\ 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}8\index{46 1 Corinthians:ch.  11, v.  28, pg. ``f30!}.} of
their being in Christ,\footnote{\ 2 Corinthians 13:5\index{47 2 Corinthians:ch.  13, v.   5, pg. ``f30!}.} of their
sins and wants;\footnote{\ 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7, pg. ``f30!}; Exodus 12:15\index{02
Exodus:ch.  12, v.  15, pg. ``f30!}.} of the truth and\index{Lord's Supper:evidence of true faith required ``f10!}
measure of their knowledge,\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}9\index{46
1 Corinthians:ch.  11, v.  29, pg. ``f30!}.} faith,\footnote{\ 1 Corinthians 13:5\index{46 1 Corinthians:ch.  13, v.  
5, pg. ``f30!}; Matthew 26:28\index{40 Matthew:ch.  26, v.  28, pg. ``f30!}.} repentance;\index{Repentance:for all sins
``f10!}\footnote{\ Zechariah 12:10\index{38 Zechariah:ch.  12, v.  10, pg. ``f30!}; 1 Corinthians 11:31\index{46 1
Corinthians:ch.  11, v.  31, pg. ``f30!}.} love to God\index{Love toward God:for God ``f10!} and the
brethren\index{Believers:to build up fellow believers ``f10!},\footnote{\ 1 Corinthians 10:16\index{46 1
Corinthians:ch.  10, v.  16, pg. ``f30!}{}-17\index{46 1 Corinthians:ch.  10, v.  16-17, pg. ``f30!}; Acts
2:46-47\index{44 Acts:ch.   2, v.  46-47, pg. ``f30!}.} charity\index{Good works:charity ``f10!} to all
men,\footnote{\ 1 Corinthians 5:8\index{46 1 Corinthians:ch.   5, v.   8, pg. ``f30!}; 1 Corinthians 11:18, 20\index{46
1 Corinthians:ch.  11, v.  18, 20, pg. ``f30!}.} forgiving those\index{Forgiveness:forgiving ``f10!} that have done
them wrong;\footnote{\ Matthew 5:23-24\index{40 Matthew:ch.   5, v.  23-24, pg. ``f30!}.} of their desires
after\index{Lord's Supper:desire to be in Christ ``f10!} Christ,\footnote{\ Isaiah 55:1\index{23 Isaiah:ch.  55, v.  
1, pg. ``f30!}; John 7:37\index{43 John:ch.   7, v.  37, pg. ``f30!}.} and of their new
obedience\index{Believers:obedience:strive for new ``f10!};\footnote{\ 1 Corinthians 5:7\index{46 1 Corinthians:ch.  
5, v.   7, pg. ``f30!}{}-8\index{46 1 Corinthians:ch.   5, v.   7-8, pg. ``f30!}.} and by renewing
the\index{Sanctification:renewed in image of God ``f10!} exercise of these graces\index{Grace:exhibited in sacraments
``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch. 
11, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  25-26, pg. ``f30!}, 28\index{46 1 Corinthians:ch.  11,
v.  25-26, 28, pg. ``f30!}; Hebrews 10:21-22, 24\index{58 Hebrews:ch.  10, v.  21-22, 24, pg. ``f30!}; Psalm
26:6\index{19 Psalms:ch.  26, v.   6, pg. ``f30!}.} by serious\index{Sacraments:improving by serious reflection ``f10!}
meditation\index{meditation ``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  24-25, pg.
``f30!}.} and fervent\index{Prayer:with fervency ``f10!} prayer.\footnote{\ 2 Chronicles 30:18-19\index{14 2
Chronicles:ch.  30, v.  18-19, pg. ``f30!}; Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}.}


\bigskip

Q. 172 \textit{May one who doubts}\index{Assurance:doubts ``f10!}\textit{ of his being in Christ, or of his due
preparation, come}\index{Lord's Supper:preparation:actions ``f10!}\textit{ to the Lord's supper}\index{Lord's
Supper:sacrament ``f10!}\textit{?}

A. One who doubts\index{Assurance:doubts ``f10!} of his being in Christ, or of his due preparation to the\index{Lord's
Supper:preparation:actions ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's supper\index{Lord's
Supper:sacrament ``f10!}, may have true interest in Christ\index{Believers:interest (share) in Christ ``f10!}, though
he be not yet assured thereof\index{Assurance:believer may await ``f10!};\footnote{\ Isaiah 1:10\index{23 Isaiah:ch.  
1, v.  10, pg. ``f30!}; 1 John 5:13\index{62 1 John:ch.   5, v.  13, pg. ``f30!}; Psalm 77:1-12\index{19 Psalms:ch. 
77, v.   1-12, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}; Jonah 2:4, 7\index{32
Jonah:ch.   2, v.   4, 7, pg. ``f30!}.} and in God's account has it, if he be duly affected with the
apprehension\index{Lord's Supper:apprehension of interest in Christ ``f10!} of the want of it,\footnote{\ Isaiah
54:7-10\index{23 Isaiah:ch.  54, v.   7-10, pg. ``f30!}; Matthew 5:3-4\index{40 Matthew:ch.   5, v.   3-4, pg. ``f30!};
Psalm 31:22\index{19 Psalms:ch.  31, v.  22, pg. ``f30!}; Psalm 73:13, 22-23\index{19 Psalms:ch.  73, v.  13, 22-23,
pg. ``f30!}.} and unfeignedly desires to\index{Lord's Supper:desire to be in Christ ``f10!} be found in
Christ,\footnote{\ Philippians 3:8-9\index{50 Philippians:ch.   3, v.   8-9, pg. ``f30!}; Psalm 10:17\index{19
Psalms:ch.  10, v.  17, pg. ``f30!}; Psalm 42:1-2, 5, 11\index{19 Psalms:ch.  42, v.   1-2, 5, 11, pg. ``f30!}.} and to
depart\index{Lord's Supper:preparation:depart from sin ``f10!} from iniquity:\footnote{\ 2 Timothy 2:19\index{55 2
Timothy:ch.   2, v.  19, pg. ``f30!}; Isaiah 1:10\index{23 Isaiah:ch.   1, v.  10, pg. ``f30!}; Psalm 66:18-20\index{19
Psalms:ch.  66, v.  18-20, pg. ``f30!}.} in which case (because promises\index{Lord's Supper:promises of ``f10!} are
made, and this sacrament is appointed, for the relief\index{Lord's Supper:relief for weak and doubting faith ``f10!}
even of weak and doubting\index{doubting Christians ``f10!} Christians)\footnote{\ Isaiah 40:11, 29, 31\index{23
Isaiah:ch.  40, v.  11, 29, 31, pg. ``f30!}; Matthew 11:28\index{40 Matthew:ch.  11, v.  28, pg. ``f30!}; Matthew
12:20\index{40 Matthew:ch.  12, v.  20, pg. ``f30!}; Matthew 26:28\index{40 Matthew:ch.  26, v.  28, pg. ``f30!}.} he
is to bewail\index{Lord's Supper:preparation:bewail sin ``f10!} his unbelief\index{unbelief (misbelief)
``f10!},\footnote{\ Mark 9:24\index{41 Mark:ch.   9, v.  24, pg. ``f30!}.} and labor to have his doubts
resolved;\index{Lord's Supper:preparation:resolve doubts ``f10!}\footnote{\ Acts 2:37\index{44 Acts:ch.   2, v.  37,
pg. ``f30!}; Acts 16:30\index{44 Acts:ch.  16, v.  30, pg. ``f30!}.} and, so doing, he may and ought to come to the
Lord's supper, that he may be further strengthened\index{Sacraments:strengthening ordinances ``f10!}.\footnote{\ Romans
4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}8\index{46 1 Corinthians:ch.  11, v.  28, pg. ``f30!}.}


\bigskip

Q. 173 \textit{May any who profess the faith}\index{Faith:profession of ``f10!}\textit{, and desire to
come}\index{Lord's Supper:desire to attend ``f10!}\textit{ to the Lord's supper}\index{Lord's Supper:sacrament
``f10!}\textit{, be kept from it?}

A. Such as are found to be ignorant\index{ignorance ``f10!} or scandalous\index{Scandalous behaviour:behaviour ``f10!},
notwithstanding their profession\index{Faith:profession of ``f10!} of the faith, and desire to come\index{Lord's
Supper:desire to attend ``f10!} to the Lord's supper\index{Lord's Supper:sacrament ``f10!}, may and ought to be kept
from that sacrament\index{Sacraments:holy:ordinances ``f10!}, by the power which\index{Church:power delegated by Christ
``f10!} Christ has left in his church\index{Church:Christ's ``f10!},\footnote{\ 1 Corinthians 5; 1 Corinthians
11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v.  27, pg.
``f30!}{}-31\index{46 1 Corinthians:ch.  11, v.  27-31, pg. ``f30!}; Matthew 7:6\index{40 Matthew:ch.   7, v.   6, pg.
``f30!}; Jude 1:23\index{65 Jude:v.   1, v.  23, pg. ``f30!}; 1 Timothy 5:22\index{54 1 Timothy:ch.   5, v.  22, pg.
``f30!}.} until they receive instruction\index{Lord's Supper:unworthy participation ``f10!}, and manifest
their\index{Lord's Supper:ungodly not to be admitted ``f10!} reformation\index{Church discipline:suspension from Lord's
Supper ``f10!}.\footnote{\ 2 Corinthians 2:7\index{47 2 Corinthians:ch.   2, v.   7, pg. ``f30!}.}


\bigskip

Q. 174 \textit{What is required of them that receive the sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{ of
the Lord's supper}\index{Lord's Supper:sacrament ``f10!}\textit{ in the time of the administration}\index{Lord's
Supper:proper administration ``f10!}\textit{ of it?}

A. It is required of them that receive the sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's
supper\index{Lord's Supper:sacrament ``f10!}, that, during the time of the administration\index{Lord's Supper:proper
administration ``f10!} of it, with all holy\index{Lord's Supper:elements:holy use ``f10!} reverence and attention they
wait upon\index{Lord's Supper:waiting on God ``f10!} God in that ordinance,\footnote{\ Leviticus 10:3\index{03
Leviticus:ch.  10, v.   3, pg. ``f30!}; Hebrews 12:1\index{58 Hebrews:ch.  12, v.   1, pg. ``f30!}8\index{58
Hebrews:ch.  12, v.  18, pg. ``f30!}; Psalm 5:7\index{19 Psalms:ch.   5, v.   7, pg. ``f30!}; 1 Corinthians 11:17,
26-27\index{46 1 Corinthians:ch.  11, v.  17, 26-27, pg. ``f30!}.} diligently observe the\index{Lord's
Supper:elements:diligently observe ``f10!} sacramental\index{Lord's Supper:elements:union between sign and signified
``f10!} elements\index{Lord's Supper:elements:bread and wine ``f10!} and actions,\footnote{\ Exodus 24:8\index{02
Exodus:ch.  24, v.   8, pg. ``f30!}; Matthew 26:28\index{40 Matthew:ch.  26, v.  28, pg. ``f30!}.} heedfully discern
the\index{Lord's Supper:discern Lord's body ``f10!} Lord's body,\footnote{\ 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}9\index{46 1 Corinthians:ch.  11, v.  29, pg. ``f30!}.} and affectionately
meditate\index{meditation ``f10!} on his death\index{Jesus:death:death ``f10!} and sufferings,\footnote{\ Luke
22:19\index{42 Luke:ch.  22, v.  19, pg. ``f30!}.} and thereby stir\index{Grace:in believers:to be stirred up ``f10!}
up themselves to a vigorous exercise of their graces;\index{Grace:exhibited in sacraments ``f10!}\footnote{\ 1
Corinthians 10:3-5, 11, 14\index{46 1 Corinthians:ch.  10, v.   3-5, 11, 14, pg. ``f30!}; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}6\index{46 1 Corinthians:ch.  11, v.  26, pg. ``f30!}.} in judging
themselves\index{Judgement:personal ``f10!},\footnote{\ 1 Corinthians 11:31\index{46 1 Corinthians:ch.  11, v.  31, pg.
``f30!}.} and sorrowing for sin\index{Confession:sorrow for sin ``f10!};\footnote{\ Zechariah 12:10\index{38
Zechariah:ch.  12, v.  10, pg. ``f30!}.} in earnest hungering\index{Believers:to hunger after Christ in Lord's Supper
``f10!} and thirsting\index{Lord's Supper:desire to be in Christ ``f10!} after Christ,\footnote{\ Revelation
22:17\index{66 Revelation:ch.  22, v.  17, pg. ``f30!}.} feeding\index{Lord's Supper:spiritual feeding:in faith ``f10!}
on him by faith,\footnote{\ John 6:35\index{43 John:ch.   6, v.  35, pg. ``f30!}.} receiving of
his\index{Faith:instrument:for receiving Christ ``f10!} fulness,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}6\index{43 John:ch.   1, v.  16, pg. ``f30!}.} trusting in his\index{Lord's Supper:trusting in Christ's merits
``f10!} merits\index{Faith:trusts in the merits of Christ ``f10!},\footnote{\ Philippians 1:1\index{50 Philippians:ch. 
 1, v.   1, pg. ``f30!}6\index{50 Philippians:ch.   1, v.  16, pg. ``f30!}.} rejoicing in his\index{Lord's
Supper:rejoice in Christ ``f10!} love,\footnote{\ Psalm 63:4-5\index{19 Psalms:ch.  63, v.   4-5, pg. ``f30!}; 2
Chronicles 30:21\index{14 2 Chronicles:ch.  30, v.  21, pg. ``f30!}.} giving thanks\index{Prayer:thanking God ``f10!}
for his grace\index{Grace:thanks for ``f10!};\footnote{\ Psalm 22:26\index{19 Psalms:ch.  22, v.  26, pg. ``f30!}.} in
renewing of their covenant with\index{Covenant:renewal with God ``f10!} God, and love to all\index{Love for others:for
saints ``f10!} the saints\index{Believers:saints ``f10!}.\footnote{\ Jeremiah 1:5\index{24 Jeremiah:ch.   1, v.   5,
pg. ``f30!}; Psalm 1:5\index{19 Psalms:ch.   1, v.   5, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg.
``f30!}.}


\bigskip

Q. 175 \textit{What is the duty of Christians, after}\index{Lord's Supper:after participation:duty ``f10!}\textit{ they
have received the sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{ of the Lord's supper}\index{Lord's
Supper:sacrament ``f10!}\textit{?}

A. The duty of Christians, after\index{Lord's Supper:after participation:duty ``f10!} they have received the
sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's supper\index{Lord's Supper:sacrament ``f10!}, is
seriously\index{Lord's Supper:after participation:seriously consider behaviour ``f10!} to consider how they have
behaved\index{Lord's Supper:post assessment of behaviour during ``f10!} themselves therein, and with what
success;\footnote{\ Psalm 28:7\index{19 Psalms:ch.  28, v.   7, pg. ``f30!}; Psalm 85:8\index{19 Psalms:ch.  85, v.  
8, pg. ``f30!}; 1 Corinthians 11:17, 30-31\index{46 1 Corinthians:ch.  11, v.  17, 30-31, pg. ``f30!}.} if they find
quickening and comfort, to\index{Lord's Supper:benefits believers receive ``f10!} bless God\index{Lord's Supper:bless
God for it ``f10!} for it,\footnote{\ 2 Chronicles 30:21\index{14 2 Chronicles:ch.  30, v.  21, pg.
``f30!}{}-16\index{14 2 Chronicles:ch.  30, v.  21-16, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg.
``f30!}, 46\index{44 Acts:ch.   2, v.  42, 46, pg. ``f30!}.} beg the continuance of it,\footnote{\ Psalm 36:10\index{19
Psalms:ch.  36, v.  10, pg. ``f30!}; Song of Solomon 3:4\index{22 Song of Solomon:ch.   3, v.   4, pg. ``f30!}; 1
Chronicles 29:18\index{13 1 Chronicles:ch.  29, v.  18, pg. ``f30!}.} watch against\index{Temptation:watchfulness
required to avoid ``f10!} relapses\index{Lord's Supper:after participation:watch against relapses ``f10!},\footnote{\ 1
Corinthians 10:3-5, 12\index{46 1 Corinthians:ch.  10, v.   3-5, 12, pg. ``f30!}.} fulfil their\index{Oaths and
Vows:duty to fulfill ``f10!} vows,\footnote{\ Psalm 50:14\index{19 Psalms:ch.  50, v.  14, pg. ``f30!}.} and encourage
themselves to a frequent attendance\index{Lord's Supper:attendance to be frequent ``f10!} on that
ordinance:\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1
Corinthians:ch.  11, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  25-26, pg. ``f30!}; Acts
2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}, 46\index{44 Acts:ch.   2, v.  42, 46, pg. ``f30!}.} but if they find
no present benefit, more exactly to review their preparation to, and\index{Lord's Supper:preparation:actions ``f10!}
carriage at, the sacrament;\footnote{\ Song of Solomon 5:1-6\index{22 Song of Solomon:ch.   5, v.   1-6, pg. ``f30!};
Ecclesiastes 5:1\index{21 Ecclesiastes:ch.   5, v.   1, pg. ``f30!}{}-6\index{21 Ecclesiastes:ch.   5, v.   1-6, pg.
``f30!}.} in both which, if they can approve\index{Lord's Supper:after participation:application ``f10!} themselves to
God and their own consciences, they\index{Conscience:approved ``f10!} are to wait for\index{Lord's Supper:waiting on
God ``f10!} the fruit of it in due time:\footnote{\ Psalm 42:5, 8\index{19 Psalms:ch.  42, v.   5, 8, pg. ``f30!};
Psalm 43:3-5\index{19 Psalms:ch.  43, v.   3-5, pg. ``f30!}; Psalm 123:1\index{19 Psalms:ch. 123, v.   1, pg.
``f30!}{}-2\index{19 Psalms:ch. 123, v.   1-2, pg. ``f30!}.} but, if they see they have failed\index{God's law:broken
daily by believers ``f10!} in either, they are to be humbled\index{Humility:before God ``f10!},\footnote{\ 2 Chronicles
30:18-19\index{14 2 Chronicles:ch.  30, v.  18-19, pg. ``f30!}; Isaiah 1:16, 18\index{23 Isaiah:ch.   1, v.  16, 18,
pg. ``f30!}.} and to attend upon it afterwards with more care and diligence\index{Lord's Supper:diligence in
application ``f10!}.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}1\index{47 2
Corinthians:ch.   7, v.  11, pg. ``f30!}; 1 Chronicles 15:12-14\index{13 1 Chronicles:ch.  15, v.  12-14, pg. ``f30!}.}


\bigskip

Q. 176 \textit{Wherein do the sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ of baptism
and}\index{Baptism:sacrament ``f10!}\textit{ the Lord's supper}\index{Lord's Supper:sacrament ``f10!}\textit{ agree?}

A. The sacraments\index{Sacraments:holy:ordinances ``f10!} of baptism and\index{Baptism:sacrament ``f10!} the Lord's
supper\index{Lord's Supper:sacrament ``f10!} agree, in that the author of both\index{Sacraments:God is author ``f10!}
is God;\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}.} the spiritual
part\index{Sacraments:spiritual part is Christ ``f10!} of both is Christ and his
benefits\index{Sacraments:benefits:worthy believers receive ``f10!};\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6,
v.   3-4, pg. ``f30!}; 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}.} both are
seals\index{Sacraments:covenant seals ``f10!} of the same covenant,\index{Covenant:new:seals ``f10!}\footnote{\ Romans
4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Colossians 2:12\index{51 Colossians:ch.   2, v.  12, pg. ``f30!};
Matthew 26:27-28\index{40 Matthew:ch.  26, v.  27-28, pg. ``f30!}.} are to be dispensed\index{Sacraments:dispensed by
minister of the Word ``f10!} by ministers\index{Church officers:minsters ``f10!} of the gospel\index{Gospel:gospel
``f10!}, and by none other;\footnote{\ John 1:3\index{43 John:ch.   1, v.   3, pg. ``f30!}3\index{43 John:ch.   1, v. 
33, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 4:1\index{46 1
Corinthians:ch.   4, v.   1, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}; Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg.
``f30!}.} and to be continued in the church of Christ\index{Church:Christ's ``f10!} until his second
coming\index{Jesus:second coming ``f10!}.\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg.
``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.  
2, pg. ``f30!}6\index{46 1 Corinthians:ch.  11, v.  26, pg. ``f30!}.}


\bigskip

Q. 177 \textit{Wherein do the sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ of baptism
and}\index{Baptism:sacrament ``f10!}\textit{ the Lord's supper}\index{Lord's Supper:sacrament ``f10!}\textit{ differ?}

A. The sacraments\index{Sacraments:holy:ordinances ``f10!} of baptism and\index{Baptism:sacrament ``f10!} the Lord's
supper\index{Lord's Supper:sacrament ``f10!} differ, in that baptism is to be administered but once\index{Baptism:to be
administered once ``f10!}, with water, to\index{Baptism:to use water ``f10!} be a sign\index{Baptism:sign of salvation
``f10!} and seal of\index{Baptism:seal of salvationof ``f10!} our regeneration\index{Regeneration:regeneration ``f10!}
and ingrafting\index{Believers:ingrafted into Christ ``f10!} into Christ,\footnote{\ Matthew 3:11\index{40 Matthew:ch. 
 3, v.  11, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}; Galatians 3:27\index{48 Galatians:ch.  
3, v.  27, pg. ``f30!}.} and that even to infants\index{Infants:of believing parents, to be baptized
``f10!};\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 9\index{01 Genesis:ch.  17, v.   7, 9,
pg. ``f30!}; Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!}; 1
Corinthians 7:14\index{46 1 Corinthians:ch.   7, v.  14, pg. ``f30!}.} whereas the Lord's supper is to be administered
often\index{Lord's Supper:administer often ``f10!}, in the elements\index{Lord's Supper:elements:bread and wine ``f10!}
of bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}, to represent and\index{Lord's
Supper:elements:represent Christ ``f10!} exhibit Christ as spiritual nourishment\index{Lord's Supper:spiritual
feeding:nourishment ``f10!} to the soul,\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  23-26, pg.
``f30!}.} and to confirm\index{Sacraments:confirm union with Christ ``f10!} our continuance and growth in
him\index{Sanctification:spiritual growth:in Christ ``f10!},\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch. 
10, v.  16, pg. ``f30!}.} and that only to such as are of years and ability to examine themselves\index{Lord's
Supper:preparation:self-examination ``f10!}.\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}8\index{46 1 Corinthians:ch.  11, v.  28, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  28-29, pg.
``f30!}.}


\bigskip

Q. 178 \textit{What is prayer}\index{Prayer:defined ``f10!}\textit{?}

A. Prayer is an\index{Prayer:defined ``f10!} offering up of our\index{Prayer:offering desires to God ``f10!} desires
unto God,\footnote{\ Psalm 62:8\index{19 Psalms:ch.  62, v.   8, pg. ``f30!}.} in the name of Christ,\footnote{\ John
16:2\index{43 John:ch.  16, v.   2, pg. ``f30!}3\index{43 John:ch.  16, v.  23, pg. ``f30!}.} by the help of
his\index{Prayer:helped by Holy Spirit ``f10!} Spirit;\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg.
``f30!}6\index{45 Romans:ch.   8, v.  26, pg. ``f30!}.} with confession\index{Confession:of sin ``f10!} of our
sins,\footnote{\ Psalm 32:5\index{19 Psalms:ch.  32, v.   5, pg. ``f30!}{}-6\index{19 Psalms:ch.  32, v.   5-6, pg.
``f30!}; Daniel 9:4\index{27 Daniel:ch.   9, v.   4, pg. ``f30!}.} and thankful\index{Prayer:thanking God ``f10!}
acknowledgment\index{Prayer:acknowledgment of:God's mercies ``f10!} of his mercies\index{God's attributes:merciful
``f10!}.\footnote{\ Philippians 4:6\index{50 Philippians:ch.   4, v.   6, pg. ``f30!}.}


\bigskip


Q. 179 \textit{Are we to pray unto}\index{Prayer:to God alone ``f10!}\textit{ God only?}

A. God only being able to search the\index{Heart:God alone able to search ``f10!} hearts,\footnote{\ 1 Kings
8:39\index{11 1 Kings:ch.   8, v.  39, pg. ``f30!}; Acts 1:24\index{44 Acts:ch.   1, v.  24, pg. ``f30!}; Romans
8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}7\index{45 Romans:ch.   8, v.  27, pg. ``f30!}.} hear the
requests\index{Prayer:assurance of being heard ``f10!},\footnote{\ Psalm 65:2\index{19 Psalms:ch.  65, v.   2, pg.
``f30!}.} pardon the\index{Forgiveness:God alone pardons ``f10!} sins,\footnote{\ Micah 7:18\index{33 Micah:ch.   7, v.
 18, pg. ``f30!}.} and fulfil the desires of\index{Prayer:offering desires to God ``f10!} all;\footnote{\ Psalm
145:18\index{19 Psalms:ch. 145, v.  18, pg. ``f30!}{}-19\index{19 Psalms:ch. 145, v.  18-19, pg. ``f30!}.} and only to
be believed in\index{Faith:belief to be in God alone ``f10!},\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1,
pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg. ``f30!}.} and worshiped\index{Worship:adoration of God ``f10!} with
religious worship\index{Worship:of God, required of all men ``f10!};\footnote{\ Matthew 4:10\index{40 Matthew:ch.   4,
v.  10, pg. ``f30!}.} prayer, which is a\index{Prayer:act of worship ``f10!} special part thereof,\footnote{\ 1
Corinthians 1:2\index{46 1 Corinthians:ch.   1, v.   2, pg. ``f30!}.} is to be made by all to him
alone,\footnote{\ Psalm 50:15\index{19 Psalms:ch.  50, v.  15, pg. ``f30!}.} and to none other.\footnote{\ Romans
10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg. ``f30!}.}


\bigskip

Q. 180 \textit{What is it to pray in the}\index{Prayer:in the name of Christ ``f10!}\textit{ name of Christ?}

A. To pray in the\index{Prayer:in the name of Christ ``f10!} name of Christ is, in obedience\index{God's
law:commandments:obedience:required ``f10!} to his command, and\index{Prayer:commanded ``f10!} in
confidence\index{Prayer:with confidence ``f10!} on his promises\index{Prayer:confidence on God's promises ``f10!}, to
ask mercy for\index{Mercy:prayer for ``f10!} his sake;\footnote{\ John 14:13-14\index{43 John:ch.  14, v.  13-14, pg.
``f30!}; John 16:2\index{43 John:ch.  16, v.   2, pg. ``f30!}4\index{43 John:ch.  16, v.  24, pg. ``f30!}; Daniel
9:17\index{27 Daniel:ch.   9, v.  17, pg. ``f30!}.} not by bare mentioning of his name,\footnote{\ Matthew
7:21\index{40 Matthew:ch.   7, v.  21, pg. ``f30!}.} but by drawing our\index{Prayer:drawing encouragement ``f10!}
encouragement to pray, and our boldness\index{Prayer:with boldness ``f10!}, strength, and hope\index{Prayer:with
strength and hope ``f10!} of acceptance in\index{Prayer:accepted in Christ ``f10!} prayer, from Christ and his
mediation\index{Jesus:mediator:of covenant ``f10!}.\footnote{\ Hebrews 4:14-16\index{58 Hebrews:ch.   4, v.  14-16, pg.
``f30!}; 1 John 5:13\index{62 1 John:ch.   5, v.  13, pg. ``f30!}{}-15\index{62 1 John:ch.   5, v.  13-15, pg.
``f30!}.}


\bigskip

Q. 181 \textit{Why are we to pray in the}\index{Prayer:in the name of Christ ``f10!}\textit{ name of Christ?}

A. The sinfulness of man, and his distance\index{Mankind:great distance between him and God ``f10!} from God by reason
thereof, being so great, as that we can have no access into\index{Believers:access to God only through Christ ``f10!}
his presence without a mediator\index{Jesus:mediator:of covenant ``f10!};\footnote{\ John 14:6\index{43 John:ch.  14,
v.   6, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}; Ephesians 3:12\index{49 Ephesians:ch.  
3, v.  12, pg. ``f30!}.} and there being none in heaven\index{Jesus:mediator:only one ``f10!} or earth appointed to, or
fit for, that glorious work\index{Jesus:glorious work ``f10!} but Christ alone,\footnote{\ John 6:27\index{43 John:ch. 
 6, v.  27, pg. ``f30!}; Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}{}-27\index{58 Hebrews:ch.   7, v. 
25-27, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.} we are to pray in
no\index{Prayer:in the name of Christ ``f10!} other name but his only.\footnote{\ Colossians 3:17\index{51
Colossians:ch.   3, v.  17, pg. ``f30!}; Hebrews 13:15\index{58 Hebrews:ch.  13, v.  15, pg. ``f30!}.}


\bigskip

Q. 182 \textit{How does the Spirit help us to}\index{Prayer:helped by Holy Spirit ``f10!}\textit{ pray?}

A. We not knowing\index{Prayer:not know what we ought to pray ``f10!}: What to pray for as\index{Prayer:helped by Holy
Spirit ``f10!} we ought, the Spirit helps our infirmities, by\index{Believers:helped in infirmities by the Holy Spirit
``f10!} enabling us\index{Prayer:enabled by Holy Spirit ``f10!} to understand both for whom, and: What, and: How
prayer\index{Prayer:direction for ``f10!} is to be made; and by working and quickening in
our\index{Conversion:quickened to new life ``f10!} hearts (although not in all persons, nor at all times, in the same
measure) those apprehensions, affections, and graces\index{Prayer:right affections ``f10!} which are requisite
for\index{God's law:commandments:duties:required by God ``f10!} the right performance of that duty\index{Prayer:grace
required to perform ``f10!}.\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}6\index{45 Romans:ch.  
8, v.  26, pg. ``f30!}{}-27\index{45 Romans:ch.   8, v.  26-27, pg. ``f30!}; Psalm 10:17\index{19 Psalms:ch.  10, v. 
17, pg. ``f30!}; Zechariah 12:10\index{38 Zechariah:ch.  12, v.  10, pg. ``f30!}.}


\bigskip


\bigskip


Q. 183 \textit{For whom are we to pray}\index{Prayer:for others ``f10!}\textit{?}

A. We are to pray for the\index{Prayer:for whole Church ``f10!} whole church\index{Church:universal ``f10!} of Christ
upon earth;\footnote{\ Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}; Psalm 28:9\index{19 Psalms:ch. 
28, v.   9, pg. ``f30!}.} for magistrates\index{Civil magistrate:pray for ``f10!},\footnote{\ 1 Timothy 2:1\index{54 1
Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}.} and ministers\index{Church
officers:minsters ``f10!};\footnote{\ Colossians 4:3\index{51 Colossians:ch.   4, v.   3, pg. ``f30!}.} for
ourselves,\footnote{\ Genesis 32:11\index{01 Genesis:ch.  32, v.  11, pg. ``f30!}.} our brethren\index{Prayer:for
brothers in Christ ``f10!},\footnote{\ James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}.} yea, our
enemies\index{Enemies:believers delivered from ``f10!};\footnote{\ Matthew 5:44\index{40 Matthew:ch.   5, v.  44, pg.
``f30!}.} and for all sorts of men living,\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg.
``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}.} or that shall live hereafter;\footnote{\ John
17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}0\index{43 John:ch.  17, v.  20, pg. ``f30!}; 2 Samuel 7:29\index{10 2
Samuel:ch.   7, v.  29, pg. ``f30!}.} but not for the dead\index{Prayer:not to be offered for the dead
``f10!},\footnote{\ 2 Samuel 12:21-23\index{10 2 Samuel:ch.  12, v.  21-23, pg. ``f30!}.} nor for those that are known
to have sinned the sin\index{Sin:sinned the sin unto death ``f10!} unto death\index{Death:committed sin unto
``f10!}.\footnote{\ 1 John 5:16\index{62 1 John:ch.   5, v.  16, pg. ``f30!}.}


\bigskip

Q. 184 \textit{For what things are we to pray}\index{Prayer:for lawful things ``f10!}\textit{?}

A. We are to pray for all\index{Prayer:for glory of God ``f10!} things tending to the glory of God,\index{Mankind:to
glorify God in thought, word, deed ``f10!}\footnote{\ Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}.} the
welfare\index{Prayer:for welfare of the Church ``f10!} of the church,\footnote{\ Psalm 51:1\index{19 Psalms:ch.  51, v.
  1, pg. ``f30!}8\index{19 Psalms:ch.  51, v.  18, pg. ``f30!}; Psalm 122:6\index{19 Psalms:ch. 122, v.   6, pg.
``f30!}.} our own\footnote{\ Matthew 7:1\index{40 Matthew:ch.   7, v.   1, pg. ``f30!}1\index{40 Matthew:ch.   7, v. 
11, pg. ``f30!}.} or others good;\footnote{\ Psalm 125:4\index{19 Psalms:ch. 125, v.   4, pg. ``f30!}.} but not for
anything that is unlawful\index{Prayer:not for unlawful ``f10!}.\footnote{\ 1 John 5:14\index{62 1 John:ch.   5, v. 
14, pg. ``f30!}.}


\bigskip


\bigskip


Q. 185 \textit{How are we to pray}\index{Prayer:direction for ``f10!}\textit{?}

A. We are to pray with an\index{Prayer:praising God ``f10!} awful apprehension\index{Prayer:apprehension of God's
majesty ``f10!} of the majesty of God,\footnote{\ Ecclesiastes 5:1\index{21 Ecclesiastes:ch.   5, v.   1, pg. ``f30!}.}
and deep sense of our own unworthiness\index{Prayer:acknowledgment of:deep sense of unworthiness
``f10!},\footnote{\ Genesis 18:27\index{01 Genesis:ch.  18, v.  27, pg. ``f30!}; Genesis 32:10\index{01 Genesis:ch. 
32, v.  10, pg. ``f30!}.} necessities,\index{Prayer:for necessities ``f10!}\footnote{\ Luke 15:17-19\index{42 Luke:ch. 
15, v.  17-19, pg. ``f30!}.} and sins;\footnote{\ Luke 18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}3-14\index{42
Luke:ch.  18, v.  13-14, pg. ``f30!}.} with penitent\index{Repentance:penitence ``f10!},\footnote{\ Psalm 51:1\index{19
Psalms:ch.  51, v.   1, pg. ``f30!}7\index{19 Psalms:ch.  51, v.  17, pg. ``f30!}.} thankful\index{Prayer:thanking God
``f10!},\footnote{\ Philippians 4:6\index{50 Philippians:ch.   4, v.   6, pg. ``f30!}.} and enlarged
hearts\index{Prayer:with understanding and faith ``f10!};\footnote{\ 1 Samuel 1:15\index{09 1 Samuel:ch.   1, v.  15,
pg. ``f30!}; 1 Samuel 2:1\index{09 1 Samuel:ch.   2, v.   1, pg. ``f30!}.} with understanding,\footnote{\ 1 Corinthians
14:15\index{46 1 Corinthians:ch.  14, v.  15, pg. ``f30!}.} faith,\footnote{\ Mark 11:24\index{41 Mark:ch.  11, v.  24,
pg. ``f30!}; James 1:6\index{59 James:ch.   1, v.   6, pg. ``f30!}.} sincerity,\footnote{\ Psalm 17:1\index{19
Psalms:ch.  17, v.   1, pg. ``f30!}; Psalm 145:18\index{19 Psalms:ch. 145, v.  18, pg. ``f30!}.}
fervency\index{Prayer:with fervency ``f10!},\footnote{\ James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}.}
love,\footnote{\ 1 Timothy 2:8\index{54 1 Timothy:ch.   2, v.   8, pg. ``f30!}.} and perseverance,\index{Prayer:with
perseverance ``f10!}\footnote{\ Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}.} waiting upon
him,\index{Patience:waiting on God ``f10!}\footnote{\ Micah 7:7\index{33 Micah:ch.   7, v.   7, pg. ``f30!}.} with
humble submission\index{Humility:submission before God ``f10!} to his will.\footnote{\ Matthew 26:39\index{40
Matthew:ch.  26, v.  39, pg. ``f30!}.}


\bigskip

Q. 186 \textit{What rule has God given for}\index{Prayer:direction for ``f10!}\textit{ our direction in the duty of
prayer}\index{Prayer:duty of ``f10!}\textit{?}

A. The whole word of God is of use to direct us in the duty of prayer\index{Prayer:duty of ``f10!};\footnote{\ 1 John
5:14\index{62 1 John:ch.   5, v.  14, pg. ``f30!}.} but the special rule of direction\index{Prayer:direction for
``f10!} is that form of prayer\index{Prayer:Lord's:pattern for our prayers ``f10!} which our Savior\index{Jesus:saviour
of the Church ``f10!} Christ taught his\index{Prayer:Lord's:Jesus taught his disciples ``f10!}
disciples\index{disciples of Jesus ``f10!}, commonly called The Lord's prayer.\footnote{\ Matthew 6:2-13\index{40
Matthew:ch.   6, v.   2-13, pg. ``f30!}; Luke 11:2\index{42 Luke:ch.  11, v.   2, pg. ``f30!}{}-4\index{42 Luke:ch. 
11, v.   2-4, pg. ``f30!}.}


\bigskip

Q. 187 \textit{How is the Lord's prayer}\index{Prayer:Lord's:pattern for our prayers ``f10!}\textit{ to be used?}

A. The Lord's prayer\index{Prayer:Lord's:pattern for our prayers ``f10!} is not only for
direction\index{Prayer:direction for ``f10!}, as a pattern, according to which we are to make other prayers; but may
also be used as a prayer, so that it be done with understanding\index{Prayer:with understanding and faith ``f10!},
faith, reverence, and other graces\index{Prayer:accompanied by graces ``f10!} necessary to the right performance of the
duty of prayer\index{Prayer:duty of ``f10!}.\footnote{\ Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}; Luke
11:2\index{42 Luke:ch.  11, v.   2, pg. ``f30!}.}


\bigskip

Q. 188 \textit{Of how many parts does the Lord's prayer}\index{Prayer:Lord's:pattern for our prayers ``f10!}\textit{
consist?}

A. The Lord's prayer\index{Prayer:Lord's:pattern for our prayers ``f10!} consists of three parts; a
preface\index{Prayer:Lord's:preface ``f10!}, petitions\index{Prayer:Lord's:petitions:petitions ``f10!}, and a
conclusion\index{Prayer:Lord's:conclusion ``f10!}.


\bigskip

Q. 189 \textit{What does the preface of}\index{Prayer:Lord's:preface ``f10!}\textit{ the Lord's
prayer}\index{Prayer:Lord's:pattern for our prayers ``f10!}\textit{ teach us?}

A. The preface of\index{Prayer:Lord's:preface ``f10!} the Lord's prayer\index{Prayer:Lord's:pattern for our prayers
``f10!} (contained in these words, Our Father\index{Father:God the ``f10!} which art in heaven\index{Heaven:God's abode
``f10!},)\footnote{\ Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}.} teaches us, when we pray, to draw
near\index{Prayer:draw near to God ``f10!} to God with confidence\index{Prayer:with confidence ``f10!} of his fatherly
goodness\index{God's attributes:goodness ``f10!}, and our interest therein\index{Prayer:acknowledgment of:interest
(share) in God as Father ``f10!};\footnote{\ Luke 11:13\index{42 Luke:ch.  11, v.  13, pg. ``f30!}; Romans 8:1\index{45
Romans:ch.   8, v.   1, pg. ``f30!}5\index{45 Romans:ch.   8, v.  15, pg. ``f30!}.} with reverence, and
all\index{Prayer:with reverence ``f10!} other childlike\index{Love toward God:childlike ``f10!}
dispositions,\footnote{\ Isaiah 64:9\index{23 Isaiah:ch.  64, v.   9, pg. ``f30!}.} heavenly
affections,\footnote{\ Psalm 123:1\index{19 Psalms:ch. 123, v.   1, pg. ``f30!}; Lamentations 3:41\index{25
Lamentations:ch.   3, v.  41, pg. ``f30!}.} and due\index{Worship:affections directed to God ``f10!} apprehensions of
his sovereign\index{God's attributes:sovereign lordship ``f10!} power, majesty, and\index{Prayer:apprehension of God's
majesty ``f10!} gracious\index{Grace:God's free ``f10!} condescension\index{God's attributes:condescension, voluntary
``f10!}:\footnote{\ Isaiah 63:15-16\index{23 Isaiah:ch.  63, v.  15-16, pg. ``f30!}; Nehemiah 1:4\index{16 Nehemiah:ch.
  1, v.   4, pg. ``f30!}{}-6\index{16 Nehemiah:ch.   1, v.   4-6, pg. ``f30!}.} as also, to pray with
an\index{Prayer:praising God ``f10!}d\index{Prayer:for others ``f10!} for others.\footnote{\ Acts 12:5\index{44
Acts:ch.  12, v.   5, pg. ``f30!}.}


\bigskip

Q. 190 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the first
petition}\index{Prayer:Lord's:petitions:first ``f10!}\textit{?}

A. In the first petition\index{Prayer:Lord's:petitions:first ``f10!} (which is, Hallowed\index{God's name:hallowed
``f10!} be thy name,)\footnote{\ Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}.} acknowledging the
utter\index{Prayer:acknowledgment of:inability to honour God rightly ``f10!} inability\index{Prayer:acknowledgment
of:reliance on God ``f10!} and indisposition\index{Prayer:acknowledgment of:indisposition ``f10!} that is in ourselves
and all men to honor God\index{Mankind:to honor God ``f10!} aright\index{Sin:original:inability to:honour God rightly
``f10!},\footnote{\ 2 Corinthians 3:5\index{47 2 Corinthians:ch.   3, v.   5, pg. ``f30!}; Psalm 51:1\index{19
Psalms:ch.  51, v.   1, pg. ``f30!}5\index{19 Psalms:ch.  51, v.  15, pg. ``f30!}.} pray,
that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God would by his grace\index{Grace:God's free ``f10!}
enable and\index{Worship:correct:enabled by Holy Spirit ``f10!} incline us and others to know, to acknowledge,
and\index{Worship:acknowledge God ``f10!} highly to esteem him\index{Worship:esteem God ``f10!},\footnote{\ Psalm
67:2-3\index{19 Psalms:ch.  67, v.   2-3, pg. ``f30!}.} his titles\index{God's name:titles to be honoured
``f10!},\footnote{\ Psalm 83:18\index{19 Psalms:ch.  83, v.  18, pg. ``f30!}.} attributes\index{God's attributes:to
be:revered ``f10!},\footnote{\ Psalm 86:10-13, 15\index{19 Psalms:ch.  86, v.  10-13, 15, pg. ``f30!}.} ordinances,
word\index{Worship:ordinances:to be:esteemed ``f10!},\footnote{\ 2 Thessalonians 3:1\index{53 2 Thessalonians:ch.   3,
v.   1, pg. ``f30!}; Psalm 138:1-3\index{19 Psalms:ch. 138, v.   1-3, pg. ``f30!}; Psalm 147:19-20\index{19 Psalms:ch.
147, v.  19-20, pg. ``f30!}; 2 Corinthians 2:14-15\index{47 2 Corinthians:ch.   2, v.  14-15, pg. ``f30!}.} works, and
whatsoever\index{Providence:God's works ``f10!} he is pleased\index{Providence:at God's pleasure ``f10!} to make
himself known by;\footnote{\ Psalm 8; Psalm 145.} and to glorify\index{Worship:glorifying God ``f10!} him in
thought\index{God's name:to be used:reverently in thoughts and words ``f10!}, word,\footnote{\ Psalm 19:14\index{19
Psalms:ch.  19, v.  14, pg. ``f30!}; Psalm 103:1\index{19 Psalms:ch. 103, v.   1, pg. ``f30!}.} and
deed:\index{Mankind:to glorify God in thought, word, deed ``f10!}\footnote{\ Philippians 1:9, 11\index{50
Philippians:ch.   1, v.   9, 11, pg. ``f30!}.} that he would prevent and\index{Prayer:that God would prevent and remove
atheism ``f10!} remove atheism\index{atheism ``f10!},\footnote{\ Psalm 67:1-4\index{19 Psalms:ch.  67, v.   1-4, pg.
``f30!}.} ignorance\index{ignorance ``f10!},\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}7-18\index{49 Ephesians:ch.   1, v.  17-18, pg. ``f30!}.} idolatry\index{Idolatry:idolatry
``f10!},\footnote{\ Psalm 97:7\index{19 Psalms:ch.  97, v.   7, pg. ``f30!}.} profaneness\index{God's name:not to
be:profaned ``f10!},\footnote{\ Psalm 74:18, 22-23\index{19 Psalms:ch.  74, v.  18, 22-23, pg. ``f30!}.} and:
whatsoever is dishonorable\index{Prayer:for removal of all that is dishonorable to God ``f10!} to him;\footnote{\ 2
Kings 19:15-16\index{12 2 Kings:ch.  19, v.  15-16, pg. ``f30!}.} and, by his overruling\index{Providence:overruling
``f10!} providence\index{Providence:providence ``f10!}, direct and dispose of\index{Providence:God disposes all things
for his glory ``f10!} all things to his own glory\index{God's attributes:glory:his own ``f10!}.\footnote{\ 2 Chronicles
20:6, 10-12\index{14 2 Chronicles:ch.  20, v.   6, 10-12, pg. ``f30!}; Psalm 83; Psalm 140:4, 8\index{19 Psalms:ch.
140, v.   4, 8, pg. ``f30!}.}


\bigskip

Q. 191 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the second
petition}\index{Prayer:Lord's:petitions:second ``f10!}\textit{?}

A. In the second petition\index{Prayer:Lord's:petitions:second ``f10!} (which is, Thy kingdom come\index{Kingdom of
God:prayer for advancement ``f10!},)\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}0\index{40
Matthew:ch.   6, v.  10, pg. ``f30!}.} acknowledging ourselves and\index{Prayer:acknowledgment of:sin ``f10!} all
mankind to be\index{Sin:original:all mankind descending from Adam ``f10!} by nature under the dominion of
sin\index{Christian liberty:dominion of sin destroyed ``f10!} and Satan\index{Satan:bondage to
``f10!},\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   2, v.  
2-3, pg. ``f30!}.} we pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} the kingdom of
sin\index{Kingdom:of sin ``f10!} and Satan may be destroyed\index{Prayer:for kingdom of sin to be destroyed
``f10!},\footnote{\ Psalm 68:1, 18\index{19 Psalms:ch.  68, v.   1, 18, pg. ``f30!}; Revelation 12:10-11\index{66
Revelation:ch.  12, v.  10-11, pg. ``f30!}.} the gospel\index{Gospel:gospel ``f10!} propagated\index{Gospel:to all
nations ``f10!} throughout the world,\footnote{\ 2 Thessalonians 3:1\index{53 2 Thessalonians:ch.   3, v.   1, pg.
``f30!}.} the Jews called\index{Jews:called ``f10!},\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg.
``f30!}.} the fulness of the\index{Gentiles:fulness of ``f10!} Gentiles\index{Gentiles:called ``f10!} brought
in;\footnote{\ John 17:9\index{43 John:ch.  17, v.   9, pg. ``f30!}, 20\index{43 John:ch.  17, v.   9, 20, pg. ``f30!};
Romans 11:25-26\index{45 Romans:ch.  11, v.  25-26, pg. ``f30!}; Psalm 67.} the church
furnished\index{Church:institution ``f10!} with all gospel officers\index{Church officers:officers ``f10!} and
ordinances,\footnote{\ Matthew 9:38\index{40 Matthew:ch.   9, v.  38, pg. ``f30!}; 2 Thessalonians 3:1\index{53 2
Thessalonians:ch.   3, v.   1, pg. ``f30!}.} purged from corruption\index{Prayer:for Church ``f10!},\footnote{\ Malachi
1:11\index{39 Malachi:ch.   1, v.  11, pg. ``f30!}; Zephaniah 3:9\index{36 Zephaniah:ch.   3, v.   9, pg. ``f30!}.}
countenanced\index{Civil magistrate:to support Church ``f10!} and maintained by the civil magistrate:\index{Civil
magistrate:to ensure ordinances of God administrated ``f10!}\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.  
1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}.} that the ordinances of Christ may be purely
dispensed, and\index{Prayer:for ordinances to be dispensed purely ``f10!} made effectual to the
converting\index{Conversion:conversion ``f10!} of those that are yet in their sins, and the confirming\index{Love for
others:confirming ``f10!}, comforting\index{Love for others:comforting others ``f10!}, and building up\index{Holy
Spirit:builds up the elect ``f10!} of those that are already converted:\footnote{\ Acts 4:29-30\index{44 Acts:ch.   4,
v.  29-30, pg. ``f30!}; Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}{}-20\index{49 Ephesians:ch.   6,
v.  18-20, pg. ``f30!}; Romans 15:29-30, 32\index{45 Romans:ch.  15, v.  29-30, 32, pg. ``f30!}; 2 Thessalonians
1:11\index{53 2 Thessalonians:ch.   1, v.  11, pg. ``f30!}; 2 Thessalonians 2:16-17\index{53 2 Thessalonians:ch.   2,
v.  16-17, pg. ``f30!}.} that Christ would rule in our hearts here\index{Heart:new:ruled by Jesus
``f10!},\footnote{\ Ephesians 3:14-20\index{49 Ephesians:ch.   3, v.  14-20, pg. ``f30!}.} and
hasten\index{Jesus:return:desire to see hastened ``f10!} the time of his second coming\index{Jesus:second coming
``f10!}, and our reigning\index{Resurrection:believers:reigning with Christ ``f10!} with him
forever:\footnote{\ Revelation 22:20\index{66 Revelation:ch.  22, v.  20, pg. ``f30!}.} and that he would be
pleased\index{Providence:at God's pleasure ``f10!} so to exercise the kingdom of his\index{Kingdom:of Christ ``f10!}
power in all\index{Jesus:almighty power ``f10!} the world, as may best conduce to these ends.\footnote{\ Isaiah
64:1-2\index{23 Isaiah:ch.  64, v.   1-2, pg. ``f30!}; Revelation 4:8\index{66 Revelation:ch.   4, v.   8, pg.
``f30!}{}-11\index{66 Revelation:ch.   4, v.   8-11, pg. ``f30!}.}


\bigskip

Q. 192 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the third
petition}\index{Prayer:Lord's:petitions:third ``f10!}\textit{?}

A. In the third petition\index{Prayer:Lord's:petitions:third ``f10!} (which is, Thy will be done in earth, as it is in
heaven\index{Heaven:God's will done ``f10!},)\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg.
``f30!}0\index{40 Matthew:ch.   6, v.  10, pg. ``f30!}.} acknowledging, that by\index{Prayer:acknowledgment of:sin
``f10!} nature we and all men are not only utterly unable\index{Sin:original:utterly unable and unwilling to do will of
God ``f10!} and unwilling to know and do the will of God,\footnote{\ Romans 7:18\index{45 Romans:ch.   7, v.  18, pg.
``f30!}; Job 21:14\index{18 Job:ch.  21, v.  14, pg. ``f30!}; 1 Corinthians 2:14\index{46 1 Corinthians:ch.   2, v. 
14, pg. ``f30!}.} but prone to rebel\index{Scripture:people prone to rebel against ``f10!} against his
word,\footnote{\ Romans 8:7\index{45 Romans:ch.   8, v.   7, pg. ``f30!}.} to repine\index{Providence:repine and murmur
against ``f10!} and murmur\index{murmuring ``f10!} against his providence\index{Providence:providence
``f10!},\footnote{\ Exodus 17:7\index{02 Exodus:ch.  17, v.   7, pg. ``f30!}; Numbers 14:2\index{04 Numbers:ch.  14, v.
  2, pg. ``f30!}.} and wholly inclined\index{Sin:original:inclined to do evil ``f10!} to do the will of the flesh, and
of the devil\index{Devil:devil ``f10!}:\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}.} we
pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God would by his Spirit take away from ourselves and
others all blindness\index{blindness, spiritual ``f10!},\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1,
pg. ``f30!}7-18\index{49 Ephesians:ch.   1, v.  17-18, pg. ``f30!}.} weakness,\footnote{\ Ephesians 3:16\index{49
Ephesians:ch.   3, v.  16, pg. ``f30!}.} indisposedness\index{Sin:original:utterly indisposed
``f10!},\footnote{\ Matthew 26:40-41\index{40 Matthew:ch.  26, v.  40-41, pg. ``f30!}.} and perverseness of
heart;\index{Heart:perverse ``f10!}\footnote{\ Jeremiah 31:18\index{24 Jeremiah:ch.  31, v.  18, pg.
``f30!}{}-19\index{24 Jeremiah:ch.  31, v.  18-19, pg. ``f30!}.} and by his grace\index{Grace:God's free ``f10!} make
us able and willing to know\index{Conversion:renewal of the will ``f10!}, do, and submit\index{Holy Spirit:subdues will
of man ``f10!} to his will in all things,\footnote{\ Psalm 119:1, 8, 35-36\index{19 Psalms:ch. 119, v.   1, 8, 35-36,
pg. ``f30!}; Acts 21:14\index{44 Acts:ch.  21, v.  14, pg. ``f30!}.} with the like humility\index{Humility:before God
``f10!},\footnote{\ Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}.} cheerfulness\index{cheerfulness
``f10!},\footnote{\ Psalm 100:2\index{19 Psalms:ch. 100, v.   2, pg. ``f30!}; Job 1:21\index{18 Job:ch.   1, v.  21,
pg. ``f30!}; 2 Samuel 15:25-26\index{10 2 Samuel:ch.  15, v.  25-26, pg. ``f30!}.} faithfulness\index{faithfulness
``f10!},\footnote{\ Isaiah 38:3\index{23 Isaiah:ch.  38, v.   3, pg. ``f30!}.} diligence,\footnote{\ Psalm
119:4-5\index{19 Psalms:ch. 119, v.   4-5, pg. ``f30!}.} zeal,\footnote{\ Romans 12:11\index{45 Romans:ch.  12, v.  11,
pg. ``f30!}.} sincerity,\footnote{\ Psalm 119:80\index{19 Psalms:ch. 119, v.  80, pg. ``f30!}.} and
constancy\index{Believers:desire to obey God ``f10!},\footnote{\ Psalm 119:11\index{19 Psalms:ch. 119, v.  11, pg.
``f30!}2\index{19 Psalms:ch. 119, v. 112, pg. ``f30!}.} as the angels\index{Angels:angels ``f10!} do in
heaven\index{Angels:works in heaven ``f10!}.\footnote{\ Isaiah 6:2-3\index{23 Isaiah:ch.   6, v.   2-3, pg. ``f30!};
Psalm 103:20-21\index{19 Psalms:ch. 103, v.  20-21, pg. ``f30!}; Matthew 18:10\index{40 Matthew:ch.  18, v.  10, pg.
``f30!}.}

Q. 193 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the fourth
petition}\index{Prayer:Lord's:petitions:fourth ``f10!}\textit{?}

A. In the fourth petition\index{Prayer:Lord's:petitions:fourth ``f10!} (which is, Give us this day our daily
bread\index{Prayer:for daily needs ``f10!},)\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg.
``f30!}1\index{40 Matthew:ch.   6, v.  11, pg. ``f30!}.} acknowledging, that in\index{Prayer:acknowledgment of:sin
``f10!} Adam, and by our own sin, we have forfeited\index{Sin:forfeited right to outward blessings in life ``f10!} our
right to all the outward blessings of\index{Material things:temporal blessings ``f10!} this life, and
deserve\index{Sin:deserves damnation ``f10!} to be wholly deprived of them by God, and to have them
cursed\index{Curse:deserved for sin ``f10!} to us in the use of them;\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2,
v.  17, pg. ``f30!}; Genesis 3:17\index{01 Genesis:ch.   3, v.  17, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.
  2, pg. ``f30!}0\index{45 Romans:ch.   8, v.  20, pg. ``f30!}{}-22\index{45 Romans:ch.   8, v.  20-22, pg. ``f30!};
Jeremiah 5:25\index{24 Jeremiah:ch.   5, v.  25, pg. ``f30!}; Deuteronomy 28:15-68\index{05 Deuteronomy:ch.  28, v. 
15-68, pg. ``f30!}.} and that neither they of themselves are able to sustain us\index{Worldly:goods cannot sustain us
``f10!},\footnote{\ Deuteronomy 8:3\index{05 Deuteronomy:ch.   8, v.   3, pg. ``f30!}.} nor we to
merit,\index{Salvation:not by:merit ``f10!}\footnote{\ Genesis 32:10\index{01 Genesis:ch.  32, v.  10, pg. ``f30!}.} or
by our own industry\index{Salvation:not by:industry ``f10!} to procure them;\footnote{\ Deuteronomy 8:17\index{05
Deuteronomy:ch.   8, v.  17, pg. ``f30!}{}-18\index{05 Deuteronomy:ch.   8, v.  17-18, pg. ``f30!}.} but prone to
desire\index{Worldly:desires ``f10!},\footnote{\ Jeremiah 6:13\index{24 Jeremiah:ch.   6, v.  13, pg. ``f30!}; Mark
7:21-22\index{41 Mark:ch.   7, v.  21-22, pg. ``f30!}.} get,\footnote{\ Hosea 12:7\index{28 Hosea:ch.  12, v.   7, pg.
``f30!}.} and use them unlawfully\index{Material things:unlawful use ``f10!}:\footnote{\ James 4:3\index{59 James:ch.  
4, v.   3, pg. ``f30!}.} we pray for ourselves\index{Prayer:for ourselves ``f10!} and others, that both they and we,
waiting upon the\index{Patience:waiting on God ``f10!} providence\index{Providence:providence ``f10!} of God from day
to day in the use of lawful means\index{Material things:to be lawfully procured ``f10!}, may, of his free gift,
and\index{Providence:provides good gifts ``f10!} as to his fatherly\index{Father:God the ``f10!} wisdom
shall\index{God's attributes:wisdom ``f10!} seem best, enjoy a competent portion\index{Prayer:for necessities ``f10!}
of them;\footnote{\ Genesis 28:20\index{01 Genesis:ch.  28, v.  20, pg. ``f30!}; Genesis 43:12-14\index{01 Genesis:ch. 
43, v.  12-14, pg. ``f30!}; Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}; 2 Thessalonians
3:1\index{53 2 Thessalonians:ch.   3, v.   1, pg. ``f30!}1\index{53 2 Thessalonians:ch.   3, v.  11, pg.
``f30!}{}-12\index{53 2 Thessalonians:ch.   3, v.  11-12, pg. ``f30!}; Philippians 4:6\index{50 Philippians:ch.   4, v.
  6, pg. ``f30!}.} and have the same continued and blessed unto us in our holy\index{Believers:holy:being perfected
``f10!} and comfortable use of them,\footnote{\ 2 Timothy 4:3\index{55 2 Timothy:ch.   4, v.   3, pg.
``f30!}{}-5\index{55 2 Timothy:ch.   4, v.   3-5, pg. ``f30!}.} and contentment\index{contentment ``f10!} in
them;\footnote{\ 1 Timothy 6:6\index{54 1 Timothy:ch.   6, v.   6, pg. ``f30!}{}-8\index{54 1 Timothy:ch.   6, v.  
6-8, pg. ``f30!}.} and be kept from all things that are contrary to our temporal support\index{Prayer:for temporal
support ``f10!} and comfort.\footnote{\ Proverbs 30:8-9\index{20 Proverbs:ch.  30, v.   8-9, pg. ``f30!}.}


\bigskip

Q. 194 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the fifth
petition}\index{Prayer:Lord's:petitions:fifth ``f10!}\textit{?}

A. In the fifth petition\index{Prayer:Lord's:petitions:fifth ``f10!} (which is, Forgive\index{Forgiveness:beg pardon for
``f10!} us our debts\index{Forgiveness:of debtors ``f10!}, as we forgive our debtors,)\footnote{\ Matthew 6:1\index{40
Matthew:ch.   6, v.   1, pg. ``f30!}2\index{40 Matthew:ch.   6, v.  12, pg. ``f30!}.} acknowledging, that
we\index{Prayer:acknowledgment of:sin ``f10!} and all others are guilty\index{guilt ``f10!} both of original and actual
sin, and thereby become debtors to the justice of God\index{God's attributes:administers justice ``f10!}; and that
neither we, nor any other creature\index{creatures ``f10!}, can make the least satisfaction for
that\index{Jesus:satisfied divine justice ``f10!} debt:\footnote{\ Romans 3:9\index{45 Romans:ch.   3, v.   9, pg.
``f30!}{}-22\index{45 Romans:ch.   3, v.   9-22, pg. ``f30!}; Matthew 18:24-25\index{40 Matthew:ch.  18, v.  24-25, pg.
``f30!}; Psalm 130:3\index{19 Psalms:ch. 130, v.   3, pg. ``f30!}{}-4\index{19 Psalms:ch. 130, v.   3-4, pg. ``f30!}.}
we pray for ourselves\index{Prayer:for ourselves ``f10!} and others, that God of his free grace\index{Grace:God's free
``f10!} would, through the obedience and\index{Jesus:obedience:imputed to believers ``f10!} satisfaction of Christ,
apprehended\index{Faith:instrument:which applies Christ to believers ``f10!} and applied by faith\index{Faith:in Jesus
Christ ``f10!}, acquit\index{Judgement day:believers:acquitted by God ``f10!} us both from the guilt and
punishment\index{Punishment:of sin ``f10!} of sin,\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg.
``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}{}-26\index{45 Romans:ch.   3, v.  24-26, pg. ``f30!}; Hebrews
9:22\index{58 Hebrews:ch.   9, v.  22, pg. ``f30!}.} accept us in his Beloved\index{Jesus:beloved of the Father
``f10!};\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}{}-7\index{49 Ephesians:ch.   1, v.  
6-7, pg. ``f30!}.} continue his favor and grace to us,\footnote{\ 2 Peter 1:2\index{61 2 Peter:ch.   1, v.   2, pg.
``f30!}.} pardon our daily failings\index{God's law:broken daily by all men (thought, word, deed)
``f10!},\footnote{\ Hosea 14:2\index{28 Hosea:ch.  14, v.   2, pg. ``f30!}; Jeremiah 14:7\index{24 Jeremiah:ch.  14, v.
  7, pg. ``f30!}.} and fill us with peace\index{Prayer:for peace and joy ``f10!} and joy, in\index{Joy:in Holy Spirit
``f10!} giving us daily more and more assurance\index{Assurance:of forgiveness ``f10!} of
forgiveness;\footnote{\ Romans 15:13\index{45 Romans:ch.  15, v.  13, pg. ``f30!}; Psalm 51:7-10, 12\index{19
Psalms:ch.  51, v.   7-10, 12, pg. ``f30!}.} which we are the rather emboldened\index{Prayer:with boldness ``f10!} to
ask, and encouraged\index{Prayer:assurance of being heard ``f10!} to expect, when we have this testimony in ourselves,
that we from the heart\index{Forgiveness:from the heart ``f10!} forgive others their offenses\index{Forgiveness:of
offenses ``f10!}.\footnote{\ Luke 11:4\index{42 Luke:ch.  11, v.   4, pg. ``f30!}; Matthew 6:1\index{40 Matthew:ch.  
6, v.   1, pg. ``f30!}4-15\index{40 Matthew:ch.   6, v.  14-15, pg. ``f30!}; Matthew 18:35\index{40 Matthew:ch.  18, v.
 35, pg. ``f30!}.}

Q. 195 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the sixth
petition}\index{Prayer:Lord's:petitions:sixth ``f10!}\textit{?}

A. In the sixth petition\index{Prayer:Lord's:petitions:sixth ``f10!} (which is, And lead us not into
temptation\index{Temptation:temptation ``f10!}, but deliver us from\index{Salvation:delivered from thraldom to sin
``f10!} evil,)\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}3\index{40 Matthew:ch.   6, v.  13,
pg. ``f30!}.} acknowledging, that the\index{Prayer:acknowledgment of:God's mercies ``f10!} most wise\index{God's
attributes:wisdom ``f10!}, righteous, and gracious\index{Grace:God's free ``f10!} God, for divers holy\index{God's
attributes:holy:ends ``f10!} and just ends, may so order things, that we may be assaulted\index{Temptation:assaults
believers ``f10!}, foiled, and for a time led captive\index{Temptation:captive to ``f10!} by temptations;\footnote{\ 2
Chronicles 32:31\index{14 2 Chronicles:ch.  32, v.  31, pg. ``f30!}.} that Satan,\index{Satan:tempter
``f10!}\footnote{\ 1 Chronicles 21:1\index{13 1 Chronicles:ch.  21, v.   1, pg. ``f30!}.} the world,\footnote{\ Luke
21:34\index{42 Luke:ch.  21, v.  34, pg. ``f30!}; Mark 4:19\index{41 Mark:ch.   4, v.  19, pg. ``f30!}.} and the flesh,
are ready powerfully to draw us\index{Temptation:draws us aside ``f10!} aside, and ensnare\index{Prayer:for freedom
from temptation ``f10!} us;\footnote{\ James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}.} and that we, even after
the pardon of our\index{Forgiveness:God alone pardons ``f10!} sins, by reason of our
corruption\index{Sanctification:remnants of corruption remain ``f10!},\footnote{\ Galatians 5:1\index{48 Galatians:ch. 
 5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}.} weakness,
and\index{Sanctification:weaknesses and imperfections remain ``f10!} want of watchfulness\index{Temptation:watchfulness
required to avoid ``f10!},\footnote{\ Matthew 26:41\index{40 Matthew:ch.  26, v.  41, pg. ``f30!}.} are not only
subject to be tempted, and forward to expose\index{Temptation:tendency to expose self to ``f10!} ourselves unto
temptations,\footnote{\ Matthew 26:69-72\index{40 Matthew:ch.  26, v.  69-72, pg. ``f30!}; Galatians 2:11-14\index{48
Galatians:ch.   2, v.  11-14, pg. ``f30!}; 2 Chronicles 18:3\index{14 2 Chronicles:ch.  18, v.   3, pg. ``f30!}; 2
Chronicles 19:2\index{14 2 Chronicles:ch.  19, v.   2, pg. ``f30!}.} but also of ourselves
unable\index{Temptation:unable and unwilling to resist ``f10!} and unwilling to resist them\index{Temptation:unwilling
to resist ``f10!}, to recover out\index{Temptation:recovery from ``f10!} of them, and to improve
them;\footnote{\ Romans 7:23\index{45 Romans:ch.   7, v.  23, pg. ``f30!}{}-24\index{45 Romans:ch.   7, v.  23-24, pg.
``f30!}; 1 Chronicles 21:1\index{13 1 Chronicles:ch.  21, v.   1, pg. ``f30!}{}-4\index{13 1 Chronicles:ch.  21, v.  
1-4, pg. ``f30!}; 2 Chronicles 16:7-10\index{14 2 Chronicles:ch.  16, v.   7-10, pg. ``f30!}.} and worthy to be left
under the power of them\index{Temptation:under the power of ``f10!}:\footnote{\ Psalm 81:11\index{19 Psalms:ch.  81, v.
 11, pg. ``f30!}{}-12\index{19 Psalms:ch.  81, v.  11-12, pg. ``f30!}.} we pray,
that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God would so overrule\index{Providence:overruling ``f10!}
the world and all in it,\footnote{\ John 17:15\index{43 John:ch.  17, v.  15, pg. ``f30!}.} subdue the\index{Prayer:for
God to subdue the flesh (sin in the world) ``f10!} flesh,\footnote{\ Psalm 51:1\index{19 Psalms:ch.  51, v.   1, pg.
``f30!}0\index{19 Psalms:ch.  51, v.  10, pg. ``f30!}; Psalm 119:133\index{19 Psalms:ch. 119, v. 133, pg. ``f30!}.} and
restrain\index{Prayer:for restraining Satan ``f10!} Satan,\footnote{\ 2 Corinthians 12:7-8\index{47 2 Corinthians:ch. 
12, v.   7-8, pg. ``f30!}.} order all things,\footnote{\ 1 Corinthians 10:12-13\index{46 1 Corinthians:ch.  10, v. 
12-13, pg. ``f30!}.} bestow and bless all\index{Believers:blessed in/by God ``f10!} means of grace\index{Grace:means of
``f10!},\footnote{\ Hebrews 13:20-21\index{58 Hebrews:ch.  13, v.  20-21, pg. ``f30!}.} and quicken us to watchfulness
in the use of them, that we and all his people may by his providence\index{Providence:providence ``f10!} be kept from
being tempted to sin;\footnote{\ Matthew 26:41\index{40 Matthew:ch.  26, v.  41, pg. ``f30!}; Psalm 19:13\index{19
Psalms:ch.  19, v.  13, pg. ``f30!}.} or, if tempted, that by his Spirit we may be powerfully supported and\index{Holy
Spirit:strengthens believers against temptation ``f10!} enabled to stand\index{Temptation:enabled to stand by Holy
Spirit ``f10!} in the hour of\index{Temptation:delivery from ``f10!} temptation;\footnote{\ Ephesians 3:14-17\index{49
Ephesians:ch.   3, v.  14-17, pg. ``f30!}; 1 Thessalonians 3:13\index{52 1 Thessalonians:ch.   3, v.  13, pg. ``f30!};
Jude 1:24\index{65 Jude:v.   1, v.  24, pg. ``f30!}.} or when fallen\index{Prayer:when fallen ``f10!}, raised again and
recovered out of it,\footnote{\ Psalm 51:1\index{19 Psalms:ch.  51, v.   1, pg. ``f30!}2\index{19 Psalms:ch.  51, v. 
12, pg. ``f30!}.} and have a sanctified\index{Sanctification:sanctification ``f10!} use and improvement
thereof:\footnote{\ 1 Peter 5:8-10\index{60 1 Peter:ch.   5, v.   8-10, pg. ``f30!}.} that our sanctification and
salvation\index{Salvation:belief in Jesus by faith ``f10!} may be perfected\index{Sanctification:perfected
``f10!},\footnote{\ 2 Corinthians 13:7, 9\index{47 2 Corinthians:ch.  13, v.   7, 9, pg. ``f30!}.} Satan
trodden\index{Satan:trodden under our feet ``f10!} under our feet,\footnote{\ Romans 16:2\index{45 Romans:ch.  16, v.  
2, pg. ``f30!}0\index{45 Romans:ch.  16, v.  20, pg. ``f30!}; Zechariah 3:2\index{38 Zechariah:ch.   3, v.   2, pg.
``f30!}; Luke 22:31-32\index{42 Luke:ch.  22, v.  31-32, pg. ``f30!}.} and we fully freed from sin\index{Prayer:for
freedom from sin ``f10!}, temptation, and all evil\index{Prayer:for deliverance from evil ``f10!},
forever.\footnote{\ John 17:15\index{43 John:ch.  17, v.  15, pg. ``f30!}; 1 Thessalonians 5:23\index{52 1
Thessalonians:ch.   5, v.  23, pg. ``f30!}.}


\bigskip


\bigskip

Q. 196 \textit{What does the conclusion}\index{Prayer:Lord's:conclusion ``f10!}\textit{ of the Lord's
prayer}\index{Prayer:Lord's:pattern for our prayers ``f10!}\textit{ teach us?}\footnote{\ Although this
conclusion\index{Prayer:Lord's:conclusion ``f10!} is not found in the oldest yet discovered manuscripts, it is found in
the majority of Byzantine texts, and its content perfectly harmonizes with what is said about the kingdom in Daniel
7:14\index{27 Daniel:ch.   7, v.  14, pg. ``f30!}.}

A. The conclusion\index{Prayer:Lord's:conclusion ``f10!} of the Lord's prayer\index{Prayer:Lord's:pattern for our
prayers ``f10!} (which is, For thine is the kingdom, and the\index{Kingdom of God:prayer ascribing power and glory
``f10!} power, and the glory, forever\index{Prayer:for glory of God ``f10!}. Amen\index{Amen, testimony of our true
desire for God's glory ``f10!}.),\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}3\index{40
Matthew:ch.   6, v.  13, pg. ``f30!}.} teaches us to enforce our\index{Prayer:enforce petitions with arguments ``f10!}
petitions\index{Prayer:Lord's:petitions:petitions ``f10!} with arguments,\footnote{\ Romans 15:30\index{45 Romans:ch. 
15, v.  30, pg. ``f30!}.} which are to be taken, not from any worthiness\index{Prayer:accepted not by worthiness on our
part ``f10!} in ourselves, or in any other creature\index{creatures ``f10!}, but from God;\footnote{\ Daniel
9:4\index{27 Daniel:ch.   9, v.   4, pg. ``f30!}, 7-9, 16-19\index{27 Daniel:ch.   9, v.   4, 7-9, 16-19, pg. ``f30!}.}
and with our prayers\index{Prayer:praising God ``f10!} to join praises,\footnote{\ Philippians 4:6\index{50
Philippians:ch.   4, v.   6, pg. ``f30!}.} ascribing to\index{Worship:praising God ``f10!} God alone eternal
sovereignty\index{God's attributes:sovereign lordship ``f10!}, omnipotency\index{God's attributes:omnipotency ``f10!},
and glorious excellency\index{God's attributes:glory:excellency ``f10!};\footnote{\ 1 Chronicles 29:10-13\index{13 1
Chronicles:ch.  29, v.  10-13, pg. ``f30!}.} in regard whereof, as he is able and willing to help\index{Prayer:helped
by Holy Spirit ``f10!} us,\footnote{\ Ephesians 3:20\index{49 Ephesians:ch.   3, v.  20, pg. ``f30!}{}-21\index{49
Ephesians:ch.   3, v.  20-21, pg. ``f30!}; Luke 11:13\index{42 Luke:ch.  11, v.  13, pg. ``f30!}.} so we by
faith\index{Faith:in prayer ``f10!} are emboldened\index{Prayer:with boldness ``f10!} to plead with him that he
would,\footnote{\ 2 Chronicles 20:6, 11\index{14 2 Chronicles:ch.  20, v.   6, 11, pg. ``f30!}.} and
quietly\index{Prayer:quiet reliance on God ``f10!} to rely upon him, that he will fulfil our\index{Prayer:assurance of
being heard ``f10!} requests.\footnote{\ 2 Chronicles 14:11\index{14 2 Chronicles:ch.  14, v.  11, pg. ``f30!}.} And,
to testify\index{Sacraments:testimony through participation ``f10!} this our desire and\index{Prayer:offering desires
to God ``f10!} assurance, we\index{Assurance:of faith ``f10!} say, Amen.\footnote{\ 1 Corinthians 14:16\index{46 1
Corinthians:ch.  14, v.  16, pg. ``f30!}; Revelation 22:20\index{66 Revelation:ch.  22, v.  20, pg.
``f30!}{}-21\index{66 Revelation:ch.  22, v.  20-21, pg. ``f30!}.}



\subsubsection[The Westminster Shorter Catechism of 1648]{{The Westminster Shorter Catechism of
1648}}
Q 1. \textit{What is the chief}\index{Mankind:chief end/purpose ``f10!}\textit{ end of man?}

A. Man's chief\index{Mankind:chief end/purpose ``f10!} end is to glorify\index{Worship:glorifying God ``f10!}
God,\footnote{\ Psalm 86:9\index{19 Psalms:ch.  86, v.   9, pg. ``f30!}; Isaiah 60:21\index{23 Isaiah:ch.  60, v.  21,
pg. ``f30!}; Romans 11:36\index{45 Romans:ch.  11, v.  36, pg. ``f30!}; 1 Corinthians 6:20\index{46 1 Corinthians:ch.  
6, v.  20, pg. ``f30!}; 1 Corinthians 10:31\index{46 1 Corinthians:ch.  10, v.  31, pg. ``f30!}; Revelation
4:11\index{66 Revelation:ch.   4, v.  11, pg. ``f30!}.} and to enjoy him for ever.\footnote{\ Psalm 16:5-11\index{19
Psalms:ch.  16, v.   5-11, pg. ``f30!}; Psalm 144:15\index{19 Psalms:ch. 144, v.  15, pg. ``f30!}; Isaiah 12:2\index{23
Isaiah:ch.  12, v.   2, pg. ``f30!}; Luke 2:10\index{42 Luke:ch.   2, v.  10, pg. ``f30!}; Philippians 4:4\index{50
Philippians:ch.   4, v.   4, pg. ``f30!}; Revelation 21:3-4\index{66 Revelation:ch.  21, v.   3-4, pg. ``f30!}.}


\bigskip

Q. 2 \textit{What rule }has \textit{God given to}\index{Scripture:direction for how to glorify God ``f10!}\textit{
direct us how we may glorify}\index{Worship:glorifying God ``f10!}\textit{ and enjoy him?}

A. The Word of God\index{Scripture:word of God:word of God ``f10!}, which is contained in the
Scripture\index{Scripture:word of God:Holy Spirit speaking ``f10!}s of the Old and New Testament\index{Scripture:New
Testament ``f10!}s,\footnote{\ Matthew 19:4\index{40 Matthew:ch.  19, v.   4, pg. ``f30!}{}-5\index{40 Matthew:ch.  19,
v.   4-5, pg. ``f30!} with Genesis 2:24\index{01 Genesis:ch.   2, v.  24, pg. ``f30!}; Luke 24:27, 44\index{42 Luke:ch.
 24, v.  27, 44, pg. ``f30!}; 1 Corinthians 2:13\index{46 1 Corinthians:ch.   2, v.  13, pg. ``f30!}; 1 Corinthians
14:37\index{46 1 Corinthians:ch.  14, v.  37, pg. ``f30!}; 2 Peter 1:2\index{61 2 Peter:ch.   1, v.   2, pg.
``f30!}0-21\index{61 2 Peter:ch.   1, v.  20-21, pg. ``f30!}; 2 Peter 3:2, 15-16\index{61 2 Peter:ch.   3, v.   2,
15-16, pg. ``f30!}.} is the only rule to direct\index{Scripture:direction for how to glorify God ``f10!} us how we may
glorify\index{Worship:glorifying God ``f10!} and enjoy him.\footnote{\ Deuteronomy 4:2\index{05 Deuteronomy:ch.   4, v.
  2, pg. ``f30!}; Psalm 19:7\index{19 Psalms:ch.  19, v.   7, pg. ``f30!}{}-11\index{19 Psalms:ch.  19, v.   7-11, pg.
``f30!}; Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg. ``f30!}; John 15:11\index{43 John:ch.  15, v.  11, pg.
``f30!}; John 20:30-31\index{43 John:ch.  20, v.  30-31, pg. ``f30!}; Acts 17:11\index{44 Acts:ch.  17, v.  11, pg.
``f30!}; 2 Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  15-17, pg.
``f30!}; 1 John 1:4\index{62 1 John:ch.   1, v.   4, pg. ``f30!}.}


\bigskip

Q. 3 \textit{What do the Scriptures principally}\index{Scripture:principal teaching ``f10!}\textit{ teach?}

A. The Scriptures principally\index{Scripture:principal teaching ``f10!} teach what man is to believe concerning
God,\footnote{\ Genesis 1:1\index{01 Genesis:ch.   1, v.   1, pg. ``f30!}; John 5:39\index{43 John:ch.   5, v.  39, pg.
``f30!}; John 20:31\index{43 John:ch.  20, v.  31, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg.
``f30!}7\index{45 Romans:ch.  10, v.  17, pg. ``f30!}; 2 Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}.}
and what duty God\index{God's law:commandments:duties:required by God ``f10!} requires of man.\footnote{\ Deuteronomy
10:12-13\index{05 Deuteronomy:ch.  10, v.  12-13, pg. ``f30!}; Joshua 1:8\index{06 Joshua:ch.   1, v.   8, pg. ``f30!};
Psalm 119:105\index{19 Psalms:ch. 119, v. 105, pg. ``f30!}; Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; 2
Timothy 3:16\index{55 2 Timothy:ch.   3, v.  16, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  16-17, pg. ``f30!}.}


\bigskip

Q. 4 \textit{What is God?}

A. God is a\index{God's attributes:listed (definition of God) ``f10!} Spirit,\footnote{\ Deuteronomy 4:15-19\index{05
Deuteronomy:ch.   4, v.  15-19, pg. ``f30!}; Luke 24:39\index{42 Luke:ch.  24, v.  39, pg. ``f30!}; John 1:1\index{43
John:ch.   1, v.   1, pg. ``f30!}8\index{43 John:ch.   1, v.  18, pg. ``f30!}; John 4:24\index{43 John:ch.   4, v.  24,
pg. ``f30!}; Acts 17:29\index{44 Acts:ch.  17, v.  29, pg. ``f30!}.} infinite,\footnote{\ 1 Kings 8:27\index{11 1
Kings:ch.   8, v.  27, pg. ``f30!}; Psalm 139:7-10\index{19 Psalms:ch. 139, v.   7-10, pg. ``f30!}; Psalm
145:3\index{19 Psalms:ch. 145, v.   3, pg. ``f30!}; Psalm 147:5\index{19 Psalms:ch. 147, v.   5, pg. ``f30!}; Jeremiah
23:24\index{24 Jeremiah:ch.  23, v.  24, pg. ``f30!}; Romans 11:33\index{45 Romans:ch.  11, v.  33, pg.
``f30!}{}-36\index{45 Romans:ch.  11, v.  33-36, pg. ``f30!}.} eternal,\index{God's attributes:eternal
``f10!}\footnote{\ Deuteronomy 33:27\index{05 Deuteronomy:ch.  33, v.  27, pg. ``f30!}; Psalm 90:2\index{19 Psalms:ch. 
90, v.   2, pg. ``f30!}; Psalm 102:12, 24-27\index{19 Psalms:ch. 102, v.  12, 24-27, pg. ``f30!}; Revelation 1:4,
8\index{66 Revelation:ch.   1, v.   4, 8, pg. ``f30!}.} and unchangeable\index{God's attributes:unchangeable
``f10!},\footnote{\ Psalm 33:11\index{19 Psalms:ch.  33, v.  11, pg. ``f30!}; Malachi 3:6\index{39 Malachi:ch.   3, v. 
 6, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}2\index{58 Hebrews:ch.   1, v.  12, pg.
``f30!}; Hebrews 6:17\index{58 Hebrews:ch.   6, v.  17, pg. ``f30!}{}-18\index{58 Hebrews:ch.   6, v.  17-18, pg.
``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}; James 1:17\index{59 James:ch.   1, v.  17, pg.
``f30!}.} in his being,\footnote{\ Exodus 3:14\index{02 Exodus:ch.   3, v.  14, pg. ``f30!}; Psalm 115:2-3\index{19
Psalms:ch. 115, v.   2-3, pg. ``f30!}; 1 Timothy 1:17\index{54 1 Timothy:ch.   1, v.  17, pg. ``f30!}; 1 Timothy
6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}{}-16\index{54 1
Timothy:ch.   6, v.  15-16, pg. ``f30!}.} wisdom,\footnote{\ Psalm 104:24\index{19 Psalms:ch. 104, v.  24, pg. ``f30!};
Romans 11:33\index{45 Romans:ch.  11, v.  33, pg. ``f30!}{}-34\index{45 Romans:ch.  11, v.  33-34, pg. ``f30!}; Hebrews
4:13\index{58 Hebrews:ch.   4, v.  13, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg. ``f30!}0\index{62 1
John:ch.   3, v.  20, pg. ``f30!}.} power,\footnote{\ Genesis 17:1\index{01 Genesis:ch.  17, v.   1, pg. ``f30!}; Psalm
62:11\index{19 Psalms:ch.  62, v.  11, pg. ``f30!}; Jeremiah 32:17\index{24 Jeremiah:ch.  32, v.  17, pg. ``f30!};
Matthew 19:26\index{40 Matthew:ch.  19, v.  26, pg. ``f30!}; Revelation 1:8\index{66 Revelation:ch.   1, v.   8, pg.
``f30!}.} holiness,\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}3\index{58 Hebrews:ch.   1, v. 
13, pg. ``f30!}; 1 Peter 1:15-16\index{60 1 Peter:ch.   1, v.  15-16, pg. ``f30!}; 1 John 3:3, 5\index{62 1 John:ch.  
3, v.   3, 5, pg. ``f30!}; Revelation 15:4\index{66 Revelation:ch.  15, v.   4, pg. ``f30!}.}
justice,\footnote{\ Genesis 18:25\index{01 Genesis:ch.  18, v.  25, pg. ``f30!}; Exodus 34:6\index{02 Exodus:ch.  34,
v.   6, pg. ``f30!}{}-7\index{02 Exodus:ch.  34, v.   6-7, pg. ``f30!}; Deuteronomy 32:4\index{05 Deuteronomy:ch.  32,
v.   4, pg. ``f30!}; Psalm 96:13\index{19 Psalms:ch.  96, v.  13, pg. ``f30!}; Romans 3:5, 26\index{45 Romans:ch.   3,
v.   5, 26, pg. ``f30!}.} goodness\index{God's attributes:goodness ``f10!},\footnote{\ Psalm 103:5\index{19 Psalms:ch.
103, v.   5, pg. ``f30!}; Psalm 107:8\index{19 Psalms:ch. 107, v.   8, pg. ``f30!}; Matthew 19:17\index{40 Matthew:ch. 
19, v.  17, pg. ``f30!}; Romans 2:4\index{45 Romans:ch.   2, v.   4, pg. ``f30!}.} and truth\index{God's
attributes:truth ``f10!}.\footnote{\ Exodus 34:6\index{02 Exodus:ch.  34, v.   6, pg. ``f30!}; Deuteronomy
32:4\index{05 Deuteronomy:ch.  32, v.   4, pg. ``f30!}; Psalm 86:15\index{19 Psalms:ch.  86, v.  15, pg. ``f30!}; Psalm
117:2\index{19 Psalms:ch. 117, v.   2, pg. ``f30!}; Hebrews 6:18\index{58 Hebrews:ch.   6, v.  18, pg. ``f30!}.}


\bigskip

Q. 5 \textit{Are }there\textit{ }more\textit{ Gods than one?}

A. There is but one only,\footnote{\ Deuteronomy 6:4\index{05 Deuteronomy:ch.   6, v.   4, pg. ``f30!}; Isaiah
44:6\index{23 Isaiah:ch.  44, v.   6, pg. ``f30!}; Isaiah 45:21-22\index{23 Isaiah:ch.  45, v.  21-22, pg. ``f30!}; 1
Corinthians 8:4-6\index{46 1 Corinthians:ch.   8, v.   4-6, pg. ``f30!}.} the living and true God .\footnote{\ Jeremiah
10:10\index{24 Jeremiah:ch.  10, v.  10, pg. ``f30!}; John 17:3\index{43 John:ch.  17, v.   3, pg. ``f30!}; 1
Thessalonians 1:9\index{52 1 Thessalonians:ch.   1, v.   9, pg. ``f30!}; 1 John 5:20\index{62 1 John:ch.   5, v.  20,
pg. ``f30!}.}


\bigskip


\bigskip

Q. 6 \textit{How many persons}\index{God's attributes:trinity:three persons ``f10!}\textit{ are there in the
Godhead}\index{God's attributes:Godhead ``f10!}\textit{?}

A. There are three persons\index{God's attributes:trinity:three persons ``f10!} in the Godhead\index{God's
attributes:Godhead ``f10!}; the Father\index{Father:God the ``f10!}, the Son\index{Jesus:Son of God ``f10!}, and the
Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!};\footnote{\ Matthew 3:16-17\index{40 Matthew:ch.   3, v.  16-17, pg.
``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 2 Corinthians 13:1\index{47 2 Corinthians:ch. 
13, v.   1, pg. ``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}; 1 Peter 1:2\index{60 1 Peter:ch.   1, v. 
 2, pg. ``f30!}.} and these three are one God, the same in substance\index{God's attributes:trinity:one substance
``f10!}, equal in power and glory\index{God's attributes:trinity:persons equal ``f10!}.\footnote{\ Psalm 45:6\index{19
Psalms:ch.  45, v.   6, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}; John 17:5\index{43 John:ch. 
17, v.   5, pg. ``f30!}; Acts 5:3\index{44 Acts:ch.   5, v.   3, pg. ``f30!}{}-4\index{44 Acts:ch.   5, v.   3-4, pg.
``f30!}; Romans 9:5\index{45 Romans:ch.   9, v.   5, pg. ``f30!}; Colossians 2:9\index{51 Colossians:ch.   2, v.   9,
pg. ``f30!}; \textrm{sundry judicial laws} \par Jude 24\index{65 Jude:v.  24, pg. ``f30!}{}-25\index{65 Jude:v. 24-25,
pg. ``f30!}.}


\bigskip

Q. 7 \textit{What }are\textit{ the decrees}\index{Predestination:eternal decrees ``f10!}\textit{ of God?}

A. The decrees\index{Predestination:eternal decrees ``f10!} of God are, his eternal purpose\index{Predestination:eternal
purpose ``f10!}, according to the counsel of his\index{Causation:God's own will ``f10!} will, whereby, for his own
glory\index{God's attributes:glory:his own ``f10!}, he has foreordained\index{Predestination:fore-ordination ``f10!}
whatsoever comes to pass.\footnote{\ Psalm 33:11\index{19 Psalms:ch.  33, v.  11, pg. ``f30!}; Isaiah 14:24\index{23
Isaiah:ch.  14, v.  24, pg. ``f30!}; Acts 2:23\index{44 Acts:ch.   2, v.  23, pg. ``f30!}; Ephesians 1:1\index{49
Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}{}-12\index{49 Ephesians:ch.  
1, v.  11-12, pg. ``f30!}.}


\bigskip

Q. 8 \textit{How does God execute his decrees}\index{Predestination:eternal decrees ``f10!}\textit{?}

A. God executes his decrees\index{Predestination:eternal decrees ``f10!} in the works of creation\index{God's
attributes:creator:works of ``f10!} and providence\index{Providence:providence ``f10!}.\footnote{\ Psalm 148:8\index{19
Psalms:ch. 148, v.   8, pg. ``f30!}; Isaiah 40:26\index{23 Isaiah:ch.  40, v.  26, pg. ``f30!}; Daniel 4:35\index{27
Daniel:ch.   4, v.  35, pg. ``f30!}; Acts 4:24-28\index{44 Acts:ch.   4, v.  24-28, pg. ``f30!}; Revelation
4:11\index{66 Revelation:ch.   4, v.  11, pg. ``f30!}.}


\bigskip

Q. 9 What\textit{ is the work of creation}\index{God's attributes:creator:works of ``f10!}\textit{?}

A. The work of creation\index{God's attributes:creator:works of ``f10!} is, God's making all things of nothing, by the
word of his power\index{Creation:by word of God's power ``f10!},\footnote{\ Genesis 1:1\index{01 Genesis:ch.   1, v.  
1, pg. ``f30!}; Psalm 33:6, 9\index{19 Psalms:ch.  33, v.   6, 9, pg. ``f30!}; Hebrews 11:3\index{58 Hebrews:ch.  11,
v.   3, pg. ``f30!}.} in the space of six\index{Creation:in six days ``f10!} days, and all very
good.\footnote{\ Genesis 1:31\index{01 Genesis:ch.   1, v.  31, pg. ``f30!}.}


\bigskip

Q. 10 \textit{How did God create man}\index{Mankind:created:by God ``f10!}\textit{?}

A. God created man\index{Mankind:created:by God ``f10!} male\index{Mankind:male and female ``f10!} and
female\index{Mankind:created:male and female ``f10!}, after his own image\index{Mankind:image of God ``f10!}
[a],\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}7\index{01 Genesis:ch.   1, v.  27, pg.
``f30!}.} in knowledge,\footnote{\ Colossians 3:10\index{51 Colossians:ch.   3, v.  10, pg. ``f30!}.} righteousness,
and holiness,\index{Mankind:created:holy ``f10!}\footnote{\ Ephesians 4:24\index{49 Ephesians:ch.   4, v.  24, pg.
``f30!}.} with dominion over the\index{Mankind:dominion over creation ``f10!} creatures\index{creatures
``f10!}.\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}8\index{01 Genesis:ch.   1, v.  28, pg.
``f30!}; Psalm 8.}


\bigskip

Q. 11 \textit{What }are\textit{ God's works}\index{Providence:God's works ``f10!}\textit{ of
providence}\index{Providence:providence ``f10!}\textit{?} 

A. God's works of providence\index{Providence:providence ``f10!} are, his most holy\index{God's attributes:holy:most
``f10!},\footnote{\ Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg. ``f30!}.} wise,\footnote{\ Psalm 104:24\index{19
Psalms:ch. 104, v.  24, pg. ``f30!}.} and powerful\footnote{\ Hebrews 1:3\index{58 Hebrews:ch.   1, v.   3, pg.
``f30!}.} preserving\footnote{\ Nehemiah 9:6\index{16 Nehemiah:ch.   9, v.   6, pg. ``f30!}.} and
governing\index{Providence:governing all ``f10!}\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg.
``f30!}9\index{49 Ephesians:ch.   1, v.  19, pg. ``f30!}{}-22\index{49 Ephesians:ch.   1, v.  19-22, pg. ``f30!}.} all
his creatures\index{creatures ``f10!}, and all their actions.\footnote{\ Psalm 36:6\index{19 Psalms:ch.  36, v.   6,
pg. ``f30!}; Proverbs 16:33\index{20 Proverbs:ch.  16, v.  33, pg. ``f30!}; Matthew 10:30\index{40 Matthew:ch.  10, v. 
30, pg. ``f30!}.}


\bigskip

Q. 12 \textit{What special act of providence}\index{Providence:providence ``f10!}\textit{ did God exercise toward man in
the estate wherein}\index{Mankind:created:state of innocence ``f10!}\textit{ he was
created}\index{Mankind:created:state ``f10!}\textit{?}

A. When God had created man\index{Mankind:created:by God ``f10!}, he entered into a covenant of life\index{Covenant:life
``f10!} with him, upon condition\index{Covenant:works:conditional on perfect obedience ``f10!} of perfect
obedience\index{Jesus:obedience:to God's law ``f10!}; forbidding\index{God's law:commandments:given to Adam ``f10!} him
to eat of the tree of the\index{Tree:of the knowledge of good and evil ``f10!} knowledge of good and evil, upon pain of
death\index{Death:punishment for sin ``f10!}.\footnote{\ Genesis 2:16-17\index{01 Genesis:ch.   2, v.  16-17, pg.
``f30!}; James 2:10\index{59 James:ch.   2, v.  10, pg. ``f30!}.}


\bigskip

Q. 13 \textit{Did our first parents}\index{Parents:first, sinned against God ``f10!}\textit{ continue in the estate
wherein}\index{Mankind:created:state of innocence ``f10!}\textit{ they were created}\index{Mankind:created:state
``f10!}\textit{?}

A. Our first parents\index{Parents:first, sinned against God ``f10!}, being left to the freedom of\index{Causation:will
of creatures:free ``f10!} their own will, fell\index{Sin:original:all mankind descending from Adam ``f10!} from the
estate wherein they were created\index{Mankind:created:state ``f10!}, by sinning against God.\footnote{\ Genesis
3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}{}-8\index{01 Genesis:ch.   3, v.   6-8, pg. ``f30!}, 13\index{01
Genesis:ch.   3, v.   6-8, 13, pg. ``f30!}; 2 Corinthians 11:3\index{47 2 Corinthians:ch.  11, v.   3, pg. ``f30!}.}


\bigskip

Q. 14 \textit{What is sin?}

A. Sin is\index{Sin:defined ``f10!} any want of conformity unto, or transgression of\index{God's
law:commandments:transgressions ``f10!}, the law of God.\footnote{\ Leviticus 5:1\index{03 Leviticus:ch.   5, v.   1,
pg. ``f30!}7\index{03 Leviticus:ch.   5, v.  17, pg. ``f30!}; James 4:17\index{59 James:ch.   4, v.  17, pg. ``f30!}; 1
John 3:4\index{62 1 John:ch.   3, v.   4, pg. ``f30!}.}


\bigskip

Q. 15 \textit{What was the sin whereby our first parents}\index{Parents:first, sinned against God ``f10!}\textit{
fell}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{ from the estate wherein they were
created}\index{Mankind:created:state ``f10!}\textit{?}

A. The sin whereby our first parents\index{Parents:first, sinned against God ``f10!} fell\index{Sin:original:all mankind
descending from Adam ``f10!} from the estate wherein thy were created\index{Mankind:created:state ``f10!}, was their
eating the forbidden\index{Sin:original:first transgression (of Adam) ``f10!} fruit.\footnote{\ Genesis 3:6\index{01
Genesis:ch.   3, v.   6, pg. ``f30!}.}


\bigskip

Q. 16 \textit{Did all mankind fall}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{ in Adam's first
transgression}\index{Sin:original:first transgression (of Adam) ``f10!}\textit{?}

A. The covenant being\index{Covenant:first:made with Adam ``f10!} made with Adam\index{Adam:covenant made with him
``f10!},\footnote{\ Genesis 2:16-17\index{01 Genesis:ch.   2, v.  16-17, pg. ``f30!}; James 2:10\index{59 James:ch.  
2, v.  10, pg. ``f30!}.} not only for himself, but for his posterity\index{Sin:original:all mankind descending from
Adam ``f10!}; all mankind, descending from him by ordinary\index{Sin:original:by ordinary generation ``f10!}
generation, sinned in him, and fell with him, in his first transgression\index{Sin:original:first transgression (of
Adam) ``f10!}.\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg.
``f30!}{}-21\index{45 Romans:ch.   5, v.  12-21, pg. ``f30!}; 1 Corinthians 15:22\index{46 1 Corinthians:ch.  15, v. 
22, pg. ``f30!}.}


\bigskip

Q. 17 \textit{Into what }estate\index{Sin:original:an estate of sin and misery ``f10!}\textit{ did the
fall}\index{Sin:original:first transgression (of Adam) ``f10!}\textit{ bring mankind}\index{Sin:original:all mankind
descending from Adam ``f10!}\textit{?}

A. The fall\index{Sin:original:first transgression (of Adam) ``f10!} brought mankind\index{Sin:original:all mankind
descending from Adam ``f10!} into an estate of sin\index{Sin:original:an estate of sin and misery ``f10!} and
misery\index{Punishment:misery ``f10!}.\footnote{\ Genesis 3:16-19\index{01 Genesis:ch.   3, v.  16-19, pg. ``f30!},
23\index{01 Genesis:ch.   3, v.  16-19, 23, pg. ``f30!}; Romans 3:16\index{45 Romans:ch.   3, v.  16, pg. ``f30!};
Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}; Ephesians
2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}.}


\bigskip

Q. 18 \textit{Wherein consists the sinfulness of that estate whereinto}\index{Sin:original:an estate of sin and misery
``f10!}\textit{ man fell}\index{Sin:original:all mankind descending from Adam ``f10!}\textit{?}

A. The sinfulness of that estate whereinto\index{Sin:original:an estate of sin and misery ``f10!} man
fell\index{Sin:original:all mankind descending from Adam ``f10!}, consists in the guilt\index{guilt ``f10!} of
Adam\index{Adam:guilt of first sin on all mankind ``f10!}{}'s first sin\index{Sin:original:first transgression (of
Adam) ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg.
``f30!}, 19\index{45 Romans:ch.   5, v.  12, 19, pg. ``f30!}.} the want of original
righteousness,\index{Sin:original:original righteousness corrupted ``f10!}\footnote{\ Romans 3:10\index{45 Romans:ch.  
3, v.  10, pg. ``f30!}; Colossians 3:10\index{51 Colossians:ch.   3, v.  10, pg. ``f30!}; Ephesians 4:24\index{49
Ephesians:ch.   4, v.  24, pg. ``f30!}.} and the corruption of his\index{Sin:original:corrupt nature:whole ``f10!}
whole nature,\footnote{\ Psalm 51:5\index{19 Psalms:ch.  51, v.   5, pg. ``f30!}; John 3:6\index{43 John:ch.   3, v.  
6, pg. ``f30!}; Romans 3:18\index{45 Romans:ch.   3, v.  18, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7,
pg. ``f30!}{}-8\index{45 Romans:ch.   8, v.   7-8, pg. ``f30!}; Ephesians 2:3\index{49 Ephesians:ch.   2, v.   3, pg.
``f30!}.} which is commonly called original sin; together with all actual transgressions\index{God's
law:commandments:transgressions ``f10!} which proceed from it\index{Sin:original:proceeds all transgressions
``f10!}.\footnote{\ Genesis 6:5\index{01 Genesis:ch.   6, v.   5, pg. ``f30!}; Psalm 53:1-3\index{19 Psalms:ch.  53, v.
  1-3, pg. ``f30!}; Matthew 15:19\index{40 Matthew:ch.  15, v.  19, pg. ``f30!}; Romans 3:10\index{45 Romans:ch.   3,
v.  10, pg. ``f30!}{}-18, 23\index{45 Romans:ch.   3, v.  10-18, 23, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch. 
 5, v.   1, pg. ``f30!}9\index{48 Galatians:ch.   5, v.  19, pg. ``f30!}{}-21\index{48 Galatians:ch.   5, v.  19-21,
pg. ``f30!}; James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}{}-15\index{59 James:ch.   1, v.  14-15, pg.
``f30!}.}


\bigskip

Q. 19 \textit{What is the misery}\index{Punishment:misery ``f10!}\textit{ of that estate
whereinto}\index{Sin:original:an estate of sin and misery ``f10!}\textit{ man fell}\index{Sin:original:all mankind
descending from Adam ``f10!}\textit{?}

A. All mankind by\index{Sin:original:all mankind descending from Adam ``f10!} their fall lost\index{Communion:with God,
lost ``f10!} communion with God\index{Communion:with God ``f10!},\footnote{\ Genesis 3:8\index{01 Genesis:ch.   3, v.  
8, pg. ``f30!}, 24\index{01 Genesis:ch.   3, v.   8, 24, pg. ``f30!}; John 8:34\index{43 John:ch.   8, v.  34, pg.
``f30!}, 42, 44\index{43 John:ch.   8, v.  34, 42, 44, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1,
pg. ``f30!}2\index{49 Ephesians:ch.   2, v.  12, pg. ``f30!}; Ephesians 4:18\index{49 Ephesians:ch.   4, v.  18, pg.
``f30!}.} are under his wrath\footnote{\ John 3:3\index{43 John:ch.   3, v.   3, pg. ``f30!}6\index{43 John:ch.   3, v.
 36, pg. ``f30!}; Romans 1:18\index{45 Romans:ch.   1, v.  18, pg. ``f30!}; Ephesians 2:3\index{49 Ephesians:ch.   2,
v.   3, pg. ``f30!}; Ephesians 5:6\index{49 Ephesians:ch.   5, v.   6, pg. ``f30!}.} and curse,\index{Curse:of God
``f10!}\footnote{\ Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}; Revelation 22:3\index{66
Revelation:ch.  22, v.   3, pg. ``f30!}.} and so made liable\index{Sin:original:liable to all miseries ``f10!} to all
miseries\index{Punishment:misery ``f10!} in this life,\footnote{\ Genesis 3:16-19\index{01 Genesis:ch.   3, v.  16-19,
pg. ``f30!}; Job 5:7\index{18 Job:ch.   5, v.   7, pg. ``f30!}; Ecclesiastes 2:22-23\index{21 Ecclesiastes:ch.   2, v. 
22-23, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}8\index{45 Romans:ch.   8, v.  18, pg.
``f30!}{}-23\index{45 Romans:ch.   8, v.  18-23, pg. ``f30!}.} to death\index{Death:punishment for sin
``f10!}\footnote{\ Ezekiel 18:4\index{26 Ezekiel:ch.  18, v.   4, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.  
1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}; Romans 6:23\index{45 Romans:ch.   6, v.  23, pg.
``f30!}.} itself, and to the pains\index{Hell:pains ``f10!} of hell\index{Hell:hell ``f10!} for
ever.\footnote{\ Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}, 46\index{40 Matthew:ch.  25, v.  41, 46,
pg. ``f30!}; 2 Thessalonians 1:9\index{53 2 Thessalonians:ch.   1, v.   9, pg. ``f30!}; Revelation 14:9-11\index{66
Revelation:ch.  14, v.   9-11, pg. ``f30!}.}


\bigskip

Q. 20 \textit{Did God leave all mankind}\index{Mankind:not all left to perish in sin and misery ``f10!}\textit{ to
perish}\index{Redemption:not all perish ``f10!}\textit{ in the estate of sin}\index{Sin:original:an estate of sin and
misery ``f10!}\textit{ and misery}\index{Punishment:misery ``f10!}\textit{?}

A. God having, out of his mere good pleasure\index{Providence:at God's pleasure ``f10!}, from all eternity\index{God's
attributes:eternal ``f10!}, elected some to\index{Election:chosen by God ``f10!} everlasting life\index{everlasting
life ``f10!},\footnote{\ Acts 13:48\index{44 Acts:ch.  13, v.  48, pg. ``f30!}; Ephesians 1:4\index{49 Ephesians:ch.  
1, v.   4, pg. ``f30!}{}-5\index{49 Ephesians:ch.   1, v.   4-5, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2
Thessalonians:ch.   2, v.  13, pg. ``f30!}{}-14\index{53 2 Thessalonians:ch.   2, v.  13-14, pg. ``f30!}.} did enter
into a covenant of grace\index{Covenant:grace:grace ``f10!}, to deliver them\index{Salvation:delivered from thraldom to
sin ``f10!} out of the estate\index{Sin:original:an estate of sin and misery ``f10!} of sin and
misery\index{Punishment:misery ``f10!}, and to bring them into an estate of salvation\index{Salvation:belief in Jesus
by faith ``f10!} by a Redeemer\index{Jesus:redeemer of God's elect ``f10!}.\footnote{\ Genesis 3:15\index{01
Genesis:ch.   3, v.  15, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}; Exodus 19:5\index{02
Exodus:ch.  19, v.   5, pg. ``f30!}{}-6\index{02 Exodus:ch.  19, v.   5-6, pg. ``f30!}; Jeremiah 31:3\index{24
Jeremiah:ch.  31, v.   3, pg. ``f30!}1-34\index{24 Jeremiah:ch.  31, v.  31-34, pg. ``f30!}; Matthew 20:28\index{40
Matthew:ch.  20, v.  28, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46
1 Corinthians:ch.  11, v.  25, pg. ``f30!}; Hebrews 9:15\index{58 Hebrews:ch.   9, v.  15, pg. ``f30!}.}


\bigskip

Q. 21 \textit{Who }is\textit{ the Redeemer}\index{Jesus:redeemer of God's elect ``f10!}\textit{ of God's
elect}\index{Election:elect of God ``f10!}\textit{?}

A. The only Redeemer\index{Jesus:redeemer of God's elect ``f10!} of God's elect\index{Election:elect of God ``f10!} is
the Lord Jesus Christ,\footnote{\ John 14:6\index{43 John:ch.  14, v.   6, pg. ``f30!}; Acts 4:12\index{44 Acts:ch.  
4, v.  12, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}{}-6\index{54 1 Timothy:ch.   2,
v.   5-6, pg. ``f30!}.} who, being the eternal\index{Jesus:eternal Son ``f10!} Son\index{Jesus:Son of God ``f10!} of
God,\footnote{\ Psalm 2:7\index{19 Psalms:ch.   2, v.   7, pg. ``f30!}; Matthew 3:17\index{40 Matthew:ch.   3, v.  17,
pg. ``f30!}; Matthew 17:5\index{40 Matthew:ch.  17, v.   5, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}8\index{43 John:ch.   1, v.  18, pg. ``f30!}.} became man,\footnote{\ Isaiah 9:6\index{23 Isaiah:ch.   9, v.  
6, pg. ``f30!}; Matthew 1:23\index{40 Matthew:ch.   1, v.  23, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg.
``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.}
and so was, and continues\index{Jesus:God-man ``f10!} to be, God and man in two distinct natures\index{Jesus:two
distinct natures ``f10!}, and one person, forever.\footnote{\ Acts 1:11\index{44 Acts:ch.   1, v.  11, pg. ``f30!};
Hebrews 7:24-25\index{58 Hebrews:ch.   7, v.  24-25, pg. ``f30!}.}


\bigskip

Q. 22 \textit{How did Christ, being the Son}\index{Jesus:Son of God ``f10!}\textit{ of God, become man?}

A. Christ, the Son\index{Jesus:Son of God ``f10!} of God, became man, by taking to himself a true body, and a
reasonable\index{Jesus:human nature:reasonable soul ``f10!} soul,\footnote{\ Philippians 2:7\index{50 Philippians:ch.  
2, v.   7, pg. ``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   2, v.  14,
pg. ``f30!}, 17\index{58 Hebrews:ch.   2, v.  14, 17, pg. ``f30!}.} being conceived by\index{Jesus:conceived:by the
power of the Holy Spirit ``f10!} the power of the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, in the
womb\index{Jesus:conceived:in womb of virgin Mary ``f10!} of the virgin\index{Jesus:born of virgin Mary ``f10!}
Mary\index{Mary, mother of Jesus ``f10!}, and born of her,\footnote{\ Luke 1:27, 31, 35\index{42 Luke:ch.   1, v.  27,
31, 35, pg. ``f30!}.} yet without sin.\footnote{\ 2 Corinthians 5:21\index{47 2 Corinthians:ch.   5, v.  21, pg.
``f30!}; Hebrews 4:15\index{58 Hebrews:ch.   4, v.  15, pg. ``f30!}; Hebrews 7:26\index{58 Hebrews:ch.   7, v.  26, pg.
``f30!}; 1 John 3:5\index{62 1 John:ch.   3, v.   5, pg. ``f30!}.}


\bigskip

Q. 23 \textit{What offices}\index{Jesus:offices ``f10!}\textit{ does Christ execute as our
Redeemer}\index{Jesus:redeemer of God's elect ``f10!}\textit{?}

A. Christ, as our Redeemer\index{Jesus:redeemer of God's elect ``f10!}, executes the office\index{Jesus:offices ``f10!}s
of a prophet\index{Jesus:prophet:office ``f10!},\footnote{\ Deuteronomy 18:18\index{05 Deuteronomy:ch.  18, v.  18, pg.
``f30!}; Acts 2:33\index{44 Acts:ch.   2, v.  33, pg. ``f30!}; Acts 3:22\index{44 Acts:ch.   3, v.  22, pg.
``f30!}{}-23\index{44 Acts:ch.   3, v.  22-23, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg.
``f30!}{}-2\index{58 Hebrews:ch.   1, v.   1-2, pg. ``f30!}.} of a priest,\footnote{\ Hebrews 4:14-15\index{58
Hebrews:ch.   4, v.  14-15, pg. ``f30!}; Hebrews 5:5-6\index{58 Hebrews:ch.   5, v.   5-6, pg. ``f30!}.} and of a
king\index{Jesus:King ``f10!},\footnote{\ Isaiah 9:6\index{23 Isaiah:ch.   9, v.   6, pg. ``f30!}{}-7\index{23
Isaiah:ch.   9, v.   6-7, pg. ``f30!}; Luke 1:32-33\index{42 Luke:ch.   1, v.  32-33, pg. ``f30!}; John 18:37\index{43
John:ch.  18, v.  37, pg. ``f30!}; 1 Corinthians 15:25\index{46 1 Corinthians:ch.  15, v.  25, pg. ``f30!}.} both in
his estate of humiliation and\index{Jesus:humiliation:aspects of ``f10!} exaltation\index{Jesus:exaltation ``f10!}.


\bigskip

Q. 24 \textit{How does Christ execute the office}\index{Jesus:offices ``f10!}\textit{ of a
prophet}\index{Jesus:prophet:office ``f10!}\textit{?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a prophet\index{Jesus:prophet:office ``f10!}, in revealing
to us\index{Revelation:God's word ``f10!}, by his Word\footnote{\ Luke 4:18-19, 21\index{42 Luke:ch.   4, v.  18-19,
21, pg. ``f30!}; Acts 1:1-2\index{44 Acts:ch.   1, v.   1-2, pg. ``f30!}; Hebrews 2:3\index{58 Hebrews:ch.   2, v.   3,
pg. ``f30!}.} and Spirit,\footnote{\ John 15:26\index{43 John:ch.  15, v.  26, pg. ``f30!}{}-27\index{43 John:ch.  15,
v.  26-27, pg. ``f30!}; Acts 1:8\index{44 Acts:ch.   1, v.   8, pg. ``f30!}; 1 Peter 1:11\index{60 1 Peter:ch.   1, v. 
11, pg. ``f30!}.} the will of God for our salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ John
4:41-42\index{43 John:ch.   4, v.  41-42, pg. ``f30!}; John 20:30-31\index{43 John:ch.  20, v.  30-31, pg. ``f30!}.}


\bigskip

Q. 25 \textit{How does Christ }execute\textit{ the office}\index{Jesus:offices ``f10!}\textit{ of a priest?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a priest, in his\index{Jesus:priest ``f10!} once offering
up of himself a sacrifice to\index{Jesus:sacrifice:of himself ``f10!} satisfy divine justice\index{God's
attributes:administers justice ``f10!},\footnote{\ Isaiah 53; Acts 8:32-35\index{44 Acts:ch.   8, v.  32-35, pg.
``f30!}; Hebrews 9:26-28\index{58 Hebrews:ch.   9, v.  26-28, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.  
1, pg. ``f30!}2\index{58 Hebrews:ch.  10, v.  12, pg. ``f30!}.} and reconcile\index{Jesus:purchased reconciliation
``f10!} us to God;\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}0-11\index{45 Romans:ch.   5, v. 
10-11, pg. ``f30!}; 2 Corinthians 5:18\index{47 2 Corinthians:ch.   5, v.  18, pg. ``f30!}; Colossians 1:21\index{51
Colossians:ch.   1, v.  21, pg. ``f30!}{}-22\index{51 Colossians:ch.   1, v.  21-22, pg. ``f30!}.} and in making
continual intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} for us.\footnote{\ Romans
8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}; Hebrews 7:25\index{58
Hebrews:ch.   7, v.  25, pg. ``f30!}; Hebrews 9:24\index{58 Hebrews:ch.   9, v.  24, pg. ``f30!}.}


\bigskip

Q. 26 \textit{How does Christ }execute\textit{ the office}\index{Jesus:offices ``f10!}\textit{ of a
king}\index{Jesus:King ``f10!}\textit{?}

A. Christ executes the office\index{Jesus:offices ``f10!} of a king\index{Jesus:King ``f10!}, in subduing
us\index{Jesus:subdues will of man ``f10!} to himself, in ruling\index{Jesus:ruling and defending his people ``f10!}
and defending us\index{Jesus:defends his Church ``f10!},\footnote{\ Psalm 110:3\index{19 Psalms:ch. 110, v.   3, pg.
``f30!}; Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  18-20, pg.
``f30!}; John 17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}; Colossians 1:13\index{51 Colossians:ch.   1, v.  13, pg.
``f30!}.} and in restraining\index{Enemies:of believers restrained ``f10!} and conquering\index{Jesus:conqueror of his
and our enemies ``f10!} all his and our enemies\index{Enemies:believers delivered from ``f10!}.\footnote{\ Psalm
2:6\index{19 Psalms:ch.   2, v.   6, pg. ``f30!}{}-9\index{19 Psalms:ch.   2, v.   6-9, pg. ``f30!}; Psalm
110:1\index{19 Psalms:ch. 110, v.   1, pg. ``f30!}{}-2\index{19 Psalms:ch. 110, v.   1-2, pg. ``f30!}; Matthew
12:28\index{40 Matthew:ch.  12, v.  28, pg. ``f30!}; 1 Corinthians 15:24-26\index{46 1 Corinthians:ch.  15, v.  24-26,
pg. ``f30!}; Colossians 2:15\index{51 Colossians:ch.   2, v.  15, pg. ``f30!}.}


\bigskip

Q. 27 \textit{Wherein did Christ's humiliation consist}\index{Jesus:humiliation:aspects of ``f10!}\textit{?}

A. Christ's humiliation consist\index{Jesus:humiliation:aspects of ``f10!}ed in his being born, and\index{Jesus:born of
virgin Mary ``f10!} that in a low condition\index{Jesus:humiliation:low condition ``f10!},\footnote{\ Luke 2:7\index{42
Luke:ch.   2, v.   7, pg. ``f30!}; 2 Corinthians 8:9\index{47 2 Corinthians:ch.   8, v.   9, pg. ``f30!}; Galatians
4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} made under the law,\footnote{\ Galatians 4:4\index{48
Galatians:ch.   4, v.   4, pg. ``f30!}.} undergoing\index{Jesus:humiliation:undergoing miseries of this life ``f10!}
the miseries\index{Punishment:misery ``f10!} of this life,\footnote{\ Isaiah 53:3\index{23 Isaiah:ch.  53, v.   3, pg.
``f30!}; Luke 9:58\index{42 Luke:ch.   9, v.  58, pg. ``f30!}; John 4:6\index{43 John:ch.   4, v.   6, pg. ``f30!};
John 11:35\index{43 John:ch.  11, v.  35, pg. ``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg.
``f30!}8\index{58 Hebrews:ch.   2, v.  18, pg. ``f30!}.} the wrath of God\index{Wrath:of God:for sin
``f10!},\footnote{\ Psalm 22:1\index{19 Psalms:ch.  22, v.   1, pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v. 
 4, pg. ``f30!}6\index{40 Matthew:ch.  27, v.  46, pg. ``f30!}; Isaiah 53:10\index{23 Isaiah:ch.  53, v.  10, pg.
``f30!}; 1 John 2:2\index{62 1 John:ch.   2, v.   2, pg. ``f30!}.} and the cursed\index{Jesus:death:cursed ``f10!}
death of the cross\index{Jesus:death:on cross ``f10!};\footnote{\ Galatians 3:13\index{48 Galatians:ch.   3, v.  13,
pg. ``f30!}; Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg. ``f30!}.} in being buried\index{Jesus:buried
``f10!}, and continuing\index{Jesus:death:continuing under power for time ``f10!} under the power of
death\index{Jesus:death:under power of ``f10!} for a time.\footnote{\ Matthew 12:40\index{40 Matthew:ch.  12, v.  40,
pg. ``f30!}; 1 Corinthians 15:3-4\index{46 1 Corinthians:ch.  15, v.   3-4, pg. ``f30!}.}


\bigskip

Q. 28 \textit{Wherein consists Christ's exaltation}\index{Jesus:exaltation ``f10!}\textit{?}

A. Christ's exaltation\index{Jesus:exaltation ``f10!} consists in his rising\index{Jesus:resurrection:from the dead
``f10!} again from the dead on the third day\index{Jesus:resurrection:third day ``f10!},\footnote{\ 1 Corinthians
15:4\index{46 1 Corinthians:ch.  15, v.   4, pg. ``f30!}.} in ascending\index{Jesus:ascension:into heaven ``f10!} up
into heaven,\footnote{\ Psalm 68:18\index{19 Psalms:ch.  68, v.  18, pg. ``f30!}; Acts 1:11\index{44 Acts:ch.   1, v. 
11, pg. ``f30!}; Ephesians 4:8\index{49 Ephesians:ch.   4, v.   8, pg. ``f30!}.} in sitting\index{Jesus:in
heaven:sitting at right hand of Father ``f10!} at the right hand\footnote{\ Psalm 110:1\index{19 Psalms:ch. 110, v.  
1, pg. ``f30!}; Acts 2:33\index{44 Acts:ch.   2, v.  33, pg. ``f30!}{}-34\index{44 Acts:ch.   2, v.  33-34, pg.
``f30!}; Hebrews 1:3\index{58 Hebrews:ch.   1, v.   3, pg. ``f30!}.} of God the Father\index{Father:God the ``f10!},
and in coming to judge the\index{Jesus:judge:world ``f10!} world at the last day\index{Last day:last day
``f10!}.\footnote{\ Matthew 16:27\index{40 Matthew:ch.  16, v.  27, pg. ``f30!}; Acts 17:31\index{44 Acts:ch.  17, v. 
31, pg. ``f30!}.}


\bigskip


\bigskip

Q. 29 \textit{How are }we\textit{ made partakers}\index{Redemption:partakers through Christ ``f10!}\textit{ of the
redemption purchased}\index{Redemption:purchased by Christ ``f10!}\textit{ by Christ?}

A. We are made partakers\index{Redemption:partakers through Christ ``f10!} of the redemption
purchased\index{Redemption:purchased by Christ ``f10!} by Christ, by the effectual
application\index{Redemption:effectually:applied ``f10!} of it to us by his Holy Spirit\index{Holy Spirit:Holy Spirit
``f10!}.\footnote{\ Titus 3:4-7\index{56 Titus:ch.   3, v.   4-7, pg. ``f30!}.}


\bigskip

Q. 30 \textit{How does the Spirit apply}\index{Holy Spirit:applies:redemption ``f10!}\textit{ to us the redemption
purchased}\index{Redemption:purchased by Christ ``f10!}\textit{ by Christ?}

A. The Spirit\index{Holy Spirit:Holy Spirit ``f10!} applies\index{Faith:instrument:which applies Christ to believers
``f10!} to us the redemption purchased\index{Redemption:purchased by Christ ``f10!} by Christ, by
working\index{Faith:worked in believers by the Holy Spirit ``f10!} faith in us,\footnote{\ Romans 10:1\index{45
Romans:ch.  10, v.   1, pg. ``f30!}7\index{45 Romans:ch.  10, v.  17, pg. ``f30!}; 1 Corinthians 2:12\index{46 1
Corinthians:ch.   2, v.  12, pg. ``f30!}{}-16\index{46 1 Corinthians:ch.   2, v.  12-16, pg. ``f30!}; Ephesians
2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}; Philippians 1:29\index{50 Philippians:ch.   1, v.  29, pg.
``f30!}.} and thereby uniting\index{Believers:union with Christ:mystical ``f10!} us to Christ in our effectual
calling\index{Effectual call:calling ``f10!}.\footnote{\ John 15:5\index{43 John:ch.  15, v.   5, pg. ``f30!}; 1
Corinthians 1:9\index{46 1 Corinthians:ch.   1, v.   9, pg. ``f30!}; Ephesians 3:17\index{49 Ephesians:ch.   3, v.  17,
pg. ``f30!}.}


\bigskip

Q. 31 \textit{What is effectual calling}\index{Effectual call:calling ``f10!}\textit{?}

A. Effectual\index{Effectual call:calling ``f10!} calling is the work of God's Spirit\index{Holy Spirit:God's Spirit
``f10!}, whereby, convincing us\index{Effectual call:convincing of sin and misery ``f10!} of our sin and
misery\index{Punishment:misery ``f10!}, enlightening\index{Effectual call:enlightening minds spiritually ``f10!} our
minds in the knowledge of Christ\index{Effectual call:knowledge of Christ given ``f10!},\footnote{\ Acts 26:18\index{44
Acts:ch.  26, v.  18, pg. ``f30!}; 1 Corinthians 2:10, 12\index{46 1 Corinthians:ch.   2, v.  10, 12, pg. ``f30!}; 2
Corinthians 4:6\index{47 2 Corinthians:ch.   4, v.   6, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1,
pg. ``f30!}7-18\index{49 Ephesians:ch.   1, v.  17-18, pg. ``f30!}.} and renewing our\index{Conversion:renewed by Holy
Spirit ``f10!} wills,\footnote{\ Deuteronomy 30:6\index{05 Deuteronomy:ch.  30, v.   6, pg. ``f30!}; Ezekiel
36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36, v.  26-27, pg. ``f30!}; John
3:5\index{43 John:ch.   3, v.   5, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.} he does
persuade and\index{Effectual call:Holy Spirit persuades and enables belief ``f10!} enable us to embrace
Jesus\index{Effectual call:enables person to:embrace Jesus ``f10!} Christ,\footnote{\ John 6:44\index{43 John:ch.   6,
v.  44, pg. ``f30!}{}-45\index{43 John:ch.   6, v.  44-45, pg. ``f30!}; Acts 16:14\index{44 Acts:ch.  16, v.  14, pg.
``f30!}.} freely offered\index{Salvation:freely offered ``f10!} to us in the gospel\index{Gospel:gospel
``f10!}.\footnote{\ Isaiah 45:22\index{23 Isaiah:ch.  45, v.  22, pg. ``f30!}; Matthew 11:28\index{40 Matthew:ch.  11,
v.  28, pg. ``f30!}{}-30\index{40 Matthew:ch.  11, v.  28-30, pg. ``f30!}; Revelation 22:17\index{66 Revelation:ch. 
22, v.  17, pg. ``f30!}.}


\bigskip

Q. 32 \textit{What }benefits\textit{ do the}\index{Believers:benefits:as members of the Church
``f10!}\textit{y}\index{Believers:benefits:in this life ``f10!}\textit{ that are effectually called}\index{Effectual
call:calling ``f10!}\textit{ partake of in}\index{Effectual call:benefits in this life ``f10!}\textit{ this life?}

A. They that are effectually called\index{Effectual call:calling ``f10!} do in this life partake\index{Effectual
call:benefits in this life ``f10!} of justification\index{Justification:justification ``f10!},
adoption\index{Adoption:in salvation ``f10!}, and sanctification\index{Sanctification:sanctification ``f10!}, and the
several benefits which in\index{Lord's Supper:benefits believers receive ``f10!} this life do either accompany or flow
from them.\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg.
``f30!}; 1 Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg. ``f30!}; 1 Corinthians 6:11\index{46 1
Corinthians:ch.   6, v.  11, pg. ``f30!}; Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}.}


\bigskip

Q. 33 \textit{What is justification}\index{Justification:justification ``f10!}\textit{?}

A. Justification\index{Justification:justification ``f10!} is an act of God's free grace\index{Grace:God's free
``f10!},\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg.
``f30!}.} wherein he pardons\index{Justification:pardoned of sin ``f10!} all our sins,\footnote{\ Romans 4:6-8\index{45
Romans:ch.   4, v.   6-8, pg. ``f30!}; 2 Corinthians 5:19\index{47 2 Corinthians:ch.   5, v.  19, pg. ``f30!}.} and
accepts us\index{Believers:accepted in Christ ``f10!} as righteous in\index{Jesus:righteousness imputed to believers
``f10!} his sight,\footnote{\ 2 Corinthians 5:21\index{47 2 Corinthians:ch.   5, v.  21, pg. ``f30!}.} only for the
righteousness of\index{Jesus:righteous ``f10!} Christ imputed to\index{Imputation:of Christ's righteousness by faith
``f10!} us,\footnote{\ Romans 4:6, 11\index{45 Romans:ch.   4, v.   6, 11, pg. ``f30!}; Romans 5:1\index{45 Romans:ch. 
 5, v.   1, pg. ``f30!}9\index{45 Romans:ch.   5, v.  19, pg. ``f30!}.} and received by faith\index{Faith:instrument:of
justification ``f10!} alone.\footnote{\ Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Philippians
3:9\index{50 Philippians:ch.   3, v.   9, pg. ``f30!}.}


\bigskip

Q. 34 \textit{What is adoption}\index{Adoption:in salvation ``f10!}\textit{?}

A. Adoption\index{Adoption:in salvation ``f10!} is an act of God's free grace\index{Grace:God's free
``f10!},\footnote{\ 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}.} whereby we are received into the number,
and have a right to all the privileges\index{Believers:privileges ``f10!}, of the sons of God.\footnote{\ John
1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}; Romans 8:1\index{45
Romans:ch.   8, v.   1, pg. ``f30!}7\index{45 Romans:ch.   8, v.  17, pg. ``f30!}.}


\bigskip

Q. 35 \textit{What is }sanctification\index{Sanctification:sanctification ``f10!}\textit{?} 

A. Sanctification\index{Sanctification:sanctification ``f10!} is the work of God's free grace\index{Grace:God's free
``f10!},\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}; Philippians 2:13\index{50
Philippians:ch.   2, v.  13, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.}
whereby we are renewed in\index{Sanctification:renewed in image of God ``f10!} the whole man after the
image\index{Mankind:image of God ``f10!} of God,\footnote{\ 2 Corinthians 5:17\index{47 2 Corinthians:ch.   5, v.  17,
pg. ``f30!}; Ephesians 4:23-24\index{49 Ephesians:ch.   4, v.  23-24, pg. ``f30!}; 1 Thessalonians 5:23\index{52 1
Thessalonians:ch.   5, v.  23, pg. ``f30!}.} and are enabled more\index{Sanctification:enabled by:Holy Spirit ``f10!}
and more to die unto\index{Sanctification:die unto sin ``f10!} sin, and live unto
righteousness\index{Sanctification:spiritual growth:unto righteousness ``f10!}.\footnote{\ Ezekiel 36:25-27\index{26
Ezekiel:ch.  36, v.  25-27, pg. ``f30!}; Romans 6:4\index{45 Romans:ch.   6, v.   4, pg. ``f30!}, 6, 12-14\index{45
Romans:ch.   6, v.   4, 6, 12-14, pg. ``f30!}; 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; 1
Peter 2:2\index{60 1 Peter:ch.   2, v.   2, pg. ``f30!}4\index{60 1 Peter:ch.   2, v.  24, pg. ``f30!}.}


\bigskip

Q. 36 \textit{What }are\textit{ the benefits which in}\index{Lord's Supper:benefits believers receive ``f10!}\textit{
this life do accompany or flow from justification}\index{Justification:justification ``f10!}\textit{,
adoption}\index{Adoption:in salvation ``f10!}\textit{, and sanctification}\index{Sanctification:sanctification
``f10!}\textit{?}

A. The benefits which in\index{Lord's Supper:benefits believers receive ``f10!} this life do accompany or flow from
justification\index{Justification:justification ``f10!}, adoption\index{Adoption:in salvation ``f10!}, and
sanctification\index{Sanctification:sanctification ``f10!}, are, assurance\index{Assurance:of God's love, benefit of
justification ``f10!} of God's love\index{God's attributes:love ``f10!},\footnote{\ Romans 5:5\index{45 Romans:ch.   5,
v.   5, pg. ``f30!}.} peace of conscience\index{Conscience:peace of ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch. 
 5, v.   1, pg. ``f30!}.} joy in\index{Joy:in Holy Spirit ``f10!} the Holy Ghost\index{Holy Spirit:Holy Ghost
``f10!},\footnote{\ Romans 14:17\index{45 Romans:ch.  14, v.  17, pg. ``f30!}.} increase of grace\index{Grace:in
believers:by Holy Spirit ``f10!},\footnote{\ 2 Peter 3:18\index{61 2 Peter:ch.   3, v.  18, pg. ``f30!}.} and
perseverance therein\index{Perseverance:of the saints ``f10!} to the end.\footnote{\ Philippians 1:6\index{50
Philippians:ch.   1, v.   6, pg. ``f30!}; 1 Peter 1:5\index{60 1 Peter:ch.   1, v.   5, pg. ``f30!}.}


\bigskip

Q. 37 \textit{What benefits do believers receive}\index{Believers:benefits:received at death and resurrection
``f10!}\textit{ from Christ at death}\index{Death:believers:benefits ``f10!}\textit{?}

A. The souls\index{Believers:souls made perfect at death ``f10!} of believers are\index{Believers:made perfect in
holiness at death ``f10!} at their death made\index{Death:believers:immediate perfection ``f10!} perfect in
holiness\index{Believers:holy:made perfect at death ``f10!},\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2,
pg. ``f30!}3\index{58 Hebrews:ch.  12, v.  23, pg. ``f30!}.} and do immediately pass into
glory\index{Resurrection:believers:into glory:glorification ``f10!};\footnote{\ Luke 23:43\index{42 Luke:ch.  23, v. 
43, pg. ``f30!}; 2 Corinthians 5:6, 8\index{47 2 Corinthians:ch.   5, v.   6, 8, pg. ``f30!}; Philippians 1:23\index{50
Philippians:ch.   1, v.  23, pg. ``f30!}.} and their bodies, being\index{Bodies in graves:believers ``f10!} still
united to Christ\index{Believers:union with Christ:mystical ``f10!},\footnote{\ 1 Thessalonians 4:14\index{52 1
Thessalonians:ch.   4, v.  14, pg. ``f30!}.} do rest in their graves\index{Resurrection:bodies:await in graves ``f10!}
till the resurrection\index{Resurrection:general, of dead ``f10!}.\footnote{\ Daniel 12:2\index{27 Daniel:ch.  12, v.  
2, pg. ``f30!}; John 5:28-29\index{43 John:ch.   5, v.  28-29, pg. ``f30!}; Acts 24:15\index{44 Acts:ch.  24, v.  15,
pg. ``f30!}.}


\bigskip

Q. 38 \textit{What benefits do believers receive}\index{Believers:benefits:received at death and resurrection
``f10!}\textit{ from Christ at the resurrection}\index{Resurrection:believers:into glory:with Christ ``f10!}\textit{?}

A. At the resurrection\index{Resurrection:believers:into glory:glorification ``f10!}, believers being raised up in
glory,\footnote{\ 1 Corinthians 15:4\index{46 1 Corinthians:ch.  15, v.   4, pg. ``f30!}2-43\index{46 1 Corinthians:ch.
 15, v.  42-43, pg. ``f30!}.} shall be openly acknowledged\index{Judgement day:believers:acknowledged by God ``f10!}
and acquitted\index{Judgement day:believers:acquitted by God ``f10!} in the day of judgment\index{Judgement
day:believers:perfected ``f10!},\footnote{\ Matthew 25:33\index{40 Matthew:ch.  25, v.  33, pg. ``f30!}{}-34,
46\index{40 Matthew:ch.  25, v.  33-34, 46, pg. ``f30!}.} and made perfectly blessed in\index{Believers:blessed in/by
God ``f10!} the full enjoying of\index{Resurrection:believers:enjoy God ``f10!} God\footnote{\ Romans 8:2\index{45
Romans:ch.   8, v.   2, pg. ``f30!}9\index{45 Romans:ch.   8, v.  29, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3,
v.   2, pg. ``f30!}.} to all eternity\index{Believers:blessed to all eternity ``f10!}.\footnote{\ Psalm 16:11\index{19
Psalms:ch.  16, v.  11, pg. ``f30!}; 1 Thessalonians 4:17\index{52 1 Thessalonians:ch.   4, v.  17, pg. ``f30!}.}


\bigskip

Q. 39 \textit{What is the duty which God}\index{God's law:commandments:duties:required by God ``f10!}\textit{ requires
of man?}

A. The duty which God\index{God's law:commandments:duties:required by God ``f10!} requires of man, is obedience to his
revealed will\index{Revelation:God's will ``f10!}.\footnote{\ Deuteronomy 29:29\index{05 Deuteronomy:ch.  29, v.  29,
pg. ``f30!}; Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; 1 John 5:2-3\index{62 1 John:ch.   5, v.   2-3, pg.
``f30!}.}


\bigskip


\bigskip


\bigskip

Q. 40 \textit{What did God at first reveal to}\index{God's law:commandments:first revelation ``f10!}\textit{ man for the
rule of his}\index{God's law:moral:rule for obedience ``f10!}\textit{ obedience?}

A. The rule which\index{God's law:moral:rule for obedience ``f10!} God at first revealed to\index{Revelation:God's law
``f10!} man for his obedience, was the moral\index{God's law:commandments:moral ``f10!} law.\footnote{\ Romans
2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}; Romans
10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}.}


\bigskip

Q. 41 \textit{Wherein is the moral}\index{God's law:commandments:moral ``f10!}\textit{ law summarily
comprehended}\index{God's law:moral:summarily comprehended in ten commandments ``f10!}\textit{?}

A. The moral\index{God's law:commandments:moral ``f10!} law is summarily comprehended\index{God's law:moral:summarily
comprehended in ten commandments ``f10!} in the ten\index{God's law:commandments:ten ``f10!}
commandments.\footnote{\ Deuteronomy 4:13\index{05 Deuteronomy:ch.   4, v.  13, pg. ``f30!}; Matthew 19:17\index{40
Matthew:ch.  19, v.  17, pg. ``f30!}{}-19\index{40 Matthew:ch.  19, v.  17-19, pg. ``f30!}.}


\bigskip

Q. 42 \textit{What is the sum of the ten}\index{God's law:commandments:ten ``f10!}\textit{ commandments?}

A. The sum of the ten\index{God's law:commandments:ten ``f10!} commandments is, To love the Lord\index{Love toward
God:for God ``f10!} our God with all our heart, with all our soul\index{Love toward God:with all our:soul ``f10!}, with
all our strength\index{Love toward God:with all our:strength ``f10!}, and with all our mind; and our
neighbour\index{Love for others:neighbour as self ``f10!} as ourselves.\footnote{\ Matthew 22:37-40\index{40
Matthew:ch.  22, v.  37-40, pg. ``f30!}.}


\bigskip

Q. 43 \textit{What is the preface to}\index{God's law:commandments:preface ``f10!}\textit{ the ten}\index{God's
law:commandments:ten ``f10!}\textit{ commandments?}

A. The preface to\index{God's law:commandments:preface ``f10!} the ten\index{God's law:commandments:ten ``f10!}
commandments is in these words, I am the Lord thy God, which have brought you out of the land of Egypt\index{Egypt
``f10!}, out of the house of bondage\index{bondage, slavery in Egypt ``f10!}.\footnote{\ Exodus 20:2\index{02
Exodus:ch.  20, v.   2, pg. ``f30!}; Deuteronomy 5:6\index{05 Deuteronomy:ch.   5, v.   6, pg. ``f30!}.}


\bigskip

Q. 44 \textit{What does }the\textit{ preface to}\index{God's law:commandments:preface ``f10!}\textit{ the
ten}\index{God's law:commandments:ten ``f10!}\textit{ commandments teach us?}

A. The preface to\index{God's law:commandments:preface ``f10!} the ten\index{God's law:commandments:ten ``f10!}
commandments teaches us, That because God is the Lord, and our God, and Redeemer\index{Jesus:redeemer of God's elect
``f10!}, therefore we are bound to keep all his commandments\index{God's law:commandments:obedience:required
``f10!}.\footnote{\ Luke 1:74-75\index{42 Luke:ch.   1, v.  74-75, pg. ``f30!}; 1 Peter 1:14-19\index{60 1 Peter:ch.  
1, v.  14-19, pg. ``f30!}.}


\bigskip

Q. 45 \textit{Which is the first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The first commandment\index{God's law:commandments:first ``f10!} is, You shalt have no other
gods\index{Idolatry:worshiping gods ``f10!} before me.\footnote{\ Exodus 20:3\index{02 Exodus:ch.  20, v.   3, pg.
``f30!}; Deuteronomy 5:7\index{05 Deuteronomy:ch.   5, v.   7, pg. ``f30!}.}


\bigskip

Q. 46 \textit{What is required in the first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The first commandment\index{God's law:commandments:first ``f10!} requires us to know and acknowledge
God\index{Worship:acknowledge God ``f10!} to be the only true God, and our God; and to worship
and\index{Worship:glorifying God ``f10!} glorify him accordingly.\footnote{\ 1 Chronicles 28:9\index{13 1
Chronicles:ch.  28, v.   9, pg. ``f30!}; Isaiah 45:20-25\index{23 Isaiah:ch.  45, v.  20-25, pg. ``f30!}; Matthew
4:10\index{40 Matthew:ch.   4, v.  10, pg. ``f30!}.}


\bigskip

Q. 47 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the first
commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. The first commandment\index{God's law:commandments:first ``f10!} forbids the denying,\index{Idolatry:denying the true
God ``f10!}\footnote{\ Psalm 14:1\index{19 Psalms:ch.  14, v.   1, pg. ``f30!}.} or not worshipping and
glorifying\index{Worship:glorifying God ``f10!} the true God as God,\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.
 20, pg. ``f30!}{}-21\index{45 Romans:ch.   1, v.  20-21, pg. ``f30!}.} and our God;\footnote{\ Psalm 81:10-11\index{19
Psalms:ch.  81, v.  10-11, pg. ``f30!}.} and the giving of that worship and glory to any\index{Idolatry:giving glory to
another than God ``f10!} other, which is due to him alone.\footnote{\ Ezekiel 8:16-18\index{26 Ezekiel:ch.   8, v. 
16-18, pg. ``f30!}; Romans 1:25\index{45 Romans:ch.   1, v.  25, pg. ``f30!}.}


\bigskip


\bigskip

Q. 48 \textit{What are we specially taught by}\index{God's law:commandments:first, words 'before me' ``f10!}\textit{
these words, ``before me,'' in the first commandment}\index{God's law:commandments:first ``f10!}\textit{?}

A. These words, before me, in the first commandment\index{God's law:commandments:first ``f10!} teach us,
that\index{God's law:commandments:first, words 'before me' ``f10!} God, who sees all things, takes
notice\index{Idolatry:God takes special notice of this sin ``f10!} of, and is much displeased\index{Idolatry:displeases
God greatly ``f10!} with, the sin of having any other God.\footnote{\ Deuteronomy 30:17-18\index{05 Deuteronomy:ch. 
30, v.  17-18, pg. ``f30!}; Psalm 44:20-21\index{19 Psalms:ch.  44, v.  20-21, pg. ``f30!}; Ezekiel 8:12\index{26
Ezekiel:ch.   8, v.  12, pg. ``f30!}.}


\bigskip

Q. 49 \textit{Which is the second commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The second commandment\index{God's law:commandments:second ``f10!} is, You shalt not make unto you any graven image,
or\index{Idolatry:graven image ``f10!} any likeness\index{Idolatry:worship of likeness of any created thing ``f10!} of
anything that is in heaven above, or that is in the earth beneath, or that is in the water under\index{Idolatry:bowing
to created entity ``f10!} the earth: you shalt not bow down thy self to them, nor serve the\index{Love toward God:in
service ``f10!}m\index{Idolatry:serve images ``f10!}: for I the Lord thy God am a jealous\index{God's
attributes:jealous ``f10!} God, visiting\index{Punishment:visiting iniquity ``f10!} the iniquity of the fathers upon
the children unto the third and\index{Punishment:to third and fourth generation ``f10!} fourth generation of them that
hate\index{Sin:hatred of God ``f10!} me; and showing mercy unto\index{God's attributes:merciful ``f10!} thousands of
them that love me\index{Love toward God:for God ``f10!}, and keep my commandments\index{God's law:commandments:ten
``f10!}.\footnote{\ Exodus 20:4-6\index{02 Exodus:ch.  20, v.   4-6, pg. ``f30!}; Deuteronomy 5:8-10\index{05
Deuteronomy:ch.   5, v.   8-10, pg. ``f30!}.}


\bigskip

Q. 50 \textit{What is required in the second commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The second commandment\index{God's law:commandments:second ``f10!} requires the receiving, observing, and keeping
pure\index{Worship:to be observed and kept pure ``f10!} and entire, all such religious worship and ordinances
as\index{Worship:ordinances:to be:kept pure ``f10!} God has appointed in his Word\index{Scripture:word of
God:standard:for worship ``f10!}.\footnote{\ Deuteronomy 12:32\index{05 Deuteronomy:ch.  12, v.  32, pg. ``f30!};
Matthew 28:20\index{40 Matthew:ch.  28, v.  20, pg. ``f30!}.}


\bigskip

Q. 51 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the second
commandment}\index{God's law:commandments:second ``f10!}\textit{?}

A. The second commandment\index{God's law:commandments:second ``f10!} forbids the worshipping of\index{Worship:false:by
images ``f10!} God by images\index{Idolatry:images ``f10!},\footnote{\ Deuteronomy 4:15-19\index{05 Deuteronomy:ch.  
4, v.  15-19, pg. ``f30!}; Romans 1:22-23\index{45 Romans:ch.   1, v.  22-23, pg. ``f30!}.} or any other way not
appointed in his Word\index{Scripture:word of God:standard:for worship ``f10!}.\footnote{\ Leviticus 10:1-2\index{03
Leviticus:ch.  10, v.   1-2, pg. ``f30!}; Jeremiah 19:4-5\index{24 Jeremiah:ch.  19, v.   4-5, pg. ``f30!}; Colossians
2:18\index{51 Colossians:ch.   2, v.  18, pg. ``f30!}{}-23\index{51 Colossians:ch.   2, v.  18-23, pg. ``f30!}.}


\bigskip

Q. 52 \textit{What are the reasons annexed to the second commandment}\index{God's law:commandments:second
``f10!}\textit{?}

A. The reasons annexed to the second commandment\index{God's law:commandments:second ``f10!} are, God's
sovereignty\index{God's attributes:sovereign lordship ``f10!} over us,\footnote{\ Psalm 95:2-3, 6-7\index{19 Psalms:ch.
 95, v.   2-3, 6-7, pg. ``f30!}; Psalm 96:9-10\index{19 Psalms:ch.  96, v.   9-10, pg. ``f30!}.} his propriety in
us\index{Providence:God's propriety over mankind ``f10!},\footnote{\ Exodus 19:5\index{02 Exodus:ch.  19, v.   5, pg.
``f30!}; Psalm 45:11\index{19 Psalms:ch.  45, v.  11, pg. ``f30!}; Isaiah 54:5\index{23 Isaiah:ch.  54, v.   5, pg.
``f30!}.} and the zeal he\index{Worship:God's zeal for proper ``f10!} has to his own worship\index{Worship:of God
``f10!}.\footnote{\ Exodus 34:1\index{02 Exodus:ch.  34, v.   1, pg. ``f30!}4\index{02 Exodus:ch.  34, v.  14, pg.
``f30!}; 1 Corinthians 10:22\index{46 1 Corinthians:ch.  10, v.  22, pg. ``f30!}.}


\bigskip

Q. 53 \textit{Which is the third commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The third commandment\index{God's law:commandments:third ``f10!} is, You shalt not take the name of the Lord thy God
in vain\index{God's name:not to be:taken in vain ``f10!}; for the Lord will not hold him guiltless\index{God's name:not
guiltless if take in vain ``f10!} that takes his name in vain.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.   7,
pg. ``f30!}; Deuteronomy 5:11\index{05 Deuteronomy:ch.   5, v.  11, pg. ``f30!}.}


\bigskip


\bigskip


Q. 54 \textit{What is required in the third commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The third commandment\index{God's law:commandments:third ``f10!} requires the holy\index{God's name:holy ``f10!} and
reverend\index{God's name:to be used:reverently in thoughts and words ``f10!} use of God's names, titles\index{God's
name:titles to be honoured ``f10!},\footnote{\ Deuteronomy 10:20\index{05 Deuteronomy:ch.  10, v.  20, pg. ``f30!};
Psalm 29:2\index{19 Psalms:ch.  29, v.   2, pg. ``f30!}; Matthew 6:9\index{40 Matthew:ch.   6, v.   9, pg. ``f30!}.}
attributes\index{God's attributes:to be:revered ``f10!},\footnote{\ 1 Chronicles 29:10-13\index{13 1 Chronicles:ch. 
29, v.  10-13, pg. ``f30!}; Revelation 15:3-4\index{66 Revelation:ch.  15, v.   3-4, pg. ``f30!}.}
ordinances,\footnote{\ Acts 2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}; 1 Corinthians 11:2\index{46 1
Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v.  27, pg. ``f30!}{}-28\index{46 1
Corinthians:ch.  11, v.  27-28, pg. ``f30!}.} Word,\footnote{\ Psalm 138:2\index{19 Psalms:ch. 138, v.   2, pg.
``f30!}; Revelation 22:18-19\index{66 Revelation:ch.  22, v.  18-19, pg. ``f30!}.} and works\index{Providence:God's
works ``f10!}.\footnote{\ Psalm 107:21-22\index{19 Psalms:ch. 107, v.  21-22, pg. ``f30!}; Revelation 4:11\index{66
Revelation:ch.   4, v.  11, pg. ``f30!}.}


\bigskip

Q. 55 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the third
commandment}\index{God's law:commandments:third ``f10!}\textit{?}

A. The third commandment\index{God's law:commandments:third ``f10!} forbids all profaning or\index{God's name:not to
be:profaned ``f10!} abusing\index{God's name:not to be:abused ``f10!} of anything whereby God makes himself
known.\footnote{\ Leviticus 19:12\index{03 Leviticus:ch.  19, v.  12, pg. ``f30!}; Matthew 5:33-37\index{40 Matthew:ch.
  5, v.  33-37, pg. ``f30!}; James 5:12\index{59 James:ch.   5, v.  12, pg. ``f30!}.}


\bigskip

Q. 56 \textit{What is }the\textit{ reason annexed to the third commandment}\index{God's law:commandments:third
``f10!}\textit{?}

A. The reason annexed to the third commandment\index{God's law:commandments:third ``f10!} is, that however the breakers
of this commandment may escape punishment from\index{Punishment:escape from man's ``f10!} men, yet the Lord our God
will not suffer them to escape\index{Punishment:no escape ``f10!} his righteous judgment\index{Judgement:righteous
``f10!}.\footnote{\ Deuteronomy 28:58\index{05 Deuteronomy:ch.  28, v.  58, pg. ``f30!}{}-59\index{05 Deuteronomy:ch. 
28, v.  58-59, pg. ``f30!}; 1 Samuel 3:13\index{09 1 Samuel:ch.   3, v.  13, pg. ``f30!}; 1 Samuel 4:11\index{09 1
Samuel:ch.   4, v.  11, pg. ``f30!}.}


\bigskip

Q. 57 \textit{Which is the }fourth\textit{ commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The fourth commandment\index{God's law:commandments:fourth ``f10!} is, Remember\index{Sabbath:to be:remembered
``f10!} the sabbath day, to keep it holy\index{Sabbath:to be:kept holy ``f10!}. Six\index{Sabbath:six days:for labour
``f10!} days shalt you labor, and do\index{Sabbath:labour six days, rest on Sabbath ``f10!} all thy work; but the
seven\index{Sabbath:one whole day in seven ``f10!}th day is the sabbath of\index{Sabbath:instituted by God ``f10!} the
Lord thy God: in it you shalt not do any work, thou, nor thy son, nor thy daughter, thy manservant, nor thy
maidservant\index{Sabbath:servants not to be made to work ``f10!}, nor thy cattle, nor thy stranger that is within thy
gates. For in six\index{Creation:in six days ``f10!} days the Lord made heaven\index{God's attributes:creator:of heaven
and earth ``f10!} and earth, the sea\index{God's attributes:creator:of sea ``f10!}, and all that in them is, and rested
the\index{Sabbath:God rested the seventh day ``f10!} seventh day\index{Sabbath:seventh day ``f10!}: wherefore the Lord
blessed the\index{Sabbath:blessed ``f10!} sabbath day, and hallowed\index{God's name:hallowed ``f10!}
it.\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}{}-11\index{02 Exodus:ch.  20, v.   8-11, pg.
``f30!}; Deuteronomy 5:12-15\index{05 Deuteronomy:ch.   5, v.  12-15, pg. ``f30!}.}


\bigskip

Q. 58 \textit{What is required in the fourth commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The fourth commandment\index{God's law:commandments:fourth ``f10!} requires the keeping holy\index{Sabbath:to be:kept
holy ``f10!} to God such set times as he has appointed in his Word\index{Scripture:word of God:standard:for worship
``f10!}; expressly one\index{Sabbath:one whole day in seven ``f10!} whole day in seven, to be a holy sabbath to
himself.\footnote{\ Exodus 31:13, 16-17\index{02 Exodus:ch.  31, v.  13, 16-17, pg. ``f30!}.}


\bigskip

Q. 59 \textit{Which day of the seven}\index{Sabbath:one whole day in seven ``f10!}\textit{ has God appointed to be the
weekly}\index{Sabbath:weekly ``f10!}\textit{ sabbath?}

A. From the beginning of the world to the resurrection of Christ\index{Jesus:resurrection:from the dead ``f10!}, God
appointed the seven\index{Sabbath:one whole day in seven ``f10!}th day of the week to be the
weekly\index{Sabbath:weekly ``f10!} sabbath;\footnote{\ Genesis 2:2-3\index{01 Genesis:ch.   2, v.   2-3, pg. ``f30!};
Exodus 20:11\index{02 Exodus:ch.  20, v.  11, pg. ``f30!}.} and the first day\index{Sabbath:first day of week ``f10!}
of the week ever since, to continue to the end of the world, which is the Christian sabbath.\footnote{\ Mark
2:27-28\index{41 Mark:ch.   2, v.  27-28, pg. ``f30!}; Acts 20:7\index{44 Acts:ch.  20, v.   7, pg. ``f30!}; 1
Corinthians 16:2\index{46 1 Corinthians:ch.  16, v.   2, pg. ``f30!}; Revelation 1:10\index{66 Revelation:ch.   1, v. 
10, pg. ``f30!}.}

Q. 60 \textit{How is the sabbath to}\index{Sabbath:to be:kept holy ``f10!}\textit{ be
sanctified}\index{Sanctification:sanctification ``f10!}\textit{?}

A. The sabbath is to\index{Sabbath:to be:kept holy ``f10!} be sanctified\index{Sanctification:sanctification ``f10!} by
a holy resting all that day, even from such worldly employments and recreations\index{Sabbath:cease:worldly employments
and recreations ``f10!} as are lawful on other\index{Sabbath:cease:activities lawful on other days ``f10!}
days;\footnote{\ Exodus 20:10\index{02 Exodus:ch.  20, v.  10, pg. ``f30!}; Nehemiah 13:15-22\index{16 Nehemiah:ch. 
13, v.  15-22, pg. ``f30!}; Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}{}-14\index{23 Isaiah:ch.  58, v. 
13-14, pg. ``f30!}.} and spending\index{Sabbath:spend bulk of time in public and private worship ``f10!} the whole time
in the public and private exercises\index{Worship:private exercises ``f10!} of God's worship\index{Worship:of God
``f10!},\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}; Leviticus 23:3\index{03 Leviticus:ch. 
23, v.   3, pg. ``f30!}; Luke 4:16\index{42 Luke:ch.   4, v.  16, pg. ``f30!}; Acts 20:7\index{44 Acts:ch.  20, v.   7,
pg. ``f30!}.} except so much as is to be taken up in the works of necessity and\index{Sabbath:works of necessity and
mercy ``f10!} mercy.\footnote{\ Matthew 12:1-13\index{40 Matthew:ch.  12, v.   1-13, pg. ``f30!}.}


\bigskip

Q. 61 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the fourth
commandment}\index{God's law:commandments:fourth ``f10!}\textit{?}

A. The fourth commandment\index{God's law:commandments:fourth ``f10!} forbids the omission\index{Sin:omission of duties
``f10!} or careless performance of the duties required\index{God's law:commandments:duties:required by God ``f10!}, and
the profaning the\index{Sabbath:not to be profaned ``f10!} day by idleness\index{idleness ``f10!}, or doing that which
is in itself sinful, or by unnecessary thoughts\index{Sabbath:avoid unnecessary, worldly thoughts ``f10!}, words, or
works, about\index{Sabbath:cease:own works ``f10!} our worldly employments or recreations\index{Sabbath:cease:worldly
employments and recreations ``f10!}.\footnote{\ Nehemiah 13:15-22\index{16 Nehemiah:ch.  13, v.  15-22, pg. ``f30!};
Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}{}-14\index{23 Isaiah:ch.  58, v.  13-14, pg. ``f30!}; Amos
8:4-6\index{30 Amos:ch.   8, v.   4-6, pg. ``f30!}.}


\bigskip

Q. 62 \textit{What are the reasons annexed to the fourth commandment}\index{God's law:commandments:fourth
``f10!}\textit{?}

A. The reasons annexed to the fourth commandment\index{God's law:commandments:fourth ``f10!} are, God's allowing us
six\index{Sabbath:six days:for own affairs ``f10!} days of the week for our own employments\index{Sabbath:cease:worldly
employments and recreations ``f10!},\footnote{\ Exodus 20:9\index{02 Exodus:ch.  20, v.   9, pg. ``f30!}; Exodus
31:15\index{02 Exodus:ch.  31, v.  15, pg. ``f30!}; Leviticus 23:3\index{03 Leviticus:ch.  23, v.   3, pg. ``f30!}.}
his challenging a special propriety in the\index{Sabbath:God's propriety over ``f10!} seventh,
his\index{Sabbath:seventh day ``f10!} own example, and his blessing the sabbath day\index{Sabbath:blessed
``f10!}.\footnote{\ Genesis 2:2-3\index{01 Genesis:ch.   2, v.   2-3, pg. ``f30!}; Exodus 20:11\index{02 Exodus:ch. 
20, v.  11, pg. ``f30!}; Exodus 31:17\index{02 Exodus:ch.  31, v.  17, pg. ``f30!}.}


\bigskip

Q. 63 \textit{Which is the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. The fifth commandment\index{God's law:commandments:fifth ``f10!} is, Honour\index{Parents:to be honoured ``f10!} thy
father and thy mother\index{Superiors:styled as fathers and mothers ``f10!}; that thy days may be long upon the land
which the Lord thy God gives you.\footnote{\ Exodus 20:12\index{02 Exodus:ch.  20, v.  12, pg. ``f30!}; Deuteronomy
5:16\index{05 Deuteronomy:ch.   5, v.  16, pg. ``f30!}.}


\bigskip

Q. 64 \textit{What is required in the fifth commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. The fifth commandment\index{God's law:commandments:fifth ``f10!} requires the preserving\index{Equals:preserve honour
of others ``f10!} the honor, and\index{Love for others:seek neighbours' honour ``f10!} performing the
duties\index{God's law:commandments:duties:to others ``f10!}, belonging to everyone in their several places and
relations, as superiors\index{Superiors:styled as fathers and mothers ``f10!}, inferiors, or equals\index{Equals:duties
of ``f10!}.\footnote{\ Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}, 7\index{45 Romans:ch.  13, v.   1, 7,
pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}1\index{49 Ephesians:ch.   5, v.  21, pg.
``f30!}{}-22, 24\index{49 Ephesians:ch.   5, v.  21-22, 24, pg. ``f30!}; Ephesians 6:1, 4-5, 9\index{49 Ephesians:ch.  
6, v.   1, 4-5, 9, pg. ``f30!}; 1 Peter 2:17\index{60 1 Peter:ch.   2, v.  17, pg. ``f30!}.}


\bigskip

Q. 65 \textit{What is }forbidden\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the fifth
commandment}\index{God's law:commandments:fifth ``f10!}\textit{?}

A. The fifth commandment\index{God's law:commandments:fifth ``f10!} forbids the neglecting of\index{Love for
others:mutual:honour ``f10!}, or doing anything against, the honor and\index{Love for others:advancement of others
``f10!} duty which belongs\index{God's law:commandments:duties:to others ``f10!} to everyone in their several places
and relations.\footnote{\ Matthew 15:4-6\index{40 Matthew:ch.  15, v.   4-6, pg. ``f30!}; Romans 13:8\index{45
Romans:ch.  13, v.   8, pg. ``f30!}.}


\bigskip


\bigskip


\bigskip

Q. 66 \textit{What is the }reason\textit{ annexed to the fifth commandment}\index{God's law:commandments:fifth
``f10!}\textit{?}

A. The reason annexed to the fifth commandment\index{God's law:commandments:fifth ``f10!} is, a promise of
long\index{God's law:commandments:promises included ``f10!} life and prosperity (as far as it shall serve for God's
glory\index{God's law:commandments:obedience:for God's glory ``f10!} and their own good) to all such as keep this
commandment.\footnote{\ Exodus 20:12\index{02 Exodus:ch.  20, v.  12, pg. ``f30!}; Deuteronomy 5:16\index{05
Deuteronomy:ch.   5, v.  16, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2, pg. ``f30!}{}-3\index{49
Ephesians:ch.   6, v.   2-3, pg. ``f30!}.}


\bigskip

Q. 67 \textit{Which is the sixth commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The sixth commandment\index{God's law:commandments:sixth ``f10!} is, You shalt not kill\index{murder, unjust killing
``f10!}.\footnote{\ Exodus 20:13\index{02 Exodus:ch.  20, v.  13, pg. ``f30!}; Deuteronomy 5:17\index{05
Deuteronomy:ch.   5, v.  17, pg. ``f30!}.}


\bigskip

Q. 68 \textit{What is required in the sixth commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The sixth commandment\index{God's law:commandments:sixth ``f10!} requires all lawful endeavors\index{Human
life:protect with:lawful endeavors ``f10!} to preserve our\index{Human life:preservation ``f10!} own life, and the life
of others.\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}8-29\index{49 Ephesians:ch.   5, v. 
28-29, pg. ``f30!}.}


\bigskip

Q. 69 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the sixth
commandment}\index{God's law:commandments:sixth ``f10!}\textit{?}

A. The sixth commandment\index{God's law:commandments:sixth ``f10!} forbids the taking away of our own life, or the life
of our neighbour\index{Love for others:neighbour as self ``f10!}, unjustly, or whatsoever tends\index{Human life:defend
``f10!} thereunto.\footnote{\ Genesis 9:6\index{01 Genesis:ch.   9, v.   6, pg. ``f30!}; Matthew 5:22\index{40
Matthew:ch.   5, v.  22, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}5\index{62 1 John:ch.   3,
v.  15, pg. ``f30!}.}


\bigskip

Q. 70 \textit{Which }is\textit{ the seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The seventh commandment\index{God's law:commandments:seventh ``f10!} is, You shalt not commit adultery\index{Sexual
immorality:adultery ``f10!}.\footnote{\ Exodus 20:14\index{02 Exodus:ch.  20, v.  14, pg. ``f30!}; Deuteronomy
5:18\index{05 Deuteronomy:ch.   5, v.  18, pg. ``f30!}.}


\bigskip

Q. 71 \textit{What is required in the seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The seventh commandment\index{God's law:commandments:seventh ``f10!} requires the preservation of our\index{Sexual
purity:preservation ``f10!} own and our neighbour\index{Love for others:neighbour as self ``f10!}{}'s
chastity\index{Chastity:heart, speech, behaviour ``f10!}, in heart\index{Love for others:from heart ``f10!}, speech,
and behavior\index{Good works:behavior ``f10!}.\footnote{\ 1 Corinthians 7:2-3, 5\index{46 1 Corinthians:ch.   7, v.  
2-3, 5, pg. ``f30!}; 1 Thessalonians 4:3-5\index{52 1 Thessalonians:ch.   4, v.   3-5, pg. ``f30!}.}


\bigskip

Q. 72 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
seven}\index{Sabbath:one whole day in seven ``f10!}\textit{th commandment?}

A. The seventh commandment\index{God's law:commandments:seventh ``f10!} forbids all unchaste\index{Sexual
immorality:unchaste company and thoughts ``f10!} thoughts, words, and actions.\footnote{\ Matthew 5:28\index{40
Matthew:ch.   5, v.  28, pg. ``f30!}; Ephesians 5:3-4\index{49 Ephesians:ch.   5, v.   3-4, pg. ``f30!}.}


\bigskip

Q. 73 \textit{Which is the eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The eighth\index{God's law:commandments:eighth ``f10!} commandment is, You shalt not steal\index{Theft:stealing
``f10!}.\footnote{\ Exodus 20:15\index{02 Exodus:ch.  20, v.  15, pg. ``f30!}; Deuteronomy 5:19\index{05
Deuteronomy:ch.   5, v.  19, pg. ``f30!}.}


\bigskip


\bigskip


\bigskip


\bigskip

Q. 74 \textit{What is required in the eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The eighth\index{God's law:commandments:eighth ``f10!} commandment requires the lawful procuring\index{Material
things:to be lawfully procured ``f10!} and furthering\index{Love for others:advancement of others ``f10!} the wealth
and outward estate of ourselves and others.\footnote{\ Leviticus 25:35\index{03 Leviticus:ch.  25, v.  35, pg. ``f30!};
Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}b\index{49 Ephesians:ch.   4, v.  28b, pg. ``f30!};
Philippians 2:4\index{50 Philippians:ch.   2, v.   4, pg. ``f30!}.}


\bigskip

Q. 75 \textit{What }is\textit{ forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the
eighth}\index{God's law:commandments:eighth ``f10!}\textit{ commandment?}

A. The eighth\index{God's law:commandments:eighth ``f10!} commandment forbids whatsoever does, or may, unjustly hinder
our\index{Love for others:advancement of others ``f10!} own, or our neighbour\index{Love for others:neighbour as self
``f10!}{}'s, wealth or outward estate.\footnote{\ Proverbs 28:19\index{20 Proverbs:ch.  28, v.  19, pg. ``f30!}ff,
Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}a\index{49 Ephesians:ch.   4, v.  28a, pg. ``f30!}; 1
Thessalonians 3:10\index{52 1 Thessalonians:ch.   3, v.  10, pg. ``f30!}; 1 Timothy 5:8\index{54 1 Timothy:ch.   5, v. 
 8, pg. ``f30!}.}


\bigskip

Q. 76 \textit{Which is the ninth}\index{God's law:commandments:ninth ``f10!}\textit{ commandment?}

A. The ninth\index{God's law:commandments:ninth ``f10!} commandment is, You shalt not bear false witness
against\index{Falsehood:false witness against neighbour ``f10!} thy neighbour\index{Love for others:neighbour as self
``f10!}.\footnote{\ Exodus 20:16\index{02 Exodus:ch.  20, v.  16, pg. ``f30!}; Deuteronomy 5:20\index{05
Deuteronomy:ch.   5, v.  20, pg. ``f30!}.}


\bigskip

Q. 77 \textit{What is required in the ninth}\index{God's law:commandments:ninth ``f10!}\textit{ commandment?}

A. The ninth\index{God's law:commandments:ninth ``f10!} commandment requires the maintaining and
promoting\index{Truth:maintain and promote ``f10!} of truth between man and man, and of our own and our
neighbour\index{Love for others:neighbour as self ``f10!}{}'s good name,\footnote{\ Zechariah 8:16\index{38
Zechariah:ch.   8, v.  16, pg. ``f30!}; Acts 25:10\index{44 Acts:ch.  25, v.  10, pg. ``f30!}; 3 John 12\index{64 3
John:v.  12, pg. ``f30!}.} especially in witness bearing\index{Truth:in witness bearing ``f10!}.\footnote{\ Proverbs
14:5, 25\index{20 Proverbs:ch.  14, v.   5, 25, pg. ``f30!}.}


\bigskip

Q. 78 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the ninth}\index{God's
law:commandments:ninth ``f10!}\textit{ commandment?}

A. The ninth\index{God's law:commandments:ninth ``f10!} commandment forbids whatsoever is
prejudicial\index{Falsehood:prejudicing truth or justice ``f10!} to truth, or injurious to our own, or our
neighbour\index{Love for others:neighbour as self ``f10!}{}'s, good name.\footnote{\ Leviticus 19:16\index{03
Leviticus:ch.  19, v.  16, pg. ``f30!}; Psalm 15:3\index{19 Psalms:ch.  15, v.   3, pg. ``f30!}; Proverbs
6:16-19\index{20 Proverbs:ch.   6, v.  16-19, pg. ``f30!}; Luke 3:14\index{42 Luke:ch.   3, v.  14, pg. ``f30!}.}


\bigskip

Q. 79 \textit{Which is the tenth}\index{God's law:commandments:tenth ``f10!}\textit{ commandment?}

A. The tenth\index{God's law:commandments:tenth ``f10!} commandment is, You shalt not
covet\index{Covetousness:covetousness ``f10!} thy neighbour\index{Love for others:neighbour as self ``f10!}{}'s house,
you\index{Covetousness:nothing of neighbour ``f10!} shalt not covet thy neighbour's wife, nor his manservant, nor his
maidservant\index{Sabbath:servants not to be made to work ``f10!}, nor his ox, nor his ass, nor anything that is thy
neighbour's.\footnote{\ Exodus 20:17\index{02 Exodus:ch.  20, v.  17, pg. ``f30!}; Deuteronomy 5:21\index{05
Deuteronomy:ch.   5, v.  21, pg. ``f30!}.}


\bigskip

Q. 80 \textit{What is required in the tenth}\index{God's law:commandments:tenth ``f10!}\textit{ commandment?}

A. The tenth\index{God's law:commandments:tenth ``f10!} commandment requires full contentment\index{contentment ``f10!}
with our own condition,\footnote{\ Psalm 34:1\index{19 Psalms:ch.  34, v.   1, pg. ``f30!}; Philippians 4:11\index{50
Philippians:ch.   4, v.  11, pg. ``f30!}; 1 Timothy 6:6\index{54 1 Timothy:ch.   6, v.   6, pg. ``f30!}; Hebrews
13:5\index{58 Hebrews:ch.  13, v.   5, pg. ``f30!}.} with a right and charitable\index{Love for others:charitable
thoughts ``f10!} frame of spirit toward our neighbour\index{Love for others:neighbour as self ``f10!}, and all that is
his.\footnote{\ Luke 15:6, 9, 11-32\index{42 Luke:ch.  15, v.   6, 9, 11-32, pg. ``f30!}; Romans 12:15\index{45
Romans:ch.  12, v.  15, pg. ``f30!}; Philippians 2:4\index{50 Philippians:ch.   2, v.   4, pg. ``f30!}.}


\bigskip


\bigskip


\bigskip

Q. 81 \textit{What is forbidden}\index{God's law:commandments:sins forbidden ``f10!}\textit{ in the tenth}\index{God's
law:commandments:tenth ``f10!}\textit{ commandment?}

A. The tenth\index{God's law:commandments:tenth ``f10!} commandment forbids all discontentment\index{discontentment
``f10!} with our own estate,\footnote{\ 1 Corinthians 10:10\index{46 1 Corinthians:ch.  10, v.  10, pg. ``f30!}; James
3:1\index{59 James:ch.   3, v.   1, pg. ``f30!}4-16\index{59 James:ch.   3, v.  14-16, pg. ``f30!}.}
envying\index{Envy:envy ``f10!} or grieving at\index{Envy:grieving at success of others ``f10!} the good of our
neighbour\index{Love for others:neighbour as self ``f10!}, and all inordinate motions\index{Covetousness:inordinate
affections to others' goods ``f10!} and affections to anything that is his.\footnote{\ Galatians 5:2\index{48
Galatians:ch.   5, v.   2, pg. ``f30!}6\index{48 Galatians:ch.   5, v.  26, pg. ``f30!}; Colossians 3:5\index{51
Colossians:ch.   3, v.   5, pg. ``f30!}.}


\bigskip

Q. 82 \textit{Is any man able perfectly to keep the commandment}\index{God's law:commandments:given to Adam
``f10!}\textit{s}\index{God's law:commandments:ten ``f10!}\textit{ of God?}

A. No mere man, since the fall\index{Sin:original:first transgression (of Adam) ``f10!}, is able in this life perfectly
to keep the commandment\index{God's law:commandments:given to Adam ``f10!}s\index{God's law:commandments:ten ``f10!} of
God, but does daily break the\index{Lord's Supper:bread broken ``f10!}m\index{God's law:broken daily by all men
(thought, word, deed) ``f10!} in thought, word, and deed.\footnote{\ Genesis 8:21\index{01 Genesis:ch.   8, v.  21, pg.
``f30!}; Romans 3:9\index{45 Romans:ch.   3, v.   9, pg. ``f30!}ff, 23.}


\bigskip

Q. 83 \textit{Are all transgressions}\index{God's law:commandments:transgressions ``f10!}\textit{ of the law equally
heinous}\index{God's law:commandments:all transgression equally heinous ``f10!}\textit{?}

A. Some sins in themselves, and by reason of several aggravations\index{Sin:aggravations:which compound ``f10!}, are
more heinous\index{Sin:aggravations:make more heinous ``f10!} in the sight of God than others.\footnote{\ Ezekiel 8:6,
13, 15\index{26 Ezekiel:ch.   8, v.   6, 13, 15, pg. ``f30!}; Matthew 11:20-24\index{40 Matthew:ch.  11, v.  20-24, pg.
``f30!}; John 19:11\index{43 John:ch.  19, v.  11, pg. ``f30!}.}


\bigskip

Q. 84 \textit{What does every sin deserve}\index{Sin:deserves damnation ``f10!}\textit{?}

A. Every sin deserves\index{Sin:deserves damnation ``f10!} God's wrath and\index{Wrath:of God:for sin ``f10!}
curse,\index{Curse:of God ``f10!} both in this life, and that which is to come.\footnote{\ Matthew 25:41\index{40
Matthew:ch.  25, v.  41, pg. ``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}; Ephesians
5:6\index{49 Ephesians:ch.   5, v.   6, pg. ``f30!}; James 2:10\index{59 James:ch.   2, v.  10, pg. ``f30!}.}


\bigskip

Q. 85 \textit{What does God require of us}\index{God's law:commandments:duties:required by God ``f10!}\textit{, that we
may escape}\index{Wrath:of God:requirements to escape ``f10!}\textit{ his wrath and}\index{Wrath:of God:for sin
``f10!}\textit{ curse,}\index{Curse:of God ``f10!}\textit{ due to us for sin?}

A. To escape\index{Wrath:of God:requirements to escape ``f10!} the wrath and\index{Wrath:of God:for sin ``f10!} curse of
God\index{Curse:of God ``f10!}, due to us for sin, God requires of us\index{Repentance:required for pardon ``f10!}
faith in Jesus\index{Faith:in Jesus Christ ``f10!} Christ, repentance unto\index{Repentance:unto life ``f10!}
life,\footnote{\ Mark 1:15\index{41 Mark:ch.   1, v.  15, pg. ``f30!}; Acts 20:21\index{44 Acts:ch.  20, v.  21, pg.
``f30!}.} with the diligent use\index{Grace:diligent use of means ``f10!} of all the outward means whereby Christ
communicates to us the benefits of redemption\index{Believers:benefits:of redemption ``f10!}.\footnote{\ Acts
2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg.
``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  24-25, pg.
``f30!}; Colossians 3:16\index{51 Colossians:ch.   3, v.  16, pg. ``f30!}.}


\bigskip

Q. 86 \textit{What is faith in Jesus}\index{Faith:in Jesus Christ ``f10!}\textit{ Christ?}

A. Faith\index{Faith:saving ``f10!} in Jesus Christ is a saving grace\index{Grace:of salvation
``f10!},\footnote{\ Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}{}-9\index{49 Ephesians:ch.   2, v.  
8-9, pg. ``f30!}; cf. Romans 4:16\index{45 Romans:ch.   4, v.  16, pg. ``f30!}.} whereby we receive and rest upon him
alone for salvation\index{Salvation:belief in Jesus by faith ``f10!}, as he is offered to us\index{Faith:receives Jesus
as offered in Gospel ``f10!} in the gospel\index{Gospel:gospel ``f10!}.\footnote{\ John 20:30-31\index{43 John:ch.  20,
v.  30-31, pg. ``f30!}; Galatians 2:15-16\index{48 Galatians:ch.   2, v.  15-16, pg. ``f30!}; Philippians
3:3-11\index{50 Philippians:ch.   3, v.   3-11, pg. ``f30!}.}


\bigskip


\bigskip


\bigskip


Q. 87 \textit{What is repentance unto}\index{Repentance:unto life ``f10!}\textit{ life?}

A. Repentance\index{Repentance:unto life ``f10!} unto life is a saving grace\index{Grace:of salvation
``f10!},\footnote{\ Acts 11:18\index{44 Acts:ch.  11, v.  18, pg. ``f30!}; 2 Timothy 2:2\index{55 2 Timothy:ch.   2, v.
  2, pg. ``f30!}5\index{55 2 Timothy:ch.   2, v.  25, pg. ``f30!}.} whereby a sinner, out of a true sense of his sin,
and apprehension\index{Repentance:apprehension of God's mercy ``f10!} of the mercy of\index{God's attributes:merciful
``f10!} God in Christ,\footnote{\ Psalm 51:1\index{19 Psalms:ch.  51, v.   1, pg. ``f30!}{}-4\index{19 Psalms:ch.  51,
v.   1-4, pg. ``f30!}; Joel 2:13\index{29 Joel:ch.   2, v.  13, pg. ``f30!}; Luke 15:7, 10\index{42 Luke:ch.  15, v.  
7, 10, pg. ``f30!}; Acts 2:37\index{44 Acts:ch.   2, v.  37, pg. ``f30!}.} does, with grief\index{Sin:grief and hatred
of ``f10!} and hatred of his\index{Sin:to be hated ``f10!} sin, turn from it unto God,\footnote{\ Jeremiah
31:18\index{24 Jeremiah:ch.  31, v.  18, pg. ``f30!}{}-19\index{24 Jeremiah:ch.  31, v.  18-19, pg. ``f30!}; Luke
1:16-17\index{42 Luke:ch.   1, v.  16-17, pg. ``f30!}; 1 Thessalonians 1:9\index{52 1 Thessalonians:ch.   1, v.   9,
pg. ``f30!}.} with full purpose of, and endeavour after\index{God's law:commandments:obedience:endeavoring ``f10!}, new
obedience\index{Believers:obedience:strive for new ``f10!}.\footnote{\ 2 Chronicles 7:14\index{14 2 Chronicles:ch.   7,
v.  14, pg. ``f30!}; Psalm 119:5\index{19 Psalms:ch. 119, v.   5, pg. ``f30!}7-64\index{19 Psalms:ch. 119, v.  57-64,
pg. ``f30!}; Matthew 3:8\index{40 Matthew:ch.   3, v.   8, pg. ``f30!}; 2 Corinthians 7:1\index{47 2 Corinthians:ch.  
7, v.   1, pg. ``f30!}0\index{47 2 Corinthians:ch.   7, v.  10, pg. ``f30!}.}


\bigskip

Q. 88 \textit{What are the outward and ordinary means whereby Christ communicates to us the }benefits\textit{ of
redemption}\index{Believers:benefits:of redemption ``f10!}\textit{?}

A. The outward and ordinary means whereby Christ communicates to us the benefits of
redemption\index{Believers:benefits:of redemption ``f10!} are, his ordinances, especially\index{Church:furnished
with:ordinances of God ``f10!} the Word, sacraments\index{Sacraments:holy:ordinances ``f10!}, and prayer;
all\index{Prayer:act of worship ``f10!} which are made effectual\index{Worship:ordinances:made effectual for salvation
``f10!} to the elect\index{Election:elect of God ``f10!} for salvation\index{Salvation:belief in Jesus by faith
``f10!}.\footnote{\ Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v. 
18-20, pg. ``f30!}; Acts 2:41\index{44 Acts:ch.   2, v.  41, pg. ``f30!}{}-42\index{44 Acts:ch.   2, v.  41-42, pg.
``f30!}.}


\bigskip

Q. 89 \textit{How is the Word made effectual}\index{Scripture:word of God:for salvation ``f10!}\textit{ to
salvation}\index{Salvation:belief in Jesus by faith ``f10!}\textit{?}

A. The Spirit\index{Holy Spirit:Holy Spirit ``f10!} of God makes the reading\index{Scripture:to be:read and heard
``f10!}, but especially the preaching of the Word\index{Preaching:ministry of the Word ``f10!}, an
effectual\index{Preaching:effectual means of convincing and enlightening ``f10!} means of convincing
and\index{Preaching:convincing and converting sinners ``f10!} converting\index{Conversion:conversion ``f10!} sinners,
and of building them\index{Holy Spirit:builds up the elect ``f10!} up in holiness and\index{Believers:holy:being
perfected ``f10!} comfort, through\index{Faith:source of comfort ``f10!} faith, unto salvation\index{Salvation:belief
in Jesus by faith ``f10!} [a].\footnote{\ Nehemiah 8:8\index{16 Nehemiah:ch.   8, v.   8, pg. ``f30!}{}-9\index{16
Nehemiah:ch.   8, v.   8-9, pg. ``f30!}; Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}; Romans 10:1\index{45
Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg. ``f30!}{}-17\index{45 Romans:ch.  10, v. 
14-17, pg. ``f30!}; 2 Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v. 
15-17, pg. ``f30!}.}


\bigskip

Q. 90 \textit{How is the Word }to\textit{ be read}\index{Scripture:to be:read and heard ``f10!}\textit{ and heard, that
it may become effectual}\index{Sacraments:effectual means of salvation ``f10!}\textit{ to
salvation}\index{Salvation:belief in Jesus by faith ``f10!}\textit{?}

A. That the Word may\index{Scripture:word of God:for salvation ``f10!} become effectual\index{Sacraments:effectual means
of salvation ``f10!} to salvation\index{Salvation:belief in Jesus by faith ``f10!}, we must attend thereunto with
diligence\index{Preaching:diligence in application ``f10!}, preparation, and\index{Preaching:preparation for ``f10!}
prayer;\footnote{\ Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg. ``f30!}ff, Psalm 119:18\index{19
Psalms:ch. 119, v.  18, pg. ``f30!}; 1 Peter 2:1-2\index{60 1 Peter:ch.   2, v.   1-2, pg. ``f30!}.} receive it with
faith\index{Faith:in prayer ``f10!} and love, lay\index{Scripture:lay up in heart ``f10!} it up in our
hearts\index{Heart:new:through ministry of Word ``f10!}, and practice it in our lives.\footnote{\ Psalm 119:11\index{19
Psalms:ch. 119, v.  11, pg. ``f30!}; 2 Thessalonians 2:10\index{53 2 Thessalonians:ch.   2, v.  10, pg. ``f30!};
Hebrews 4:2\index{58 Hebrews:ch.   4, v.   2, pg. ``f30!}; James 1:22\index{59 James:ch.   1, v.  22, pg.
``f30!}{}-25\index{59 James:ch.   1, v.  22-25, pg. ``f30!}.}


\bigskip

Q. 91 \textit{How do the sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ become
effectual}\index{Sacraments:effectual means of salvation ``f10!}\textit{ means of salvation}\index{Salvation:belief in
Jesus by faith ``f10!}\textit{?}

A. The sacraments\index{Sacraments:holy:ordinances ``f10!} become effectual\index{Sacraments:effectual means of
salvation ``f10!} means of salvation\index{Salvation:belief in Jesus by faith ``f10!}, not from any
virtue\index{Sacraments:grace:not conferred by elements ``f10!} in them, or in him that does administer them; but only
by the blessing of Christ, and the working of his\index{Salvation:by working of Holy Spirit ``f10!} Spirit in them that
by faith\index{Faith:worked in believers by the Holy Spirit ``f10!} receive them.\footnote{\ 1 Corinthians 3:7\index{46
1 Corinthians:ch.   3, v.   7, pg. ``f30!} cf. 1 Corinthians 1:12-17\index{46 1 Corinthians:ch.   1, v.  12-17, pg.
``f30!}.}


\bigskip

Q. 92 \textit{What is a sacrament}\index{Sacraments:holy:ordinances ``f10!}\textit{?}

A. A sacrament\index{Sacraments:holy:ordinances ``f10!} is an holy ordinance instituted by
Christ\index{Sacraments:instituted by:Christ ``f10!};\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg.
``f30!}; Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg.
``f30!}; Mark 14:22-25\index{41 Mark:ch.  14, v.  22-25, pg. ``f30!}; Luke 22:19\index{42 Luke:ch.  22, v.  19, pg.
``f30!}{}-20\index{42 Luke:ch.  22, v.  19-20, pg. ``f30!}; 1 Corinthians 1:2\index{46 1 Corinthians:ch.   1, v.   2,
pg. ``f30!}2-26\index{46 1 Corinthians:ch.   1, v.  22-26, pg. ``f30!}.} wherein, by sensible
signs\index{Sacraments:holy:signs ``f10!}, Christ, and the benefits of the\index{Covenant:grace:benefits ``f10!} new
covenant, are\index{Covenant:new:benefits ``f10!} represented\index{Sacraments:represent Christ ``f10!}, sealed,
and\index{Sacraments:benefits:sealed ``f10!} applied to believers\index{Sacraments:applied to believers
``f10!}.\footnote{\ Galatians 3:27\index{48 Galatians:ch.   3, v.  27, pg. ``f30!}; 1 Corinthians 10:16\index{46 1
Corinthians:ch.  10, v.  16, pg. ``f30!}{}-17\index{46 1 Corinthians:ch.  10, v.  16-17, pg. ``f30!}.}

Q. 93 \textit{Which are the sacraments}\index{Sacraments:holy:ordinances ``f10!}\textit{ of the New
Testament}\index{Scripture:New Testament ``f10!}\textit{?}

A. The sacraments\index{Sacraments:holy:ordinances ``f10!} of the New Testament\index{Scripture:New Testament ``f10!}
are, Baptism\index{Baptism:sacrament ``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.}
and the Lord's Supper\index{Lord's Supper:sacrament ``f10!}.\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch. 
11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v. 
23-26, pg. ``f30!}.}


\bigskip

Q. 94 \textit{What is Baptism}\index{Baptism:definition ``f10!}\textit{?}

A. Baptism is a\index{Baptism:sacrament ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!}, wherein the
washing\index{Baptism:washing with water ``f10!} with water in\index{Baptism:to use water ``f10!} the name of the
Father\index{Father:God the ``f10!}, and of the Son\index{Jesus:Son of God ``f10!}, and of the Holy Ghost\index{Holy
Spirit:Holy Ghost ``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.} does
signify\index{Baptism:sign of ingrafting into Christ ``f10!} and seal our\index{Baptism:seal of salvation ``f10!}
ingrafting\index{Believers:ingrafted into Christ ``f10!} into Christ, and partaking of the benefits of
the\index{Covenant:grace:benefits ``f10!} covenant of grace\index{Covenant:grace:grace ``f10!}, and our engagement to
be\index{Oaths and Vows:engagements to God ``f10!} the Lord's.\footnote{\ Acts 2:38\index{44 Acts:ch.   2, v.  38, pg.
``f30!}{}-42\index{44 Acts:ch.   2, v.  38-42, pg. ``f30!}; Acts 22:16\index{44 Acts:ch.  22, v.  16, pg. ``f30!};
Romans 6:3-4\index{45 Romans:ch.   6, v.   3-4, pg. ``f30!}; Galatians 3:26-27\index{48 Galatians:ch.   3, v.  26-27,
pg. ``f30!}; 1 Peter 3:21\index{60 1 Peter:ch.   3, v.  21, pg. ``f30!}.}


\bigskip

Q. 95 \textit{To whom is Baptism to be administered}\index{Baptism:recipients ``f10!}\textit{?}

A. Baptism is not to be administered to any\index{Baptism:not to be administered outside those in visible church ``f10!}
that are out of the visible church\index{Church:visible ``f10!}, till they profess their faith\index{Faith:profession
of ``f10!} in Christ, and obedience\index{God's law:commandments:obedience:required ``f10!} to him;\footnote{\ Acts.
2:41; Acts 8:12, 36\index{44 Acts:ch.   8, v.  12, 36, pg. ``f30!}, 38\index{44 Acts:ch.   8, v.  12, 36, 38, pg.
``f30!}; Acts 18:8\index{44 Acts:ch.  18, v.   8, pg. ``f30!}.} but the infants of such\index{Infants:of believing
parents, to be baptized ``f10!} as are members of the\index{Church members ``f10!} visible church are to be
baptized\index{Baptism:infants ``f10!}.\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!},
9\index{01 Genesis:ch.  17, v.   7, 9, pg. ``f30!}{}-11\index{01 Genesis:ch.  17, v.   7, 9-11, pg. ``f30!}; Acts
2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!}; Acts 16:32\index{44
Acts:ch.  16, v.  32, pg. ``f30!}{}-33\index{44 Acts:ch.  16, v.  32-33, pg. ``f30!}; Colossians 2:11-12\index{51
Colossians:ch.   2, v.  11-12, pg. ``f30!}.}


\bigskip

Q. 96 \textit{What is }the\textit{ Lord's Supper}\index{Lord's Supper:sacrament ``f10!}\textit{?}

A. The Lord's Supper\index{Lord's Supper:sacrament ``f10!} is a sacrament\index{Sacraments:holy:ordinances ``f10!},
wherein, by giving and receiving\index{Lord's Supper:defined ``f10!} bread and\index{Lord's Supper:bread ``f10!}
wine\index{Lord's Supper:wine ``f10!}, according to Christ's appointment\index{Lord's Supper:proper administration
``f10!}, his death\index{Jesus:death:death ``f10!} is showed forth;\footnote{\ Luke 22:19\index{42 Luke:ch.  22, v. 
19, pg. ``f30!}{}-20\index{42 Luke:ch.  22, v.  19-20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11,
v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v. 
23-26, pg. ``f30!}.} and the worthy\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!} receivers are, not
after a corporal\index{Lord's Supper:spiritual feeding:not corporal ``f10!} and carnal manner\index{Lord's
Supper:spiritual feeding:not carnal ``f10!}, but by faith, made partakers of his\index{Lord's Supper:partakers of
Christ through faith ``f10!} body and blood\index{Lord's Supper:body and blood of Jesus ``f10!}, with all his benefits,
to\index{Lord's Supper:benefits believers receive ``f10!} their spiritual nourishment\index{Lord's Supper:spiritual
feeding:nourishment ``f10!}, and growth in grace\index{Sanctification:spiritual growth:in grace ``f10!}.\footnote{\ 1
Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}{}-17\index{46 1 Corinthians:ch.  10, v.  16-17,
pg. ``f30!}.}


\bigskip

Q. 97 \textit{What is required for the worthy}\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!}\textit{
receiving of the}\index{Worship:administration of sacraments ``f10!}\textit{ Lord's Supper}\index{Lord's
Supper:sacrament ``f10!}\textit{?}

A. It is required of them that would worthily\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!} partake of
the Lord's Supper\index{Lord's Supper:sacrament ``f10!}, that they examine themselves\index{Lord's
Supper:preparation:self-examination ``f10!} of their knowledge to discern the\index{Lord's Supper:discern Lord's body
``f10!} Lord's body, of their faith to\index{Faith:applied in Lord's Supper ``f10!} feed\index{Lord's Supper:spiritual
feeding:in faith ``f10!} upon him, of their repentance\index{Repentance:for all sins ``f10!}, love, and new
obedience\index{Believers:obedience:strive for new ``f10!}; lest, coming unworthily\index{Lord's Supper:unworthy
participation ``f10!}, they eat and drink judgment to\index{Judgement:temporal ``f10!} themselves.\footnote{\ 1
Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v.  27, pg.
``f30!}{}-32\index{46 1 Corinthians:ch.  11, v.  27-32, pg. ``f30!}.}


\bigskip


\bigskip


Q. 98 \textit{What is prayer}\index{Prayer:defined ``f10!}\textit{?}

A. Prayer is an\index{Prayer:defined ``f10!} offering up of our\index{Prayer:offering desires to God ``f10!} desires
unto God,\footnote{\ Psalm 10:17\index{19 Psalms:ch.  10, v.  17, pg. ``f30!}; Psalm 62:8\index{19 Psalms:ch.  62, v.  
8, pg. ``f30!}; Matthew 7:7-8\index{40 Matthew:ch.   7, v.   7-8, pg. ``f30!}.} for things
agreeable\index{Prayer:requests agreeable to God's will ``f10!} to his will,\footnote{\ 1 John 5:14\index{62 1 John:ch.
  5, v.  14, pg. ``f30!}.} in the name of Christ,\footnote{\ John 16:2\index{43 John:ch.  16, v.   2, pg.
``f30!}3\index{43 John:ch.  16, v.  23, pg. ``f30!}{}-24\index{43 John:ch.  16, v.  23-24, pg. ``f30!}.} with
confession\index{Confession:of sin ``f10!} of our sins,\footnote{\ Psalm 32:5\index{19 Psalms:ch.  32, v.   5, pg.
``f30!}{}-6\index{19 Psalms:ch.  32, v.   5-6, pg. ``f30!}; Daniel 9:4\index{27 Daniel:ch.   9, v.   4, pg.
``f30!}{}-19\index{27 Daniel:ch.   9, v.   4-19, pg. ``f30!}; 1 John 1:9\index{62 1 John:ch.   1, v.   9, pg. ``f30!}.}
and thankful\index{Prayer:thanking God ``f10!} acknowledgement\index{Judgement day:believers:acknowledged by God
``f10!} of his mercies\index{God's attributes:merciful ``f10!}.\footnote{\ Psalm 103:1\index{19 Psalms:ch. 103, v.   1,
pg. ``f30!}{}-5\index{19 Psalms:ch. 103, v.   1-5, pg. ``f30!}; Psalm 136; Philippians 4:6\index{50 Philippians:ch.  
4, v.   6, pg. ``f30!}.}


\bigskip

Q. 99 \textit{What rule has God given for}\index{Prayer:direction for ``f10!}\textit{ our direction in prayer?}

A. The whole Word of God\index{Scripture:word of God:word of God ``f10!} is of use to direct us in
prayer\index{Prayer:direction for ``f10!};\footnote{\ 1 John 5:14\index{62 1 John:ch.   5, v.  14, pg. ``f30!}.} but
the special rule of direction is that form of prayer\index{Prayer:Lord's:pattern for our prayers ``f10!} which Christ
taught his\index{Prayer:Lord's:Jesus taught his disciples ``f10!} disciples\index{disciples of Jesus ``f10!}, commonly
called The Lord's Prayer\index{Prayer:Lord's:The Lord's Prayer ``f10!}.\footnote{\ Matthew 6:9\index{40 Matthew:ch.  
6, v.   9, pg. ``f30!}{}-13\index{40 Matthew:ch.   6, v.   9-13, pg. ``f30!}.}


\bigskip

Q. 100 \textit{What does the preface of}\index{Prayer:Lord's:preface ``f10!}\textit{ }the\textit{ Lord's
Prayer}\index{Prayer:Lord's:The Lord's Prayer ``f10!}\textit{ teach us?}

A. The preface of\index{Prayer:Lord's:preface ``f10!} the Lord's Prayer\index{Prayer:Lord's:The Lord's Prayer ``f10!},
which is, Our Father\index{Father:God the ``f10!} which art in heaven\index{Heaven:God's abode ``f10!}, teaches us to
draw near\index{Prayer:draw near to God ``f10!} to God with all holy\index{Prayer:holy reverence ``f10!}
reverence\footnote{\ Psalm 95:6\index{19 Psalms:ch.  95, v.   6, pg. ``f30!}.} and confidence\index{Prayer:with
confidence ``f10!},\footnote{\ Ephesians 3:12\index{49 Ephesians:ch.   3, v.  12, pg. ``f30!}.} as children to a
father,\footnote{\ Matthew 7:9-11\index{40 Matthew:ch.   7, v.   9-11, pg. ``f30!}; cf. Luke 11:11-13\index{42 Luke:ch.
 11, v.  11-13, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}5\index{45 Romans:ch.   8, v.  15,
pg. ``f30!}.} able and ready to help\index{Prayer:helped by Holy Spirit ``f10!} us;\footnote{\ Ephesians 3:20\index{49
Ephesians:ch.   3, v.  20, pg. ``f30!}.} and that we should pray with an\index{Prayer:praising God
``f10!}d\index{Prayer:for others ``f10!} for others.\footnote{\ Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg.
``f30!}; 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg.
``f30!}.}


\bigskip

Q. 101 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the first
petition}\index{Prayer:Lord's:petitions:first ``f10!}\textit{?}

A. In the first petition\index{Prayer:Lord's:petitions:first ``f10!}, which is, Hallowed\index{God's name:hallowed
``f10!} be thy name, we pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God would
enable\index{Worship:correct:enabled by Holy Spirit ``f10!} us, and others, to glorify\index{Worship:glorifying God
``f10!} him in all that whereby he makes himself known;\footnote{\ Psalm 67:1-3\index{19 Psalms:ch.  67, v.   1-3, pg.
``f30!}; Psalm 99:3\index{19 Psalms:ch.  99, v.   3, pg. ``f30!}; Psalm 100:3-4\index{19 Psalms:ch. 100, v.   3-4, pg.
``f30!}.} and that he would dispose all\index{Providence:God disposes all things for his glory ``f10!} things to his
own glory\index{God's attributes:glory:his own ``f10!}.\footnote{\ Romans 11:33\index{45 Romans:ch.  11, v.  33, pg.
``f30!}{}-36\index{45 Romans:ch.  11, v.  33-36, pg. ``f30!}; Revelation 4:11\index{66 Revelation:ch.   4, v.  11, pg.
``f30!}.}


\bigskip

Q. 102 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the second
petition}\index{Prayer:Lord's:petitions:second ``f10!}\textit{?}

A. In the second petition\index{Prayer:Lord's:petitions:second ``f10!}, which is, Thy kingdom come\index{Kingdom of
God:prayer for advancement ``f10!}, we pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} Satan's
kingdom may be destroyed\index{Prayer:for kingdom of sin to be destroyed ``f10!};\footnote{\ Matthew 12:25\index{40
Matthew:ch.  12, v.  25, pg. ``f30!}{}-28\index{40 Matthew:ch.  12, v.  25-28, pg. ``f30!}; Romans 16:2\index{45
Romans:ch.  16, v.   2, pg. ``f30!}0\index{45 Romans:ch.  16, v.  20, pg. ``f30!}; 1 John 3:8\index{62 1 John:ch.   3,
v.   8, pg. ``f30!}.} and that the kingdom of grace may be\index{Grace:kingdom of (advanced) ``f10!}
advanced,\footnote{\ Psalm 72:8-11\index{19 Psalms:ch.  72, v.   8-11, pg. ``f30!}; Matthew 24:14\index{40 Matthew:ch. 
24, v.  14, pg. ``f30!}; 1 Corinthians 15:24-25\index{46 1 Corinthians:ch.  15, v.  24-25, pg. ``f30!}.} ourselves and
others brought into it, and kept in it;\footnote{\ Psalm 119:5\index{19 Psalms:ch. 119, v.   5, pg. ``f30!}; Luke
22:32\index{42 Luke:ch.  22, v.  32, pg. ``f30!}; 2 Thessalonians 3:1\index{53 2 Thessalonians:ch.   3, v.   1, pg.
``f30!}{}-5\index{53 2 Thessalonians:ch.   3, v.   1-5, pg. ``f30!}.} and that the kingdom of glory
may\index{Prayer:for glory of God ``f10!} be hastened\index{Kingdom:of glory, desire to see hastened
``f10!}.\footnote{\ Revelation 22:20\index{66 Revelation:ch.  22, v.  20, pg. ``f30!}.}


\bigskip

Q. 103 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the third
petition}\index{Prayer:Lord's:petitions:third ``f10!}\textit{?}

A. In the third petition\index{Prayer:Lord's:petitions:third ``f10!}, which is, Thy will be done in earth, as it is in
heaven\index{Heaven:God's will done ``f10!}, we pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God,
by his grace\index{Grace:God's free ``f10!}, would make us able and willing to know\index{Conversion:renewal of the
will ``f10!}, obey, and\index{God's law:obligation to obey:in all things ``f10!} submit\index{Holy Spirit:subdues will
of man ``f10!} to his will in all things,\footnote{\ Psalm 19:14\index{19 Psalms:ch.  19, v.  14, pg. ``f30!}; Psalm
119; 1 Thessalonians 5:23\index{52 1 Thessalonians:ch.   5, v.  23, pg. ``f30!}; Hebrews 13:20-21\index{58 Hebrews:ch. 
13, v.  20-21, pg. ``f30!}.} as the angels\index{Angels:angels ``f10!} do in heaven\index{Angels:works in heaven
``f10!}.\footnote{\ Psalm 103:20-21\index{19 Psalms:ch. 103, v.  20-21, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.  
1, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   1, v.  14, pg. ``f30!}.}


\bigskip

Q. 104 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the fourth
petition}\index{Prayer:Lord's:petitions:fourth ``f10!}\textit{?}

A. In the fourth petition\index{Prayer:Lord's:petitions:fourth ``f10!}, which is, Give us this day our daily
bread\index{Prayer:for daily needs ``f10!}, we pray that of God's free gift we\index{Providence:provides good gifts
``f10!} may receive a competent portion\index{Prayer:for necessities ``f10!} of the good things of this life, and enjoy
his blessing with them.\footnote{\ Proverbs 30:8-9\index{20 Proverbs:ch.  30, v.   8-9, pg. ``f30!}; Matthew
6:31-34\index{40 Matthew:ch.   6, v.  31-34, pg. ``f30!}; Philippians 4:11\index{50 Philippians:ch.   4, v.  11, pg.
``f30!}, 19\index{50 Philippians:ch.   4, v.  11, 19, pg. ``f30!}; 1 Timothy 6:6\index{54 1 Timothy:ch.   6, v.   6,
pg. ``f30!}{}-8\index{54 1 Timothy:ch.   6, v.   6-8, pg. ``f30!}.}


\bigskip

Q. 105 \textit{What do we pray for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the fifth
petition}\index{Prayer:Lord's:petitions:fifth ``f10!}\textit{?}

A. In the fifth petition\index{Prayer:Lord's:petitions:fifth ``f10!}, which is, And forgive us\index{Forgiveness:beg
pardon for ``f10!} our debts\index{Forgiveness:of debtors ``f10!}, as we forgive our debtors, we pray that
God\index{Prayer:for pardon of sin ``f10!}, for Christ's sake, would freely\index{Salvation:freely offered ``f10!}
pardon all our sins;\footnote{\ Psalm 51:1\index{19 Psalms:ch.  51, v.   1, pg. ``f30!}{}-2, 7, 9\index{19 Psalms:ch. 
51, v.   1-2, 7, 9, pg. ``f30!}; Daniel 9:17\index{27 Daniel:ch.   9, v.  17, pg. ``f30!}{}-19\index{27 Daniel:ch.   9,
v.  17-19, pg. ``f30!}; 1 John 1:7\index{62 1 John:ch.   1, v.   7, pg. ``f30!}.} which we are the rather
encouraged\index{Prayer:assurance of being heard ``f10!} to ask, because by his grace\index{Grace:God's free ``f10!} we
are enabled from\index{Forgiveness:enabled by Holy Spirit ``f10!} the heart to forgive others.\footnote{\ Matthew
18:21-35\index{40 Matthew:ch.  18, v.  21-35, pg. ``f30!}; Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg.
``f30!}2\index{49 Ephesians:ch.   4, v.  32, pg. ``f30!}; Colossians 3:13\index{51 Colossians:ch.   3, v.  13, pg.
``f30!}.}


\bigskip

Q. 106 \textit{What do we }pray\textit{ for in}\index{Prayer:Lord's:petitions:items to pray for ``f10!}\textit{ the
sixth petition}\index{Prayer:Lord's:petitions:sixth ``f10!}\textit{?}

A. In the sixth petition\index{Prayer:Lord's:petitions:sixth ``f10!}, which is, And lead us not into
temptation\index{Temptation:temptation ``f10!}, but deliver us from\index{Salvation:delivered from thraldom to sin
``f10!} evil, we pray, that\index{Prayer:Lord's:petitions:items to pray for ``f10!} God would either keep us from being
tempted to sin,\footnote{\ Psalm 19:13\index{19 Psalms:ch.  19, v.  13, pg. ``f30!}; Matthew 26:41\index{40 Matthew:ch.
 26, v.  41, pg. ``f30!}; John 17:15\index{43 John:ch.  17, v.  15, pg. ``f30!}.} or support and deliver us
when\index{Temptation:delivery from ``f10!} we are tempted.\footnote{\ Luke 22:31-32\index{42 Luke:ch.  22, v.  31-32,
pg. ``f30!}; 1 Corinthians 10:13\index{46 1 Corinthians:ch.  10, v.  13, pg. ``f30!}; 2 Corinthians 12:7-9\index{47 2
Corinthians:ch.  12, v.   7-9, pg. ``f30!}; Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}8\index{58
Hebrews:ch.   2, v.  18, pg. ``f30!}.}


\bigskip

Q. 107 \textit{What }does \textit{the conclusion}\index{Prayer:Lord's:conclusion ``f10!}\textit{ of the Lord's
Prayer}\index{Prayer:Lord's:The Lord's Prayer ``f10!}\textit{ teach us?}

A. The conclusion\index{Prayer:Lord's:conclusion ``f10!} of the Lord's Prayer\index{Prayer:Lord's:The Lord's Prayer
``f10!}, which is, For thine is the kingdom, and the\index{Kingdom of God:prayer ascribing power and glory ``f10!}
power, and the glory, for\index{Prayer:for glory of God ``f10!} ever, Amen\index{Amen, testimony of our true desire for
God's glory ``f10!}. teaches us to take our encouragement in\index{Prayer:drawing encouragement ``f10!} prayer from God
only,\footnote{\ Daniel 9:4\index{27 Daniel:ch.   9, v.   4, pg. ``f30!}, 7-9, 16-19\index{27 Daniel:ch.   9, v.   4,
7-9, 16-19, pg. ``f30!}; Luke 18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}, 7-8\index{42 Luke:ch.  18, v.   1, 7-8,
pg. ``f30!}.} and in our prayers\index{Prayer:praising God ``f10!} to praise him, ascribing kingdom, power, and glory
to him;\footnote{\ 1 Chronicles 29:10-13\index{13 1 Chronicles:ch.  29, v.  10-13, pg. ``f30!}; 1 Timothy 1:17\index{54
1 Timothy:ch.   1, v.  17, pg. ``f30!}; Revelation 5:11-13\index{66 Revelation:ch.   5, v.  11-13, pg. ``f30!}.} and,
in testimony of our desire, and\index{Prayer:offering desires to God ``f10!} assurance to be\index{Prayer:assurance of
being heard ``f10!} heard, we say, Amen.\footnote{\ 1 Corinthians 14:16\index{46 1 Corinthians:ch.  14, v.  16, pg.
``f30!}; Revelation 22:20\index{66 Revelation:ch.  22, v.  20, pg. ``f30!}.}
\end{document}
