\chapter{Official Vows}

\section[1. Vows of Church Membership]{1. Vows of Church Membership} \index{Church membership:vows:official ``f20!}

\begin{enumerate}
  \item Do you believe the Scriptures\index{Church membership:vows:believe the Scriptures ``f20!} of the Old and\index{Bible:Old and New Testaments ``f20!} New Testaments to be the Word\index{Bible:Word of God:both Testaments ``f20!} of God, the only infallible\index{Bible:Word of God:only infallible rule for faith and life ``f20!} rule for faith and life?
  \item Do you believe in the one living and true God---Father\index{Church membership:vows:listed ``f20!}, Son\index{Jesus:Son ``f20!}, and Holy Spirit\index{Holy Spirit ``f20!}, as revealed in the Scriptures\index{Bible:Word of God:standard:for the Church ``f20!}?
  \item Do you repent\index{repentance ``f20!} of your sin\index{Sin:admonishment to:repent ``f20!}; confess your guilt and helplessness as a sinner against\index{Sin:confession of ``f20!} God; profess Jesus\index{profession of faith ``f20!} Christ, Son\index{Jesus:Son ``f20!} of God, as your Saviour\index{Jesus:Saviour ``f20!} and Lord; and\index{Church membership:vows:Jesus as Saviour and Lord ``f20!} dedicate yourself to his service\index{Church membership:vows:dedication to Christ's service ``f20!}: Do you promise that you will endeavour to forsake\index{Church membership:vows:forsake sin ``f20!} all sin, and to conform your life to his teaching and example?
  \item To the end that you may grow in the Christian life, do you promise that you will:

    \begin{enumerate} 
    \item Diligently read the Bible, engage\index{Worship:private:Bible reading ``f20!} in private prayer\index{Prayer:private ``f20!}, keep the Lord's Day\index{Sabbath:to be kept holy ``f20!}, participate in public worship\index{Worship:public:Lord's Day ``f20!}, observe the sacraments, and\index{Sacraments:to be observed by the Church ``f20!} give back to the Lord as he\index{Worship:public:tithes:biblical requirement ``f20!} biblically requires?
    \item Seek first the kingdom\index{Kingdom of God/Christ ``f20!} of God and his righteousness in\index{Righteousness:God's ``f20!} all the relationships of life, faithfully to perform\index{Christian:duty:walk:faithfully ``f20!} your whole duty as\index{Christian:duty:vow to perform ``f20!} a true servant\index{Pastor:servant of Christ ``f20!} of Jesus Christ, and seek to win\index{Office bearers:draw souls to Christ ``f20!} others to him? 
    \item Accept your responsibility to work with others in the Church, and do you promise to support and encourage them in their service to the Lord\index{Church membership:vows:encourage others in service to the Lord ``f20!}?
    \item Submit in the Lord to\index{Church membership:vows:submit in the Lord ``f20!} the teaching and government of this Church as being based upon the Scriptures\index{Church government:must conform to the Scriptures ``f20!} and set forth in the Constitution\index{Constitution:constitution ``f20!} of the Reformed Presbyterian Church of Canada?
    \item Respect the authority and discipline\index{Church discipline:pastoral ``f20!} of the Church, should you need correction\index{Church discipline:respect authority in correction in doctrine or life ``f20!} in doctrine or life?
    \end{enumerate}

  \item Do you make this profession of faith\index{profession of faith ``f20!} and state these promises\index{Church membership:vows:promises ``f20!} in the presence of God, in humble\index{humility ``f20!} reliance upon his grace\index{Grace of God:Christians' reliance upon ``f20!}, as you desire to give your account with joy at the Last Great Day?

\end{enumerate}

\section[2. Parental/Legal Guardian Baptismal Vows]{2. Parental/Legal Guardian Baptismal Vows} \index{Baptism:vows:parent ``f20!}\index{Baptism:guardians in place of parents ``f20!}

\begin{enumerate} 
  \item Do you believe this child is God's possession entrusted to your care? Do you promise therefore to provide for his/her physical and emotional wellbeing? 
  \item Do you promise to teach\index{Parents:teach children:God's truth ``f20!} him/her to love God and his Word\index{Bible:Word of God:to be loved ``f20!}, and to provide him/her with a God-centered education\index{Parents:provide God-centered education ``f20!}? 
  \item Do you promise to teach\index{Parents:teach children:God's truth ``f20!} him/her of his/her sinful nature\index{Sinful:nature ``f20!}, of the plan of\index{Salvation:plan of ``f20!} salvation\index{Salvation:salvation ``f20!} which centers on Jesus Christ, of his/her own personal need of a relationship with Christ, and the need for daily repentance\index{repentance ``f20!} of sin? 
  \item To the end that he/she may grow in the Christian life, do you promise to pray\index{Parents:promise to pray for baptized child ``f20!} for him/her, and to train him/her to read the Bible, to\index{Parents:teach children:to read Bible ``f20!} pray, to\index{Prayer:promise to engage in ``f20!} keep the Lord's Day\index{Sabbath:to be kept holy ``f20!}, and to understand the nature of the Church, the value of its worship\index{Worship:value ``f20!} and fellowship, and\index{Congregation:fellowship ``f20!} his/her need to seek communicant\index{Church membership:communicant ``f20!} membership in the Church? 
  \item Do you promise to lead him/her by your example and parental discipline\index{Church discipline:pastoral ``f20!}, exercised in love, to seek first the Kingdom\index{Kingdom of God/Christ ``f20!} of God and his righteousness in\index{Righteousness:God's ``f20!} all his/her relationships? 
  \item Do you make these promises\index{Baptism:vows:promises ``f20!} in the presence of God, in humble\index{humility ``f20!} reliance upon his grace\index{Grace of God:Christians' reliance upon ``f20!}, as you desire to give your account with joy at the Last Great Day?
\end{enumerate}

\section[3. Statement Concerning Subscription]{3. Statement Concerning Subscription} \index{Office bearers:subscription:requirements ``f20!}

Subscribing\index{Office bearers:subscription:before God ``f20!}, in the presence of Almighty God, to the \textit{Doctrinal Standards}\index{Office bearers:subscription:doctrinal standards ``f20!} of the Reformed Presbyterian Church of Canada is a solemn act and\index{Office bearers:subscription:solemn act ``f20!} should only be entered into following prayerful deliberation.   A man who has any doubts or scruples\index{Office bearers:subscription:scruples ``f20!} about subscribing to the \textit{Doctrinal Standards}\index{Office bearers:subscription:doctrinal standards ``f20!} of the Reformed Presbyterian Church of Canada shall not be admitted to a Church office. No\index{Office bearers:subscription:changed views concerning ``f20!} man is at liberty to subscribe to doctrines\index{Office bearers:subscription:only to doctrines believed ``f20!} which he does not truly believe.  If during the fulfilment of his office a\index{Office bearers:subscription:changed views concerning ``f20!} man alters his views concerning any part of the doctrine, worship, or order of the Reformed Presbyterian Church of Canada, he must:

\begin{enumerate} 
  \item Reflect on the fact that no church constitution will\index{Constitution:cannot express every biblical truth ``f20!} express every biblical truth in a way that will satisfy everyone, and that the purest churches are subject to error. 
  \item Decide if he can conscientiously\index{conscience ``f20!} remain silent about them. If he can, then he may remain in peace. However, if he cannot and feels compelled to speak, publish, or propagate them, then he must inform the session and the presbytery immediately with a view to peaceably joining\index{Pastor:credentials:transfer:to another denomination ``f20!} another branch\index{Church:branches of the visible ``f20!} of the visible\index{Church:visible ``f20!} Church, with which he can walk harmoniously\index{Christian:duty:walk:harmoniously ``f20!}.
\end{enumerate} 

\par Date: [\textit{date}]  
\\ Signature: [\textit{signature}\index{signatures on documents ``f20!}] 


\subsection[Subscription Exceptions]{\textbf{Subscription Exceptions}} \index{Office bearers:subscription:requirements ``f20!}\index{Office bearers:subscription:exceptions ``f20!}

\textbf{The RP Church of Canada permits her office bearers}\index{Office bearers:office-bearers ``f20!}\textbf{, if they deem it necessary, to hold to two exceptions to the }\textbf{\textit{Westminster Confession}}\index{Westminster:Confession of Faith ``f20!}\textbf{\textit{ of Faith}}\textbf{:}

\textit{Exception}\index{Office bearers:subscription:exceptions ``f20!}\textit{ 1 }\\Chapter 24:4 of the \textit{Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ s}tates that, ``The man may not marry any of his wife\index{wives ``f20!}{'s kindred}\index{Marriage:consanguinity ``f20!} nearer in blood than he may of his own; nor the woman of her husband's kindred nearer in blood than of her own.'' In view of the uncertainty among many students of Scripture as to the true interpretation of the injunction\index{Marriage:wife's kindred ``f20!} laid down in Leviticus 18:18\index{03 Leviticus:ch.  18, v.  18, pg. ``f30!}, any office bearer\index{Office bearers:office-bearers ``f20!} having reservations\index{Office bearers:subscription:reservations ``f20!} on this subject may take exception\index{Office bearers:subscription:exceptions ``f20!} to subscribing to this clause.  

\textit{Exception}\index{Office bearers:subscription:exceptions ``f20!}\textit{ 2}\\Chapter 25:6 of the \textit{Confession}\index{Westminster:Confession of Faith ``f20!} states that, ``The Pope\index{Pope of Rome, antichrist ``f20!} of Rome ... is that Antichrist, that man of sin\index{Sin:man of ``f20!}, and son of perdition, that exalts himself, in the Church, against Christ and all that is called God.'' In view of the manifest uncertainty in interpreting prophetic\index{Bible:prophetic passages, interpretation ``f20!} passages of Scripture, any office bearer\index{Office bearers:office-bearers ``f20!} having reservations\index{Office bearers:subscription:reservations ``f20!} about specifically identifying the papacy with the antichrist rather than an antichrist, may take an exception\index{Office bearers:subscription:exceptions ``f20!} to this clause.

\subsection[Subscription Clarifications]{\textbf{Subscription}\textbf{ Clarifications}} \index{Office bearers:subscription:requirements ``f20!}

\textbf{Subscription}\index{Office bearers:subscription:requirements ``f20!}\textbf{ to the confessional standards of the RP Church of Canada requires the acceptance of the following clarifying}\index{Office bearers:subscription:clarifying statements ``f20!}\textbf{ statements to the }\textbf{\textit{Westminster Confession}}\index{Westminster:Confession of Faith ``f20!}\textbf{\textit{ of Faith}}\textbf{:}

\textit{Clarification 1}\\The RP Church of Canada maintains that the phrase ``in the space of six days\index{Creation:six natural days ``f20!},'' and ``all very good'' in Chapter 4.1 of the \textit{Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ }requires the plain sense interpretation of Genesis. The duration of the creation period was six days of ordinary length, and the state of the original creation prior\index{Creation:original:perfect before curse ``f20!} to the curse was perfect and without death\index{Creation:original:without death ``f20!}.\textsuperscript{ \footnotemark{}}\footnotetext{\ Genesis 1:21, 24\index{01 Genesis:ch.   1, v.  21, 24, pg. ``f30!}; Genesis 2:7\index{01 Genesis:ch.   2, v.   7, pg. ``f30!}.}

\textit{Clarification 2}\\In the \textit{Westminster Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ of Faith:}

\par Chapter 23:3: This portion of article 3 harmonizes with its first sentence, as well as chapters 30 and 31. It refers to a civil magistrate's responsibility to uphold the first table of the moral law\index{moral law ``f20!} and not to support Erastian\index{Church government:Erastian ``f20!} (State\index{Civil magistrates:State ``f20!}) authority, jurisdiction, or interference\index{Civil magistrates:no authority over the Church ``f20!} regarding the affairs of the Church. Furthermore, no part of the Westminster standards countenances persecution\index{Westminster:standards:no persecution for conscience's sake ``f20!} for conscience\index{conscience ``f20!}{'s sake.}\footnote{\ Psalm 2:10-12\index{19 Psalms:ch.   2, v.  10-12, pg. ``f30!}; Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}.}

\par Chapter 31:2: This second article refers to times when the visible\index{Church:visible ``f20!} Church is not settled, or when schisms and factions\index{civil schisms and factions ``f20!} prevail within a nation\index{National:schisms ``f20!}. Civil\index{Civil magistrates:uphold first table of law ``f20!} magistrates are to uphold the first table of the moral law\index{moral law ``f20!}, but they do not have authority or jurisdiction over the\index{Civil magistrates:no jurisdiction over the Church ``f20!} Church. The word `synod\index{Synod:historic councils ``f20!}{' in this section, does not refer to the church court}\index{Synod:church court ``f20!} commonly called the `Synod', but to an advisory committee\index{Church courts:committees:advisory ``f20!}, such as the Council of Nicaea\index{Council of Nicaea ``f20!}, the Synod of Dort\index{Synod of Dort ``f20!}, or the Westminster Assembly\index{Westminster:Assembly ``f20!}. Church government, being distinct from civil government, is always free to assemble as\index{Church courts:assemble as often as required ``f20!} often as needed for the good of the Church.

\par Therefore, we express our reservations\index{Office bearers:subscription:reservations ``f20!} regarding these three sections of the \textit{Westminster Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ of Faith} and understand that they should be interpreted in accordance with the decision of the General Assembly\index{Church courts:General Assembly of Church of Scotland ``f20!} of the Church of Scotland\index{Church of Scotland, decision of General Assembly ``f20!}, upon receiving the \textit{Confession of Faith} in 1647, stated:  

\par {``The Assembly}\index{Westminster:Assembly ``f20!} understandeth some parts of the second article of the thirty-first chapter only of kirks\index{Congregation:not yet constituted ``f20!} not settled or constituted in\index{Church courts:constitution procedures ``f20!} point of Government\index{Church government:Scottish Assembly ``f20!} and that, although in such kirks a Synod\index{Synod:church court ``f20!} of Pastors\index{Pastor:pastor ``f20!}, and other fit persons, may be called by the Magistrate\index{Civil magistrates:request church courts to assemble ``f20!}{'s authority and nomination without any other call, to consult and advise with, about matters of religion}\index{Civil magistrates:Church to advise in matters of religion ``f20!}; and although, likewise, the Pastors of Christ, without delegation\index{Church courts:congregational delegates ``f20!} from their Churches, may of themselves, and by virtue of their office, meet\index{Office bearers:meet together in synods ``f20!} together synodically in such kirks not yet constituted, yet neither of these ought to be done in kirks constituted and settled.''

\par The RP Church of Canada's acceptance of this interpretation does not imply\index{Civil magistrates:no authority over the Church ``f20!} the granting of any authority to the magistrate other than the requesting of pastors and other qualified persons to assemble together\index{Civil magistrates:request church courts to assemble ``f20!}.     

\section[4. Vows for Licensure, Ordination, and Installation]{4. Vows for Licensure, Ordination, and Installation} \index{Licensure:vows ``f20!}\index{Office bearers:ordination:ordination ``f20!}\index{Office bearers:installation ``f20!}

\par {``I [}\textit{name}], do sincerely believe and own all the doctrine\index{Office bearers:subscription:doctrinal standards ``f20!} set forth in the \textit{Westminster Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ of Faith}, and the \textit{Larger Catechism}\index{Westminster:Larger Catechism ``f20!} and \textit{Shorter Catechism}\index{Westminster:Shorter Catechism ``f20!} to be the teaching of the Word\index{Westminster:standards:systematic teaching of God's revelation ``f20!} of God. I declare these as the confession of my faith. I do seriously promise and vow, in the\index{Office bearers:subscription:vows ``f20!} presence of Almighty God, that I will consistently adhere to, maintain, and defend this\index{Office bearers:subscription:vow to defend doctrine ``f20!} doctrine. I also vow to keep and uphold the \textit{Directory for Worship of God} and the \textit{Directory for Church Government}\index{Church government:Directory ``f20!} of the Reformed Presbyterian Church of Canada.

\par If I should ever find myself unable sincerely to own and believe any doctrine\index{Office bearers:subscription:changed views concerning ``f20!} set forth in the \textit{Westminster Confession}\index{Westminster:Confession of Faith ``f20!}\textit{ of Faith}, and the \textit{Larger Catechism}\index{Westminster:Larger Catechism ``f20!} and \textit{Shorter Catechism}\index{Westminster:Shorter Catechism ``f20!}, then I promise, in the presence of Almighty God, that I will not propagate or defend these sentiments, either by preaching\index{Preaching:God's Word ``f20!}, teaching, or writing. I will reveal and explain my new convictions\index{Office bearers:subscription:profession of convictions ``f20!} to the session, the presbytery, or a higher court\index{Church courts:graded ``f20!}, where they may be examined\index{Office bearers:examination for suitability ``f20!}. If necessary, I promise cheerfully to submit to the judgment\index{Office bearers:subscription:vows ``f20!} of the court and/or that\index{Church courts:submission to judgements ``f20!} I will peaceably withdraw\index{Congregation:disjoining ``f20!} and join some\index{Pastor:credentials:transfer:to another denomination ``f20!} other branch\index{Church:branches of the visible ``f20!} of the visible\index{Church:visible ``f20!} Church, with which I can walk harmoniously\index{Christian:duty:walk:harmoniously ``f20!}.''

\par Date: [\textit{date}]
\\Signature: [\textit{signature}\index{signatures on documents ``f20!}]

\par\textit{For the ordination of ruling elders add the following:} \index{Elders:elders ``f20!}

\par So far as I know in my heart, it is the call of Christ, the glory of God\index{Office bearers:service:for glory of God ``f20!}, and the welfare\index{Office bearers:subscription:vows ``f20!} of the Church, and not any selfish object, that has moved me\index{Office bearers:ordination:disavow selfish object ``f20!} to undertake this sacred office\index{Office bearers:sacred duty ``f20!}.

\par I am determined to seek the Holy Spirit\index{Holy Spirit ``f20!}{'s guidance. I}\index{Office bearers:seek guidance from Holy Spirit ``f20!} promise, in his strength, to live a holy\index{Office bearers:subscription:vow to live holy life ``f20!} and exemplary\index{Office bearers:subscription:requirements ``f20!} life, to study and promote the purity, peace, unity\index{Congregation:unity ``f20!}, and progress of the Church, to watch over the spiritual growth of\index{Office bearers:watch over spiritual growth of members ``f20!} the members of the congregation, to endeavour to win\index{Office bearers:draw souls to Christ ``f20!} others to Christ, to visit the afflicted\index{Elders:duties:visit:afflicted ``f20!}, and to attend the meetings\index{Office bearers:attend meetings ``f20!} associated with my office.

\par Date: [\textit{date}]
\\Signature: [\textit{signature}\index{signatures on documents ``f20!}]

\par\textit{For the ordination of pastors/teachers add the following:} \index{Pastor:ordination:vow ``f20!}\index{Teacher:teachers ``f20!}

\par So far as I know in my heart, it is the call of Christ, the glory of God\index{Office bearers:service:for glory of God ``f20!}, and the welfare\index{Office bearers:subscription:vows ``f20!} of the Church, and not any selfish object, that has moved me\index{Office bearers:ordination:disavow selfish object ``f20!} to undertake this sacred office\index{Office bearers:sacred duty ``f20!}.

\par I promise, in the strength of the Holy Spirit\index{Holy Spirit ``f20!}, to live a holy\index{Office bearers:subscription:vow to live holy life ``f20!} and exemplary\index{Office bearers:subscription:requirements ``f20!} life, to study and promote the purity, peace, unity\index{Congregation:unity ``f20!}, and progress of the Church, to watch over the spiritual growth of\index{Office bearers:watch over spiritual growth of members ``f20!} the members of the congregation, to visit the afflicted\index{Elders:duties:visit:afflicted ``f20!}, and to attend the meetings\index{Office bearers:attend meetings ``f20!} associated with my office. I promise, in his strength, to bring to the congregation the fruits of earnest study of the Word\index{Bible:Word of God:to be studied ``f20!}, to watch for souls as\index{Pastor:draw souls to Christ ``f20!} one who must give an account, to maintain a testimony for\index{Office bearers:maintain testimony for God's kingdom ``f20!} God's Kingdom\index{Kingdom of God/Christ ``f20!}, and to endeavour to win\index{Office bearers:draw souls to Christ ``f20!} others to Christ.

\par Date: [\textit{date}]
\\Signature: [\textit{signature}\index{signatures on documents ``f20!}]

\par\textit{For the ordination of deacons add:} \index{Deacons:deacons ``f20!}

\par {``I [name] do sincerely believe and own all the doctrine}\index{Office bearers:subscription:doctrinal standards ``f20!} set forth in the Westminster Confession\index{Westminster:Confession of Faith ``f20!} of Faith, and the Larger Catechism\index{Westminster:Larger Catechism ``f20!} and Shorter Catechism\index{Westminster:Shorter Catechism ``f20!} to be the teaching of the Word\index{Westminster:standards:systematic teaching of God's revelation ``f20!} of God. I declare these as the confession of my faith. I do seriously promise and vow, in the\index{Office bearers:subscription:vows ``f20!} presence of Almighty God, that I will consistently adhere to, maintain, and defend this\index{Office bearers:subscription:vow to defend doctrine ``f20!} doctrine. I also vow to submit in the Lord to\index{Church membership:vows:submit in the Lord ``f20!} the teaching and government of this Church as being based upon the Scriptures\index{Church government:must conform to the Scriptures ``f20!} and set forth in the Constitution\index{Constitution:constitution ``f20!} of the Reformed Presbyterian Church of Canada.

\par If I should ever find myself unable sincerely to own and believe any doctrine\index{Office bearers:subscription:changed views concerning ``f20!} set forth in the Westminster Confession\index{Westminster:Confession of Faith ``f20!} of Faith, and the Larger Catechism\index{Westminster:Larger Catechism ``f20!} and Shorter Catechism\index{Westminster:Shorter Catechism ``f20!}, then I promise, in the presence of Almighty God, that I will not propagate or defend these sentiments, either by preaching\index{Preaching:God's Word ``f20!}, teaching, or writing. I will reveal and explain my new convictions\index{Office bearers:subscription:profession of convictions ``f20!} to the session, the presbytery, or a higher court\index{Church courts:graded ``f20!}, where they may be examined\index{Office bearers:examination for suitability ``f20!}. If necessary, I promise cheerfully to submit to the judgment\index{Office bearers:subscription:vows ``f20!} of the court and/or that\index{Church courts:submission to judgements ``f20!} I will peaceably withdraw\index{Congregation:disjoining ``f20!} and join some\index{Pastor:credentials:transfer:to another denomination ``f20!} other branch\index{Church:branches of the visible ``f20!} of the visible\index{Church:visible ``f20!} Church with which I can walk harmoniously\index{Christian:duty:walk:harmoniously ``f20!}.

\par So far as I know in my heart, it is the call of Christ, the glory of God\index{Office bearers:service:for glory of God ``f20!}, and the welfare\index{Office bearers:subscription:vows ``f20!} of the Church, and not any selfish object, that has moved me\index{Office bearers:ordination:disavow selfish object ``f20!} to undertake this sacred office\index{Office bearers:sacred duty ``f20!}.

\par I am determined to seek the Holy Spirit\index{Holy Spirit ``f20!}{'s guidance. I}\index{Office bearers:seek guidance from Holy Spirit ``f20!} promise, in his strength, faithfully to keep\index{Deacons:financial faithfulness ``f20!} and distribute the tithes\index{Worship:public:tithes:tithes ``f20!} and offerings\index{Worship:public:freewill offerings ``f20!} given to the Church according to the sincere needs of God's people and in the interest of the Kingdom\index{Kingdom of God/Christ ``f20!} of Jesus Christ. And I endeavour to handle the Lord's money\index{Deacons:handle Lord's money in worthy manner ``f20!} without covetousness\index{Deacons:endeavour not to covet ``f20!} and in a manner worthy of Christ's service\index{Deacons:fulfillment of Church's service ``f20!}.

\par Date: [\textit{date}]
\\Signature: [\textit{signature}\index{signatures on documents ``f20!}]
