\subsubsection[The Westminster Confession of Faith of 1647]{{The Westminster Confession of
Faith}

\index{Faith:saving ``f10!}

\section{Chapter 1 -- Of the Holy Scripture} \index{Scripture:holy ``f10!}1.1\ \ Although the light of nature and the works of creation\index{God's  attributes:creator:works of ``f10!} and providence\index{Providence:providence ``f10!} do so far manifest the\index{Providence:manifests goodness of God ``f10!} goodness\index{God's attributes:goodness ``f10!}, wisdom, and power\index{God's attributes:almighty power ``f10!} of God, as to leave men unexcusable\index{Unbelievers:unexcusable ``f10!};\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}; Romans 1:19-20\index{45 Romans:ch.   1, v.  19-20, pg. ``f30!}; Psalm 19:1-3\index{19 Psalms:ch.  19, v.   1-3, pg. ``f30!}; Romans 1:32\index{45 Romans:ch.   1, v.  32, pg. ``f30!} with Romans 2:1.} yet are they not sufficient to\index{Covenant:OT ordinances:sufficient for that time ``f10!} give that knowledge of God\index{Revelation:natural:not sufficient knowledge of God ``f10!} and of His will, which is necessary unto salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Corinthians 1:2\index{46 1 Corinthians:ch.   1, v.   2, pg. ``f30!}1\index{46 1 Corinthians:ch.   1, v.  21, pg. ``f30!}; 1 Corinthians 2:13\index{46 1 Corinthians:ch.   2, v.  13, pg. ``f30!}{}-14\index{46 1 Corinthians:ch.   2, v.  13-14, pg. ``f30!}.} Therefore it pleased\index{Providence:at God's pleasure ``f10!} the Lord, at sundry times, and in divers manners\index{Revelation:sundry times and divers manners ``f10!}, to reveal Himself, and to declare that\index{Revelation:special declaration ``f10!} His will unto His Church\index{Church:Christ's ``f10!};\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}.} and afterwards, for the better preserving\index{Church:to preserve truth ``f10!} and propagating\index{Scripture:for propagating truth ``f10!} of the truth, and\index{Oaths and Vows:assertion of truth ``f10!} for the more sure establishment\index{Scripture:for establishment and comfort of Church ``f10!} and comfort of the Church against\index{Church:institution ``f10!} the corruption of the\index{Sin:original:corrupt nature:flesh ``f10!} flesh, and the\index{flesh, sinful nature ``f10!} malice of\index{Satan:malice of ``f10!} Satan and of\index{Satan:tempter ``f10!} the world, to commit the same wholly unto writing:\index{Scripture:written ``f10!}\footnote{\ Proverbs 22:1\index{20 Proverbs:ch.  22, v.   1, pg. ``f30!}9-21\index{20 Proverbs:ch.  22, v.  19-21, pg. ``f30!}; Luke 1:3-4\index{42 Luke:ch.   1, v.   3-4, pg. ``f30!}; Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}; Matthew 4:4\index{40 Matthew:ch.   4, v.   4, pg. ``f30!}, 7, 10\index{40 Matthew:ch.   4, v.   4, 7, 10, pg. ``f30!}; Isaiah 8:19-20\index{23 Isaiah:ch.   8, v.  19-20, pg. ``f30!}.} which makes the Holy Scripture\index{Scripture:holy ``f10!} to be most necessary;\footnote{\ 2 Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}; 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg. ``f30!}.} those former ways of God's revealing His\index{Revelation:God's will ``f10!} will unto His people being now ceased\index{Scripture:other forms of revelation have ceased ``f10!}.\footnote{\ Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}{}-2\index{58 Hebrews:ch.   1, v.   1-2, pg. ``f30!}.}  
\bigskip
1.2\ \ Under the name of Holy Scripture\index{Scripture:holy ``f10!}, or the Word of God\index{Scripture:word of God:word of God ``f10!} written, are now contained all the books of the Old and New Testament\index{Scripture:New Testament ``f10!}, which are these\index{Scripture:Old and New Testaments ``f10!}:  
\bigskip
{\centering \textit{Of the Old Testament}\index{Scripture:Old Testament ``f10!}\textit{:} \par}  \begin{multicols}{3} Genesis  Exodus  Leviticus  Numbers  Deuteronomy  Joshua  Judges  Ruth  1 Samuel  2 Samuel  1 Kings  2 Kings  1 Chronicles  2 Chronicles  Ezra  Nehemiah  Esther  Job  Psalms  Proverbs  Ecclesiastes  Song of Songs  Isaiah  Jeremiah  Lamentations  Ezekiel  Daniel  Hosea  Joel  Amos  Obadiah  Jonah  Micah  Nahum  Habakkuk  Zephaniah  Haggai  Zechariah  Malachi  \end{multicols}  
\bigskip
{\centering \textit{Of the New Testament}\index{Scripture:New Testament ``f10!}\textit{:} \par} \begin{multicols}{3} Matthew  Mark  Luke  John  Acts  Romans  1 Corinthians  2 Corinthians  Galatians  Ephesians  Philippians  Colossians  1 Thessalonians  2 Thessalonians  1 Timothy  2 Timothy  Titus  Philemon  Hebrews  James  1 Peter  2 Peter  1 John  2 John  3 John  Jude  Revelation \end{multicols}  
\bigskip
All which are given by inspiration of\index{Scripture:inspired by God ``f10!} God, to be the rule of faith\index{Faith:rule of ``f10!} and life.\footnote{\ Luke 16:29, 31\index{42 Luke:ch.  16, v.  29, 31, pg. ``f30!}; Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!}; Revelation 22:18-19\index{66 Revelation:ch.  22, v.  18-19, pg. ``f30!}; 2 Timothy 3:16\index{55 2 Timothy:ch.   3, v.  16, pg. ``f30!}.}   
\bigskip
1.3\ \ The books commonly called Apocrypha\index{Apocrypha:not God breathed ``f10!}, not being of divine inspiration, are no part of the canon\index{Scripture:canon ``f10!} of the Scripture; and\index{Scripture:Apocrypha not part of canon ``f10!} therefore are of no authority in the Church of God\index{Church:God's ``f10!}, nor to be any otherwise approved, or made use of, than other human writings\index{Apocrypha:human writings ``f10!}.\footnote{\ Luke 24:27, 44\index{42 Luke:ch.  24, v.  27, 44, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}; 2 Peter 1:2\index{61 2 Peter:ch.   1, v.   2, pg. ``f30!}1\index{61 2 Peter:ch.   1, v.  21, pg. ``f30!}.}   
\bigskip
1.4\ \ The authority of the Holy Scripture\index{Scripture:holy ``f10!}, for which it ought to be believed and\index{Scripture:to be:believed ``f10!} obeyed\index{Scripture:to be:obeyed:because word of God ``f10!}, depends not upon the testimony\index{Scripture:authority not from testimony of any man ``f10!} of any man, or Church\index{Church:institution ``f10!}; but wholly upon God (who is truth\index{God's attributes:truth ``f10!} itself) the author thereof\index{Scripture:authored by God ``f10!}: and therefore it is to be received because it is the Word of God\index{Scripture:word of God:word of God ``f10!}.\footnote{\ 2 Peter 1:19\index{61 2 Peter:ch.   1, v.  19, pg. ``f30!}, 21\index{61 2 Peter:ch.   1, v.  19, 21, pg. ``f30!}; 2 Timothy 3:16\index{55 2 Timothy:ch.   3, v.  16, pg. ``f30!}; 1 John 5:9\index{62 1 John:ch.   5, v.   9, pg. ``f30!}; 1 Thessalonians 2:13\index{52 1 Thessalonians:ch.   2, v.  13, pg. ``f30!}.}  
\bigskip
1.5\ \ We may be moved and induced by the testimony\index{Scripture:testimony of the Church ``f10!} of the Church to\index{Church:institution ``f10!} a high and reverent\index{Scripture:to be:treated with reverent esteem ``f10!} esteem of the\index{Scripture:to be:esteemed ``f10!} Holy Scripture\index{Scripture:holy ``f10!}.\footnote{\ 1 Timothy 3:1\index{54 1 Timothy:ch.   3, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   3, v.  15, pg. ``f30!}.} And the heavenliness\index{Scripture:heavenly matter ``f10!} of the matter, the efficacy\index{Scripture:efficacy of doctrine ``f10!} of the doctrine, the\index{Doctrine:of Scripture ``f10!} majesty of the\index{Scripture:majesty of the style ``f10!} style, the consent of all\index{Scripture:consent of all parts ``f10!} the parts, the scope of the whole (which is, to give all glory\index{Scripture:gives glory to God ``f10!} to God), the full discovery\index{Scripture:makes full discovery of the only way of salvation ``f10!} it makes of the only way of man's salvation\index{Salvation:belief in Jesus by faith ``f10!}, the many other incomparable excellencies\index{Scripture:excellency ``f10!}, and the entire perfection thereof\index{Scripture:perfection of ``f10!}, are arguments whereby it does abundantly evidence itself to be the Word of God\index{Scripture:word of God:word of God ``f10!}: yet notwithstanding, our full persuasion\index{Scripture:Holy Spirit persuades us that it is word of God ``f10!} and assurance of the\index{Assurance:that God's word is true ``f10!} infallible truth\index{Scripture:infallible truth ``f10!} and divine authority\index{Scripture:divine authority ``f10!} thereof, is from the inward\index{Scripture:witnessed inwardly by Holy Spirit ``f10!} work of the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!} bearing witness by\index{Scripture:witness to heart of man ``f10!} and with the Word in our hearts\index{Heart:new:through ministry of Word ``f10!}.\footnote{\ 1 John 2:2\index{62 1 John:ch.   2, v.   2, pg. ``f30!}0, 27\index{62 1 John:ch.   2, v.  20, 27, pg. ``f30!}; John 16:13-14\index{43 John:ch.  16, v.  13-14, pg. ``f30!}; 1 Corinthians 2:10, 11-12\index{46 1 Corinthians:ch.   2, v.  10, 11-12, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}1\index{23 Isaiah:ch.  59, v.  21, pg. ``f30!}.}   
\bigskip
1.6\ \ The whole counsel of God\index{Causation:God's own will ``f10!} concerning all things necessary for His own glory\index{God's attributes:glory:his own ``f10!}, man's salvation\index{Salvation:belief in Jesus by faith ``f10!}, faith, and life, is either expressly set\index{Scripture:interpretation:some things expressly set down ``f10!} down in Scripture, or by good and necessary consequence\index{Scripture:interpretation:deduced by good and necessary consequence ``f10!} may be deduced from Scripture: unto which nothing at any time is to be added, whether by new revelations\index{Revelation:new, not added to Scripture ``f10!} of the Spirit, or\index{Revelation:Word and Spirit ``f10!} traditions\index{Scripture:supersedes traditions of men ``f10!} of men.\footnote{\ 2 Timothy 3:15\index{55 2 Timothy:ch.   3, v.  15, pg. ``f30!}{}-17\index{55 2 Timothy:ch.   3, v.  15-17, pg. ``f30!}; Galatians 1:8-9\index{48 Galatians:ch.   1, v.   8-9, pg. ``f30!}; 2 Thessalonians 2:2\index{53 2 Thessalonians:ch.   2, v.   2, pg. ``f30!}.} Nevertheless we acknowledge the inward illumination\index{Conversion:Holy Spirit:provides inner illumination ``f10!} of the Spirit of God\index{Holy Spirit:God's Spirit ``f10!} to be necessary for the saving understanding\index{Salvation:by working of Holy Spirit ``f10!} of such things as are revealed in the Word:\index{Scripture:word of God:standard for:life ``f10!}\footnote{\ John 6:45\index{43 John:ch.   6, v.  45, pg. ``f30!}; 1 Corinthians 2:9-12\index{46 1 Corinthians:ch.   2, v.   9-12, pg. ``f30!}.} and that there are some circumstances\index{Worship:circumstances ``f10!} concerning the worship of God, and\index{Worship:correct:instituted by God ``f10!} government of\index{Church government:government ``f10!} the Church, common\index{Church:institution ``f10!} to human actions and societies\index{Scripture:interpretation:circumstances of societies ``f10!}, which are to be ordered\index{Scripture:interpretation:by light of nature according rules of Word ``f10!} by the light of nature and Christian prudence, according\index{Scripture:interpretation:by prudence according rules of Word ``f10!} to the general rules of the Word, which are always to be observed\index{Scripture:interpretation:general rules to be observed ``f10!}.\footnote{\ 1 Corinthians 11:13-14\index{46 1 Corinthians:ch.  11, v.  13-14, pg. ``f30!}; 1 Corinthians 14:26\index{46 1 Corinthians:ch.  14, v.  26, pg. ``f30!}, 40\index{46 1 Corinthians:ch.  14, v.  26, 40, pg. ``f30!}.}  
\bigskip
1.7\ \ All things in Scripture are not alike plain in\index{Scripture:not all parts plain in themselves ``f10!} themselves, nor alike clear unto all:\footnote{\ 2 Peter 3:16\index{61 2 Peter:ch.   3, v.  16, pg. ``f30!}.} yet those things which are necessary to be known, believed, and\index{Scripture:to be:believed ``f10!} observed for salvation\index{Salvation:belief in Jesus by faith ``f10!}, are so clearly propounded\index{Scripture:what necessary to know, clearly propounded ``f10!} and opened in some place of Scripture or\index{Scripture:interpretation:scripture with scripture ``f10!} other, that not only the learned\index{Scripture:useful for learned and unlearned ``f10!}, but the unlearned, in a due use of the ordinary means, may\index{Scripture:understandable:through ordinary means ``f10!} attain unto a sufficient\index{Scripture:understandable:by unlearned ``f10!} understanding of them.\footnote{\ Psalm 119:105\index{19 Psalms:ch. 119, v. 105, pg. ``f30!}, 130\index{19 Psalms:ch. 119, v. 105, 130, pg. ``f30!}.}  
\bigskip
1.8\ \ The Old Testament\index{Scripture:Old Testament ``f10!} in Hebrew\index{Scripture:Old Testament Hebrew ``f10!} (which was the native language of the people of God of old), and the New Testament\index{Scripture:New Testament ``f10!} in Greek\index{Scripture:New Testament Greek ``f10!} (which, at the time of the writing of\index{Scripture:written ``f10!} it was most generally known to the nations\index{Scripture:written in languages known to nations ``f10!}), being immediately inspired\index{Scripture:inspired by God ``f10!} by God, and, by His singular care and providence\index{Providence:providence ``f10!} kept pure in\index{Scripture:kept pure in all ages ``f10!} all ages, are therefore authentical\index{Scripture:original language:versions authentic ``f10!};\footnote{\ Matthew 5:18\index{40 Matthew:ch.   5, v.  18, pg. ``f30!}.} so as, in all controversies\index{Scripture:determine all controversies ``f10!} of religion, the Church is\index{Church:institution ``f10!} finally to appeal unto them.\footnote{\ Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg. ``f30!} ,Acts 15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}; John 5:39\index{43 John:ch.   5, v.  39, pg. ``f30!}, 46\index{43 John:ch.   5, v.  39, 46, pg. ``f30!}.} But, because these original tongues\index{Scripture:original language:not known to all ``f10!} are not known to all the people of God, who have right unto, and interest in the\index{Scripture:all people have interest (share) in ``f10!} Scriptures, and are commanded, in the\index{Scripture:to be:read and heard ``f10!} fear of\index{Fear:of God:reverence ``f10!} God, to read and search the\index{Heart:God alone able to search ``f10!}m\index{Scripture:to be:searched ``f10!},\footnote{\ John 5:39\index{43 John:ch.   5, v.  39, pg. ``f30!}.} therefore they are to be translated\index{Scripture:to be:translated into common (vulgar) languages ``f10!} into the vulgar language of every nation unto which they come,\footnote{\ 1 Corinthians 14:6, 9, 11-12, 24, 27-28\index{46 1 Corinthians:ch.  14, v.   6, 9, 11-12, 24, 27-28, pg. ``f30!}.} that the Word of God\index{Scripture:word of God:word of God ``f10!} dwelling plentifully in all, they may worship Him\index{Worship:acceptable manner ``f10!} in an acceptable manner;\footnote{\ Colossians 3:16\index{51 Colossians:ch.   3, v.  16, pg. ``f30!}.} and, through patience and comfort of the Scriptures\index{Scripture:comforts and builds up believers ``f10!}, may have hope\index{Scripture:provides hope ``f10!}.\footnote{\ Romans 15:4\index{45 Romans:ch.  15, v.   4, pg. ``f30!}.}  
\bigskip
1.9\ \ The infallible\index{Scripture:interpretation:scripture with scripture ``f10!} rule of interpretation of Scripture is the Scripture itself: and therefore, when there is a question about the true and full sense of any Scripture (which is not manifold, but\index{Scripture:interpretation:not manifold, but on sense ``f10!} one), it must be searched\index{Scripture:to be:searched ``f10!} and known by other places that speak more clearly.\footnote{\ 2 Peter 1:2\index{61 2 Peter:ch.   1, v.   2, pg. ``f30!}0-21\index{61 2 Peter:ch.   1, v.  20-21, pg. ``f30!}; Acts 15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}{}-16\index{44 Acts:ch.  15, v.  15-16, pg. ``f30!}.}  
\bigskip
1.10\ \ The supreme judge by\index{Scripture:supreme judge of controversies ``f10!} which all controversies\index{Scripture:determine all controversies ``f10!} of religion are to be determined, and all decrees\index{Church councils:decrees ``f10!} of councils\index{Church government:councils ``f10!}, opinions of\index{Scripture:final authority, not opinions of ancient writers ``f10!} ancient writers, doctrines\index{Doctrine:false:of men ``f10!} of men, and private spirits, are to be examined\index{Scripture:to be:used to examine all doctrines ``f10!}; and in whose sentence we are to rest; can be no other but the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!} speaking in the Scripture\index{Scripture:word of God:Holy Spirit speaking ``f10!}.\footnote{\ Matthew 22:29, 31\index{40 Matthew:ch.  22, v.  29, 31, pg. ``f30!}; Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!} with Acts 28:25\index{44 Acts:ch.  28, v.  25, pg. ``f30!}.}  

\section{Chapter 2 -- Of God, and of the Holy Trinity} \index{God's attributes:trinity:one substance ``f10!} 2.1\ \ There is but one only,\footnote{\ Deuteronomy 6:4\index{05 Deuteronomy:ch.   6, v.   4, pg. ``f30!}; 1 Corinthians 8:4, 6\index{46 1 Corinthians:ch.   8, v.   4, 6, pg. ``f30!}.} living, and true God:\footnote{\ 1 Thessalonians 1:9\index{52 1 Thessalonians:ch.   1, v.   9, pg. ``f30!}; Jeremiah 10:10\index{24 Jeremiah:ch.  10, v.  10, pg. ``f30!}.} who is infinite\index{God's attributes:infinite:knowledge ``f10!} in being and perfection\index{God's attributes:perfect blessedness ``f10!},\footnote{\ Job 11:7-9\index{18 Job:ch.  11, v.   7-9, pg. ``f30!}; Job 26:14\index{18 Job:ch.  26, v.  14, pg. ``f30!}.} a most pure\index{God's attributes:a spirit, most pure ``f10!} spirit,\footnote{\ John 4:24\index{43 John:ch.   4, v.  24, pg. ``f30!}.} invisible,\footnote{\ 1 Timothy 1:17\index{54 1 Timothy:ch.   1, v.  17, pg. ``f30!}.} without body, parts,\footnote{\ Deuteronomy 4:15-16\index{05 Deuteronomy:ch.   4, v.  15-16, pg. ``f30!}; John 4:24\index{43 John:ch.   4, v.  24, pg. ``f30!} with Luke 24:39\index{42 Luke:ch.  24, v.  39, pg. ``f30!}.} or passions\index{God's attributes:no passions ``f10!},\footnote{\ Acts 14:11, 15\index{44 Acts:ch.  14, v.  11, 15, pg. ``f30!}.} immutable\index{God's attributes:immutable ``f10!},\footnote{\ James 1:17\index{59 James:ch.   1, v.  17, pg. ``f30!}; Malachi 3:6\index{39 Malachi:ch.   3, v.   6, pg. ``f30!}.} immense\index{God's attributes:immense ``f10!},\footnote{\ 1 Kings 8:27\index{11 1 Kings:ch.   8, v.  27, pg. ``f30!}; Jeremiah 23:23-24\index{24 Jeremiah:ch.  23, v.  23-24, pg. ``f30!}.} eternal,\index{God's attributes:eternal ``f10!}\footnote{\ Psalm 90:2\index{19 Psalms:ch.  90, v.   2, pg. ``f30!}; 1 Timothy 1:17\index{54 1 Timothy:ch.   1, v.  17, pg. ``f30!}.} incomprehensible\index{God's attributes:incomprehensible ``f10!},\footnote{\ Psalm 145:3\index{19 Psalms:ch. 145, v.   3, pg. ``f30!}.} almighty\index{God's attributes:almighty power ``f10!},\footnote{\ Genesis 17:1\index{01 Genesis:ch.  17, v.   1, pg. ``f30!}; Revelation 4:8\index{66 Revelation:ch.   4, v.   8, pg. ``f30!}.} most wise\index{God's attributes:wisdom ``f10!},\footnote{\index{45 Romans:ch.  16, v.   2, pg. ``f30!} most holy\index{God's attributes:holy:most ``f10!},\footnote{\index{23 Isaiah:ch.   6, v.   3, pg. ``f30!}\index{66 Revelation:ch.   4, v.   8, pg. ``f30!}\index{19 Psalms:ch. 115, v.   3, pg. ``f30!} most absolute\index{God's attributes:absolute ``f10!},\footnote{\index{02 Exodus:ch.   3, v.  14, pg. ``f30!} things according to the counsel of His\index{Causation:God's own will ``f10!} own immutable and most righteous will\index{God's attributes:righteousness ``f10!},\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}.} for His own glory\index{God's attributes:glory:his own ``f10!};\footnote{\ Proverbs 16:4\index{20 Proverbs:ch.  16, v.   4, pg. ``f30!}; Romans 11:36\index{45 Romans:ch.  11, v.  36, pg. ``f30!}.} most loving\index{God's attributes:love ``f10!},\footnote{\ 1 John 4:8, 16\index{62 1 John:ch.   4, v.   8, 16, pg. ``f30!}.} gracious\index{Grace:God's free ``f10!}, merciful\index{God's attributes:merciful ``f10!}, long-suffering\index{God's attributes:long-suffering ``f10!}, abundant in goodness\index{God's attributes:goodness ``f10!} and truth\index{God's attributes:truth ``f10!}, forgiving iniquity\index{God's attributes:forgiving iniquity ``f10!}, transgression, and\index{God's law:commandments:transgressions ``f10!} sin;\footnote{\ Exodus 34:6\index{02 Exodus:ch.  34, v.   6, pg. ``f30!}{}-7\index{02 Exodus:ch.  34, v.   6-7, pg. ``f30!}.} the rewarder\index{Believers:rewarded by God ``f10!} of them that diligently seek\index{Salvation:for those who diligently seek God ``f10!} Him;\footnote{\ Hebrews 11:6\index{58 Hebrews:ch.  11, v.   6, pg. ``f30!}.} and withal, most just and terrible\index{Judgement:just and terrible ``f10!} in His judgments\index{God's attributes:judge ``f10!},\footnote{\ Nehemiah 9:32-33\index{16 Nehemiah:ch.   9, v.  32-33, pg. ``f30!}.} hating\index{God's attributes:hates sin ``f10!} all sin,\footnote{\ Psalm 5:5-6\index{19 Psalms:ch.   5, v.   5-6, pg. ``f30!}.} and who will by no means clear the guilty\index{guilt ``f10!}.\footnote{\ Nahum 1:2-3\index{34 Nahum:ch.   1, v.   2-3, pg. ``f30!}; Exodus 34:7\index{02 Exodus:ch.  34, v.   7, pg. ``f30!}.}     
\bigskip
2.2\ \ God has all life\index{God's attributes:glory:his own ``f10!},\footnote{\ John 5:26\index{43 John:ch.   5, v.  26, pg. ``f30!}.} glory,\footnote{\ Acts 7:2\index{44 Acts:ch.   7, v.   2, pg. ``f30!}.} goodness\index{God's attributes:goodness ``f10!},\footnote{\ Psalm 119:68\index{19 Psalms:ch. 119, v.  68, pg. ``f30!}.} blessedness,\index{God's attributes:perfect blessedness ``f10!}\footnote{\ 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}; Romans 9:5\index{45 Romans:ch.   9, v.   5, pg. ``f30!}.} in and of Himself; and is alone in and unto Himself all-sufficient\index{God's attributes:all-sufficient ``f10!}, not standing in need of any creatures\index{creatures ``f10!} which He has made,\footnote{\ Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}{}-25\index{44 Acts:ch.  17, v.  24-25, pg. ``f30!}.} nor deriving\index{God's attributes:glory:not derived from creatures ``f10!} any glory from them,\footnote{\ Job 22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}.} but only manifesting\index{God's attributes:glory:manifested ``f10!} His own glory in, by, unto, and upon them: He is the alone fountain\index{God's attributes:creator:fountain of existence ``f10!} of all being, of whom, through whom, and to whom are all things;\footnote{\ Rom 11:36\index{45 Romans:ch.  11, v.  36, pg. ``f30!}.} and has most sovereign\index{God's attributes:sovereign lordship ``f10!} dominion over the\index{Mankind:dominion over creation ``f10!}m\index{God's attributes:sovereign dominion ``f10!}, to do by them, for them, or upon them whatsoever Himself pleases\index{Providence:at God's pleasure ``f10!}.\footnote{\ Revelation 4:11\index{66 Revelation:ch.   4, v.  11, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}; Daniel 4:25, 35\index{27 Daniel:ch.   4, v.  25, 35, pg. ``f30!}.} In His sight all things are open and manifest;\footnote{\ Hebrews 4:13\index{58 Hebrews:ch.   4, v.  13, pg. ``f30!}.} His knowledge\index{God's attributes:omniscience ``f10!} is infinite\index{God's attributes:infinite:knowledge ``f10!}, infallible, and\index{God's attributes:infallible fore-knowledge ``f10!} independent\index{God's attributes:knowledge independent of creature ``f10!} upon the creature,\footnote{\ Romans 11:33\index{45 Romans:ch.  11, v.  33, pg. ``f30!}{}-34\index{45 Romans:ch.  11, v.  33-34, pg. ``f30!}; Psalm 147:5\index{19 Psalms:ch. 147, v.   5, pg. ``f30!}.} so as nothing is to Him contingent\index{Causation:no contingency with God ``f10!}, or uncertain.\footnote{\ Acts 15:18\index{44 Acts:ch.  15, v.  18, pg. ``f30!}; Ezekiel 11:5\index{26 Ezekiel:ch.  11, v.   5, pg. ``f30!}.} He is most holy\index{God's attributes:holy:most ``f10!} in all His counsels\index{God's attributes:holy:counsels ``f10!}, in all His works\index{Providence:God's works ``f10!}, and in all His commands\index{God's law:commandments:ten ``f10!}.\footnote{\ Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg. ``f30!}; Romans 7:12\index{45 Romans:ch.   7, v.  12, pg. ``f30!}.} To Him is due from angels\index{Angels:angels ``f10!} and men, and every other creature, whatsoever worship\index{Worship:of God ``f10!}, service, or obedience\index{God's law:commandments:obedience:required ``f10!} He is pleased to require of them.\footnote{\ Revelation 5:12-14\index{66 Revelation:ch.   5, v.  12-14, pg. ``f30!}.}     
\bigskip
2.3\ \ In the unity\index{God's attributes:unity of persons ``f10!} of the Godhead\index{God's attributes:Godhead ``f10!} there be three persons\index{God's attributes:trinity:three persons ``f10!}, of one substance\index{God's attributes:trinity:one substance ``f10!}, power, and eternity\index{God's attributes:eternal ``f10!}; God the Father\index{Father:God the ``f10!}, God the Son\index{Jesus:Son of God ``f10!}, and God the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}.\footnote{\ 1 John 5:7\index{62 1 John:ch.   5, v.   7, pg. ``f30!}; Matthew 3:16-17\index{40 Matthew:ch.   3, v.  16-17, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 2 Corinthians 13:1\index{47 2 Corinthians:ch.  13, v.   1, pg. ``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}.} The Father is of none, neither begotten\index{Holy Spirit:not begotten ``f10!}, nor proceeding: the\index{Father:God the, no procession ``f10!} Son is eternally\index{Jesus:begotten eternally ``f10!} begotten of the\index{Jesus:begotten by the Father ``f10!} Father:\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}, 18\index{43 John:ch.   1, v.  14, 18, pg. ``f30!}.} the Holy Ghost eternally proceeding from\index{Holy Spirit:procession from Father and Son ``f10!} the Father and the Son.\footnote{\ John 15:26\index{43 John:ch.  15, v.  26, pg. ``f30!}; Galatians 4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.}   
\bigskip
  

\section{Chapter 3 -- Of God's Eternal Decree} \index{Predestination:eternal decrees ``f10!}3.1\ \ God from all eternity\index{God's attributes:eternal ``f10!} did, by the most wise\index{God's attributes:wisdom ``f10!} and holy counsel of His\index{Causation:God's own will ``f10!} own will, freely, and unchangeably ordain whatsoever\index{Predestination:all that happens ``f10!} comes to pass:\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}; Romans 11:33\index{45 Romans:ch.  11, v.  33, pg. ``f30!}; Hebrews 6:17\index{58 Hebrews:ch.   6, v.  17, pg. ``f30!}; Romans 9:15, 18\index{45 Romans:ch.   9, v.  15, 18, pg. ``f30!}.} yet so, as thereby neither is God the author\index{Sin:God not author or approver of ``f10!} of sin,\footnote{\ James 1:13, 17\index{59 James:ch.   1, v.  13, 17, pg. ``f30!}; 1 John 1:5\index{62 1 John:ch.   1, v.   5, pg. ``f30!}.} nor is violence\index{Causation:will of creatures:not destroyed by predestination ``f10!} offered to the will of the creatures\index{creatures ``f10!}, nor is the liberty or\index{Liberty:natural ``f10!} contingency\index{Causation:contingency ``f10!} of second causes\index{Causation:second causes ``f10!} taken away, but rather established.\footnote{\ Acts 2:23\index{44 Acts:ch.   2, v.  23, pg. ``f30!}; Matthew 17:12\index{40 Matthew:ch.  17, v.  12, pg. ``f30!}; Acts 4:27-28\index{44 Acts:ch.   4, v.  27-28, pg. ``f30!}; John 19:11\index{43 John:ch.  19, v.  11, pg. ``f30!}; Proverbs 16:33\index{20 Proverbs:ch.  16, v.  33, pg. ``f30!}.}   
\bigskip
3.2\ \ Although God knows whatsoever\index{God's attributes:omniscience ``f10!} may or can come to pass upon all supposed conditions\index{Predestination:not conditional ``f10!},\footnote{\ Acts 15:18\index{44 Acts:ch.  15, v.  18, pg. ``f30!}; 1 Samuel 23:11-12\index{09 1 Samuel:ch.  23, v.  11-12, pg. ``f30!}; Matthew 11:21, 23\index{40 Matthew:ch.  11, v.  21, 23, pg. ``f30!}.} yet has He not decreed\index{Predestination:eternal decrees ``f10!} anything because He foresaw\index{Predestination:not based on God's foresight:of events ``f10!} it as future, or as that which would come to pass upon such conditions.\footnote{\ Romans 9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}, 13, 16\index{45 Romans:ch.   9, v.  11, 13, 16, pg. ``f30!}, 18\index{45 Romans:ch.   9, v.  11, 13, 16, 18, pg. ``f30!}.}   
\bigskip
3.3\ \ By the decree\index{Predestination:eternal decrees ``f10!} of God, for the manifestation of His glory\index{God's attributes:glory:his own ``f10!}, some men and angels\index{Angels:angels ``f10!}\footnote{\ 1 Timothy 5:21\index{54 1 Timothy:ch.   5, v.  21, pg. ``f30!}; Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}.} are predestinated\index{Predestination:predestination ``f10!} unto everlasting life\index{everlasting life ``f10!}, and others fore-ordained\index{Predestination:fore-ordination ``f10!} to everlasting death\index{Death:everlasting ``f10!}.\footnote{\ Romans 9:22, 23\index{45 Romans:ch.   9, v.  22, 23, pg. ``f30!}; Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}{}-6\index{49 Ephesians:ch.   1, v.   5-6, pg. ``f30!}; Proverbs 16:4\index{20 Proverbs:ch.  16, v.   4, pg. ``f30!}.}   
\bigskip
3.4\ \ These angels\index{Angels:angels ``f10!} and men, thus predestinated\index{Predestination:predestination ``f10!}, and fore-ordained\index{Predestination:fore-ordination ``f10!}, are particularly and unchangeably designed\index{Angels:unchangeably designed ``f10!}, and their number so certain and definite\index{Angels:definite number, no increase or decrease ``f10!}, that it cannot be either increased or diminished.\footnote{\ 2 Timothy 2:19\index{55 2 Timothy:ch.   2, v.  19, pg. ``f30!}; John 13:18\index{43 John:ch.  13, v.  18, pg. ``f30!}.}   
\bigskip
3.5\ \ Those of mankind\index{Predestination:mankind:unto life ``f10!} that are predestinated\index{Predestination:predestination ``f10!} unto life, God, before the foundation\index{Predestination:before foundation of world. ``f10!} of the world was laid, according to His eternal and immutable\index{God's attributes:immutable ``f10!} purpose, and the secret counsel and\index{Causation:God's own will ``f10!} good pleasure\index{Providence:at God's pleasure ``f10!} of His will, has chosen\index{Election:chosen by God ``f10!}, in Christ, unto everlasting glory,\index{Resurrection:believers:into glory:everlasting ``f10!}\footnote{\ Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}, 9\index{49 Ephesians:ch.   1, v.   4, 9, pg. ``f30!}, 11\index{49 Ephesians:ch.   1, v.   4, 9, 11, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}; 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.   9, pg. ``f30!}; 1 Thessalonians 5:9\index{52 1 Thessalonians:ch.   5, v.   9, pg. ``f30!}.} out of His mere free grace\index{Grace:God's free ``f10!} and love, without\index{God's attributes:love ``f10!} any foresight\index{Predestination:not based on God's foresight:of faith ``f10!} of faith or good works\index{Good works:adorn profession of faith ``f10!}, or perseverance\index{Good works:not foreseen by God as cause of election ``f10!} in either of them, or any other thing in the creature\index{creatures ``f10!}, as conditions\index{Predestination:not conditional ``f10!}, or causes moving Him thereunto:\footnote{\ Romans 9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}, 13, 16\index{45 Romans:ch.   9, v.  11, 13, 16, pg. ``f30!}; Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}, 9\index{49 Ephesians:ch.   1, v.   4, 9, pg. ``f30!}.} and all to the praise of His\index{Worship:praising God ``f10!} glorious grace\index{Grace:glorious ``f10!}.\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}, 12\index{49 Ephesians:ch.   1, v.   6, 12, pg. ``f30!}.}   
\bigskip
3.6.\ \ As God has appointed the elect\index{Election:elect of God ``f10!} unto glory\index{Believers:elected unto glory ``f10!}, so has He, by the eternal and most\index{Causation:God's own will ``f10!} free purpose of His will, fore-ordained\index{Predestination:fore-ordination ``f10!} all the means thereunto.\footnote{\ 1 Peter 1:2\index{60 1 Peter:ch.   1, v.   2, pg. ``f30!}; Ephesians 1:4\index{49 Ephesians:ch.   1, v.   4, pg. ``f30!}{}-5\index{49 Ephesians:ch.   1, v.   4-5, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  10, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.} Wherefore they who are elected\index{Election:chosen by God ``f10!}, being fallen in\index{Sin:original:first transgression (of Adam) ``f10!} Adam, are\index{Adam:guilt of first sin on all mankind ``f10!} redeemed\index{Redemption:purchased by Christ ``f10!} by Christ,\footnote{\ 1 Thessalonians 5:9\index{52 1 Thessalonians:ch.   5, v.   9, pg. ``f30!}{}-10\index{52 1 Thessalonians:ch.   5, v.   9-10, pg. ``f30!}; Titus 2:14\index{56 Titus:ch.   2, v.  14, pg. ``f30!}.} are effectually called\index{Effectual call:calling ``f10!} unto faith\index{Faith:worked in believers by the Holy Spirit ``f10!} in Christ by His Spirit working in\index{Salvation:by working of Holy Spirit ``f10!} due season, are justified\index{Justification:justified ``f10!}, adopted\index{Adoption:in salvation ``f10!}, sanctified\index{Sanctification:sanctification ``f10!},\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}; Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.} and kept by His power\index{Perseverance:kept by God's power ``f10!} through faith, unto salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Peter 1:5\index{60 1 Peter:ch.   1, v.   5, pg. ``f30!}.} Neither are any other redeemed by Christ, effectually called, justified, adopted, sanctified, and saved, but\index{Salvation:only the elect ``f10!} the elect only.\footnote{\ John 17:9\index{43 John:ch.  17, v.   9, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!} to the end, John 6:64-65\index{43 John:ch.   6, v.  64-65, pg. ``f30!}; John 10:26\index{43 John:ch.  10, v.  26, pg. ``f30!}; John 8:47\index{43 John:ch.   8, v.  47, pg. ``f30!}; 1 John 2:19\index{62 1 John:ch.   2, v.  19, pg. ``f30!}.}   
\bigskip
3.7\ \ The rest of mankind\index{Predestination:mankind:unto reprobation ``f10!} God was pleased\index{Providence:at God's pleasure ``f10!}, according to the unsearchable\index{Providence:unsearchable wise counsel ``f10!} counsel of His\index{Causation:God's own will ``f10!} own will, whereby He extends or withholds\index{Providence:God withholds mercy, grace, favour as he wills ``f10!} mercy, as\index{God's attributes:merciful ``f10!} He pleases, for the glory of His\index{God's attributes:glory:his own ``f10!} sovereign\index{God's attributes:sovereign lordship ``f10!} power over His creatures\index{creatures ``f10!}, to pass by; and to ordain the\index{Predestination:Jesus as mediator ``f10!}m\index{Predestination:unto wrath ``f10!} to dishonour\index{Resurrection:wicked raised in dishonour ``f10!} and wrath, for\index{Wrath:of God:for sin ``f10!} their sin, to the praise of His\index{Worship:praising God ``f10!} glorious justice\index{God's attributes:administers justice ``f10!}.\footnote{\ Matthew 11:25-26\index{40 Matthew:ch.  11, v.  25-26, pg. ``f30!}; Romans 9:17\index{45 Romans:ch.   9, v.  17, pg. ``f30!}, 18, 21-22\index{45 Romans:ch.   9, v.  17, 18, 21-22, pg. ``f30!}; 2 Timothy 2:19\index{55 2 Timothy:ch.   2, v.  19, pg. ``f30!}{}-20\index{55 2 Timothy:ch.   2, v.  19-20, pg. ``f30!}; Jude 4\index{65 Jude:v.   4, pg. ``f30!}; 1 Peter 2:8\index{60 1 Peter:ch.   2, v.   8, pg. ``f30!}.}   
\bigskip
3.8\ \ The doctrine of this\index{Doctrine:of predestination ``f10!} high mystery\index{Predestination:mystery of ``f10!} of predestination\index{Predestination:predestination ``f10!} is to be handled with special prudence and\index{Predestination:doctrine to be handled with prudence ``f10!} care,\footnote{{\ Romans 9:20}\index{45 Romans:ch.   9, v.  20, pg. ``f30!}\index{45 Romans:ch.  11, v.  33, pg. ``f30!}\index{05 Deuteronomy:ch.  29, v.  29, pg. ``f30!} that men attending the will of God revealed in His Word, and\index{Scripture:word of God:standard for:life ``f10!} yielding\index{God's law:commandments:obedience:required ``f10!} obedience thereunto, may, from the certainty of their effectual vocation\index{Redemption:effectually:assured ``f10!}, be assured of\index{Assurance:of election ``f10!} their eternal election.\index{Election:eternal ``f10!}\footnote{\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}, reverence, and admiration\index{Predestination:to elicit admiration of God ``f10!} of God,\footnote{\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}\index{45 Romans:ch.  11, v.  33, pg. ``f30!}, diligence, and abundant consolation to\index{Predestination:consolation for believers ``f10!} all that sincerely obey\index{Gospel:to be sincerely obeyed ``f10!} the Gospel\index{Gospel:gospel ``f10!}.\footnote{\ Romans 11:5-6, 20\index{45 Romans:ch.  11, v.   5-6, 20, pg. ``f30!}; 2 Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}3\index{45 Romans:ch.   8, v.  33, pg. ``f30!}; Luke 10:20\index{42 Luke:ch.  10, v.  20, pg. ``f30!}.}  

\section{Chapter 4 -- Of Creation} 4.1\ \ It pleased\index{Providence:at God's pleasure ``f10!} God the Father\index{Father:God the ``f10!}, Son\index{Jesus:Son of God ``f10!}, and Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!},\footnote{\ Hebrews 1:2\index{58 Hebrews:ch.   1, v.   2, pg. ``f30!}; John 1:2-3\index{43 John:ch.   1, v.   2-3, pg. ``f30!}; Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}; Job 26:13\index{18 Job:ch.  26, v.  13, pg. ``f30!}; Job 33:4\index{18 Job:ch.  33, v.   4, pg. ``f30!}.} for the manifestation of the glory of His\index{God's attributes:glory:his own ``f10!} eternal power\index{God's attributes:eternal power ``f10!}, wisdom, and goodness\index{God's attributes:goodness ``f10!},\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.  20, pg. ``f30!}; Jeremiah 10:12\index{24 Jeremiah:ch.  10, v.  12, pg. ``f30!}; Psalm 104:24\index{19 Psalms:ch. 104, v.  24, pg. ``f30!}; Psalm 33:5-6\index{19 Psalms:ch.  33, v.   5-6, pg. ``f30!}.} in the beginning, to create, or\index{God's attributes:creator:from nothing ``f10!} make of nothing, the world, and all things therein whether visible or\index{God's attributes:creator:of visible or invisible ``f10!} invisible, in\index{Creation:visible and invisible ``f10!} the space of six\index{Creation:in six days ``f10!} days; and all very good.\footnote{\ Genesis 1 chap. Hebrews 11:3\index{58 Hebrews:ch.  11, v.   3, pg. ``f30!}; Colossians 1:16\index{51 Colossians:ch.   1, v.  16, pg. ``f30!}; Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Clarifications}}\textcolor{black}{).}}   
\bigskip
4.2\ \ After God\index{Mankind:created:attributes ``f10!} had made all other creatures\index{creatures ``f10!}, He created man\index{Mankind:created:by God ``f10!}, male\index{Mankind:male and female ``f10!} and female\index{Mankind:created:male and female ``f10!},\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}.} with reasonable\index{Mankind:created:reasonable soul ``f10!} and immortal\index{Mankind:immortal soul ``f10!} souls,\footnote{\ Genesis 2:7\index{01 Genesis:ch.   2, v.   7, pg. ``f30!} with Ecclesiastes 12:7\index{21 Ecclesiastes:ch.  12, v.   7, pg. ``f30!} \& Luke 23:43\index{42 Luke:ch.  23, v.  43, pg. ``f30!} and Matthew 10:28\index{40 Matthew:ch.  10, v.  28, pg. ``f30!}.} endued with knowledge\index{Mankind:created:with knowledge, righteousness, and holiness ``f10!}, righteousness, and true holiness\index{Mankind:created:holy ``f10!}, after His own image\index{Mankind:image of God ``f10!};\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}; Colossians 3:10\index{51 Colossians:ch.   3, v.  10, pg. ``f30!}; Ephesians 4:24\index{49 Ephesians:ch.   4, v.  24, pg. ``f30!}.} having the law of God written in their hearts\index{God's law:written on man's heart ``f10!},\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}.} and power to fulfil it\index{Covenant:works:conditional on perfect obedience ``f10!}:\footnote{\ Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.} and yet under a possibility of transgressing\index{Mankind:created:with possibility of transgressing ``f10!}, being left to the liberty of their\index{Liberty:of own will ``f10!} own will, which was subject unto change.\footnote{\ Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}.} Beside this law written in their hearts, they received a command, not\index{God's law:commandments:given to Adam ``f10!} to eat of the tree of the\index{Tree:of the knowledge of good and evil ``f10!} knowledge of good and evil, which while they kept, they were happy\index{happiness ``f10!} in their communion with God\index{Communion:with God ``f10!},\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Genesis 3:8\index{01 Genesis:ch.   3, v.   8, pg. ``f30!}{}-11, 23\index{01 Genesis:ch.   3, v.   8-11, 23, pg. ``f30!}.} and had dominion over the\index{Mankind:dominion over creation ``f10!} creatures.\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}, 28\index{01 Genesis:ch.   1, v.  26, 28, pg. ``f30!}.}  

\section{Chapter 5 -- Of Providence} \index{Providence:providence ``f10!}5.1\ \ God the great Creator\index{God's attributes:creator:of all things ``f10!} of all things does uphold\index{Providence:upholds all creatures ``f10!},\footnote{\ Hebrews 1:3\index{58 Hebrews:ch.   1, v.   3, pg. ``f30!}.} direct, dispose, and\index{Providence:God disposes all things for his glory ``f10!} govern\index{Providence:governing all ``f10!} all creatures\index{creatures ``f10!}, actions, and things,\footnote{\ Daniel 4:34-35\index{27 Daniel:ch.   4, v.  34-35, pg. ``f30!}; Psalm 135:6\index{19 Psalms:ch. 135, v.   6, pg. ``f30!}; Acts 17:25\index{44 Acts:ch.  17, v.  25, pg. ``f30!}{}-26, 28\index{44 Acts:ch.  17, v.  25-26, 28, pg. ``f30!}; Job 38 to 41 chapters.} from the greatest even to the least,\footnote{\ Matthew 10:29, 30-31\index{40 Matthew:ch.  10, v.  29, 30-31, pg. ``f30!}.} by His most wise\index{God's attributes:wisdom ``f10!} and holy providence\index{Providence:providence ``f10!},\footnote{\ Proverbs 15:3\index{20 Proverbs:ch.  15, v.   3, pg. ``f30!}; Psalm 104:24\index{19 Psalms:ch. 104, v.  24, pg. ``f30!}; Psalm 145:17\index{19 Psalms:ch. 145, v.  17, pg. ``f30!}.} according to His infallible\index{God's attributes:infallible fore-knowledge ``f10!} fore-knowledge\index{God's attributes:foreknowledge ``f10!},\footnote{\ Acts 15:18\index{44 Acts:ch.  15, v.  18, pg. ``f30!}; Psalm 94:8, 11\index{19 Psalms:ch.  94, v.   8, 11, pg. ``f30!}.} and the free and immutable\index{God's attributes:immutable ``f10!} counsel of His\index{Causation:God's own will ``f10!} own will,\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}; Psalm 33:10-11\index{19 Psalms:ch.  33, v.  10-11, pg. ``f30!}.} to the praise of the\index{Worship:praising God ``f10!} glory of His\index{God's attributes:glory:his own ``f10!} wisdom, power\index{God's attributes:listed (definition of God) ``f10!}, justice, goodness\index{God's attributes:goodness ``f10!}, and mercy.\footnote{\ Isaiah 63:14\index{23 Isaiah:ch.  63, v.  14, pg. ``f30!}; Ephesians 3:10\index{49 Ephesians:ch.   3, v.  10, pg. ``f30!}; Romans 9:17\index{45 Romans:ch.   9, v.  17, pg. ``f30!}; Genesis 45:7\index{01 Genesis:ch.  45, v.   7, pg. ``f30!}; Psalm 145:7\index{19 Psalms:ch. 145, v.   7, pg. ``f30!}.}   
\bigskip
5.2\ \ Although, in relation to the\index{Causation:relation to fore-knowledge ``f10!} fore-knowledge\index{God's attributes:foreknowledge ``f10!} and decree\index{Predestination:eternal decrees ``f10!} of God, the first Cause\index{Causation:first cause ``f10!}, all things come to pass immutably\index{God's attributes:immutable ``f10!}, and infallibly\index{Providence:infallible ``f10!}:\footnote{\ Acts 2:23\index{44 Acts:ch.   2, v.  23, pg. ``f30!}.} yet, by the same providence\index{Providence:providence ``f10!}, He orders them to fall out, according to the nature of second causes\index{Causation:second causes ``f10!}, either necessarily, freely, or\index{Causation:God's own will ``f10!} contingently\index{Causation:no contingency with God ``f10!}.\footnote{\ Genesis 8:22\index{01 Genesis:ch.   8, v.  22, pg. ``f30!}; Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}5\index{24 Jeremiah:ch.  31, v.  35, pg. ``f30!}; Exodus 21:13\index{02 Exodus:ch.  21, v.  13, pg. ``f30!} with Deuteronomy 19:5\index{05 Deuteronomy:ch.  19, v.   5, pg. ``f30!}; 1 Kings 22:28, 34\index{11 1 Kings:ch.  22, v.  28, 34, pg. ``f30!}; Isaiah 10:6-7\index{23 Isaiah:ch.  10, v.   6-7, pg. ``f30!}.}   
\bigskip
5.3\ \ God in His ordinary\index{Providence:ordinary means not binding on God ``f10!} providence\index{Providence:providence ``f10!} makes use of means,\footnote{\ Acts 27:31, 44\index{44 Acts:ch.  27, v.  31, 44, pg. ``f30!}; Isaiah 55:1\index{23 Isaiah:ch.  55, v.   1, pg. ``f30!}0-11\index{23 Isaiah:ch.  55, v.  10-11, pg. ``f30!}; Hosea 2:21-22\index{28 Hosea:ch.   2, v.  21-22, pg. ``f30!}.} yet is free to work without,\footnote{\ Hosea 1:7\index{28 Hosea:ch.   1, v.   7, pg. ``f30!}; Matthew 4:4\index{40 Matthew:ch.   4, v.   4, pg. ``f30!}; Job 34:20\index{18 Job:ch.  34, v.  20, pg. ``f30!}.} above,\footnote{\ Romans 4:19-21\index{45 Romans:ch.   4, v.  19-21, pg. ``f30!}.} and against them at His pleasure\index{Providence:at God's pleasure ``f10!}.\footnote{\ 2 Kings 6:6\index{12 2 Kings:ch.   6, v.   6, pg. ``f30!}; Daniel 3:27\index{27 Daniel:ch.   3, v.  27, pg. ``f30!}.}   
\bigskip
5.4\ \ The almighty\index{God's attributes:almighty power ``f10!} power, unsearchable\index{Providence:unsearchable wise counsel ``f10!} wisdom, and infinite\index{God's attributes:infinite:goodness ``f10!} goodness\index{God's attributes:goodness ``f10!} of God so far manifest themselves in\index{Providence:manifests goodness of God ``f10!} His providence\index{Providence:providence ``f10!}, that it extends itself\index{Providence:extends even to first sin ``f10!} even to the first fall\index{Sin:original:first transgression (of Adam) ``f10!}, and all other sins of angels\index{Angels:angels ``f10!} and men;\footnote{\ Romans 11:32\index{45 Romans:ch.  11, v.  32, pg. ``f30!}{}-34\index{45 Romans:ch.  11, v.  32-34, pg. ``f30!}; 2 Samuel 24:1\index{10 2 Samuel:ch.  24, v.   1, pg. ``f30!} with 1 Chronicles 21:1\index{13 1 Chronicles:ch.  21, v.   1, pg. ``f30!}; 1 Kings 22:22-23\index{11 1 Kings:ch.  22, v.  22-23, pg. ``f30!}; 1 Chronicles 10:4, 13-14\index{13 1 Chronicles:ch.  10, v.   4, 13-14, pg. ``f30!}; 2 Samuel 16:10\index{10 2 Samuel:ch.  16, v.  10, pg. ``f30!}; Acts 2:23\index{44 Acts:ch.   2, v.  23, pg. ``f30!}; Acts 4:27-28\index{44 Acts:ch.   4, v.  27-28, pg. ``f30!}.} and that not by a bare permission\index{Sin:God purposes and permits for his own glory ``f10!},\footnote{\ Acts 14:16\index{44 Acts:ch.  14, v.  16, pg. ``f30!}.} but such as has joined with it a most wise\index{God's attributes:wisdom ``f10!} and powerful bounding\index{Providence:governing all ``f10!},\footnote{\ Psalm 76:10\index{19 Psalms:ch.  76, v.  10, pg. ``f30!}; 2 Kings 19:28\index{12 2 Kings:ch.  19, v.  28, pg. ``f30!}.} and otherwise ordering and\index{Providence:ordering:all things for his glory and believers' good ``f10!} governing of them, in a manifold dispensation\index{Providence:God's dispensation ``f10!}, to His own holy\index{God's attributes:holy:ends ``f10!} ends;\footnote{\ Genesis 50:20\index{01 Genesis:ch.  50, v.  20, pg. ``f30!}; Isaiah 10:6-7\index{23 Isaiah:ch.  10, v.   6-7, pg. ``f30!}, 12\index{23 Isaiah:ch.  10, v.   6-7, 12, pg. ``f30!}.} yet so, as the sinfulness thereof proceeded only\index{Sin:proceeds only from man, not God ``f10!} from the creature\index{creatures ``f10!}, and not from God, who, being most holy\index{God's attributes:holy:most ``f10!} and righteous, neither\index{God's attributes:righteousness ``f10!} is, nor can be, the author or approver\index{Sin:God not author or approver of ``f10!} of sin.\footnote{\ James 1:13-14, 17\index{59 James:ch.   1, v.  13-14, 17, pg. ``f30!}; 1 John 2:16\index{62 1 John:ch.   2, v.  16, pg. ``f30!}; Psalm 50:21\index{19 Psalms:ch.  50, v.  21, pg. ``f30!}.} 
\bigskip
5.5\ \ The most wise\index{God's attributes:wisdom ``f10!}, righteous, and gracious\index{Grace:God's free ``f10!} God does oftentimes leave for a season His\index{Temptation:left by God for a season ``f10!} own children to manifold temptations\index{Temptation:temptation ``f10!}, and the corruption of the\index{Sin:original:corrupt nature:flesh ``f10!}ir\index{Sin:original:corrupt nature:hearts ``f10!} own hearts\index{Heart:corrupt ``f10!}, to chastise\index{Believers:chastened by God ``f10!} them for their former sins, or to discover\index{Temptation:exposes (discovers) corruption ``f10!} unto them the hidden strength of\index{Sin:original:hidden strength of corruption ``f10!} corruption, and\index{Sin:original:corrupt nature:hidden ``f10!} deceitfulness\index{Deceitfulness:of heart ``f10!} of their hearts, that\index{Heart:deceitful ``f10!} they may be humbled\index{Humility:before God ``f10!};\footnote{\ 2 Chronicles 32:25-26, 31\index{14 2 Chronicles:ch.  32, v.  25-26, 31, pg. ``f30!}; 2 Samuel 24:1\index{10 2 Samuel:ch.  24, v.   1, pg. ``f30!}.} and, to raise them to a more close and constant dependence\index{Temptation:increase dependence on God ``f10!} for their support upon Himself, and to make them more watchful against\index{Temptation:watchfulness required to avoid ``f10!} all future occasions of sin, and for sundry other just and holy\index{God's attributes:holy:nature ``f10!} ends.\footnote{\ 2 Corinthians 12:7-9\index{47 2 Corinthians:ch.  12, v.   7-9, pg. ``f30!}; Psalm 73 throughout, Psalm 77:1-12\index{19 Psalms:ch.  77, v.   1-12, pg. ``f30!}; Mark 14:66 to the end, with John 21:15-17\index{43 John:ch.  21, v.  15-17, pg. ``f30!}.}   
\bigskip
5.6\ \ As for those wicked\index{blindness, spiritual ``f10!} and ungodly\index{Unbelievers:ungodly ``f10!} men whom God, as a righteous Judge, for\index{God's attributes:judge ``f10!} former sins, does blind and harden\index{Heart:hardened ``f10!},\footnote{\ Romans 1:24, 26, 28\index{45 Romans:ch.   1, v.  24, 26, 28, pg. ``f30!}; Romans 11:7\index{45 Romans:ch.  11, v.   7, pg. ``f30!}{}-8\index{45 Romans:ch.  11, v.   7-8, pg. ``f30!}.} from them He not only withholds\index{Providence:God withholds mercy, grace, favour as he wills ``f10!} His grace, whereby they might have been enlightened\index{Providence:God withholds enlightening from wicked ``f10!} in their understandings\index{Unbelievers:understandings darkened ``f10!}, and wrought upon\index{Salvation:by working of Holy Spirit ``f10!} in their hearts;\index{Heart:new:enlightened ``f10!}\footnote{\ Deuteronomy 29:4\index{05 Deuteronomy:ch.  29, v.   4, pg. ``f30!}.} but sometimes also withdraws\index{Punishment:God withdraws gifts ``f10!} the gifts which\index{Punishment:gifts from God withdrawn ``f10!} they had,\footnote{\ Matthew 13:12\index{40 Matthew:ch.  13, v.  12, pg. ``f30!}; Matthew 25:29\index{40 Matthew:ch.  25, v.  29, pg. ``f30!}.} and exposes\index{Punishment:exposure to consequences of sin ``f10!} them to such objects as their corruption makes\index{Sin:original:corrupt nature:lustful ``f10!} occasions of sin;\footnote{\ Deuteronomy 2:30\index{05 Deuteronomy:ch.   2, v.  30, pg. ``f30!}; 2 Kings 8:12-13\index{12 2 Kings:ch.   8, v.  12-13, pg. ``f30!}.} and, withal, gives them over to their own lusts\index{Sexual immorality:lust ``f10!}, the temptations\index{Temptation:temptation ``f10!} of the world, and the power of Satan\index{Satan:power of ``f10!}:\footnote{\ Psalm 81:11\index{19 Psalms:ch.  81, v.  11, pg. ``f30!}{}-12\index{19 Psalms:ch.  81, v.  11-12, pg. ``f30!}; 2 Thessalonians 2:10\index{53 2 Thessalonians:ch.   2, v.  10, pg. ``f30!}{}-12\index{53 2 Thessalonians:ch.   2, v.  10-12, pg. ``f30!}.} whereby it comes to pass that they harden themselves, even under those means which God uses for the softening\index{Providence:hardens some, softens others ``f10!} of others.\footnote{\ Exodus 7:3\index{02 Exodus:ch.   7, v.   3, pg. ``f30!} with Exodus 8:15, 32\index{02 Exodus:ch.   8, v.  15, 32, pg. ``f30!}; 2 Corinthians 2:15-16\index{47 2 Corinthians:ch.   2, v.  15-16, pg. ``f30!}; Isaiah 8:14\index{23 Isaiah:ch.   8, v.  14, pg. ``f30!}; 1 Peter 2:7-8\index{60 1 Peter:ch.   2, v.   7-8, pg. ``f30!}; Isaiah 6:9-10\index{23 Isaiah:ch.   6, v.   9-10, pg. ``f30!} with Acts 28:26-27\index{44 Acts:ch.  28, v.  26-27, pg. ``f30!}.}   
\bigskip
5.7\ \ As the providence\index{Providence:providence ``f10!} of God does in general reach to all creatures\index{creatures ``f10!}, so after a most special manner, it takes care of His Church\index{Church:Christ's ``f10!}, and disposes\index{Providence:God disposes all for good of his Church ``f10!} all things to the good thereof.\footnote{\ 1 Timothy 4:1\index{54 1 Timothy:ch.   4, v.   1, pg. ``f30!}0\index{54 1 Timothy:ch.   4, v.  10, pg. ``f30!}; Amos 9:8-9\index{30 Amos:ch.   9, v.   8-9, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!}; Isaiah 43:3-5, 14\index{23 Isaiah:ch.  43, v.   3-5, 14, pg. ``f30!}.}  

\section{Chapter 6 -- Of the Fall of Man, of Sin, and of the Punishment thereof} \index{Sin:original:first transgression (of Adam) ``f10!}\index{Punishment:of sin ``f10!}6.1\ \ Our first parents\index{Parents:first, sinned against God ``f10!}, being seduced\index{Temptation:Satan's seduction of our first parents ``f10!} by the subtilty\index{Satan:tempter ``f10!} and temptation\index{Temptation:temptation ``f10!} of Satan, sinned, in eating the forbidden\index{Sin:original:first transgression (of Adam) ``f10!} fruit.\footnote{\ Genesis 3:13\index{01 Genesis:ch.   3, v.  13, pg. ``f30!}; 2 Corinthians 11:3\index{47 2 Corinthians:ch.  11, v.   3, pg. ``f30!}.} This their sin God was pleased\index{Providence:at God's pleasure ``f10!}, according to His wise\index{God's attributes:wisdom ``f10!} and holy counsel, to\index{Causation:God's own will ``f10!} permit\index{Sin:God purposes and permits for his own glory ``f10!}, having purposed to order it to His own glory\index{God's attributes:glory:his own ``f10!}.\footnote{\ Romans 11:32\index{45 Romans:ch.  11, v.  32, pg. ``f30!}.}   
\bigskip
6.2\ \ By this sin they fell\index{Sin:original:all mankind descending from Adam ``f10!} from their original righteousness and\index{Mankind:created:original righteousness ``f10!} communion, with\index{Communion:with God ``f10!} God,\footnote{\ Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}{}-8\index{01 Genesis:ch.   3, v.   6-8, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}3\index{45 Romans:ch.   3, v.  23, pg. ``f30!}.} and so became dead\index{Death:spiritual, in sin ``f10!} in sin,\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}.} and wholly defiled in all\index{Sin:original:corrupt nature:defiled in all parts and faculties ``f10!} the parts and faculties of soul\index{Sin:original:corrupt nature:faculties of the soul ``f10!} and body.\footnote{\ Titus 1:15\index{56 Titus:ch.   1, v.  15, pg. ``f30!}; Genesis 6:5\index{01 Genesis:ch.   6, v.   5, pg. ``f30!}; Jeremiah 17:9\index{24 Jeremiah:ch.  17, v.   9, pg. ``f30!}; Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}{}-19\index{45 Romans:ch.   3, v.  10-19, pg. ``f30!}.} 
\bigskip
6.3\ \ They being the root\index{Sin:original:first transgression (of Adam) ``f10!} of all mankind, the\index{Sin:original:all mankind descending from Adam ``f10!} guilt\index{guilt ``f10!} of this sin was imputed,\index{Imputation:of Adam's sin ``f10!}\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}7\index{01 Genesis:ch.   1, v.  27, pg. ``f30!}{}-28\index{01 Genesis:ch.   1, v.  27-28, pg. ``f30!} and Genesis 2:16-17\index{01 Genesis:ch.   2, v.  16-17, pg. ``f30!} and Acts 17:26\index{44 Acts:ch.  17, v.  26, pg. ``f30!} with Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}, 15-19\index{45 Romans:ch.   5, v.  12, 15-19, pg. ``f30!} and 1 Corinthians 15:21-22\index{46 1 Corinthians:ch.  15, v.  21-22, pg. ``f30!}, 49\index{46 1 Corinthians:ch.  15, v.  21-22, 49, pg. ``f30!}.} and the same death\index{Death:spiritual, in sin ``f10!} in sin and corrupted\index{Sin:original:corrupt nature:conveyed from Adam ``f10!} nature conveyed, to all their posterity descending from them by ordinary\index{Sin:original:by ordinary generation ``f10!} generation.\footnote{\ Psalm 51:5\index{19 Psalms:ch.  51, v.   5, pg. ``f30!}; Genesis 5:3\index{01 Genesis:ch.   5, v.   3, pg. ``f30!}; Job 14:4\index{18 Job:ch.  14, v.   4, pg. ``f30!}; Job 15:14\index{18 Job:ch.  15, v.  14, pg. ``f30!}.}   
\bigskip
6.4\ \ From this original corruption, whereby\index{Sin:original:corrupt nature:whole ``f10!} we are utterly indisposed\index{Sin:original:utterly indisposed ``f10!}, disabled\index{Sin:original:utterly unable and unwilling to do will of God ``f10!}, and made opposite to all good,\footnote{\ Romans 5:6\index{45 Romans:ch.   5, v.   6, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7, pg. ``f30!}; Romans 7:18\index{45 Romans:ch.   7, v.  18, pg. ``f30!}; Colossians 1:21\index{51 Colossians:ch.   1, v.  21, pg. ``f30!}.} and wholly inclined\index{Sin:original:inclined to do evil ``f10!} to all evil,\footnote{\ Genesis 6:5\index{01 Genesis:ch.   6, v.   5, pg. ``f30!}; Genesis 8:21\index{01 Genesis:ch.   8, v.  21, pg. ``f30!}; Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}{}-12\index{45 Romans:ch.   3, v.  10-12, pg. ``f30!}.} do proceed all\index{Sin:original:proceeds all transgressions ``f10!} actual transgressions\index{God's law:commandments:transgressions ``f10!}.\footnote{\ James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}{}-15\index{59 James:ch.   1, v.  14-15, pg. ``f30!}; Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   2, v.   2-3, pg. ``f30!}; Matthew 15:19\index{40 Matthew:ch.  15, v.  19, pg. ``f30!}.}   
\bigskip
6.5\ \ This corruption\index{Sanctification:remnants of corruption remain ``f10!} of nature, during this life, does remain in those that are regenerated\index{Regeneration:regeneration ``f10!};\footnote{\ 1 John 1:8, 10\index{62 1 John:ch.   1, v.   8, 10, pg. ``f30!}; Romans 7:14\index{45 Romans:ch.   7, v.  14, pg. ``f30!}, 17-18, 23\index{45 Romans:ch.   7, v.  14, 17-18, 23, pg. ``f30!}; James 3:2\index{59 James:ch.   3, v.   2, pg. ``f30!}; Proverbs 20:9\index{20 Proverbs:ch.  20, v.   9, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}0\index{21 Ecclesiastes:ch.   7, v.  20, pg. ``f30!}.} and although it be, through Christ, pardoned and\index{Regeneration:pardoned of sin ``f10!} mortified\index{Sin:mortified ``f10!}, yet both itself and all the motions thereof are truly\index{Sin:original:corrupt nature:whole ``f10!} and properly sin.\footnote{\ Romans 7:5, 7-8, 25\index{45 Romans:ch.   7, v.   5, 7-8, 25, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}.}   
\bigskip
6.6\ \ Every sin, both original and actual, being a transgression of\index{God's law:commandments:transgressions ``f10!} the righteous law\index{God's law:righteous requirements ``f10!} of God, and contrary thereunto,\footnote{\ 1 John 3:4\index{62 1 John:ch.   3, v.   4, pg. ``f30!}.} does, in its own nature, bring guilt\index{guilt ``f10!} upon the sinner;\footnote{\ Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}5\index{45 Romans:ch.   2, v.  15, pg. ``f30!}; Romans 3:9\index{45 Romans:ch.   3, v.   9, pg. ``f30!}, 19\index{45 Romans:ch.   3, v.   9, 19, pg. ``f30!}.} whereby he is bound over to the wrath of God\index{Wrath:of God:for sin ``f10!},\footnote{\ Ephesians 2:3\index{49 Ephesians:ch.   2, v.   3, pg. ``f30!}.} and curse of the\index{Curse:of the law ``f10!} law,\footnote{\ Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}.} and so made subject to death\index{Death:punishment for sin ``f10!},\footnote{\ Romans 6:23\index{45 Romans:ch.   6, v.  23, pg. ``f30!}.} with all miseries\index{Punishment:misery ``f10!} spiritual,\footnote{\ Ephesians 4:18\index{49 Ephesians:ch.   4, v.  18, pg. ``f30!}.} temporal,\index{Judgement:temporal ``f10!}\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}0\index{45 Romans:ch.   8, v.  20, pg. ``f30!}; Lamentations 3:39\index{25 Lamentations:ch.   3, v.  39, pg. ``f30!}.} and eternal.\footnote{\ Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}; 2 Thessalonians 1:9\index{53 2 Thessalonians:ch.   1, v.   9, pg. ``f30!}.}  

\section{Chapter 7 -- Of God's Covenant with Man} \index{Covenant:of God with man ``f10!}7.1\ \ The distance\index{Mankind:great distance between him and God ``f10!} between God and the creature\index{creatures ``f10!} is so great, that although reasonable\index{Mankind:created:reasonable soul ``f10!} creatures do owe obedience\index{God's law:commandments:obedience:required ``f10!} unto Him as their Creator\index{God's attributes:creator:of all things ``f10!}, yet they could never have any fruition of Him as their blessedness and\index{Believers:blessed in/by God ``f10!} reward\index{Believers:rewarded by God ``f10!}, but by some voluntary condescension\index{God's attributes:condescension, voluntary ``f10!} on God's part, which He has been pleased\index{Providence:at God's pleasure ``f10!} to express by way of covenant\index{Covenant:administrations ``f10!}.\footnote{\ Isaiah 40:13-17\index{23 Isaiah:ch.  40, v.  13-17, pg. ``f30!}; Job 9:32-33\index{18 Job:ch.   9, v.  32-33, pg. ``f30!}; 1 Samuel 2:25\index{09 1 Samuel:ch.   2, v.  25, pg. ``f30!}; Psalm 113:5-6\index{19 Psalms:ch. 113, v.   5-6, pg. ``f30!}; Psalm 100:2\index{19 Psalms:ch. 100, v.   2, pg. ``f30!}{}-3\index{19 Psalms:ch. 100, v.   2-3, pg. ``f30!}; Job 22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}; Job 35:7-8\index{18 Job:ch.  35, v.   7-8, pg. ``f30!}; Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}; Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}{}-25\index{44 Acts:ch.  17, v.  24-25, pg. ``f30!}.}   
\bigskip
 7.2\ \ The first covenant made\index{Covenant:first:made with man ``f10!} with man was a covenant of works\index{Covenant:works:works ``f10!},\footnote{\ Galatians 3:12\index{48 Galatians:ch.   3, v.  12, pg. ``f30!}.} wherein life was promised to\index{Covenant:first:promised life ``f10!} Adam, and in him to\index{Adam:covenant made with him ``f10!} his posterity\index{Sin:original:all mankind descending from Adam ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}{}-20\index{45 Romans:ch.   5, v.  12-20, pg. ``f30!}; Romans 10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}.} upon condition\index{Covenant:works:conditional on perfect obedience ``f10!} of perfect and personal obedience\index{Covenant:works:perfect obedience required ``f10!}.\footnote{\ Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}.}   
\bigskip
7.3\ \ Man by his fall\index{Sin:original:first transgression (of Adam) ``f10!} having made himself incapable\index{Covenant:works:Adam:incapable of life after his sin ``f10!} of life by that covenant, the\index{Covenant:first:broken ``f10!} Lord was pleased\index{Providence:at God's pleasure ``f10!} to make a second\index{Covenant:second ``f10!},\footnote{\ Galatians 3:21\index{48 Galatians:ch.   3, v.  21, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}{}-21\index{45 Romans:ch.   3, v.  20-21, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}; Genesis 3:15\index{01 Genesis:ch.   3, v.  15, pg. ``f30!}; Isaiah 42:6\index{23 Isaiah:ch.  42, v.   6, pg. ``f30!}.} commonly called the covenant of grace\index{Covenant:grace:grace ``f10!}; wherein He freely\index{Salvation:freely offered ``f10!} offers unto sinners life and salvation\index{Salvation:belief in Jesus by faith ``f10!} by Jesus Christ, requiring\index{Salvation:faith alone required ``f10!} of them faith\index{Faith:in Jesus Christ ``f10!} in Him, that they may be saved,\footnote{\ Mark 16:15\index{41 Mark:ch.  16, v.  15, pg. ``f30!}{}-16\index{41 Mark:ch.  16, v.  15-16, pg. ``f30!}; John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}; Romans 10:6, 9\index{45 Romans:ch.  10, v.   6, 9, pg. ``f30!}; Galatians 3:11\index{48 Galatians:ch.   3, v.  11, pg. ``f30!}.} and promising to\index{Holy Spirit:promising to believers ``f10!} give unto all those that are ordained unto\index{Predestination:unto life ``f10!} life His Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}, to make them willing and able to\index{Conversion:renewal of the will ``f10!} believe.\footnote{\ Ezekiel 36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36, v.  26-27, pg. ``f30!}; John 6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}{}-45\index{43 John:ch.   6, v.  44-45, pg. ``f30!}.}   
\bigskip
7.4\ \ This covenant of grace\index{Covenant:grace:grace ``f10!} is frequently set forth in Scripture by the name of a Testament\index{Covenant:testament ``f10!}, in reference to the death\index{Jesus:death:death ``f10!} of Jesus Christ the Testator\index{Covenant:grace:Jesus testator ``f10!}, and to the everlasting inheritance\index{Believers:inheritance:everlasting ``f10!}, with all things belonging to it, therein bequeathed.\footnote{\ Hebrews 9:15\index{58 Hebrews:ch.   9, v.  15, pg. ``f30!}{}-17\index{58 Hebrews:ch.   9, v.  15-17, pg. ``f30!}; Hebrews 7:22\index{58 Hebrews:ch.   7, v.  22, pg. ``f30!}; Luke 22:20\index{42 Luke:ch.  22, v.  20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch.  11, v.  25, pg. ``f30!}.}   
\bigskip
7.5\ \ This covenant was\index{Covenant:administrations ``f10!} differently administered in the time of the law, and in the time of the gospel\index{Gospel:gospel ``f10!}:\footnote{\ 2 Corinthians 3:6\index{47 2 Corinthians:ch.   3, v.   6, pg. ``f30!}{}-9\index{47 2 Corinthians:ch.   3, v.   6-9, pg. ``f30!}.} under the law, it was administered by promises\index{Covenant:promises ``f10!}, prophecies\index{Covenant:prophecies ``f10!}, sacrifices\index{Ceremonial law:sacrifices ``f10!}, circumcision\index{circumcision ``f10!}, the paschal\index{paschal lamb ``f10!} lamb\index{lamb, paschal ``f10!}, and other types\index{types, OT ``f10!} and ordinances delivered\index{Covenant:OT ordinances:signified Christ ``f10!} to the people of the Jews, all\index{Jews:OT ordinances ``f10!} fore-signifying\index{Ceremonial law:fore-signifying Christ ``f10!} Christ to come:\footnote{\ Hebrews 8-10\index{58 Hebrews:ch. 8-10, pg. ``f30!} chapters, Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Colossians 2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg. ``f30!}; 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7, pg. ``f30!}.} which were, for that time, sufficient and\index{Covenant:OT ordinances:sufficient for that time ``f10!} efficacious\index{Covenant:OT ordinances:efficacious ``f10!}, through the operation\index{Holy Spirit:operations:saving ``f10!} of the Spirit, to\index{Holy Spirit:instructs the elect ``f10!} instruct and build up the\index{Holy Spirit:builds up the elect ``f10!} elect\index{Election:elect of God ``f10!} in faith\index{Faith:in Jesus Christ, as Messiah ``f10!} in the promised\index{Jesus:Messiah promised ``f10!} Messiah,\footnote{\ 1 Corinthians 10:1-4\index{46 1 Corinthians:ch.  10, v.   1-4, pg. ``f30!}; Hebrews 11:13\index{58 Hebrews:ch.  11, v.  13, pg. ``f30!}; John 8:56\index{43 John:ch.   8, v.  56, pg. ``f30!}.} by whom they had full remission\index{Redemption:remission of sins ``f10!} of sins, and eternal salvation\index{Salvation:belief in Jesus by faith ``f10!}; and is called, the Old Testament\index{Scripture:Old Testament ``f10!}.\footnote{\ Galatians 3:7-9, 14\index{48 Galatians:ch.   3, v.   7-9, 14, pg. ``f30!}.}   
\bigskip
7.6\ \ Under the gospel\index{Gospel:gospel ``f10!}, when Christ, the substance,\footnote{\ Colossians 2:17\index{51 Colossians:ch.   2, v.  17, pg. ``f30!}.} was exhibited, the\index{Covenant:grace:exhibited ``f10!} ordinances in\index{Church:furnished with:ordinances of God ``f10!} which this covenant is dispensed are\index{Covenant:new:ways dispensed ``f10!} the preaching of the Word\index{Preaching:ministry of the Word ``f10!}, and the administration\index{Church:true:sacraments administered ``f10!} of the sacraments\index{Sacraments:holy:ordinances ``f10!} of Baptism and\index{Baptism:sacrament ``f10!} the Lord's Supper\index{Lord's Supper:sacrament ``f10!}:\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-25\index{46 1 Corinthians:ch.  11, v.  23-25, pg. ``f30!}.} which, though fewer in number, and administered with\index{Sacraments:to be:administered with simplicity ``f10!} more simplicity, and less outward\index{Sacraments:NT less outward glory than OT ``f10!} glory; yet, in them, it is held forth in more fulness\index{Covenant:new:more fulness ``f10!}, evidence, and spiritual efficacy,\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}2\index{58 Hebrews:ch.  12, v.  22, pg. ``f30!} to 28; Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}3\index{24 Jeremiah:ch.  31, v.  33, pg. ``f30!}{}-34\index{24 Jeremiah:ch.  31, v.  33-34, pg. ``f30!}.} to all nations\index{Gospel:to all nations ``f10!}, both Jews and\index{Jews:called ``f10!} Gentiles\index{Gentiles:called ``f10!};\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}5-19\index{49 Ephesians:ch.   2, v.  15-19, pg. ``f30!}.} and is called the New Testament\index{Scripture:New Testament ``f10!}.\footnote{\ Luke 22:20\index{42 Luke:ch.  22, v.  20, pg. ``f30!}.} There are not therefore two covenants of\index{Covenant:grace:one only ``f10!} grace, differing in substance\index{Covenant:grace:one, but differ in substance ``f10!}, but one and the same, under various dispensations\index{Covenant:dispensations ``f10!}.\footnote{\ Galatians 3:14, 16\index{48 Galatians:ch.   3, v.  14, 16, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}1-23, 30\index{45 Romans:ch.   3, v.  21-23, 30, pg. ``f30!}; Psalm 32:1\index{19 Psalms:ch.  32, v.   1, pg. ``f30!} with Romans 4:3, 6, 16-17, 23-24\index{45 Romans:ch.   4, v.   3, 6, 16-17, 23-24, pg. ``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}; Acts 15:11\index{44 Acts:ch.  15, v.  11, pg. ``f30!}.}  

\section{Chapter 8 -- Of Christ the Mediator} \index{Jesus:mediator:of covenant ``f10!}8.1\ \ It pleased\index{Providence:at God's pleasure ``f10!} God, in His eternal purpose\index{Predestination:eternal purpose ``f10!}, to choose and ordain the\index{Predestination:Jesus as mediator ``f10!} Lord Jesus, His only begotten\index{Jesus:begotten by the Father ``f10!} Son\index{Jesus:Son of God ``f10!}, to be the Mediator\index{Jesus:mediator:of covenant ``f10!} between God and man;\footnote{\ Isaiah 42:1\index{23 Isaiah:ch.  42, v.   1, pg. ``f30!}; 1 Peter 1:19-20\index{60 1 Peter:ch.   1, v.  19-20, pg. ``f30!}; John 3:16\index{43 John:ch.   3, v.  16, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.} the Prophet\index{Jesus:prophet:office ``f10!},\footnote{\ Acts 3:22\index{44 Acts:ch.   3, v.  22, pg. ``f30!}.} Priest\index{Jesus:priest ``f10!},\footnote{\ Hebrews 5:5-6\index{58 Hebrews:ch.   5, v.   5-6, pg. ``f30!}.} and King\index{Jesus:King ``f10!},\footnote{\ Psalm 2:6\index{19 Psalms:ch.   2, v.   6, pg. ``f30!}; Luke 1:33\index{42 Luke:ch.   1, v.  33, pg. ``f30!}.} the Head\index{Church:Christ as head ``f10!} and Saviour\index{Jesus:saviour of the Church ``f10!} of His Church\index{Church:Christ's ``f10!},\footnote{\ Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}3\index{49 Ephesians:ch.   5, v.  23, pg. ``f30!}.} the Heir\index{Jesus:heir of all things ``f10!} of all things,\footnote{\ Hebrews 1:2\index{58 Hebrews:ch.   1, v.   2, pg. ``f30!}.} and Judge of\index{Jesus:judge:of world ``f10!} the world:\footnote{\ Acts 17:31\index{44 Acts:ch.  17, v.  31, pg. ``f30!}.} unto whom He did from all eternity\index{God's attributes:eternal ``f10!} give a people, to be His seed\index{Election:God's seed ``f10!},\footnote{\ John 17:6\index{43 John:ch.  17, v.   6, pg. ``f30!}; Psalm 22:30\index{19 Psalms:ch.  22, v.  30, pg. ``f30!}; Isaiah 53:10\index{23 Isaiah:ch.  53, v.  10, pg. ``f30!}.} and to be by Him in time redeemed\index{Redemption:purchased by Christ ``f10!}, called, justified\index{Justification:justified ``f10!}, sanctified\index{Sanctification:sanctification ``f10!}, and glorified\index{Resurrection:believers:into glory:glorification ``f10!}.\footnote{\ 1 Timothy 2:6\index{54 1 Timothy:ch.   2, v.   6, pg. ``f30!}; Isaiah 55:4-5\index{23 Isaiah:ch.  55, v.   4-5, pg. ``f30!}; 1 Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg. ``f30!}.}   
\bigskip
8.2\ \ The Son\index{Jesus:Son of God ``f10!} of God, the second person in\index{Jesus:second person in Trinity ``f10!} the Trinity\index{God's attributes:trinity:one substance ``f10!}, being very and eternal God\index{God's attributes:eternal ``f10!}, of one substance and equal with the\index{God's attributes:trinity:persons equal ``f10!} Father\index{Father:God the ``f10!}, did, when the fulness of time\index{Jesus:incarnation:in fulness of time ``f10!} was come, take upon Him man's nature,\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}, 14\index{43 John:ch.   1, v.   1, 14, pg. ``f30!}; 1 John 5:20\index{62 1 John:ch.   5, v.  20, pg. ``f30!}; Philippians 2:6\index{50 Philippians:ch.   2, v.   6, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} with all the essential properties and\index{Jesus:human nature:essential properties ``f10!} common infirmities thereof\index{Jesus:humiliation:suffered:infirmities ``f10!}, yet without sin:\footnote{\ Hebrews 2:1\index{58 Hebrews:ch.   2, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   2, v.  14, pg. ``f30!}, 16-17\index{58 Hebrews:ch.   2, v.  14, 16-17, pg. ``f30!}; Hebrews 4:15\index{58 Hebrews:ch.   4, v.  15, pg. ``f30!}.} being conceived by\index{Jesus:conceived:by the power of the Holy Spirit ``f10!} the power of the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, in the womb\index{Jesus:conceived:in womb of virgin Mary ``f10!} of the virgin\index{Jesus:born of virgin Mary ``f10!} Mary\index{Mary, mother of Jesus ``f10!}, of her substance\index{Jesus:conceived:of Mary's substance ``f10!}.\footnote{\ Luke 1:27, 31, 35\index{42 Luke:ch.   1, v.  27, 31, 35, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} So that two whole, perfect, and distinct natures\index{Jesus:two distinct natures ``f10!}, the Godhead\index{God's attributes:Godhead ``f10!} and the manhood\index{Jesus:human nature:manhood ``f10!}, were inseparably\index{Jesus:human nature:inseparable from divine, one person ``f10!} joined together in one person, without conversion, composition\index{God's attributes:being:no composition ``f10!}, or confusion\index{God's attributes:being:no confusion ``f10!}.\footnote{\ Luke 1:35\index{42 Luke:ch.   1, v.  35, pg. ``f30!}; Colossians 2:9\index{51 Colossians:ch.   2, v.   9, pg. ``f30!}; Romans 9:5\index{45 Romans:ch.   9, v.   5, pg. ``f30!}; 1 Peter 3:18\index{60 1 Peter:ch.   3, v.  18, pg. ``f30!}; 1 Timothy 3:1\index{54 1 Timothy:ch.   3, v.   1, pg. ``f30!}6\index{54 1 Timothy:ch.   3, v.  16, pg. ``f30!}.} Which person is very God, and very man, yet one Christ, the only Mediator\index{Jesus:mediator:of covenant ``f10!} between God and man.\footnote{\ Romans 1:3-4\index{45 Romans:ch.   1, v.   3-4, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}.}   
\bigskip
8.3\ \ The Lord Jesus, in His human nature thus\index{Jesus:human nature:united with divine ``f10!} united to the divine, was sanctified\index{Sanctification:sanctification ``f10!} and anointed\index{Jesus:anointed with Holy Spirit ``f10!} with the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}, above measure,\footnote{\ Psalm 45:7\index{19 Psalms:ch.  45, v.   7, pg. ``f30!}; John 3:3\index{43 John:ch.   3, v.   3, pg. ``f30!}4\index{43 John:ch.   3, v.  34, pg. ``f30!}.} having in Him all the treasures of wisdom and\index{Jesus:wisdom, treasures of ``f10!} knowledge;\index{Jesus:human nature:all knowledge ``f10!}\footnote{\ Colossians 2:3\index{51 Colossians:ch.   2, v.   3, pg. ``f30!}.} in whom it pleased\index{Providence:at God's pleasure ``f10!} the Father\index{Father:God the ``f10!} that all fulness should\index{Jesus:fulness of Godhead dwells ``f10!} dwell;\footnote{\ Colossians 1:19\index{51 Colossians:ch.   1, v.  19, pg. ``f30!}.} to the end that, being holy\index{Jesus:holy ``f10!}, harmless\index{Jesus:harmless ``f10!}, undefiled\index{Jesus:undefiled ``f10!}, and full of grace\index{Jesus:full of grace ``f10!} and truth\index{God's attributes:truth ``f10!},\footnote{\ Hebrews 7:26\index{58 Hebrews:ch.   7, v.  26, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}4\index{43 John:ch.   1, v.  14, pg. ``f30!}.} He might be thoroughly furnished\index{Jesus:mediator:fully furnished to execute office ``f10!} to execute the office\index{Jesus:offices ``f10!} of a mediator\index{Jesus:mediator:of covenant ``f10!} and surety\index{Jesus:surety for the elect ``f10!}.\footnote{\ Acts 10:38\index{44 Acts:ch.  10, v.  38, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}4\index{58 Hebrews:ch.  12, v.  24, pg. ``f30!}; Hebrews 7:22\index{58 Hebrews:ch.   7, v.  22, pg. ``f30!}.} Which office He took not unto Himself, but was thereunto called by His Father,\footnote{\ Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}{}-5\index{58 Hebrews:ch.   5, v.   4-5, pg. ``f30!}.} who put all power\index{Jesus:almighty power ``f10!} and judgment into\index{Jesus:judge:assigned by Father ``f10!} His hand, and gave Him commandment\index{Jesus:commanded by the Father ``f10!} to execute the same\index{Jesus:obedience:to God's law ``f10!}.\footnote{\ John 5:22, 27\index{43 John:ch.   5, v.  22, 27, pg. ``f30!}; Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}; Acts 2:36\index{44 Acts:ch.   2, v.  36, pg. ``f30!}.} 
\bigskip
8.4\ \ This office\index{Jesus:offices ``f10!} the Lord Jesus did most willingly\index{Jesus:humiliation:willingly undertaken ``f10!} undertake;\footnote{\ Psalm 40:7-8\index{19 Psalms:ch.  40, v.   7-8, pg. ``f30!} with Hebrews 10:5-10\index{58 Hebrews:ch.  10, v.   5-10, pg. ``f30!}; John 10:18\index{43 John:ch.  10, v.  18, pg. ``f30!}; Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg. ``f30!}.} which that He might discharge, He\index{Jesus:discharged:his duties ``f10!} was made under the law,\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}.} and did perfectly fulfil\index{Jesus:obedience:to God's law ``f10!} it,\footnote{\ Matthew 3:15\index{40 Matthew:ch.   3, v.  15, pg. ``f30!}; Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}.} endured most grievous torments immediately\index{Jesus:humiliation:suffered:torments ``f10!} in His soul,\footnote{\ Matthew 26:37-38\index{40 Matthew:ch.  26, v.  37-38, pg. ``f30!}; Luke 22:44\index{42 Luke:ch.  22, v.  44, pg. ``f30!}; Matthew 27:4\index{40 Matthew:ch.  27, v.   4, pg. ``f30!}6\index{40 Matthew:ch.  27, v.  46, pg. ``f30!}.} and most painful\index{Jesus:humiliation:suffered pain ``f10!} sufferings in\index{Jesus:death:sufferings and death ``f10!} His body;\footnote{\ Matthew Chapters 26-27.} was crucified\index{Jesus:crucified ``f10!}, and died\index{Jesus:died ``f10!};\footnote{\ Philippians 2:8\index{50 Philippians:ch.   2, v.   8, pg. ``f30!}.} was buried\index{Jesus:buried ``f10!}, and remained under the power of death\index{Jesus:death:under power of ``f10!}; yet saw no corruption\index{Jesus:death:no corruption ``f10!}.\footnote{\ Acts. 2:23-24, 27; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}7\index{44 Acts:ch.  13, v.  37, pg. ``f30!}; Romans 6:9\index{45 Romans:ch.   6, v.   9, pg. ``f30!}.} On the third day\index{Jesus:resurrection:third day ``f10!} He arose\index{Jesus:resurrection:historical fact ``f10!} from the dead\index{Jesus:resurrection:from the dead ``f10!},\footnote{\ 1 Corinthians 15:3-4\index{46 1 Corinthians:ch.  15, v.   3-4, pg. ``f30!}.} with the same body in which He suffered\index{Jesus:resurrection:body same:in which he suffered ``f10!},\footnote{\ John 20:25, 27\index{43 John:ch.  20, v.  25, 27, pg. ``f30!}.} with which also he ascended\index{Jesus:ascension:into heaven ``f10!} into heaven, and there sits\index{Jesus:in heaven:sitting at right hand of Fathe ``f10!} at the right hand of His Father\index{Father:God the ``f10!},\footnote{\ Mark 16:19\index{41 Mark:ch.  16, v.  19, pg. ``f30!}.} making intercession\index{Jesus:in heaven:intercession made for all believers ``f10!},\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}; Hebrews 9:24\index{58 Hebrews:ch.   9, v.  24, pg. ``f30!}; Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}.} and shall return to judge men\index{Jesus:judge:of mankind ``f10!} and angels\index{Angels:angels ``f10!} at the end of the world.\footnote{\ Romans 14:9\index{45 Romans:ch.  14, v.   9, pg. ``f30!}{}-10\index{45 Romans:ch.  14, v.   9-10, pg. ``f30!}; Acts 1:11\index{44 Acts:ch.   1, v.  11, pg. ``f30!}; Acts 10:42\index{44 Acts:ch.  10, v.  42, pg. ``f30!}; Matthew 13:40, 41-42\index{40 Matthew:ch.  13, v.  40, 41-42, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg. ``f30!}; 2 Peter 2:4\index{61 2 Peter:ch.   2, v.   4, pg. ``f30!}.}   
\bigskip
8.5\ \ The Lord Jesus, by His perfect obedience\index{Jesus:obedience:to God's law ``f10!}, and sacrifice of Himself\index{Jesus:sacrifice:of himself ``f10!}, which He, through the eternal\index{Holy Spirit:eternal ``f10!} Spirit, once offered up unto God, has fully satisfied the justice of His\index{Jesus:satisfied divine justice ``f10!} Father\index{Father:God the ``f10!};\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}9\index{45 Romans:ch.   5, v.  19, pg. ``f30!}; Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}, 16\index{58 Hebrews:ch.   9, v.  14, 16, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.  10, v.  14, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}5-26\index{45 Romans:ch.   3, v.  25-26, pg. ``f30!}.} and purchased\index{Redemption:purchased by Christ ``f10!}, not only reconciliation\index{Jesus:purchased reconciliation ``f10!}, but an everlasting inheritance\index{Believers:inheritance:everlasting ``f10!} in the kingdom of heaven\index{Kingdom:of heaven ``f10!}, for all those whom the Father has given unto Him.\footnote{\ Daniel 9:24, 26\index{27 Daniel:ch.   9, v.  24, 26, pg. ``f30!}; Colossians 1:19\index{51 Colossians:ch.   1, v.  19, pg. ``f30!}{}-20\index{51 Colossians:ch.   1, v.  19-20, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}1\index{49 Ephesians:ch.   1, v.  11, pg. ``f30!}, 14\index{49 Ephesians:ch.   1, v.  11, 14, pg. ``f30!}; John 17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}; Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!}, 15\index{58 Hebrews:ch.   9, v.  12, 15, pg. ``f30!}.}   
\bigskip
8.6\ \ Although the work of redemption\index{Redemption:work of by Christ ``f10!} was not actually wrought by Christ\index{Jesus:redeemer of God's elect ``f10!} till after His incarnation\index{Jesus:incarnation:appearing in our nature ``f10!}, yet the virtue\index{Jesus:vicarious atonement ``f10!}, efficacy, and\index{Redemption:efficacy of Christ's incarnation ``f10!} benefits thereof were\index{Believers:benefits:of redemption ``f10!} communicated unto the elect\index{Election:elect of God ``f10!} in all ages successively\index{Redemption:effectually:communicated ``f10!} from the beginning of the world, in and by those promises\index{Salvation:promises made ``f10!}, types\index{types, OT ``f10!}, and sacrifices\index{Ceremonial law:sacrifices ``f10!}, wherein He was revealed, and\index{Revelation:God's word ``f10!} signified to be the seed of the\index{Jesus:seed of the woman ``f10!} woman which should bruise\index{Jesus:bruised serpent's head ``f10!} the serpent\index{Satan:serpent, head to be bruised ``f10!}{}'s head; and the Lamb\index{Jesus:Lamb slain ``f10!} slain\index{Jesus:slain from beginning of world ``f10!} from the beginning of the world: being yesterday\index{Jesus:unchanging ``f10!} and to-day the same, and forever.\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}; Genesis 3:15\index{01 Genesis:ch.   3, v.  15, pg. ``f30!}; Revelation 13:8\index{66 Revelation:ch.  13, v.   8, pg. ``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}.}   
\bigskip
8.7\ \ Christ, in the work of mediation\index{Jesus:mediator:of covenant ``f10!}, acts according to both natures\index{Jesus:two distinct natures ``f10!}, by each nature doing that which is proper to itself:\footnote{\ Hebrews 9:14\index{58 Hebrews:ch.   9, v.  14, pg. ``f30!}; 1 Peter 3:18\index{60 1 Peter:ch.   3, v.  18, pg. ``f30!}.} yet, by reason of the unity\index{Jesus:human nature:united with divine ``f10!} of the person, that which is proper to one nature, is sometimes in Scripture attributed to the person\index{Jesus:one person, two natures ``f10!} denominated by the other nature.\footnote{\ Acts 20:28\index{44 Acts:ch.  20, v.  28, pg. ``f30!}; John 3:13\index{43 John:ch.   3, v.  13, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}6\index{62 1 John:ch.   3, v.  16, pg. ``f30!}.}   
\bigskip
 8.8\ \ To all those for whom Christ has purchased\index{Redemption:purchased by Christ ``f10!} redemption, He does certainly and effectually apply\index{Redemption:effectually:applied ``f10!} and communicate the same,\footnote{\ John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}, 39\index{43 John:ch.   6, v.  37, 39, pg. ``f30!}; John 10:15-16\index{43 John:ch.  10, v.  15-16, pg. ``f30!}.} making intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} for them,\footnote{\ 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}4\index{45 Romans:ch.   8, v.  34, pg. ``f30!}.} and revealing unto\index{Revelation:God's word ``f10!} them, in and by the Word, the\index{Scripture:word of God:for salvation ``f10!} mysteries of\index{Salvation:mystery of ``f10!} salvation\index{Salvation:belief in Jesus by faith ``f10!},\footnote{\ John 15:13, 15\index{43 John:ch.  15, v.  13, 15, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}{}-9\index{49 Ephesians:ch.   1, v.   7-9, pg. ``f30!}; John 17:6\index{43 John:ch.  17, v.   6, pg. ``f30!}.} effectually persuading\index{Effectual call:persuading to believe and obey ``f10!} them by His Spirit to\index{Holy Spirit:applies:faith ``f10!} believe and obey, and\index{God's law:obligation to obey:in all things ``f10!} governing\index{Providence:governing all ``f10!} their hearts by\index{Heart:new:governed by Holy Spirit ``f10!} His Word and Spirit\index{Revelation:Word and Spirit ``f10!};\footnote{\ John 14:26\index{43 John:ch.  14, v.  26, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}; 2 Corinthians 4:13\index{47 2 Corinthians:ch.   4, v.  13, pg. ``f30!}; Romans 8:9\index{45 Romans:ch.   8, v.   9, pg. ``f30!}, 14\index{45 Romans:ch.   8, v.   9, 14, pg. ``f30!}; Romans 15:18-19\index{45 Romans:ch.  15, v.  18-19, pg. ``f30!}; John 17:17\index{43 John:ch.  17, v.  17, pg. ``f30!}.} overcoming\index{Jesus:conqueror of his and our enemies ``f10!} all their enemies\index{Enemies:of believers overcome ``f10!} by His almighty\index{God's attributes:almighty power ``f10!} power and wisdom, in\index{God's attributes:wisdom ``f10!} such manner, and ways, as are most consonant to His wonderful and unsearchable\index{Providence:unsearchable wise counsel ``f10!} dispensation\index{Providence:God's dispensation ``f10!}.\footnote{\ Psalm 110:1\index{19 Psalms:ch. 110, v.   1, pg. ``f30!}; 1 Corinthians 15:25\index{46 1 Corinthians:ch.  15, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  15, v.  25-26, pg. ``f30!}; Malachi 4:2-3\index{39 Malachi:ch.   4, v.   2-3, pg. ``f30!}; Colossians 2:15\index{51 Colossians:ch.   2, v.  15, pg. ``f30!}.}  

\section{Chapter 9 -- Of Free Will} \index{Causation:will of creatures:free ``f10!}9.1\ \ God has endued the\index{Causation:will of creatures:natural liberty ``f10!} will of man with that natural liberty, that\index{Liberty:natural ``f10!} is neither forced, nor by any absolute\index{Mankind:human will, natural liberty ``f10!} necessity of\index{Liberty:not absolute necessity to good or evil ``f10!} nature determined to good or evil.\footnote{\ Matthew 17:12\index{40 Matthew:ch.  17, v.  12, pg. ``f30!}; James 1:14\index{59 James:ch.   1, v.  14, pg. ``f30!}; Deuteronomy 30:19\index{05 Deuteronomy:ch.  30, v.  19, pg. ``f30!}.}   
\bigskip
9.2\ \ Man, in his state of innocency, had\index{Mankind:created:state of innocence ``f10!} freedom and\index{Causation:will of creatures:free ``f10!} power to will and to do that which was good, and well pleasing\index{Mankind:created:initially able to please God ``f10!} to God;\footnote{\ Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}; Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}.} but yet, mutably\index{Mankind:created:mutable ``f10!}, so that he might fall\index{Sin:original:first transgression (of Adam) ``f10!} from it.\footnote{\ Genesis 2:16-17\index{01 Genesis:ch.   2, v.  16-17, pg. ``f30!}; Genesis 3:6\index{01 Genesis:ch.   3, v.   6, pg. ``f30!}.}   
\bigskip
9.3\ \ Man, by his fall\index{Sin:original:first transgression (of Adam) ``f10!} into a state of sin, has wholly lost all ability of\index{Sin:original:inability to:do anything good ``f10!} will to any spiritual good\index{Sin:original:lost all spiritual good ``f10!} accompanying salvation\index{Salvation:belief in Jesus by faith ``f10!}:\footnote{\ Romans 5:6\index{45 Romans:ch.   5, v.   6, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7, pg. ``f30!}; John 15:5\index{43 John:ch.  15, v.   5, pg. ``f30!}.} so as, a natural man, being altogether averse from that good,\footnote{\ Romans 3:10\index{45 Romans:ch.   3, v.  10, pg. ``f30!}, 12\index{45 Romans:ch.   3, v.  10, 12, pg. ``f30!}.} and dead in sin\index{Death:spiritual, in sin ``f10!},\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}, 5\index{49 Ephesians:ch.   2, v.   1, 5, pg. ``f30!}; Colossians 2:13\index{51 Colossians:ch.   2, v.  13, pg. ``f30!}.} is not able, by his own strength, to\index{Conversion:not by own strength ``f10!} convert himself\index{Conversion:not of self ``f10!}, or to prepare himself thereunto.\footnote{\ John 6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}, 65\index{43 John:ch.   6, v.  44, 65, pg. ``f30!}; Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}{}-5\index{49 Ephesians:ch.   2, v.   2-5, pg. ``f30!}; 1 Corinthians 2:14\index{46 1 Corinthians:ch.   2, v.  14, pg. ``f30!}; Titus 3:3-5\index{56 Titus:ch.   3, v.   3-5, pg. ``f30!}.}    
\bigskip
9.4\ \ When God converts\index{Conversion:conversion ``f10!} a sinner, and translates\index{Conversion:translated into state of grace ``f10!} him into the state of grace\index{Grace:state:translation into at conversion ``f10!}, He frees\index{Believers:freed from:guilt of sin ``f10!} him from his natural bondage under\index{Sin:bondage to ``f10!} sin;\footnote{\ Colossians 1:13\index{51 Colossians:ch.   1, v.  13, pg. ``f30!}; John 8:34\index{43 John:ch.   8, v.  34, pg. ``f30!}, 36\index{43 John:ch.   8, v.  34, 36, pg. ``f30!}.} and, by His grace\index{Grace:God's free ``f10!} alone, enables\index{Sanctification:enabled by:grace ``f10!} him freely\index{Causation:will of creatures:free ``f10!} to will and to do that which is spiritually good;\index{Good works:spiritual services ``f10!}\footnote{\ Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Romans 6:18, 22\index{45 Romans:ch.   6, v.  18, 22, pg. ``f30!}.} yet so, as that by reason of his remaining corruption, he\index{Sanctification:remnants of corruption remain ``f10!} does not perfectly, nor only, will that which is good, but does also will that which is evil.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}; Romans 7:15, 18\index{45 Romans:ch.   7, v.  15, 18, pg. ``f30!}{}-19, 21, 23\index{45 Romans:ch.   7, v.  15, 18-19, 21, 23, pg. ``f30!}.}    
\bigskip
9.5\ \ The will of man is made perfectly and immutably\index{God's attributes:immutable ``f10!} free to do good alone, in the state of glory\index{Resurrection:believers:into glory:light ``f10!} only.\footnote{\ Ephesians 4:13\index{49 Ephesians:ch.   4, v.  13, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}3\index{58 Hebrews:ch.  12, v.  23, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg. ``f30!}; Jude 24\index{65 Jude:v.  24, pg. ``f30!}.}  

\section{Chapter 10 -- Of Effectual Calling} \index{Effectual call:calling ``f10!}10.1\ \ All those whom God has predestinated\index{Predestination:predestination ``f10!} unto life, and those only, He is pleased\index{Providence:at God's pleasure ``f10!} in His appointed and accepted time\index{Salvation:at appointed time ``f10!} effectually to\index{Effectual call:calling ``f10!} call,\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}; Romans 11:7\index{45 Romans:ch.  11, v.   7, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}0-11\index{49 Ephesians:ch.   1, v.  10-11, pg. ``f30!}.} by His Word and Spirit\index{Revelation:Word and Spirit ``f10!},\footnote{\ 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}{}-14\index{53 2 Thessalonians:ch.   2, v.  13-14, pg. ``f30!}; 2 Corinthians 3:3, 6\index{47 2 Corinthians:ch.   3, v.   3, 6, pg. ``f30!}.} out of that state of sin and death\index{Death:spiritual, in sin ``f10!}, in which they are by nature, to grace\index{Grace:of salvation ``f10!} and salvation\index{Salvation:belief in Jesus by faith ``f10!} by Jesus Christ;\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}{}-5\index{49 Ephesians:ch.   2, v.   1-5, pg. ``f30!}; 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.   9, pg. ``f30!}{}-10\index{55 2 Timothy:ch.   1, v.   9-10, pg. ``f30!}.} enlightening\index{Effectual call:enlightening minds spiritually ``f10!} their minds spiritually and savingly to understand the things of God,\footnote{\ Acts 26:18\index{44 Acts:ch.  26, v.  18, pg. ``f30!}; 1 Corinthians 2:10, 12\index{46 1 Corinthians:ch.   2, v.  10, 12, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}7-18\index{49 Ephesians:ch.   1, v.  17-18, pg. ``f30!}.} taking away their heart of stone\index{Heart:stone ``f10!}, and giving unto them a heart of flesh\index{Heart:flesh, sinful nature ``f10!};\footnote{\ Ezekiel 36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}.} renewing the\index{Sanctification:renewed in image of God ``f10!}ir\index{Conversion:renewed by Holy Spirit ``f10!} wills, and, by\index{Conversion:renewal of the will ``f10!} His almighty\index{God's attributes:almighty power ``f10!} power determining them\index{Predestination:determining wills ``f10!} to that which is good,\footnote{\ Ezekiel 11:19\index{26 Ezekiel:ch.  11, v.  19, pg. ``f30!}; Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Deuteronomy 30:6\index{05 Deuteronomy:ch.  30, v.   6, pg. ``f30!}; Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}.} and effectually drawing them\index{Effectual call:drawing to Christ ``f10!} to Jesus Christ:\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}9\index{49 Ephesians:ch.   1, v.  19, pg. ``f30!}; John 6:44\index{43 John:ch.   6, v.  44, pg. ``f30!}{}-45\index{43 John:ch.   6, v.  44-45, pg. ``f30!}.} yet so, as they come most freely\index{Effectual call:enables person to:freely answer call ``f10!}, being made willing by His grace\index{Grace:God's free ``f10!}.\footnote{\ Song of Solomon 1:4\index{22 Song of Solomon:ch.   1, v.   4, pg. ``f30!}; Psalm 110:3\index{19 Psalms:ch. 110, v.   3, pg. ``f30!}; John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}; Romans 6:16-18\index{45 Romans:ch.   6, v.  16-18, pg. ``f30!}.}   
\bigskip
10.2\ \ This effectual call\index{Effectual call:calling ``f10!} is of God's free and special grace\index{Grace:God's free ``f10!} alone, not from anything at all foreseen\index{Predestination:not based on God's foresight:of anything in man ``f10!} in man,\footnote{\ 2 Timothy 1:9\index{55 2 Timothy:ch.   1, v.   9, pg. ``f30!}; Titus 3:4-5\index{56 Titus:ch.   3, v.   4-5, pg. ``f30!}; Ephesians 2:4-5, 8-9\index{49 Ephesians:ch.   2, v.   4-5, 8-9, pg. ``f30!}; Romans 9:11\index{45 Romans:ch.   9, v.  11, pg. ``f30!}.} who is altogether passive\index{Effectual call:mankind passive ``f10!} therein, until being quickened\index{Conversion:quickened to new life ``f10!} and renewed by\index{Conversion:renewed by Holy Spirit ``f10!} the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!},\footnote{\ 1 Corinthians 2:14\index{46 1 Corinthians:ch.   2, v.  14, pg. ``f30!}; Romans 8:7\index{45 Romans:ch.   8, v.   7, pg. ``f30!}; Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5, pg. ``f30!}.} he is thereby enabled to answer\index{Effectual call:enables person to:freely answer call ``f10!} this call, and to embrace the grace\index{Grace:in believers:by Holy Spirit ``f10!} offered and\index{Effectual call:enables person to:embrace grace ``f10!} conveyed in it.\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}; John 5:25\index{43 John:ch.   5, v.  25, pg. ``f30!}; John 6:37\index{43 John:ch.   6, v.  37, pg. ``f30!}.}   
\bigskip
10.3\ \ Elect\index{Infants:elect, dying in infancy ``f10!} infants, dying in infancy, are regenerated\index{Regeneration:regeneration ``f10!}, and saved by Christ\index{Salvation:by Christ ``f10!} through the Spirit\index{Holy Spirit:applies:regeneration ``f10!},\footnote{\ Luke 18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}5\index{42 Luke:ch.  18, v.  15, pg. ``f30!}{}-16\index{42 Luke:ch.  18, v.  15-16, pg. ``f30!} and Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!}; John 3:3\index{43 John:ch.   3, v.   3, pg. ``f30!}, 5\index{43 John:ch.   3, v.   3, 5, pg. ``f30!}; 1 John 5:12\index{62 1 John:ch.   5, v.  12, pg. ``f30!}; and Romans 8:9\index{45 Romans:ch.   8, v.   9, pg. ``f30!} compared.} who works\index{Providence:God's works ``f10!} when, and where, and how He pleases\index{Providence:at God's pleasure ``f10!}:\footnote{\ John 3:8\index{43 John:ch.   3, v.   8, pg. ``f30!}.} so also, are all other elect\index{Election:elect of God ``f10!} persons who are uncapable\index{Infants:incapable of being outwardly called ``f10!} of being outwardly called\index{Preaching:outward call ``f10!} by the ministry of the Word\index{Preaching:ministry of the Word ``f10!}.\footnote{\ 1 John 5:12\index{62 1 John:ch.   5, v.  12, pg. ``f30!}; Acts 4:12\index{44 Acts:ch.   4, v.  12, pg. ``f30!}.}   
\bigskip
10.4\ \ Others, not elected\index{Election:some not chosen ``f10!}, although they may be called by the ministry of the Word\index{Preaching:ministry of the Word ``f10!},\footnote{\ Matthew 22:14\index{40 Matthew:ch.  22, v.  14, pg. ``f30!}.} and may have some common operations\index{Holy Spirit:operations:common ``f10!} of the Spirit,\footnote{\ Matthew 7:22\index{40 Matthew:ch.   7, v.  22, pg. ``f30!}; Matthew 13:20-21\index{40 Matthew:ch.  13, v.  20-21, pg. ``f30!}; Hebrews 6:4-5\index{58 Hebrews:ch.   6, v.   4-5, pg. ``f30!}.} yet they never truly\index{Unbelievers:non-elect never truly come to Jesus ``f10!} come unto Christ, and therefore cannot be saved:\index{Salvation:not in other religions ``f10!}\footnote{\ John 6:64-66\index{43 John:ch.   6, v.  64-66, pg. ``f30!}; John 8:24\index{43 John:ch.   8, v.  24, pg. ``f30!}.} much less can men, not professing the\index{Profession:absent ``f10!} Christian religion\index{Profession:of false religion ``f10!}, be saved in any other way whatsoever, be they never so diligent to\index{God's law:diligence in application ``f10!} frame their\index{Salvation:not by:obedience ``f10!} lives according to the light of nature, and the law of that religion they do profess.\footnote{\ Acts 4:12\index{44 Acts:ch.   4, v.  12, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}2\index{49 Ephesians:ch.   2, v.  12, pg. ``f30!}; John 4:22\index{43 John:ch.   4, v.  22, pg. ``f30!}; John 14:6\index{43 John:ch.  14, v.   6, pg. ``f30!}; John 17:3\index{43 John:ch.  17, v.   3, pg. ``f30!}.} And to assert and maintain that they may, is very pernicious\index{Salvation:outside of Christ, pernicious and detestable claim ``f10!}, and to be detested.\footnote{\ 2 John 9-11\index{63 2 John:v. 9-11, pg. ``f30!}; 1 Corinthians 16:2\index{46 1 Corinthians:ch.  16, v.   2, pg. ``f30!}2\index{46 1 Corinthians:ch.  16, v.  22, pg. ``f30!}; Galatians 1:6-8\index{48 Galatians:ch.   1, v.   6-8, pg. ``f30!}.}  

\section{Chapter 11 -- Of Justification} \index{Justification:justification ``f10!}11.1\ \ Those whom God effectually calls\index{Effectual call:calling ``f10!}, He also freely justifies\index{Justification:freely by God ``f10!};\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}.} not by infusing righteousness into\index{Jesus:righteousness imputed to believers ``f10!} them, but by pardoning\index{Justification:pardoned of sin ``f10!} their sins, and by accounting and\index{Justification:accounted by God as righteous ``f10!} accepting their\index{Justification:accepted by God as righteous ``f10!} persons as righteous, not for anything wrought in them\index{Salvation:not by:works ``f10!}, or done by them, but for Christ's sake alone; nor by imputing\index{Imputation:of Christ's righteousness by faith ``f10!} faith itself, the act of believing, or\index{Effectual call:not based on act of believing ``f10!} any other evangelical obedience\index{God's law:commandments:obedience:required ``f10!} to them, as their righteousness, but by imputing the obedience and\index{Jesus:obedience:imputed to believers ``f10!} satisfaction of Christ\index{Jesus:satisfied divine justice ``f10!} unto them,\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}2, 24-25\index{45 Romans:ch.   3, v.  22, 24-25, pg. ``f30!}, 27-28\index{45 Romans:ch.   3, v.  22, 24-25, 27-28, pg. ``f30!}; Romans 4:5\index{45 Romans:ch.   4, v.   5, pg. ``f30!}{}-8\index{45 Romans:ch.   4, v.   5-8, pg. ``f30!}; 2 Corinthians 5:19\index{47 2 Corinthians:ch.   5, v.  19, pg. ``f30!}, 21\index{47 2 Corinthians:ch.   5, v.  19, 21, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}, 7\index{56 Titus:ch.   3, v.   5, 7, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}; Jeremiah 23:6\index{24 Jeremiah:ch.  23, v.   6, pg. ``f30!}; 1 Corinthians 1:30\index{46 1 Corinthians:ch.   1, v.  30, pg. ``f30!}{}-31\index{46 1 Corinthians:ch.   1, v.  30-31, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}7-19\index{45 Romans:ch.   5, v.  17-19, pg. ``f30!}.} they receiving and\index{Faith:instrument:for receiving Christ ``f10!} resting on\index{Faith:resting on Christ ``f10!} Him and His righteousness\index{Jesus:righteous ``f10!} by faith; which\index{Faith:in Jesus Christ ``f10!} faith they have not of themselves, it is the gift of God\index{Faith:gift of God ``f10!}.\footnote{\ Acts 10:43\index{44 Acts:ch.  10, v.  43, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Philippians 3:19\index{50 Philippians:ch.   3, v.  19, pg. ``f30!}; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}8-39\index{44 Acts:ch.  13, v.  38-39, pg. ``f30!}; Ephesians 2:7\index{49 Ephesians:ch.   2, v.   7, pg. ``f30!}{}-8\index{49 Ephesians:ch.   2, v.   7-8, pg. ``f30!}.}   
\bigskip
11.2\ \ Faith\index{Faith:saving ``f10!}, thus receiving and\index{Faith:instrument:for receiving Christ ``f10!} resting on\index{Faith:resting on Christ ``f10!} Christ and His righteousness\index{Jesus:righteous ``f10!}, is the alone instrument\index{Faith:instrument:of justification ``f10!} of justification\index{Justification:justification ``f10!};\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}8\index{45 Romans:ch.   3, v.  28, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}.} yet is it not alone in the person justified\index{Justification:justified ``f10!}, but is ever accompanied with all other saving graces\index{Grace:of salvation ``f10!}, and is no dead\index{Faith:not dead when resting on Christ ``f10!} faith, but works\index{Faith:works by love ``f10!} by love.\footnote{\ James 2:17, 22, 26\index{59 James:ch.   2, v.  17, 22, 26, pg. ``f30!}; Galatians 5:6\index{48 Galatians:ch.   5, v.   6, pg. ``f30!}.}   
\bigskip
11.3\ \ Christ, by His obedience\index{Jesus:obedience:to God's law ``f10!} and death, did\index{Jesus:death:death ``f10!} fully discharge the\index{Jesus:discharged:debt of sin for his people ``f10!} debt\index{Forgiveness:of debtors ``f10!} of all those that are thus justified\index{Justification:justified ``f10!}, and did make a proper, real, and full satisfaction to His\index{Jesus:satisfied divine justice ``f10!} Father\index{Father:God the ``f10!}{}'s justice in their behalf.\footnote{\ Romans 5:8-10, 19\index{45 Romans:ch.   5, v.   8-10, 19, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}{}-6\index{54 1 Timothy:ch.   2, v.   5-6, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}0\index{58 Hebrews:ch.  10, v.  10, pg. ``f30!}, 14\index{58 Hebrews:ch.  10, v.  10, 14, pg. ``f30!}; Daniel 9:24, 26\index{27 Daniel:ch.   9, v.  24, 26, pg. ``f30!}; Isaiah 53:4-6, 10-12\index{23 Isaiah:ch.  53, v.   4-6, 10-12, pg. ``f30!}.} Yet, inasmuch as He was given by the Father for them;\footnote{\ Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}2\index{45 Romans:ch.   8, v.  32, pg. ``f30!}.} and His obedience and satisfaction accepted in their stead\index{Jesus:vicarious atonement ``f10!};\footnote{\ 2 Corinthians 5:21\index{47 2 Corinthians:ch.   5, v.  21, pg. ``f30!}; Matthew 3:17\index{40 Matthew:ch.   3, v.  17, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}.} and both freely\index{Salvation:freely offered ``f10!}, not for anything in them; their justification\index{Justification:justification ``f10!} is only of free grace\index{Grace:God's free ``f10!};\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}.} that both the exact justice\index{God's attributes:administers justice ``f10!}, and rich grace\index{Grace:God's rich ``f10!} of God, might be glorified\index{God's attributes:glory:his own ``f10!} in the justification of sinners.\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}6\index{45 Romans:ch.   3, v.  26, pg. ``f30!}; Ephesians 2:7\index{49 Ephesians:ch.   2, v.   7, pg. ``f30!}.}   
\bigskip
11.4\ \ God did, from all eternity\index{God's attributes:eternal ``f10!}, decree\index{Predestination:eternal decrees ``f10!} to justify a\index{Justification:by faith ``f10!}ll\index{Justification:by God's decree ``f10!} the elect\index{Election:elect of God ``f10!},\footnote{\ Galatians 3:8\index{48 Galatians:ch.   3, v.   8, pg. ``f30!}; 1 Peter 1:2\index{60 1 Peter:ch.   1, v.   2, pg. ``f30!}, 19-20\index{60 1 Peter:ch.   1, v.   2, 19-20, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}0\index{45 Romans:ch.   8, v.  30, pg. ``f30!}.} and Christ did, in the fulness of time\index{Jesus:incarnation:in fulness of time ``f10!}, die for\index{Death:punishment for sin ``f10!} their sins, and rise again\index{Jesus:resurrection:for justification ``f10!} for their justification\index{Justification:justification ``f10!}:\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}; 1 Timothy 2:6\index{54 1 Timothy:ch.   2, v.   6, pg. ``f30!}; Romans 4:25\index{45 Romans:ch.   4, v.  25, pg. ``f30!}.} nevertheless, they are not justified\index{Justification:justified ``f10!}, until the Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!} does, in due time, actually apply Christ unto them.\footnote{\ Colossians 1:21\index{51 Colossians:ch.   1, v.  21, pg. ``f30!}{}-22\index{51 Colossians:ch.   1, v.  21-22, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Titus 3:3-7\index{56 Titus:ch.   3, v.   3-7, pg. ``f30!}.}   
\bigskip
11.5\ \ God does continue to forgive the\index{Forgiveness:God forgives sin ``f10!} sins of those that are justified\index{Justification:justified ``f10!}:\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}2\index{40 Matthew:ch.   6, v.  12, pg. ``f30!}; 1 John 1:7\index{62 1 John:ch.   1, v.   7, pg. ``f30!}, 9\index{62 1 John:ch.   1, v.   7, 9, pg. ``f30!}; 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}.} and although they can never fall\index{Assurance:never fall from being saved ``f10!} from the state of justification\index{Justification:justification ``f10!};\footnote{\ Luke 22:32\index{42 Luke:ch.  22, v.  32, pg. ``f30!}; John 10:28\index{43 John:ch.  10, v.  28, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.  10, v.  14, pg. ``f30!}.} yet they may, by their sins, fall under God's fatherly\index{Father:God the ``f10!} displeasure, and not\index{Believers:may experience God's displeasure for sin ``f10!} have the light of His countenance\index{Believers:chastened by God ``f10!} restored unto them, until they humble themselves\index{Humility:before God ``f10!}, confess\index{Confession:of sin ``f10!} their sins, beg pardon\index{Forgiveness:beg pardon for ``f10!}, and renew their faith\index{Faith:renewal of ``f10!} and repentance\index{Repentance:for all sins ``f10!}.\footnote{\ Psalm 89:31-33\index{19 Psalms:ch.  89, v.  31-33, pg. ``f30!}; Psalm 51:7-12\index{19 Psalms:ch.  51, v.   7-12, pg. ``f30!}; Psalm 32:5\index{19 Psalms:ch.  32, v.   5, pg. ``f30!}; Matthew 26:75\index{40 Matthew:ch.  26, v.  75, pg. ``f30!}; 1 Corinthians 11:30, 32\index{46 1 Corinthians:ch.  11, v.  30, 32, pg. ``f30!}; Luke 1:20\index{42 Luke:ch.   1, v.  20, pg. ``f30!}.}  
\bigskip
11.6\ \ The justification\index{Justification:justification ``f10!} of believers under\index{Justification:of believers, same under OT and NT ``f10!} the old testament\index{Scripture:Old Testament ``f10!} was, in all these respects, one and the same with the justification of believers under the new testament\index{Scripture:New Testament ``f10!}.\footnote{\ Galatians 3:9, 13-14\index{48 Galatians:ch.   3, v.   9, 13-14, pg. ``f30!}; Romans 4:22-24\index{45 Romans:ch.   4, v.  22-24, pg. ``f30!}; Hebrews 13:8\index{58 Hebrews:ch.  13, v.   8, pg. ``f30!}.}  

\section{Chapter 12 -- Of Adoption} \index{Adoption:in salvation ``f10!}12.1\ \ All those that are justified\index{Justification:justified ``f10!}, God vouchsafes\index{Justification:God vouchsafes in Jesus ``f10!}, in and for His only Son\index{Jesus:Son of God ``f10!} Jesus Christ, to make partakers of the grace of\index{Grace:of adoption ``f10!} adoption\index{Adoption:in salvation ``f10!}:\footnote{\ Ephesians 1:5\index{49 Ephesians:ch.   1, v.   5, pg. ``f30!}.} by which they are taken into the number, and enjoy the\index{Believers:enjoy:sense of God's love ``f10!} liberties\index{Christian liberty:privileges of children of God ``f10!} and privileges\index{Believers:privileges ``f10!} of the children of God,\footnote{\ Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}7\index{45 Romans:ch.   8, v.  17, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}.} have His name put upon them,\footnote{\ Jeremiah 14:9\index{24 Jeremiah:ch.  14, v.   9, pg. ``f30!}; 2 Corinthians 6:18\index{47 2 Corinthians:ch.   6, v.  18, pg. ``f30!}; Revelation 3:1\index{66 Revelation:ch.   3, v.   1, pg. ``f30!}2\index{66 Revelation:ch.   3, v.  12, pg. ``f30!}.} receive the spirit\index{Justification:receive the spirit of adoption ``f10!} of adoption,\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}5\index{45 Romans:ch.   8, v.  15, pg. ``f30!}.} have access to the\index{Believers:bold access to God's throne ``f10!} throne of grace\index{Grace:throne of ``f10!} with boldness\index{Prayer:with boldness ``f10!},\footnote{\ Ephesians 3:12\index{49 Ephesians:ch.   3, v.  12, pg. ``f30!}; Romans 5:2\index{45 Romans:ch.   5, v.   2, pg. ``f30!}.} are enabled to cry\index{Prayer:cry Abba Father ``f10!}, Abba\index{God's attributes:Abba ``f10!}, Father\index{Father:God the ``f10!},\footnote{\ Galatians 4:6\index{48 Galatians:ch.   4, v.   6, pg. ``f30!}.} are pitied\index{Justification:pitied, protected, provided for ``f10!},\footnote{\ Psalm 103:1\index{19 Psalms:ch. 103, v.   1, pg. ``f30!}3\index{19 Psalms:ch. 103, v.  13, pg. ``f30!}.} protected,\footnote{\ Proverbs 14:26\index{20 Proverbs:ch.  14, v.  26, pg. ``f30!}.} provided for,\index{Believers:protected and provided for by God ``f10!}\footnote{\ Matthew 6:30\index{40 Matthew:ch.   6, v.  30, pg. ``f30!}, 32\index{40 Matthew:ch.   6, v.  30, 32, pg. ``f30!}; 1 Peter 5:7\index{60 1 Peter:ch.   5, v.   7, pg. ``f30!}.} and chastened\index{Believers:chastened by God ``f10!} by Him as by a Father;\footnote{\ Hebrews 12:6\index{58 Hebrews:ch.  12, v.   6, pg. ``f30!}.} yet never cast off\index{Believers:never cast off ``f10!},\footnote{\ Lamentations 3:31\index{25 Lamentations:ch.   3, v.  31, pg. ``f30!}.} but sealed to the\index{Salvation:sealed to the day of redemption ``f10!} day of redemption\index{Redemption:day of ``f10!},\footnote{\ Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}.} and inherit\index{Believers:inheritance:everlasting ``f10!} the promises,\footnote{\ Hebrews 6:12\index{58 Hebrews:ch.   6, v.  12, pg. ``f30!}.} as heirs\index{Believers:heirs (salvation, promises, glory) ``f10!} of everlasting\index{Salvation:everlasting ``f10!} salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ 1 Peter 1:3-4\index{60 1 Peter:ch.   1, v.   3-4, pg. ``f30!}; Hebrews 1:1\index{58 Hebrews:ch.   1, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.   1, v.  14, pg. ``f30!}.}  

\section{Chapter 13 -- Of Sanctification} \index{Sanctification:sanctification ``f10!}13.1\ \ They who are once effectually called\index{Effectual call:calling ``f10!} and regenerated\index{Regeneration:regeneration ``f10!}, having a new heart\index{Heart:new:regenerate ``f10!} and a new spirit created in them\index{Conversion:created in Christ ``f10!}, are further sanctified\index{Sanctification:sanctification ``f10!}, really and personally\index{Sanctification:personal ``f10!}, through the virtue\index{Jesus:vicarious atonement ``f10!} of Christ's death\index{Jesus:death:death ``f10!} and resurrection,\index{Jesus:resurrection:from the dead ``f10!}\footnote{\ 1 Corinthians 6:11\index{46 1 Corinthians:ch.   6, v.  11, pg. ``f30!}; Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}; Philippians 3:10\index{50 Philippians:ch.   3, v.  10, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}{}-6\index{45 Romans:ch.   6, v.   5-6, pg. ``f30!}.} by His Word and Spirit\index{Revelation:Word and Spirit ``f10!} dwelling in\index{Holy Spirit:abiding in believers ``f10!} them:\footnote{\ John 17:17\index{43 John:ch.  17, v.  17, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}6\index{49 Ephesians:ch.   5, v.  26, pg. ``f30!}; 2 Thessalonians 2:13\index{53 2 Thessalonians:ch.   2, v.  13, pg. ``f30!}.} the dominion of the\index{Regeneration:dominion of sin destroyed ``f10!} whole body of sin is destroyed,\footnote{\ Romans 6:6, 14\index{45 Romans:ch.   6, v.   6, 14, pg. ``f30!}.} and the several lusts\index{Sexual immorality:lust ``f10!} thereof are more and more weakened\index{Sanctification:lusts weakened and mortified ``f10!} and mortified\index{Sin:mortified ``f10!};\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}4\index{48 Galatians:ch.   5, v.  24, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}3\index{45 Romans:ch.   8, v.  13, pg. ``f30!}.} and they more and more quickened\index{Conversion:quickened to new life ``f10!} and strengthened in\index{Sanctification:supply of strength from Holy Spirit ``f10!} all saving graces\index{Grace:of salvation ``f10!},\footnote{\ Colossians 1:11\index{51 Colossians:ch.   1, v.  11, pg. ``f30!}; Ephesians 3:16\index{49 Ephesians:ch.   3, v.  16, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  16-19, pg. ``f30!}.} to the practice of true holiness\index{Believers:holy:being perfected ``f10!}, without which no man shall see the Lord.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; Hebrews 12:1\index{58 Hebrews:ch.  12, v.   1, pg. ``f30!}4\index{58 Hebrews:ch.  12, v.  14, pg. ``f30!}.}   
\bigskip
13.2\ \ This sanctification\index{Sanctification:sanctification ``f10!} is throughout, in the whole man;\footnote{\ 1 Thessalonians 5:23\index{52 1 Thessalonians:ch.   5, v.  23, pg. ``f30!}.} yet imperfect in\index{Sanctification:imperfection in believers ``f10!} this life, there abiding still\index{Sin:remnants in believers ``f10!} some remnants\index{Sanctification:remnants of corruption remain ``f10!} of corruption in every part:\footnote{\ 1 John 1:10\index{62 1 John:ch.   1, v.  10, pg. ``f30!}; Romans 7:18\index{45 Romans:ch.   7, v.  18, pg. ``f30!}, 23\index{45 Romans:ch.   7, v.  18, 23, pg. ``f30!}; Philippians 3:12\index{50 Philippians:ch.   3, v.  12, pg. ``f30!}.} whence arises a continual and irreconcilable war; the flesh lusting\index{Sexual immorality:lust ``f10!} against the Spirit, and the Spirit against\index{Holy Spirit:at war with believers' lusts and flesh ``f10!} the flesh.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}; 1 Peter 2:11\index{60 1 Peter:ch.   2, v.  11, pg. ``f30!}.}   
\bigskip
13.3\ \ In which war, although the remaining corruption, for\index{Sanctification:remnants of corruption remain ``f10!} a time, may much prevail\index{Sanctification:imperfection in believers ``f10!};\footnote{\ Romans 7:23\index{45 Romans:ch.   7, v.  23, pg. ``f30!}.} yet through the continual supply\index{Sanctification:supply of strength from Holy Spirit ``f10!} of strength from the sanctifying\index{Sanctification:sanctification ``f10!} Spirit of Christ\index{Holy Spirit:Spirit of Christ ``f10!}, the regenerate\index{Regeneration:regeneration ``f10!} part does overcome\index{Sin:overcoming ``f10!};\footnote{\ Romans 6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}; 1 John 5:4\index{62 1 John:ch.   5, v.   4, pg. ``f30!}; Ephesians 4:15-16\index{49 Ephesians:ch.   4, v.  15-16, pg. ``f30!}.} and so, the saints\index{Believers:saints ``f10!} grow in\index{Sanctification:spiritual growth:in grace ``f10!} grace,\footnote{\ 2 Peter 3:18\index{61 2 Peter:ch.   3, v.  18, pg. ``f30!}; 2 Corinthians 3:18\index{47 2 Corinthians:ch.   3, v.  18, pg. ``f30!}.} perfecting holiness\index{Believers:holy:being perfected ``f10!} in the fear of\index{Fear:of God:reverence ``f10!} God.\footnote{\ 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}.}  

\section{Chapter 14 -- Of Saving Faith} \index{Salvation:belief in Jesus by faith ``f10!}\index{Faith:saving ``f10!}14.1\ \ The grace of faith\index{Faith:given by grace ``f10!}, whereby the elect\index{Election:elect of God ``f10!} are enabled to believe\index{Conversion:Holy Spirit:enables belief ``f10!} to the saving of\index{Salvation:by working of Holy Spirit ``f10!} their souls,\index{Holy Spirit:applies:faith ``f10!}\footnote{\ Hebrews 10:39\index{58 Hebrews:ch.  10, v.  39, pg. ``f30!}.} is the work of the Spirit of Christ\index{Holy Spirit:Spirit of Christ ``f10!} in their hearts;\index{Heart:new:enlightened ``f10!}\footnote{\ 2 Corinthians 4:13\index{47 2 Corinthians:ch.   4, v.  13, pg. ``f30!}; Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}7-19\index{49 Ephesians:ch.   1, v.  17-19, pg. ``f30!}; Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}.} and is ordinarily wrought by the\index{Preaching:convincing and converting sinners ``f10!} ministry of the Word\index{Preaching:ministry of the Word ``f10!}:\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}4\index{45 Romans:ch.  10, v.  14, pg. ``f30!}, 17\index{45 Romans:ch.  10, v.  14, 17, pg. ``f30!}.} by which also, and by the administration\index{Sacraments:grace:administered ``f10!} of the sacraments\index{Sacraments:holy:ordinances ``f10!}, and prayer, it\index{Prayer:act of worship ``f10!} is increased and strengthened\index{Sacraments:strengthening ordinances ``f10!}.\footnote{\ 1 Peter 2:2\index{60 1 Peter:ch.   2, v.   2, pg. ``f30!}; Acts 20:32\index{44 Acts:ch.  20, v.  32, pg. ``f30!}; Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Luke 17:5\index{42 Luke:ch.  17, v.   5, pg. ``f30!}; Romans 1:16\index{45 Romans:ch.   1, v.  16, pg. ``f30!}{}-17\index{45 Romans:ch.   1, v.  16-17, pg. ``f30!}.}   
\bigskip
14.2\ \ By this faith\index{Faith:believes Scripture to be word of God ``f10!}, a Christian believes to\index{Scripture:to be:believed ``f10!} be true whatsoever is revealed in the Word, for\index{Scripture:word of God:standard for:life ``f10!} the authority of God Himself speaking therein;\footnote{\ John 4:42\index{43 John:ch.   4, v.  42, pg. ``f30!}; 1 Thessalonians 2:13\index{52 1 Thessalonians:ch.   2, v.  13, pg. ``f30!}; 1 John 5:10\index{62 1 John:ch.   5, v.  10, pg. ``f30!}; Acts 24:14\index{44 Acts:ch.  24, v.  14, pg. ``f30!}.} and acts differently upon that which each particular passage thereof contains; yielding\index{God's law:commandments:obedience:required ``f10!} obedience to the commands\index{God's law:commandments:ten ``f10!},\footnote{\ Romans 16:2\index{45 Romans:ch.  16, v.   2, pg. ``f30!}6\index{45 Romans:ch.  16, v.  26, pg. ``f30!}.} trembling at the threatenings\index{God's law:threatenings for disobedience ``f10!},\footnote{\ Isaiah 66:2\index{23 Isaiah:ch.  66, v.   2, pg. ``f30!}.} and embracing the promises\index{Salvation:promises made ``f10!} of God for this life, and that which is to come.\footnote{\ Hebrews 11:13\index{58 Hebrews:ch.  11, v.  13, pg. ``f30!}; 1 Timothy 4:8\index{54 1 Timothy:ch.   4, v.   8, pg. ``f30!}.} But the principal\index{Faith:principal acts of ``f10!} acts of saving faith\index{Faith:saving ``f10!} are accepting\index{Salvation:belief in Jesus by faith ``f10!}, receiving, and\index{Faith:instrument:for receiving Christ ``f10!} resting upon\index{Faith:resting on Christ ``f10!} Christ alone for justification\index{Justification:justification ``f10!}, sanctification\index{Sanctification:sanctification ``f10!}, and eternal life\index{eternal life ``f10!}, by virtue of the covenant of grace\index{Covenant:grace:grace ``f10!}.\footnote{\ John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}2\index{43 John:ch.   1, v.  12, pg. ``f30!}; Acts 16:31\index{44 Acts:ch.  16, v.  31, pg. ``f30!}; Galatians 2:20\index{48 Galatians:ch.   2, v.  20, pg. ``f30!}; Acts 15:11\index{44 Acts:ch.  15, v.  11, pg. ``f30!}.}   
\bigskip
14.3\ \ This faith is different in degrees\index{Faith:degrees of ``f10!}, weak or strong;\footnote{\ Hebrews 5:13-14\index{58 Hebrews:ch.   5, v.  13-14, pg. ``f30!}; Romans 4:19-20\index{45 Romans:ch.   4, v.  19-20, pg. ``f30!}; Matthew 6:30\index{40 Matthew:ch.   6, v.  30, pg. ``f30!}; Matthew 8:10\index{40 Matthew:ch.   8, v.  10, pg. ``f30!}.} may be often and many ways assailed\index{Faith:assailed ``f10!}, and weakened\index{Faith:weak ``f10!}, but gets the victory\index{Faith:victorious ``f10!};\footnote{\ Luke 22:31-32\index{42 Luke:ch.  22, v.  31-32, pg. ``f30!}; Ephesians 6:16\index{49 Ephesians:ch.   6, v.  16, pg. ``f30!}; 1 John 5:4\index{62 1 John:ch.   5, v.   4, pg. ``f30!}{}-5\index{62 1 John:ch.   5, v.   4-5, pg. ``f30!}.} growing up in\index{Assurance:growing ``f10!} many to the attainment of a full assurance through\index{Assurance:of faith ``f10!} Christ,\footnote{\ Hebrews 6:11-12\index{58 Hebrews:ch.   6, v.  11-12, pg. ``f30!}; Hebrews 10:22\index{58 Hebrews:ch.  10, v.  22, pg. ``f30!}; Colossians 2:2\index{51 Colossians:ch.   2, v.   2, pg. ``f30!}.} who is both the author\index{Jesus:author of a believer's faith ``f10!} and finisher\index{Jesus:finisher of a believer's faith ``f10!} of our faith.\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}.}  

\section{Chapter 15 -- Of Repentance unto Life} \index{Repentance:unto life ``f10!}15.1\ \ Repentance\index{Repentance:unto life ``f10!} unto life is an evangelical grace\index{Grace:evangelical ``f10!},\footnote{\ Zechariah 12:10\index{38 Zechariah:ch.  12, v.  10, pg. ``f30!}; Acts 11:18\index{44 Acts:ch.  11, v.  18, pg. ``f30!}.} the doctrine whereof\index{Doctrine:Gospel ``f10!} is to be preached by every\index{Preaching:call to repentance ``f10!} minister\index{Church officers:minsters ``f10!} of the Gospel\index{Gospel:gospel ``f10!}, as well as that of faith\index{Faith:in Jesus Christ ``f10!} in Christ.\footnote{\ Luke 24:47\index{42 Luke:ch.  24, v.  47, pg. ``f30!}; Mark 1:15\index{41 Mark:ch.   1, v.  15, pg. ``f30!}; Acts 20:21\index{44 Acts:ch.  20, v.  21, pg. ``f30!}.}   
\bigskip
15.2\ \ By it, a sinner, out of the sight and sense not only of the danger\index{Repentance:apprehension of danger of sin ``f10!}, but also of the filthiness\index{Confession:of sin ``f10!} and odiousness\index{Sin:odious before God ``f10!} of his sins, as contrary to the holy\index{God's law:holy ``f10!} nature and righteous law\index{God's law:righteous requirements ``f10!} of God; and upon the apprehension\index{Repentance:apprehension of God's mercy ``f10!} of his mercy in\index{God's attributes:merciful ``f10!} Christ to such as are penitent\index{Repentance:penitence ``f10!}, so grieves for\index{Sin:grief and hatred of ``f10!}, and hates\index{Sin:to be hated ``f10!} his sins, as to turn from them all unto God,\footnote{\ Ezekiel 18:30-31\index{26 Ezekiel:ch.  18, v.  30-31, pg. ``f30!}; Ezekiel 36:31\index{26 Ezekiel:ch.  36, v.  31, pg. ``f30!}; Isaiah 30:22\index{23 Isaiah:ch.  30, v.  22, pg. ``f30!}; Psalm 51:4\index{19 Psalms:ch.  51, v.   4, pg. ``f30!}; Jeremiah 31:18\index{24 Jeremiah:ch.  31, v.  18, pg. ``f30!}{}-19\index{24 Jeremiah:ch.  31, v.  18-19, pg. ``f30!}; Joel 2:12\index{29 Joel:ch.   2, v.  12, pg. ``f30!}{}-13\index{29 Joel:ch.   2, v.  12-13, pg. ``f30!}; Amos 5:15\index{30 Amos:ch.   5, v.  15, pg. ``f30!}; Psalm 119:128\index{19 Psalms:ch. 119, v. 128, pg. ``f30!}; 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}1\index{47 2 Corinthians:ch.   7, v.  11, pg. ``f30!}.} purposing\index{Repentance:purposing to walk with God ``f10!} and endeavouring to walk with\index{God's law:commandments:obedience:endeavoring ``f10!} Him in all the ways of His commandments\index{God's law:commandments:ten ``f10!}.\footnote{\ Psalm 119:6, 59, 106\index{19 Psalms:ch. 119, v.   6, 59, 106, pg. ``f30!}; Luke 1:6\index{42 Luke:ch.   1, v.   6, pg. ``f30!}; 2 Kings 23:25\index{12 2 Kings:ch.  23, v.  25, pg. ``f30!}.}   
\bigskip
15.3\ \ Although repentance be\index{Repentance:required for pardon ``f10!} not to be rested in, as any satisfaction for sin\index{Jesus:satisfied divine justice ``f10!}, or any cause of the pardon the\index{Forgiveness:God alone pardons ``f10!}reof,\footnote{\ Ezekiel 36:31\index{26 Ezekiel:ch.  36, v.  31, pg. ``f30!}{}-32\index{26 Ezekiel:ch.  36, v.  31-32, pg. ``f30!}; Ezekiel 16:61-63\index{26 Ezekiel:ch.  16, v.  61-63, pg. ``f30!}.} which is the act of God's free grace\index{Grace:God's free ``f10!} in Christ;\footnote{\ Hosea 14:2\index{28 Hosea:ch.  14, v.   2, pg. ``f30!}, 4\index{28 Hosea:ch.  14, v.   2, 4, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}4\index{45 Romans:ch.   3, v.  24, pg. ``f30!}; Ephesians 1:7\index{49 Ephesians:ch.   1, v.   7, pg. ``f30!}.} yet is it of such necessity to\index{Repentance:necessity ``f10!} all sinners, that none may expect pardon without it.\footnote{\ Luke 13:3, 5\index{42 Luke:ch.  13, v.   3, 5, pg. ``f30!}; Acts 17:30-31\index{44 Acts:ch.  17, v.  30-31, pg. ``f30!}.}   
\bigskip
15.4\ \ As there is no sin so small, but it deserves\index{Sin:deserves damnation ``f10!} damnation\index{Punishment:everlasting:damnation ``f10!},\footnote{\ Romans 6:23\index{45 Romans:ch.   6, v.  23, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}; Matthew 12:36\index{40 Matthew:ch.  12, v.  36, pg. ``f30!}.} so there is no sin so great, that it can bring damnation upon those who truly\index{Repentance:for all sins ``f10!} repent.\footnote{\ Isaiah 55:7\index{23 Isaiah:ch.  55, v.   7, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}; Isaiah 1:16, 18\index{23 Isaiah:ch.   1, v.  16, 18, pg. ``f30!}.}   
\bigskip
15.5\ \ Men ought not to content themselves with a general repentance, but\index{Repentance:for all sins ``f10!} it is every man's duty to endeavour\index{Repentance:duty to endeavour ``f10!} to repent of his particular sins, particularly.\footnote{\ Psalm 19:13\index{19 Psalms:ch.  19, v.  13, pg. ``f30!}; Luke 19:8\index{42 Luke:ch.  19, v.   8, pg. ``f30!}; 1 Timothy 1:13, 15\index{54 1 Timothy:ch.   1, v.  13, 15, pg. ``f30!}.}   
\bigskip
15.6\ \ As every man is bound to make private confession\index{Confession:of sin ``f10!} of his sins to God, praying for\index{Prayer:for pardon of sin ``f10!} the pardon the\index{Forgiveness:God alone pardons ``f10!}reof;\index{Forgiveness:beg pardon for ``f10!}\footnote{\ Psalm 51:4\index{19 Psalms:ch.  51, v.   4, pg. ``f30!}{}-5, 7, 9, 14\index{19 Psalms:ch.  51, v.   4-5, 7, 9, 14, pg. ``f30!}; Psalm 32:5\index{19 Psalms:ch.  32, v.   5, pg. ``f30!}{}-6\index{19 Psalms:ch.  32, v.   5-6, pg. ``f30!}.} upon which, and the forsaking\index{Repentance:forsaking sin ``f10!} of them, he shall find mercy:\index{Mercy:on confession ``f10!}\footnote{\ Proverbs 28:13\index{20 Proverbs:ch.  28, v.  13, pg. ``f30!}; 1 John 1:9\index{62 1 John:ch.   1, v.   9, pg. ``f30!}.} so, he that scandalizes\index{Scandalous behaviour:to be confessed ``f10!} his brother, or the Church of Christ\index{Church:Christ's ``f10!}, ought to be willing, by a private or public confession, and sorrow\index{Confession:sorrow for sin ``f10!} for his sin, to declare his\index{Repentance:public declaration ``f10!} repentance to\index{Repentance:to those offended ``f10!} those that are offended,\index{Confession:to those offended ``f10!}\footnote{\ James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}; Luke 17:3-4\index{42 Luke:ch.  17, v.   3-4, pg. ``f30!}; Joshua 7:19\index{06 Joshua:ch.   7, v.  19, pg. ``f30!}; Psalm 51.} who are thereupon to be reconciled\index{Confession:to be reconciled to God ``f10!} to him, and in love to\index{Love toward God:for God ``f10!} receive him.\footnote{\ 2 Corinthians 2:8\index{47 2 Corinthians:ch.   2, v.   8, pg. ``f30!}.}  

\section{Chapter 16 -- Of Good Works} \index{Good works:definition ``f10!}16.1\ \ Good works\index{Good works:definition ``f10!} are only such as God has commanded in His\index{God's law:commandments:ten ``f10!} holy Word\index{Scripture:holy ``f10!},\footnote{\ Micah 6:8\index{33 Micah:ch.   6, v.   8, pg. ``f30!}; Romans 12:2\index{45 Romans:ch.  12, v.   2, pg. ``f30!}; Hebrews 13:21\index{58 Hebrews:ch.  13, v.  21, pg. ``f30!}.} and not such as, without the warrant\index{Good works:not devised by men without warrant ``f10!} thereof, are devised\index{Good works:not devised by men ``f10!} by men, out of blind zeal, or\index{Good works:not out of blind zeal ``f10!} upon any pretence\index{pretence ``f10!} of good intention\index{Good works:not by good intentions ``f10!}.\footnote{\ Matthew 15:9\index{40 Matthew:ch.  15, v.   9, pg. ``f30!}; Isaiah 29:13\index{23 Isaiah:ch.  29, v.  13, pg. ``f30!}; 1 Peter 1:18\index{60 1 Peter:ch.   1, v.  18, pg. ``f30!}; Romans 10:2\index{45 Romans:ch.  10, v.   2, pg. ``f30!}; John 16:2\index{43 John:ch.  16, v.   2, pg. ``f30!}; 1 Samuel 15:21\index{09 1 Samuel:ch.  15, v.  21, pg. ``f30!}{}-23\index{09 1 Samuel:ch.  15, v.  21-23, pg. ``f30!}.}    
\bigskip
16.2\ \ These good works\index{Good works:adorn profession of faith ``f10!}, done in obedience\index{God's law:commandments:obedience:required ``f10!} to God's commandments\index{God's law:commandments:ten ``f10!}, are the fruits\index{Good works:fruit of salvation ``f10!} and evidences of a true and lively\index{Faith:lively ``f10!} faith:\footnote{\ James 2:18, 22\index{59 James:ch.   2, v.  18, 22, pg. ``f30!}.} and by them believers manifest\index{Good works:believers manifest thankfulness ``f10!} their thankfulness\index{Prayer:thanking God ``f10!},\footnote{\ Psalm 116:12-13\index{19 Psalms:ch. 116, v.  12-13, pg. ``f30!}; 1 Peter 2:9\index{60 1 Peter:ch.   2, v.   9, pg. ``f30!}.} strengthen their assurance\index{Assurance:strengthened ``f10!},\footnote{\ 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}, 5\index{62 1 John:ch.   2, v.   3, 5, pg. ``f30!}; 2 Peter 1:5-10\index{61 2 Peter:ch.   1, v.   5-10, pg. ``f30!}.} edify\index{Good works:edify other Christians ``f10!} their brethren\index{Believers:to build up fellow believers ``f10!},\footnote{\ 2 Corinthians 9:2\index{47 2 Corinthians:ch.   9, v.   2, pg. ``f30!}; Matthew 5:16\index{40 Matthew:ch.   5, v.  16, pg. ``f30!}.} adorn the profession\index{Faith:profession of ``f10!} of the Gospel\index{Gospel:gospel ``f10!},\footnote{\ Titus 2:5, 9-12\index{56 Titus:ch.   2, v.   5, 9-12, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}.} stop the mouths\index{Good works:stop mouths of adversaries ``f10!} of the adversaries,\footnote{\ 1 Peter 2:15\index{60 1 Peter:ch.   2, v.  15, pg. ``f30!}.} and glorify\index{Worship:glorifying God ``f10!} God,\footnote{\ 1 Peter 2:12\index{60 1 Peter:ch.   2, v.  12, pg. ``f30!}; Philippians 1:1\index{50 Philippians:ch.   1, v.   1, pg. ``f30!}1\index{50 Philippians:ch.   1, v.  11, pg. ``f30!}; John 15:8\index{43 John:ch.  15, v.   8, pg. ``f30!}.} whose workmanship\index{Believers:God's workmanship ``f10!} they are, created in Christ\index{Conversion:created in Christ ``f10!} Jesus thereunto;\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  10, pg. ``f30!}.} that, having their fruit unto holiness\index{Believers:holy:made perfect at death ``f10!}, they may have the end, eternal life\index{eternal life ``f10!}.\footnote{\ Romans 6:22\index{45 Romans:ch.   6, v.  22, pg. ``f30!}.}   
\bigskip
16.3\ \ Their ability to do\index{Good works:only in Holy Spirit ``f10!} good works\index{Good works:adorn profession of faith ``f10!} is not at all of themselves, but wholly from the Spirit of Christ\index{Holy Spirit:Spirit of Christ ``f10!}.\footnote{\ John 15:4-5\index{43 John:ch.  15, v.   4-5, pg. ``f30!}; Ezekiel 36:26\index{26 Ezekiel:ch.  36, v.  26, pg. ``f30!}{}-27\index{26 Ezekiel:ch.  36, v.  26-27, pg. ``f30!}.} And that they may be enabled thereunto\index{Good works:enabled by Holy Spirit ``f10!}, besides the graces\index{Grace:in believers:by Holy Spirit ``f10!} they have already received, there is required an actual influence of the same Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!}, to work in them to will and to do of His good pleasure\index{Providence:at God's pleasure ``f10!}:\footnote{\ Philippians 2:13\index{50 Philippians:ch.   2, v.  13, pg. ``f30!}; Philippians 4:13\index{50 Philippians:ch.   4, v.  13, pg. ``f30!}; 2 Corinthians 3:5\index{47 2 Corinthians:ch.   3, v.   5, pg. ``f30!}.} yet are they not hereupon to grow negligent, as\index{Good works:not be negligent ``f10!} if they were not bound to perform any duty, unless\index{God's law:commandments:duties:required by God ``f10!} upon a special motion of the Spirit; but\index{Holy Spirit:helps believers do good works ``f10!} they ought to be diligent in\index{Grace:diligent use of means ``f10!} stirring up the grace of\index{Grace:in believers:to be stirred up ``f10!} God that is in them.\footnote{\ Philippians 2:12\index{50 Philippians:ch.   2, v.  12, pg. ``f30!}; Hebrews 6:11-12\index{58 Hebrews:ch.   6, v.  11-12, pg. ``f30!}; 2 Peter 1:3, 5, 10-11\index{61 2 Peter:ch.   1, v.   3, 5, 10-11, pg. ``f30!}; Isaiah 64:7\index{23 Isaiah:ch.  64, v.   7, pg. ``f30!}; 2 Timothy 1:6\index{55 2 Timothy:ch.   1, v.   6, pg. ``f30!}; Acts 26:6-7\index{44 Acts:ch.  26, v.   6-7, pg. ``f30!}; Jude 20-21\index{65 Jude:v. 20-21, pg. ``f30!}.}   
\bigskip
16.4\ \ They, who in their obedience\index{Salvation:not by:obedience ``f10!} attain to the greatest height which is possible in this life, are so far from being able to supererogate\index{Good works:believers cannot supererogate (do more than required) ``f10!}, and to do more than God requires, as\index{God's law:commandments:duties:required by God ``f10!} that they fall short\index{Sanctification:imperfection in believers ``f10!} of much which in duty they are bound to do.\footnote{\ Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}; Nehemiah 13:22\index{16 Nehemiah:ch.  13, v.  22, pg. ``f30!}; Job 9:2-3\index{18 Job:ch.   9, v.   2-3, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}.}   
\bigskip
16.5\ \ We cannot, by our best works, merit\index{Salvation:not by:works ``f10!} pardon of sin, or\index{Good works:not merit pardon of sin ``f10!} eternal life\index{eternal life ``f10!} at the hand of God, by reason of the great disproportion\index{Good works:great disproportion compared with glory ``f10!} that is between them and the glory to\index{Resurrection:believers:into glory:state of ``f10!} come; and the infinite\index{God's attributes:infinite:separation from creatures ``f10!} distance\index{Mankind:great distance between him and God ``f10!} that is between us and God, whom, by them, we can neither profit, nor satisfy for\index{Jesus:satisfied divine justice ``f10!} the debt\index{Forgiveness:of debtors ``f10!} of our former sins,\footnote{\ Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}; Romans 4:2, 4, 6\index{45 Romans:ch.   4, v.   2, 4, 6, pg. ``f30!}; Ephesians 2:8\index{49 Ephesians:ch.   2, v.   8, pg. ``f30!}{}-9\index{49 Ephesians:ch.   2, v.   8-9, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}{}-7\index{56 Titus:ch.   3, v.   5-7, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}8\index{45 Romans:ch.   8, v.  18, pg. ``f30!}; Psalm 16:2\index{19 Psalms:ch.  16, v.   2, pg. ``f30!}; Job 22:2-3\index{18 Job:ch.  22, v.   2-3, pg. ``f30!}; Job 35:7-8\index{18 Job:ch.  35, v.   7-8, pg. ``f30!}.} but when we have done all we can, we have done but our duty, and are\index{God's law:commandments:duties:required by God ``f10!} unprofitable servants\index{Good works:duty of unprofitable servants ``f10!};\footnote{\ Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}.} and because, as they are good, they proceed from His\index{Good works:only in Holy Spirit ``f10!} Spirit;\footnote{\ Galatians 5:2\index{48 Galatians:ch.   5, v.   2, pg. ``f30!}2-23\index{48 Galatians:ch.   5, v.  22-23, pg. ``f30!}.} and as they are wrought by us, they are defiled, and\index{Good works:defiled ``f10!} mixed\index{Good works:mixed with weakness and imperfection ``f10!} with so much weakness and\index{Sanctification:weaknesses and imperfections remain ``f10!} imperfection, that they cannot endure the severity\index{Good works:cannot endure severity of God's judgement ``f10!} of God's judgment\index{Judgement:severe ``f10!}.\footnote{\ Isaiah 64:6\index{23 Isaiah:ch.  64, v.   6, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}7\index{48 Galatians:ch.   5, v.  17, pg. ``f30!}; Romans 7:15, 18\index{45 Romans:ch.   7, v.  15, 18, pg. ``f30!}; Psalm 143:2\index{19 Psalms:ch. 143, v.   2, pg. ``f30!}; Psalm 130:3\index{19 Psalms:ch. 130, v.   3, pg. ``f30!}.}    
\bigskip
16.6\ \ Yet notwithstanding, the persons of believers being accepted through\index{Believers:accepted in Christ ``f10!} Christ, their good works\index{Good works:adorn profession of faith ``f10!} also are accepted in Him\index{Good works:accepted in Christ ``f10!},\footnote{\ Ephesians 1:6\index{49 Ephesians:ch.   1, v.   6, pg. ``f30!}; 1 Peter 2:5\index{60 1 Peter:ch.   2, v.   5, pg. ``f30!}; Exodus 28:38\index{02 Exodus:ch.  28, v.  38, pg. ``f30!}; Genesis 4:4\index{01 Genesis:ch.   4, v.   4, pg. ``f30!} with Hebrews 11:4\index{58 Hebrews:ch.  11, v.   4, pg. ``f30!}.} not as though they were in this life wholly unblameable and unreproveable\index{Good works:not unreproveable ``f10!} in God's sight;\footnote{\ Job 9:20\index{18 Job:ch.   9, v.  20, pg. ``f30!}; Psalm 143:2\index{19 Psalms:ch. 143, v.   2, pg. ``f30!}.} but that He, looking upon them in His Son\index{Jesus:Son of God ``f10!}, is pleased\index{Providence:at God's pleasure ``f10!} to accept and reward\index{Believers:rewarded by God ``f10!} that which is sincere\index{Good works:sincere accepted by God ``f10!}, although accompanied with many weaknesses\index{Sanctification:weaknesses and imperfections remain ``f10!} and imperfections\index{Sanctification:imperfection in believers ``f10!}.\footnote{\ Hebrews 13:20-21\index{58 Hebrews:ch.  13, v.  20-21, pg. ``f30!}; 2 Corinthians 8:1\index{47 2 Corinthians:ch.   8, v.   1, pg. ``f30!}2\index{47 2 Corinthians:ch.   8, v.  12, pg. ``f30!}; Hebrews 6:10\index{58 Hebrews:ch.   6, v.  10, pg. ``f30!}; Matthew 25:21\index{40 Matthew:ch.  25, v.  21, pg. ``f30!}, 23\index{40 Matthew:ch.  25, v.  21, 23, pg. ``f30!}.}   
\bigskip
16.7\ \ Works done\index{Good works:not merit pardon of sin ``f10!} by unregenerate\index{Unbelievers:unregenerate ``f10!} men, although, for the matter of them, they may be things which God commands, and of\index{God's law:commandments:obedience:not salvific ``f10!} good use both to themselves and others:\footnote{\ 2 Kings 10:30-31\index{12 2 Kings:ch.  10, v.  30-31, pg. ``f30!}; 1 Kings 21:27, 29\index{11 1 Kings:ch.  21, v.  27, 29, pg. ``f30!}; Philippians 1:1\index{50 Philippians:ch.   1, v.   1, pg. ``f30!}5-16, 18\index{50 Philippians:ch.   1, v.  15-16, 18, pg. ``f30!}.} yet, because they proceed not\index{Good works:not from unregenerate ``f10!} from a heart\index{Heart:new:purified ``f10!} purified by faith\index{Faith:false ``f10!};\footnote{\ Genesis 4:5\index{01 Genesis:ch.   4, v.   5, pg. ``f30!} with Hebrews 11:4\index{58 Hebrews:ch.  11, v.   4, pg. ``f30!}, 6\index{58 Hebrews:ch.  11, v.   4, 6, pg. ``f30!}.} nor are done in a right manner according to the Word;\index{Scripture:word of God:standard for:life ``f10!}\footnote{\ 1 Corinthians 13:3\index{46 1 Corinthians:ch.  13, v.   3, pg. ``f30!}; Isaiah 1:12\index{23 Isaiah:ch.   1, v.  12, pg. ``f30!}.} nor to a right end, the glory of God;\index{Good works:to be done for the glory of God ``f10!}\footnote{\ Matthew 6:2, 5, 16\index{40 Matthew:ch.   6, v.   2, 5, 16, pg. ``f30!}.} they are therefore sinful, and cannot please God\index{Good works:cannot please God ``f10!}, or make a man meet to receive grace\index{Grace:received from God ``f10!} from God.\footnote{\ Haggai 2:14\index{37 Haggai:ch.   2, v.  14, pg. ``f30!}; Titus 1:15\index{56 Titus:ch.   1, v.  15, pg. ``f30!}; Amos 5:22-23\index{30 Amos:ch.   5, v.  22-23, pg. ``f30!}; Hosea 1:4\index{28 Hosea:ch.   1, v.   4, pg. ``f30!}; Romans 9:16\index{45 Romans:ch.   9, v.  16, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.} And yet, their neglect of them\index{Good works:not be negligent ``f10!} is more sinful, and displeasing unto God.\footnote{\ Psalm 14:4\index{19 Psalms:ch.  14, v.   4, pg. ``f30!}; Psalm 36:3\index{19 Psalms:ch.  36, v.   3, pg. ``f30!}; Job 21:14\index{18 Job:ch.  21, v.  14, pg. ``f30!}{}-15\index{18 Job:ch.  21, v.  14-15, pg. ``f30!}; Matthew 25:41\index{40 Matthew:ch.  25, v.  41, pg. ``f30!}{}-43\index{40 Matthew:ch.  25, v.  41-43, pg. ``f30!}, 45\index{40 Matthew:ch.  25, v.  41-43, 45, pg. ``f30!}; Matthew 23:23\index{40 Matthew:ch.  23, v.  23, pg. ``f30!}.}  

\section{Chapter 17 -- Of the Perseverance of the Saints} \index{Perseverance:of the saints ``f10!}\index{Believers:saints ``f10!}17.1\ \ They, whom God has accepted in His\index{Believers:accepted in Christ ``f10!} Beloved\index{Jesus:beloved of the Father ``f10!}, effectually called\index{Effectual call:calling ``f10!}, and sanctified\index{Sanctification:sanctification ``f10!} by His Spirit, can\index{Holy Spirit:applies:sanctification ``f10!} neither totally\index{Assurance:never fall from being saved ``f10!}, nor finally, fall away from the state of grace\index{Grace:state:fall from ``f10!}: but shall certainly persevere\index{Perseverance:of the saints ``f10!} therein to the end, and be eternally\index{Salvation:eternal for those effectually called ``f10!} saved.\footnote{\ Philippians 1:6\index{50 Philippians:ch.   1, v.   6, pg. ``f30!}; 2 Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}; John 10:28\index{43 John:ch.  10, v.  28, pg. ``f30!}{}-29\index{43 John:ch.  10, v.  28-29, pg. ``f30!}; 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}; 1 Peter 1:5\index{60 1 Peter:ch.   1, v.   5, pg. ``f30!}, 9\index{60 1 Peter:ch.   1, v.   5, 9, pg. ``f30!}.}   
\bigskip
17.2\ \ This perseverance\index{Perseverance:of the saints ``f10!} of the saints\index{Believers:saints ``f10!} depends not upon their\index{Perseverance:depends not on person's will ``f10!} own free will, but upon the immutability\index{God's attributes:immutable ``f10!} of the decree\index{Predestination:eternal decrees ``f10!} of election\index{Election:decree of ``f10!}, flowing from the free and unchangeable\index{God's attributes:unchangeable ``f10!} love of God\index{Love toward God:for God ``f10!} the Father\index{Father:God the ``f10!};\footnote{\ 2 Timothy 2:18-19\index{55 2 Timothy:ch.   2, v.  18-19, pg. ``f30!}; Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}.} upon the efficacy\index{Perseverance:enabled by Holy Spirit ``f10!} of the merit and\index{Jesus:obedience:merit of ``f10!} intercession\index{Jesus:in heaven:intercession made for all believers ``f10!} of Jesus Christ;\footnote{\ Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}0\index{58 Hebrews:ch.  10, v.  10, pg. ``f30!}, 14\index{58 Hebrews:ch.  10, v.  10, 14, pg. ``f30!}; Hebrews 13:20-21\index{58 Hebrews:ch.  13, v.  20-21, pg. ``f30!}; Hebrews 9:12\index{58 Hebrews:ch.   9, v.  12, pg. ``f30!}{}-15\index{58 Hebrews:ch.   9, v.  12-15, pg. ``f30!}; Romans 8:3\index{45 Romans:ch.   8, v.   3, pg. ``f30!}3\index{45 Romans:ch.   8, v.  33, pg. ``f30!} to the end, John 17:11, 24\index{43 John:ch.  17, v.  11, 24, pg. ``f30!}; Luke 22:32\index{42 Luke:ch.  22, v.  32, pg. ``f30!}; Hebrews 7:25\index{58 Hebrews:ch.   7, v.  25, pg. ``f30!}.} the abiding of\index{Holy Spirit:abiding in believers ``f10!} the Spirit, and of the seed of God\index{Believers:seed of God planted within ``f10!} within them;\footnote{\ John 14:16-17\index{43 John:ch.  14, v.  16-17, pg. ``f30!}; 1 John 2:2\index{62 1 John:ch.   2, v.   2, pg. ``f30!}7\index{62 1 John:ch.   2, v.  27, pg. ``f30!}; 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}.} and the nature of the covenant of grace\index{Covenant:grace:grace ``f10!}:\footnote{\ Jeremiah 32:40\index{24 Jeremiah:ch.  32, v.  40, pg. ``f30!}.} from all which arises also the certainty and infallibility\index{Covenant:grace:infallible ``f10!} thereof.\footnote{\ John 10:28\index{43 John:ch.  10, v.  28, pg. ``f30!}; 2 Thessalonians 3:3\index{53 2 Thessalonians:ch.   3, v.   3, pg. ``f30!}; 1 John 2:19\index{62 1 John:ch.   2, v.  19, pg. ``f30!}.}   
\bigskip
17.3\ \ Nevertheless, they may, through the temptations\index{Temptation:temptation ``f10!} of Satan and of\index{Satan:tempter ``f10!} the world, the prevalence of corruption remaining\index{Sanctification:remnants of corruption remain ``f10!} in them, and the neglect of the means\index{Love for others:neighbour as self ``f10!} of their preservation\index{Perseverance:means neglected ``f10!}, fall into grievous sins\index{Temptation:fall into grievous sins ``f10!};\footnote{\ Matthew 26:70, 72, 74\index{40 Matthew:ch.  26, v.  70, 72, 74, pg. ``f30!}.} and, for a time, continue therein:\footnote{\ Psalm 51 title and 14.} whereby they incur\index{Believers:may experience God's displeasure for sin ``f10!} God's displeasure,\footnote{\ Isaiah 64:5, 7, 9\index{23 Isaiah:ch.  64, v.   5, 7, 9, pg. ``f30!}; 2 Samuel 11:27\index{10 2 Samuel:ch.  11, v.  27, pg. ``f30!}.} and grieve\index{Sin:grieves the Holy Spirit ``f10!} His Holy Spirit\index{Holy Spirit:Holy Spirit ``f10!},\footnote{\ Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}.} come to be deprived of some measure of their graces and\index{Grace:withheld ``f10!} comforts\index{Temptation:comforts removed ``f10!},\footnote{\ Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}, 10, 12\index{19 Psalms:ch.  51, v.   8, 10, 12, pg. ``f30!}; Revelation 2:4\index{66 Revelation:ch.   2, v.   4, pg. ``f30!}; Song of Solomon 5:2-4, 6\index{22 Song of Solomon:ch.   5, v.   2-4, 6, pg. ``f30!}.} have their hearts hardened\index{Heart:hardened ``f10!},\footnote{\ Isaiah 63:17\index{23 Isaiah:ch.  63, v.  17, pg. ``f30!}; Mark 6:52\index{41 Mark:ch.   6, v.  52, pg. ``f30!}; Mark 16:14\index{41 Mark:ch.  16, v.  14, pg. ``f30!}.} and their consciences\index{Conscience:wounded ``f10!} wounded,\footnote{\ Psalm 32:3-4\index{19 Psalms:ch.  32, v.   3-4, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}.} hurt and scandalize\index{Scandalous behaviour:scandalize others ``f10!} others,\footnote{\ 2 Samuel 12:14\index{10 2 Samuel:ch.  12, v.  14, pg. ``f30!}.} and bring temporal judgments upon\index{Judgement:temporal ``f10!} themselves.\footnote{\ Psalm 89:31-32\index{19 Psalms:ch.  89, v.  31-32, pg. ``f30!}; 1 Corinthians 11:32\index{46 1 Corinthians:ch.  11, v.  32, pg. ``f30!}.}  

\section{Chapter 18 -- Of the Assurance of Grace and Salvation} \index{Assurance:of state of grace ``f10!}\index{Salvation:assurance of ``f10!}18.1\ \ Although hypocrites\index{hypocrisy ``f10!} and other unregenerate\index{Unbelievers:unregenerate ``f10!} men may vainly deceive\index{Deceitfulness:of self ``f10!} themselves with false hopes\index{Hope:false ``f10!}, and carnal presumptions\index{presumption ``f10!} of being in the favour\index{Unbelievers:presumptuous about favour with God ``f10!} of God, and estate of salvation\index{Salvation:belief in Jesus by faith ``f10!};\footnote{\ Job 8:13-14\index{18 Job:ch.   8, v.  13-14, pg. ``f30!}; Micah 3:11\index{33 Micah:ch.   3, v.  11, pg. ``f30!}; Deuteronomy 29:19\index{05 Deuteronomy:ch.  29, v.  19, pg. ``f30!}; John 8:41\index{43 John:ch.   8, v.  41, pg. ``f30!}.} which hope of theirs shall perish:\footnote{\ Matthew 7:22\index{40 Matthew:ch.   7, v.  22, pg. ``f30!}{}-23\index{40 Matthew:ch.   7, v.  22-23, pg. ``f30!}.} yet such as truly believe in the Lord Jesus, and love Him\index{Love toward God:for God ``f10!} in sincerity\index{Love:of Christ ``f10!}, endeavouring to walk in all\index{Believers:strive to walk in good conscience ``f10!} good conscience\index{Conscience:good, endeavour for ``f10!} before Him, may, in this life, be certainly assured that\index{Assurance:of state of grace ``f10!} they are in the state of grace\index{Grace:state:believers in ``f10!},\footnote{\ 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}4\index{62 1 John:ch.   3, v.  14, pg. ``f30!}, 18-19, 21, 24\index{62 1 John:ch.   3, v.  14, 18-19, 21, 24, pg. ``f30!}; John 5:13\index{43 John:ch.   5, v.  13, pg. ``f30!}.} and may rejoice in the hope of\index{Hope:of glory of God, rejoice in ``f10!} the glory of God, which\index{Resurrection:believers:into glory:everlasting ``f10!} hope shall\index{Hope:certain ``f10!} never make them ashamed\index{Believers:never made ashamed by God ``f10!}.\footnote{\ Romans 5:2\index{45 Romans:ch.   5, v.   2, pg. ``f30!}, 5\index{45 Romans:ch.   5, v.   2, 5, pg. ``f30!}.}   
\bigskip
18.2\ \ This certainty is not a bare conjectural\index{Assurance:not bare conjecture ``f10!} and probable persuasion\index{Assurance:not probable persuasion ``f10!}, grounded upon a fallible hope;\index{Hope:fallible ``f10!}\footnote{\ Hebrews 6:11, 19\index{58 Hebrews:ch.   6, v.  11, 19, pg. ``f30!}.} but an infallible\index{Assurance:infallible ``f10!} assurance of faith\index{Assurance:of faith ``f10!}, founded\index{Assurance:founded on God's promises ``f10!} upon the divine\index{Salvation:founded on divine truth ``f10!} truth of the promises\index{Salvation:promises made ``f10!} of salvation\index{Salvation:belief in Jesus by faith ``f10!},\footnote{\ Hebrews 6:17\index{58 Hebrews:ch.   6, v.  17, pg. ``f30!}{}-18\index{58 Hebrews:ch.   6, v.  17-18, pg. ``f30!}.} the inward evidence\index{Assurance:inward evidence ``f10!} of those graces\index{Grace:in believers:by Holy Spirit ``f10!} unto which these promises\index{Assurance:promises of God ``f10!} are made,\footnote{\ 2 Peter 1:4-5, 10-11\index{61 2 Peter:ch.   1, v.   4-5, 10-11, pg. ``f30!}; 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}4\index{62 1 John:ch.   3, v.  14, pg. ``f30!}; 2 Corinthians 1:1\index{47 2 Corinthians:ch.   1, v.   1, pg. ``f30!}2\index{47 2 Corinthians:ch.   1, v.  12, pg. ``f30!}.} the testimony of the Spirit\index{Assurance:Holy Spirit witnesses with our spirits ``f10!} of adoption\index{Adoption:in salvation ``f10!} witnessing with our spirits\index{Conversion:Holy Spirit:bears witness to our spirits ``f10!} that we are the children of God:\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}5\index{45 Romans:ch.   8, v.  15, pg. ``f30!}{}-16\index{45 Romans:ch.   8, v.  15-16, pg. ``f30!}.} which Spirit\index{Holy Spirit:earnest of believers' inheritance ``f10!} is the earnest of\index{Believers:inheritance:earnest of ``f10!} our inheritance\index{Believers:inheritance:everlasting ``f10!}, whereby we are sealed to the\index{Salvation:sealed to the day of redemption ``f10!} day of redemption\index{Redemption:day of ``f10!}.\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}3\index{49 Ephesians:ch.   1, v.  13, pg. ``f30!}{}-14\index{49 Ephesians:ch.   1, v.  13-14, pg. ``f30!}; Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}; 2 Corinthians 1:21-22\index{47 2 Corinthians:ch.   1, v.  21-22, pg. ``f30!}.} 
\bigskip
18.3\ \ This infallible\index{Assurance:infallible ``f10!} assurance does\index{Assurance:believer may await ``f10!} not so belong to the essence of faith\index{Faith:essence of ``f10!}, but that a true believer may wait long, and conflict with many difficulties\index{Assurance:believer may experience difficulties ``f10!} before he be partaker of it:\footnote{\ 1 John 5:13\index{62 1 John:ch.   5, v.  13, pg. ``f30!}; Isaiah 50:10\index{23 Isaiah:ch.  50, v.  10, pg. ``f30!}; Mark 9:24\index{41 Mark:ch.   9, v.  24, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}; Psalm 77:1-12\index{19 Psalms:ch.  77, v.   1-12, pg. ``f30!}.} yet, being enabled by\index{Assurance:enabled by Holy Spirit ``f10!} the Spirit to\index{Revelation:Word and Spirit ``f10!} know the things which are freely\index{Salvation:freely offered ``f10!} given him of God, he may without extraordinary revelation\index{Revelation:extraordinary ``f10!}, in the right use of ordinary means, attain thereunto.\footnote{\ 1 Corinthians 2:12\index{46 1 Corinthians:ch.   2, v.  12, pg. ``f30!}; 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg. ``f30!}3\index{62 1 John:ch.   4, v.  13, pg. ``f30!}; Hebrews 6:11-12\index{58 Hebrews:ch.   6, v.  11-12, pg. ``f30!}; Ephesians 3:17\index{49 Ephesians:ch.   3, v.  17, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  17-19, pg. ``f30!}.} And therefore it is the duty of everyone\index{Election:duty to make calling sure ``f10!} to give all diligence to\index{Election:diligence to make calling sure ``f10!} make his calling and\index{Effectual call:elect ``f10!} election sure;\footnote{\ 2 Peter 1:10\index{61 2 Peter:ch.   1, v.  10, pg. ``f30!}.} that thereby his heart may be enlarged\index{Heart:enlarged in peace and joy ``f10!} in peace\index{Assurance:peace and joy ``f10!} and joy in\index{Joy:in Holy Spirit ``f10!} the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, in love and\index{Love toward God:for God ``f10!} thankfulness\index{Prayer:thanking God ``f10!} to God, and in strength\index{Sanctification:supply of strength from Holy Spirit ``f10!} and cheerfulness\index{cheerfulness ``f10!} in the duties of obedience, the\index{God's law:commandments:obedience:required ``f10!} proper fruits\index{Good works:fruit of salvation ``f10!} of this assurance:\index{Assurance:not to be used to justify sin ``f10!}\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}{}-2\index{45 Romans:ch.   5, v.   1-2, pg. ``f30!}, 5\index{45 Romans:ch.   5, v.   1-2, 5, pg. ``f30!}; Romans 14:17\index{45 Romans:ch.  14, v.  17, pg. ``f30!}; Romans 15:13\index{45 Romans:ch.  15, v.  13, pg. ``f30!}; Ephesians 1:3-4\index{49 Ephesians:ch.   1, v.   3-4, pg. ``f30!}; Psalm 4:6-7\index{19 Psalms:ch.   4, v.   6-7, pg. ``f30!}; Psalm 119:32\index{19 Psalms:ch. 119, v.  32, pg. ``f30!}.} so far is it from inclining men to looseness.\footnote{\ 1 John 2:1-2\index{62 1 John:ch.   2, v.   1-2, pg. ``f30!}; Romans 6:1-2\index{45 Romans:ch.   6, v.   1-2, pg. ``f30!}; Titus 2:11-12, 14\index{56 Titus:ch.   2, v.  11-12, 14, pg. ``f30!}; 2 Corinthians 7:1\index{47 2 Corinthians:ch.   7, v.   1, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}, 12\index{45 Romans:ch.   8, v.   1, 12, pg. ``f30!}; 1 John 3:2\index{62 1 John:ch.   3, v.   2, pg. ``f30!}{}-3\index{62 1 John:ch.   3, v.   2-3, pg. ``f30!}; Psalm 130:4\index{19 Psalms:ch. 130, v.   4, pg. ``f30!}; 1 John 1:6-7\index{62 1 John:ch.   1, v.   6-7, pg. ``f30!}.}   
\bigskip
18.4\ \ True believers may\index{Assurance:believer may not have at all times ``f10!} have the assurance of their salvation\index{Salvation:belief in Jesus by faith ``f10!} divers ways shaken\index{Assurance:shaken at times ``f10!}, diminished, and intermitted; as, by negligence\index{Assurance:negligence in preserving ``f10!} in preserving\index{Perseverance:means neglected ``f10!} of it, by falling\index{God's law:broken daily by believers ``f10!} into some special sin, which wounds the conscience\index{Conscience:wounded ``f10!} and grieves the\index{Sin:grieves the Holy Spirit ``f10!} Spirit; by some sudden\index{Temptation:sudden ``f10!} or vehement\index{Temptation:vehement ``f10!} temptation\index{Temptation:temptation ``f10!}, by God's withdrawing the light of His countenance\index{Believers:chastened by God ``f10!}, and suffering even such as fear Him\index{Fear:of God:reverence ``f10!} to walk in darkness\index{darkness, spiritual ``f10!} and to have no light:\footnote{\ Song of Solomon 5:2-3, 6\index{22 Song of Solomon:ch.   5, v.   2-3, 6, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}, 12\index{19 Psalms:ch.  51, v.   8, 12, pg. ``f30!}, 14\index{19 Psalms:ch.  51, v.   8, 12, 14, pg. ``f30!}; Ephesians 4:3\index{49 Ephesians:ch.   4, v.   3, pg. ``f30!}0\index{49 Ephesians:ch.   4, v.  30, pg. ``f30!}{}-31\index{49 Ephesians:ch.   4, v.  30-31, pg. ``f30!}; Psalm 77:1-10\index{19 Psalms:ch.  77, v.   1-10, pg. ``f30!}; Matthew 26:69-72\index{40 Matthew:ch.  26, v.  69-72, pg. ``f30!}; Psalm 31:22\index{19 Psalms:ch.  31, v.  22, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}; Isaiah 50:10\index{23 Isaiah:ch.  50, v.  10, pg. ``f30!}.} yet are they never so utterly destitute\index{Assurance:never destitute ``f10!} of that seed of God\index{Believers:seed of God planted within ``f10!}, and life of faith\index{Faith:living by ``f10!}, that love of Christ\index{Love:of Christ ``f10!} and the brethren\index{Believers:to build up fellow believers ``f10!}, that sincerity of\index{Heart:sincerity of ``f10!} heart, and conscience of\index{Believers:desire to obey God ``f10!} duty, out\index{God's law:commandments:duties:required by God ``f10!} of which, by the operation\index{Holy Spirit:operations:saving ``f10!} of the Spirit, this assurance may, in due time, be revived\index{Assurance:revived ``f10!};\footnote{\ 1 John 3:9\index{62 1 John:ch.   3, v.   9, pg. ``f30!}; Luke 22:32\index{42 Luke:ch.  22, v.  32, pg. ``f30!}; Job 13:15\index{18 Job:ch.  13, v.  15, pg. ``f30!}; Psalm 73:15\index{19 Psalms:ch.  73, v.  15, pg. ``f30!}; Psalm 51:8\index{19 Psalms:ch.  51, v.   8, pg. ``f30!}, 12\index{19 Psalms:ch.  51, v.   8, 12, pg. ``f30!}; Isaiah 50:10\index{23 Isaiah:ch.  50, v.  10, pg. ``f30!}.} and by the which, in the mean time, they are supported from\index{Assurance:Holy Spirit keeps believer from utter despair ``f10!} utter despair\index{despair ``f10!}.\footnote{\ Micah 7:7\index{33 Micah:ch.   7, v.   7, pg. ``f30!}{}-9\index{33 Micah:ch.   7, v.   7-9, pg. ``f30!}; Jeremiah 32:40\index{24 Jeremiah:ch.  32, v.  40, pg. ``f30!}; Isaiah 54:7-10\index{23 Isaiah:ch.  54, v.   7-10, pg. ``f30!}; Psalm 22:1\index{19 Psalms:ch.  22, v.   1, pg. ``f30!}; Psalm 88:1-18\index{19 Psalms:ch.  88, v.   1-18, pg. ``f30!}.}  

\section{Chapter 19 -- Of the Law of God} 19.1\ \ God gave to Adam a law\index{Adam:given a law ``f10!}, as a covenant of works\index{Covenant:works:works ``f10!}, by which He bound him and all his posterity\index{Sin:original:all mankind descending from Adam ``f10!} to personal, entire, exact, and perpetual obedience\index{Covenant:works:perpetual obedience required ``f10!}; promised life\index{Covenant:first:promised life ``f10!} upon the fulfilling, and\index{Covenant:works:conditional on perfect obedience ``f10!} threatened death\index{Death:punishment for sin ``f10!} upon the breach of it\index{God's law:breach:death threatened ``f10!}: and endued him\index{Mankind:created:attributes ``f10!} with power and ability to keep\index{Covenant:works:Adam:original ability and power to keep ``f10!} it.\footnote{\ Genesis 1:2\index{01 Genesis:ch.   1, v.   2, pg. ``f30!}6\index{01 Genesis:ch.   1, v.  26, pg. ``f30!}{}-27\index{01 Genesis:ch.   1, v.  26-27, pg. ``f30!} with Genesis 2:17\index{01 Genesis:ch.   2, v.  17, pg. ``f30!}; Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   2, v.  14-15, pg. ``f30!}; Romans 10:5\index{45 Romans:ch.  10, v.   5, pg. ``f30!}; Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}2\index{45 Romans:ch.   5, v.  12, pg. ``f30!}, 19\index{45 Romans:ch.   5, v.  12, 19, pg. ``f30!}; Galatians 3:10\index{48 Galatians:ch.   3, v.  10, pg. ``f30!}, 12\index{48 Galatians:ch.   3, v.  10, 12, pg. ``f30!}; Ecclesiastes 7:2\index{21 Ecclesiastes:ch.   7, v.   2, pg. ``f30!}9\index{21 Ecclesiastes:ch.   7, v.  29, pg. ``f30!}; Job 28:28\index{18 Job:ch.  28, v.  28, pg. ``f30!}.}   
\bigskip
19.2\ \ This law, after his fall\index{Sin:original:first transgression (of Adam) ``f10!}, continued to be a perfect rule of righteousness, and, as\index{God's law:righteous requirements ``f10!} such, was delivered by\index{God's law:commandments:delivered on Sinai ``f10!} God upon Mount Sinai\index{Sinai, mount ``f10!}, in ten\index{God's law:commandments:ten ``f10!} commandments, and written in two tables\index{God's law:commandments:two tables ``f10!}:\footnote{\index{59 James:ch.   1, v.  25, pg. ``f30!}{, 10-12}\index{59 James:ch.   2, v.   8, 10-12, pg. ``f30!}\index{45 Romans:ch.  13, v.   8, pg. ``f30!}\index{45 Romans:ch.  13, v.   8-9, pg. ``f30!}\index{05 Deuteronomy:ch.   5, v.  32, pg. ``f30!}\index{05 Deuteronomy:ch.  10, v.   4, pg. ``f30!} the four first commandment\index{God's law:commandments:first ``f10!}s\index{God's law:commandments:first four, duty to God ``f10!} containing our duty towards\index{God's law:commandments:duties:required by God ``f10!} God; and the other six\index{Sabbath:six days:for own affairs ``f10!} our duty to man\index{God's law:commandments:duties:to others ``f10!}.\footnote{\ Matthew 22:37-40\index{40 Matthew:ch.  22, v.  37-40, pg. ``f30!}.}   
\bigskip
19.3\ \ Beside this law, commonly called moral\index{God's law:commandments:moral ``f10!}, God was pleased\index{Providence:at God's pleasure ``f10!} to give to the people of Israel\index{Israel ``f10!}, as a church\index{Church:OT ``f10!} under age, ceremonial\index{Ceremonial law:fore-signifying Christ ``f10!} laws, containing several typical\index{types, OT ``f10!} ordinances, partly\index{Covenant:OT ordinances:signified Christ ``f10!} of worship\index{Worship:of God ``f10!}, prefiguring\index{Jesus:prefigured in ceremonial ordinances ``f10!} Christ, His graces\index{Jesus:full of grace ``f10!}, actions, sufferings, and\index{Jesus:death:sufferings and death ``f10!} benefits;\footnote{\ Hebrews chap. 9; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}; Galatians 4:1-3\index{48 Galatians:ch.   4, v.   1-3, pg. ``f30!}; Colossians 2:17\index{51 Colossians:ch.   2, v.  17, pg. ``f30!}.} and partly holding\index{God's law:benefits ``f10!} forth divers instructions of moral duties.\index{God's law:commandments:duties:required by God ``f10!}\footnote{\ 1 Corinthians 5:7\index{46 1 Corinthians:ch.   5, v.   7, pg. ``f30!}; 2 Corinthians 6:17\index{47 2 Corinthians:ch.   6, v.  17, pg. ``f30!}; Jude 23\index{65 Jude:v.  23, pg. ``f30!}.} All which ceremonial laws are now abrogated\index{Ceremonial law:abrogated ``f10!}, under the new testament\index{Scripture:New Testament ``f10!}.\footnote{\ Colossians 2:14, 16-17\index{51 Colossians:ch.   2, v.  14, 16-17, pg. ``f30!}; Daniel 9:27\index{27 Daniel:ch.   9, v.  27, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}5-16\index{49 Ephesians:ch.   2, v.  15-16, pg. ``f30!}.}   
\bigskip
19.4\ \ To them also, as a body politic, He gave sundry judicial laws, which expired\index{God's law:judicial:expired ``f10!} together with the State of that people; not obliging any other now, further than the general equity thereof\index{God's law:judicial:apply equity today ``f10!} may require.\footnote{\ Exodus chap. 21; Exodus 22:1-29\index{02 Exodus:ch.  22, v.   1-29, pg. ``f30!}; Genesis 49:10\index{01 Genesis:ch.  49, v.  10, pg. ``f30!} with 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}; Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!} with Matthew 5:38-39\index{40 Matthew:ch.   5, v.  38-39, pg. ``f30!}; 1 Corinthians 9:8-10\index{46 1 Corinthians:ch.   9, v.   8-10, pg. ``f30!}.}   
\bigskip
19.5\ \ The moral\index{God's law:commandments:moral ``f10!} law does for ever bind all\index{God's law:binding on all men ``f10!}, as well justified\index{Justification:justified ``f10!} persons as others, to the obedience thereof\index{God's law:commandments:obedience:required ``f10!};\footnote{\ Romans 13:8\index{45 Romans:ch.  13, v.   8, pg. ``f30!}{}-10\index{45 Romans:ch.  13, v.   8-10, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2, pg. ``f30!}; 1 John 2:3\index{62 1 John:ch.   2, v.   3, pg. ``f30!}{}-4, 7-8\index{62 1 John:ch.   2, v.   3-4, 7-8, pg. ``f30!}.} and that, not only in regard of the matter contained in it, but also in respect of the authority of God the Creator\index{God's attributes:creator:of all things ``f10!}, who gave it:\footnote{\ James 2:10\index{59 James:ch.   2, v.  10, pg. ``f30!}{}-11\index{59 James:ch.   2, v.  10-11, pg. ``f30!}.} neither does Christ, in the Gospel\index{Gospel:gospel ``f10!}, any way dissolve, but much strengthen this obligation\index{God's law:obligation to obey:strengthened by Gospel ``f10!}.\footnote{\ Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}{}-19\index{40 Matthew:ch.   5, v.  17-19, pg. ``f30!}; James 2:8\index{59 James:ch.   2, v.   8, pg. ``f30!}; Romans 3:31\index{45 Romans:ch.   3, v.  31, pg. ``f30!}.}   
\bigskip
19.6\ \ Although true believers\index{Covenant:works:believers not under ``f10!} be not under the law, as a covenant of works\index{Covenant:works:works ``f10!}, to be thereby justified\index{Justification:justified ``f10!}, or condemned\index{Covenant:works:believers delivered from ``f10!};\footnote{\ Romans 6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}; Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Galatians 3:13\index{48 Galatians:ch.   3, v.  13, pg. ``f30!}; Galatians 4:4\index{48 Galatians:ch.   4, v.   4, pg. ``f30!}{}-5\index{48 Galatians:ch.   4, v.   4-5, pg. ``f30!}; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}9\index{44 Acts:ch.  13, v.  39, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}.} yet is it of great use to them, as well as to others; in that, as a rule of life\index{God's law:moral:rule for life ``f10!} informing them of the will of God, and their duty, it\index{God's law:commandments:duties:required by God ``f10!} directs, and binds them\index{God's law:binding on all men ``f10!} to walk accordingly;\footnote{\ Romans 7:12\index{45 Romans:ch.   7, v.  12, pg. ``f30!}, 22, 25\index{45 Romans:ch.   7, v.  12, 22, 25, pg. ``f30!}; Psalm 119:4-6\index{19 Psalms:ch. 119, v.   4-6, pg. ``f30!}; 1 Corinthians 7:19\index{46 1 Corinthians:ch.   7, v.  19, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}4, 16, 18-23\index{48 Galatians:ch.   5, v.  14, 16, 18-23, pg. ``f30!}.} discovering also the sinful pollutions\index{Sin:original:corrupt nature:polluted ``f10!} of their nature, hearts, and lives\index{Heart:polluted ``f10!};\footnote{\ Romans 7:7\index{45 Romans:ch.   7, v.   7, pg. ``f30!}; Romans 3:2\index{45 Romans:ch.   3, v.   2, pg. ``f30!}0\index{45 Romans:ch.   3, v.  20, pg. ``f30!}.} so as, examining\index{God's law:examine self against ``f10!} themselves thereby, they may come to further conviction\index{God's law:conviction of sin furthered ``f10!} of, humiliation for\index{Humility:before God ``f10!}, and hatred against\index{Sin:to be hated ``f10!} sin;\footnote{\ James 1:23-25\index{59 James:ch.   1, v.  23-25, pg. ``f30!}; Romans 7:9, 14, 24\index{45 Romans:ch.   7, v.   9, 14, 24, pg. ``f30!}.} together with a clearer sight of the need they have of Christ, and the perfection of His obedience\index{Jesus:obedience:to God's law ``f10!}.\footnote{\index{48 Galatians:ch.   3, v.  24, pg. ``f30!}\index{45 Romans:ch.   7, v.  24-25, pg. ``f30!}\index{45 Romans:ch.   8, v.   3, pg. ``f30!} It is likewise of use to the regenerate\index{Regeneration:regeneration ``f10!}, to restrain their corruptions, in\index{God's law:restrains corruptions ``f10!} that it forbids\index{God's law:forbids sin ``f10!} sin:\footnote{\index{59 James:ch.   2, v.  11, pg. ``f30!}\index{19 Psalms:ch. 119, v. 101, 104, 128, pg. ``f30!} of it serve to show what even their sins deserve\index{Sin:deserves damnation ``f10!}; and what afflictions, in\index{God's law:breach:define afflictions imposed on sin ``f10!} this life, they may expect for them, although freed\index{Believers:freed from:curse of the law ``f10!} from the curse thereof\index{Curse:of the law ``f10!} threatened in\index{God's law:commandments:threatenings for disobedience ``f10!} the law.\footnote{\ Ezra 9:13-14\index{15 Ezra:ch.   9, v.  13-14, pg. ``f30!}; Psalm 89:30-34\index{19 Psalms:ch.  89, v.  30-34, pg. ``f30!}.} The promises\index{God's law:promises ``f10!} of it, in like manner, show them God's approbation\index{God's law:approbation for obedience ``f10!} of obedience, and\index{Believers:obedience:rewarded ``f10!} what blessings they\index{Believers:blessed in/by God ``f10!} may expect upon the performance thereof;\footnote{\ Leviticus 26:1-14\index{03 Leviticus:ch.  26, v.   1-14, pg. ``f30!} with 2 Corinthians 6:16\index{47 2 Corinthians:ch.   6, v.  16, pg. ``f30!}; Ephesians 6:2\index{49 Ephesians:ch.   6, v.   2, pg. ``f30!}{}-3\index{49 Ephesians:ch.   6, v.   2-3, pg. ``f30!}; Psalm 37:11\index{19 Psalms:ch.  37, v.  11, pg. ``f30!} with Matthew 5:5\index{40 Matthew:ch.   5, v.   5, pg. ``f30!}; Psalm 19:11\index{19 Psalms:ch.  19, v.  11, pg. ``f30!}.} although not as due to them by the law, as a covenant of works.\footnote{\ Galatians 2:16\index{48 Galatians:ch.   2, v.  16, pg. ``f30!}; Luke 17:10\index{42 Luke:ch.  17, v.  10, pg. ``f30!}.} So as, a man's doing good, and refraining from evil\index{God's law:commandments:obedience:refraining from evil ``f10!}, because the law encourages to the one and deterrers\index{God's law:deterrence from sin ``f10!} from the other, is no evidence of his being under the law; and not under grace\index{Grace:in believers:by Holy Spirit ``f10!}.\footnote{\ Romans 6:12, 14\index{45 Romans:ch.   6, v.  12, 14, pg. ``f30!}; 1 Peter 3:8-12\index{60 1 Peter:ch.   3, v.   8-12, pg. ``f30!} with Psalm 34:1\index{19 Psalms:ch.  34, v.   1, pg. ``f30!}2-16\index{19 Psalms:ch.  34, v.  12-16, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}{}-29\index{58 Hebrews:ch.  12, v.  28-29, pg. ``f30!}.}   
\bigskip
19.7\ \ Neither are the forementioned uses of the law contrary to the grace of\index{Grace:of Gospel ``f10!} the Gospel\index{Gospel:gospel ``f10!}, but do sweetly comply\index{God's law:compliance with help of Holy Spirit ``f10!} with it;\footnote{\ Galatians 3:21\index{48 Galatians:ch.   3, v.  21, pg. ``f30!}.} the Spirit of Christ\index{Holy Spirit:Spirit of Christ ``f10!} subduing and\index{Holy Spirit:subdues will of man ``f10!} enabling the\index{God's law:commandments:obedience:enabled by Holy Spirit ``f10!} will of man to do that, freely\index{Causation:will of creatures:free ``f10!} and cheerfully\index{cheerfulness ``f10!}, which the will of God, revealed in the law\index{Revelation:God's will ``f10!}, requires to be\index{God's law:commandments:duties:required by God ``f10!} done.\footnote{\ Ezekiel 36:27\index{26 Ezekiel:ch.  36, v.  27, pg. ``f30!}; Hebrews 8:10\index{58 Hebrews:ch.   8, v.  10, pg. ``f30!} with Jeremiah 31:3\index{24 Jeremiah:ch.  31, v.   3, pg. ``f30!}3\index{24 Jeremiah:ch.  31, v.  33, pg. ``f30!}.}  

\section{Chapter 20 -- Of Christian Liberty, and Liberty of Conscience} \index{Christian liberty:defined ``f10!}\index{Conscience:liberty of:defined ``f10!}20.1\ \ The liberty which\index{Christian liberty:purchased by Christ ``f10!} Christ has purchased\index{Redemption:purchased by Christ ``f10!} for believers under\index{Christian liberty:believers under the Gospel ``f10!} the Gospel\index{Gospel:gospel ``f10!} consists in their freedom\index{Believers:freed from:guilt of sin ``f10!} from the guilt\index{guilt ``f10!} of sin, and condemning\index{Punishment:condemnation ``f10!} wrath of God\index{Wrath:of God:for sin ``f10!}, the curse of the\index{Curse:of the law ``f10!} moral\index{God's law:commandments:moral ``f10!} law;\footnote{\ Titus 2:14\index{56 Titus:ch.   2, v.  14, pg. ``f30!}; 1 Thessalonians 1:10\index{52 1 Thessalonians:ch.   1, v.  10, pg. ``f30!}; Galatians 3:13\index{48 Galatians:ch.   3, v.  13, pg. ``f30!}.} and, in their being delivered from this\index{Believers:delivered from:this evil world ``f10!} present evil\index{Christian liberty:deliverance from evil world ``f10!} world, bondage to Satan\index{Satan:bondage to ``f10!}, and dominion of sin\index{Christian liberty:dominion of sin destroyed ``f10!};\footnote{\ Galatians 1:4\index{48 Galatians:ch.   1, v.   4, pg. ``f30!}; Colossians 1:13\index{51 Colossians:ch.   1, v.  13, pg. ``f30!}; Acts 26:18\index{44 Acts:ch.  26, v.  18, pg. ``f30!}; Romans 6:14\index{45 Romans:ch.   6, v.  14, pg. ``f30!}.} from the evil of afflictions, the\index{Salvation:release from evil afflictions ``f10!} sting of death\index{Death:believers:delivered from sting of ``f10!}, the victory of\index{Death:defeated by Christ ``f10!} the grave, and everlasting\index{Punishment:everlasting:damnation ``f10!} damnation;\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}8\index{45 Romans:ch.   8, v.  28, pg. ``f30!}; Psalm 119:71\index{19 Psalms:ch. 119, v.  71, pg. ``f30!}; 1 Corinthians 15:54-57\index{46 1 Corinthians:ch.  15, v.  54-57, pg. ``f30!}; Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}.} as also, in their free access to God\index{Believers:access to God only through Christ ``f10!},\footnote{\ Romans 5:1\index{45 Romans:ch.   5, v.   1, pg. ``f30!}{}-2\index{45 Romans:ch.   5, v.   1-2, pg. ``f30!}.} and their yielding\index{God's law:commandments:obedience:required ``f10!} obedience unto Him, not out of slavish\index{God's law:commandments:obedience:not of slavish fear ``f10!} fear, but\index{Fear:slavish ``f10!} a child-like\index{Love toward God:childlike ``f10!} love and willing mind.\index{God's law:commandments:obedience:willing mind ``f10!}\footnote{\ Romans 8:1\index{45 Romans:ch.   8, v.   1, pg. ``f30!}4-15\index{45 Romans:ch.   8, v.  14-15, pg. ``f30!}; 1 John 4:1\index{62 1 John:ch.   4, v.   1, pg. ``f30!}8\index{62 1 John:ch.   4, v.  18, pg. ``f30!}.} All which were common also to believers under\index{Christian liberty:believers under the law ``f10!} the law.\footnote{\ Galatians 3:9, 14\index{48 Galatians:ch.   3, v.   9, 14, pg. ``f30!}.} But, under the new testament\index{Scripture:New Testament ``f10!}, the liberty of Christians\index{Christian liberty:enlarged under NT economy ``f10!} is further enlarged, in their\index{Believers:freed from:yoke of the law ``f10!} freedom from the yoke\index{Ceremonial law:yoke ``f10!} of the ceremonial\index{Ceremonial law:fore-signifying Christ ``f10!} law, to which the Jewish\index{Church:Jewish ``f10!} Church was subjected\index{Ceremonial law:subjected Jewish Church ``f10!};\footnote{\ Galatians 4:1-3\index{48 Galatians:ch.   4, v.   1-3, pg. ``f30!}, 6-7\index{48 Galatians:ch.   4, v.   1-3, 6-7, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}; Acts 15:10-11\index{44 Acts:ch.  15, v.  10-11, pg. ``f30!}.} and in greater boldness\index{Prayer:with boldness ``f10!} of access to the\index{Believers:bold access to God's throne ``f10!} throne of grace\index{Grace:throne of ``f10!},\footnote{\ Hebrews 4:14, 16\index{58 Hebrews:ch.   4, v.  14, 16, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}9-22\index{58 Hebrews:ch.  10, v.  19-22, pg. ``f30!}.} and in fuller communications of the free Spirit\index{Holy Spirit:instructs the elect ``f10!} of God, than believers under the law did ordinarily partake\index{Believers:benefits:in this life ``f10!} of.\footnote{\ John 7:38-39\index{43 John:ch.   7, v.  38-39, pg. ``f30!}; 2 Corinthians 3:13, 17-18\index{47 2 Corinthians:ch.   3, v.  13, 17-18, pg. ``f30!}.}   
\bigskip
20.2\ \ God alone is Lord of the conscience\index{Conscience:God alone lord of ``f10!},\footnote{\ James 4:12\index{59 James:ch.   4, v.  12, pg. ``f30!}; Romans 14:4\index{45 Romans:ch.  14, v.   4, pg. ``f30!}.} and has left it free from the doctrines\index{Doctrine:false:of men ``f10!} and commandments\index{Commandments of men:false ``f10!} of men, which are in any thing contrary to His Word; or beside it, in matters of faith\index{Faith:contrary opinions to the ``f10!} or worship\index{Worship:of God ``f10!}.\footnote{\ Acts 4:19\index{44 Acts:ch.   4, v.  19, pg. ``f30!}; Acts 5:2\index{44 Acts:ch.   5, v.   2, pg. ``f30!}9\index{44 Acts:ch.   5, v.  29, pg. ``f30!}; 1 Corinthians 7:23\index{46 1 Corinthians:ch.   7, v.  23, pg. ``f30!}; Matthew 23:8-10\index{40 Matthew:ch.  23, v.   8-10, pg. ``f30!}; 2 Corinthians 1:24\index{47 2 Corinthians:ch.   1, v.  24, pg. ``f30!}; Matthew 15:9\index{40 Matthew:ch.  15, v.   9, pg. ``f30!}.} So that, to believe such doctrines, or to obey such commands, out of conscience\index{Conscience:influenced falsely by others ``f10!},\footnote{\ Colossians 2:2\index{51 Colossians:ch.   2, v.   2, pg. ``f30!}0, 22-23\index{51 Colossians:ch.   2, v.  20, 22-23, pg. ``f30!}; Galatians 1:10\index{48 Galatians:ch.   1, v.  10, pg. ``f30!}; Galatians 2:4-5\index{48 Galatians:ch.   2, v.   4-5, pg. ``f30!}; Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}.} is to betray true liberty of conscience\index{Conscience:liberty of:defined ``f10!}: and the requiring\index{Salvation:faith alone required ``f10!} of an implicit\index{Faith:implicit ``f10!} faith, and an absolute\index{Conscience:liberty of:not absolute obedience when contrary to God's Word ``f10!} and blind obedience is to destroy\index{Conscience:liberty of:destroyed ``f10!} liberty of conscience, and reason also.\footnote{\ Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}7\index{45 Romans:ch.  10, v.  17, pg. ``f30!}; Romans 14:23\index{45 Romans:ch.  14, v.  23, pg. ``f30!}; Isaiah 8:20\index{23 Isaiah:ch.   8, v.  20, pg. ``f30!}; Acts 17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}; John 4:22\index{43 John:ch.   4, v.  22, pg. ``f30!}; Hosea 5:11\index{28 Hosea:ch.   5, v.  11, pg. ``f30!}; Revelation 13:12, 16-17\index{66 Revelation:ch.  13, v.  12, 16-17, pg. ``f30!}; Jeremiah 8:9\index{24 Jeremiah:ch.   8, v.   9, pg. ``f30!}.}   
\bigskip
20.3\ \ They who, upon pretence\index{pretence ``f10!} of Christian liberty\index{Christian liberty:defined ``f10!}, do practice any sin, or cherish any lust\index{Sexual immorality:lust ``f10!}, do thereby destroy the\index{Conscience:liberty of:destroyed ``f10!} end of Christian liberty, which is, that being delivered out\index{Believers:delivered from:hands of enemies ``f10!} of the hands of our enemies\index{Enemies:believers delivered from ``f10!}, we might serve the\index{Love toward God:in service ``f10!} Lord, without fear\index{Fear:none ``f10!}, in holiness and\index{Believers:holy:being perfected ``f10!} righteousness before\index{Believers:to serve God in righteousness ``f10!} Him, all the days of our life.\footnote{\ Galatians 5:1\index{48 Galatians:ch.   5, v.   1, pg. ``f30!}3\index{48 Galatians:ch.   5, v.  13, pg. ``f30!}; 1 Peter 2:16\index{60 1 Peter:ch.   2, v.  16, pg. ``f30!}; 2 Peter 2:19\index{61 2 Peter:ch.   2, v.  19, pg. ``f30!}; John 8:34\index{43 John:ch.   8, v.  34, pg. ``f30!}; Luke 1:74-75\index{42 Luke:ch.   1, v.  74-75, pg. ``f30!}.}   
\bigskip
20.4\ \ And because the powers which God has ordained, and\index{Civil magistrate:ordained by God ``f10!} the liberty which\index{Christian liberty:purchased by Christ ``f10!} Christ has purchased\index{Redemption:purchased by Christ ``f10!}, are not intended by God to destroy\index{Conscience:liberty of:destroyed ``f10!}, but mutually to uphold and\index{Christian liberty:upholds just laws of men ``f10!} preserve one\index{Human life:preservation ``f10!} another; they who, upon pretence\index{pretence ``f10!} of Christian liberty\index{Christian liberty:defined ``f10!}, shall oppose\index{Liberty:not oppose lawful governance ``f10!} any lawful power, or the lawful exercise of it, whether it be civil or ecclesiastical\index{Church government:ecclesiastical powers not to be resisted ``f10!}, resist the ordinance of\index{Worship:ordinances:to be:esteemed ``f10!} God.\footnote{\ Matthew 12:25\index{40 Matthew:ch.  12, v.  25, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}, 16\index{60 1 Peter:ch.   2, v.  13-14, 16, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}{}-8\index{45 Romans:ch.  13, v.   1-8, pg. ``f30!}; Hebrews 13:17\index{58 Hebrews:ch.  13, v.  17, pg. ``f30!}.} And, for their publishing\index{Church discipline:for opinions contrary to truth ``f10!} of such opinions, or maintaining of such practices, as are contrary to the light of nature, or to the known principles of Christianity\index{Christianity ``f10!}, whether concerning faith\index{Faith:contrary opinions to the ``f10!}, worship, or\index{Worship:of God ``f10!} conversation\index{Believers:to walk faithfully in obedience (conversation) ``f10!}; or, to the power of godliness\index{God's attributes:power of godliness ``f10!}; or, such erroneous\index{Falsehood:false opinions ``f10!} opinions or practices, as either in their own nature, or in the manner of publishing or maintaining them, are destructive\index{Heresy:heresies ``f10!} to the external\index{Heresy:destructive or external peace of Church ``f10!} peace and order\index{Church:to maintain peace and order ``f10!} which Christ has established\index{Church:order established by Christ ``f10!} in the Church, they\index{Church:institution ``f10!} may lawfully be\index{Church discipline:heresies lawfully censured ``f10!} called to account, and proceeded against by the censures of\index{Church discipline:censures ``f10!} the Church,\footnote{\ Romans 1:32\index{45 Romans:ch.   1, v.  32, pg. ``f30!} with 1 Corinthians 5:1\index{46 1 Corinthians:ch.   5, v.   1, pg. ``f30!}, 5, 11, 13\index{46 1 Corinthians:ch.   5, v.   1, 5, 11, 13, pg. ``f30!}; 2 John 10-11\index{63 2 John:v. 10-11, pg. ``f30!} and 2 Thessalonians 3:1\index{53 2 Thessalonians:ch.   3, v.   1, pg. ``f30!}4\index{53 2 Thessalonians:ch.   3, v.  14, pg. ``f30!}; 1 Timothy 6:3\index{54 1 Timothy:ch.   6, v.   3, pg. ``f30!}{}-5\index{54 1 Timothy:ch.   6, v.   3-5, pg. ``f30!}; Titus 1:10-11, 13\index{56 Titus:ch.   1, v.  10-11, 13, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!} with Matthew 18:15\index{40 Matthew:ch.  18, v.  15, pg. ``f30!}{}-17\index{40 Matthew:ch.  18, v.  15-17, pg. ``f30!}; 1 Timothy 1:19-20\index{54 1 Timothy:ch.   1, v.  19-20, pg. ``f30!}; Revelation 2:2, 14-15, 20\index{66 Revelation:ch.   2, v.   2, 14-15, 20, pg. ``f30!}; Revelation 3:9\index{66 Revelation:ch.   3, v.   9, pg. ``f30!}.} and by the power of the civil magistrate\index{Civil magistrate:to punish evil doers ``f10!}.\footnote{\ Deuteronomy 13:6-12\index{05 Deuteronomy:ch.  13, v.   6-12, pg. ``f30!}; Romans 13:3\index{45 Romans:ch.  13, v.   3, pg. ``f30!}{}-4\index{45 Romans:ch.  13, v.   3-4, pg. ``f30!} with 2 John 10-11\index{63 2 John:v. 10-11, pg. ``f30!}; Ezra 7:23, 25-28\index{15 Ezra:ch.   7, v.  23, 25-28, pg. ``f30!}; Revelation 17:12, 16-17\index{66 Revelation:ch.  17, v.  12, 16-17, pg. ``f30!}; Nehemiah 13:15, 17\index{16 Nehemiah:ch.  13, v.  15, 17, pg. ``f30!}, 21-22, 25, 30\index{16 Nehemiah:ch.  13, v.  15, 17, 21-22, 25, 30, pg. ``f30!}; 2 Kings 23:5-6, 9, 20-21\index{12 2 Kings:ch.  23, v.   5-6, 9, 20-21, pg. ``f30!}; 2 Chronicles 34:33\index{14 2 Chronicles:ch.  34, v.  33, pg. ``f30!}; 2 Chronicles 15:12-13\index{14 2 Chronicles:ch.  15, v.  12-13, pg. ``f30!}, 16\index{14 2 Chronicles:ch.  15, v.  12-13, 16, pg. ``f30!}; Daniel 3:29\index{27 Daniel:ch.   3, v.  29, pg. ``f30!}; 1 Timothy 2:2\index{54 1 Timothy:ch.   2, v.   2, pg. ``f30!}; Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; Zechariah 13:2, 3\index{38 Zechariah:ch.  13, v.   2, 3, pg. ``f30!}.}  

\section{Chapter 21 -- Of Religious Worship and the Sabbath Day} \index{Worship:glorifying God ``f10!}\index{Worship:on the Sabbath ``f10!}\index{Sabbath:for worship ``f10!}21.1\ \ The light of nature shows\index{Revelation:natural:declares there is a God ``f10!} that there is a God, who has lordship\index{God's attributes:sovereign lordship ``f10!} and sovereignty over all, is good, and does good unto all, and is therefore to be feared\index{Fear:of God:reverence ``f10!}, loved\index{Love toward God:for God ``f10!}, praised\index{Worship:praising God ``f10!}, called upon, trusted\index{trust in God ``f10!} in, and served\index{Love toward God:in service ``f10!}, with all the heart, and with all the soul\index{Love toward God:with all our:soul ``f10!}, and with all the might.\footnote{\ Romans 1:20\index{45 Romans:ch.   1, v.  20, pg. ``f30!}; Acts 17:24\index{44 Acts:ch.  17, v.  24, pg. ``f30!}; Psalm 119:68\index{19 Psalms:ch. 119, v.  68, pg. ``f30!}; Jeremiah 10:7\index{24 Jeremiah:ch.  10, v.   7, pg. ``f30!}; Psalm 31:23\index{19 Psalms:ch.  31, v.  23, pg. ``f30!}; Psalm 18:3\index{19 Psalms:ch.  18, v.   3, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}2\index{45 Romans:ch.  10, v.  12, pg. ``f30!}; Psalm 62:8\index{19 Psalms:ch.  62, v.   8, pg. ``f30!}; Joshua 24:14\index{06 Joshua:ch.  24, v.  14, pg. ``f30!}; Mark 12:33\index{41 Mark:ch.  12, v.  33, pg. ``f30!}.} But the acceptable way of worship\index{Worship:of God ``f10!}ping the true God is instituted by Himself\index{Worship:correct:instituted by God ``f10!}, and so limited\index{Worship:only acts approved by God ``f10!} by His own revealed will\index{Revelation:God's will ``f10!}, that He may not be worshipped according to the imaginations and\index{Worship:false:imaginations and devices of men ``f10!} devices of men, or the suggestions of Satan\index{Satan:tempter ``f10!}, under any visible\index{Idolatry:worship of likeness of any created thing ``f10!} representation\index{Idolatry:any visible representation of God or gods ``f10!}, or any other way not prescribed in the holy\index{Scripture:holy ``f10!} Scripture.\footnote{\ Deuteronomy 12:32\index{05 Deuteronomy:ch.  12, v.  32, pg. ``f30!}; Matthew 15:9\index{40 Matthew:ch.  15, v.   9, pg. ``f30!}; Acts 17:25\index{44 Acts:ch.  17, v.  25, pg. ``f30!}; Matthew 4:9-10\index{40 Matthew:ch.   4, v.   9-10, pg. ``f30!}; Deuteronomy 4:15-20\index{05 Deuteronomy:ch.   4, v.  15-20, pg. ``f30!}; Exodus 20:4-6\index{02 Exodus:ch.  20, v.   4-6, pg. ``f30!}; Colossians 2:2\index{51 Colossians:ch.   2, v.   2, pg. ``f30!}3\index{51 Colossians:ch.   2, v.  23, pg. ``f30!}.}   
\bigskip
21.2\ \ Religious\index{Worship:glorifying God ``f10!} worship is to\index{Worship:of Trinity ``f10!} be given to God, the Father\index{Father:God the ``f10!}, Son\index{Jesus:Son of God ``f10!}, and Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}; and to Him alone;\footnote{\ Matthew 4:10\index{40 Matthew:ch.   4, v.  10, pg. ``f30!} with John 5:23\index{43 John:ch.   5, v.  23, pg. ``f30!} and 2 Corinthians 13:1\index{47 2 Corinthians:ch.  13, v.   1, pg. ``f30!}4\index{47 2 Corinthians:ch.  13, v.  14, pg. ``f30!}.} not to angels\index{Angels:angels ``f10!}, saints\index{Believers:saints ``f10!}, or any other creature\index{creatures ``f10!}:\footnote{\ Colossians 2:18\index{51 Colossians:ch.   2, v.  18, pg. ``f30!}; Revelation 19:10\index{66 Revelation:ch.  19, v.  10, pg. ``f30!}; Romans 1:25\index{45 Romans:ch.   1, v.  25, pg. ``f30!}.} and since the fall\index{Sin:original:first transgression (of Adam) ``f10!}, not without a Mediator\index{Jesus:mediator:of covenant ``f10!}; nor in the mediation of any other but of Christ alone.\footnote{\ John 14:6\index{43 John:ch.  14, v.   6, pg. ``f30!}; 1 Timothy 2:5\index{54 1 Timothy:ch.   2, v.   5, pg. ``f30!}; Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}8\index{49 Ephesians:ch.   2, v.  18, pg. ``f30!}; Colossians 3:17\index{51 Colossians:ch.   3, v.  17, pg. ``f30!}.}   
\bigskip
21.3\ \ Prayer, with\index{Prayer:with thanksgiving ``f10!} thanksgiving\index{Prayer:thanking God ``f10!}, being one special part of religious worship,\index{Worship:elements ``f10!}\footnote{\ Philippians 4:6\index{50 Philippians:ch.   4, v.   6, pg. ``f30!}.} is by God required of all men:\footnote{\ Psalm 65:2\index{19 Psalms:ch.  65, v.   2, pg. ``f30!}.} and that it may be accepted, it is to be made in\index{Prayer:accepted in Christ ``f10!} the name of the Son\index{Jesus:Son of God ``f10!},\footnote{\ John 14:13-14\index{43 John:ch.  14, v.  13-14, pg. ``f30!}; 1 Peter 2:5\index{60 1 Peter:ch.   2, v.   5, pg. ``f30!}.} by the help of His\index{Prayer:helped by Holy Spirit ``f10!} Spirit,\footnote{\ Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}6\index{45 Romans:ch.   8, v.  26, pg. ``f30!}.} according to His will,\footnote{\ 1 John 5:14\index{62 1 John:ch.   5, v.  14, pg. ``f30!}.} with understanding, reverence, humility\index{Humility:before God ``f10!}, fervency\index{Prayer:with fervency ``f10!}, faith, love\index{Scripture:to be:loved in faith ``f10!}, and perseverance;\index{Prayer:with perseverance ``f10!}\footnote{\ Psalm 47:7\index{19 Psalms:ch.  47, v.   7, pg. ``f30!}; Ecclesiastes 5:1\index{21 Ecclesiastes:ch.   5, v.   1, pg. ``f30!}{}-2\index{21 Ecclesiastes:ch.   5, v.   1-2, pg. ``f30!}; Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}; Genesis 18:27\index{01 Genesis:ch.  18, v.  27, pg. ``f30!}; James 5:16\index{59 James:ch.   5, v.  16, pg. ``f30!}; James 1:6\index{59 James:ch.   1, v.   6, pg. ``f30!}{}-7\index{59 James:ch.   1, v.   6-7, pg. ``f30!}; Mark 11:24\index{41 Mark:ch.  11, v.  24, pg. ``f30!}; Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}2\index{40 Matthew:ch.   6, v.  12, pg. ``f30!}, 14-15\index{40 Matthew:ch.   6, v.  12, 14-15, pg. ``f30!}; Colossians 4:2\index{51 Colossians:ch.   4, v.   2, pg. ``f30!}; Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}.} and, if vocal\index{Prayer:vocal in a known language ``f10!}, in a known tongue.\footnote{\ 1 Corinthians 14:14\index{46 1 Corinthians:ch.  14, v.  14, pg. ``f30!}.}   
\bigskip
21.4\ \ Prayer is to be made for things lawful\index{Prayer:for lawful things ``f10!};\footnote{\ 1 John 5:14\index{62 1 John:ch.   5, v.  14, pg. ``f30!}.} and for all sorts of men living, or that shall live hereafter:\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}; John 17:2\index{43 John:ch.  17, v.   2, pg. ``f30!}0\index{43 John:ch.  17, v.  20, pg. ``f30!}; 2 Samuel 7:29\index{10 2 Samuel:ch.   7, v.  29, pg. ``f30!}; Ruth 4:12\index{08 Ruth:ch.   4, v.  12, pg. ``f30!}.} but not for the dead\index{Prayer:not to be offered for the dead ``f10!},\footnote{\ 2 Samuel 12:21-23\index{10 2 Samuel:ch.  12, v.  21-23, pg. ``f30!} with Luke 16:25-26\index{42 Luke:ch.  16, v.  25-26, pg. ``f30!}; Revelation 14:13\index{66 Revelation:ch.  14, v.  13, pg. ``f30!}.} nor for those of whom it may be known that they have sinned the sin\index{Sin:sinned the sin unto death ``f10!} unto death\index{Death:committed sin unto ``f10!}.\footnote{\ 1 John 5:16\index{62 1 John:ch.   5, v.  16, pg. ``f30!}.}   
\bigskip
21.5\ \ The reading\index{Scripture:to be:read and heard ``f10!} of the Scriptures with godly fear,\footnote{\ Acts 15:2\index{44 Acts:ch.  15, v.   2, pg. ``f30!}1\index{44 Acts:ch.  15, v.  21, pg. ``f30!}; Revelation 1:3\index{66 Revelation:ch.   1, v.   3, pg. ``f30!}.} the sound preaching\index{Preaching:sound doctrine ``f10!}\footnote{\ 2 Timothy 4:2\index{55 2 Timothy:ch.   4, v.   2, pg. ``f30!}.} and conscionable\index{Preaching:hearing:conscionable ``f10!} hearing of the Word, in obedience\index{God's law:commandments:obedience:required ``f10!} unto God, with understanding\index{Scripture:to be:read with understanding ``f10!}, faith and reverence;\footnote{\ James 1:22\index{59 James:ch.   1, v.  22, pg. ``f30!}; Acts 10:33\index{44 Acts:ch.  10, v.  33, pg. ``f30!}; Matthew 13:19\index{40 Matthew:ch.  13, v.  19, pg. ``f30!}; Hebrews 4:2\index{58 Hebrews:ch.   4, v.   2, pg. ``f30!}; Isaiah 66:2\index{23 Isaiah:ch.  66, v.   2, pg. ``f30!}.} singing of psalms\index{Psalms:singing ``f10!} with grace\index{Psalms:to be sung with grace ``f10!} in the heart;\index{Worship:grace in heart, as singing Psalms ``f10!}\footnote{\ Colossians 3:16\index{51 Colossians:ch.   3, v.  16, pg. ``f30!}; Ephesians 5:19\index{49 Ephesians:ch.   5, v.  19, pg. ``f30!}; James 5:13\index{59 James:ch.   5, v.  13, pg. ``f30!}.} as also, the due administration\index{Worship:administration of sacraments ``f10!} and worthy receiving of the sacraments\index{Sacraments:holy:ordinances ``f10!} instituted by Christ\index{Sacraments:instituted by:Christ ``f10!}; are all parts of the ordinary\index{Sacraments:part of ordinary worship ``f10!} religious worship of God\index{Worship:of God ``f10!}:\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  23-29, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}.} beside religious\index{Oaths and Vows:religious worship ``f10!} oaths\index{Oaths and Vows:oaths ``f10!},\footnote{\ Deuteronomy 6:13\index{05 Deuteronomy:ch.   6, v.  13, pg. ``f30!} with Nehemiah 10:29\index{16 Nehemiah:ch.  10, v.  29, pg. ``f30!}.} vows,\footnote{\ Isaiah 19:21\index{23 Isaiah:ch.  19, v.  21, pg. ``f30!} with Ecclesiastes 5:4-5\index{21 Ecclesiastes:ch.   5, v.   4-5, pg. ``f30!}.} solemn fastings\index{fasting ``f10!},\footnote{\ Joel 2:12\index{29 Joel:ch.   2, v.  12, pg. ``f30!}; Esther 4:16\index{17 Esther:ch.   4, v.  16, pg. ``f30!}; Matthew 9:15\index{40 Matthew:ch.   9, v.  15, pg. ``f30!}; 1 Corinthians 7:5\index{46 1 Corinthians:ch.   7, v.   5, pg. ``f30!}.} and thanksgivings\index{Prayer:thanking God ``f10!}, upon special occasions,\footnote{\ Psalm 107; Esther 9:22\index{17 Esther:ch.   9, v.  22, pg. ``f30!}.} which are, in their several times and seasons\index{Worship:elements in times and seasons ``f10!}, to be used in a holy\index{Scripture:holy ``f10!} and religious manner\index{Worship:acceptable manner ``f10!}.\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}8\index{58 Hebrews:ch.  12, v.  28, pg. ``f30!}.}   
\bigskip
21.6\ \ Neither prayer\index{Prayer:act of worship ``f10!}, nor any other part of religious worship,\index{Worship:elements ``f10!} is now under the Gospel\index{Gospel:gospel ``f10!} either tied unto, or made more acceptable by any place in which it is performed, or towards which it is directed:\footnote{\ John 4:21\index{43 John:ch.   4, v.  21, pg. ``f30!}.} but God is to be worshipped everywhere\index{Worship:in spirit and truth ``f10!},\footnote{\ Malachi 1:11\index{39 Malachi:ch.   1, v.  11, pg. ``f30!}; 1 Timothy 2:8\index{54 1 Timothy:ch.   2, v.   8, pg. ``f30!}.} in spirit and truth\index{God's attributes:truth ``f10!};\footnote{\ John 4:23-24\index{43 John:ch.   4, v.  23-24, pg. ``f30!}.} as in private families\index{family, natural ``f10!}\footnote{\ Jeremiah 10:25\index{24 Jeremiah:ch.  10, v.  25, pg. ``f30!}; Deuteronomy 6:6\index{05 Deuteronomy:ch.   6, v.   6, pg. ``f30!}{}-7\index{05 Deuteronomy:ch.   6, v.   6-7, pg. ``f30!}; Job 1:5\index{18 Job:ch.   1, v.   5, pg. ``f30!}; 2 Samuel 6:18, 20\index{10 2 Samuel:ch.   6, v.  18, 20, pg. ``f30!}; 1 Peter 3:7\index{60 1 Peter:ch.   3, v.   7, pg. ``f30!}; Acts 10:2\index{44 Acts:ch.  10, v.   2, pg. ``f30!}.} daily,\index{Prayer:daily in private families, ``f10!}\footnote{\ Matthew 6:1\index{40 Matthew:ch.   6, v.   1, pg. ``f30!}.} and in secret each\index{Prayer:secret ``f10!} one by himself;\footnote{\ Matthew 6:6\index{40 Matthew:ch.   6, v.   6, pg. ``f30!}; Ephesians 6:18\index{49 Ephesians:ch.   6, v.  18, pg. ``f30!}.} so, more solemnly, in\index{Prayer:solemn ``f10!} the public assemblies, which\index{Worship:assemblies not to be neglected ``f10!} are not carelessly or wilfully to be neglected, or forsaken, when\index{Worship:not to be:forsaken ``f10!} God, by His Word or providence\index{Providence:providence ``f10!}, calls thereunto\index{Worship:assembly called together ``f10!}.\footnote{\ Isaiah 56:6-7\index{23 Isaiah:ch.  56, v.   6-7, pg. ``f30!}; Hebrews 10:25\index{58 Hebrews:ch.  10, v.  25, pg. ``f30!}; Proverbs 1:20-21, 24\index{20 Proverbs:ch.   1, v.  20-21, 24, pg. ``f30!}; Proverbs 8:34\index{20 Proverbs:ch.   8, v.  34, pg. ``f30!}; Acts 13:42\index{44 Acts:ch.  13, v.  42, pg. ``f30!}; Luke 4:16\index{42 Luke:ch.   4, v.  16, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}.}   
\bigskip
21.7\ \ As it is the law of nature, that, in general, a due proportion\index{Sabbath:proportion of time set apart for worship ``f10!} of time be set apart for the worship of God; so\index{Worship:times of ``f10!}, in His Word, by a positive\index{Sabbath:positive, moral, perpetual command ``f10!}, moral\index{God's law:commandments:moral ``f10!}, and perpetual commandment\index{Sabbath:perpetual commandment ``f10!}, binding all\index{Sabbath:binding on all men ``f10!} men, in all ages, He has particularly appointed one day in seven\index{Sabbath:one whole day in seven ``f10!}, for a Sabbath, to\index{Sabbath:seventh day ``f10!} be kept holy\index{Sabbath:to be:kept holy ``f10!} unto Him:\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}, 10\index{02 Exodus:ch.  20, v.   8, 10, pg. ``f30!}{}-11\index{02 Exodus:ch.  20, v.   8, 10-11, pg. ``f30!}; Isaiah 56:2, 4, 6-7\index{23 Isaiah:ch.  56, v.   2, 4, 6-7, pg. ``f30!}.} which, from the beginning of the world to the resurrection of Christ\index{Jesus:resurrection:from the dead ``f10!}, was the last day of\index{Sabbath:last day of week ``f10!} the week; and, from the resurrection of Christ, was changed into the first day\index{Sabbath:first day of week ``f10!} of the week,\footnote{\ Genesis 2:2-3\index{01 Genesis:ch.   2, v.   2-3, pg. ``f30!}; 1 Corinthians 16:1-2\index{46 1 Corinthians:ch.  16, v.   1-2, pg. ``f30!}; Acts 20:7\index{44 Acts:ch.  20, v.   7, pg. ``f30!}.} which, in Scripture, is\index{Sabbath:Lord's Day ``f10!} called the Lord's Day,\footnote{\ Revelation 1:10\index{66 Revelation:ch.   1, v.  10, pg. ``f30!}.} and is to be continued to the end of the world, as the Christian Sabbath\index{Sabbath:Christian ``f10!}.\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}, 10\index{02 Exodus:ch.  20, v.   8, 10, pg. ``f30!} with Matthew 5:17\index{40 Matthew:ch.   5, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.   5, v.  17-18, pg. ``f30!}.}   
\bigskip
21.8\ \ This Sabbath is\index{Sabbath:to be:kept holy ``f10!} then kept holy unto the Lord, when men, after a due preparing\index{Heart:prepared for Sabbath ``f10!} of their hearts, and ordering of the\index{Church councils:directions for ordering public worship ``f10!}ir common affairs beforehand\index{Sabbath:order common affairs before ``f10!}, do not only observe an holy rest, all the day, from their own works\index{Sabbath:cease:own works ``f10!}, words, and thoughts about\index{Sabbath:avoid unnecessary, worldly thoughts ``f10!} their worldly employments and recreations\index{Sabbath:cease:worldly employments and recreations ``f10!},\footnote{\ Exodus 20:8\index{02 Exodus:ch.  20, v.   8, pg. ``f30!}; Exodus 16:23\index{02 Exodus:ch.  16, v.  23, pg. ``f30!}, 25-26, 29-30\index{02 Exodus:ch.  16, v.  23, 25-26, 29-30, pg. ``f30!}; Exodus 31:15\index{02 Exodus:ch.  31, v.  15, pg. ``f30!}{}-17\index{02 Exodus:ch.  31, v.  15-17, pg. ``f30!}; Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}; Nehemiah 13:15-19, 21-22\index{16 Nehemiah:ch.  13, v.  15-19, 21-22, pg. ``f30!}.} but also are taken up the whole time in the public and private exercises\index{Worship:private exercises ``f10!} of His worship, and in\index{Worship:of God ``f10!} the duties of necessity and\index{Sabbath:works of necessity and mercy ``f10!} mercy.\footnote{\ Isaiah 58:13\index{23 Isaiah:ch.  58, v.  13, pg. ``f30!}; Matthew 12:1-13\index{40 Matthew:ch.  12, v.   1-13, pg. ``f30!}.}  

\section{Chapter 22 -- Of Lawful Oaths and Vows} \index{Oaths and Vows:lawful ``f10!}\index{Oaths and Vows:vows ``f10!}22.1\ \ A lawful\index{Oaths and Vows:lawful ``f10!} oath\index{Oaths and Vows:oaths ``f10!} is a part of religious worship,\index{Worship:elements ``f10!}\footnote{\ Deuteronomy 10:20\index{05 Deuteronomy:ch.  10, v.  20, pg. ``f30!}.} wherein, upon just occasion, the person swearing\index{Oaths and Vows:swear only by God's name ``f10!} solemnly calls God\index{Oaths and Vows:solemnly calls God to witness ``f10!} to witness what he asserts\index{Oaths and Vows:assertion with God as witness ``f10!}, or promises\index{Oaths and Vows:promises made ``f10!}, and to judge him according to the truth or\index{Oaths and Vows:assertion of truth ``f10!} falsehood\index{Oaths and Vows:consistent with truth or falsehood ``f10!} of what he swears.\footnote{\index{02 Exodus:ch.  20, v.   7, pg. ``f30!}\index{03 Leviticus:ch.  19, v.  12, pg. ``f30!}\index{47 2 Corinthians:ch.   1, v.  23, pg. ``f30!}\index{14 2 Chronicles:ch.   6, v.  22-23, pg. ``f30!}   
\bigskip
22.2\ \ The name of God only is that by which men ought to swear\index{Oaths and Vows:swear only by God's name ``f10!}; and therein it is to be used with all holy\index{God's name:holy ``f10!} fear and\index{Fear:of God:reverence ``f10!} reverence.\footnote{\index{05 Deuteronomy:ch.   6, v.  13, pg. ``f30!} rashly\index{rashness ``f10!}, by that glorious and\index{God's name:glorious and dreadful ``f10!} dreadful Name; or, to swear at all by any other thing, is sinful, and to be abhorred.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.   7, pg. ``f30!}; Jeremiah 5:7\index{24 Jeremiah:ch.   5, v.   7, pg. ``f30!}; Matthew 5:34, 37\index{40 Matthew:ch.   5, v.  34, 37, pg. ``f30!}; James 5:12\index{59 James:ch.   5, v.  12, pg. ``f30!}.} Yet, as in matters of weight and moment, an oath\index{Oaths and Vows:oaths ``f10!} is warranted\index{Oaths and Vows:warranted by the Word of God ``f10!} by the Word of God\index{Scripture:word of God:word of God ``f10!}, under the New Testament\index{Scripture:New Testament ``f10!}, as well as under the Old;\footnote{\ Hebrews 6:16\index{58 Hebrews:ch.   6, v.  16, pg. ``f30!}; 2 Corinthians 1:23\index{47 2 Corinthians:ch.   1, v.  23, pg. ``f30!}; Isaiah 65:16\index{23 Isaiah:ch.  65, v.  16, pg. ``f30!}.} so a lawful\index{Oaths and Vows:lawful ``f10!} oath, being imposed by lawful authority, in such matters ought to be taken.\footnote{\ I Kings 8:31; Nehemiah 13:25\index{16 Nehemiah:ch.  13, v.  25, pg. ``f30!}; Ezra 10:5\index{15 Ezra:ch.  10, v.   5, pg. ``f30!}.}   
\bigskip
22.3\ \ Whosoever takes an oath\index{Oaths and Vows:oaths ``f10!} ought duly to consider the weightiness\index{Oaths and Vows:consider the weightiness ``f10!} of so solemn an\index{Oaths and Vows:solemn act ``f10!} act; and therein to avouch\index{Oaths and Vows:avouch only truth ``f10!} nothing, but what he is fully persuaded is the truth\index{Oaths and Vows:assertion of truth ``f10!}.\footnote{\ Exodus 20:7\index{02 Exodus:ch.  20, v.   7, pg. ``f30!}; Jeremiah 4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}.} Neither may any man bind himself\index{Oaths and Vows:bind:only to good and just ``f10!} by oath to anything but what is good and just, and what he believes so to be, and what he is able and resolved to\index{Oaths and Vows:resolve to perform ``f10!} perform.\footnote{\ Genesis 24:2-3, 5-6, 8-9\index{01 Genesis:ch.  24, v.   2-3, 5-6, 8-9, pg. ``f30!}.} Yet is it a sin to refuse\index{Oaths and Vows:not to:refuse lawful ``f10!} an oath touching anything that is good and just, being imposed by lawful authority\index{Oaths and Vows:lawful ``f10!}.\footnote{\ Numbers 5:19, 21\index{04 Numbers:ch.   5, v.  19, 21, pg. ``f30!}; Nehemiah 5:12\index{16 Nehemiah:ch.   5, v.  12, pg. ``f30!}; Exodus 22:7-11\index{02 Exodus:ch.  22, v.   7-11, pg. ``f30!}.}   
\bigskip
22.4\ \ An oath\index{Oaths and Vows:oaths ``f10!} is to be taken in the plain and\index{Oaths and Vows:take in plain and common sense of words ``f10!} common sense of the words, without equivocation\index{Oaths and Vows:not with equivocation ``f10!}, or mental reservation\index{Oaths and Vows:no mental reservations ``f10!}.\footnote{\ Jeremiah 4:2\index{24 Jeremiah:ch.   4, v.   2, pg. ``f30!}; Psalm 24:4\index{19 Psalms:ch.  24, v.   4, pg. ``f30!}.} It cannot oblige to sin: but in anything not sinful, being taken, it binds to\index{Oaths and Vows:bind:performance ``f10!} performance, although to a man's own hurt.\footnote{\ 1 Samuel 25:22, 32-34\index{09 1 Samuel:ch.  25, v.  22, 32-34, pg. ``f30!}; Psalm 15:4\index{19 Psalms:ch.  15, v.   4, pg. ``f30!}.} Not is it to be violated\index{Oaths and Vows:not to:be violated ``f10!}, although made to heretics\index{heretics ``f10!}, or infidels\index{Oaths and Vows:to be kept, even if made to infidels ``f10!}.\footnote{\ Ezekiel 17:16, 18-19\index{26 Ezekiel:ch.  17, v.  16, 18-19, pg. ``f30!}; Joshua 9:18-19\index{06 Joshua:ch.   9, v.  18-19, pg. ``f30!} with 2 Samuel 21:1\index{10 2 Samuel:ch.  21, v.   1, pg. ``f30!}.}   
\bigskip
22.5\ \ A vow is\index{Oaths and Vows:vow defined ``f10!} of the like nature with a promissory\index{Oaths and Vows:promissory ``f10!} oath\index{Oaths and Vows:oaths ``f10!}, and ought to be made with the like religious care\index{Oaths and Vows:religious worship ``f10!}, and to be performed with the like faithfulness\index{faithfulness ``f10!}.\footnote{\ Isaiah 19:21\index{23 Isaiah:ch.  19, v.  21, pg. ``f30!}; Ecclesiastes 5:4-6\index{21 Ecclesiastes:ch.   5, v.   4-6, pg. ``f30!}; Psalm 61:8\index{19 Psalms:ch.  61, v.   8, pg. ``f30!}; Psalm 66:13-14\index{19 Psalms:ch.  66, v.  13-14, pg. ``f30!}.}   
\bigskip
22.6\ \ It is not to be made to any creature\index{creatures ``f10!}, but to God alone:\footnote{\ Psalm 76:11\index{19 Psalms:ch.  76, v.  11, pg. ``f30!}; Jeremiah 44:25-26\index{24 Jeremiah:ch.  44, v.  25-26, pg. ``f30!}.} and that it may be accepted, it is to be made voluntarily\index{Oaths and Vows:voluntary to be accepted ``f10!}, out of faith\index{Faith:when making oath or vow ``f10!}, and conscience of\index{Believers:desire to obey God ``f10!} duty, in way of thankfulness\index{Prayer:thanking God ``f10!} for mercy received\index{Mercy:thankfulness for ``f10!}, or for the obtaining\index{Prayer:faith for obtaining wants ``f10!} of what we want; whereby we more strictly bind ourselves\index{Oaths and Vows:bind:only to good and just ``f10!} to necessary duties\index{Oaths and Vows:bind:only to necessary duties ``f10!}; or to other things, so far and so long as they may fitly conduce thereunto.\footnote{\ Deuteronomy 23:21-23\index{05 Deuteronomy:ch.  23, v.  21-23, pg. ``f30!}; Psalm 50:14\index{19 Psalms:ch.  50, v.  14, pg. ``f30!}; Genesis 28:20\index{01 Genesis:ch.  28, v.  20, pg. ``f30!}{}-22\index{01 Genesis:ch.  28, v.  20-22, pg. ``f30!}; 1 Samuel 1:11\index{09 1 Samuel:ch.   1, v.  11, pg. ``f30!}; Psalm 66:13-14\index{19 Psalms:ch.  66, v.  13-14, pg. ``f30!}; Psalm 132:2-5\index{19 Psalms:ch. 132, v.   2-5, pg. ``f30!}.}   
\bigskip
22.7\ \ No man may vow to do anything forbidden\index{Oaths and Vows:not for things:forbidden by Word of God ``f10!} in the Word of God\index{Scripture:word of God:word of God ``f10!}, or what would hinder any\index{Oaths and Vows:bind:only to good and just ``f10!} duty therein\index{God's law:commandments:duties:required by God ``f10!} commanded, or\index{God's law:commandments:duties:not to hinder ``f10!} which is not in his own power\index{Oaths and Vows:not for things:not in person's power ``f10!}, and for the performance whereof he has no promise of ability\index{Oaths and Vows:not for things:unable to perform ``f10!} from God.\footnote{\ Acts 23:12, 14\index{44 Acts:ch.  23, v.  12, 14, pg. ``f30!}; Mark 6:26\index{41 Mark:ch.   6, v.  26, pg. ``f30!}; Numbers 30:5, 8, 12-13\index{04 Numbers:ch.  30, v.   5, 8, 12-13, pg. ``f30!}.} In which respects, Popish\index{Popish rituals ``f10!} monastical\index{Oaths and Vows:not to:be Popish monastical ``f10!} vows of perpetual single\index{Oaths and Vows:not to:perpetual single life ``f10!} life, professed poverty\index{Oaths and Vows:not to:professed poverty ``f10!}, and regular obedience\index{Salvation:not by:obedience ``f10!}, are so far from being degrees\index{Oaths and Vows:none of higher degrees ``f10!} of higher perfection, that\index{perfectionism ``f10!} they are superstitious\index{Superstitions:superstitions ``f10!} and sinful snares, in which no Christian may entangle himself.\footnote{\ Matthew 19:11-12\index{40 Matthew:ch.  19, v.  11-12, pg. ``f30!}; 1 Corinthians 7:2, 9\index{46 1 Corinthians:ch.   7, v.   2, 9, pg. ``f30!}; Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}; 1 Peter 4:2\index{60 1 Peter:ch.   4, v.   2, pg. ``f30!}; 1 Corinthians 7:23\index{46 1 Corinthians:ch.   7, v.  23, pg. ``f30!}.}  

\section{Chapter 23 -- Of the Civil Magistrate} \index{Civil magistrate:office of ``f10!}23.1\ \ God, the supreme\index{God's attributes:sovereign lordship ``f10!} Lord and King\index{Jesus:King ``f10!} of all the world, has ordained\index{Civil magistrate:ordained by God ``f10!} civil magistrates, to be, under Him, over the people, for His own glory\index{God's attributes:glory:his own ``f10!}, and the public good: and, to this end, has armed them with the power of the sword\index{Civil magistrate:power of sword ``f10!}, for the defence\index{Civil magistrate:to defend citizens ``f10!} and encouragement of\index{Civil magistrate:to encourage those doing good ``f10!} them that are good, and for the punishment\index{Punishment:of sin ``f10!} of evil doers\index{Civil magistrate:to punish evil doers ``f10!}.\footnote{\ Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}{}-4\index{45 Romans:ch.  13, v.   1-4, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}.}   
\bigskip
23.2\ \ It is lawful for Christians\index{Civil magistrate:lawful for Christians to execute office ``f10!} to accept and execute the office of a magistrate, when called thereunto;\footnote{\ Proverbs 8:15-16\index{20 Proverbs:ch.   8, v.  15-16, pg. ``f30!}; Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}, 2, 4\index{45 Romans:ch.  13, v.   1, 2, 4, pg. ``f30!}.} in the managing whereof, as they ought especially to maintain piety, justice\index{Civil magistrate:to maintain justice ``f10!}, and peace, according\index{Civil magistrate:to maintain peace:in society ``f10!} to the wholesome\index{Commandments of men:wholesome ``f10!} laws of each commonwealth\index{commonwealth ``f10!};\footnote{\ Psalm 2:10-12\index{19 Psalms:ch.   2, v.  10-12, pg. ``f30!}; 1 Timothy 2:2\index{54 1 Timothy:ch.   2, v.   2, pg. ``f30!}; Psalm 82:3-4\index{19 Psalms:ch.  82, v.   3-4, pg. ``f30!}; 2 Samuel 23:3\index{10 2 Samuel:ch.  23, v.   3, pg. ``f30!}; 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}.} so for that end, they may lawfully now\index{war, just, lawful ``f10!}, under the New Testament\index{Scripture:New Testament ``f10!}, wage war, upon just and necessary occasion.\footnote{\ Luke 3:14\index{42 Luke:ch.   3, v.  14, pg. ``f30!}; Romans 13:4\index{45 Romans:ch.  13, v.   4, pg. ``f30!}; Matthew 8:9-10\index{40 Matthew:ch.   8, v.   9-10, pg. ``f30!}; Acts 10:1-2\index{44 Acts:ch.  10, v.   1-2, pg. ``f30!}; Revelation 17:14, 16\index{66 Revelation:ch.  17, v.  14, 16, pg. ``f30!}.} 
\bigskip
23.3\ \ The civil magistrate may not assume to himself the administration\index{Civil magistrate:not to administer word or sacraments ``f10!} of the Word and sacraments\index{Sacraments:holy:ordinances ``f10!}, or the power\index{Church:power delegated by Christ ``f10!} of the keys\index{Kingdom:keys of ``f10!} of the kingdom of heaven\index{Kingdom:of heaven ``f10!}:\footnote{\ 2 Chronicles 26:18\index{14 2 Chronicles:ch.  26, v.  18, pg. ``f30!} with Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!} and Matthew 16:19\index{40 Matthew:ch.  16, v.  19, pg. ``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  12, v.  28-29, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-12\index{49 Ephesians:ch.   4, v.  11-12, pg. ``f30!}; 1 Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.   1, pg. ``f30!}{}-2\index{46 1 Corinthians:ch.   4, v.   1-2, pg. ``f30!}; Romans 10:1\index{45 Romans:ch.  10, v.   1, pg. ``f30!}5\index{45 Romans:ch.  10, v.  15, pg. ``f30!}; Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}.} yet he has authority, and it is his duty, to take order, that unity\index{Civil magistrate:to ensure unity and peace in Church ``f10!} and peace be\index{Civil magistrate:to maintain peace:in Church ``f10!} preserved in the Church, that\index{Church:institution ``f10!} the truth of God be\index{Truth:to be kept pure by civil magistrate ``f10!} kept pure and entire; that\index{Civil magistrate:to keep truth of God pure ``f10!} all blasphemies\index{blasphemy ``f10!} and heresies\index{Heresy:heresies ``f10!} be suppressed\index{Civil magistrate:to suppress blasphemies and heresies ``f10!}; all corruptions and\index{Worship:false:corruptions in ``f10!} abuses\index{Worship:not to be:abused ``f10!} in worship and\index{Worship:abuses ``f10!} discipline\index{Church discipline:discipline ``f10!} prevented\index{Civil magistrate:to keep worship of God pure ``f10!} or reformed; and all the ordinances of God duly\index{Civil magistrate:to ensure ordinances of God administrated ``f10!} settled, administrated, and observed.\footnote{\ Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; Psalm 122:9\index{19 Psalms:ch. 122, v.   9, pg. ``f30!}; Ezra 7:23, 25-28\index{15 Ezra:ch.   7, v.  23, 25-28, pg. ``f30!}; Leviticus 24:16\index{03 Leviticus:ch.  24, v.  16, pg. ``f30!}; Deuteronomy 13:5-6, 12\index{05 Deuteronomy:ch.  13, v.   5-6, 12, pg. ``f30!}; 1 Kings 18:4\index{11 1 Kings:ch.  18, v.   4, pg. ``f30!}; 1 Chronicles 13:1-9\index{13 1 Chronicles:ch.  13, v.   1-9, pg. ``f30!}; 2 Kings 23:1-26\index{12 2 Kings:ch.  23, v.   1-26, pg. ``f30!}; 2 Chronicles 34:33\index{14 2 Chronicles:ch.  34, v.  33, pg. ``f30!}; 2 Chronicles 15:12-13\index{14 2 Chronicles:ch.  15, v.  12-13, pg. ``f30!}.} For the better effecting whereof, he has power\index{Civil magistrate:to call synods to consult and advise ``f10!} to call synods\index{Church government:synods ``f10!}, to be present at them, and to provide that whatsoever is transacted in them be according to the mind of God.\footnote{\ 2 Chronicles 19:8-11\index{14 2 Chronicles:ch.  19, v.   8-11, pg. ``f30!}; 2 Chronicles 29; 2 Chronicles 30; Matthew 2:4-5\index{40 Matthew:ch.   2, v.   4-5, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Clarifications}}\textcolor{black}{).}}   
\bigskip
23.4\ \ It is the duty of people\index{Prayer:for magistrates ``f10!} to pray for magistrates\index{Civil magistrate:pray for ``f10!},\footnote{\ 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}.} to honour their\index{Civil magistrate:to be honoured ``f10!} persons,\footnote{\ 1 Peter 2:17\index{60 1 Peter:ch.   2, v.  17, pg. ``f30!}.} to pay them tribute\index{Civil magistrate:to be paid tribute and dues ``f10!} or other dues,\footnote{\ Romans 13:6-7\index{45 Romans:ch.  13, v.   6-7, pg. ``f30!}.} to obey their\index{Commandments of men:lawful to be obeyed ``f10!} lawful commands, and to be subject to their authority, for conscience\index{Conscience:for sake of ``f10!}{}' sake.\footnote{\ Romans 13:5\index{45 Romans:ch.  13, v.   5, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}.} Infidelity\index{Civil magistrate:legal authority even if:infidels ``f10!}, or difference in religion, does\index{Civil magistrate:legal authority even if:different religion ``f10!} not make void the magistrates' just and legal\index{Civil magistrate:legal authority even if:unbelievers ``f10!} authority, nor free the people from their due obedience to them:\footnote{\ 1 Peter 2:13\index{60 1 Peter:ch.   2, v.  13, pg. ``f30!}{}-14\index{60 1 Peter:ch.   2, v.  13-14, pg. ``f30!}, 16\index{60 1 Peter:ch.   2, v.  13-14, 16, pg. ``f30!}.} from which ecclesiastical\index{Church officers:to obey legal laws of civil authorities ``f10!} persons are not exempted,\footnote{\ Romans 13:1\index{45 Romans:ch.  13, v.   1, pg. ``f30!}; 1 Kings 2:35\index{11 1 Kings:ch.   2, v.  35, pg. ``f30!}; Acts 25:9-11\index{44 Acts:ch.  25, v.   9-11, pg. ``f30!}; 2 Peter 2:1, 10-11\index{61 2 Peter:ch.   2, v.   1, 10-11, pg. ``f30!}; Jude 8-11\index{65 Jude:v. 8-11, pg. ``f30!}.} much less has the Pope\index{Pope:of Rome ``f10!} any power and jurisdiction\index{Civil magistrate:pope has no jurisdiction over ``f10!} over them in their dominions\index{Pope:no jurisdiction over people in their dominions ``f10!}, or over any of their people; and, least of all, to deprive them of their dominions, or lives, if he shall judge the\index{Jesus:judge:world ``f10!}m\index{Pope:not to judge magistrates as heretics ``f10!} to be heretics\index{heretics ``f10!}, or upon any other pretence\index{pretence ``f10!} whatsoever.\footnote{\ 2 Thessalonians 2:4\index{53 2 Thessalonians:ch.   2, v.   4, pg. ``f30!}; Revelation 13:15-17\index{66 Revelation:ch.  13, v.  15-17, pg. ``f30!}.}  

\section{Chapter 24 -- Of Marriage and Divorce} \index{Marriage:Biblical definition ``f10!}\index{Divorce:divorce ``f10!}24.1\ \ Marriage\index{Marriage:Biblical definition ``f10!} is to be between one man and one woman: neither is it lawful for any\index{Polygamy:unlawful ``f10!} man to have more than one wife\index{Marriage:one spouse ``f10!}, nor for any woman to have more than one husband; at the same time.\footnote{\ Genesis 2:24\index{01 Genesis:ch.   2, v.  24, pg. ``f30!}; Matthew 19:5\index{40 Matthew:ch.  19, v.   5, pg. ``f30!}{}-6\index{40 Matthew:ch.  19, v.   5-6, pg. ``f30!}; Proverbs 2:1\index{20 Proverbs:ch.   2, v.   1, pg. ``f30!}7\index{20 Proverbs:ch.   2, v.  17, pg. ``f30!}.}   
\bigskip
24.2\ \ Marriage\index{Marriage:Biblical definition ``f10!} was ordained for the mutual help\index{Marriage:mutual help ``f10!} of husband and wife,\footnote{\ Genesis 2:18\index{01 Genesis:ch.   2, v.  18, pg. ``f30!}.} for the increase of mankind with\index{Marriage:to increase mankind ``f10!} a legitimate\index{Marriage:to produce legitimate issue (offspring) ``f10!} issue, and of the Church with\index{Church:home of believers ``f10!} an holy seed\index{Marriage:produce holy seed for Church ``f10!};\footnote{\ Malachi 2:15\index{39 Malachi:ch.   2, v.  15, pg. ``f30!}.} and for preventing of uncleanness\index{Marriage:to prevent uncleanness ``f10!}.\footnote{\ 1 Corinthians 7:2, 9\index{46 1 Corinthians:ch.   7, v.   2, 9, pg. ``f30!}.}   
\bigskip
24.3\ \ It is lawful for all\index{Marriage:lawful:for all ``f10!} sorts of people to marry, who are able with judgment to\index{Judgement:temporal ``f10!} give their consent\index{Marriage:consent of parties ``f10!}.\footnote{\ Hebrews 13:4\index{58 Hebrews:ch.  13, v.   4, pg. ``f30!}; 1 Timothy 4:3\index{54 1 Timothy:ch.   4, v.   3, pg. ``f30!}; 1 Corinthians 7:36-38\index{46 1 Corinthians:ch.   7, v.  36-38, pg. ``f30!}; Genesis 24:57-58\index{01 Genesis:ch.  24, v.  57-58, pg. ``f30!}.} Yet is it the duty of Christians to marry only\index{Marriage:Christians:only in the Lord ``f10!} in the Lord:\footnote{\ 1 Corinthians 7:39\index{46 1 Corinthians:ch.   7, v.  39, pg. ``f30!}.} and therefore such as profess the true\index{Profession:of true religion ``f10!} reformed religion should not marry with\index{Marriage:Christians:not with infidels or papists ``f10!} infidels, papists\index{papists ``f10!}, or other idolaters\index{Idolatry:idolatry ``f10!}: neither should such as are godly be unequally yoked\index{Marriage:not unequally yoked ``f10!}, by marrying with such as are notoriously\index{Marriage:not with notoriously wicked ``f10!} wicked in their life, or maintain damnable heresies\index{Heresy:heresies ``f10!}.\footnote{\ Genesis 34:14\index{01 Genesis:ch.  34, v.  14, pg. ``f30!}; Exodus 34:1\index{02 Exodus:ch.  34, v.   1, pg. ``f30!}6\index{02 Exodus:ch.  34, v.  16, pg. ``f30!}; Deuteronomy 7:3-4\index{05 Deuteronomy:ch.   7, v.   3-4, pg. ``f30!}; I Kings 11:4; Nehemiah 13:25\index{16 Nehemiah:ch.  13, v.  25, pg. ``f30!}{}-26, 27\index{16 Nehemiah:ch.  13, v.  25-26, 27, pg. ``f30!}; Malachi 2:11-12\index{39 Malachi:ch.   2, v.  11-12, pg. ``f30!}; 2 Corinthians 6:14\index{47 2 Corinthians:ch.   6, v.  14, pg. ``f30!}.}    
\bigskip
24.4\ \ Marriage\index{Marriage:Biblical definition ``f10!} ought not to be within the degrees of consanguinity\index{Consanguinity:consanguinity ``f10!} or affinity forbidden\index{Consanguinity:forbidden affinity ``f10!} by the Word;\footnote{\ Leviticus 18; 1 Corinthians 5:1\index{46 1 Corinthians:ch.   5, v.   1, pg. ``f30!}; Amos 2:7\index{30 Amos:ch.   2, v.   7, pg. ``f30!}.} nor can such incestuous marriages\index{Marriage:unlawful ``f10!} ever be made lawful by any\index{Consanguinity:incestuous marriage not lawful ``f10!} law of man or consent of parties\index{Marriage:consent of parties ``f10!}, so as those persons may live together as man and wife\index{Marriage:live together as man and wife ``f10!}.\footnote{\ Mark 6:18\index{41 Mark:ch.   6, v.  18, pg. ``f30!}; Leviticus 18:24-28\index{03 Leviticus:ch.  18, v.  24-28, pg. ``f30!}.} The man may not marry any\index{Marriage:not near kindred of wife ``f10!} of his wife's kindred\index{Consanguinity:near kindred ``f10!} nearer in blood\index{Consanguinity:nearness of blood ``f10!} than he may of his own; nor the woman of her husband's kindred nearer in blood than of her own.\footnote{\ Leviticus 20:19-21\index{03 Leviticus:ch.  20, v.  19-21, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).} }   
\bigskip
24.5\ \ Adultery\index{Sexual immorality:adultery ``f10!} or fornication\index{Sexual immorality:fornication ``f10!} committed after a contract\index{Contracts:marriage ``f10!}, being detected\index{Marriage:valid reasons for ending ``f10!} before marriage, gives just occasion to the innocent party to dissolve that contract.\footnote{\ Matthew 1:18-20\index{40 Matthew:ch.   1, v.  18-20, pg. ``f30!}.} In the case of adultery after marriage, it is lawful for the innocent party to sue out a divorce\index{Divorce:divorce ``f10!}:\footnote{\ Matthew 5:31-32\index{40 Matthew:ch.   5, v.  31-32, pg. ``f30!}.} and, after the divorce, to marry another\index{Marriage:remarriage permitted for innocent party after divorce ``f10!}, as if the offending\index{Divorce:offending treated as dead ``f10!} party were dead\index{Death:divorced marriage partner viewed as dead ``f10!}.\footnote{\ Matthew 19:9\index{40 Matthew:ch.  19, v.   9, pg. ``f30!}; Romans 7:2-3\index{45 Romans:ch.   7, v.   2-3, pg. ``f30!}.}   
\bigskip
24.6\ \ Although the corruption of man\index{Sin:original:corrupt nature:whole ``f10!} be such as is apt to study arguments unduly to put asunder\index{Marriage:not to be broken, except if adultery ``f10!} those whom God has joined together in marriage: yet nothing but adultery\index{Sexual immorality:adultery ``f10!}, or such wilful desertion\index{Marriage:desertion ``f10!} as can no way be remedied by the Church or\index{Church:institution ``f10!} civil magistrate, is\index{Civil magistrate:role in divorce cases ``f10!} cause sufficient of dissolving\index{Marriage:valid reasons for ending ``f10!} the bond of marriage:\footnote{\ Matthew 19:8-9\index{40 Matthew:ch.  19, v.   8-9, pg. ``f30!}; 1 Corinthians 7:15\index{46 1 Corinthians:ch.   7, v.  15, pg. ``f30!}; Matthew 19:6\index{40 Matthew:ch.  19, v.   6, pg. ``f30!}.} wherein, a public and orderly course of proceeding is to be observed; and the persons concerned in it not left to their own wills and\index{Causation:will of creatures:not to be exercised to dissolve marriage ``f10!} discretion, in their own case.\footnote{\ Deuteronomy 24:1-4\index{05 Deuteronomy:ch.  24, v.   1-4, pg. ``f30!}.}  

\section{Chapter 25 -- Of the Church} \index{Church:institution ``f10!}25.1\ \ The catholic\index{Church:catholic ``f10!} or universal Church\index{Church:universal ``f10!} which is invisible, consists\index{Church:invisible ``f10!} of the whole number of the elect\index{Election:elect of God ``f10!}, that have been, are, or shall be gathered into one, under Christ the Head\index{Church:Christ as head ``f10!} thereof; and is the spouse\index{Church:Christ's spouse ``f10!}, the body, the fulness of Him\index{Jesus:fulness of all things ``f10!} that fills\index{Jesus:fills all in all ``f10!} all in all.\footnote{\ Ephesians 1:1\index{49 Ephesians:ch.   1, v.   1, pg. ``f30!}0, 22-23\index{49 Ephesians:ch.   1, v.  10, 22-23, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}3\index{49 Ephesians:ch.   5, v.  23, pg. ``f30!}, 27, 32\index{49 Ephesians:ch.   5, v.  23, 27, 32, pg. ``f30!}; Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg. ``f30!}.}   
\bigskip
25.2\ \ The visible Church\index{Church:visible ``f10!}, which is also catholic\index{Church:catholic ``f10!} or universal under\index{Church:universal ``f10!} the Gospel\index{Gospel:gospel ``f10!} (not confined to one nation as\index{Gospel:to all nations ``f10!} before under the law), consists of all those throughout the world that profess the true\index{Profession:of true religion ``f10!} religion;\footnote{\ 1 Corinthians 1:2\index{46 1 Corinthians:ch.   1, v.   2, pg. ``f30!}; 1 Corinthians 12:12-13\index{46 1 Corinthians:ch.  12, v.  12-13, pg. ``f30!}; Psalm 2:8\index{19 Psalms:ch.   2, v.   8, pg. ``f30!}; Revelation 7:9\index{66 Revelation:ch.   7, v.   9, pg. ``f30!}; Romans 15:9, 10-12\index{45 Romans:ch.  15, v.   9, 10-12, pg. ``f30!}.} and of their children:\footnote{\ 1 Corinthians 7:14\index{46 1 Corinthians:ch.   7, v.  14, pg. ``f30!}; Acts 2:39\index{44 Acts:ch.   2, v.  39, pg. ``f30!}; Ezekiel 16:20-21\index{26 Ezekiel:ch.  16, v.  20-21, pg. ``f30!}; Romans 11:16\index{45 Romans:ch.  11, v.  16, pg. ``f30!}; Genesis 3:15\index{01 Genesis:ch.   3, v.  15, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}.} and is the kingdom of the\index{Kingdom:of Christ ``f10!} Lord Jesus Christ,\footnote{\ Matthew 13:47\index{40 Matthew:ch.  13, v.  47, pg. ``f30!}; Isaiah 9:7\index{23 Isaiah:ch.   9, v.   7, pg. ``f30!}.} the house and\index{Church:home of believers ``f10!} family of\index{Church:family of God ``f10!} God,\footnote{\ Ephesians 2:1\index{49 Ephesians:ch.   2, v.   1, pg. ``f30!}9\index{49 Ephesians:ch.   2, v.  19, pg. ``f30!}; Ephesians 3:15\index{49 Ephesians:ch.   3, v.  15, pg. ``f30!}.} out of which there is no ordinary\index{Church:ordinary means of salvation ``f10!} possibility of salvation\index{Salvation:belief in Jesus by faith ``f10!}.\footnote{\ Acts 2:47\index{44 Acts:ch.   2, v.  47, pg. ``f30!}.}   
\bigskip
25.3\ \ Unto this catholic\index{Church:catholic ``f10!} visible Church\index{Church:visible ``f10!} Christ has given the ministry\index{Church:given the ministry of God ``f10!}, oracles\index{Church:endowed with oracles of God ``f10!}, and ordinances of God, for\index{Church:furnished with:ordinances of God ``f10!} the gathering\index{Church:universal ``f10!} and perfecting of\index{Church:perfecting of saints ``f10!} the saints\index{Believers:saints ``f10!}, in this life, to the end of the world: and does by His own presence and Spirit\index{Holy Spirit:works in the Church ``f10!}, according to His promise\index{Church:promised presence of God ``f10!}, make them effectual thereunto\index{Worship:ordinances:made effectual for salvation ``f10!}.\footnote{\ 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!}; Ephesians 4:11\index{49 Ephesians:ch.   4, v.  11, pg. ``f30!}{}-13\index{49 Ephesians:ch.   4, v.  11-13, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}; Isaiah 59:2\index{23 Isaiah:ch.  59, v.   2, pg. ``f30!}1\index{23 Isaiah:ch.  59, v.  21, pg. ``f30!}.}   
\bigskip
25.4\ \ This catholic\index{Church:catholic ``f10!} Church has been sometimes more, sometimes less visible.\footnote{\ Romans 11:3-4\index{45 Romans:ch.  11, v.   3-4, pg. ``f30!}; Revelation 12:6, 14\index{66 Revelation:ch.  12, v.   6, 14, pg. ``f30!}.} And particular Churches, which\index{Church:mix of truth and error ``f10!} are members thereof\index{Church members ``f10!}, are more or less pure\index{Church:true:marks ``f10!}, according as the doctrine of the\index{Doctrine:Gospel ``f10!} Gospel\index{Gospel:gospel ``f10!} is taught and\index{Gospel:to be taught and embraced ``f10!} embraced, ordinances administered, and\index{Church:true:ordinances administered ``f10!} public worship\index{Worship:public ``f10!} performed more or less purely in them.\footnote{\ Revelation 2; Revelation 3; 1 Corinthians 5:6-7\index{46 1 Corinthians:ch.   5, v.   6-7, pg. ``f30!}.}   
\bigskip
25.5\ \ The purest Churches under\index{Church:mix of truth and error ``f10!} heaven are subject both to mixture and error:\footnote{\ 1 Corinthians 13:12\index{46 1 Corinthians:ch.  13, v.  12, pg. ``f30!}; Revelation 2; Revelation 3; Matthew 13:24-30, 47\index{40 Matthew:ch.  13, v.  24-30, 47, pg. ``f30!}.} and some have so degenerated\index{Church:degenerated churches ``f10!}, as to become no Churches of\index{Church:false, synagogues of Satan ``f10!} Christ, but synagogues of Satan.\footnote{\ Revelation 18:2\index{66 Revelation:ch.  18, v.   2, pg. ``f30!}; Romans 11:18-22\index{45 Romans:ch.  11, v.  18-22, pg. ``f30!}.} Nevertheless, there shall be always a Church\index{Church:indestructible ``f10!} on earth, to worship God\index{Worship:always be a Church to worship ``f10!} according to His will.\footnote{\ Matthew 16:18\index{40 Matthew:ch.  16, v.  18, pg. ``f30!}; Psalm 72:17\index{19 Psalms:ch.  72, v.  17, pg. ``f30!}; Psalm 102:28\index{19 Psalms:ch. 102, v.  28, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}   
\bigskip
25.6\ \ There is no other head of the\index{Church:Christ as head ``f10!} Church, but the Lord Jesus Christ;\footnote{\ Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg. ``f30!}; Ephesians 1:22\index{49 Ephesians:ch.   1, v.  22, pg. ``f30!}.} nor can the Pope\index{Pope:of Rome ``f10!} of Rome, in any sense, be head thereof\index{Pope:not head of the Church ``f10!}; but is that Antichrist\index{Antichrist (Pope) ``f10!}, that man of sin, and son of perdition\index{Pope:son of perdition ``f10!}, that exalts\index{Pope:exalts self ``f10!} himself, in the Church, against\index{Church:institution ``f10!} Christ and all that is called God.\footnote{\ Matthew 23:8-10\index{40 Matthew:ch.  23, v.   8-10, pg. ``f30!}; 2 Thessalonians 2:3-4, 8-9\index{53 2 Thessalonians:ch.   2, v.   3-4, 8-9, pg. ``f30!}; Revelation 13:6\index{66 Revelation:ch.  13, v.   6, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).}}  

\section{Chapter 26 -- Of the Communion of the Saints} \index{Communion:of saints ``f10!}\index{Believers:saints ``f10!}26.1\ \ All saints\index{Believers:saints ``f10!}, that are united to Jesus\index{Believers:union with Christ:mystical ``f10!} Christ their Head\index{Church:Christ as head ``f10!} by His Spirit and by faith\index{Faith:worked in believers by the Holy Spirit ``f10!}, have fellowship with\index{Fellowship:with Christ ``f10!} Him in His grace\index{Grace:fellowship with Jesus in ``f10!}, sufferings, death\index{Jesus:death:sufferings and death ``f10!}, resurrection, and glory:\index{Resurrection:believers:into glory:with Christ ``f10!}\footnote{\ John 1:3\index{43 John:ch.   1, v.   3, pg. ``f30!}; Ephesians 3:16\index{49 Ephesians:ch.   3, v.  16, pg. ``f30!}{}-19\index{49 Ephesians:ch.   3, v.  16-19, pg. ``f30!}; John 1:1\index{43 John:ch.   1, v.   1, pg. ``f30!}6\index{43 John:ch.   1, v.  16, pg. ``f30!}; Ephesians 2:5\index{49 Ephesians:ch.   2, v.   5, pg. ``f30!}{}-6\index{49 Ephesians:ch.   2, v.   5-6, pg. ``f30!}; Philippians 3:10\index{50 Philippians:ch.   3, v.  10, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}{}-6\index{45 Romans:ch.   6, v.   5-6, pg. ``f30!}; 2 Timothy 2:12\index{55 2 Timothy:ch.   2, v.  12, pg. ``f30!}.} and, being united to one\index{Believers:united to one another ``f10!} another in love, they\index{Love for others:united ``f10!} have communion\index{Communion:of saints ``f10!} in each other's gifts and graces,\index{Love for others:acknowledge gifts and graces of others ``f10!}\footnote{\ Ephesians 4:15-16\index{49 Ephesians:ch.   4, v.  15-16, pg. ``f30!}; 1 Corinthians 12:7\index{46 1 Corinthians:ch.  12, v.   7, pg. ``f30!}; 1 Corinthians 3:2\index{46 1 Corinthians:ch.   3, v.   2, pg. ``f30!}1-23\index{46 1 Corinthians:ch.   3, v.  21-23, pg. ``f30!}; Colossians 2:19\index{51 Colossians:ch.   2, v.  19, pg. ``f30!}.} and are obliged to the performance of such duties\index{God's law:commandments:duties:to others ``f10!}, public and private, as do conduce to their mutual good\index{Good works:for mutual good ``f10!}, both in the inward\index{Love for others:inward thoughts further others' good ``f10!} and outward man\index{Love for others:for saints ``f10!}.\footnote{\ 1 Thessalonians 5:11\index{52 1 Thessalonians:ch.   5, v.  11, pg. ``f30!}, 14\index{52 1 Thessalonians:ch.   5, v.  11, 14, pg. ``f30!}; Romans 1:11-12, 14\index{45 Romans:ch.   1, v.  11-12, 14, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}6\index{62 1 John:ch.   3, v.  16, pg. ``f30!}{}-18\index{62 1 John:ch.   3, v.  16-18, pg. ``f30!}; Galatians 6:10\index{48 Galatians:ch.   6, v.  10, pg. ``f30!}.}   
\bigskip
26.2\ \ Saints\index{Believers:saints ``f10!} by profession are\index{Faith:profession of ``f10!} bound to maintain a holy\index{Worship:holy fellowship to be maintained ``f10!} fellowship and\index{Fellowship:with fellow believers ``f10!} communion in the\index{Communion:in worship ``f10!} worship of God; and in performing such other spiritual services\index{Good works:spiritual services ``f10!} as tend to their mutual edification\index{Love for others:mutual:edification ``f10!};\footnote{\ Hebrews 10:24-25\index{58 Hebrews:ch.  10, v.  24-25, pg. ``f30!}; Acts 2:42\index{44 Acts:ch.   2, v.  42, pg. ``f30!}, 46\index{44 Acts:ch.   2, v.  42, 46, pg. ``f30!}; Isaiah 2:3\index{23 Isaiah:ch.   2, v.   3, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}.} as also in relieving\index{Love for others:mutual:relieving in outward things ``f10!} each other in outward things\index{Love for others:for saints ``f10!}, according to their several abilities, and necessities. Which communion, as\index{Communion:with God ``f10!} God offers\index{Church:communion with all who call upon Jesus ``f10!} opportunity\index{Communion:of saints ``f10!}, is to be extended unto all those who, in every place, call upon the name of the Lord Jesus.\footnote{\ Acts 2:44-45\index{44 Acts:ch.   2, v.  44-45, pg. ``f30!}; 1 John 3:1\index{62 1 John:ch.   3, v.   1, pg. ``f30!}7\index{62 1 John:ch.   3, v.  17, pg. ``f30!}; 2 Corinthians 8; 2 Corinthians 9; Acts 11:29-30\index{44 Acts:ch.  11, v.  29-30, pg. ``f30!}.}   
\bigskip
26.3\ \ This communion which\index{Communion:with Christ ``f10!} the saints\index{Believers:saints ``f10!} have with Christ, does not make them, in any wise, partakers\index{Believers:union with Christ:not partakers of Godhead ``f10!} of the substance of His Godhead\index{God's attributes:Godhead ``f10!}; or to be equal with Christ\index{Communion:with Christ, does not make believers equal to him ``f10!}, in any respect: either of which to affirm is impious and blasphemous\index{blasphemy ``f10!}.\footnote{\ Colossians 1:18\index{51 Colossians:ch.   1, v.  18, pg. ``f30!}{}-19\index{51 Colossians:ch.   1, v.  18-19, pg. ``f30!}; 1 Corinthians 8:6\index{46 1 Corinthians:ch.   8, v.   6, pg. ``f30!}; Isaiah 42:8\index{23 Isaiah:ch.  42, v.   8, pg. ``f30!}; 1 Timothy 6:1\index{54 1 Timothy:ch.   6, v.   1, pg. ``f30!}5\index{54 1 Timothy:ch.   6, v.  15, pg. ``f30!}{}-16\index{54 1 Timothy:ch.   6, v.  15-16, pg. ``f30!}; Psalm 45:7\index{19 Psalms:ch.  45, v.   7, pg. ``f30!} with Hebrews 1:8\index{58 Hebrews:ch.   1, v.   8, pg. ``f30!}{}-9\index{58 Hebrews:ch.   1, v.   8-9, pg. ``f30!}.} Nor does their communion one\index{Communion:of saints ``f10!} with another, as saints, take away, or infringe\index{property, rights not to be infringed ``f10!} the title or propriety which each man has in his goods\index{Theft:infringe on property of others ``f10!} and possessions.\footnote{\ Exodus 20:15\index{02 Exodus:ch.  20, v.  15, pg. ``f30!}; Ephesians 4:28\index{49 Ephesians:ch.   4, v.  28, pg. ``f30!}; Acts 5:4\index{44 Acts:ch.   5, v.   4, pg. ``f30!}.}  

\section{Chapter 27 -- Of the Sacraments} \index{Sacraments:holy:ordinances ``f10!}27.1\ \ Sacraments\index{Sacraments:holy:ordinances ``f10!} are holy\index{Sacraments:signs and seals:holy ``f10!} signs\index{Sacraments:holy:signs ``f10!} and seals\index{Sacraments:covenant seals ``f10!} of the covenant of grace\index{Covenant:grace:grace ``f10!},\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 10\index{01 Genesis:ch.  17, v.   7, 10, pg. ``f30!}.} immediately instituted by God\index{Sacraments:instituted by:God ``f10!},\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}.} to represent Christ\index{Sacraments:represent Christ ``f10!} and His benefits; and to\index{Sacraments:benefits:worthy believers receive ``f10!} confirm\index{Sacraments:confirm union with Christ ``f10!} our interest in Him\index{Believers:interest (share) in Christ ``f10!};\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch.  11, v.  25, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  25-26, pg. ``f30!}; Galatians 3:17\index{48 Galatians:ch.   3, v.  17, pg. ``f30!}.} as also, to put a visible\index{Sacraments:signs and seals:visible difference ``f10!} difference between those that belong unto the Church, and\index{Church:institution ``f10!} the rest of the world;\footnote{\ Romans 15:8\index{45 Romans:ch.  15, v.   8, pg. ``f30!}; Exodus 12:48\index{02 Exodus:ch.  12, v.  48, pg. ``f30!}; Genesis 34:14\index{01 Genesis:ch.  34, v.  14, pg. ``f30!}.} and solemnly to\index{Sacraments:solemn engagement to God's service ``f10!} engage\index{Sacraments:engage participants to service of God ``f10!} them to the service of\index{Love toward God:in service ``f10!} God in Christ, according to His Word.\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6, v.   3-4, pg. ``f30!}; 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}, 21\index{46 1 Corinthians:ch.  10, v.  16, 21, pg. ``f30!}.}   
\bigskip
27.2\ \ There is in every sacrament\index{Sacraments:holy:ordinances ``f10!} a spiritual relation\index{Sacraments:spiritual relation between sign and signified ``f10!}, or sacramental\index{Lord's Supper:elements:union between sign and signified ``f10!} union, between\index{Sacraments:union between sign and signified ``f10!} the sign and the thing signified: whence it comes to pass, that the names and effects of the one are attributed to the other.\footnote{\ Genesis 17:1\index{01 Genesis:ch.  17, v.   1, pg. ``f30!}0\index{01 Genesis:ch.  17, v.  10, pg. ``f30!}; Matthew 26:27-28\index{40 Matthew:ch.  26, v.  27-28, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.}   
\bigskip
27.3\ \ The grace which\index{Grace:exhibited in sacraments ``f10!} is exhibited in\index{Sacraments:grace:exhibited ``f10!} or by the sacraments\index{Sacraments:holy:ordinances ``f10!} rightly used\index{Sacraments:grace:conferred ``f10!}, is not conferred by any\index{Sacraments:grace:not conferred by elements ``f10!} power in them; neither does the efficacy of a\index{Sacraments:spiritual efficacy ``f10!} sacrament depend upon the piety or\index{Lord's Supper:efficacy, not dependent on piety or intention of person administering ``f10!} intention of him that does administer it:\footnote{\ Romans 2:28-29\index{45 Romans:ch.   2, v.  28-29, pg. ``f30!}; 1 Peter 3:21\index{60 1 Peter:ch.   3, v.  21, pg. ``f30!}.} but upon the work of the Spirit,\index{Holy Spirit:Spirit and the word ``f10!}\footnote{\ Matthew 3:11\index{40 Matthew:ch.   3, v.  11, pg. ``f30!}; 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.} and the word of institution\index{Sacraments:instituted by:Christ ``f10!}, which contains, together with a precept authorizing the use thereof, a promise of benefit to\index{Sacraments:benefits:worthy believers receive ``f10!} worthy receivers.\footnote{\ Matthew 26:27-28\index{40 Matthew:ch.  26, v.  27-28, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}   
\bigskip
27.4\ \ There are only two sacraments\index{Sacraments:holy:ordinances ``f10!} ordained by\index{Sacraments:ordained by Jesus ``f10!} Christ our Lord in the Gospel\index{Gospel:gospel ``f10!}; that is to say, Baptism and\index{Baptism:sacrament ``f10!} the Supper\index{Lord's Supper:sacrament ``f10!} of the Lord: neither of which may be dispensed\index{Sacraments:to be:dispensed by minister of the Word ``f10!} by any but by a minister\index{Church officers:minsters ``f10!} of the Word lawfully\index{Preachers:to be:lawfully ordained ``f10!} ordained.\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}, 23\index{46 1 Corinthians:ch.  11, v.  20, 23, pg. ``f30!}; 1 Corinthians 4:1\index{46 1 Corinthians:ch.   4, v.   1, pg. ``f30!}; Hebrews 5:4\index{58 Hebrews:ch.   5, v.   4, pg. ``f30!}.}   
\bigskip
27.5\ \ The sacraments\index{Sacraments:holy:ordinances ``f10!} of the Old Testament\index{Scripture:Old Testament ``f10!}, in regard to the spiritual things\index{Sacraments:spiritual things signified, same in both Testaments ``f10!} thereby signified and exhibited, were, for substance\index{Sacraments:OT and NT same substance ``f10!}, the same with those of the New.\footnote{\ 1 Corinthians 10:1-4\index{46 1 Corinthians:ch.  10, v.   1-4, pg. ``f30!}.}  

\section{Chapter 28 -- Of Baptism} \index{Baptism:sacrament ``f10!}28.1\ \ Baptism is a\index{Baptism:sacrament ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} of the New Testament\index{Scripture:New Testament ``f10!}, ordained by\index{Sacraments:ordained by Jesus ``f10!} Jesus Christ,\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}.} not only for the solemn admission\index{Baptism:admission into visible Church ``f10!} of the party baptized into the visible Church\index{Church:visible ``f10!};\footnote{\ 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.} but also, to be unto him a sign\index{Baptism:sign of salvation ``f10!} and seal of\index{Baptism:seal of salvationof ``f10!} the covenant of grace\index{Covenant:grace:grace ``f10!},\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!} with Colossians 2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg. ``f30!}.} of his ingrafting\index{Believers:ingrafted into Christ ``f10!} into Christ,\footnote{\ Galatians 3:27\index{48 Galatians:ch.   3, v.  27, pg. ``f30!}; Romans 6:5\index{45 Romans:ch.   6, v.   5, pg. ``f30!}.} of regeneration\index{Regeneration:regeneration ``f10!},\footnote{\ Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.} of remission\index{Redemption:remission of sins ``f10!} of sins,\footnote{\ Mark 1:4\index{41 Mark:ch.   1, v.   4, pg. ``f30!}.} and of his giving up unto God through Jesus Christ, to walk in the newness\index{Baptism:walk in newness of life ``f10!} of life.\footnote{\ Romans 6:3-4\index{45 Romans:ch.   6, v.   3-4, pg. ``f30!}.} Which sacrament is, by Christ's own appointment, to be\index{Baptism:appointed by Christ for Church ``f10!} continued in His Church\index{Church:Christ's ``f10!} until the end of the world.\footnote{\ Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}   
\bigskip
28.2\ \ The outward element to be used in this sacrament\index{Sacraments:holy:ordinances ``f10!} is water\index{Baptism:to use water ``f10!}, wherewith the party is to be baptized\index{Baptism:in name of Trinity ``f10!}, in the name of the Father\index{Father:God the ``f10!}, and of the Son\index{Jesus:Son of God ``f10!}, and of the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, by a minister\index{Church officers:minsters ``f10!} of the Gospel\index{Gospel:gospel ``f10!}, lawfully call\index{Civil magistrate:to call synods to consult and advise ``f10!}ed\index{Preachers:to be:lawfully called ``f10!} thereunto.\footnote{\ Matthew 3:11\index{40 Matthew:ch.   3, v.  11, pg. ``f30!}; John 1:3\index{43 John:ch.   1, v.   3, pg. ``f30!}3\index{43 John:ch.   1, v.  33, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  19-20, pg. ``f30!}.}   
\bigskip
28.3\ \ Dipping\index{Baptism:modes:dipping ``f10!} of the person into the water is\index{Baptism:to use water ``f10!} not necessary; but Baptism is rightly administered\index{Baptism:modes:valid ``f10!} by pouring\index{Baptism:modes:pouring ``f10!} or sprinkling\index{Baptism:modes:sprinkling ``f10!} water upon the person.\footnote{\ Hebrews 9:10, 19-22\index{58 Hebrews:ch.   9, v.  10, 19-22, pg. ``f30!}; Acts 2:41\index{44 Acts:ch.   2, v.  41, pg. ``f30!}; Acts 16:33\index{44 Acts:ch.  16, v.  33, pg. ``f30!}; Mark 7:4\index{41 Mark:ch.   7, v.   4, pg. ``f30!}.} 
\bigskip
28.4\ \ Not only those that do actually profess faith\index{Faith:profession of ``f10!} in and obedience\index{God's law:commandments:obedience:required ``f10!} unto Christ,\footnote{\ Mark 16:15\index{41 Mark:ch.  16, v.  15, pg. ``f30!}{}-16\index{41 Mark:ch.  16, v.  15-16, pg. ``f30!}; Acts 8:37-38\index{44 Acts:ch.   8, v.  37-38, pg. ``f30!}.} but also the infants of one\index{Infants:of believing parents, to be baptized ``f10!} or both believing\index{Baptism:infants of believing parents ``f10!} parents, are to be baptized\index{Baptism:infants ``f10!}.\footnote{\ Genesis 17:7\index{01 Genesis:ch.  17, v.   7, pg. ``f30!}, 9\index{01 Genesis:ch.  17, v.   7, 9, pg. ``f30!}{}-10\index{01 Genesis:ch.  17, v.   7, 9-10, pg. ``f30!} with Galatians 3:9, 14\index{48 Galatians:ch.   3, v.   9, 14, pg. ``f30!}; Colossians 2:11-12\index{51 Colossians:ch.   2, v.  11-12, pg. ``f30!}; Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}{}-39\index{44 Acts:ch.   2, v.  38-39, pg. ``f30!} and Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}{}-12\index{45 Romans:ch.   4, v.  11-12, pg. ``f30!}; 1 Corinthians 7:14\index{46 1 Corinthians:ch.   7, v.  14, pg. ``f30!}; Matthew 28:19\index{40 Matthew:ch.  28, v.  19, pg. ``f30!}; Mark 10:13-16\index{41 Mark:ch.  10, v.  13-16, pg. ``f30!}; Luke 18:1\index{42 Luke:ch.  18, v.   1, pg. ``f30!}5\index{42 Luke:ch.  18, v.  15, pg. ``f30!}.}   
\bigskip
28.5\ \ Although it be a great sin to contemn\index{Baptism:not to be treated with contempt ``f10!} or neglect this\index{Baptism:not to neglect ``f10!} ordinance,\footnote{\ Luke 7:30\index{42 Luke:ch.   7, v.  30, pg. ``f30!} with Exodus 4:24-26\index{02 Exodus:ch.   4, v.  24-26, pg. ``f30!}.} yet grace\index{Grace:exhibited in sacraments ``f10!} and salvation\index{Salvation:belief in Jesus by faith ``f10!} are not so inseparably\index{Sanctification:inseparable from:grace ``f10!} annexed unto it, as that no person can be regenerated\index{Regeneration:regeneration ``f10!} or saved without\index{Salvation:not without grace ``f10!} it;\footnote{\ Romans 4:11\index{45 Romans:ch.   4, v.  11, pg. ``f30!}; Acts 10:2\index{44 Acts:ch.  10, v.   2, pg. ``f30!}, 4, 22, 31, 45, 47\index{44 Acts:ch.  10, v.   2, 4, 22, 31, 45, 47, pg. ``f30!}.} or, that all that are baptized\index{Baptism:not all baptized are regenerated ``f10!} are undoubtedly regenerated.\footnote{\ Acts 8:13, 23\index{44 Acts:ch.   8, v.  13, 23, pg. ``f30!}.}   
\bigskip
28.6\ \ The efficacy\index{Baptism:spiritual efficacy ``f10!} of Baptism is not tied to that moment of\index{Baptism:efficacy not time dependent ``f10!} time wherein it is administered;\footnote{\ John 3:5\index{43 John:ch.   3, v.   5, pg. ``f30!}, 8\index{43 John:ch.   3, v.   5, 8, pg. ``f30!}.} yet notwithstanding, by the right use of this ordinance\index{Baptism:ordinance, to be used rightly ``f10!}, the grace promised is\index{Baptism:means of grace ``f10!} not only offered, but really exhibited and\index{Baptism:grace:exhibited ``f10!} conferred, by\index{Baptism:grace:conferred by Holy Spirit ``f10!} the Holy Ghost\index{Holy Spirit:Holy Ghost ``f10!}, to such (whether of age or infants) as\index{Infants:of believing parents, to be baptized ``f10!} that grace\index{Grace:in believers:by Holy Spirit ``f10!} belongs unto, according to the counsel of God's\index{Causation:God's own will ``f10!} own will, in His appointed time.\footnote{\ Galatians 3:27\index{48 Galatians:ch.   3, v.  27, pg. ``f30!}; Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}; Ephesians 5:2\index{49 Ephesians:ch.   5, v.   2, pg. ``f30!}5-26\index{49 Ephesians:ch.   5, v.  25-26, pg. ``f30!}; Acts 2:38\index{44 Acts:ch.   2, v.  38, pg. ``f30!}, 41\index{44 Acts:ch.   2, v.  38, 41, pg. ``f30!}.}   
\bigskip
28.7\ \ The sacrament\index{Sacraments:holy:ordinances ``f10!} of Baptism is but\index{Baptism:to be administered once ``f10!} once to be administered unto any person.\footnote{\ Titus 3:5\index{56 Titus:ch.   3, v.   5, pg. ``f30!}.}  

\section{Chapter 29 -- Of the Lord's Supper} \index{Lord's Supper:sacrament ``f10!}29.1\ \ Our Lord Jesus, in the night wherein He was betrayed\index{Jesus:betrayed ``f10!}, instituted the\index{Sacraments:instituted by:Christ ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} of His body and blood\index{Lord's Supper:body and blood of Jesus ``f10!}, called the Lord's Supper\index{Lord's Supper:sacrament ``f10!}, to be observed in\index{Lord's Supper:to be observed to end of world ``f10!} His Church\index{Church:Christ's ``f10!}, unto the end of the world, for the perpetual remembrance\index{Lord's Supper:remembrance of Christ's sacrifice ``f10!} of the sacrifice of Himself\index{Jesus:sacrifice:of himself ``f10!} in His death\index{Jesus:death:death ``f10!}; the sealing\index{Lord's Supper:benefits sealed ``f10!} all benefits thereof unto\index{Lord's Supper:benefits believers receive ``f10!} true believers, their spiritual nourishment\index{Lord's Supper:spiritual feeding:nourishment ``f10!} and growth in Him\index{Sanctification:spiritual growth:in Christ ``f10!}, their further engagement in and\index{God's law:commandments:obedience:required ``f10!} to all duties which\index{God's law:commandments:duties:required by God ``f10!} they owe unto Him; and to be a bond and pledge\index{Lord's Supper:pledge of communion with Christ ``f10!} of their communion with Him, and\index{Communion:with Christ ``f10!} with each other, as members of His\index{Church members ``f10!} mystical\index{Believers:union with Christ:mystical ``f10!} body.\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  23-26, pg. ``f30!}; 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}{}-17\index{46 1 Corinthians:ch.  10, v.  16-17, pg. ``f30!}, 21\index{46 1 Corinthians:ch.  10, v.  16-17, 21, pg. ``f30!}; 1 Corinthians 12:13\index{46 1 Corinthians:ch.  12, v.  13, pg. ``f30!}.}   
\bigskip
29.2\ \ In this sacrament\index{Sacraments:holy:ordinances ``f10!}, Christ is not offered up to\index{Lord's Supper:not a sacrifice:of Jesus ``f10!} His Father\index{Father:God the ``f10!}; nor any real sacrifice made at all for remission\index{Redemption:remission of sins ``f10!} of sins of the quick or dead\index{Lord's Supper:not a sacrifice:for quick or dead ``f10!};\footnote{\ Hebrews 9:22\index{58 Hebrews:ch.   9, v.  22, pg. ``f30!}, 25-26, 28\index{58 Hebrews:ch.   9, v.  22, 25-26, 28, pg. ``f30!}.} but only a commemoration\index{Lord's Supper:commemoration ``f10!} of that one offering up of Himself\index{Jesus:sacrifice:of himself ``f10!}, by Himself, upon the cross\index{Jesus:death:on cross ``f10!}, once for all: and a spiritual oblation\index{Lord's Supper:remembrance of Christ's sacrifice ``f10!} of all possible praise unto\index{Worship:praising God ``f10!} God for the same:\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  24-26, pg. ``f30!}; Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-27\index{40 Matthew:ch.  26, v.  26-27, pg. ``f30!}.} so that the Popish\index{Popish rituals ``f10!} sacrifice of the\index{Lord's Supper:not a sacrifice:of the mass ``f10!} mass\index{Mass:Roman Catholic ``f10!} (as they call it) is most abominably injurious to Christ's one, only sacrifice, the\index{Jesus:sacrifice:once only ``f10!} alone propitiation\index{Jesus:sacrifice:propitiation ``f10!} for all the sins of His elect\index{Election:elect of God ``f10!}.\footnote{\ Hebrews 7:23-24, 27\index{58 Hebrews:ch.   7, v.  23-24, 27, pg. ``f30!}; Hebrews 10:1\index{58 Hebrews:ch.  10, v.   1, pg. ``f30!}1-12, 14, 18\index{58 Hebrews:ch.  10, v.  11-12, 14, 18, pg. ``f30!}.}   
\bigskip
29.3\ \ The Lord Jesus has, in this ordinance, appointed His ministers\index{Church officers:minsters ``f10!} to declare His\index{Preachers:declare word of institution at Lord's Supper ``f10!} word of institution\index{Sacraments:instituted by:Christ ``f10!} to the people; to pray, and bless the\index{Lord's Supper:elements:Jesus blessed ``f10!} elements\index{Lord's Supper:elements:bread and wine ``f10!} of bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}, and thereby to set them apart from a common to a holy\index{Lord's Supper:elements:holy use ``f10!} use; and to take and break the\index{Lord's Supper:bread broken ``f10!} bread, to take the cup\index{Lord's Supper:cup ``f10!}, and (they communicating\index{Lord's Supper:communion ``f10!} also themselves) to give both to the communicants\index{Lord's Supper:both bread and wine to communicants ``f10!};\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg. ``f30!} and Mark 14:22-24\index{41 Mark:ch.  14, v.  22-24, pg. ``f30!} and Luke 22:19\index{42 Luke:ch.  22, v.  19, pg. ``f30!}{}-20\index{42 Luke:ch.  22, v.  19-20, pg. ``f30!} with 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}3\index{46 1 Corinthians:ch.  11, v.  23, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  23-26, pg. ``f30!}.} but to none who are not then present in the congregation\index{Lord's Supper:only to those present in congregation ``f10!}.\footnote{\ Acts. 20:7; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}0\index{46 1 Corinthians:ch.  11, v.  20, pg. ``f30!}.} 
\bigskip
29.4\ \ Private\index{Mass:private ``f10!} masses\index{Mass:Roman Catholic ``f10!}, or receiving this\index{Lord's Supper:not private ``f10!} sacrament\index{Sacraments:holy:ordinances ``f10!} by a priest or\index{Lord's Supper:not consecrated by a priest ``f10!} any other alone;\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}.} as likewise, the denial\index{Lord's Supper:both bread and wine to communicants ``f10!} of the cup\index{Lord's Supper:cup ``f10!} to the people,\footnote{\ Mark 14:23\index{41 Mark:ch.  14, v.  23, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}5\index{46 1 Corinthians:ch.  11, v.  25, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  25-29, pg. ``f30!}.} worshipping the elements\index{Lord's Supper:elements:bread and wine ``f10!}, the lifting\index{Lord's Supper:elements:not to be lifted for adoration ``f10!} them up or carrying them about for adoration\index{Idolatry:adoration of eucharistic elements ``f10!}, and the reserving them for any pretended religious use\index{Lord's Supper:elements:not to be used incorrectly ``f10!}; are all contrary to the nature of this sacrament, and to the institution\index{Sacraments:instituted by:Christ ``f10!} of Christ.\footnote{\ Matthew 15:9\index{40 Matthew:ch.  15, v.   9, pg. ``f30!}.}   
\bigskip
29.5\ \ The outward elements\index{Lord's Supper:elements:bread and wine ``f10!} in this sacrament\index{Sacraments:holy:ordinances ``f10!}, duly set apart to the uses ordained by\index{Sacraments:ordained by Jesus ``f10!} Christ, have such relation to Him\index{Lord's Supper:elements:relation to Christ crucified ``f10!} crucified\index{Jesus:crucified ``f10!}, as that, truly, yet sacramentally\index{Lord's Supper:elements:union between sign and signified ``f10!} only, they are sometimes called by the name of the things they represent, to\index{Lord's Supper:elements:represent Christ ``f10!} wit, the body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of Christ;\footnote{\ Matthew 26:26\index{40 Matthew:ch.  26, v.  26, pg. ``f30!}{}-28\index{40 Matthew:ch.  26, v.  26-28, pg. ``f30!}.} albeit in substance and nature they still remain truly\index{Lord's Supper:elements:remain truly and only natural ``f10!} and only bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}, as they were before.\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}6\index{46 1 Corinthians:ch.  11, v.  26, pg. ``f30!}{}-28\index{46 1 Corinthians:ch.  11, v.  26-28, pg. ``f30!}; Matthew 26:29\index{40 Matthew:ch.  26, v.  29, pg. ``f30!}.}   
\bigskip
29.6\ \ That doctrine\index{Doctrine:false:transubstantiation ``f10!} which maintains\index{Lord's Supper:elements:transubstantiation:repugnant ``f10!} a change of the substance of bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!} into the substance of Christ's body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} (commonly called transubstantiation\index{Lord's Supper:elements:transubstantiation:superstition ``f10!}) by consecration\index{Lord's Supper:elements:not changed by consecration of priest ``f10!} of a priest, or by\index{Lord's Supper:not consecrated by a priest ``f10!} any other way, is repugnant, not to Scripture alone, but even to common sense and reason; overthrows the nature of the sacrament\index{Sacraments:holy:ordinances ``f10!}, and has been, and is the cause of manifold superstitions\index{Superstitions:superstitions ``f10!}; yea, of gross idolatries\index{Idolatry:idolatry ``f10!}.\footnote{\ Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!} with 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}4\index{46 1 Corinthians:ch.  11, v.  24, pg. ``f30!}{}-26\index{46 1 Corinthians:ch.  11, v.  24-26, pg. ``f30!}; Luke 24:6, 39\index{42 Luke:ch.  24, v.   6, 39, pg. ``f30!}.}   
\bigskip
29.7\ \ Worthy\index{Lord's Supper:spiritual feeding:worthy receivers ``f10!} receivers outwardly partaking\index{Lord's Supper:elements:outward partaking ``f10!} of the visible elements\index{Lord's Supper:elements:bread and wine ``f10!} in this sacrament\index{Sacraments:holy:ordinances ``f10!},\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}8\index{46 1 Corinthians:ch.  11, v.  28, pg. ``f30!}.} do then also, inwardly by faith, really and indeed, yet not carnally\index{Lord's Supper:spiritual feeding:not carnal ``f10!} and corporally\index{Lord's Supper:spiritual feeding:not corporal ``f10!}, but spiritually, receive and feed\index{Lord's Supper:spiritual feeding:in faith ``f10!} upon Christ crucified\index{Jesus:crucified ``f10!}, and all benefits of\index{Lord's Supper:benefits believers receive ``f10!} His death\index{Jesus:death:death ``f10!}: the body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of Christ being then, not corporally or carnally, in, with, or under the bread and\index{Lord's Supper:bread ``f10!} wine\index{Lord's Supper:wine ``f10!}; yet, as really, but spiritually, present to the faith\index{Faith:profession of ``f10!} of believers in that ordinance, as the elements themselves are to their outward\index{Lord's Supper:elements:outward senses ``f10!} senses.\footnote{\ 1 Corinthians 10:16\index{46 1 Corinthians:ch.  10, v.  16, pg. ``f30!}.}   
\bigskip
29.8\ \ Although ignorant\index{ignorance ``f10!} and wicked men receive\index{Lord's Supper:wicked do not receive thing signified ``f10!} the outward elements\index{Lord's Supper:elements:bread and wine ``f10!} in this sacrament\index{Sacraments:holy:ordinances ``f10!}: yet they receive not the thing signified thereby, but by their unworthy coming\index{Lord's Supper:unworthy participation ``f10!} thereunto are guilty\index{guilt ``f10!} of the body and blood\index{Lord's Supper:body and blood of Jesus ``f10!} of the Lord to their own damnation\index{Punishment:everlasting:damnation ``f10!}. Wherefore, all ignorant and ungodly\index{Unbelievers:ungodly ``f10!} persons, as they are unfit\index{Lord's Supper:ungodly not to be admitted ``f10!} to enjoy communion with Him, so are they unworthy of the Lord's table; and cannot, without great sin against Christ while they remain such, partake of these holy\index{Lord's Supper:elements:holy mysteries ``f10!} mysteries,\index{Lord's Supper:mystery of ``f10!}\footnote{\ 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v.  27, pg. ``f30!}{}-29\index{46 1 Corinthians:ch.  11, v.  27-29, pg. ``f30!}; 2 Corinthians 6:14\index{47 2 Corinthians:ch.   6, v.  14, pg. ``f30!}{}-16\index{47 2 Corinthians:ch.   6, v.  14-16, pg. ``f30!}.} or be admitted thereunto.\footnote{\ 1 Corinthians 5:6-7\index{46 1 Corinthians:ch.   5, v.   6-7, pg. ``f30!}, 13\index{46 1 Corinthians:ch.   5, v.   6-7, 13, pg. ``f30!}; 2 Thessalonians 3:6, 14-15\index{53 2 Thessalonians:ch.   3, v.   6, 14-15, pg. ``f30!}; Matthew 7:6\index{40 Matthew:ch.   7, v.   6, pg. ``f30!}.}  

\section{Chapter 30 -- Of Church Censures} \index{Church discipline:censures ``f10!}30.1\ \ The Lord Jesus, as King\index{Jesus:King ``f10!} and Head\index{Church:Christ as head ``f10!} of His Church\index{Church:Christ's ``f10!}, has therein appointed a government, in\index{Church government:government ``f10!} the hand of Church\index{Church:institution ``f10!} officers\index{Church officers:officers ``f10!}, distinct from\index{Church government:distinct from civil ``f10!} the civil magistrate.\footnote{\ Isaiah 9:6\index{23 Isaiah:ch.   9, v.   6, pg. ``f30!}{}-7\index{23 Isaiah:ch.   9, v.   6-7, pg. ``f30!}; 1 Timothy 5:1\index{54 1 Timothy:ch.   5, v.   1, pg. ``f30!}7\index{54 1 Timothy:ch.   5, v.  17, pg. ``f30!}; 1 Thessalonians 5:12\index{52 1 Thessalonians:ch.   5, v.  12, pg. ``f30!}; Acts 20:17, 28\index{44 Acts:ch.  20, v.  17, 28, pg. ``f30!}; Hebrews 13:7\index{58 Hebrews:ch.  13, v.   7, pg. ``f30!}, 17, 24\index{58 Hebrews:ch.  13, v.   7, 17, 24, pg. ``f30!}; 1 Corinthians 12:28\index{46 1 Corinthians:ch.  12, v.  28, pg. ``f30!}; Matthew 28:18\index{40 Matthew:ch.  28, v.  18, pg. ``f30!}{}-20\index{40 Matthew:ch.  28, v.  18-20, pg. ``f30!}.}   
\bigskip
30.2\ \ To these officers\index{Church officers:officers ``f10!} the keys\index{Kingdom:keys of ``f10!} of the kingdom of heaven\index{Kingdom:of heaven ``f10!} are committed: by virtue whereof, they have power\index{Church:power delegated by Christ ``f10!} respectively to retain, and remit\index{Church officers:power to retain or remit sins ``f10!} sins; to shut that kingdom against the impenitent\index{Punishment:of impenitent ``f10!}, both by the Word and\index{Scripture:word of God:standard for:life ``f10!} censures; and\index{Church discipline:censures ``f10!} to open it unto penitent\index{Repentance:penitence ``f10!} sinners, by the ministry of the Gospel\index{Gospel:gospel ``f10!}, and by absolution\index{Church discipline:absolution ``f10!} from censures, as occasion shall require.\footnote{\ Matthew 16:19\index{40 Matthew:ch.  16, v.  19, pg. ``f30!}; Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}{}-18\index{40 Matthew:ch.  18, v.  17-18, pg. ``f30!}; John 20:21-23\index{43 John:ch.  20, v.  21-23, pg. ``f30!}; 2 Corinthians 2:6-8\index{47 2 Corinthians:ch.   2, v.   6-8, pg. ``f30!}.}   
\bigskip
30.3\ \ Church censures are\index{Church discipline:censures ``f10!} necessary, for the reclaiming\index{Church discipline:for reclaiming offending brothers ``f10!} and gaining of offending brethren, for\index{Church discipline:for gaining offending brothers ``f10!} deterring\index{Church discipline:deterrence from sin ``f10!} of others from the like offences, for purging\index{Church discipline:purging leaven ``f10!} out of that leaven\index{leaven, spiritual ``f10!} which might infect the whole lump, for vindicating the honour of\index{Church discipline:vindicate honour of Christ ``f10!} Christ, and the holy\index{Preaching:holy profession of Gospel ``f10!} profession of the Gospel\index{Gospel:gospel ``f10!}, and for preventing the wrath of God\index{Wrath:of God:for sin ``f10!}, which might justly fall\index{Punishment:just:on Church for failure of discipline ``f10!} upon the Church, if\index{Church:institution ``f10!} they should suffer His covenant and\index{Covenant:new:seals ``f10!} the seals\index{Sacraments:covenant seals ``f10!} thereof to be profaned by\index{Covenant:not to be profaned ``f10!} notorious\index{Church discipline:notorious offenders ``f10!} and obstinate\index{Church discipline:obstinate offenders ``f10!} offenders.\footnote{\ 1 Corinthians chap. 5; 1 Timothy 5:20\index{54 1 Timothy:ch.   5, v.  20, pg. ``f30!}; Matthew 7:6\index{40 Matthew:ch.   7, v.   6, pg. ``f30!}; 1 Timothy 1:20\index{54 1 Timothy:ch.   1, v.  20, pg. ``f30!}; 1 Corinthians 11:2\index{46 1 Corinthians:ch.  11, v.   2, pg. ``f30!}7\index{46 1 Corinthians:ch.  11, v.  27, pg. ``f30!} to the end, with Jude 23\index{65 Jude:v.  23, pg. ``f30!}.}   
\bigskip
30.4\ \ For the better attaining of these ends, the officers\index{Church officers:officers ``f10!} of the Church are\index{Church:institution ``f10!} to proceed by admonition\index{Church discipline:admonition ``f10!}; suspension\index{Church discipline:suspension from Lord's Supper ``f10!} from the sacrament\index{Sacraments:holy:ordinances ``f10!} of the Lord's Supper\index{Lord's Supper:sacrament ``f10!} for a season; and\index{Lord's Supper:suspension for a season ``f10!} by excommunication\index{Church discipline:excommunication ``f10!} from the Church; according to the nature of the crime\index{Church discipline:consistent with:nature of crime ``f10!}, and demerit\index{Church discipline:consistent with:demerit of person ``f10!} of the person.\footnote{\ 1 Thessalonians 5:12\index{52 1 Thessalonians:ch.   5, v.  12, pg. ``f30!}; 2 Thessalonians 3:6, 14-15\index{53 2 Thessalonians:ch.   3, v.   6, 14-15, pg. ``f30!}; 1 Corinthians 5:4-5\index{46 1 Corinthians:ch.   5, v.   4-5, pg. ``f30!}, 13\index{46 1 Corinthians:ch.   5, v.   4-5, 13, pg. ``f30!}; Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}; Titus 3:1\index{56 Titus:ch.   3, v.   1, pg. ``f30!}0\index{56 Titus:ch.   3, v.  10, pg. ``f30!}.}  

\section{Chapter 31 -- Of Synods and Councils} \index{Church government:synods ``f10!}\index{Church government:councils ``f10!}31.1\ \ For the better government, and\index{Church government:government ``f10!} further edification of\index{Church councils:edification of Church ``f10!} the Church, there\index{Church:institution ``f10!} ought to be such assemblies\index{Church councils:assemblies to be held ``f10!} as are commonly called synods\index{Church government:synods ``f10!} or councils\index{Church government:councils ``f10!}.\footnote{\ Acts 15:2\index{44 Acts:ch.  15, v.   2, pg. ``f30!}, 4, 6\index{44 Acts:ch.  15, v.   2, 4, 6, pg. ``f30!}.}   
\bigskip
31.2\ \ As magistrates may lawfully call\index{Civil magistrate:to call synods to consult and advise ``f10!} a synod of ministers\index{Church officers:minsters ``f10!}, and other fit persons, to consult and advise with, about matters of religion\index{Civil magistrate:to call synods to consult about matters of religion ``f10!};\footnote{\ Isaiah 49:23\index{23 Isaiah:ch.  49, v.  23, pg. ``f30!}; 1 Timothy 2:1\index{54 1 Timothy:ch.   2, v.   1, pg. ``f30!}{}-2\index{54 1 Timothy:ch.   2, v.   1-2, pg. ``f30!}; 2 Chronicles 19:8-11\index{14 2 Chronicles:ch.  19, v.   8-11, pg. ``f30!}; 2 Chronicles 29; 2 Chronicles 30; Matthew 2:4-5\index{40 Matthew:ch.   2, v.   4-5, pg. ``f30!}; Proverbs 11:1\index{20 Proverbs:ch.  11, v.   1, pg. ``f30!}4\index{20 Proverbs:ch.  11, v.  14, pg. ``f30!}.} so, if magistrates be\index{Civil magistrate:legal authority even if:unbelievers ``f10!} open enemies\index{Enemies:of the Church ``f10!} to the Church, the\index{Church:institution ``f10!} ministers of Christ of themselves, by virtue of their office\index{Church officers:officers ``f10!}, or they, with other fit persons upon delegation\index{Church government:delegate representatives to meet with government ``f10!} from their Churches, may meet together in such assemblies\index{Church councils:assemblies to be held ``f10!}.\footnote{\ Acts 15:2\index{44 Acts:ch.  15, v.   2, pg. ``f30!}, 4, 22-23, 25\index{44 Acts:ch.  15, v.   2, 4, 22-23, 25, pg. ``f30!}.}\textsuperscript{,} \footnote{\ \textcolor{black}{Refer to the }\textit{\textcolor{black}{Statement Concerning Subscription }}\textcolor{black}{(}\textit{\textcolor{black}{Subscription Exceptions}}\textcolor{black}{).}}  
\bigskip
31.3\ \ It belongs to synods\index{Church government:synods ``f10!} and councils\index{Church government:councils ``f10!}, ministerially to determine controversies\index{Church councils:determine controversies of faith ``f10!} of faith and cases of conscience\index{Conscience:cases of, in controversies ``f10!}; to set down rules and\index{Church councils:set rules for public worship ``f10!} directions for the better ordering of the\index{Church councils:directions for ordering public worship ``f10!} public worship\index{Worship:public ``f10!} of God, and government of\index{Church government:government ``f10!} his Church\index{Church:Christ's ``f10!}; to receive complaints\index{Church discipline:synods to receive complaints ``f10!} in cases of maladministration\index{Church discipline:synods to address maladministration ``f10!}, and authoritatively to determine the same: which decrees\index{Church councils:decrees ``f10!} and determinations\index{Church councils:Biblical determinations to be received ``f10!}, if consonant to the Word of God\index{Scripture:word of God:word of God ``f10!}, are to be received with reverence and\index{Church councils:submission to ``f10!} submission; not only for their agreement with the Word, but\index{Scripture:word of God:standard:for Church ``f10!} also for the power\index{Church:power delegated by Christ ``f10!} whereby they are made, as being an ordinance of\index{Church:furnished with:ordinances of God ``f10!} God appointed thereunto in His Word.\footnote{\ Acts 15:15\index{44 Acts:ch.  15, v.  15, pg. ``f30!}, 19, 24, 27-31\index{44 Acts:ch.  15, v.  15, 19, 24, 27-31, pg. ``f30!}; Acts 16:4\index{44 Acts:ch.  16, v.   4, pg. ``f30!}; Matthew 18:17\index{40 Matthew:ch.  18, v.  17, pg. ``f30!}{}-20\index{40 Matthew:ch.  18, v.  17-20, pg. ``f30!}.}   
\bigskip
31.4\ \ All synods\index{Church government:synods ``f10!} or councils\index{Church government:councils ``f10!}, since the Apostles\index{Apostles:apostles ``f10!}{}' times, whether general or particular, may err\index{Church councils:may err ``f10!}; and many have erred\index{Church councils:many have erred ``f10!}. Therefore they are not to be made the rule of faith\index{Faith:rule of ``f10!}, or practice; but to be used as a help in\index{Church councils:not the rule of faith or practice, but help ``f10!} both.\footnote{\ Ephesians 2:2\index{49 Ephesians:ch.   2, v.   2, pg. ``f30!}0\index{49 Ephesians:ch.   2, v.  20, pg. ``f30!}; Acts 17:11\index{44 Acts:ch.  17, v.  11, pg. ``f30!}; 1 Corinthians 2:5\index{46 1 Corinthians:ch.   2, v.   5, pg. ``f30!}; 2 Corinthians 1:24\index{47 2 Corinthians:ch.   1, v.  24, pg. ``f30!}.}   
\bigskip
31.5\ \ Synods\index{Church government:synods ``f10!} and councils\index{Church government:councils ``f10!} are to handle, or conclude, nothing, but that which is ecclesiastical\index{Church councils:to address only ecclesiastical matters ``f10!}: and are not to intermeddle with civil affairs which\index{Church councils:not to meddle with civil affairs ``f10!} concern the commonwealth\index{commonwealth ``f10!}; unless by way of humble\index{Church councils:to receive humble petitions ``f10!} petition, in cases extraordinary; or by way of advice, for satisfaction of conscience, if\index{Conscience:satisfaction of ``f10!} they be thereunto required by the civil\index{Civil magistrate:to call synods to consult and advise ``f10!} magistrate.\footnote{\ Luke 12:13-14\index{42 Luke:ch.  12, v.  13-14, pg. ``f30!}; John 18:36\index{43 John:ch.  18, v.  36, pg. ``f30!}.}  

\section{Chapter 32 -- Of the State of Man After Death, and of the Resurrection of the Dead} \index{Death:following:state of man ``f10!}\index{Resurrection:general, of dead ``f10!}32.1\ \ The bodies of men, after death\index{Death:bodies:return to dust ``f10!}, return to dust\index{Adam:formed of dust ``f10!} and see corruption:\index{Death:corruption of body ``f10!}\footnote{\ Genesis 3:19\index{01 Genesis:ch.   3, v.  19, pg. ``f30!}; Acts 13:3\index{44 Acts:ch.  13, v.   3, pg. ``f30!}6\index{44 Acts:ch.  13, v.  36, pg. ``f30!}.} but their souls (which neither die nor\index{Mankind:immortal soul ``f10!} sleep) having an immortal subsistence, immediately return to God who gave them:\index{Believers:souls made perfect at death ``f10!}\footnote{\ Luke 23:43\index{42 Luke:ch.  23, v.  43, pg. ``f30!}; Ecclesiastes 12:7\index{21 Ecclesiastes:ch.  12, v.   7, pg. ``f30!}.} the souls of the righteous, being then\index{Resurrection:believers:perfected ``f10!} made perfect in holiness\index{Believers:holy:made perfect at death ``f10!}, are received into the highest heavens, where\index{Heaven:God's abode ``f10!} they behold\index{Believers:behold face of God in glory ``f10!} the face of\index{Resurrection:believers:behold face of God ``f10!} God, in light and glory\index{Resurrection:believers:into glory:light ``f10!}, waiting for\index{Resurrection:believers:waiting expectantly ``f10!} the full redemption\index{Redemption:full ``f10!} of their bodies.\index{Bodies in graves:believers ``f10!}\footnote{\ Hebrews 12:2\index{58 Hebrews:ch.  12, v.   2, pg. ``f30!}3\index{58 Hebrews:ch.  12, v.  23, pg. ``f30!}; 2 Corinthians 5:1, 6, 8\index{47 2 Corinthians:ch.   5, v.   1, 6, 8, pg. ``f30!}; Philippians 1:23\index{50 Philippians:ch.   1, v.  23, pg. ``f30!} with Acts 3:21\index{44 Acts:ch.   3, v.  21, pg. ``f30!} and Ephesians 4:10\index{49 Ephesians:ch.   4, v.  10, pg. ``f30!}.} And the souls of the wicked are\index{Judgement day:wicked:cast into:hell ``f10!} cast into\index{Punishment:hell:wicked cast into ``f10!} hell\index{Hell:hell ``f10!}, where they remain in torments and\index{Punishment:everlasting:torments (body and soul) ``f10!} utter darkness\index{darkness, spiritual ``f10!}, reserved to the judgment of the\index{Judgement:final ``f10!} great day.\footnote{\ Luke 16:23-24\index{42 Luke:ch.  16, v.  23-24, pg. ``f30!}; Acts 1:25\index{44 Acts:ch.   1, v.  25, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg. ``f30!}{}-7\index{65 Jude:v. 6-7, pg. ``f30!}; Peter 3:19.} Beside these two places, for soul\index{Superiors:to:provide inferiors with things necessary for soul and body ``f10!}s separated\index{Death:soul separated from body ``f10!} from their bodies, the\index{Bodies in graves:believers and unbelievers ``f10!} Scripture acknowledges\index{Last day:heaven or hell, no other option ``f10!} none.   
\bigskip
32.2\ \ At the last day\index{Last day:last day ``f10!}, such as are found alive\index{Last day:those alive on earth ``f10!} shall not die, but\index{Last day:not all shall die ``f10!} be changed:\footnote{\ 1 Thessalonians 4:17\index{52 1 Thessalonians:ch.   4, v.  17, pg. ``f30!}; 1 Corinthians 15:51-52\index{46 1 Corinthians:ch.  15, v.  51-52, pg. ``f30!}.} and all the dead\index{Death:following:all men raised ``f10!} shall be raised up, with the selfsame bodies, and none other, although with different qualities\index{Resurrection:bodies:different qualities ``f10!}, which shall be united again\index{Resurrection:bodies:united to souls ``f10!} to their souls for ever.\footnote{\ Job 19:26-27\index{18 Job:ch.  19, v.  26-27, pg. ``f30!}; 1 Corinthians 15:4\index{46 1 Corinthians:ch.  15, v.   4, pg. ``f30!}2-44\index{46 1 Corinthians:ch.  15, v.  42-44, pg. ``f30!}.}   
\bigskip
32.3\ \ The bodies of the\index{Resurrection:bodies:same ones as before death, for just and unjust ``f10!} unjust shall, by the power\index{Resurrection:believers:by the power of Christ ``f10!} of Christ, be raised to dishonour\index{Resurrection:wicked raised in dishonour ``f10!}: the bodies of the just, by His Spirit, unto\index{Holy Spirit:raises bodies of just to honour ``f10!} honour; and\index{Resurrection:believers:raised in honour ``f10!} be made conformable\index{Resurrection:bodies:conformable to Christ's ``f10!} to His own glorious body\index{Jesus:resurrection:body:glorious ``f10!}.\footnote{\ Acts 24:15\index{44 Acts:ch.  24, v.  15, pg. ``f30!}; John 5:28-29\index{43 John:ch.   5, v.  28-29, pg. ``f30!}; 1 Corinthians 15:4\index{46 1 Corinthians:ch.  15, v.   4, pg. ``f30!}3\index{46 1 Corinthians:ch.  15, v.  43, pg. ``f30!}; Philippians 3:21\index{50 Philippians:ch.   3, v.  21, pg. ``f30!}.}  

\section{Chapter 33 -- Of the Last Judgment} \index{Judgement:final ``f10!}33.1\ \ God has appointed a day, wherein He will judge the\index{Jesus:judge:world ``f10!} world in righteousness\index{Judgement day:in righteousness ``f10!}, by Jesus Christ,\footnote{\ Acts 17:31\index{44 Acts:ch.  17, v.  31, pg. ``f30!}.} to whom all power\index{Jesus:almighty power ``f10!} and judgment is given\index{Jesus:judge:assigned by Father ``f10!} of the Father\index{Father:God the ``f10!}.\footnote{\ John 5:22, 27\index{43 John:ch.   5, v.  22, 27, pg. ``f30!}.} In which day, not only the apostate angels\index{Angels:angels ``f10!} shall be judged\index{Angels:apostate, to be judged ``f10!},\footnote{\ 1 Corinthians 6:3\index{46 1 Corinthians:ch.   6, v.   3, pg. ``f30!}; Jude 6\index{65 Jude:v.   6, pg. ``f30!}; 2 Peter 2:4\index{61 2 Peter:ch.   2, v.   4, pg. ``f30!}.} but likewise all persons that have lived upon earth shall appear before the tribunal\index{Judgement day:tribunal of Christ ``f10!} of Christ, to give an account of their thoughts, words, and deeds\index{Judgement day:account for thoughts, words, and deeds ``f10!}; and to receive according to what they have done in the body, whether good or evil\index{Judgement day:give account of deeds, whether good or evil ``f10!}.\footnote{\ 2 Corinthians 5:10\index{47 2 Corinthians:ch.   5, v.  10, pg. ``f30!}; Ecclesiastes 12:1\index{21 Ecclesiastes:ch.  12, v.   1, pg. ``f30!}4\index{21 Ecclesiastes:ch.  12, v.  14, pg. ``f30!}; Romans 2:1\index{45 Romans:ch.   2, v.   1, pg. ``f30!}6\index{45 Romans:ch.   2, v.  16, pg. ``f30!}; Romans 14:10, 12\index{45 Romans:ch.  14, v.  10, 12, pg. ``f30!}; Matthew 12:36\index{40 Matthew:ch.  12, v.  36, pg. ``f30!}{}-37\index{40 Matthew:ch.  12, v.  36-37, pg. ``f30!}.}   
\bigskip
33.2\ \ The end of God's appointing this day is for the manifestation of the glory of His\index{God's attributes:glory:his own ``f10!} mercy, in\index{God's attributes:merciful ``f10!} the eternal salvation\index{Salvation:belief in Jesus by faith ``f10!} of the elect\index{Election:elect of God ``f10!}; and of His justice\index{God's attributes:administers justice ``f10!}, in the damnation\index{Punishment:everlasting:damnation ``f10!} of the reprobate\index{Punishment:of reprobate ``f10!} who are wicked and\index{Judgement day:wicked:damned ``f10!} disobedient\index{Judgement day:disobedient cast into hell ``f10!}. For then shall the righteous go\index{Resurrection:believers:to everlasting life ``f10!} into everlasting life\index{everlasting life ``f10!}, and receive that fulness of joy\index{Joy:everlasting ``f10!} and refreshing\index{Judgement day:believers:joy and refreshing ``f10!}, which shall come from the presence of the Lord: but the wicked who\index{Judgement day:wicked:cast into:eternal torments ``f10!} know not God, and obey not\index{Punishment:those who do not obey the Gospel ``f10!} the Gospel\index{Gospel:gospel ``f10!} of Jesus Christ, shall be cast into\index{Punishment:hell:wicked cast into ``f10!} eternal torments, and\index{Punishment:everlasting:torments (body and soul) ``f10!} be punished with everlasting destruction from\index{Punishment:everlasting:destruction ``f10!} the presence of the Lord, and from the glory of His power.\footnote{\ Matthew 25:31\index{40 Matthew:ch.  25, v.  31, pg. ``f30!} to the end, Romans 2:5\index{45 Romans:ch.   2, v.   5, pg. ``f30!}{}-6\index{45 Romans:ch.   2, v.   5-6, pg. ``f30!}; Romans 9:22-23\index{45 Romans:ch.   9, v.  22-23, pg. ``f30!}; Matthew 25:21\index{40 Matthew:ch.  25, v.  21, pg. ``f30!}; Acts 3:19\index{44 Acts:ch.   3, v.  19, pg. ``f30!}; 2 Thessalonians 1:7\index{53 2 Thessalonians:ch.   1, v.   7, pg. ``f30!}{}-10\index{53 2 Thessalonians:ch.   1, v.   7-10, pg. ``f30!}.}   
\bigskip
33.3\ \ As Christ would have us to be certainly persuaded that\index{Judgement day:be persuaded that there is ``f10!} there shall be a day of judgment\index{Judgement day:believers:perfected ``f10!}, both to deter\index{Judgement day:knowledge as deterrence from sin ``f10!} all men from sin, and for the greater consolation of the godly in their adversity\index{Judgement day:encourage godly in adversity ``f10!};\footnote{\ 2 Peter 3:11, 14\index{61 2 Peter:ch.   3, v.  11, 14, pg. ``f30!}; 2 Corinthians 5:10\index{47 2 Corinthians:ch.   5, v.  10, pg. ``f30!}{}-11\index{47 2 Corinthians:ch.   5, v.  10-11, pg. ``f30!}; 2 Thessalonians 1:5-7\index{53 2 Thessalonians:ch.   1, v.   5-7, pg. ``f30!}; Luke 21:27-28\index{42 Luke:ch.  21, v.  27-28, pg. ``f30!}; Romans 8:2\index{45 Romans:ch.   8, v.   2, pg. ``f30!}3\index{45 Romans:ch.   8, v.  23, pg. ``f30!}{}-25\index{45 Romans:ch.   8, v.  23-25, pg. ``f30!}.} so will He have that day unknown\index{Judgement day:day unknown to men ``f10!} to men, that they may shake off all carnal security\index{Idolatry:carnal:security ``f10!}, and be always watchful, because\index{Jesus:return:watchfulness ``f10!} they know not at what hour the\index{Jesus:return:no one knows hour ``f10!} Lord will come; and may be ever prepared\index{Last day:be prepared to say, 'Come, Lord Jesus' ``f10!} to say, Come, Lord Jesus, come quickly, Amen\index{Amen, testimony of our true desire for God's glory ``f10!}.\footnote{\ Matthew 24:36\index{40 Matthew:ch.  24, v.  36, pg. ``f30!}, 42-44\index{40 Matthew:ch.  24, v.  36, 42-44, pg. ``f30!}; Mark 13:35-37\index{41 Mark:ch.  13, v.  35-37, pg. ``f30!}; Luke 12:35-36\index{42 Luke:ch.  12, v.  35-36, pg. ``f30!}; Revelation 22:20\index{66 Revelation:ch.  22, v.  20, pg. ``f30!}.}
